### 2019-09-23 notes/analysis_notes/analysis_notes.tex
init

\documentclass{article}

\usepackage{geometry}  % Set margins
	\geometry{left=2.2cm,
		right=2.2cm,
		top=2.2cm,
		bottom=2cm}
\parskip 0.15cm
\setlength{\parindent}{0cm}

\usepackage{graphics}
\usepackage{graphicx}
\graphicspath{ {img/} }
\usepackage{float}  % Fancy graphics placement [H] [H!] arguments

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{textcomp}
\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}}

\usepackage{listings}
\usepackage{xcolor}

\definecolor{codegreen}{rgb}{0,0.6,0}
\definecolor{codegray}{rgb}{0.5,0.5,0.5}
\definecolor{codepurple}{rgb}{0.58,0,0.82}
\definecolor{backcolour}{rgb}{0.95,0.95,0.92}

\lstdefinestyle{mystyle}{
    backgroundcolor=\color{backcolour},
    commentstyle=\color{codegreen},
    keywordstyle=\color{magenta},
    numberstyle=\tiny\color{codegray},
    stringstyle=\color{codepurple},
    basicstyle=\ttfamily\footnotesize,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}
\lstset{style=mystyle}

\usepackage{cite}
\usepackage{natbib}
\bibliographystyle{agsmnourl}  % Use my custom agsm bibliography template which never includes URLs in articles
\usepackage{url}

\renewcommand{\refname}{\vspace{-1cm}}

\title{\vspace{-1cm}Progress in analyses - Regional Biodiversity - Biomass relationship}
\date{}

\begin{document}

\maketitle{}

\section{Hypotheses and research question}

\begin{itemize}
	\item{Higher species richness will result in higher biomass stocks due to niche complementarity.}
	\item{Species composition will have more effect on biomass than species richness \textit{per se} due to selection effects.}
	\item{Increased aridity will result in a strong richness - biomass relationship}
	\item{There will be a positive effect of abundance evenness oon biomass stocks.}
\end{itemize}

\section{Data description}

\begin{itemize}
	\item{Above ground biomass}
		\begin{itemize}
			\item{SEOSAW} 
			\item{DBH, height, species, wood density, \verb|BIOMASS| package}
			\item{\citet{SEOSAW}}
		\end{itemize}
	\item{Species composition of stems}
		\begin{itemize}
			\item{SEOSAW}
			\item{Species richness, species abundance}
		\end{itemize}
	\item{Aridity}
		\begin{itemize}
			\item{Global Aridity Index (\verb|Global-Aridity_ET0|)}
			\item{Calculated from Worldclim data}
			\item{\citet{Trabucco2009}}
		\end{itemize}
	\item{Fire frequency}
		\begin{itemize}
			\item{MODIS fire return interval}
			\item{\citet{}}
		\end{itemize}
	\item{Stand structural complexity}
		\begin{itemize}
			\item{SEOSAW}
			\item{DBH coef. var.}
			\item{Height coef. var.}
		\end{itemize}
\end{itemize}

\section{Method}

\subsection{Data cleaning}

\begin{enumerate}
	\item{Aggregate Zambia plots}
	\item{Remove stems \textless 5 cm DBH, \textless 1 m height}
	\item{Remove plots with \textless 10 stems ha\textsuperscript{-1}}
	\item{Remove plots with a small area}
	\item{Remove plots that have:}
		\begin{itemize}
			\item{Wood harvesting}
			\item{Farming}
			\item{Livestock grazing}
			\item{Experimental treatments}
		\end{itemize}
	\item{Remove plots which are composition outliers}
\end{enumerate}

\subsection{Data analysis}

Data was log or cube tranformed where appropriate to achieve a normal distribution. All data was standardized to allow direct comparison of effect sizes within path analyses of the Structural Equation Models (SEMs).

Linear models of each bivariate relationship present in the full SEM were analysed to specify the direction of paths.

A full SEM was specified, after which variables were incrementally removed in order to assess which variables were redundant in explaining variation in Biomass

\end{document}

% Trabucco, A., and Zomer, R.J. 2009. Global Aridity Index (Global-Aridity) and Global Potential Evapo-Transpiration (Global-PET) Geospatial Database. CGIAR Consortium for Spatial Information. Published online, available from the CGIAR-CSI GeoPortal at: http://www.csi.cgiar.org/

### 2019-09-23 manuscript/chapter_1.tex
Main commit

\documentclass[12pt,english,a4paper]{report}

\usepackage{natbib}
\usepackage{amsmath}

\setlength{\textheight}{24cm}
\setlength{\headsep}{0cm}

\usepackage{textcomp}
\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}}

\newcommand{\dir}{chapter}

\begin{document}
\title{A regional scale assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
\author{John L. Godlee}
%\date{
%  \today\\
%  \vspace{2cm}
%  {\centering\footnotesize\begin{tabular}[c]{c}
%      \epsfig{file=../preface/2Line2ColCMYK_CS3.eps,height=3cm}\\
%      School of GeoSciences\\
%      Grant Institute\\
%      West Mains Road\\
%      Edinburgh, EH9 3JW\\
%      Scotland
%    \end{tabular}}
%  }
\maketitle
\tableofcontents
%\clearpage
%\listoffigures
%\listoftables
%\listofalgorithms
%\newpage
%\bibliography{}

\section{Introduction}

A global observed decline in biodiversity may lead to a decrease in ecosystem functioning and health, with knock-on effects for the benefits humans derive from ecosystems as ecosystem services \citep{}. Carbon storage as biomass within ecosystems is a commonly studied ecosystem function, as is Gross Primary Productivity (GPP). The storage of atmospheric CO\textsubscript{2} as biomass buffers increases in atmospheric CO\textsubscript{2} caused by human actions. Concentration of biomass in vegetation and soils also indirectly benefits other ecosystem functions, being correlated with ecosystem stability and resilience to disturbance \citep{}. Understanding how ecosystem structure and diversity affects biomass storage is important in order to predict how changes to communities may impact the global carbon balance, with potential impacts for future climate change.

Savannas and sparse woodlands are the dominant vegetation type across southern Africa, spanning \textgreater{}4 million km\textsuperscript{2} across the region. Climatic conditions and biogeography vary across the region, resulting in a diverse range of woodland tree species compositions, which retain the common features of a non-closed canopy and an understorey generalyl dominated by C4 grass species. Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are \textgreater{}300 tree species \citep{Frost1996}.

Unlike temperate and wet tropical forests, where ecosystem assemblage is often determined by competition between tree species in an equilibrium state, southern African woodlands are highly structured by disturbance in the form of fire and herbivory, which limit the growth of trees which would otherwise form a closed canopy forest \citep{}. At the dry end of the woodland gradient, woodlands are limited by low precipitation, which allows C4 grasses to compete with tree seedlings.

Other forces acting upon an ecosystem to influence its functionality include climatic factors and the frequency and intensity of disturbances.

Precipitation acts as a key determinant of woodland structure

In southern African woodlands, disturbance as a result of seasonal fires .... Precipitation interacts with disturbance to increase the prevalence of fires by adding grass fuel load and increasing the probability of a fire occurring, and the intensity of the fire when a fire occurs.

Grimes, mass ratio, selection effects and complementarity effects

In this study, we made the first known attempt to make a regional estimation of the Biodiversity Ecosystem Function Relationship in southern African woodlands. We used aboveground woody biomass and estimates of ecosystem productivity as measures of ecosystem functionality to understand the co-relationship between tree species biodiversity and ecosystem functionality. We compared the relative effects of tree species biodiversity with that of other environmental factors known to affect ecosystem productivity and biomass accumulation such as precipitation and temperature. In acknowledgement of the wide variation in biogeographically determined community composition across the region, we also used previously determined biogeographical clusters \citep{} to understand how species composition as well as species biodiversity \textit{per se} affected ecosystem functionality. Initially, we made four hypotheses: 1) across the region, after other environmental factors had been accounted for, woodland plots with a higher species richness would have a higher aboveground biomass, 2) increased aridity would increase the strength of the biomass tree species richness relationship, 3) species composition would have a greater affect on aboveground biomass than species diversity \textit{per se}. We used Structural Equation Modelling as a preferred method to simultaneously account for environmental factors and biogeographic factors, which often interact in their effect on ecosystem structure and therefore biomass.

\subsection{Hypotheses}

We hypothesised that plots with a higher species richness will maintain higher biomass stocks.

Species composition will influence above-ground productivity more than species richness \textit{per se}.

Plots in more arid regions will feature a stronger positive effect of tree species richness on above ground biomass stocks, due to abiotic facilitation effects, despite lower overall tree species richness.

Structural characteristics of the woodland will interact with species composition and richness to provide an indirect path of influecne between species composition and biomass stocks.

\section{Methods}

All analyses were performed in R version 3.6.0

Plot data were collated from a network of plots.

Plots exist along a temperature and precipitation gradient (Figure ).

In each of these plots, each living tree stem \textgreater{}5 cm Diameter at Breast Height (DBH) was measured. For each tree, species, DBH and height to the highest branch material were recorded.

Plots varied in size across the region.

Rarefied species richness accounted for variation in plot size and therefore sampling effort across the region.

To ensure that all plots represented true woodland, plots with fewer than 20 stems per Ha were discarded.

Structural Equation Models were compared using different combinations of variables. Path analysis investigated the

Species richness and the Shannon index (Shannon \& Weaver 1949) (Equation 1) were calculated to assess species diversity.

\begin{equation}
	H = -\sum{}P_i \ln{P_i} / \ln{S} 
\end{equation}

\section{Results}

\section{Discussion}

\end{document}

### 2019-09-23 analysis/aridity_index_seosaw.R

@@ -1,0 +1,54 @@
+# Get Aridity Index values for SEOSAW plots
+# John Godlee (johngodlee@gmail.com)
+# 2018_12_10
+# 2019_09_17
+
+# Preamble ----
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(rgdal)
+library(raster)
+library(maps)
+library(ggplot2)
+library(dplyr)
+
+# Import data ----
+
+# Aridity Index
+ai <- readGDAL("data/ai_et0_crop/ai_et0_crop.tif")
+
+# SEOSAW plot locations
+load("data/seosaw_plot_summary5Apr2019.Rdata")
+
+# Convert plot locations to spatial ----
+plot_xy <- data.frame(ssaw8$plotInfoFull$longitude_of_centre, 
+	ssaw8$plotInfoFull$latitude_of_centre)
+
+plot_loc_spdf <- SpatialPointsDataFrame(coords = plot_xy, 
+  data = as.data.frame(ssaw8$plotInfoFull),
+	proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
+
+# Convert aridity index to raster ----
+ai_raster <- raster(ai)
+
+# Get aridity index for each plot ----
+ai_seosaw <- extract(ai_raster, plot_loc_spdf)
+
+ssaw8$plotInfoFull$ai <- ai_seosaw
+
+# Export data ----
+
+# Select relevant columns
+seosaw_ai <- ssaw8$plotInfoFull %>%
+	dplyr::select(plotcode, longitude_of_centre, latitude_of_centre, ai)
+
+# Export
+write.csv(seosaw_ai, file = "data/seosaw_aridity_index.csv", row.names = FALSE)
+

### 2019-09-23 analysis/bivar_rel_norm.R

@@ -1,0 +1,523 @@
+# Bivariate relationships between variables, transform, normalise and standardize the data
+# John Godlee (johngodlee@gmail.com)
+# 2018_12_10
+
+# Preamble ----
+
+# Source files
+# source("sem_data_collation.R")
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(psych)  # pairs.panels()
+library(gridExtra)  # grid.arrange()
+library(lme4)
+library(MuMIn)
+
+
+# Functions
+# A ggplot2 theme
+theme.bivar <- function(base_size = 12, base_font_family = "Helvetica"){
+  theme_bw(base_size = base_size, base_family = base_font_family) %+replace%
+    theme(
+      axis.ticks = element_line(colour = "black", 
+        size = rel(1)),
+      axis.text = element_text(angle = 0,
+        margin = margin(t = rel(1)),
+        size = rel(0.8)),
+      axis.title = element_text(size = rel(1), 
+        face = "plain"),
+      axis.title.x = element_text(angle = 0,
+        margin = margin(t = rel(1.2))),
+      axis.title.y = element_text(angle = 90, 
+        margin = margin(r = rel(1.2))),
+      
+      plot.margin = unit(rep(0.04, times = 4), 
+        units = "native"),
+      plot.title = element_text(size = rel(1.8), 
+        margin = margin(b = rel(12)),
+        hjust = rel(0.05)),
+      legend.text = element_text(size = rel(1), 
+        face = "italic"),
+      legend.title = element_text(size = rel(1.5), 
+        vjust = rel(0.8)),
+      legend.position = "bottom",
+      
+      panel.grid.major = element_line(colour="black", 
+        size = rel(0.2)),
+      panel.grid.minor = element_blank(),
+      panel.background = element_blank(),
[…]

### 2019-09-23 analysis/clust_pal.R

@@ -1,0 +1,2 @@
+# Define a 5 point colour palette for vegetation clusters
+clust_pal <- c("#5cb930", "#e6c332", "#6e49cc", "#49b9da", "#da2c57")

### 2019-09-23 analysis/compositional_outliers.R

@@ -1,0 +1,148 @@
+# Identifying plots with tree species compositional outliers so they can be removed from the dataset 
+# John Godlee (johngodlee@gmail.com)
+# 2019_07_11
+
+# Preamble ----
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+library(dave)
+
+# Functions
+
+# Import data ----
+plot_data <- read.csv("data/plot_data_fil_agg_norm_std.csv")
+plotcode_plot_group_lookup <- read.csv("data/plotcode_plot_group_lookup.csv")
+load("data/clean_input_data.Rdata")
+
+# Filter stem data to only include suitable plots and living stems
+##; Suitable plots idetified in data_collation.R
+s_fil <- s %>%
+  left_join(., plotcode_plot_group_lookup, by = c("plotcode" = "plotcode_vec")) %>%
+  inner_join(., data.frame(plot_group = plot_data$plot_group), 
+    by = c("plot_group" = "plot_group")) %>%
+  filter(alive %in% c("A", NA),
+    !is.na(gen_sp))
+
+# Create community composition matrix
+stem_ab_mat <- s_fil %>%
+  group_by(plot_group, gen_sp, .drop = FALSE) %>%
+	tally() %>%
+	spread(gen_sp, n, fill = 0) %>%
+  ungroup() %>%
+  data.frame() 
+
+# Set rownames
+rownames(stem_ab_mat) <- stem_ab_mat$plot_group
+
+stem_ab_mat <- select(stem_ab_mat, -plot_group)
+
+stem_ab_mat_dist <- vegdist(stem_ab_mat,  method = "bray")
+
+# Define function NMDS.scree() 
+##' performs a NMDS for 1-10 dimensions,
+##' plots number of dimensions vs stress
+NMDS.scree <- function(x) { #where x is the name of the data frame variable
+  plot(rep(1, 10), 
+    replicate(10, metaMDS(x, autotransform = F, k = 1)$stress), 
+    xlim = c(1, 10),ylim = c(0, 0.30), 
+    xlab = "# of Dimensions", ylab = "Stress", 
[…]

### 2019-09-23 analysis/lmm.R

@@ -1,0 +1,181 @@
+# Linear Mixed Effects models predicting biomass to determine optimal model structure for SEM
+# John Godlee (johngodlee@gmail.com)
+# 2019_07_11
+
+# Preamble ----
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(lme4)
+library(colortools)
+
+# Cluster colour palette
+source("clust_pal.R")
+
+# Import data ----
+
+# Aggregated and filtered data
+plot_data <- read.csv("data/plot_data_fil_agg_norm_std.csv")
+
+# Make cluster categorical
+plot_data$clust5 <- as.character(plot_data$clust5)
+
+# Remove plots not in a cluster
+plot_data <- plot_data %>%
+  filter(!is.na(clust5))
+
+# Plot all variables against bchave_log in a facet wrap ----
+
+# Climatic variables
+plot_data_gather_clim <- plot_data %>%
+  select(clust5, bchave_log, aridity_index, mean_temp, temp_seasonality, 
+    fire_return_mean_log, firecount_2001_2018,
+    ocdens, sand_per, cation_ex_cap) %>%
+  gather(key = "variable", value = "value", 3:9) %>%
+  mutate(facet_label = factor(variable, 
+    levels = c("aridity_index", "mean_temp", "temp_seasonality", 
+      "fire_return_mean_log", "firecount_2001_2018", "ocdens", "sand_per"),
+    labels = c(
+      expression("Aridity" ~ "index"),
+      expression("Mean" ~ "annual" ~ "temp." ~ (degree * C)), 
+      expression("Temp." ~ "seasonality" ~ (degree * C)),
+      expression("log(Mean" ~ "fire" ~ "return" ~ "interval)" ~ (yr)),
+      expression("Number" ~ "of" ~ "fires" ~ "2001-18"),
+      expression("Organic" ~ "C" ~ "%"),
+      expression("Sand" ~ "%"))))
+
+
+pdf(file = "img/biomass_clim_lm_clust.pdf", width = 12, height = 10)
+ggplot(plot_data_gather_clim, aes(x = value, y = bchave_log)) + 
[…]

### 2019-09-23 analysis/plot_data_collation.R

@@ -1,0 +1,311 @@
+# Creating a dataset for use in Chapter 1 - Regional analysis of the factors affecting woody AGB and productivity
+# John Godlee (johngodlee@gmail.com)
+# 2018_12_10
+
+# Preamble ----
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(dplyr)
+library(ggplot2)
+library(tidyr)
+library(reshape2)
+library(vegan)
+library(gridExtra)
+
+# Import data ----
+
+# Base SEOSAW plot summary dataset
+load("data/seosaw_plot_summary5Apr2019.Rdata")
+##' AGB
+##' Shannon / Sp. R
+##' Stem density
+##' Soil nutrients (CEC)
+##' Plot location
+
+# Per stem data 
+load("data/clean_input_data.Rdata")
+##' Coefficient of variation height
+##' Coefficient of variation DBH
+
+# Aridity data 
+aridity_index <- read.csv("data/seosaw_aridity_index.csv")
+##' Aridity index from WorldClim calculations
+
+# Soil organic carbon
+ocdens <- read.csv("data/seosaw_ocdens.csv")
+
+# MODIS fire data 
+fire_return <- read.csv("data/fire_return.csv")
+##' Data copied from burn /exports/...
+
+# Clean data ----
+
+# Create large raw dataframe
+plot_data <- ssaw8$struct %>%
+  left_join(., ssaw8$structPerHa, by = c("plotcode" = "plotcode")) %>%
+  left_join(., ssaw8$plotInfoFull, by = c("plotcode" = "plotcode")) %>%
+  left_join(., ssaw8$floristicsGT5, by = c("plotcode" = "plotcode")) %>%
+  left_join(., ssaw8$soil, by = c("plotcode" = "plotcode")) %>%
+  left_join(., ssaw8$climate, by = c("plotcode" = "plotcode")) %>%
+  left_join(., ssaw8$env, by = c("plotcode" = "plotcode")) %>%
+  left_join(., ssaw8$cluster, by = c("plotcode" = "plotcode")) %>%
+  left_join(., aridity_index, by = c("plotcode" = "plotcode")) %>%
+  left_join(., ocdens, by = c("plotcode" = "plotcode")) %>%
[…]

### 2019-09-23 analysis/productivity_calculations/acm_luke/ACM_GPP_ET_single_site_test.r

@@ -1,0 +1,281 @@
+###
+## Define useful functions
+###
+
+closest2d <- function (id,lat,long,lat_in,long_in,nos_dim) {
+
+    ###
+    ## function to return x,y coordinate from an array which is nearest to a provided lat / long value
+    ###
+  
+    ###
+    ## Define arguments
+  
+    # id = location from the lat_in and long_in vector you wish to find (can be equal to 1 if *_in are scalers)
+    # lat = array / vector of latitudes (-90 -> 90) to search for nearest location
+    # long = array / vector of longitudes (-180 -> 180) to search for nearest location
+    # lat_in = scaler / vector of latitudes for locations you wish to find
+    # long_in = scaler / vector of longitudes for locations you wish to find
+    # nos_dim = defines the way in which the searchable lat / long information has been provided. 
+    #           See code below for definitions
+
+    # extract needed lat / long 
+    lat1=lat_in[id] ; long1=long_in[id]
+
+    # calculate the distance between two points by Spherical law of the cosine
+    # mean radius of earth in km
+    R=6371  # 6378137 m (R source equitorial)
+    # convert degrees to radians
+    deg_to_rad = pi/180
+    
+    # check lat / long system has been provided in the correct bounds
+    if (length(which(as.vector(long) > 180)) > 1) {stop("Input error closest2d: longitude should be -180 to +180")}
+
+    if (nos_dim == 1) {
+
+      	## lat long are in single vectors repeating i.e. lat[1:10]=89.9,89.9,89.9... ; long[1:10]=-180,-160,-140....
+
+      	# loop through to find the smallest distance
+      	d_old=1e6
+      	# check for locations which exactly coincide, complete calculatuion and finally remove NaN generated by value "1"
+      	d = acos(sin(lat1*deg_to_rad)*sin(lat*deg_to_rad)+cos(lat1*deg_to_rad)*cos(lat*deg_to_rad)*cos((long*deg_to_rad)-(long1*deg_to_rad)))*R
+      	match = which(is.na(d) == TRUE) ; d[match] = 0 ; d = pmax(1e-6,d,na.rm=TRUE)
+      	output=which(d == min(d))[1] ; d_old=d[output] ; rm(d_old)
+      	
+    } else if (nos_dim == 2) {
+
[…]

### 2019-09-23 analysis/productivity_calculations/acm_luke/src/ACM_GPP_ET_R_interface.f90

@@ -1,0 +1,92 @@
+
+subroutine racmgppet(output_dim,met,pars,out_var,lat  &
+                    ,nopars,nomet,nofluxes,nopools    &
+                    ,nodays,deltat,nos_iter           &
+                    ,soil_frac_clay_in,soil_frac_sand_in)
+
+  use CARBON_MODEL_MOD, only: CARBON_MODEL, wSWP_time, nos_soil_layers, &
+                              soil_frac_clay,soil_frac_sand
+
+  ! subroutine specificially deals with the calling of the fortran code model by
+  ! R
+
+  implicit none
+
+  ! declare input variables
+  integer, intent(in) :: nopars         & ! number of paremeters in vector
+                        ,output_dim     & !
+                        ,nos_iter       & !
+                        ,nomet          & ! number of meteorological fields
+                        ,nofluxes       & ! number of model fluxes
+                        ,nopools        & ! number of model pools
+                        ,nodays           ! number of days in simulation
+
+  double precision, intent(in) :: met(nomet,nodays)   & ! met drivers, note reverse of needed
+                       ,pars(nopars,nos_iter)         & ! number of parameters
+                       ,soil_frac_clay_in(nos_soil_layers) & ! clay in soil (%)
+                       ,soil_frac_sand_in(nos_soil_layers) & ! sand in soil (%)
+                       ,lat                 ! site latitude (degrees)
+
+  double precision, intent(inout) :: deltat(nodays) ! time step in decimal days
+
+  ! output declaration
+  double precision, intent(out), dimension(nos_iter,nodays,output_dim) :: out_var
+
+  ! local variables
+  integer i
+  ! vector of ecosystem pools
+  double precision, dimension((nodays+1),nopools) :: POOLS
+  ! vector of ecosystem fluxes
+  double precision, dimension(nodays,nofluxes) :: FLUXES
+
+  ! zero initial conditions
+  POOLS = 0d0 ; FLUXES = 0d0
+  out_var = 0d0
+
+  ! update soil parameters
+  soil_frac_clay=soil_frac_clay_in
+  soil_frac_sand=soil_frac_sand_in
[…]

### 2019-09-23 analysis/productivity_calculations/prod_data_collation/ecmwf_extract_array_compile.R

@@ -1,0 +1,147 @@
+# Extract data from ECMWF netCDF files for SEOSAW plots
+# John Godlee (johngodlee@gmail.com)
+# 2018_10_30
+
+# Preamble ----
+
+# Remove old crap
+rm(list=ls())
+
+# Set working directory to the location of the source file
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(dplyr)
+library(ncdf4)
+library(purrr)
+
+# Import netcdf files ----
+# Get file names
+nc_list <- list.files(paste0(getwd(),"/raw_data/ecmwf/ecmwf_nc"), 
+                      pattern="nc$", 
+                      full.names=FALSE)
+
+# Import files directly into list
+nc_file_list <- list()
+
+for(i in nc_list){ 
+  nc_file_list[[i]] <- nc_open(paste0("raw_data/ecmwf/ecmwf_nc/", i))
+} 
+
+# Import SEOSAW plot locations ----
+load("raw_data/seosaw_plot_summary24Sept_v1.Rdata")
+
+plot_loc <- ssaw7$plotInfoFull %>%
+  dplyr::select(plotcode, longitude_of_centre, latitude_of_centre)
+
+# For testing, isolate 3 sets of plot coords ----
+plot_match <- c("ABG-001", "ZLT-009", "MAR-030")
+plot_test <- plot_loc[grep(paste(plot_match, collapse = "|"), plot_loc$plotcode),]
+plot_test$plotcode <- as.character(plot_test$plotcode)
+plot_loc_list <- split(plot_test, seq(nrow(plot_test)))
+
+
+
+# Automate extraction of many variables in a for loop ---- 
+##' Read variable for each timestep across all months for SEOSAW plots
+##' List of dataframes
+##' Each dataframe contains data for one plot
+##' Columns for plotcode, lon, lat, timestep, sol
+
+
+
+
+
+extracted_ecmwf_all <- list()
+for(i in 1:length(plot_loc_list)){
+  all_df <- NULL  # temp df for each plot
+  lon_match <- plot_loc_list[[i]]$longitude_of_centre
+  lat_match <- plot_loc_list[[i]]$latitude_of_centre
+  plot_match <- plot_loc_list[[i]]$plotcode
+  
+  for(j in 1:length(nc_file_list)){
+    # Get grid 
[…]

### 2019-09-23 analysis/productivity_calculations/prod_data_collation/merge_extracted.R

@@ -1,0 +1,68 @@
+# Aggregate all data sources 
+# John Godlee (johngodlee@gmail.com)
+# 2018_10_30
+
+# Preamble ----
+
+# Source these files to rebuild data inputs - 
+ source("lai_extract.R")
+ source("ecmwf_extract.R")
+ source("trmm_extract.R")
+ source("soil_extract.R")
+
+# Remove old crap
+rm(list=ls())
+
+# Set working directory to the location of the source file
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(plyr)
+library(dplyr)
+
+# Load data ----
+
+load("extracted_data/extracted_ecmwf_all_daily.Rdata")
+
+load("extracted_data/extracted_trmm_all.Rdata")
+
+soil <- read.csv("extracted_data/soil_data_plots.csv")
+
+# Subset for testing to three plots
+soil_test <- soil %>%
+  filter(plotcode %in% c("ABG-001", "MAR-030", "ZLT-009"))
+
+load("extracted_data/lai_interp_all.Rdata")
+
+# Make sure all are flat dataframes ----
+
+# ECMWF
+extracted_ecmwf_df <- rbind.fill(extracted_ecmwf_all_daily)
+
+# LAI
+extracted_lai_df <- rbind.fill(predict_list)
+
+# TRMM
+extracted_trmm_df <- rbind.fill(extracted_trmm_all)
+
+
+# Join datasets
+
+all_data <- extracted_ecmwf_df %>%
+  left_join(., extracted_lai_df, by = c("plot_match" = "plotcode", "date" = "date")) %>%
+  left_join(., soil_test, by = c("plot_match" = "plotcode")) %>%
+  left_join(., extracted_trmm_df, by = c("plot_match" = "plot_match", "date" = "date"))
+
+# Clean joined dataset
+str(all_data)
+
+all_data_clean <- all_data %>%
+  dplyr::select(date, plot_match, lon = lon_match.x, lat = lat_match.x, mean_sol, 
+                max_temp, min_temp, total_net_sol, total_net_clear_sol, 
+                precip_ecmwf = total_precip, 
+                lai_predic_loess, lay_predic_spline, CLYPPT_M_sl1_1km_ll, SLTPPT_M_sl1_1km_ll,
+                SNDPPT_M_sl1_1km_ll, ORCDRC_M_sl1_1km_ll, CECSOL_M_sl1_1km_ll, 
[…]

### 2019-09-23 analysis/productivity_calculations/prod_data_collation/trmm_extract.R

@@ -1,0 +1,91 @@
+# Extract data from TRMM precipitation netCDF files for SEOSAW plots
+# John Godlee (johngodlee@gmail.com)
+# 2018_10_30
+
+# Preamble ----
+
+# Remove old crap
+rm(list=ls())
+
+# Set working directory to the location of the source file
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(dplyr)
+library(ncdf4)
+library(stringr)
+
+# Import netcdf files ----
+# Get file names
+nc_list <- list.files(paste0(getwd(),"/raw_data/trmm_nc"), 
+                      pattern="nc4$", 
+                      full.names=FALSE)
+
+# Import files directly into list
+nc_file_list <- list()
+
+from <- match("3B42_Daily.20100101.7.nc4.nc4", nc_list)
+to <- match("3B42_Daily.20171231.7.nc4.nc4", nc_list)
+
+
+# Import SEOSAW plot locations ----
+load("raw_data/seosaw_plot_summary24Sept_v1.Rdata")
+
+plot_loc <- ssaw7$plotInfoFull %>%
+  dplyr::select(plotcode, longitude_of_centre, latitude_of_centre)
+
+# For testing, isolate 3 sets of plot coords ----
+plot_match <- c("ABG-001", "ZLT-009", "MAR-030")
+plot_test <- plot_loc[grep(paste(plot_match, collapse = "|"), plot_loc$plotcode),]
+plot_test$plotcode <- as.character(plot_test$plotcode)
+plot_loc_list <- split(plot_test, seq(nrow(plot_test)))
+
+
+# nc_file <- nc_open(paste0("raw_data/trmm_nc/", nc_list[1]))
+
+extracted_trmm_all <- list()
+for(i in 1:length(plot_loc_list)){
+  all_df <- NULL  # temp df for each plot
+  lon_match <- plot_loc_list[[i]]$longitude_of_centre
+  lat_match <- plot_loc_list[[i]]$latitude_of_centre
+  plot_match <- plot_loc_list[[i]]$plotcode
+  
+  for(j in nc_list[from:to]){ 
+    nc_file <- nc_open(paste0("raw_data/trmm_nc/", j))
+    
+    # Get grid 
+    lon <- ncvar_get(nc_file, "lon", verbose = FALSE)
+    lat <- ncvar_get(nc_file, "lat", verbose = FALSE)
+    filename <- nc_file$filename
[…]

### 2019-09-23 analysis/soil_carbon.R

@@ -1,0 +1,48 @@
+# Extract soil carbon from OCDENS
+# John Godlee (johngodlee@gmail.com)
+# 2019_09_19
+
+# Preamble ----
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages 
+library(rgdal)
+library(raster)
+library(dplyr)
+
+# Organic Carbon 
+ocdens <- readGDAL("data/OCDENS_M_sl2_250m_ll_crop.tif")
+
+# SEOSAW plot locations
+load("data/seosaw_plot_summary5Apr2019.Rdata")
+
+# Convert plot locations to spatial ----
+plot_xy <- data.frame(ssaw8$plotInfoFull$longitude_of_centre, 
+	ssaw8$plotInfoFull$latitude_of_centre)
+
+plot_loc_spdf <- SpatialPointsDataFrame(coords = plot_xy, 
+  data = as.data.frame(ssaw8$plotInfoFull),
+	proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
+
+# Convert aridity index to raster ----
+ocdens_raster <- raster(ocdens)
+
+# Get aridity index for each plot ----
+ocdens_seosaw <- raster::extract(ocdens_raster, plot_loc_spdf)
+
+ssaw8$plotInfoFull$ocdens <- ocdens_seosaw
+
+# Export data ----
+
+# Select relevant columns
+seosaw_ocdens <- ssaw8$plotInfoFull %>%
+	dplyr::select(plotcode, longitude_of_centre, latitude_of_centre, ocdens)
+
+# Export
+write.csv(seosaw_ocdens, file = "data/seosaw_ocdens.csv", row.names = FALSE)

### 2019-09-24 manuscript/chapter_1.tex
Improved productivity calcs

@@
-\documentclass[12pt,english,a4paper]{report}
+\documentclass[11pt,a4paper]{article}
 
-\usepackage{natbib}
-\usepackage{amsmath}
+\usepackage{mynotes}
 
-\setlength{\textheight}{24cm}
-\setlength{\headsep}{0cm}
-
-\usepackage{textcomp}
-\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}}
-
-\newcommand{\dir}{chapter}
+\title{A regional scale assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
+\author{John L. Godlee}
+\date{}
 
 \begin{document}
-\title{A regional scale assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
-\author{John L. Godlee}
-%\date{
-%  \today\\
-%  \vspace{2cm}
-%  {\centering\footnotesize\begin{tabular}[c]{c}
-%      \epsfig{file=../preface/2Line2ColCMYK_CS3.eps,height=3cm}\\
-%      School of GeoSciences\\
-%      Grant Institute\\
-%      West Mains Road\\
-%      Edinburgh, EH9 3JW\\
-%      Scotland
-%    \end{tabular}}
-%  }
+
 \maketitle
 \tableofcontents
-%\clearpage
-%\listoffigures
-%\listoftables
-%\listofalgorithms
-%\newpage
-%\bibliography{}
 
@@
 
-A global observed decline in biodiversity may lead to a decrease in ecosystem functioning and health, with knock-on effects for the benefits humans derive from ecosystems as ecosystem services \citep{}.
+A globally observed decline in biodiversity may lead to a decrease in ecosystem functioning and health, with knock-on effects for the benefits humans derive from ecosystems as ecosystem services \citep{}.
 Carbon storage as biomass within ecosystems is a commonly studied ecosystem function, as is Gross Primary Productivity (GPP).
 The storage of atmospheric CO\textsubscript{2} as biomass buffers increases in atmospheric CO\textsubscript{2} caused by human actions.
-Concentration of biomass in vegetation and soils also indirectly benefits other ecosystem functions, being correlated with ecosystem stability and resilience to disturbance \citep{}.
-Understanding how ecosystem structure and diversity affects biomass storage is important in order to predict how changes to communities may impact the global carbon balance, with potential impacts for future climate change.
+Accumulation of biomass in vegetation and soils indirectly benefits other ecosystem functions, being correlated with ecosystem stability and resilience to disturbance \citep{}.
+Understanding how ecosystem structure and species diversity affects biomass storage is important in order to predict how changes to communities may impact the global carbon balance which will impact future climate change \citep{}.
 
 Savannas and sparse woodlands are the dominant vegetation type across southern Africa, spanning \textgreater{}4 million km\textsuperscript{2} across the region.
-Climatic conditions and biogeography vary across the region, resulting in a diverse range of woodland tree species compositions, which retain the common features of a non-closed canopy and an understorey generalyl dominated by C4 grass species.
+Climatic conditions and biogeography vary across the region, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}.
 Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are \textgreater{}300 tree species \citep{Frost1996}.
+Southern African Woodlands have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}.
+Many conservation projects in the region currently aim to conserve biodiversity and carbon stocks simultaneously under the banner of REDD+ \citep{Hinsley2015}.
+A small number of academic studies have shown that above ground woody carbon stocks correlate positively with tree species richness \citep{McNicol2017} \todo{others}, but all have been hampered by a restricted plot network.
+This study is the first regional analysis of the BEFR in southern African woodlands, using plots which straddle multi-dimensional environmental gradients (Figure \ref{}).
 
-Unlike temperate and wet tropical forests, where ecosystem assemblage is often determined by competition between tree species in an equilibrium state, southern African woodlands are highly structured by disturbance in the form of fire and herbivory, which limit the growth of trees which would otherwise form a closed canopy forest \citep{}.
+The role of environmental factors in shaping the BEFR is poorly understood \citep{Ratcliffe2017}.
+Climatic variation is known to affect woody biomass production \citep{Michaletz2014}, so it is important to control for these factors and acknowledge their interaction with woody species diversity \textit{and} biomass stocks to effectively model the effects of biodiversity on biomass stocks in a regional analysis.
+\citet{Sankaran2005} used data from 854 African field sites to show that mean annual precipitation sets the upper limit for woody cover in these ecosystems, while other factors such as herbivory, fire and soil properties contribute to reducing woody cover below this maximum.
+\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution in Panamanian tropical forests, which may affect ecosystem level productivity and thus biomass stocks through selection effects, which may promote the growth of a certain highly productive species only under certain environmental conditions.
+In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments, suggesting variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}.
+Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential biomass stocks.
+Temperature imposes a similar physiological limit on woody growth rate by limiting metabolic process rates.
+Temperature and water availability together impact the transpiration rate of a tree, with high temperatures and low water availability limiting growth and potentially causing damage or mortality by causing cavitation of vessels within the tree \citep{Rowland2015a, Fensham2009}.
+In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}.
+The effect of extreme climatic conditions may depend on the degree to which the current biota is adapted to it.
+Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem function.
+Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness \textit{per se}.
+
+Unlike temperate and wet tropical forests, where ecosystem assemblage is often determined by competition between tree species in an equilibrium state \citep{}, southern African woodlands are highly structured by disturbance in the form of fire and herbivory, which limit the growth of trees which would otherwise form a closed canopy forest \citep{}.
 At the dry end of the woodland gradient, woodlands are limited by low precipitation, which allows C4 grasses to compete with tree seedlings.
 
-Other forces acting upon an ecosystem to influence its functionality include climatic factors and the frequency and intensity of disturbances.
+An underpinning principle of the Biodiversity-Ecosystem Function Relationship (BEFR) is that of niche complementarity.
+Species are assumed to differ in their functional niche by the fact that they may coexist in a steady-state ecological community, therefore the more species are present the greater the area of the total fundamental niche of the ecosystem is filled, allowing more efficient use of natural resouerces and higher values of productivity and biomass \citep{}.
+This theory however, which has been supported by many experiments and observational studies in temperate and wet tropical ecosystems, may not hold in savannas, which are structured by disturbance rather than competition.
 
 Precipitation acts as a key determinant of woodland structure
-
-In southern African woodlands, disturbance as a result of seasonal fires ....
-Precipitation interacts with disturbance to increase the prevalence of fires by adding grass fuel load and increasing the probability of a fire occurring, and the intensity of the fire when a fire occurs.
 
@@
 
+\subsection{Study location}
+
+\subsection{Data collection}
+
 All analyses were performed in R version 3.6.0
 
-Plot data were collated from a network of plots.
+Plot data were collated from a network of plots across southern Africa, across 12 countries (Figure \ref{plot_map}).
+Plots are managed for various monitoring and scientific purposes by a number of researchersPlots used in the study were chosen from the full plot network of 5385 plots to satisfy a number of conditions.
+Plots were \textgreater{}0.1 ha in area \todo{why}.
+Only stems \textgreater{}5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
+To ensure all plots were within woodland rather than ``grass savanna'', plots with a stem density \textless{}20 stems ha\textsuperscript{-1} were excluded.
+Grass savannas are considered a separate biome according to \citet{Parr2014} \todo{why}.
+Plots which had been farmed with livestock or for arable in the previous 30 years were excluded as were plots which had been used in experimental treatments such as fire or herbivore exclusion.
+Plots created by the Zambian Forestry Commission were originally located in clusters of four adjacent 20x50 m plots.
+Measurements from these plot clusters were aggregated prior to analysis.
+This left 1084 plots in 10 countries.
 
-Plots exist along a temperature and precipitation gradient (Figure ).
+The plots used in this study occurred across a wide cimatic gradient (Figure \ref{env_map}).
 
-In each of these plots, each living tree stem \textgreater{}5 cm Diameter at Breast Height (DBH) was measured. For each tree, species, DBH and height to the highest branch material were recorded.
+All stems \textgreater{}5 cm DBH were measured within each plot.
+For each tree, species, DBH and height to the highest branch material were recorded.
+Height was measured through a variety of means \todo{audit of height methods}
 
-Plots varied in size across the region.
-
+\subsection{Data analysis}
 Rarefied species richness accounted for variation in plot size and therefore sampling effort across the region.
-
-To ensure that all plots represented true woodland, plots with fewer than 20 stems per Ha were discarded.
-
-Structural Equation Models were compared using different combinations of variables. Path analysis investigated the
+Rarefied species richness was calculated using the \verb|vegan| package in R, using a sample size of 20 stems.
 
@@
 
+\subsubsection{Structural Equation Modelling}
+
+Structural Equation Models were compared using different combinations of variables. Path analysis investigated the
+
 \section{Results}

### 2019-09-24 analysis/bivar_rel_norm.R

@@ -17,11 +17,11 @@ rm(list=ls())
 # Packages
 library(dplyr)
 library(tidyr)
-library(ggplot2)
+library(ggplot2)  # ggplot()
 library(psych)  # pairs.panels()
 library(gridExtra)  # grid.arrange()
-library(lme4)
-library(MuMIn)
+library(lme4)  # lmer()
+library(MuMIn)  # r.squared.GLMM()
 
 
 # Functions

### 2019-09-24 analysis/prod_calc/data_collation/ecmwf_extract.R

@@ -16,7 +16,7 @@ library(ncdf4)
 
 # Import netcdf files ----
 # Get file names
-nc_list <- list.files(paste0(getwd(),"/raw_data/ecmwf/ecmwf_nc"), 
+nc_list <- list.files(paste0(getwd(),"../raw_data/era5"), 
                        pattern="nc$", 
                        full.names=FALSE)
 
@@ -38,8 +38,6 @@ plot_match <- c("ABG-001", "ZLT-009", "MAR-030")
 plot_test <- plot_loc[grep(paste(plot_match, collapse = "|"), plot_loc$plotcode),]
 plot_test$plotcode <- as.character(plot_test$plotcode)
 plot_loc_list <- split(plot_test, seq(nrow(plot_test)))
-
-
 
 # Automate extraction of sol in a for loop ---- 
 ##' Read sol for each timestep across all months for SEOSAW plots

### 2019-09-24 analysis/prod_calc/data_collation/grb_to_nc.sh

@@ -1,0 +1,4 @@
+#!/bin/bash
+
+cdo -f nc4 copy ../raw_data/era5.grb ../raw_data/era5.nc
+

### 2019-09-24 analysis/prod_calc/data_collation/soil_extract.R

@@ -12,7 +12,7 @@ setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 
 # Load data ----
 
-load("raw_data/seosaw_plot_summary24Sept_v1.Rdata")
+load("../seosaw_plot_summary5Apr_v1.Rdata")
 
 # Filter data to soil properties ----
 soil <- data.frame(plotcode = ssaw7$soil$plotcode, 
@@ -25,4 +25,4 @@ soil <- data.frame(plotcode = ssaw7$soil$plotcode,
                    CECSOL_M_sl1_1km_ll = ssaw7$soil$CECSOL_M_sl1_1km_ll)
 
 # Write data ----
-write.csv(soil, "extracted_data/soil_data_plots.csv", row.names = FALSE)
+write.csv(soil, "../extracted_data/soil_data_plots.csv", row.names = FALSE)

### 2019-09-24 analysis/productivity_calculations/prod_data_collation/lai_extract.R

@@ -1,144 +1,0 @@
-# Extract LAI pixel values from copernicus tif files
-# John Godlee (johngodlee@gmail.com)
-# 2018_10_16
-
-# Preamble ----
-
-# Remove old crap
-rm(list=ls())
-
-# Set working directory to the location of the source file
-setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
-
-# Packages
-library(raster)
-library(ggplot2)
-library(plyr)
-library(dplyr)
-
-# Import tif files ----
-# Get file names
-tif_list <- list.files(paste0(getwd(),"/raw_data/copernicus_tif"), 
-                       pattern="tif$", 
-                       full.names=FALSE)
-
-# Import files directly into list
-tif_file_list <- list()
-
-for(i in tif_list){ 
-  tif_file_list[[i]] <- raster(paste0("raw_data/copernicus_tif/", i))
-} 
-
-# Import SEOSAW plot locations ----
-load("raw_data/seosaw_plot_summary24Sept_v1.Rdata")
-
-plot_loc <- ssaw7$plotInfoFull %>%
-  dplyr::select(plotcode, longitude_of_centre, latitude_of_centre)
-
-# For testing, isolate 3 sets of plot coords ----
-plot_match <- c("ABG-001", "ZLT-009", "MAR-030")
-plot_test <- plot_loc[grep(paste(plot_match, collapse = "|"), plot_loc$plotcode),]
-plot_test$plotcode <- as.character(plot_test$plotcode)
-plot_loc_list <- split(plot_test, seq(nrow(plot_test)))
-
-# Extract LAI from all tif files for plot locations ----
-
-# Create an empty data frame which will hold all the data
-extracted_data_plot <- data.frame(plotcode = character(0), 
-                                 long = double(0), 
-                                 lat = double(0), 
-                                 pixel_value = integer(0),
-                                 date = as.Date(character(0)), stringsAsFactors = FALSE)
-
-# Create an empty object to rbind results of each loop run to
-extracted_data_all <- NULL
-
-# Nested for loop to extract LAI from each tif (j) for each plot (i)
-for(i in 1:length(plot_loc_list)){
-  for(j in 1:length(tif_file_list)){
-    
-    extracted_data_plot[j,1] <- plot_loc_list[[i]]$plotcode
-    extracted_data_plot[j,2] <- plot_loc_list[[i]]$longitude_of_centre
-    extracted_data_plot[j,3] <- plot_loc_list[[i]]$latitude_of_centre
[…]

### 2019-10-17 manuscript/chapter_1.tex
New figures

--- 2019-09-23 manuscript/chapter_1.tex
@@
 
-A global observed decline in biodiversity may lead to a decrease in ecosystem functioning and health, with knock-on effects for the benefits humans derive from ecosystems as ecosystem services \citep{}.
-Carbon storage as biomass within ecosystems is a commonly studied ecosystem function, as is Gross Primary Productivity (GPP).
-The storage of atmospheric CO\textsubscript{2} as biomass buffers increases in atmospheric CO\textsubscript{2} caused by human actions.
-Concentration of biomass in vegetation and soils also indirectly benefits other ecosystem functions, being correlated with ecosystem stability and resilience to disturbance \citep{}.
-Understanding how ecosystem structure and diversity affects biomass storage is important in order to predict how changes to communities may impact the global carbon balance, with potential impacts for future climate change.
+A number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Review}.
+The strength and direction of these observed relationships varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{}.
+Over the past decade, many observational studies of the Biodiversity-Ecosystem Function Relationship (BEFR) have been conducted in tropical and temperate forests, and grasslands \citep{Chen2011}.
+The representation of dry tropical ecosystems in the BEFR literature however, is poor.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in Costa Rica.
 
-Savannas and sparse woodlands are the dominant vegetation type across southern Africa, spanning \textgreater{}4 million km\textsuperscript{2} across the region.
-Climatic conditions and biogeography vary across the region, resulting in a diverse range of woodland tree species compositions, which retain the common features of a non-closed canopy and an understorey generalyl dominated by C4 grass species.
+In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity.
+Resource extraction directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}.
+Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation dynamics and affecting climate seasonality \citep{}.
+Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning \textgreater{}4 million km\textsuperscript{2}.
+The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and are of global importance to the carbon cycle.
+Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}.
 Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are \textgreater{}300 tree species \citep{Frost1996}.
+They have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}.
+Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the banner of REDD+ \citep{Hinsley2015}.
+A small number of academic studies in these woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2017, and, others}, but all have been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends cannot be reliably inferred.
+\citet{Fayolle2018}
 
-Unlike temperate and wet tropical forests, where ecosystem assemblage is often determined by competition between tree species in an equilibrium state, southern African woodlands are highly structured by disturbance in the form of fire and herbivory, which limit the growth of trees which would otherwise form a closed canopy forest \citep{}.
-At the dry end of the woodland gradient, woodlands are limited by low precipitation, which allows C4 grasses to compete with tree seedlings.
+The role of environmental factors in shaping the BEFR generally is poorly understood \citep{Ratcliffe2017}.
+Climatic variation is known to affect woody biomass \citep{Michaletz2014}, so it is important to account for these factors and acknowledge their interaction with woody species diversity \textit{and} biomass stocks to effectively model the effects of biodiversity on biomass stocks in a regional analysis.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in these ecosystems, which is presumably positively correlated with biomass, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
+However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.
 
-Other forces acting upon an ecosystem to influence its functionality include climatic factors and the frequency and intensity of disturbances.
+\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution in Panamanian tropical forests, which may affect ecosystem level productivity and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under certain environmental conditions.
+In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments, suggesting variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}.
+Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}.
+Temperature imposes a similar physiological limit on woody growth rate by limiting metabolic rates.
+Temperature and water availability together impact the transpiration rate of a tree, with high temperatures and low water availability limiting growth and potentially causing damage or mortality through cavitation of vessels within the tree \citep{Rowland2015a, Fensham2009}.
+In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}, so this effect may not be present in historically droughted areas.
+The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it.
+Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem functionality via two effects, the direct effect of resource availability in the form of moisture availability and soil nutrients, and indirectly through the effect of these environmental factors on species composition.
+Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness \textit{per se}.
 
-Precipitation acts as a key determinant of woodland structure
+\todo{Species abundance evenness and selection effects.}
 
-In southern African woodlands, disturbance as a result of seasonal fires ....
-Precipitation interacts with disturbance to increase the prevalence of fires by adding grass fuel load and increasing the probability of a fire occurring, and the intensity of the fire when a fire occurs.
+While southern African woodlands are species rich in the herbaceous understorey, the tree layer is relatively species poor \citep{}.
+\citet{Miombo-Book} however, writes that they are structurally diverse in the tree layer, with trees of the same species occupying distinct layers of the canopy at different growth stages.
+This structural diversity may be one mechanism through which species diversity appears to influence woody biomass.
+Trees occupying different canopy layers can create a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area.
+An underpinning principle of the Biodiversity-Ecosystem Function Relationship (BEFR) is that of niche complementarity.
+Species are assumed to differ in their functional niche by the fact that they may coexist in a steady-state ecological community, therefore the more species are present the greater the area of the total fundamental niche of the ecosystem is filled, allowing more efficient use of natural resources and higher gross primary productivity and biomass \citep{}.
+This theory however, which has been supported by many experiments and observational studies in temperate and wet tropical ecosystems, may not hold in savannas, which are structured by disturbance rather than competition.
 
-Grimes, mass ratio, selection effects and complementarity effects
+High levels of disturbance in the form of seasonal fires and herbivory from large mammals \citep{Fire is a herbivore} may moderate the strength of observable BEFRs in southern African woodlands.
+Fire disturbance in forests has been linked to abundance dependent mortality among smaller stems \citep{Roques2001}.
+A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion \citep{}.
+Such fast-growing bushy species exist in southern African woodlands and often proliferate when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{nieto2018}.
+Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
 
-In this study, we made the first known attempt to make a regional estimation of the Biodiversity Ecosystem Function Relationship in southern African woodlands.
-We used aboveground woody biomass and estimates of ecosystem productivity as measures of ecosystem functionality to understand the co-relationship between tree species biodiversity and ecosystem functionality.
+In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \ref{}).
+We used aboveground woody biomass \todo{and estimates of ecosystem productivity} as measures of ecosystem functionality to understand the co-relationship between tree species biodiversity and ecosystem functionality.
 We compared the relative effects of tree species biodiversity with that of other environmental factors known to affect ecosystem productivity and biomass accumulation such as precipitation and temperature.
-In acknowledgement of the wide variation in biogeographically determined community composition across the region, we also used previously determined biogeographical clusters \citep{} to understand how species composition as well as species biodiversity \textit{per se} affected ecosystem functionality.
+In acknowledgement of the wide variation in biogeographically determined community composition across the region, we also used biogeographical clusters \citep{} to understand how species composition as well as species biodiversity \textit{per se} affected ecosystem functionality.
 Initially, we made four hypotheses: 1) across the region, after other environmental factors had been accounted for, woodland plots with a higher species richness would have a higher aboveground biomass, 2) increased aridity would increase the strength of the biomass tree species richness relationship, 3) species composition would have a greater affect on aboveground biomass than species diversity \textit{per se}.
@@
 
-We hypothesised that plots with a higher species richness will maintain higher biomass stocks.
-
-Species composition will influence above-ground productivity more than species richness \textit{per se}.
-
-Plots in more arid regions will feature a stronger positive effect of tree species richness on above ground biomass stocks, due to abiotic facilitation effects, despite lower overall tree species richness.
-
-Structural characteristics of the woodland will interact with species composition and richness to provide an indirect path of influecne between species composition and biomass stocks.
+\begin{enumerate}
+	\item{Plots with a higher tree species diversity will maintain higher biomass stocks.}
+	\item{Species abundance evenness will influence above-ground productivity more than species richness \textit{per se}.}
+	\item{Plots in more arid regions will feature a stronger positive effect of tree species richness on above ground biomass stocks, due to abiotic facilitation effects, despite lower overall tree species richness.}
+	\item{Structural characteristics of the woodland will interact with species composition and richness to provide an indirect path of influence between species composition and biomass stocks.}
+\end{enumerate}
 
@@
 
-All analyses were performed in R version 3.6.0
+\subsection{Study location}
 
-Plot data were collated from a network of plots.
+The study used 1066 woodland monitoring from a larger network located across 10 countries within southern Africa in so-called Miombo woodland (Figure \ref{plot_loc}).
+The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in the \todo{where} and \textapprox{}1700 mm y\textsuperscript{-1} in the \todo{where}.
+The study sites straddle the climate space of the region as a whole (Figure \ref{temp_precip_hull}a).
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature of the study sites covers 92.7\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 seconds (0.86 km\textsuperscript{2} at the equator).
 
-Plots exist along a temperature and precipitation gradient (Figure ).
+\begin{figure}[H]
+	\centering
+	\subfloat[]{{\includegraphics[width=0.4\textwidth]{plot_loc}}}%
+    \qquad
+    \subfloat[]{{\includegraphics[width=0.4\textwidth]{temp_precip_hull}}}%
+    \caption{The locations of the 1066 plots used in this study, by geographic location (a) with respect to the distribution of Miombo woodland vegetation \citep{White}, and in climate space (b), showing the plot locations as points compared to the climate space of the whole region as estimated using the WorldClim dataset over the Miombo woodland vegetation extent.}
+\end{figure}
 
-In each of these plots, each living tree stem \textgreater{}5 cm Diameter at Breast Height (DBH) was measured. For each tree, species, DBH and height to the highest branch material were recorded.
+Plots were chosen from a larger pool of 5395 plots based on the quality of data collection and plot setup.
+Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna''.
+Plots with evidence of farming, human resource extraction or experimental treatements such as prescribed burning or herbivore exclusion were exlcuded from the initial pool.
+Only plots \textgreater{}0.1 hectares were used as per hectare biomass estimation from small plots is highly influenced by rare large trees.
+Only plots with a stem density \textless{}10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
 
-Plots varied in size across the region.
+Many plots provided by the Zambian Forestry Commission were arranged in clusters of four 20x50 m plots, 20 metres apart.
+These plots were aggregated for analyses and their values averaged for use in further plot level data cleaning.
 
-Rarefied species richness accounted for variation in plot size and therefore sampling effort across the region.
+A Principle Coordinate Analysis (PCoA) was conducted on tree species abundance matrix data for each plot to identify compositional outliers, which were considered not representative southern African woodlands.
+Plots within the upper 95th percentile of plot neighbour distance in the PCoA were excluded from further analyses (Figure \ref{pcoa_outlier}).
+This removed 66 plots.
 
-To ensure that all plots represented true woodland, plots with fewer than 20 stems per Ha were discarded.
+\begin{figure}[H]
+\centering
+	\includegraphics[width=0.5\textwidth]{pcoa_outlier}
+	\caption{Principle Co-ordinate Analysis of tree species diversity of the 1122 plots chosen from the larger pool of 5395 plots.
+The PCoA is constructed from a tree species abundance matrix.
+Plots which are within the upper 95th percentile of plot neighbour euclidean distance in the PCoA are marked in red, to be excluded from further analysis.}
+	\label{pcoa_outlier}
+\end{figure}
 
-Structural Equation Models were compared using different combinations of variables. Path analysis investigated the
+\subsection{Data collection}
 
-Species richness and the Shannon index (Shannon \& Weaver 1949) (Equation 1) were calculated to assess species diversity.
+We considered only trees and shrubs in our analyses of above-ground biomass, including woody species including palms and cycads but excluding lianas.
+Only stems \textgreater{}5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
+AGB estimation for trees \textless{}5 cm DBH is imprecise due to its sensitivity to many environmental and historical factors \citep{}.
+Additionally, for some species, species identification is difficult for saplings.
+
+All stems \textgreater{}5 cm DBH were measured within each plot.
+For each tree, species, DBH and height to the highest branch material were recorded.
+Height was measured through a variety of means \todo{audit of height methods}.
+When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
+The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial, parameters estimated using a test dataset \citep{Ryanunbublished}.
+
+Aboveground biomass for each plot was calculated using Equation \ref{chave_agb}, taken from \citet{Chave2014}.
+Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
+Wood density for species without species level estimates was estimated from the mean of their respective genus.
 
 \begin{equation}
-	H = -\sum{}P_i \ln{P_i} / \ln{S} 
+	AGB_{est} = 0.0673 \times (\rho D^{2} H)^{0.976}
+	\label{chave_agb}
 \end{equation}
 
+Where $\rho$ is the species level mean wood density, $D$ is the DBH at 1.3 m, and $H$ is the tree height.
+
+Climatic data was collected from the ECMWF ERA5 dataset \citep{}.
+Values were aggregated for each year between 2000 and 2018, then averaged across the years to provide single values per plot.
+Temperature and precipitation seasonality were both calculated as the coefficient of variation of these variables across the 18 years of available data.
+Soils data was extracted from the \todo{WHERE}.
+
+Fire return interval ($F_{ret}$) was calculated using the MODIS burned area product V6 (MCD46A1) \citep{}.
+Data was downloaded from January 2000 to December 2018.
+Mean fire return interval was calculated as:
+
+\begin{equation}
+	F_{ret} = \frac{\sum_{i = 1}^{n} t_{i} - t_{i-1}}{n-1}
+\end{equation}
+
+Where $t_{i}$ is the date of fire $i$, and $n$ is the total number of fires in the time period.
+Fire return interval data was binned into five ordinal categories ranging from ``frequent fire'' to ``no fire''.
+Bin values for categories where fire had occurred were defined as four bins with percentiles based on equal numbers of plots.
+Plots without a single fire recorded in the 18 year survey period were assumed to not be affected by fire.
+
+\subsection{Data analysis}
+Rarefied species richness accounted for variation in plot size and therefore sampling effort across the region.
+Rarefied species richness was calculated using the \verb|vegan| package in R, using a sample size of 20 stems.
+Plot size varied from 0.1 hectares to 10 hectares.
+
+To assess tree species abundance evenness, the Shannon Equitability index \citep{} (Equation \ref{shannon_equit}) was calculated:
+
+\begin{equation}
+	E_{H'} = \frac{-\sum_{i = 1}^{s} p_i \ln{p_i}}{\ln{S}}
+	\label{shannon_equit}
+\end{equation}
+
+Where $p$ is the proportion abundance of a tree species $i$ across all individuals in the plot and $S$ is the total number of species in the plot..
+The numerator of the Shannon Equitability index ($E_{H'}$) is the frequently used Shannon-Wiener diversity index ($H'$).
+
+\subsubsection{Biogeographical clusters}
+
+Plots were assigned to biogeographical groups after \citet{Fayolle2018}.
+Groups were identified using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data.
+Groups were identified using unconstrained correspondence analysis and ordination.
+Plot data used in this study occurred in five biogeographical groups, here referred to as: \todo{AS WHAT?}.
+
+\subsubsection{Structural Equation Modelling}
+
+Structural Equation Models (SEM) investigated the regional determinants of woody Above-Ground Biomass (AGB).
+All SEMs were constructed and analysed in the \textit{lavaan} package in R version 3.6.0.
+SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lehmann2014_161-166}.
+A key aspect to our decision to use SEMs is that they can explicitly model indirect effects, which is impossible in multiple regression.
+Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset.
+For example, moisture availability is known to affect productivity, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation, and soil composition.
+Indepedently of total precipitation, precipitation seasonality (measured as the covariance of daily precipitation) determines, simplistically, whether water arrives uniformly over a given time period or as few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
+We specified composite variables for moisture availability, temperature suitability, soil fertility, fire regime, tree species diversity and tree structural diversity.
+Latent variables.
+Although the fire regime composite only had a single indicator variable, it was nevertheless included in a latent variable to account for measurement error.
+
+A conceptual model was defined based on \textit{a priori} hypotheses of the direct and indirect effects of abiotic and biotic factors on AGB (Figure \ref{con_mod}).
+Latent variables were defined based on theoretical unmeasured factors known to affect AGB: moisture availability, temperature suitability for plant growth, soil fertility, fire regime intensity, tree species diversity, and tree structural diversity.
+Observed variables were collected in order to best inform these latent constructs.
+
+Observed variables of the latent variables were standardized to Z-scores for analysis.
+Standardization puts latent variables on the same scale, with a mean of zero and a standard deviation of one.
+It allows path regression coefficients to be easily compared between paths to assess their relative effect strength, and eliminates problems arising from the observed variables being on different scales.
+Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix.
+
+The factor loadings of each observed variable in a latent variable were set to 1 in all cases, assuming that each has an equal contribution to its latent variable.
+Due to some latent variables being regressed against both structural diversity and AGB, factor loadings from multiple regressions against either of these response variables could not be used to inform factor loadings in the SEM.
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=0.8\textwidth]{con_mod}
+	\caption{Conceptual model showing the theoretical relationships between environmental factors, structural and species tree diversity and AGB.
+Hypothesised paths of causal inference are depicted as arrows from predictor to response.}
+	\label{con_mod}
+\end{figure}
+
+The model allowed us to compare the relative effect sizes of the latent variables in determining AGB, and to assess the indirect effects of abiotic environment on AGB via species and structural tree diversity.
+Models including either abiotic or biotic factors only were compared using Comparative Fit Index (CFI), Akaike Information Criterion (AIC) and the R\textsuperscript{2} coefficient of determination for AGB.
+
+Two sets of SEMs were fitted.
+The first used only tree diversity latent constructs to estimate diversity, with the indirect effect of species diversity on AGB via structural diversity being explicitly modelled as a mediation effect to examine the strength and form of this relationship.
+The latent variable loadings differed between biogeographic clusters of plots.
+The second incorporated environmental covariates to understand the relative effects of climate, soils, disturbance on diversity and biomass.
+
 \section{Results}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=0.6\textwidth]{corr_mat}
+	\caption{Correlogram of exogenous variables used in the analysis, with correlation coefficients shaded according to sign (+ve red, -ve blue) and strength of correlation.}
+	\label{corr_mat}
+\end{figure}
+
+Simple correlations showed a positive association between Aboveground Biomass (AGB) and a composite latent variable representing tree species diversity ($r_{(1031)}$ = 0.30, $t$ = 11.3).
+
+In a simple path model describing the indirect effect of tree species diversity on AGB via the mediating effect of stand structural diversity, species diversity had a negative effect on AGB when the indirect path via stand structural diversity was accounted for (Figure \ref{struc_model_slopes}).
+Stand structural diversity had a strong positive effect on AGB (\todo{Test stats}).
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{struc_model_slopes}
+	\caption{Path coefficients (slopes) for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
+Path coefficients are +/- 1 standard error.
+Path coefficients where the standard error does not overlap zero are considered to be non-significant effects.}
+	\label{struc_model_slopes}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{full_latent_model_slopes}
+	\caption{Path coefficients (slopes) for the interactive effects of abiotic environment and tree diversity on AGB across all plots.
+Path coefficients are +/- 1 standard error.
+Path coefficients where the standard error does not overlap zero are considered to be non-significant effects.}
+	\label{full_latent_model_slopes}
+\end{figure}
+
+\subsection{Structural diversity}
+
+The latent variable of tree species diversity had a negative direct effect on AGB across the region.
 
@@
 \end{document}
+
+% Seidel 2019: Large trees tend to possess a greater structural complexity than small trees, not due to their size per se, but due to more complex architecture.

### 2019-10-17 analysis/img_transfer.sh

@@ -1,0 +1,21 @@
+#!/bin/bash
+
+OUT="../manuscript/img/"
+
+# Transfer images from analysis to manuscript
+
+cp img/plot_loc.pdf $OUT
+
+cp img/temp_precip_hull.pdf $OUT
+
+cp img/struc_model_slopes.pdf $OUT
+
+cp img/pcoa_outlier.pdf $OUT
+
+cp img/full_latent_model_slopes.pdf $OUT
+
+cp img/corr_mat.pdf $OUT
+
+cp img/biomass_clim_lm_clust.pdf $OUT
+
+cp img/pcoa.pdf $OUT

### 2019-10-17 analysis/lmm.R

@@ -16,7 +16,7 @@ library(dplyr)
 library(tidyr)
 library(ggplot2)
 library(lme4)
-library(colortools)
+library(gridExtra)
 
 # Cluster colour palette
 source("clust_pal.R")
@@ -26,155 +26,105 @@ source("clust_pal.R")
 # Aggregated and filtered data
 plot_data <- read.csv("data/plot_data_fil_agg_norm_std.csv")
 
-# Make cluster categorical
-plot_data$clust5 <- as.character(plot_data$clust5)
-
-# Remove plots not in a cluster
-plot_data <- plot_data %>%
-  filter(!is.na(clust5))
-
-# Plot all variables against bchave_log in a facet wrap ----
-
-# Climatic variables
-plot_data_gather_clim <- plot_data %>%
-  select(clust5, bchave_log, aridity_index, mean_temp, temp_seasonality, 
-    fire_return_mean_log, firecount_2001_2018,
-    ocdens, sand_per, cation_ex_cap) %>%
-  gather(key = "variable", value = "value", 3:9) %>%
-  mutate(facet_label = factor(variable, 
-    levels = c("aridity_index", "mean_temp", "temp_seasonality", 
-      "fire_return_mean_log", "firecount_2001_2018", "ocdens", "sand_per"),
-    labels = c(
-      expression("Aridity" ~ "index"),
-      expression("Mean" ~ "annual" ~ "temp." ~ (degree * C)), 
-      expression("Temp." ~ "seasonality" ~ (degree * C)),
-      expression("log(Mean" ~ "fire" ~ "return" ~ "interval)" ~ (yr)),
-      expression("Number" ~ "of" ~ "fires" ~ "2001-18"),
-      expression("Organic" ~ "C" ~ "%"),
-      expression("Sand" ~ "%"))))
-
-
-pdf(file = "img/biomass_clim_lm_clust.pdf", width = 12, height = 10)
-ggplot(plot_data_gather_clim, aes(x = value, y = bchave_log)) + 
-  geom_point(aes(fill = clust5), colour = "black", shape = 21) + 
-  geom_smooth(method = "lm", aes(colour = clust5)) +
-  scale_fill_manual(name = "Cluster", values = clust_pal) + 
-  scale_colour_manual(name = "Cluster", values = clust_pal) + 
-  theme_classic() + 
-  facet_wrap(~facet_label, scales = "free_x", labeller = label_parsed)
-dev.off()
-
[…]

### 2019-10-17 analysis/plot_data_collation.R

@@ -45,6 +45,12 @@ ocdens <- read.csv("data/seosaw_ocdens.csv")
 fire_return <- read.csv("data/fire_return.csv")
 ##' Data copied from burn /exports/...
 
+# WorldClim temp.
+temp <- read.csv("data/plot_temp.csv")
+
+# WorldClim precip.
+precip <- read.csv("data/plot_precip.csv")
+
 # Clean data ----
 
 # Create large raw dataframe
@@ -58,7 +64,9 @@ plot_data <- ssaw8$struct %>%
   left_join(., ssaw8$cluster, by = c("plotcode" = "plotcode")) %>%
   left_join(., aridity_index, by = c("plotcode" = "plotcode")) %>%
   left_join(., ocdens, by = c("plotcode" = "plotcode")) %>%
-  left_join(., fire_return, by = c("plotcode" = "plotcode")) %>%
+  left_join(., fire_return, by = c("plotcode" = "plotcode")) %>%  
+  left_join(., temp, by = c("plotcode" = "plotcode")) %>%
+  left_join(., precip, by = c("plotcode" = "plotcode")) %>%
   dplyr::select(plotcode,
     longitude_of_centre = longitude_of_centre.x.x,
     latitude_of_centre = latitude_of_centre.x.x,
@@ -81,14 +89,42 @@ plot_data <- ssaw8$struct %>%
     sand_per = SNDPPT_M_sl1_1km_ll,
     org_c_per = ORCDRC_M_sl1_1km_ll,
     aridity_index = ai,
+    total_precip = bio12,
+    wc_total_precip = p_plot,
+    precip_seasonality = bio15,
+    wc_precip_seasonality = p_cov_plot,
     ocdens = ocdens,
     mean_temp = bio1,
+    wc_mean_temp = t_plot,
     temp_seasonality = bio4,
+    wc_temp_seasonality = t_cov_plot,
     pi, 
     plot_id,
     fire_return_mean,
     firecount_2001_2018,
     clust7, clust5, clust4)
+
+# Compare SEOSAW worldclim temp with worldclim from my extraction
+ggplot(plot_data, aes(x = mean_temp, y = wc_mean_temp)) + 
+  geom_point()
+
+ggplot(plot_data, aes(x = temp_seasonality, y = wc_temp_seasonality)) + 
+  geom_point()
+
+# Compare SEOSAW worldclim precip with worldclim from my extraction
[…]

### 2019-10-17 analysis/prod_calc/acm_gpp.R

@@ -90,7 +90,7 @@ closest2d <- function (id,lat,long,lat_in,long_in,nos_dim) {
 # set to the working directory for your analysis, 
 # NOTE 1: this script assumes that within this directory contains a "src" directory containing the ACM-GPP-ET model
 # NOTE 2: this script assumes that within this directory contains a "input" directory containing the *csv file inputs
-setwd("/Users/johngodlee/google_drive/postgrad/phd/thesis/chapters/chapter_1_biomass_region_befr/analysis/productivity_calculations") ; 
+setwd("/Users/johngodlee/google_drive/postgrad/phd/thesis/chapters/chapter_1_biomass_region_befr/analysis/prod_calc") ; 
 wkdir = getwd()
 # compile fortran code into a shared object which R can load as a function
 system("gfortran src/ACM_GPP_ET_R_interface.f90 -o src/acm_gpp_et.so -fPIC -shared")
@@ -272,18 +272,24 @@ acm_gpp_et_output = list(    units = units,
 save(acm_gpp_et_output, file="./acm_output/ACM_GPP_ET_test_output.RData")
 
 ###
-## Generate simple plots for the first two sites
+## Generate simple plots for the first three sites
 
-if (n > 1) {par(mfrow=c(2,4), mar = c(4,4,3,1), omi=c(0.1,0.1,0.1,0.1))} else {par(mfrow=c(2,2), mar = c(4,4,3,1), omi=c(0.1,0.1,0.1,0.1))}
+if (n > 1) {par(mfrow=c(3,4), mar = c(4,4,3,1), omi=c(0.1,0.1,0.1,0.1))} else {par(mfrow=c(2,2), mar = c(4,4,3,1), omi=c(0.1,0.1,0.1,0.1))}
 plot(timeseries_lai[1,]~seq(1,length(drivers$doy)), ylab="LAI", xlab="Day of simulation")
 plot(timeseries_gpp[1,]~seq(1,length(drivers$doy)), ylab="GPP", xlab="Day of simulation")
 plot(timeseries_transpiration[1,]~seq(1,length(drivers$doy)), ylab="Transpiration", xlab="Day of simulation")
 plot(timeseries_soilwatermm[1,]~seq(1,length(drivers$doy)), ylab="Top soil water (mm)", xlab="Day of simulation")
 if (n > 1) {
[…]

### 2019-10-17 analysis/prod_calc/data_collation/grb_to_nc.sh

@@ -1,4 +1,0 @@
-#!/bin/bash
-
-cdo -f nc4 copy ../raw_data/era5.grb ../raw_data/era5.nc
-

### 2019-10-17 analysis/prod_calc/data_download_scripts_metadata/grb_to_nc.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+
+for file in *.grb; do
+    cdo -f nc4 copy $file "$(basename "$file" .grb).nc"
+done
+

### 2019-10-17 analysis/prod_calc/era5_cdsapi_req.py

@@ -1,0 +1,51 @@
+import cdsapi
+from datetime import date
+
+target_dir = "/Users/johngodlee/google_drive/postgrad/phd/thesis/chapters/chapter_1_biomass_region_befr/prod_calc/raw_data/era5/"
+
+start_date = date(2016, 1, 1)
+end_date = date(2018, 12, 31)
+
+date_list = [date.fromordinal(i) for i in range(start_date.toordinal(), end_date.toordinal() + 1)]
+
+def fnServerRetrieve(date):
+  year = date.strftime('%Y')
+  month = date.strftime('%m')
+  day = date.strftime('%d')
+
+  c = cdsapi.Client()
+  c.retrieve(
+	'reanalysis-era5-single-levels',
+	{
+	'product_type':'reanalysis',
+	'format':'netcdf',
+	'variable':[
+		'10m_u_component_of_wind','10m_v_component_of_wind','2m_temperature','2m_dewpoint_temperature',
+		'leaf_area_index_high_vegetation','leaf_area_index_low_vegetation','maximum_2m_temperature_since_previous_post_processing',
+		'minimum_2m_temperature_since_previous_post_processing','surface_net_solar_radiation','surface_net_solar_radiation_clear_sky',
+		'top_net_solar_radiation','top_net_solar_radiation_clear_sky','total_cloud_cover',
+		'total_precipitation'
+		],
+	'area':[40, -20,-40, 60], # North, West, South, East
+	'year':year,
+	'month':month,
+	'day':day,
+	'time':[
+		'00:00','02:00','04:00','06:00',
+		'08:00','10:00','12:00','14:00',
+		'16:00','18:00','20:00','22:00'
+		]
+	},
+	target_dir + "era5" + "_" + year + "_" + month + "_" + day + ".nc")
+	return
+
+for i in date_list:
+	year = i.strftime('%Y')
+	month = i.strftime('%m')
+	day = i.strftime('%d')
+	print("day:" + day + " month:" + month + " year:" + year)
+	fnServerRetrieve(i)
+
+
+
+

### 2019-10-17 analysis/prod_calc/merge_extracted.R

@@ -1,0 +1,68 @@
+# Aggregate all data sources
+# John Godlee (johngodlee@gmail.com)
+# 2018_10_30
+
+# Preamble ----
+
+# Remove old crap
+rm(list = ls())
+
+# Set working directory to the location of the source file
+setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(plyr)
+library(dplyr)
+
+# Load data ----
+
+extracted_ecmwf_df <- readRDS("extracted_data/ecmwf_extrac_df.rds")
+
+soil <- read.csv("extracted_data/soil_data_plots.csv")
+
+# Subset for testing to three plots
+soil_test <- soil %>%
+  filter(plotcode %in% c("ABG-001", "MAR-030", "ZLT-009"))
+
+# Join datasets
+
+all_data <- extracted_ecmwf_df %>%
+  left_join(., soil_test, by = c("plot_match" = "plotcode"))
+
+# Clean joined dataset
+str(all_data)
+
+all_data_clean <- all_data %>%
+  dplyr::select(
+    date,
+    plot_match,
+    lon = lon_match,
+    lat = lat_match,
+    wind_speed_mean,
+    lai_high_mean,
+    lai_low_mean,
+    temp_mean,
+    max_temp,
+    min_temp,
+    vpd_mean,
+    surf_net_sol_sum,
+    surf_net_sol_clear_sum,
+    top_net_sol_sum,
+    top_net_sol_clear_sum,
+    cloud_mean,
+    precip_sum,
+    CLYPPT_M_sl1_1km_ll,
+    CLYPPT_M_sl7_1km_ll,
+    SLTPPT_M_sl1_1km_ll,
+    SLTPPT_M_sl7_1km_ll,
+    SNDPPT_M_sl1_1km_ll,
+    SNDPPT_M_sl7_1km_ll,
+    ORCDRC_M_sl1_1km_ll,
+    ORCDRC_M_sl7_1km_ll,
+    CECSOL_M_sl1_1km_ll,
+    CECSOL_M_sl7_1km_ll
+  )
+
+write.csv(all_data_clean,
+  "extracted_data/all_data_compiled.csv", 
+  row.names = FALSE)

### 2019-10-17 analysis/prod_calc/plot_split.R

@@ -15,19 +15,19 @@ setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
 library(dplyr)
 
 # Load data 
-df <- read.csv("all_data_compiled.csv")
+df <- read.csv("extracted_data/all_data_compiled.csv")
 
 # Format data like Luke's data
 df$date <- as.Date(df$date, format = "%Y-%m-%d")
 df$doy <- format(df$date, format = "%j")
-df$min_T_oC <- df$min_temp
-df$max_T_oC <- df$max_temp
-df$swrad_MJ_day <- df$total_net_sol
-df$rainfall_mmday <- df$precip_trmm
-df$wind_ms <- 3.23
-df$VPD_kPa <- 1
+df$min_T_oC <- df$min_temp - 273.15
+df$max_T_oC <- df$max_temp - 273.15
+df$swrad_MJ_day <- df$surf_net_sol_clear_sum * 0.000001
+df$rainfall_mmday <- df$precip_sum * 1000
+df$wind_ms <- df$wind_speed_mean
+df$VPD_kPa <- df$vpd_mean
 df$co2 <- 400
-df$lai <- 0.469393245
+df$lai <- df$lai_high_mean
 df$latitude <- df$lat
 
 df_clean <- df %>%
@@ -40,12 +40,12 @@ df_list <- split(df_clean, df_clean$plot_match)
 # Extract latitude and save as text file
 lat_vec <- as.character(unname(sapply(df_list, function(x){mean(x$latitude)})))
 
-fileConn <- file("lat_vec.txt")
+fileConn <- file("extracted_data/lat_vec.txt")
 writeLines(lat_vec, fileConn)
 close(fileConn)
 
 # Save each element as a CSV
 for(i in 1:length(df_list)){
   csv <- df_list[[i]][,3:11]
-  write.csv(csv, paste0("../inputs/ACM_GPP_ET_", names(df_list[i]), ".csv"), row.names = FALSE)
+  write.csv(csv, paste0("acm_input/ACM_GPP_ET_", names(df_list[i]), ".csv"), row.names = FALSE)
 }

### 2019-10-22 manuscript/chapter_1.tex
Added some model outputs

@@
 A number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Review}.
-The strength and direction of these observed relationships varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{}.
+The strength and direction of these observed relationships varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}.
 Over the past decade, many observational studies of the Biodiversity-Ecosystem Function Relationship (BEFR) have been conducted in tropical and temperate forests, and grasslands \citep{Chen2011}.
 The representation of dry tropical ecosystems in the BEFR literature however, is poor.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in Costa Rica.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the highly endemic wet tropical forests Costa Rica.
 
@@
 Resource extraction directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}.
-Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation dynamics and affecting climate seasonality \citep{}.
+Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality \citep{}.
 Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning \textgreater{}4 million km\textsuperscript{2}.
@@
 Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are \textgreater{}300 tree species \citep{Frost1996}.
-They have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}.
+These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the banner of REDD+ \citep{Hinsley2015}.
@@
 
-The role of environmental factors in shaping the BEFR generally is poorly understood \citep{Ratcliffe2017}.
-Climatic variation is known to affect woody biomass \citep{Michaletz2014}, so it is important to account for these factors and acknowledge their interaction with woody species diversity \textit{and} biomass stocks to effectively model the effects of biodiversity on biomass stocks in a regional analysis.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in these ecosystems, which is presumably positively correlated with biomass, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
+The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}.
+Climatic variation is known to affect woody biomass \citep{Michaletz2014}, so it is important to account for these factors and acknowledge their interaction with both woody species diversity \textit{and} biomass stocks to effectively model the effects of biodiversity on biomass stocks in a regional analysis.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in these ecosystems, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
 However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.
 
-\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution in Panamanian tropical forests, which may affect ecosystem level productivity and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under certain environmental conditions.
+\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution in Panamanian tropical forests, which may affect ecosystem level productivity and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under specific environmental conditions.
 In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments, suggesting variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}.
 Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}.
+
 Temperature imposes a similar physiological limit on woody growth rate by limiting metabolic rates.
@@
 For each tree, species, DBH and height to the highest branch material were recorded.
-Height was measured through a variety of means \todo{audit of height methods}.
+Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which resulted in a variety of measurement uncertainty in the dataset, however, averaged over such a large number of stems this should not meaningfully alter the results of our analysis.
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
@@
 
-Plots were assigned to biogeographical groups after \citet{Fayolle2018}.
-Groups were identified using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data.
+Plots were assigned to biogeographical groups, after \citet{Fayolle2018}.
+Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data.
 Groups were identified using unconstrained correspondence analysis and ordination.
-Plot data used in this study occurred in five biogeographical groups, here referred to as: \todo{AS WHAT?}.
+Plot data used in this study occurred in five biogeographical groups, see Table \ref{clust_tabl} for a description of each biogeographical cluster.
 
@@
 Indepedently of total precipitation, precipitation seasonality (measured as the covariance of daily precipitation) determines, simplistically, whether water arrives uniformly over a given time period or as few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
-We specified composite variables for moisture availability, temperature suitability, soil fertility, fire regime, tree species diversity and tree structural diversity.
-Latent variables.
-Although the fire regime composite only had a single indicator variable, it was nevertheless included in a latent variable to account for measurement error.
 
-A conceptual model was defined based on \textit{a priori} hypotheses of the direct and indirect effects of abiotic and biotic factors on AGB (Figure \ref{con_mod}).
-Latent variables were defined based on theoretical unmeasured factors known to affect AGB: moisture availability, temperature suitability for plant growth, soil fertility, fire regime intensity, tree species diversity, and tree structural diversity.
-Observed variables were collected in order to best inform these latent constructs.
-
-Observed variables of the latent variables were standardized to Z-scores for analysis.
-Standardization puts latent variables on the same scale, with a mean of zero and a standard deviation of one.
-It allows path regression coefficients to be easily compared between paths to assess their relative effect strength, and eliminates problems arising from the observed variables being on different scales.
-Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix.
-
-The factor loadings of each observed variable in a latent variable were set to 1 in all cases, assuming that each has an equal contribution to its latent variable.
-Due to some latent variables being regressed against both structural diversity and AGB, factor loadings from multiple regressions against either of these response variables could not be used to inform factor loadings in the SEM.
+We specified a conceptual model with latent variables based on theoretical unmeasured factors known to affect AGB: moisture availability, temperature suitability for plant growth, soil fertility, fire regime intensity, tree species diversity and tree structural diversity (Figure \ref{con_mod}).
+Although the fire regime composite only had a single indicator variable, it was nevertheless included in a latent variable to soak up measurement error \citep{}.
 
@@
 
-The model allowed us to compare the relative effect sizes of the latent variables in determining AGB, and to assess the indirect effects of abiotic environment on AGB via species and structural tree diversity.
-Models including either abiotic or biotic factors only were compared using Comparative Fit Index (CFI), Akaike Information Criterion (AIC) and the R\textsuperscript{2} coefficient of determination for AGB.
+Observed variables of the latent variables were standardized to Z-scores for analysis.
+Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
+It allows path regression coefficients to be easily compared between paths to assess their relative effect strength, and eliminates problems arising from the observed variables being on different scales.
+Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix.
+
+The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention \citep{}.
+Due to some latent variables being regressed against both structural diversity and AGB, exact factor loadings from multiple regressions against either of these response variables could not be used to inform factor loadings in the SEM and were therefore estimated by the SEM model.
+
+For each biogeographical cluster, and at the regional level, we employed SEM to assess the relative importance of each latent variable in determining AGB.
+We assessed the indirect effects of abiotic environment on AGB via species and structural tree diversity to understand the interactive nature of species diversity and abiotic environment in determining AGB.
+In a model selection process, latent variable paths for each biogeographical cluster model were removed sequentially to investigate their removal on model fit.
+Model fit at each stage of the model selection process was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI) and the R\textsuperscript{2} coefficient of determination for AGB.
+The best models according to these goodness-of-fit statistics are presented in the results.
+See \todo{Appendix 1 for a full description of all models in the model description process.}
 
@@
 The first used only tree diversity latent constructs to estimate diversity, with the indirect effect of species diversity on AGB via structural diversity being explicitly modelled as a mediation effect to examine the strength and form of this relationship.
-The latent variable loadings differed between biogeographic clusters of plots.
-The second incorporated environmental covariates to understand the relative effects of climate, soils, disturbance on diversity and biomass.
+The second incorporated environmental covariates to understand the relative effects of climate, soils, disturbance on diversity and AGB.
 
@@
 
-Simple correlations showed a positive association between Aboveground Biomass (AGB) and a composite latent variable representing tree species diversity ($r_{(1031)}$ = 0.30, $t$ = 11.3).
+Simple correlations showed a positive association between Aboveground Biomass (AGB) and a latent variable representing tree species diversity ($r_{(1522)}$ = 0.30, $t$ = 12.2).
+The tree structural diversity latent variable also had a positive association with AGB ($r_{(1327)}$ = 0.18, $t$ = 15.0).
 
-In a simple path model describing the indirect effect of tree species diversity on AGB via the mediating effect of stand structural diversity, species diversity had a negative effect on AGB when the indirect path via stand structural diversity was accounted for (Figure \ref{struc_model_slopes}).
+In an SEM describing the indirect effect of tree species diversity on AGB via the mediating effect of stand structural diversity, species diversity had a negative effect on AGB when the indirect path via stand structural diversity was accounted for (Figure \ref{struc_model_slopes}).
 Stand structural diversity had a strong positive effect on AGB (\todo{Test stats}).
+This effect was observed in biogeographical clusters 1, 2 and 3.
+Clusters 4 and 5 had small sample sizes due to missing data and so path regression uncertainties were high, with confidence intervals overlapping zero.
 
@@
 \centering
-	\includegraphics[width=\textwidth]{struc_model_slopes}
+	\includegraphics[width=\textwidth]{struc_model_slopes_clust_all}
 	\caption{Path coefficients (slopes) for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
@@
 	\label{struc_model_slopes}
+\end{figure}
+
+\input{include/struc_model_fit_clust_stats}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{sem_struc_path}
+	\caption{Path diagram with regression coefficients for the structural model including plots in all five clusters.}
+	\label{sem_struc_path}
 \end{figure}
@@
 
-The latent variable of tree species diversity had a negative direct effect on AGB across the region.
+The latent variable of tree species diversity had a weak positive effect on AGB across the region.
+A stronger positive effect was seen between structural diversity and aboveground biomass, with species diversity acting \textit{through} structural diversity to effect aboveground biomass.
 
 \section{Discussion}
+
+%\bibliography{/Users/johngodlee/google_drive/bib/lib}
 

### 2019-10-22 manuscript/include/struc_model_fit_clust_stats.tex
Added some model outputs

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Tue, Oct 22, 2019 - 11:39:36
\begin{table}[!htbp] \centering 
  \caption{} 
  \label{struc_model_fit_clust_stats} 
\begin{tabular}{@{\extracolsep{5pt}} cccccccccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & npar & chisq & df & cfi & tli & logl & aic & bic & ntotal & rmsea & rmsea.ci.lower & rmsea.ci.upper & srmr \\ 
\hline \\[-1.8ex] 
C1 & $12$ & $4.660$ & $3$ & $0.997$ & $0.989$ & $$-$1854$ & $3732$ & $3777.900$ & $337$ & $0.040$ & $0$ & $0.108$ & $0.016$ \\ 
C2 & $12$ & $22.830$ & $3$ & $0.982$ & $0.939$ & $$-$2838.600$ & $5701.200$ & $5754.300$ & $619$ & $0.100$ & $0.066$ & $0.145$ & $0.017$ \\ 
C3 & $12$ & $7.910$ & $3$ & $0.961$ & $0.871$ & $$-$502.300$ & $1028.500$ & $1057$ & $79$ & $0.140$ & $0.012$ & $0.270$ & $0.073$ \\ 
C4 & $12$ & $1.690$ & $3$ & $1$ & $1.054$ & $$-$139.600$ & $303.100$ & $318.200$ & $26$ & $0$ & $0$ & $0.265$ & $0.048$ \\ 
C5 & $12$ & $1.120$ & $3$ & $1$ & $1.068$ & $$-$368.200$ & $760.400$ & $785.300$ & $59$ & $0$ & $0$ & $0.146$ & $0.034$ \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

### 2019-10-22 analysis/sem.R

@@ -36,6 +36,8 @@ library(ggcorrplot)
 library(lme4)
 library(stargazer)
 
+source("clust_pal.R")
+
 # Import data ----
 
 # Aggregated and filtered data
@@ -55,7 +57,8 @@ sem_data<- sem_data %>%
 # Create a dataframe 
 corr_df <- sem_data %>%
   select(sand_per_std, ocdens_std, cation_ex_cap_std, 
-    total_precip_std, precip_seasonality_std, mean_temp_std, temp_seasonality_std,
+    total_precip_std, precip_seasonality_std, 
+    mean_temp_std, temp_seasonality_std, isothermality_std,
     fire_return_mean_log_std, sp_rich_raref_std, shannon_equit_std,
     cov_height_std, cov_dbh_std, bchave_log) %>%
   rename(`Sand %` = sand_per_std,
@@ -65,6 +68,7 @@ corr_df <- sem_data %>%
     `MAP CoV` = precip_seasonality_std, 
     `MAT` = mean_temp_std, 
     `MAT CoV` = temp_seasonality_std,
+    `Isothermality` = isothermality_std,
     `Fire return interval` = fire_return_mean_log_std, 
     `Raref. Sp. Rich.` = sp_rich_raref_std, 
     `Shannon equit.` = shannon_equit_std,
@@ -97,7 +101,7 @@ dev.off()
 
 env_path_model_fit <- lmer(bchave_log ~ ocdens_std + sand_per_std + cation_ex_cap_std +
   aridity_index_std + total_precip_std + precip_seasonality_std +
-  mean_temp_std + temp_seasonality_std + 
+  mean_temp_std + temp_seasonality_std + isothermality_std +
   fire_return_mean_log_std + (1|clust5), data = sem_data)
 
 sink("output/env_path_model_fit.txt")
@@ -106,14 +110,16 @@ print(MuMIn::r.squaredGLMM(env_path_model_fit))
 sink()
 
 # Indirect effect of diversity on biomass via structure ----
+
 struc_model_spec <- "
 # Latent vars
-div     =~  1*sp_rich_raref_std + 1*shannon_equit_std + 1*shannon_cube_std
-struc   =~  1*cov_dbh_std + 1*cov_height_std
+div     =~  1*sp_rich_raref_std + shannon_equit_std
+struc   =~  1*cov_dbh_std + cov_height_std
+biomass =~  1*bchave_log
 
 # Regressions
[…]

### 2019-10-29 analysis/output/include/clust_summ.tex
More functions in SEM to automate clusters

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Thu, Oct 24, 2019 - 13:20:25
\begin{table}[!htbp] \centering 
  \caption{} 
  \label{clust_summ} 
\begin{tabular}{@{\extracolsep{5pt}} ccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
clust5 & c\_ind & sp\_rich\_raref & n\_stems\_ha & bchave \\ 
\hline \\[-1.8ex] 
C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 16+12.6 & 281+373.8 & 34.7+37.1 \\ 
C2 & Julbernardia.paniculata, Isoberlinia.angolensis, Albizia.antunesiana & 22+13.7 & 274+246.9 & 46.7+35.07 \\ 
C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 8+6.5 & 272+535.3 & 36+27.93 \\ 
C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 15+9.2 & 448+391.9 & 46.6+45.41 \\ 
C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 9+9.4 & 336+434.3 & 43.8+34.45 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

### 2019-10-29 manuscript/chapter_1.tex
More functions in SEM to automate clusters

@@
 \usepackage{mynotes}
+
+\usepackage{dcolumn}
+\usepackage{multirow}
 
@@
 We compared the relative effects of tree species biodiversity with that of other environmental factors known to affect ecosystem productivity and biomass accumulation such as precipitation and temperature.
-In acknowledgement of the wide variation in biogeographically determined community composition across the region, we also used biogeographical clusters \citep{} to understand how species composition as well as species biodiversity \textit{per se} affected ecosystem functionality.
+We incorporated biogeographical clusters \citep{} into our analyses to understand how species composition as well as species biodiversity \textit{per se} affected ecosystem functionality.
 Initially, we made four hypotheses: 1) across the region, after other environmental factors had been accounted for, woodland plots with a higher species richness would have a higher aboveground biomass, 2) increased aridity would increase the strength of the biomass tree species richness relationship, 3) species composition would have a greater affect on aboveground biomass than species diversity \textit{per se}.
-We used Structural Equation Modelling as a preferred method to simultaneously account for environmental factors and biogeographic factors, which often interact in their effect on ecosystem structure and therefore biomass.
+We used Structural Equation Modelling as a preferred method to simultaneously account for environmental factors and biotic factors, which often interact in their effect on ecosystem structure and therefore biomass \citep{}.
 
@@
 	\item{Plots with a higher tree species diversity will maintain higher biomass stocks.}
-	\item{Species abundance evenness will influence above-ground productivity more than species richness \textit{per se}.}
-	\item{Plots in more arid regions will feature a stronger positive effect of tree species richness on above ground biomass stocks, due to abiotic facilitation effects, despite lower overall tree species richness.}
+	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, due to abiotic facilitation effects, despite lower overall tree species richness.}
 	\item{Structural characteristics of the woodland will interact with species composition and richness to provide an indirect path of influence between species composition and biomass stocks.}
@@
 The study used 1066 woodland monitoring from a larger network located across 10 countries within southern Africa in so-called Miombo woodland (Figure \ref{plot_loc}).
-The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in the \todo{where} and \textapprox{}1700 mm y\textsuperscript{-1} in the \todo{where}.
+The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in the north of Zambia, Malawi and northern Mozambique.
 The study sites straddle the climate space of the region as a whole (Figure \ref{temp_precip_hull}a).
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature of the study sites covers 92.7\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 seconds (0.86 km\textsuperscript{2} at the equator).
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers 92.7\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 seconds (0.86 km\textsuperscript{2} at the equator).
 
@@
 
-Plots were chosen from a larger pool of 5395 plots based on the quality of data collection and plot setup.
+Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup.
 Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna''.
@@
 Only plots \textgreater{}0.1 hectares were used as per hectare biomass estimation from small plots is highly influenced by rare large trees.
-Only plots with a stem density \textless{}10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
+Only plots with a stem density \textgreater{}10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
 
-Many plots provided by the Zambian Forestry Commission were arranged in clusters of four 20x50 m plots, 20 metres apart.
+Many plots provided by the Zambian Forestry Commission were arranged in clusters of up to four 20x50 m plots, 20 metres apart.
 These plots were aggregated for analyses and their values averaged for use in further plot level data cleaning.
 
-A Principle Coordinate Analysis (PCoA) was conducted on tree species abundance matrix data for each plot to identify compositional outliers, which were considered not representative southern African woodlands.
-Plots within the upper 95th percentile of plot neighbour distance in the PCoA were excluded from further analyses (Figure \ref{pcoa_outlier}).
-This removed 66 plots.
+A Principle Coordinate Analysis (PCoA) was conducted on tree species abundance matrix data for each suitable plot to identify compositional outliers, which were considered not representative southern African woodlands.
+Plots within the upper 95th percentile of neighbour distance in the PCoA were excluded from further analyses (Figure \ref{pcoa_outlier}).
+This procedure removed 66 plots.
 
@@
 
-We considered only trees and shrubs in our analyses of above-ground biomass, including woody species including palms and cycads but excluding lianas.
+We considered only trees and shrubs in our analyses of above-ground biomass, including woody species such as palms and cycads which are functionally tree-like but excluding lianas which fill a different ecological niche.
 Only stems \textgreater{}5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
-AGB estimation for trees \textless{}5 cm DBH is imprecise due to its sensitivity to many environmental and historical factors \citep{}.
+AGB estimation for stems \textless{}5 cm DBH is imprecise due to its sensitivity to many environmental and historical factors \citep{}.
 Additionally, for some species, species identification is difficult for saplings.
 
-All stems \textgreater{}5 cm DBH were measured within each plot.
-For each tree, species, DBH and height to the highest branch material were recorded.
-Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which resulted in a variety of measurement uncertainty in the dataset, however, averaged over such a large number of stems this should not meaningfully alter the results of our analysis.
+All stems \textgreater{}5 cm DBH were measured within each plot resulting in a total of 160,076 stems with measurements.
+Note that a tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
+For each stem, species, DBH and height to the highest branch material were recorded.
+Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which may cause potentially large measurement uncertainty in the dataset.
+Averaged over such a large number of stems however, this should not meaningfully alter the results of our analysis.
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
@@
 Climatic data was collected from the ECMWF ERA5 dataset \citep{}.
-Values were aggregated for each year between 2000 and 2018, then averaged across the years to provide single values per plot.
+Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
 Temperature and precipitation seasonality were both calculated as the coefficient of variation of these variables across the 18 years of available data.
-Soils data was extracted from the \todo{WHERE}.
+Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}.
+We extracted Cation Exchange Capacity (CEC), Organic Carbon Density, and the percentage sand content of the soil.
+These data are a modelled product compiled from various remote sensed and directly measured data sources.
 
-Fire return interval ($F_{ret}$) was calculated using the MODIS burned area product V6 (MCD46A1) \citep{}.
+\todo{Fire return interval ($F_{ret}$) was calculated using the MODIS burned area product V6 (MCD46A1) \citep{}.
 Data was downloaded from January 2000 to December 2018.
-Mean fire return interval was calculated as:
+Mean fire return interval was calculated as:}
 
 \begin{equation}
-	F_{ret} = \frac{\sum_{i = 1}^{n} t_{i} - t_{i-1}}{n-1}
+	\todo{F_{ret} = \frac{\sum_{i = 1}^{n} t_{i} - t_{i-1}}{n-1}}
 \end{equation}
 
-Where $t_{i}$ is the date of fire $i$, and $n$ is the total number of fires in the time period.
-Fire return interval data was binned into five ordinal categories ranging from ``frequent fire'' to ``no fire''.
-Bin values for categories where fire had occurred were defined as four bins with percentiles based on equal numbers of plots.
-Plots without a single fire recorded in the 18 year survey period were assumed to not be affected by fire.
+\todo{Where $t_{i}$ is the date of fire $i$, and $n$ is the total number of fires in the time period.}
 
 \subsection{Data analysis}
-Rarefied species richness accounted for variation in plot size and therefore sampling effort across the region.
-Rarefied species richness was calculated using the \verb|vegan| package in R, using a sample size of 20 stems.
-Plot size varied from 0.1 hectares to 10 hectares.
+Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}.
+This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richnes value.
+These estimated species richness values are referred to as rarefied species richness throughout the paper.
+Rarefied species richness accounted for variation in plot size (0.1-10 ha) and therefore sampling effort across the region.
+Larger plots will tend to encompass more individuals, and therefore probabilistically more species, according to the species-area relationship \citep{}.
 
@@
 \begin{equation}
-	E_{H'} = \frac{-\sum_{i = 1}^{s} p_i \ln{p_i}}{\ln{S}}
+	\begin{gathered}
+		E_{H'} = \frac{H'_{e}}{\ln{S}} \\
+	\end{gathered}
 	\label{shannon_equit}
@@
 
-Where $p$ is the proportion abundance of a tree species $i$ across all individuals in the plot and $S$ is the total number of species in the plot..
-The numerator of the Shannon Equitability index ($E_{H'}$) is the frequently used Shannon-Wiener diversity index ($H'$).
+Where $H'_{e}$ is an estimation of the Shannon diversity index by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the total number of species in the plot.
 
@@
 
-Plots were assigned to biogeographical groups, after \citet{Fayolle2018}.
+Plots were assigned to biogeographical groups based on tree species composition.
 Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data.
 Groups were identified using unconstrained correspondence analysis and ordination.
-Plot data used in this study occurred in five biogeographical groups, see Table \ref{clust_tabl} for a description of each biogeographical cluster.
+Plot data used in this study occurred in five biogeographical groups, see Table \ref{clust_test} for a description of each biogeographical cluster and Map \ref{clust_map} for the spatial distribution of plots from each of these clusters.
+
+\input{include/clust_summ}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=0.6\textwidth]{clust_map}
+	\caption{}
+	\label{clust_map}
+\end{figure}
 
@@
 
-Structural Equation Models (SEM) investigated the regional determinants of woody Above-Ground Biomass (AGB).
-All SEMs were constructed and analysed in the \textit{lavaan} package in R version 3.6.0.
+Structural Equation Models (SEM) investigated the determinants of woody Above-Ground Biomass (AGB).
+All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R}.
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lehmann2014_161-166}.
@@
 Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset.
-For example, moisture availability is known to affect productivity, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation, and soil composition.
-Indepedently of total precipitation, precipitation seasonality (measured as the covariance of daily precipitation) determines, simplistically, whether water arrives uniformly over a given time period or as few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
+For example, moisture availability is known to affect productivity \citep{}, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation and the seasonality of that precipitation.
+Indepedently of total precipitation, precipitation seasonality, measured as the covariance of daily precipitation, determines whether water arrives uniformly over a given time period or as few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
 
 We specified a conceptual model with latent variables based on theoretical unmeasured factors known to affect AGB: moisture availability, temperature suitability for plant growth, soil fertility, fire regime intensity, tree species diversity and tree structural diversity (Figure \ref{con_mod}).
-Although the fire regime composite only had a single indicator variable, it was nevertheless included in a latent variable to soak up measurement error \citep{}.
+\todo{Although the fire regime composite only had a single indicator variable, it was nevertheless included in a latent variable to soak up measurement error \citep{}.}
 
@@
 	\caption{Conceptual model showing the theoretical relationships between environmental factors, structural and species tree diversity and AGB.
-Hypothesised paths of causal inference are depicted as arrows from predictor to response.}
+Hypothesised paths of causal inference are depicted as arrows from predictor to response.
+The shaded box defines the model which includes only tree diversity variables.}
 	\label{con_mod}
@@
 
-Observed variables of the latent variables were standardized to Z-scores for analysis.
+Observed variables were standardized to Z-scores for analysis.
 Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
-It allows path regression coefficients to be easily compared between paths to assess their relative effect strength, and eliminates problems arising from the observed variables being on different scales.
-Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix.
+It allows path regression coefficients to be easily compared between paths to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}.
+Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}.
 
-The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention \citep{}.
+The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention \citep{}, with other observed variables being allowed to vary.
 Due to some latent variables being regressed against both structural diversity and AGB, exact factor loadings from multiple regressions against either of these response variables could not be used to inform factor loadings in the SEM and were therefore estimated by the SEM model.
 
-For each biogeographical cluster, and at the regional level, we employed SEM to assess the relative importance of each latent variable in determining AGB.
-We assessed the indirect effects of abiotic environment on AGB via species and structural tree diversity to understand the interactive nature of species diversity and abiotic environment in determining AGB.
-In a model selection process, latent variable paths for each biogeographical cluster model were removed sequentially to investigate their removal on model fit.
-Model fit at each stage of the model selection process was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI) and the R\textsuperscript{2} coefficient of determination for AGB.
-The best models according to these goodness-of-fit statistics are presented in the results.
-See \todo{Appendix 1 for a full description of all models in the model description process.}
+For each biogeographical cluster separately, and at the regional scale, we employed SEM to assess the relative importance of each latent variable in determining AGB.
+We assessed the indirect effects of abiotic environment on AGB via the mediators of tree species diversity and tree structural diversity to understand the interactive nature of tree diversity and abiotic environment in determining AGB.
+Model fit  was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
+We follow the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI \textgreater{} 0.95, TLI \textgreater{} 0.95, RMSEA \textless{} 0.06.
 
-Two sets of SEMs were fitted.
-The first used only tree diversity latent constructs to estimate diversity, with the indirect effect of species diversity on AGB via structural diversity being explicitly modelled as a mediation effect to examine the strength and form of this relationship.
+Two sets of SEMs were fitted (Figure \ref{con_mod}).
+The first used only tree diversity latent variables to estimate diversity, with the indirect effect of species diversity on AGB via structural diversity being explicitly modelled as a mediation effect to examine the strength and form of this relationship.
 The second incorporated environmental covariates to understand the relative effects of climate, soils, disturbance on diversity and AGB.
+Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Kline2005}.
+
+For models on the whole region, standardized path coefficients are reported in order to better compare the relative contribution of different paths within the model.
+When comparing the path coefficients of models using individual biogeographical clusters, unstandardized path coefficients are used in order to more readily compare the same path coefficient across different models while accounting for variation in mean values between biogeographical clusters.
 
@@
 	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlogram of exogenous variables used in the analysis, with correlation coefficients shaded according to sign (+ve red, -ve blue) and strength of correlation.}
+	\caption{Correlogram of observed exogenous variables used in the SEMs, with correlation coefficients shaded according to sign (+ve red, -ve blue) and strength of correlation.}
 	\label{corr_mat}
@@
 
-Simple correlations showed a positive association between Aboveground Biomass (AGB) and a latent variable representing tree species diversity ($r_{(1522)}$ = 0.30, $t$ = 12.2).
-The tree structural diversity latent variable also had a positive association with AGB ($r_{(1327)}$ = 0.18, $t$ = 15.0).
+At the regional level, tree species and structural diversity observed variables had weak positive correlations with AGB (Figure \ref{corr_mat}), while environmental observed variables had very weak correlations with AGB.
 
-In an SEM describing the indirect effect of tree species diversity on AGB via the mediating effect of stand structural diversity, species diversity had a negative effect on AGB when the indirect path via stand structural diversity was accounted for (Figure \ref{struc_model_slopes}).
-Stand structural diversity had a strong positive effect on AGB (\todo{Test stats}).
-This effect was observed in biogeographical clusters 1, 2 and 3.
-Clusters 4 and 5 had small sample sizes due to missing data and so path regression uncertainties were high, with confidence intervals overlapping zero.
+In an SEM describing the effect of tree species diversity on AGB via the mediating effect of stand structural diversity (Figure \ref{sem_struc_path}, species diversity had a positive effect on AGB when the indirect path via stand structural diversity was accounted for (Figure \ref{struc_model_slopes}).
+Stand structural diversity had a positive effect on AGB (\todo{Test stats}).
+The indirect effect of diversity on biomass via structural diversity was small but significant ().
+Model fit for this model was good \todo{MORE}.
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{struc_model_slopes_clust_all}
-	\caption{Path coefficients (slopes) for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
-Path coefficients are +/- 1 standard error.
-Path coefficients where the standard error does not overlap zero are considered to be non-significant effects.}
-	\label{struc_model_slopes}
-\end{figure}
-
-\input{include/struc_model_fit_clust_stats}
+When the model was run again using only data
 
@@
 \centering
-	\includegraphics[width=\textwidth]{full_latent_model_slopes}
-	\caption{Path coefficients (slopes) for the interactive effects of abiotic environment and tree diversity on AGB across all plots.
+	\includegraphics[width=\textwidth]{struc_model_slopes}
+	\caption{Standardized path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
+Due to all observed variables being standardized and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale +/- 1 standard error.
+Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
+	\label{struc_model_slopes}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{struc_model_slopes_all}
+	\caption{Unstandardized path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
 Path coefficients are +/- 1 standard error.
-Path coefficients where the standard error does not overlap zero are considered to be non-significant effects.}
-	\label{full_latent_model_slopes}
+Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
+	\label{struc_model_slopes_all}
 \end{figure}
+
+\input{include/struc_model_fit_clust_stats}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{sem_full_path}
+	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters.}
+	\label{sem_full_path}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{full_model_slopes}
+	\caption{Standardized path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots.
+Path coefficients are +/- 1 standard error.
+Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
+	\label{full_model_slopes}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{full_model_slopes_all}
+	\caption{Unstandardized path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots.
+Path coefficients are +/- 1 standard error.
+Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
+	\label{full_model_slopes_all}
+\end{figure}
+
+\input{include/full_model_fit_clust_stats}
 
@@
 
-The latent variable of tree species diversity had a weak positive effect on AGB across the region.
+The latent variable of tree species diversity had a positive effect on AGB across the region.
 A stronger positive effect was seen between structural diversity and aboveground biomass, with species diversity acting \textit{through} structural diversity to effect aboveground biomass.
+
+\subsection{Variation between biogeographical clusters}
 
@@
 
+\subsection{Conclusion}
+
 \end{document}

### 2019-10-29 manuscript/include/clust_summ.tex
More functions in SEM to automate clusters

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Thu, Oct 24, 2019 - 13:20:25
\begin{table}[!htbp] \centering 
  \caption{} 
  \label{clust_summ} 
\begin{tabular}{@{\extracolsep{0pt}} clccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
{Cluster} & {Indicator species} & {Rarefied species richness} & {Stems ha\textsuperscript{-1}} & {AGB (t ha\textsuperscript{-1})} \\
\hline \\[-1.8ex] 
C1 & \begin{tabular}[l]{@{}l@{}l@{}} Diplorhynchus condylocarpon \\ Combretum spp. \\ Pseudolachnostylis maprouneifolia \end{tabular} & 16$\pm$12.6 & 281$\pm$373.8 & 34.7$\pm$37.1 \\ 
\hline
C2 & \begin{tabular}[l]{@{}l@{}l@{}} Julbernardia.paniculata \\ Isoberlinia.angolensis \\ Albizia.antunesiana \end{tabular} & 22$\pm$13.7 & 274$\pm$246.9 & 46.7$\pm$35.07 \\ 
\hline
C3 & \begin{tabular}[l]{@{}l@{}l@{}} Burkea africana \\ Pterocarpus angolensis \\ Baikiaea plurijuga \end{tabular} & 8$\pm$6.5 & 272$\pm$535.3 & 36$\pm$27.93 \\ 
\hline
C4 & \begin{tabular}[l]{@{}l@{}l@{}} Baikiaea plurijuga \\ Terminalia randii \\ Albizia amara \end{tabular} & 15$\pm$9.2 & 448$\pm$391.9 & 46.6$\pm$45.41 \\ 
\hline
C5 & \begin{tabular}[l]{@{}l@{}l@{}} Colophospermum mopane \\ Pseudolachnostylis maprouneifolia \end{tabular} & 9$\pm$9.4 & 336$\pm$434.3 & 43.8$\pm$34.45 \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

### 2019-10-29 analysis/bivar_rel_norm.R

@@ -61,174 +61,68 @@ theme.bivar <- function(base_size = 12, base_font_family = "Helvetica"){
     )}
 
 source("clust_pal.R")
-
-facet_levels <- c(
-  "aridity_index",
-  "total_precip",
-  "precip_seasonality",
-  "mean_temp",
-  "temp_seasonality",
-  "isothermality",
-  "bchave",
-  "cation_ex_cap",
-  "sand_per",
-  "ocdens",
-  "cov_dbh",
-  "cov_height",
-  "mean_dbh",
-  "mean_height",
-  "stems_ha",
-  "shannon",
-  "shannon_equit",
-  "sp_rich",
-  "sp_rich_raref",
-  "fire_return_mean",
-  "firecount_2001_2018",
-  "bchave_log", 
-  "mean_dbh_log",
-  "stems_ha_log", 
-  "shannon_cube",
-  "fire_return_mean_log",
-  "aridity_index_std",
-  "mean_temp_std",
-  "temp_seasonality_std", 
-  "isothermality_std",
-  "bchave_log_std", 
-  "cation_ex_cap_std",
-  "sand_per_std", 
-  "ocdens_std", 
-  "cov_dbh_std", 
-  "cov_height_std", 
-  "mean_dbh_log_std",
-  "mean_height_std", 
-  "stems_ha_log_std", 
-  "shannon_cube_std",
-  "shannon_equit_std",
-  "sp_rich_std", 
-  "sp_rich_raref_std", 
-  "fire_return_mean_log_std",
-  "firecount_2001_2018_std")
-
-facet_labels <- c(  
-  expression("Aridity" ~ "index"),
-  expression("Mean" ~ "Annual" ~ "Precipitation" ~ (mm ~ y^-1)),
-  expression("Precipitation" ~ "seasonality"),
-  expression("Mean" ~ "annual" ~ "temp." ~ (degree * C)), 
-  expression("Temp." ~ "seasonality" ~ (degree * C)),
-  expression("Isothermality"),
-  expression("AGB" ~ t ~ ha^-1),
-  expression("Cation" ~ "exchange" ~ "capacity"),
-  expression("Sand" ~ "%"), 
-  expression("Organic" ~ "C" ~ "%"),
-  expression("Coef." ~ "var." ~ "DBH"),
-  expression("Coef." ~ "var." ~ "height"),
-  expression("Mean" ~ "DBH" ~ (cm)),
-  expression("Mean" ~ "height" ~ (m)),
[…]

### 2019-10-29 analysis/cluster_species.R

@@ -1,0 +1,99 @@
+# Which species make up the vegetation clusters?
+# John Godlee (johngodlee@gmail.com)
+# 2019_10_24
+
+# Preamble ----
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Packages ----
+library(labdsv)
+library(stargazer)
+library(dplyr)
+
+# Import data ----
+plot_data <- read.csv("data/plot_data_fil_agg_norm_std.csv")
+plotcode_plot_group_lookup <- read.csv("data/plotcode_plot_group_lookup.csv")
+load("data/clean_input_data.Rdata")
+
+# Create community composition matrix of suitable plots
+s_fil <- s %>%
+  left_join(., plotcode_plot_group_lookup, by = c("plotcode" = "plotcode_vec")) %>%
+  inner_join(., data.frame(plot_group = plot_data$plot_group), 
+    by = c("plot_group" = "plot_group")) %>%
+  filter(alive %in% c("A", NA),
+    !is.na(gen_sp))
+
+# Create community composition matrix
+ab_mat <- s_fil %>%
+  group_by(plot_group, gen_sp, .drop = FALSE) %>%
+  tally() %>%
+  spread(gen_sp, n, fill = 0) %>%
+  ungroup() %>%
+  data.frame() %>%
+  select(-plot_group) %>%
+  select_if(any)
+
+# Remove species not present in any plot
+clust5_vec <- plot_data$clust5
+
+# Run Dufrene-Legendre indicator species analysis
+clust_indval <- indval(ab_mat, clustering = clust5_vec)
+
+sink("output/clust_species_indicators.txt")
+print(summary(clust_indval, p = 0.05, type = "short", digits = 2, show = p))
+sink()
+
+fileConn<-file("output/clust_species_summary.txt")
+writeLines(c(
+  "Cluster 1 - Diplorhynchus-Combretum",
+  "Combretum.zeyheri, Brachystegia.floribunda, Brachystegia.utilis",
+  "", 
+  "Cluster 2 - Miombo",
+  "Julbernardia.paniculata, Isoberlinia.angolensis, Albizia.antunesiana, Brachystegia.longifolia",
+  "",
+  "Cluster 3 - Burkea-Pterocarpus",
+  "Burkea.africana, Pterocarpus.angolensis, Baikiaea.plurijuga",
+  "",
[…]

### 2019-10-29 analysis/compositional_outliers.R

@@ -17,10 +17,13 @@ library(ggplot2)
 library(tidyr)
 library(dave)
 
+source("full_best.R")
+
 # Functions
 
 # Import data ----
-plot_data <- read.csv("data/plot_data_fil_agg_norm_std.csv")
+plot_data <- read.csv(paste0("data/plot_data_fil_agg", ext, "_norm_std.csv"))
+
 plotcode_plot_group_lookup <- read.csv("data/plotcode_plot_group_lookup.csv")
 load("data/clean_input_data.Rdata")
 
@@ -36,8 +39,8 @@ s_fil <- s %>%
 # Create community composition matrix
 stem_ab_mat <- s_fil %>%
   group_by(plot_group, gen_sp, .drop = FALSE) %>%
-	tally() %>%
-	spread(gen_sp, n, fill = 0) %>%
+  tally() %>%
+  spread(gen_sp, n, fill = 0) %>%
   ungroup() %>%
   data.frame() 
 
@@ -67,23 +70,23 @@ NMDS.scree <- function(x) { #where x is the name of the data frame variable
 #NMDS.scree(stem_ab_mat_dist)
 
 # Run NMDS
-#nmds_stems <- metaMDS(stem_ab_mat, 
+#nmds_stems <- metaMDS(stem_ab_mat,
 #  try = 20)
 
-# Re-run NMDS with noshare and more dimensions
-#nmds_stems_2 <- metaMDS(stem_ab_mat,
-#  try = 50,
-#  k = 3,
-#  noshare = 0.1, 
-#  previous.best = nmds_stems)
+#saveRDS(nmds_stems, file = "data/nmds_stems.rds")
 
 # Identify outlier threshold for ~5% of plots
-outlier_list <- list()
+# outlier_list <- list()
+# 
+# for(i in 1:100){
+#   x <- seq(from = 0.005, to = 0.5, by = 0.005)[i]
+#   outlier_list[[i]] <- outlier(stem_ab_mat, thresh = x, y = 0.5)
+# }
+# 
+# # Save to data file
+# saveRDS(outlier_list, file = paste0("data/outlier_list", ext, ".rds"))
 
-for(i in 1:100){
-  x <- seq(from = 0.005, to = 0.5, by = 0.005)[i]
-  outlier_list[[i]] <- outlier(stem_ab_mat, thresh = x, y = 0.5)
-}
+outlier_list <- readRDS(paste0("data/outlier_list", ext, ".rds"))
 
[…]

### 2019-10-29 analysis/full_best.R

@@ -1,0 +1,7 @@
+full_best <- "full"
+
+if(full_best == "full"){
+  ext <- ""
+}else{
+  ext <- "_best"
+}

### 2019-10-29 analysis/lmm.R

@@ -26,20 +26,22 @@ source("clust_pal.R")
 # Aggregated and filtered data
 plot_data <- read.csv("data/plot_data_fil_agg_norm_std.csv")
 
+plot_data$clust5 <- as.character(plot_data$clust5)
+
 # Mixed models ----
 # Model specification
 model_list <- list(
   model_precip = lmer(bchave_log ~ total_precip_std + (1|clust5), data = plot_data),
-  model_precip_s = lmer(bchave_log ~ precip_seasonality_std + (1|clust5), data = plot_data),
+  model_precip_s = lmer(bchave_log ~ precip_seasonality_log_std + (1|clust5), data = plot_data),
   
   model_temp = lmer(bchave_log ~ mean_temp_std + (1|clust5), data = plot_data),
-  model_temp_s = lmer(bchave_log ~ temp_seasonality_std + (1|clust5), data = plot_data),
+  model_temp_s = lmer(bchave_log ~ temp_seasonality_log_std + (1|clust5), data = plot_data),
   
   model_fire_return = lmer(bchave_log ~ fire_return_mean_log_std + (1|clust5), data = plot_data),
 
-  model_spr = lmer(bchave_log ~ sp_rich_raref_std + (1|clust5), data = plot_data),
-  model_shannon = lmer(bchave_log ~ shannon_cube_std + (1|clust5), data = plot_data),
-  model_equit = lmer(bchave_log ~ shannon_equit_std + (1|clust5), data = plot_data),
+  model_spr = lmer(bchave_log ~ sp_rich_raref_log_std + (1|clust5), data = plot_data),
+  model_shannon = lmer(bchave_log ~ shannon_log_std + (1|clust5), data = plot_data),
+  model_equit = lmer(bchave_log ~ shannon_equit_log_std + (1|clust5), data = plot_data),
   model_stem_n = lmer(bchave_log ~ stems_ha_log_std + (1|clust5), data = plot_data),
   
   model_cov_height = lmer(bchave_log ~ cov_height_std + (1|clust5), data = plot_data),

### 2019-10-29 analysis/plot_data_descrip.R

@@ -24,8 +24,6 @@ library(ggnewscale)
 # Plot data
 plot_data_final <- read.csv("data/plot_data_fil_agg_norm_std_outlier.csv")
 
-plot_data_final$fire_index <- factor(plot_data_final$fire_index, levels = c("Frequent", "Occassional", "Rare", "Very rare", "No fire"))
-
 # Regional tmeperature and precipitation data for the Miombo woodlands
 t_p <- read.csv("data/region_temp_precip.csv")
 
@@ -63,7 +61,7 @@ plot_map <- ggplot() +
     data = white_veg_miombo, alpha = 1) +
   # geom_point(data = ssaw8$plotInfoFull, aes(x = longitude_of_centre, y = latitude_of_centre), alpha = 0.6) + 
     geom_point(data = plot_data_final, 
-      aes(x = longitude_of_centre, y = latitude_of_centre, fill = as.character(clust5)), 
+      aes(x = longitude_of_centre, y = latitude_of_centre, fill = paste0("C", as.character(clust5))), 
     colour = "black", shape = 21, size = 2, alpha = 1) +
   map_africa_colour +
   scale_fill_manual(name = "Cluster", values = clust_pal) + 
@@ -115,15 +113,8 @@ var_map <- function(x, var, lab = expression("log(AGB) (t ha"^-1*")")){
 # Biomass
 pdf(file = "img/biomass_map.pdf", width = 5, height = 8)
 var_map(plot_data_final, 
-  var = plot_data_final$bchave_log, 
+  var = plot_data_final$bchave, 
   lab = expression("log(AGB) (t ha"^-1*")"))
-dev.off()
-
-# Aridity
-pdf(file = "img/aridity_map.pdf", width = 5, height = 8)
-var_map(plot_data_final, 
-  var = plot_data_final$aridity_index, 
-  lab = expression("Ardity" ~ "Index"))
 dev.off()
 
 # Precipitation
@@ -150,22 +141,15 @@ dev.off()
 # Shannon Diversity index
 pdf(file = "img/shannon_map.pdf", width = 5, height = 8)
 var_map(plot_data_final,
-  var = plot_data_final$shannon_cube,
+  var = plot_data_final$shannon_log,
   lab = expression("Shannon" ~ "index" ~ (H)))
 dev.off()
 
-# Fire index
+# Fire return
 pdf(file = "img/fire_index_map.pdf", width = 5, height = 8)
-ggplot() + 
-  map_africa + 
[…]

### 2019-10-29 analysis/sem.R

@@ -35,43 +35,38 @@ library(semTools)
 library(ggcorrplot)
 library(lme4)
 library(stargazer)
+library(tidyr)
 
 source("clust_pal.R")
+source("full_best.R")
 
 # Import data ----
 
 # Aggregated and filtered data
-sem_data <- read.csv("data/plot_data_fil_agg_norm_std_outlier.csv")
 
-# Recode fire index
-sem_data<- sem_data %>%
-  mutate(fire_index = case_when(
-    fire_index == "Frequent" ~ 5,
-    fire_index == "Occassional" ~ 4,
-    fire_index == "Rare" ~ 3,
-    fire_index == "Very rare" ~ 2,
-    fire_index == "No fire" ~ 1
-  ))
-  
+sem_data <- read.csv(paste0("data/plot_data_fil_agg", ext, "_norm_std_outlier.csv"))
+
 # Correlation matrix between variables ----
 # Create a dataframe 
 corr_df <- sem_data %>%
   select(sand_per_std, ocdens_std, cation_ex_cap_std, 
-    total_precip_std, precip_seasonality_std, 
-    mean_temp_std, temp_seasonality_std, isothermality_std,
-    fire_return_mean_log_std, sp_rich_raref_std, shannon_equit_std,
+    total_precip_std, precip_seasonality_log_std, 
+    mean_temp_std, temp_seasonality_log_std, isothermality_std,
+    fire_return_mean_log_std, 
+    sp_rich_raref_log_std, shannon_equit_log_std, shannon_log_std,
     cov_height_std, cov_dbh_std, bchave_log) %>%
   rename(`Sand %` = sand_per_std,
     `Organic C %` = ocdens_std,
     `CEC` = cation_ex_cap_std,
     `MAP` = total_precip_std, 
-    `MAP CoV` = precip_seasonality_std, 
+    `MAP CoV` = precip_seasonality_log_std, 
     `MAT` = mean_temp_std, 
-    `MAT CoV` = temp_seasonality_std,
+    `MAT CoV` = temp_seasonality_log_std,
     `Isothermality` = isothermality_std,
     `Fire return interval` = fire_return_mean_log_std, 
-    `Raref. Sp. Rich.` = sp_rich_raref_std, 
-    `Shannon equit.` = shannon_equit_std,
+    `Raref. Sp. Rich.` = sp_rich_raref_log_std, 
+    `Shannon equit.` = shannon_equit_log_std,
[…]

### 2019-11-05 analysis/output/include/clust_summ.tex
Updated figures and text, pre-comments from Kyle

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Oct 24, 2019 - 13:20:25
+% Date and time: Mon, Nov 04, 2019 - 14:52:48
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 16+12.6 & 281+373.8 & 34.7+37.1 \\ 
-C2 & Julbernardia.paniculata, Isoberlinia.angolensis, Albizia.antunesiana & 22+13.7 & 274+246.9 & 46.7+35.07 \\ 
-C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 8+6.5 & 272+535.3 & 36+27.93 \\ 
-C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 15+9.2 & 448+391.9 & 46.6+45.41 \\ 
-C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 9+9.4 & 336+434.3 & 43.8+34.45 \\ 
+C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 14+12.2 & 278+348.1 & 35.4+37.21 \\ 
+C2 & Julbernardia paniculata, Isoberlinia angolensis, Albizia antunesiana & 20+14.1 & 274+245.9 & 46.8+35.05 \\ 
+C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 7+6.1 & 274+536.5 & 36.3+27.86 \\ 
+C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 14+9.5 & 447+386.3 & 47+45.41 \\ 
+C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 8+9.1 & 322+390.1 & 43.8+34.45 \\ 
 \hline \\[-1.8ex] 

### 2019-11-05 manuscript/chapter_1.tex
Updated figures and text, pre-comments from Kyle

@@
 
+\input{include/n_plots}
+
 \begin{document}
@@
 A number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Review}.
-The strength and direction of these observed relationships varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}.
-Over the past decade, many observational studies of the Biodiversity-Ecosystem Function Relationship (BEFR) have been conducted in tropical and temperate forests, and grasslands \citep{Chen2011}.
-The representation of dry tropical ecosystems in the BEFR literature however, is poor.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the highly endemic wet tropical forests Costa Rica.
+Ecosystem functions can be defined in broad terms as rate processes and properties which describe the degree of biotic activity within an ecosystem \citep{Jax2005}.
+This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect aggregate measures of function such as resistance of productivity to disturbance \citep{}.
+Ecosystem functionality may also include static ecosystem properties such as forest canopy complexity and trophic complexity, which in turn influence ecosystem processes.
+The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}.
+Over the past decade, many observational studies of the BEFR have been conducted, mostly in tropical and temperate forests, and grasslands \citep{Chen2011}.
+These studies support previous findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+
+The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the highly endemic wet tropical forests of Costa Rica.
+In wet tropical forests, much of the observed effect of biodiversity on ecosystem function has been attributed to niche complementarity \citep{}, whereby competition between individuals is reduced due to them occupying different niche space, allowing diverse species assemblages to take greater advantage of natural resources \citep{}.
+In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as much of a role in the relationship between biodiversity and ecosystem function.
+Instead, stress tolerance and productivity of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{}.
+Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{}.
 
 In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity.
-Resource extraction directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}.
+This has in turn raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}.
+Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}.
 Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality \citep{}.
-Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning \textgreater{}4 million km\textsuperscript{2}.
+Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016}.
 The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and are of global importance to the carbon cycle.
 Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}.
-Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are \textgreater{}300 tree species \citep{Frost1996}.
+Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}.
 These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}.
-Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the banner of REDD+ \citep{Hinsley2015}.
-A small number of academic studies in these woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2017, and, others}, but all have been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends cannot be reliably inferred.
-\citet{Fayolle2018}
+Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
+A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, and, others}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends could not be reliably inferred.
 
 The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}.
-Climatic variation is known to affect woody biomass \citep{Michaletz2014}, so it is important to account for these factors and acknowledge their interaction with both woody species diversity \textit{and} biomass stocks to effectively model the effects of biodiversity on biomass stocks in a regional analysis.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in these ecosystems, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
+Climatic variation is known to affect both woody biomass \citep{Michaletz2014} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in dry woodlands, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
 However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.
@@
 
+% Kunz et al.
+(2019) found that crown complementarity and crown plasticity increased with species richness.
+Trees growing in species rich neighbourhoods had enhanced AGB production in trunk and branches.
+
 \todo{Species abundance evenness and selection effects.}
@@
 
-High levels of disturbance in the form of seasonal fires and herbivory from large mammals \citep{Fire is a herbivore} may moderate the strength of observable BEFRs in southern African woodlands.
+High levels of disturbance in the form of seasonal fires and herbivory from large mammals \citep{Fireherbivore} may moderate the strength of observable BEFRs in southern African woodlands.
 Fire disturbance in forests has been linked to abundance dependent mortality among smaller stems \citep{Roques2001}.
@@
 
-In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \ref{}).
-We used aboveground woody biomass \todo{and estimates of ecosystem productivity} as measures of ecosystem functionality to understand the co-relationship between tree species biodiversity and ecosystem functionality.
+In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{}).
+We investigated the relationship between aboveground woody biomass and tree biodiversity.
 We compared the relative effects of tree species biodiversity with that of other environmental factors known to affect ecosystem productivity and biomass accumulation such as precipitation and temperature.
-We incorporated biogeographical clusters \citep{} into our analyses to understand how species composition as well as species biodiversity \textit{per se} affected ecosystem functionality.
-Initially, we made four hypotheses: 1) across the region, after other environmental factors had been accounted for, woodland plots with a higher species richness would have a higher aboveground biomass, 2) increased aridity would increase the strength of the biomass tree species richness relationship, 3) species composition would have a greater affect on aboveground biomass than species diversity \textit{per se}.
-We used Structural Equation Modelling as a preferred method to simultaneously account for environmental factors and biotic factors, which often interact in their effect on ecosystem structure and therefore biomass \citep{}.
+We also investigated potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass.
+We incorporated vegetation type clusters \citep{} into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
+Initially, we made four hypotheses:
 
@@
 \begin{enumerate}
-	\item{Plots with a higher tree species diversity will maintain higher biomass stocks.}
-	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, due to abiotic facilitation effects, despite lower overall tree species richness.}
-	\item{Structural characteristics of the woodland will interact with species composition and richness to provide an indirect path of influence between species composition and biomass stocks.}
+	\item{Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.}
+	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above-ground woody biomass.} 
+	\item{There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species and structural tree diversity.}
+	\item{Structural characteristics of the woodland will interact with species diversity to provide an indirect path of influence between species composition and biomass stocks.}
+	\item{The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.}
 \end{enumerate}
+
+To address these hypotheses we used Structural Equation Models with path analysis as a preferred method to simultaneously account for environmental factors and biotic factors, which may interact in their effect on ecosystem structure and therefore biomass.
 
@@
 
-The study used 1066 woodland monitoring from a larger network located across 10 countries within southern Africa in so-called Miombo woodland (Figure \ref{plot_loc}).
-The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in the north of Zambia, Malawi and northern Mozambique.
-The study sites straddle the climate space of the region as a whole (Figure \ref{temp_precip_hull}a).
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers 92.7\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 seconds (0.86 km\textsuperscript{2} at the equator).
+The study used \nplots{} woodland monitoring plots from a larger network located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1983}).
+The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
+The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}).
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers 92.7\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 seconds (0.86 km\textsuperscript{2} at the equator) \citep{worldclim}.
 
@@
 	\centering
-	\subfloat[]{{\includegraphics[width=0.4\textwidth]{plot_loc}}}%
+	\subfloat[]{{\includegraphics[width=0.3\textwidth]{plot_loc}}\label{plot_loc}}%
     \qquad
-    \subfloat[]{{\includegraphics[width=0.4\textwidth]{temp_precip_hull}}}%
-    \caption{The locations of the 1066 plots used in this study, by geographic location (a) with respect to the distribution of Miombo woodland vegetation \citep{White}, and in climate space (b), showing the plot locations as points compared to the climate space of the whole region as estimated using the WorldClim dataset over the Miombo woodland vegetation extent.}
+\subfloat[]{{\includegraphics[width=0.4\textwidth]{temp_precip_hull}}\label{temp_precip_hull}}%
+\caption{The locations of the \nplots{} plots used in this study, by geographic location (a) with respect to the distribution of Miombo woodland vegetation according to \citet{White}, and in climate space (b), showing the plot locations as points compared to the climate space of the whole region as estimated using the WorldClim dataset over the Miombo woodland vegetation extent \citep{worldclim}.
+Note that the density colour scale is log-transformed.}
 \end{figure}
@@
 Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup.
-Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna''.
-Plots with evidence of farming, human resource extraction or experimental treatements such as prescribed burning or herbivore exclusion were exlcuded from the initial pool.
-Only plots \textgreater{}0.1 hectares were used as per hectare biomass estimation from small plots is highly influenced by rare large trees.
-Only plots with a stem density \textgreater{}10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
+Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature \citep{}.
+Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were exlcuded from the initial pool.
+Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{}, leading to inaccurate estimates.
+Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
 
 Many plots provided by the Zambian Forestry Commission were arranged in clusters of up to four 20x50 m plots, 20 metres apart.
-These plots were aggregated for analyses and their values averaged for use in further plot level data cleaning.
+Plots within a cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.
 
-A Principle Coordinate Analysis (PCoA) was conducted on tree species abundance matrix data for each suitable plot to identify compositional outliers, which were considered not representative southern African woodlands.
-Plots within the upper 95th percentile of neighbour distance in the PCoA were excluded from further analyses (Figure \ref{pcoa_outlier}).
+After the initial plot data cleaning described above, a Principle Coordinate Analysis (PCoA) was conducted on tree species abundance matrix data for each plot to identify compositional outliers, which were considered not representative southern African woodlands.
+Plots within the upper 95th percentile of neighbour distance in the PCoA were excluded from further analyses (\autoref{pcoa_outlier}).
 This procedure removed 66 plots.
@@
 	\includegraphics[width=0.5\textwidth]{pcoa_outlier}
-	\caption{Principle Co-ordinate Analysis of tree species diversity of the 1122 plots chosen from the larger pool of 5395 plots.
+	\caption{Principle Co-ordinate Analysis of tree species diversity of the \nplotspcoa{} plots chosen from the larger pool of 5395 plots.
 The PCoA is constructed from a tree species abundance matrix.
-Plots which are within the upper 95th percentile of plot neighbour euclidean distance in the PCoA are marked in red, to be excluded from further analysis.}
+Plots which are within the upper 95th percentile of plot neighbour euclidean distance in the PCoA are marked in red, and were excluded from further analysis.}
 	\label{pcoa_outlier}
@@
 
-We considered only trees and shrubs in our analyses of above-ground biomass, including woody species such as palms and cycads which are functionally tree-like but excluding lianas which fill a different ecological niche.
-Only stems \textgreater{}5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
-AGB estimation for stems \textless{}5 cm DBH is imprecise due to its sensitivity to many environmental and historical factors \citep{}.
+We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{}.
+Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
+AGB estimation for stems <5 cm DBH is imprecise due to its sensitivity to many environmental and historical factors \citep{}.
 Additionally, for some species, species identification is difficult for saplings.
 
-All stems \textgreater{}5 cm DBH were measured within each plot resulting in a total of 160,076 stems with measurements.
-Note that a tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
-For each stem, species, DBH and height to the highest branch material were recorded.
+All stems >5 cm DBH were measured within each plot resulting in a total of 160,076 stems with measurements.
+A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
+For each stem, species, DBH and tree height to the top of the highest branch material were recorded.
 Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which may cause potentially large measurement uncertainty in the dataset.
@@
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
-The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial, parameters estimated using a test dataset \citep{Ryanunbublished}.
+The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from \citet{Ryanunbublished}.
 
-Aboveground biomass for each plot was calculated using Equation \ref{chave_agb}, taken from \citet{Chave2014}.
+Aboveground biomass for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}.
 Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
@@
 Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}.
-We extracted Cation Exchange Capacity (CEC), Organic Carbon Density, and the percentage sand content of the soil.
+We extracted Cation Exchange Capacity (CEC), organic carbon density, and the percentage sand content of the soil.
 These data are a modelled product compiled from various remote sensed and directly measured data sources.
 
-\todo{Fire return interval ($F_{ret}$) was calculated using the MODIS burned area product V6 (MCD46A1) \citep{}.
+% \todo{Fire return interval ($F_{ret}$) was calculated using the MODIS burned area product V6 (MCD46A1) \citep{}.
 Data was downloaded from January 2000 to December 2018.
 Mean fire return interval was calculated as:}
-
-\begin{equation}
-	\todo{F_{ret} = \frac{\sum_{i = 1}^{n} t_{i} - t_{i-1}}{n-1}}
-\end{equation}
-
-\todo{Where $t_{i}$ is the date of fire $i$, and $n$ is the total number of fires in the time period.}
+% 
+% \begin{equation}
+% 	\todo{F_{ret} = \frac{\sum_{i = 1}^{n} t_{i} - t_{i-1}}{n-1}}
+% \end{equation}
+% 
+% \todo{Where $t_{i}$ is the date of fire $i$, and $n$ is the total number of fires in the time period.}
 
@@
 Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}.
-This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richnes value.
-These estimated species richness values are referred to as rarefied species richness throughout the paper.
-Rarefied species richness accounted for variation in plot size (0.1-10 ha) and therefore sampling effort across the region.
+This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
+Rarefied species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort across the region.
 Larger plots will tend to encompass more individuals, and therefore probabilistically more species, according to the species-area relationship \citep{}.
 
-To assess tree species abundance evenness, the Shannon Equitability index \citep{} (Equation \ref{shannon_equit}) was calculated:
+To measure tree species abundance evenness, the Shannon Equitability index \citep{} (\autoref{shannon_equit}) was calculated:
 
@@
 
-Where $H'_{e}$ is an estimation of the Shannon diversity index by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the total number of species in the plot.
+Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the rarefied and extrapolated number of tree species in the plot.
 
-\subsubsection{Biogeographical clusters}
+\subsubsection{Vegetation clusters}
 
-Plots were assigned to biogeographical groups based on tree species composition.
+Plots were assigned to vegetation type groups based on tree species composition.
 Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data.
 Groups were identified using unconstrained correspondence analysis and ordination.
-Plot data used in this study occurred in five biogeographical groups, see Table \ref{clust_test} for a description of each biogeographical cluster and Map \ref{clust_map} for the spatial distribution of plots from each of these clusters.
+Plot data used in this study occurred in five vegetation type groups.
+See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .
 
@@
 \centering
-	\includegraphics[width=0.6\textwidth]{clust_map}
-	\caption{}
+	\includegraphics[width=0.8\textwidth]{clust_map}
+	\caption{The spatial distribution of plots according to vegetation cluster within southern Africa.}
 	\label{clust_map}
@@
 
-Structural Equation Models (SEM) investigated the determinants of woody Above-Ground Biomass (AGB).
+Structural Equation Models (SEM) investigated the determinants of AGB.
 All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R}.
@@
 Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset.
-For example, moisture availability is known to affect productivity \citep{}, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation and the seasonality of that precipitation.
-Indepedently of total precipitation, precipitation seasonality, measured as the covariance of daily precipitation, determines whether water arrives uniformly over a given time period or as few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
+For example, moisture availability is known to affect productivity \citep{}, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration.
+Independently of total precipitation, precipitation seasonality, measured as the covariance of daily precipitation, determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
 
-We specified a conceptual model with latent variables based on theoretical unmeasured factors known to affect AGB: moisture availability, temperature suitability for plant growth, soil fertility, fire regime intensity, tree species diversity and tree structural diversity (Figure \ref{con_mod}).
-\todo{Although the fire regime composite only had a single indicator variable, it was nevertheless included in a latent variable to soak up measurement error \citep{}.}
+Prior to model fitting, we specified a conceptual model with latent variables based on unmeasured factors known to affect AGB: moisture availability, soil fertility, tree species diversity and tree structural diversity (\autoref{con_mod}).
 
@@
 	\includegraphics[width=0.8\textwidth]{con_mod}
-	\caption{Conceptual model showing the theoretical relationships between environmental factors, structural and species tree diversity and AGB.
+	\caption{Conceptual model showing the theoretical relationships between environmental factors, structural and species tree diversity, and AGB.
 Hypothesised paths of causal inference are depicted as arrows from predictor to response.
-The shaded box defines the model which includes only tree diversity variables.}
+Latent variables are represented as shaded boxes with their constituent observed variables within.
+The names of each latent variable are in ovals.
+The pale orange dashed box defines a nested model which includes only tree diversity variables.}
 	\label{con_mod}
@@
 
-Observed variables were standardized to Z-scores for analysis.
+Observed variables were standardized to Z-scores for the SEMs.
 Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
-It allows path regression coefficients to be easily compared between paths to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}.
+Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}.
 Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}.
 
-The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention \citep{}, with other observed variables being allowed to vary.
-Due to some latent variables being regressed against both structural diversity and AGB, exact factor loadings from multiple regressions against either of these response variables could not be used to inform factor loadings in the SEM and were therefore estimated by the SEM model.
+The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{}.
+While it is recommended by some to set exact factor loadings in the SEM from the regression coefficients of multiple regressions \citep{}, because some latent variables were regressed against both structural diversity and AGB, exact factor loadings from simple multiple regressions could not be used.
+We therefore allowed factor loadings to be estimated by the SEM itself.
 
-For each biogeographical cluster separately, and at the regional scale, we employed SEM to assess the relative importance of each latent variable in determining AGB.
-We assessed the indirect effects of abiotic environment on AGB via the mediators of tree species diversity and tree structural diversity to understand the interactive nature of tree diversity and abiotic environment in determining AGB.
+First, we assessed the interacting factors of structural diversity and species diversity in determining AGB.
+We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity (see orange box \autoref{con_mod}).
+To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately.
+We compared path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between structural diversity, species diversity and AGB.
 Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
-We follow the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI \textgreater{} 0.95, TLI \textgreater{} 0.95, RMSEA \textless{} 0.06.
+We follow the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.95, TLI >0.95, RMSEA <0.06.
 
-Two sets of SEMs were fitted (Figure \ref{con_mod}).
-The first used only tree diversity latent variables to estimate diversity, with the indirect effect of species diversity on AGB via structural diversity being explicitly modelled as a mediation effect to examine the strength and form of this relationship.
-The second incorporated environmental covariates to understand the relative effects of climate, soils, disturbance on diversity and AGB.
+Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability, and soil fertility, on AGB both directly and via species diversity.
+
+For models of the whole region which included all suitable plots, standardised path coefficients are reported in order to better compare the relative contribution of different paths within the model.
+When comparing the path coefficients of models using individual vegetation clusters, unstandardised path coefficients are used in order to more readily compare the same path coefficient across different models while accounting for variation in mean values between biogeographical clusters.
 Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Kline2005}.
 
-For models on the whole region, standardized path coefficients are reported in order to better compare the relative contribution of different paths within the model.
-When comparing the path coefficients of models using individual biogeographical clusters, unstandardized path coefficients are used in order to more readily compare the same path coefficient across different models while accounting for variation in mean values between biogeographical clusters.
+To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create 900 datasets of similar size with varying median stem density.
+We used each of these datasets to fit the model including only tree species and structural diversity latent variables to predict AGB (see orange in \autoref{con_mod}).
+We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of the dataset.
+
+We fitted separate moderation models to investigate whether there was an interaction effect whereby the strength of the relationship between species diversity and AGB was influenced by moisture availability and soil fertility.
+The \verb|lavaan| R package does not natively support moderation of latent variables in its model specification.
+Instead we manually calculated interaction variables for both soil fertility and moisture availability from the product of predicted values of these latent variables in a Confirmatory Factor Analysis (CFA).
+Interaction variables were the products: species diversity $\times$ moisture availability, and species diversity $\times$ soil fertility.
+These interaction variables were included as explanatory variables in multiple regressions alongside the latent variable of species diversity to predict AGB.
+Regression coefficients and model fit were analysed to determine the presence of interaction effects.
 
@@
 	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlogram of observed exogenous variables used in the SEMs, with correlation coefficients shaded according to sign (+ve red, -ve blue) and strength of correlation.}
+	\caption{Correlogram of standardised observed variables used in the SEMs, with correlation coefficients shaded according to sign ($+$ve red, $-$ve blue) and strength of correlation.}
 	\label{corr_mat}
@@
 
-At the regional level, tree species and structural diversity observed variables had weak positive correlations with AGB (Figure \ref{corr_mat}), while environmental observed variables had very weak correlations with AGB.
+At the regional level, tree species diversity and structural diversity observed variables all had moderate positive correlations with AGB, while environmental observed variables had weak correlations with AGB.
+Only MAP and soil fertility observed variables (CEC, organic C \%, sand \%) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero.
+Stronger correlations were seen between certain environmental covariates and measures of tree species diversity, namely, MAP, MAT, soil organic C \%, and soil sand \% (\autoref{corr_mat}).
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effect of stand structural diversity (Figure \ref{sem_struc_path}, species diversity had a positive effect on AGB when the indirect path via stand structural diversity was accounted for (Figure \ref{struc_model_slopes}).
-Stand structural diversity had a positive effect on AGB (\todo{Test stats}).
-The indirect effect of diversity on biomass via structural diversity was small but significant ().
-Model fit for this model was good \todo{MORE}.
+\subsection{Structural diversity and species diversity models}
 
-When the model was run again using only data
+In an SEM describing the effect of tree species diversity on AGB via the mediating effect of stand structural diversity (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}).
+Stand structural diversity had a positive direct effect on AGB independent of species diversity (\todo{STATS}), but this effect was not as strong as the effect of species diversity itself.
+Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}).
+
+Model fit was also good when the model was refitted separately using data from each of the 5 vegetation clusters.
+It should be noted that model fit was not as good as when using regional level data, which may have arisen due to the small sample sizes available for some vegetation clusters, most notably C4, which only had 46 plots.
+While the strength of the path coefficients varied between vegetation clusters, the sign of the path coefficients remained positive and significant in all cases except for the direct effect of diversity on biomass in cluster 3 (\autoref{struc_model_slopes_all}).
 
@@
 \centering
-	\includegraphics[width=\textwidth]{sem_struc_path}
+	\includegraphics[width=\textwidth]{struc_mod}
 	\caption{Path diagram with regression coefficients for the structural model including plots in all five clusters.}
-	\label{sem_struc_path}
+	\label{struc_mod}
 \end{figure}
@@
 	\includegraphics[width=\textwidth]{struc_model_slopes_all}
-	\caption{Unstandardized path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
+	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
 Path coefficients are +/- 1 standard error.
@@
 
+\subsubsection{Mediation by stem density}
+
+We repeatedly sub-sampled the available plots to build many datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
+The same SEM specification was used as above, incorporating the effects of species diversity and structural diversity on AGB (\autoref{con_mod}).
+\autoref{sem_struc_stems_ha} shows that there is a peak in the effect of diversity on biomass at \textapprox{}765 stems ha\textsuperscript{-1}.
+There appears to be a minimum stem density threshold below which there appears to be no effect of diversity on biomass.
+This threshold is at \textapprox{}295 stems ha\textsuperscript{-1}.
+Interestingly, at high stem densities, the effect of species diversity on biomass declines.
+The effect of structural diversity on biomass appears to decrease with increasing stem density, though the relationship is less clear.
+
 \begin{figure}[H]
 \centering
-	\includegraphics[width=\textwidth]{sem_full_path}
+	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
+	\caption{Line charts showing the variation in path coefficients in the SEM, using datasets with different mean stem density.}
+	\label{sem_struc_stems_ha}
+\end{figure}
+
+\subsection{Models with environmental covariates}
+
+A model incorporating moisture availability and soil showed that at the regional level, the effect of diversity on biomass was greater than that of moisture and soil.
+Surprisingly, the effect of moisture availability on biomass was negligible, with all of the observed effect coming from the indirect path of moisture availability on biomass via species diversity.
+
+Soil fertility had a strong positive effect on species diversity. Soil fertility also had a positive effect on AGB.
+
+The effect of species diversity had the strongest total effect on biomass, with some indirect effect via structural diversity.
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{full_mod}
 	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters.}
-	\label{sem_full_path}
+	\label{full_mod}
 \end{figure}
@@
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{full_model_slopes_all}
-	\caption{Unstandardized path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots.
-Path coefficients are +/- 1 standard error.
-Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
-	\label{full_model_slopes_all}
-\end{figure}
-
 \input{include/full_model_fit_clust_stats}
@@
 
-\subsection{Variation between biogeographical clusters}
+\subsection{Interaction effects of moisture availability and soil}
+
+To further address our hypothesis (H\textsubscript{2}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit a separate model with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
+
+Moderating effects of moisture availability were not observed.
+
+Soil fertility had a small positive effect on the relationship between species diversity and biomass.
+As soil fertility increased, the strength of the relationship between species diversity and biomass also increased.
+This effect was small in comparison to the direct effect of soil fertility on AGB.
 
@@
 
-%\bibliography{/Users/johngodlee/google_drive/bib/lib}
+\bibliography{/Users/johngodlee/google_drive/bib/lib}
 
@@
 % Seidel 2019: Large trees tend to possess a greater structural complexity than small trees, not due to their size per se, but due to more complex architecture.
+
+% https://jslefche.github.io/sem_book/coefficients.html

### 2019-11-05 manuscript/include/clust_summ.tex
Updated figures and text, pre-comments from Kyle

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Oct 24, 2019 - 13:20:25
+% Date and time: Mon, Nov 04, 2019 - 14:52:48
 \begin{table}[!htbp] \centering 
-  \caption{} 
+  \caption{Description of the biogeographical clusters (C1-C5) to which each plot in the study was assigned.
+Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv}.
+Numeric values are $\pm$ 1 standard deviation.}
   \label{clust_summ} 
@@
 \hline \\[-1.8ex] 
-C1 & \begin{tabular}[l]{@{}l@{}l@{}} Diplorhynchus condylocarpon \\ Combretum spp. \\ Pseudolachnostylis maprouneifolia \end{tabular} & 16$\pm$12.6 & 281$\pm$373.8 & 34.7$\pm$37.1 \\ 
+C1 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Combretum }spp.
+\\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 14$\pm$12.2 & 278$\pm$348.1 & 35.4$\pm$37.21 \\
 \hline
-C2 & \begin{tabular}[l]{@{}l@{}l@{}} Julbernardia.paniculata \\ Isoberlinia.angolensis \\ Albizia.antunesiana \end{tabular} & 22$\pm$13.7 & 274$\pm$246.9 & 46.7$\pm$35.07 \\ 
+C2 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Albizia antunesiana} \end{tabular} & 20$\pm$14.1 & 274$\pm$245.9 & 46.8$\pm$35.05 \\
 \hline
-C3 & \begin{tabular}[l]{@{}l@{}l@{}} Burkea africana \\ Pterocarpus angolensis \\ Baikiaea plurijuga \end{tabular} & 8$\pm$6.5 & 272$\pm$535.3 & 36$\pm$27.93 \\ 
+C3 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \\ \textit{Baikiaea plurijuga} \end{tabular} & 7$\pm$6.1 & 274$\pm$536.5 & 36.3$\pm$27.86 \\ 
 \hline
-C4 & \begin{tabular}[l]{@{}l@{}l@{}} Baikiaea plurijuga \\ Terminalia randii \\ Albizia amara \end{tabular} & 15$\pm$9.2 & 448$\pm$391.9 & 46.6$\pm$45.41 \\ 
+C4 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Baikiaea plurijuga} \\ \textit{Terminalia randii} \\ \textit{Albizia amara} \end{tabular} & 14$\pm$9.5 & 447$\pm$386.3 & 47$\pm$45.41 \\ 
 \hline
-C5 & \begin{tabular}[l]{@{}l@{}l@{}} Colophospermum mopane \\ Pseudolachnostylis maprouneifolia \end{tabular} & 9$\pm$9.4 & 336$\pm$434.3 & 43.8$\pm$34.45 \\ 
+C5 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Colophospermum mopane} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 8$\pm$9.1 & 322$\pm$390.1 & 43.8$\pm$34.45 \\ 
 \hline \\[-1.8ex] 

### 2019-11-05 analysis/img_transfer.sh

@@ -26,8 +26,19 @@ cp img/biomass_clim_lm_clust.pdf $OUT
 
 cp img/clust_map.pdf $OUT
 
+cp img/struc_mod.pdf $OUT
+
+cp img/full_mod.pdf $OUT
+
+cp img/sem_struc_stems_ha.pdf $OUT
+
+cp sem_path_diag/con_mod.pdf $OUT
+
+cp output/include/n_plots.tex $INC
+
 # Transfer tables from analysis to manuscript
 
+##  Structural model fit
 cp output/include/struc_model_fit_clust_stats.tex $INC
 
 sed -i 's/\$//g' "${INC}struc_model_fit_clust_stats.tex"
@@ -37,6 +48,7 @@ sed -i 's/\\extracolsep{5pt}/\\extracolsep{0pt}/g' "${INC}struc_model_fit_clust_
 sed -i "10s/.*/${STRUC_HEAD}/" "${INC}struc_model_fit_clust_stats.tex"
 
 
+## Full model fit
 cp output/include/full_model_fit_clust_stats.tex $INC
 
 sed -i 's/\$//g' "${INC}full_model_fit_clust_stats.tex"
@@ -45,6 +57,8 @@ sed -i 's/\\extracolsep{5pt}/\\extracolsep{0pt}/g' "${INC}full_model_fit_clust_s
 
 sed -i "10s/.*/${STRUC_HEAD}/" "${INC}full_model_fit_clust_stats.tex"
 
+
+## Biogeographical cluster description
 cp output/include/clust_summ.tex $INC
 
 sed -i 's/ccccc/clccc/g' "${INC}clust_summ.tex"
@@ -59,8 +73,12 @@ sed -i '12,16s/,/ \\\\/g' "${INC}clust_summ.tex"
 
 sed -i '12,16s/+/$\\pm$/g' "${INC}clust_summ.tex"
 
+sed -i -r '12,16s/\w+ \w+\.*/\\textit\{&\}/g' "${INC}clust_summ.tex"
+
+sed -i '12,16s/spp\.\}/\}spp./g' "${INC}clust_summ.tex"
+
 sed -i '10s/.*/{Cluster} \& {Indicator species} \& {Rarefied species richness} \& {Stems ha\\textsuperscript{-1}} \& {AGB (t ha\\textsuperscript{-1})} \\\\/' "${INC}clust_summ.tex"
 
 sed -i '13,16i\\\hline' "${INC}clust_summ.tex"
 
-
[…]

### 2019-11-13 analysis/output/include/clust_summ.tex
Made tikz path diagrams

--- 2019-10-29 analysis/output/include/clust_summ.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Oct 24, 2019 - 13:20:25
+% Date and time: Fri, Nov 08, 2019 - 11:40:45
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 16+12.6 & 281+373.8 & 34.7+37.1 \\ 
-C2 & Julbernardia.paniculata, Isoberlinia.angolensis, Albizia.antunesiana & 22+13.7 & 274+246.9 & 46.7+35.07 \\ 
-C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 8+6.5 & 272+535.3 & 36+27.93 \\ 
-C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 15+9.2 & 448+391.9 & 46.6+45.41 \\ 
-C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 9+9.4 & 336+434.3 & 43.8+34.45 \\ 
+C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 7+4.7 & 276+345.2 & 35.3+37.03 \\ 
+C2 & Julbernardia paniculata, Isoberlinia angolensis, Albizia antunesiana & 8+4.3 & 274+245.9 & 46.8+35.05 \\ 
+C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 4+2.3 & 274+536.5 & 36.3+27.86 \\ 
+C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 10+5.9 & 447+386.3 & 47+45.41 \\ 
+C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 4+2.9 & 322+390.1 & 43.8+34.45 \\ 
 \hline \\[-1.8ex] 

### 2019-11-13 manuscript/chapter_1.tex
Made tikz path diagrams

@@
 \usepackage{multirow}
+\usepackage{tikz}
+\usetikzlibrary{calc,arrows,positioning,shapes,shapes.gates.logic.US,trees}
 
-\title{A regional scale assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
+\title{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
 \author{John L. Godlee}
@@
 \input{include/n_plots}
+\input{include/n_outliers}
+\input{include/hull_cover}
+\input{include/path_coef_struc}
+\input{include/path_coef_full}
 
@@
 The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the highly endemic wet tropical forests of Costa Rica.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage.
 In wet tropical forests, much of the observed effect of biodiversity on ecosystem function has been attributed to niche complementarity \citep{}, whereby competition between individuals is reduced due to them occupying different niche space, allowing diverse species assemblages to take greater advantage of natural resources \citep{}.
@@
 
-In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity.
+\todo{Re-assess this paragraph -}In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity.
 This has in turn raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}.
@@
 A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, and, others}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends could not be reliably inferred.
+\todo{Say how the previous studies might have been wrong because of their limited scope, or how in order to really answer the question you need to sample long environmental gradients.}
 
 The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}.
-Climatic variation is known to affect both woody biomass \citep{Michaletz2014} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation sets the upper limit for woody cover in dry woodlands, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
+In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales in savannas.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
 However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.
@@
 
+\todo{Stand structural complexity}
+
 While southern African woodlands are species rich in the herbaceous understorey, the tree layer is relatively species poor \citep{}.
@@
 We also investigated potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type clusters \citep{} into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
+We incorporated vegetation type clusters based on tree species composition \citep{} into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
 Initially, we made four hypotheses:
@@
 	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above-ground woody biomass.} 
-	\item{There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species and structural tree diversity.}
+	\item{There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species diversity.}
 	\item{Structural characteristics of the woodland will interact with species diversity to provide an indirect path of influence between species composition and biomass stocks.}
@@
 The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}).
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers 92.7\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 seconds (0.86 km\textsuperscript{2} at the equator) \citep{worldclim}.
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{worldclim}.
 
@@
 
-After the initial plot data cleaning described above, a Principle Coordinate Analysis (PCoA) was conducted on tree species abundance matrix data for each plot to identify compositional outliers, which were considered not representative southern African woodlands.
-Plots within the upper 95th percentile of neighbour distance in the PCoA were excluded from further analyses (\autoref{pcoa_outlier}).
-This procedure removed 66 plots.
-
-\begin{figure}[H]
-\centering
-	\includegraphics[width=0.5\textwidth]{pcoa_outlier}
-	\caption{Principle Co-ordinate Analysis of tree species diversity of the \nplotspcoa{} plots chosen from the larger pool of 5395 plots.
-The PCoA is constructed from a tree species abundance matrix.
-Plots which are within the upper 95th percentile of plot neighbour euclidean distance in the PCoA are marked in red, and were excluded from further analysis.}
-	\label{pcoa_outlier}
-\end{figure}
+After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition.
+We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
+We set the threshold value to remove the top 5\% of plots with the highest nearest neighbour distances \citep{Otto}, removing \noutliers{} plots.
 
@@
 Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
-AGB estimation for stems <5 cm DBH is imprecise due to its sensitivity to many environmental and historical factors \citep{}.
-Additionally, for some species, species identification is difficult for saplings.
+Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems forming a very small proportion of the total AGB in a plot \citep{}.
 
@@
 For each stem, species, DBH and tree height to the top of the highest branch material were recorded.
-Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which may cause potentially large measurement uncertainty in the dataset.
-Averaged over such a large number of stems however, this should not meaningfully alter the results of our analysis.
+Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which may cause varying measurement uncertainty by data source in the dataset.
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
@@
 Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
-Temperature and precipitation seasonality were both calculated as the coefficient of variation of these variables across the 18 years of available data.
+Temperature and precipitation seasonality were both calculated as the coefficient of variation of the daily data across the 18 years of available data.
 Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}.
@@
 This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
-Rarefied species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort across the region.
-Larger plots will tend to encompass more individuals, and therefore probabilistically more species, according to the species-area relationship \citep{}.
+Extrapolated species richness was preferred over raw Hill numbers as they are more interpretable, representing actual species numbers.
+Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort across the region.
+Larger plots will tend to encompass more individuals, and therefore probabilistically more species \citep{}.
 
@@
 
-Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the rarefied and extrapolated number of tree species in the plot.
+Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot.
+
+We calculated tree stand structural diversity for each plot by calculating the coefficient of variation of DBH and tree height.
 
@@
 For example, moisture availability is known to affect productivity \citep{}, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration.
-Independently of total precipitation, precipitation seasonality, measured as the covariance of daily precipitation, determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
+Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
 
@@
 \centering
-	\includegraphics[width=0.8\textwidth]{con_mod}
+	\input{include/path_diagram_concept}
 	\caption{Conceptual model showing the theoretical relationships between environmental factors, structural and species tree diversity, and AGB.
@@
 Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}.
+To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1.
+For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed.
+Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way.
 
@@
 
-Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability, and soil fertility, on AGB both directly and via species diversity.
+Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density.
 
@@
 	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlogram of standardised observed variables used in the SEMs, with correlation coefficients shaded according to sign ($+$ve red, $-$ve blue) and strength of correlation.}
+	\caption{Correlogram of standardised observed variables used in the SEMs, with correlation coefficients shaded according to sign ($+$ve red, $-$ve blue) and strength of correlation.
+Variables in bold were later reversed for modelling to maintain positive correlations between observed variables within each latent variable.
+Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
+Colours of variable names group variables into latent variables: red = Soil fertility, blue = Moisture availability, green = species diversity, purple = structural diversity.}
 	\label{corr_mat}
@@
 
-At the regional level, tree species diversity and structural diversity observed variables all had moderate positive correlations with AGB, while environmental observed variables had weak correlations with AGB.
-Only MAP and soil fertility observed variables (CEC, organic C \%, sand \%) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero.
-Stronger correlations were seen between certain environmental covariates and measures of tree species diversity, namely, MAP, MAT, soil organic C \%, and soil sand \% (\autoref{corr_mat}).
+A simple correlation plot showed that all tree species diversity and structural diversity observed variables all had moderate positive correlations with AGB.
+Stem density had the strongest correlation with AGB of all variables ($\rho$ = 0.51).
+Environmental variables had weaker correlations with AGB than diversity variables.
+Only MAP ($\rho$ = 0.19), MAP seasonality ($\rho$ = -0.18), soil Organic C \% ($\rho$ = 0.19) and soil Sand \% ($\rho$ = -0.18) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero.
+(\autoref{corr_mat}).
 
-\subsection{Structural diversity and species diversity models}
+As expected, there was a positive correlation between MAP and AGB, and a negative correlation between the seasonality of precipitation and AGB.
+MAT and the seasonality of temperature variation did not correlate with AGB.
+As expected, there was a negative correlation between soil sand content and AGB, and a positive correlation between soil organic Carbon AGB.
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effect of stand structural diversity (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}).
-Stand structural diversity had a positive direct effect on AGB independent of species diversity (\todo{STATS}), but this effect was not as strong as the effect of species diversity itself.
+MAP had positive correlations with tree species diversity, tree height diversity and a positive but not significant correlation with tree stem density.
+MAT had non significant correlations with tree species and structural diversity variables.
+
+Tree species diversity variables had clear positive correlations with stem density (Species richness: $\rho$ = 0.45, Shannon equitability: $\rho$ = 0.43).
+
+\subsection{Structural and species diversity models}
+
+In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via stem density, but not via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}).
+Stand structural diversity had a positive direct effect on AGB independent of species diversity (\todo{STATS}).
 Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}).
 
-Model fit was also good when the model was refitted separately using data from each of the 5 vegetation clusters.
-It should be noted that model fit was not as good as when using regional level data, which may have arisen due to the small sample sizes available for some vegetation clusters, most notably C4, which only had 46 plots.
-While the strength of the path coefficients varied between vegetation clusters, the sign of the path coefficients remained positive and significant in all cases except for the direct effect of diversity on biomass in cluster 3 (\autoref{struc_model_slopes_all}).
+When the model was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardized path coefficients varied but generally similar relationships between diversity and AGB with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
+It should be noted that models for vegetation clusters 4 and 5 did not converge, possibly due to small sample sizes (\autoref{struc_model_slopes_all}).
 
@@
 \centering
-	\includegraphics[width=\textwidth]{struc_mod}
+	\input{include/path_diagram_struc}
 	\caption{Path diagram with regression coefficients for the structural model including plots in all five clusters.}
@@
 \end{figure}
+
+%Meaasurement errors of exogenous variables omitted for clarity
 
@@
 \end{figure}
+
+\subsection{Variation among vegetation clusters in structural and species diversity effects}
 
@@
 
-\subsubsection{Mediation by stem density}
+\subsubsection{Moderation of Diversity-AGB relationship by stem density}
 
@@
 	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line charts showing the variation in path coefficients in the SEM, using datasets with different mean stem density.}
+	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density.}
 	\label{sem_struc_stems_ha}
@@
 
-\subsection{Models with environmental covariates}
+\subsection{Mediation of environmental covariates via diversity}
 
@@
 \centering
-	\includegraphics[width=\textwidth]{full_mod}
+	\input{include/path_diagram_full}
 	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters.}
@@
 
-\input{include/full_model_fit_clust_stats}
+% \input{include/full_model_fit_clust_stats} - Excluded, low fit
 
-\subsection{Structural diversity}
+Vegetation cluster level models could not be reliably fitted for these more complex models with environmental covariates, due to sample size and convergence issus.
 
-The latent variable of tree species diversity had a positive effect on AGB across the region.
-A stronger positive effect was seen between structural diversity and aboveground biomass, with species diversity acting \textit{through} structural diversity to effect aboveground biomass.
-
-\subsection{Interaction effects of moisture availability and soil}
+\subsection{Moderation effects of moisture availability and soil}
 
@@
 
+subsection{Large trees}
+
+We analysed how the strength of the relationship between tree species richness and AGB varied for different size classes of tree.
+\autoref{bchave_quantile_slope_sp_rich} shows a saturating increase in the importance of species richness to determine AGB as the proportional size of trees increases.
+
 \section{Discussion}
 
-\bibliography{/Users/johngodlee/google_drive/bib/lib}
+\subsection{Effect of tree species diversity on AGB}
+
+\subsection{Structural diversity as a mechanism for the BEFR}
+
+\subsection{Moderation effects of moisture availability and soil fertility}
+
+\subsection{Vegetation type specific responses}
+
+\subsection{Stem density effects}
+
+\todo{Why is there a peak in the effect of species richness on AGB at a certain level of stem density?}
+
+% Seidel 2019: Large trees tend to possess a greater structural complexity than small trees, not due to their size per se, but due to more complex architecture.
+
+%\bibliography{/Users/johngodlee/google_drive/bib/lib}
 
@@
 
-% Seidel 2019: Large trees tend to possess a greater structural complexity than small trees, not due to their size per se, but due to more complex architecture.
-
 % https://jslefche.github.io/sem_book/coefficients.html

### 2019-11-13 manuscript/include/clust_summ.tex
Made tikz path diagrams

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Nov 04, 2019 - 14:52:48
+% Date and time: Fri, Nov 08, 2019 - 11:40:45
 \begin{table}[!htbp] \centering 
@@
 Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv}.
-Numeric values are $\pm$ 1 standard deviation.}
+Numeric values are means $\pm$ 1 standard deviation.}
   \label{clust_summ} 
@@
 \hline \\[-1.8ex] 
-{Cluster} & {Indicator species} & {Rarefied species richness} & {Stems ha\textsuperscript{-1}} & {AGB (t ha\textsuperscript{-1})} \\
+{Cluster} & {Indicator species} & {Species richness} & {Stems ha\textsuperscript{-1}} & {AGB (t ha\textsuperscript{-1})} \\
 \hline \\[-1.8ex] 
 C1 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Combretum }spp.
-\\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 14$\pm$12.2 & 278$\pm$348.1 & 35.4$\pm$37.21 \\
+\\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 7$\pm$4.7 & 276$\pm$345.2 & 35.3$\pm$37.03 \\
 \hline
-C2 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Albizia antunesiana} \end{tabular} & 20$\pm$14.1 & 274$\pm$245.9 & 46.8$\pm$35.05 \\
+C2 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Albizia antunesiana} \end{tabular} & 8$\pm$4.3 & 274$\pm$245.9 & 46.8$\pm$35.05 \\ 
 \hline
-C3 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \\ \textit{Baikiaea plurijuga} \end{tabular} & 7$\pm$6.1 & 274$\pm$536.5 & 36.3$\pm$27.86 \\ 
+C3 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \\ \textit{Baikiaea plurijuga} \end{tabular} & 4$\pm$2.3 & 274$\pm$536.5 & 36.3$\pm$27.86 \\ 
 \hline
-C4 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Baikiaea plurijuga} \\ \textit{Terminalia randii} \\ \textit{Albizia amara} \end{tabular} & 14$\pm$9.5 & 447$\pm$386.3 & 47$\pm$45.41 \\ 
+C4 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Baikiaea plurijuga} \\ \textit{Terminalia randii} \\ \textit{Albizia amara} \end{tabular} & 10$\pm$5.9 & 447$\pm$386.3 & 47$\pm$45.41 \\ 
 \hline
-C5 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Colophospermum mopane} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 8$\pm$9.1 & 322$\pm$390.1 & 43.8$\pm$34.45 \\ 
+C5 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Colophospermum mopane} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 4$\pm$2.9 & 322$\pm$390.1 & 43.8$\pm$34.45 \\ 
 \hline \\[-1.8ex] 

### 2019-11-13 manuscript/include/path_diagram_full.tex
Made tikz path diagrams

\begin{tikzpicture}[auto,scale=2, 
	observed/.style={rectangle,draw,thick,inner sep=5pt,minimum size=1cm, align=center},
	latent/.style={circle, draw, thick, inner sep=0pt, minimum size=2cm, align=center}, 
	path/.style={->, thick, >=stealth'},
	loading/.style={->, thick, dashed, >=stealth'}]

\tikzset{mystyle/.style={->,double=black}}
\node [latent] (m) {Moisture};
\node [latent] (d) [below = 1in of m] {Species\\Diversity}; 
\node [latent] (s) [below = 1in of d] {Soil}; 
\node [latent] (h) [right = 1in of d] {Struc.\\Diversity}; 
\node [observed] (b) [right = 1in of h] {AGB}; 
\node [observed] (i) [left = 1.5in of d] {Stem\\Density};

\coordinate[above= 0.7in of m] (momid);
\node [observed] (mp) [left = 0.8in of momid] {MAP};
\node [observed] (mpc) [left = 0.1in of momid] {MAP\\CoV};
\node [observed] (mt) [right = 0.1in of momid] {MAT};
\node [observed] (mtc) [right = 0.8in of momid] {MAT\\CoV};

\coordinate[below= 0.7in of s] (somid);
\node [observed] (sc) [left = 0.5in of somid] {SOC};
\node [observed] (ss) at (somid) {Sand};
\node [observed] (so) [right = 0.5in of somid] {CEC};

\coordinate[left= 0.3in of d] (domid);
\node [observed] (ds) [above = 0.4in of domid] {Sp.\\Rich.};
\node [observed] (de) [below = 0.42in of domid] {Shannon\\Equit.};

\coordinate[left= 0.25in of h] (homid); 
\node [observed] (hd) [above = 0.4in of homid] {DBH\\CoV};
\node [observed] (hh) [below = 0.4in of homid] {Height\\CoV};

\draw [path] (m) to node {\pcfmd{}} (d);
\draw [path] (s) to node[right] {\pcfsd{}} (d);
\draw [path] (d) to node[below] {\pcfdh{}} (h);
\draw [path] (h) to node {\pcfhb{}} (b);
\draw [path] (m) to node[above=0.1in, pos=0.5] {\pcfmi{}} (i);
\draw [path] (s) to node {\pcfsi{}} (i);
\draw [path] (d) to node {\pcfdi{}} (i);
\draw [path] (m) to node {\pcfmb{}} (b);
\draw [path] (s) to node[below=0.1in, pos=0.5] {\pcfsb{}} (b);

\draw [loading] (m) to node[left, pos=0.5] {\pcfmmp{}} (mp);
\draw [loading] (m) to node[left, pos=0.5] {\pcfmmpc{}} (mpc);
\draw [loading] (m) to node[right, pos=0.5] {\pcfmmt{}} (mt);
\draw [loading] (m) to node[right, pos=0.5] {\pcfmmtc{}} (mtc);

\draw [loading] (s) to node[left, pos=0.5] {\pcfssc{}} (sc);
\draw [loading] (s) to node[left, pos=0.5] {\pcfsss{}} (ss);
\draw [loading] (s) to node[right, pos=0.5] {\pcfsso{}} (so);

\draw [loading] (d) to node {\pcfdds{}} (ds);
\draw [loading] (d) to node[left, pos=0] {\pcfdde{}} (de);

\draw [loading] (h) to node[] {\pcfhhd{}} (hd);
\draw [loading] (h) to node {\pcfhhh{}} (hh);

\coordinate[below= 3in of i] (ibleft); 
\coordinate[below= 3in of b] (ibright); 
\draw [path] (i) -- (ibleft) to node[below] {\pcfib{}} (ibright) -- (b);

\coordinate[above= 1.3in of b] (dbright);
\coordinate[left= 2in of dbright] (dbleft);
\draw [path] (d) -- (dbleft) to node[above] {\pcfdb{}} (dbright) -- (b);

\end{tikzpicture}

### 2019-11-13 manuscript/include/path_diagram_struc.tex
Made tikz path diagrams

\begin{tikzpicture}[auto,scale=2, 
	observed/.style={rectangle,draw,thick,inner sep=5pt, outer sep=2pt, minimum size=1cm, align=center},
	latent/.style={circle, draw, thick, inner sep=0pt, outer sep=2pt, minimum size=2cm, align=center}, 
	path/.style={->, thick, >=stealth'},
	loading/.style={->, thick, dashed, >=stealth'}]

\tikzset{mystyle/.style={->,double=black}}
\node [latent] (d) {Species\\Diversity}; 
\node [latent] (h) [right = 1in of d] {Struc.\\Diversity}; 
\node [observed] (b) [below = 1in of h] {AGB}; 
\node [observed] (i) [below = 1in of d] {Stem\\Density};

\coordinate[above= 0.7in of d] (domid);
\coordinate[above= 0.7in of h] (homid);
\node [observed] (ds) [left = 0.2in of domid] {Sp.\\Rich.};
\node [observed] (de) [right = 0.2in of domid] {Shannon\\Equit.};
\node [observed] (hd) [left = 0.2in of homid] {DBH\\CoV};
\node [observed] (hh) [right = 0.2in of homid] {Height\\CoV};

\draw [loading] (d) to node[left, pos=0.5] {\pcsdsd{}} (ds);
\draw [loading] (d) to node[right, pos=0.5] {\pcsded{}} (de);
\draw [loading] (h) to node[left, pos=0.5]  {\pcshdh{}} (hd);
\draw [loading] (h) to node[right, pos=0.5] {\pcshhh{}} (hh);

\draw [path] (d) to node {\pcsdh{}} (h);
\draw [path] (h) to node {\pcshb{}} (b);
\draw [path] (d) to node {\pcsdi{}} (i);
\draw [path] (i) to node {\pcsib{}} (b);
\draw [path] (d) to node {\pcsdb{}} (b);

\end{tikzpicture}

### 2019-11-13 analysis/cluster_species.R

@@ -78,8 +78,8 @@ clust_summ <- plot_data %>%
     n_stems_ha = n_stems / area_of_plot) %>%
   summarise(bchave_mean = round(mean(bchave, na.rm = TRUE), digits = 1),
     bchave_sd = round(sd(bchave, na.rm = TRUE), digits=  2),
-    sp_rich_raref_mean = round(mean(sp_rich_raref, na.rm = TRUE), digits = 0),
-    sp_rich_raref_sd = round(sd(sp_rich_raref, na.rm = TRUE), digits = 1),
+    sp_rich_mean = round(mean(sp_rich, na.rm = TRUE), digits = 0),
+    sp_rich_sd = round(sd(sp_rich, na.rm = TRUE), digits = 1),
     n_stems_ha_mean = round(mean(n_stems_ha, na.rm = TRUE), digits = 0),
     n_stems_ha_sd = round(sd(n_stems_ha, na.rm = TRUE), digits = 1)) %>%
   mutate(c_name, c_ind,

### 2019-11-13 analysis/img_transfer.sh

@@ -6,6 +6,7 @@ STRUC_HEAD="{Cluster} \& {Params.} \& {n} \& {\$\\\\chi\^{2}\$} \& {DoF} \& {CFI
 
 # Transfer images from analysis to manuscript
 
+
 cp img/plot_loc.pdf $OUT
 
 cp img/temp_precip_hull.pdf $OUT
@@ -35,6 +36,22 @@ cp img/sem_struc_stems_ha.pdf $OUT
 cp sem_path_diag/con_mod.pdf $OUT
 
 cp output/include/n_plots.tex $INC
+
+cp output/include/n_outliers.tex $INC
+
+cp output/include/hull_cover.tex $INC
+
+cp output/include/path_coef_full.tex $INC
+
+cp output/include/path_coef_struc.tex $INC
+
+cp sem_path_diag/path_diagram_*.tex output/include/
+
+cp output/include/path_diagram_struc.tex $INC
+
+cp output/include/path_diagram_full.tex $INC
+
+cp output/include/path_diagram_concept.tex $INC
 
 # Transfer tables from analysis to manuscript
 
@@ -77,8 +94,8 @@ sed -i -r '12,16s/\w+ \w+\.*/\\textit\{&\}/g' "${INC}clust_summ.tex"
 
 sed -i '12,16s/spp\.\}/\}spp./g' "${INC}clust_summ.tex"
 
-sed -i '10s/.*/{Cluster} \& {Indicator species} \& {Rarefied species richness} \& {Stems ha\\textsuperscript{-1}} \& {AGB (t ha\\textsuperscript{-1})} \\\\/' "${INC}clust_summ.tex"
+sed -i '10s/.*/{Cluster} \& {Indicator species} \& {Species richness} \& {Stems ha\\textsuperscript{-1}} \& {AGB (t ha\\textsuperscript{-1})} \\\\/' "${INC}clust_summ.tex"
 
 sed -i '13,16i\\\hline' "${INC}clust_summ.tex"
 
-sed -i 's/caption{}/caption{Description of the biogeographical clusters (C1-C5) to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \\citep{Dufrene1997} implemented with \\texttt{indval()} from the \\texttt{labdsv} R package \\citep{labdsv}. Numeric values are $\\pm$ 1 standard deviation.}/g' "${INC}clust_summ.tex" 
[…]

### 2019-11-13 analysis/plot_data_collation.R

@@ -15,6 +15,7 @@ rm(list=ls())
 library(dplyr)
 library(tidyr)
 library(iNEXT)
+library(vegan)
 
 # Import data ----
 
@@ -166,9 +167,12 @@ plotcode_plot_group_lookup <- plot_data_agg %>%
   dplyr::select(plotcode_vec, plot_group) %>%
   unnest(plotcode_vec)
 
-write.csv(plotcode_plot_group_lookup, paste0("data/plotcode_plot_group_lookup", ext, ".csv"), row.names = FALSE)
-
 s <- left_join(s, plotcode_plot_group_lookup, by = c("plotcode" = "plotcode_vec"))
+
+s_diam_summ <- s %>% 
+  group_by(plot_group) %>%
+  summarise(min_diam = min(diam, na.rm = TRUE)) %>%
+  filter(min_diam < 10)
 
 # Calculate dbh and height covariance ----
 # Big trees only 
@@ -178,20 +182,72 @@ s_fil <- s %>%
   filter(alive %in% c("A", NA),
     !is.na(gen_sp))
 
+dbh_class_lo <- as.character(seq(from = 5, to = 180, by = 5))
+dbh_class_hi <- as.character(seq(from = 9, to = 184, by = 5))
+
+dbh_labels <- paste0(dbh_class_lo, "-", dbh_class_hi)
+
 s_fil_summ <- s_fil %>%
   group_by(plot_group) %>%
   summarise(n_stems = n(),
     var_height = var(height, na.rm = TRUE),
     mean_height = mean(height, na.rm = TRUE),
     sd_height = sd(height, na.rm = TRUE),
+    max_height = max(height, na.rm = TRUE),
     var_dbh = var(diam, na.rm = TRUE),
     mean_dbh = mean(diam, na.rm = TRUE),
-    sd_dbh = sd(diam, na.rm = TRUE)) %>%
+    sd_dbh = sd(diam, na.rm = TRUE),
+    max_dbh = max(diam, na.rm = TRUE)) %>%
   mutate(cov_height = sd_height / mean_height * 100,
     cov_dbh = sd_dbh / mean_dbh * 100)
 
+# Calculate shannon index of diameter classes
+s_fil$diam_bin <- cut(s_fil$diam, breaks = seq(from = 5, to = 185, by = 5), 
+  labels = dbh_labels, include.lowest = TRUE)
+
+diam_fil_even <- s_fil %>%
+  group_by(plot_group, diam_bin, .drop = FALSE) %>%
[…]

### 2019-11-13 analysis/plot_data_descrip.R

@@ -18,6 +18,7 @@ library(tidyr)
 library(maps)
 library(rgdal)
 library(ggnewscale)
+library(gridExtra)
 
 # Import data ----
 
@@ -41,6 +42,9 @@ white_veg_miombo <- white_veg_fort %>%
 
 # Original data
 load("data/seosaw_plot_summary5Apr2019.Rdata")
+
+# Biomass quantiles
+bchave_quantile_df <- read.csv("data/bchave_quantile.csv")
 
 source("clust_pal.R")
 
@@ -203,10 +207,10 @@ t_p_sp <- SpatialPointsDataFrame(
   coords = t_p_std_nona, data = t_p_std_nona)
 
 # Create a convex hull spatialpolygons object
-plot_hull <- chull(plot_data_final$mean_temp_std, plot_data_final$total_precip_std)
+plot_hull <- chull(plot_data_final$mean_temp_rev_std, plot_data_final$total_precip_std)
 plot_hull <- c(plot_hull, plot_hull[1])
 plot_hull_data <- plot_data_final[plot_hull,]
-plot_hull_coords <- plot_hull_data %>% dplyr::select(mean_temp_std, total_precip_std)
+plot_hull_coords <- plot_hull_data %>% dplyr::select(mean_temp_rev_std, total_precip_std)
 
 plot_hull_poly <- Polygon(plot_hull_coords, hole=F)
 plot_hull_polys <- Polygons(list(plot_hull_poly), 1)
@@ -215,8 +219,14 @@ plot_hull_polys_sp = SpatialPolygons(list(plot_hull_polys))
 # Count number of pixels covered by polygon
 t_p_hull <- over(t_p_sp, plot_hull_polys_sp)
 
-100 - length(t_p_hull[is.na(t_p_hull)]) / length(t_p_hull[!is.na(t_p_hull)]) * 100
+pixel_cover <- round(100 - length(t_p_hull[is.na(t_p_hull)]) / length(t_p_hull[!is.na(t_p_hull)]) * 100, 
+  digits = 1)
 
+fileConn <- file(paste0("output/include/hull_cover.tex"))
+writeLines(
+    paste0("\\newcommand{\\hullcover}{", pixel_cover, "}"),
+  fileConn)
+close(fileConn)
 # Plot the convex hull and the points together
 plot_hull_fort <- fortify(plot_hull_polys_sp)
 
@@ -231,7 +241,7 @@ ggplot() +
[…]

### 2019-11-15 analysis/output/include/corr_ci_tab.tex
Updated flow chart of data cleaning

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Thu, Nov 14, 2019 - 15:38:56
\begin{table}[!htbp] \centering 
  \caption{} 
  \label{corr_ci_tab} 
\begin{tabular}{@{\extracolsep{5pt}} ccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
x\_var & y\_var & raw.r & raw.lower & raw.upper & n & p \\ 
\hline \\[-1.8ex] 
Sand \% & Organic C \% & $$-$0.430$ & $$-$0.470$ & $$-$0.390$ & $1387$ & $0$ \\ 
Sand \% & CEC & $$-$0.580$ & $$-$0.610$ & $$-$0.540$ & $1387$ & $0$ \\ 
Sand \% & MAP & $$-$0.420$ & $$-$0.460$ & $$-$0.370$ & $1387$ & $0$ \\ 
Sand \% & MAP CoV & $0.380$ & $0.330$ & $0.420$ & $1387$ & $0$ \\ 
Sand \% & MAT & $0.240$ & $0.190$ & $0.290$ & $1387$ & $0$ \\ 
Sand \% & MAT CoV & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
Sand \% & Extrap. sp. rich. & $$-$0.290$ & $$-$0.340$ & $$-$0.240$ & $1387$ & $0$ \\ 
Sand \% & Shannon equit. & $$-$0.250$ & $$-$0.300$ & $$-$0.200$ & $1387$ & $0$ \\ 
Sand \% & Tree height CoV & $$-$0.070$ & $$-$0.130$ & $$-$0.010$ & $974$ & $0.024$ \\ 
Sand \% & DBH CoV & $$-$0.180$ & $$-$0.230$ & $$-$0.130$ & $1387$ & $0$ \\ 
Sand \% & Stem density & $$-$0.220$ & $$-$0.260$ & $$-$0.160$ & $1387$ & $0$ \\ 
Sand \% & AGB & $$-$0.190$ & $$-$0.240$ & $$-$0.140$ & $1387$ & $0$ \\ 
Organic C \% & CEC & $0.300$ & $0.250$ & $0.340$ & $1387$ & $0$ \\ 
Organic C \% & MAP & $0.330$ & $0.280$ & $0.380$ & $1387$ & $0$ \\ 
Organic C \% & MAP CoV & $$-$0.350$ & $$-$0.390$ & $$-$0.300$ & $1387$ & $0$ \\ 
Organic C \% & MAT & $$-$0.310$ & $$-$0.350$ & $$-$0.260$ & $1387$ & $0$ \\ 
Organic C \% & MAT CoV & $0$ & $$-$0.050$ & $0.050$ & $1387$ & $0.996$ \\ 
Organic C \% & Extrap. sp. rich. & $0.150$ & $0.100$ & $0.200$ & $1387$ & $0$ \\ 
Organic C \% & Shannon equit. & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
Organic C \% & Tree height CoV & $0.010$ & $$-$0.050$ & $0.080$ & $974$ & $0.663$ \\ 
Organic C \% & DBH CoV & $0.100$ & $0.050$ & $0.160$ & $1387$ & $0$ \\ 
Organic C \% & Stem density & $0.070$ & $0.010$ & $0.120$ & $1387$ & $0.015$ \\ 
Organic C \% & AGB & $0.200$ & $0.150$ & $0.250$ & $1387$ & $0$ \\ 
CEC & MAP & $$-$0.060$ & $$-$0.120$ & $$-$0.010$ & $1387$ & $0.016$ \\ 
CEC & MAP CoV & $$-$0.560$ & $$-$0.590$ & $$-$0.520$ & $1387$ & $0$ \\ 
CEC & MAT & $0.240$ & $0.190$ & $0.290$ & $1387$ & $0$ \\ 
CEC & MAT CoV & $0.080$ & $0.030$ & $0.130$ & $1387$ & $0.002$ \\ 
CEC & Extrap. sp. rich. & $$-$0.070$ & $$-$0.120$ & $$-$0.020$ & $1387$ & $0.011$ \\ 
CEC & Shannon equit. & $0.240$ & $0.190$ & $0.280$ & $1387$ & $0$ \\ 
CEC & Tree height CoV & $0.060$ & $0$ & $0.130$ & $974$ & $0.044$ \\ 
CEC & DBH CoV & $0.250$ & $0.200$ & $0.300$ & $1387$ & $0$ \\ 
CEC & Stem density & $0.210$ & $0.160$ & $0.260$ & $1387$ & $0$ \\ 
CEC & AGB & $0.050$ & $0$ & $0.100$ & $1387$ & $0.053$ \\ 
MAP & MAP CoV & $$-$0.080$ & $$-$0.140$ & $$-$0.030$ & $1387$ & $0.002$ \\ 
MAP & MAT & $$-$0.080$ & $$-$0.130$ & $$-$0.030$ & $1387$ & $0.003$ \\ 
MAP & MAT CoV & $$-$0.650$ & $$-$0.680$ & $$-$0.620$ & $1387$ & $0$ \\ 
MAP & Extrap. sp. rich. & $0.310$ & $0.270$ & $0.360$ & $1387$ & $0$ \\ 
MAP & Shannon equit. & $$-$0.020$ & $$-$0.080$ & $0.030$ & $1387$ & $0.382$ \\ 
MAP & Tree height CoV & $0.180$ & $0.120$ & $0.240$ & $974$ & $0$ \\ 
MAP & DBH CoV & $0.060$ & $0.010$ & $0.110$ & $1387$ & $0.025$ \\ 
MAP & Stem density & $$-$0.070$ & $$-$0.120$ & $$-$0.010$ & $1387$ & $0.013$ \\ 
MAP & AGB & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
MAP CoV & MAT & $$-$0.170$ & $$-$0.220$ & $$-$0.110$ & $1387$ & $0$ \\ 
MAP CoV & MAT CoV & $0.020$ & $$-$0.040$ & $0.070$ & $1387$ & $0.561$ \\ 
MAP CoV & Extrap. sp. rich. & $0.080$ & $0.020$ & $0.130$ & $1387$ & $0.005$ \\ 
MAP CoV & Shannon equit. & $$-$0.230$ & $$-$0.280$ & $$-$0.180$ & $1387$ & $0$ \\ 
MAP CoV & Tree height CoV & $$-$0.050$ & $$-$0.110$ & $0.010$ & $974$ & $0.102$ \\ 
MAP CoV & DBH CoV & $$-$0.350$ & $$-$0.390$ & $$-$0.300$ & $1387$ & $0$ \\ 
MAP CoV & Stem density & $$-$0.250$ & $$-$0.300$ & $$-$0.200$ & $1387$ & $0$ \\ 
MAP CoV & AGB & $$-$0.100$ & $$-$0.160$ & $$-$0.050$ & $1387$ & $0$ \\ 
MAT & MAT CoV & $$-$0.330$ & $$-$0.370$ & $$-$0.280$ & $1387$ & $0$ \\ 
MAT & Extrap. sp. rich. & $$-$0.230$ & $$-$0.280$ & $$-$0.180$ & $1387$ & $0$ \\ 
MAT & Shannon equit. & $0.010$ & $$-$0.040$ & $0.060$ & $1387$ & $0.775$ \\ 
MAT & Tree height CoV & $0.100$ & $0.040$ & $0.160$ & $974$ & $0.002$ \\ 
MAT & DBH CoV & $0.090$ & $0.040$ & $0.150$ & $1387$ & $0$ \\ 
MAT & Stem density & $0.080$ & $0.030$ & $0.130$ & $1387$ & $0.003$ \\ 
MAT & AGB & $$-$0.060$ & $$-$0.110$ & $0$ & $1387$ & $0.038$ \\ 
MAT CoV & Extrap. sp. rich. & $$-$0.240$ & $$-$0.290$ & $$-$0.190$ & $1387$ & $0$ \\ 
MAT CoV & Shannon equit. & $$-$0.030$ & $$-$0.090$ & $0.020$ & $1387$ & $0.193$ \\ 
MAT CoV & Tree height CoV & $$-$0.120$ & $$-$0.180$ & $$-$0.060$ & $974$ & $0$ \\ 
MAT CoV & DBH CoV & $$-$0.020$ & $$-$0.070$ & $0.030$ & $1387$ & $0.428$ \\ 
MAT CoV & Stem density & $$-$0.070$ & $$-$0.120$ & $$-$0.020$ & $1387$ & $0.010$ \\ 
MAT CoV & AGB & $$-$0.080$ & $$-$0.140$ & $$-$0.030$ & $1387$ & $0.002$ \\ 
Extrap. sp. rich. & Shannon equit. & $0.540$ & $0.500$ & $0.580$ & $1387$ & $0$ \\ 
Extrap. sp. rich. & Tree height CoV & $0.070$ & $0$ & $0.130$ & $974$ & $0.038$ \\ 
Extrap. sp. rich. & DBH CoV & $0.160$ & $0.110$ & $0.210$ & $1387$ & $0$ \\ 
Extrap. sp. rich. & Stem density & $0.390$ & $0.350$ & $0.440$ & $1387$ & $0$ \\ 
Extrap. sp. rich. & AGB & $0.460$ & $0.420$ & $0.500$ & $1387$ & $0$ \\ 
Shannon equit. & Tree height CoV & $$-$0.060$ & $$-$0.130$ & $0$ & $974$ & $0.046$ \\ 
Shannon equit. & DBH CoV & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
Shannon equit. & Stem density & $0.540$ & $0.500$ & $0.580$ & $1387$ & $0$ \\ 
Shannon equit. & AGB & $0.400$ & $0.350$ & $0.440$ & $1387$ & $0$ \\ 
Tree height CoV & DBH CoV & $0.530$ & $0.480$ & $0.570$ & $974$ & $0$ \\ 
Tree height CoV & Stem density & $$-$0.030$ & $$-$0.100$ & $0.030$ & $974$ & $0.311$ \\ 
Tree height CoV & AGB & $0.210$ & $0.150$ & $0.270$ & $974$ & $0$ \\ 
DBH CoV & Stem density & $0.190$ & $0.140$ & $0.240$ & $1387$ & $0$ \\ 
DBH CoV & AGB & $0.420$ & $0.380$ & $0.470$ & $1387$ & $0$ \\ 
Stem density & AGB & $0.510$ & $0.470$ & $0.540$ & $1387$ & $0$ \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

### 2019-11-15 analysis/output/include/path_diagram_full.tex
Updated flow chart of data cleaning

--- 2019-11-13 manuscript/include/path_diagram_full.tex
@@
 	path/.style={->, thick, >=stealth'},
-	loading/.style={->, thick, dashed, >=stealth'}]
+	loading/.style={->, thick, dotted, >=stealth'},
+	loadingset/.style={->, thick, dashed, >=stealth'}]
 
@@
 \coordinate[below= 0.7in of s] (somid);
-\node [observed] (sc) [left = 0.5in of somid] {SOC};
-\node [observed] (ss) at (somid) {Sand};
+\node [observed] (ss) [left = 0.5in of somid] {Sand};
+\node [observed] (sc) at (somid) {SOC};
 \node [observed] (so) [right = 0.5in of somid] {CEC};
@@
 
-\draw [loading] (m) to node[left, pos=0.5] {\pcfmmp{}} (mp);
+\draw [loadingset] (m) to node[left, pos=0.5] {\pcfmmp{}} (mp);
 \draw [loading] (m) to node[left, pos=0.5] {\pcfmmpc{}} (mpc);
@@
 
+\draw [loadingset] (s) to node[left, pos=0.5] {\pcfsss{}} (ss);
 \draw [loading] (s) to node[left, pos=0.5] {\pcfssc{}} (sc);
-\draw [loading] (s) to node[left, pos=0.5] {\pcfsss{}} (ss);
 \draw [loading] (s) to node[right, pos=0.5] {\pcfsso{}} (so);
 
-\draw [loading] (d) to node {\pcfdds{}} (ds);
+\draw [loadingset] (d) to node {\pcfdds{}} (ds);
 \draw [loading] (d) to node[left, pos=0] {\pcfdde{}} (de);
@@
 \draw [loading] (h) to node[] {\pcfhhd{}} (hd);
-\draw [loading] (h) to node {\pcfhhh{}} (hh);
+\draw [loadingset] (h) to node {\pcfhhh{}} (hh);
 

### 2019-11-15 manuscript/chapter_1.tex
Updated flow chart of data cleaning

@@
 \usetikzlibrary{calc,arrows,positioning,shapes,shapes.gates.logic.US,trees}
+\usepackage{longtable}
 
@@
 The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}.
-In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales in savannas.
+In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales in savannas.
 \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
@@
 While southern African woodlands are species rich in the herbaceous understorey, the tree layer is relatively species poor \citep{}.
-\citet{Miombo-Book} however, writes that they are structurally diverse in the tree layer, with trees of the same species occupying distinct layers of the canopy at different growth stages.
+\citet{Solbrig1996} however, writes that they are structurally diverse in the tree layer, with trees of the same species occupying distinct layers of the canopy at different growth stages.
 This structural diversity may be one mechanism through which species diversity appears to influence woody biomass.
@@
 
-High levels of disturbance in the form of seasonal fires and herbivory from large mammals \citep{Fireherbivore} may moderate the strength of observable BEFRs in southern African woodlands.
+High levels of disturbance in the form of seasonal fires and herbivory from large mammals may moderate the strength of observable BEFRs in southern African woodlands \citep{Staver2009, Bond2005}.
 Fire disturbance in forests has been linked to abundance dependent mortality among smaller stems \citep{Roques2001}.
 A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion \citep{}.
-Such fast-growing bushy species exist in southern African woodlands and often proliferate when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{nieto2018}.
+Such fast-growing bushy species exist in southern African woodlands and often proliferate when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}.
 Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
@@
 
-The study used \nplots{} woodland monitoring plots from a larger network located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1983}).
+The study used \nplots{} woodland monitoring plots from a larger network located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}).
 The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
 The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}).
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{worldclim}.
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
 
@@
 \subfloat[]{{\includegraphics[width=0.4\textwidth]{temp_precip_hull}}\label{temp_precip_hull}}%
-\caption{The locations of the \nplots{} plots used in this study, by geographic location (a) with respect to the distribution of Miombo woodland vegetation according to \citet{White}, and in climate space (b), showing the plot locations as points compared to the climate space of the whole region as estimated using the WorldClim dataset over the Miombo woodland vegetation extent \citep{worldclim}.
+\caption{The locations of the \nplots{} plots used in this study, by geographic location (a) with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}, and in climate space (b), showing the plot locations as points compared to the climate space of the whole region as estimated using the WorldClim dataset over the Miombo woodland vegetation extent \citep{Fick2017}.
 Note that the density colour scale is log-transformed.}
@@
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
-The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from \citet{Ryanunbublished}.
+The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from (Ryan, unpublished).
 
@@
 Structural Equation Models (SEM) investigated the determinants of AGB.
-All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R}.
-SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lehmann2014_161-166}.
+All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R2019}.
+SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
 A key aspect to our decision to use SEMs is that they can explicitly model indirect effects, which is impossible in multiple regression.
@@
 	\input{include/path_diagram_concept}
-	\caption{Conceptual model showing the theoretical relationships between environmental factors, structural and species tree diversity, and AGB.
-Hypothesised paths of causal inference are depicted as arrows from predictor to response.
-Latent variables are represented as shaded boxes with their constituent observed variables within.
-The names of each latent variable are in ovals.
-The pale orange dashed box defines a nested model which includes only tree diversity variables.}
+	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural iversity, tree stem density, and AGB.
+Hypothesised paths of causal inference are depicted as arrows from predictor to response.}
 	\label{con_mod}
@@
 When comparing the path coefficients of models using individual vegetation clusters, unstandardised path coefficients are used in order to more readily compare the same path coefficient across different models while accounting for variation in mean values between biogeographical clusters.
-Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Kline2005}.
+Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.
 
@@
 	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlogram of standardised observed variables used in the SEMs, with correlation coefficients shaded according to sign ($+$ve red, $-$ve blue) and strength of correlation.
-Variables in bold were later reversed for modelling to maintain positive correlations between observed variables within each latent variable.
+	\caption{Correlogram of standardised observed variables used in the SEMs, with pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
+Variables in bold on the axis labels were later reversed for SEMs to maintain positive correlations between observed variables within each latent variable.
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
-Colours of variable names group variables into latent variables: red = Soil fertility, blue = Moisture availability, green = species diversity, purple = structural diversity.}
+Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = species diversity, purple = structural diversity.
+See \autoref{corr_ci_tab} for a full table of correlation fit statistics.}
 	\label{corr_mat}
@@
 
-A simple correlation plot showed that all tree species diversity and structural diversity observed variables all had moderate positive correlations with AGB.
-Stem density had the strongest correlation with AGB of all variables ($\rho$ = 0.51).
+A simple set of correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
+Stem density had the strongest correlation with AGB of all variables ($\rho$ = 0.51, n = 1387, p <0.001).
 Environmental variables had weaker correlations with AGB than diversity variables.
-Only MAP ($\rho$ = 0.19), MAP seasonality ($\rho$ = -0.18), soil Organic C \% ($\rho$ = 0.19) and soil Sand \% ($\rho$ = -0.18) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero.
+Only MAP ($\rho$ = 0.19, n = 1387, p <0.001), MAP seasonality ($\rho$ = -0.18, n = 1387, p <0.001 ), soil Organic C \% ($\rho$ = 0.19, n = 1387, p <0.001) and soil Sand \% ($\rho$ = -0.18, n = 1387, p <0.001) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero.
 (\autoref{corr_mat}).
@@
 MAT had non significant correlations with tree species and structural diversity variables.
-
-Tree species diversity variables had clear positive correlations with stem density (Species richness: $\rho$ = 0.45, Shannon equitability: $\rho$ = 0.43).
+Tree species diversity variables had clear positive correlations with stem density (Species richness: $\rho$ = 0.45, n = 1387, p <0.001; Shannon equitability: $\rho$ = 0.43, n = 1387, p <0.001).
 
@@
 In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via stem density, but not via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}).
-Stand structural diversity had a positive direct effect on AGB independent of species diversity (\todo{STATS}).
+Stand structural diversity had a positive direct effect on AGB independent of species diversity.
+Tree species diversity had a strong positive effect on stem density.
 Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}).
-
-When the model was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardized path coefficients varied but generally similar relationships between diversity and AGB with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
-It should be noted that models for vegetation clusters 4 and 5 did not converge, possibly due to small sample sizes (\autoref{struc_model_slopes_all}).
 
@@
 	\input{include/path_diagram_struc}
-	\caption{Path diagram with regression coefficients for the structural model including plots in all five clusters.}
+	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all five vegetation clusters.
+Latent variables are shown as circles while observed variables are shown as rectangles.
+Path coefficients are solid arrows pointing from predictor to response.
+The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
+Measurement errors of exogenous variables are omitted for clarity.}
 	\label{struc_mod}
 \end{figure}
-
-%Meaasurement errors of exogenous variables omitted for clarity
 
@@
 	\caption{Standardized path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
-Due to all observed variables being standardized and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale +/- 1 standard error.
+Due to all observed variables being standardized and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
 Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
@@
 \subsection{Variation among vegetation clusters in structural and species diversity effects}
+
+When the structural diversity and species diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardized path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
+It should be noted that models for vegetation clusters 4 and 5 did not converge, possibly due to small sample sizes (\autoref{struc_model_slopes_all}).
 
@@
 
-We repeatedly sub-sampled the available plots to build many datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
-The same SEM specification was used as above, incorporating the effects of species diversity and structural diversity on AGB (\autoref{con_mod}).
-\autoref{sem_struc_stems_ha} shows that there is a peak in the effect of diversity on biomass at \textapprox{}765 stems ha\textsuperscript{-1}.
-There appears to be a minimum stem density threshold below which there appears to be no effect of diversity on biomass.
-This threshold is at \textapprox{}295 stems ha\textsuperscript{-1}.
-Interestingly, at high stem densities, the effect of species diversity on biomass declines.
-The effect of structural diversity on biomass appears to decrease with increasing stem density, though the relationship is less clear.
+We repeatedly sub-sampled the dataset of plots to build many datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
+The same SEM specification was used as above except that stem density was removed, incorporating the effects of species diversity and structural diversity on AGB (\autoref{con_mod}).
+\autoref{sem_struc_stems_ha} shows that there is a peak in the total effect of tree species diversity on biomass at \textapprox{}700 stems ha\textsuperscript{-1}.
+There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of diversity on biomass.
+Interestingly, at high stem densities, the observed effect of species diversity on biomass begins to decline.
+The effect of structural diversity on biomass appears to decrease slightly with increasing stem density, though the relationship is less clear.
+At stem densities below \textapprox{}650 stems ha\textsuperscript{-1}  there is no observable effect of tree species diversity on structural diversity.
 
@@
 	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density.}
+	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves.}
 	\label{sem_struc_stems_ha}
@@
 
-A model incorporating moisture availability and soil showed that at the regional level, the effect of diversity on biomass was greater than that of moisture and soil.
-Surprisingly, the effect of moisture availability on biomass was negligible, with all of the observed effect coming from the indirect path of moisture availability on biomass via species diversity.
+A model incorporating the latent variables of moisture availability and soil fertility showed that the effect of diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{path_diagram_full}).
+Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (\autoref{full_model_slopes}).
 
-Soil fertility had a strong positive effect on species diversity. Soil fertility also had a positive effect on AGB.
+Moisture availability and soil fertility also had negligible direct effects on stem density, but both had a positive total effect on stem density via species diversity, where species diversity itself had a strong positive effect on stem density.
+As species diversity increases, stem density also increases.
 
-The effect of species diversity had the strongest total effect on biomass, with some indirect effect via structural diversity.
+Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was very weakly positive, while structural diversity itself had a positive effect on AGB.
+Again, stem density had a positive direct effect on AGB.
 
@@
 	\input{include/path_diagram_full}
-	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters.}
+	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters.
+Latent variables are shown as circles while observed variables are shown as rectangles.
+Path coefficients are solid arrows pointing from predictor to response.
+The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
+Measurement errors of exogenous variables are omitted for clarity.}
 	\label{full_mod}
@@
 
-Vegetation cluster level models could not be reliably fitted for these more complex models with environmental covariates, due to sample size and convergence issus.
+Vegetation cluster level models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation clusters were narrow in their climate space.
+leading to a lack of variance particularly in moisture availability.
 
@@
 
-To further address our hypothesis (H\textsubscript{2}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit a separate model with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
+To further address the hypothesis (H\textsubscript{2}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
 
-Moderating effects of moisture availability were not observed.
+Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass.
+Regression coefficient for the interaction effect of tree species diversity and moisture availability was significant ($\beta$ = 0.058, t(1383) = 2.562, p <0.05), while the interaction term of tree species diversity and soil fertility was not ($\beta$ = 0.034, t(1383) = 1.452, p = 0.147).
+As moisture availability increased, the effect of tree species diversity on AGB became stronger.
+At high tree species diversity, an increase in resource availability led to a larger increase in biomass.
 
-Soil fertility had a small positive effect on the relationship between species diversity and biomass.
-As soil fertility increased, the strength of the relationship between species diversity and biomass also increased.
-This effect was small in comparison to the direct effect of soil fertility on AGB.
-
-subsection{Large trees}
-
-We analysed how the strength of the relationship between tree species richness and AGB varied for different size classes of tree.
-\autoref{bchave_quantile_slope_sp_rich} shows a saturating increase in the importance of species richness to determine AGB as the proportional size of trees increases.
+\begin{figure}[H]
+\centering
+	\subfloat[]{
+	\includegraphics[width=0.48\textwidth]{mois_int}
+	}
+	\subfloat[]{
+	\includegraphics[width=0.48\textwidth]{soil_int}
+	}
+	\caption{Interaction plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b).
+All variables are standardised and centred to a mean of zero and a standard deviation of 1.
+Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2.
+Moisture availability, tree species diversity and soil fertility are the values of latent variables the same as those used in the SEMs.}
+	\label{int_plots}
+\end{figure}
 
@@
 
+%1. Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.
+%2. More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above-ground woody biomass.
+%3. There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species diversity.
+%4. Structural characteristics of the woodland will interact with species diversity to provide an indirect path of influence between species composition and biomass stocks.
+%5. The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.
+
+In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, and [d] stem density on aboveground woody biomass (AGB) across southern African woodlands.
+Using latent variables and SEMs, we found support for a general relationship between tree species diversity and AGB, and a relationship between tree structural diversity and AGB that was independent of tree species diversity.
+Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB.
+
 \subsection{Effect of tree species diversity on AGB}
+
+The latent variable of tree species diversity had the strongest positive effect on woody above-ground biomass (AGB) in environment+diversity SEM model (\autoref{full_mod}).
+Within the savanna woodlands of southern Africa we therefore find support for our hypothesis (H\textsubscript{1}), that higher tree species richness causes higher woody AGB.
+This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{}.
+
+Most of the total species diversity effect was through the indirect effect via stem density.
+This suggests that a higher species diversity allows for a greater density of tree stems and that in southern African woodlands, this leads to an increase in total AGB.
+The effect of increased stem density within forests is variable.
+The question of whether lots of small stems or a few large stems maximises AGB has yet to be solved \citep{}.
+
+While we did not explicitly measure net primary production (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{}.
+Mechanistically, a higher standing woody biomass results in a greater stand level leaf area index (LAI) \citep{}, with leaf area being a primary determinant of productivity \citep{}.
+
+As tree species diversity increased, this allowed for greater stem density which in turn led to greater AGB.
 
@@
 
-\subsection{Moderation effects of moisture availability and soil fertility}
+We found no evidence that species diversity led to an increase in structural diversity.
+Structural diversity did however have a positive effect on AGB.
+Our original hpyothesis was that an increase in tree species richness would increase the number of functional forms of trees per unit area due to niche complementarity in growth form of coexisting species.
+However, it seems more likely that structural diversity is instead merely a result of disturbance history.
+With highly disturbed plots generally having a less structurally diverse canopy.
+In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges.
+In highly disturbed woodland plots, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained.
 
-\subsection{Vegetation type specific responses}
+\subsection{Effects of moisture availability and soil fertility}
+
+Surprisingly, moisture availability and soil fertility had only small effects on AGB compared to that of tree species diversity.
+We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater biomass per unit area.
+We expected that a greater number of stems would be able to coexist in a higher resource availability area.
+
+Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via tree species diversity and stand structure \todo{what sort of stand structure?} \citep{Ali2019a, Ali2019b, Poorter2017}.
+This was not observed in our study however.
+Compared to most tropical forests, climatic water availability is far more of a limiting factor in southern African woodlands.
+We expect that an increase in water availability in southern African woodlands would allow for greater tree growth and therefore greater AGB.
+The range of observed moisture availability in this study however may not have been able to capture this effect.
+Instead, in southern African woodlands moisture availability is much more closely linked with disturbance from seasonal fires.
+The growth of C4 grasses in wetter savannas leads to seasonal fires which limit tree growth.
+Additionally, we may see high local adaptation of tree species andf compositional turnover in response to variation in moisture availability, negating any effects of moisture availability on tree growth.
+Savannas in Africa have been dry for a long time so it is understnadable that trees may be highly adapted to low rainfall to the point where it doesn't affect their growth.
+
+In a separate set of linear multiple regressions, we found a weak positive interaction effect of moisture availabiltiy on the relationship between tree species diversity and AGB.
+As moisture availability increased, the relationship between tree species diversity and AGB became stronger.
+We could not add moderation effects to the SEM framework.
+
+\subsection{Vegetation type cluster specific responses}
 
@@
 
-\todo{Why is there a peak in the effect of species richness on AGB at a certain level of stem density?}
+We found a non linear effect of stem density on the relationship between tree species diversity and AGB.
+The effect of tree species diversity on AGB peaked at \textapprox{}700 stems ha\textsuperscript{-1}.
+At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness does not make any difference.
+
+It is possible that at very high stem density there are other effects that become more important than tree species diversity in affecting AGB.
+For example, very high stem density plots may have been found areas of thicket vegetation which are highly structured by disturbance.
+There may also be an effect whereby the high stem density plots have lots of very small stems, meaning that AGB cannot be high.
+The largest trees always ahve the bulk of the biomass in a system.
+Again, this is indicative of thicket vegetation.
+The peak in species diversity effect may occur at a given stem density because that is the balance between tree species diversity effects and the offsetting effects of stem crowding which preclude high biomass large trees.
 
@@
 
-%\bibliography{/Users/johngodlee/google_drive/bib/lib}
+\subsection{Future work}
+
+We hope that this study prompts further investigation of the relationship between woody productuivity and species richness and composition in southern African woodlands.
+While AGB is correlated with NPP,...
 
 \subsection{Conclusion}
+
+\bibliography{/Users/johngodlee/google_drive/bib/lib.bib}
+
+\newpage{}
+\appendix{}
+
+\section*{Appendix 1 - Data cleaning process}
+
+\begin{figure}[H]
+\centering
+	\input{include/data_clean_flow}
+	\caption{Flow diagram of the data filtering and cleaning process prior to analysis. Rounded boxes indicate filtering events while normal boxes indicate calculation events.}
+	\label{data_clean_flow}
+\end{figure}
+
+\section*{Appendix 2 - Table of correlation fit statistics}
+
+\input{include/corr_ci_tab}
+
+\section*{Appendix 3 - Frequency distribution of observed variables}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{histogram_raw_obs}
+	\caption{Histograms of raw untransformed observed variables used in final analyses.}
+	\label{histogram_raw_obs}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{histogram_trans_obs}
+	\caption{Histograms of observed variables transformed to achieve a normal frequency distribution.}
+	\label{histogram_trans_obs}
+\end{figure}
+
+\section*{Appendix 4 - Regression tables for moderation by environmental variables}
+
+\input{include/mois_div_int_mod}
+
+\input{include/soil_div_int_mod}
 

### 2019-11-15 manuscript/include/path_diagram_struc.tex
Updated flow chart of data cleaning

@@
 
-\draw [loading] (d) to node[left, pos=0.5] {\pcsdsd{}} (ds);
+\draw [loadingset] (d) to node[left, pos=0.5] {\pcsdsd{}} (ds);
 \draw [loading] (d) to node[right, pos=0.5] {\pcsded{}} (de);
-\draw [loading] (h) to node[left, pos=0.5]  {\pcshdh{}} (hd);
+\draw [loadingset] (h) to node[left, pos=0.5]  {\pcshdh{}} (hd);
 \draw [loading] (h) to node[right, pos=0.5] {\pcshhh{}} (hh);

### 2019-11-15 manuscript/include/struc_model_fit_clust_stats.tex
More functions in SEM to automate clusters
Updated figures and text, pre-comments from Kyle
Made tikz path diagrams
Updated flow chart of data cleaning

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Tue, Oct 22, 2019 - 11:39:36
+% Date and time: Fri, Nov 08, 2019 - 17:15:20
 \begin{table}[!htbp] \centering 
-  \caption{} 
+  \caption{Model fit statistics for SEMs investigating the effects of tree diversity and stem density on AGB (\autoref{struc_mod}).} 
   \label{struc_model_fit_clust_stats} 
-\begin{tabular}{@{\extracolsep{5pt}} cccccccccccccc} 
+\begin{tabular}{@{\extracolsep{0pt}} ccccccccccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
- & npar & chisq & df & cfi & tli & logl & aic & bic & ntotal & rmsea & rmsea.ci.lower & rmsea.ci.upper & srmr \\ 
+{Cluster} & {Params.} & {n} & {$\chi^{2}$} & {DoF} & {CFI} & {TLI} & {LogLik} & {AIC} & {RMSEA} & {SRMR} \\
 \hline \\[-1.8ex] 
-C1 & $12$ & $4.660$ & $3$ & $0.997$ & $0.989$ & $$-$1854$ & $3732$ & $3777.900$ & $337$ & $0.040$ & $0$ & $0.108$ & $0.016$ \\ 
-C2 & $12$ & $22.830$ & $3$ & $0.982$ & $0.939$ & $$-$2838.600$ & $5701.200$ & $5754.300$ & $619$ & $0.100$ & $0.066$ & $0.145$ & $0.017$ \\ 
-C3 & $12$ & $7.910$ & $3$ & $0.961$ & $0.871$ & $$-$502.300$ & $1028.500$ & $1057$ & $79$ & $0.140$ & $0.012$ & $0.270$ & $0.073$ \\ 
-C4 & $12$ & $1.690$ & $3$ & $1$ & $1.054$ & $$-$139.600$ & $303.100$ & $318.200$ & $26$ & $0$ & $0$ & $0.265$ & $0.048$ \\ 
-C5 & $12$ & $1.120$ & $3$ & $1$ & $1.068$ & $$-$368.200$ & $760.400$ & $785.300$ & $59$ & $0$ & $0$ & $0.146$ & $0.034$ \\ 
+C1 & 12 & 420 & 10.430 & 3 & 0.987 & 0.958 & -2630.700 & 5285.300 & 0.080 & 0.029 \\ 
+C2 & 12 & 671 & 50.320 & 3 & 0.942 & 0.806 & -3730.200 & 7484.300 & 0.150 & 0.052 \\ 
+C3 & 12 & 105 & 24.970 & 3 & 0.880 & 0.601 & -619.100 & 1262.300 & 0.260 & 0.142 \\ 
+C4 & 12 & 46 & 2.710 & 3 & 1 & 1.008 & -257.900 & 539.900 & 0 & 0.025 \\ 
+C5 & 12 & 84 & 5.400 & 3 & 0.981 & 0.936 & -529.500 & 1083 & 0.100 & 0.085 \\ 
+All & 12 & 1326 & 62.210 & 3 & 0.968 & 0.895 & -8110.500 & 16245 & 0.120 & 0.036 \\ 
 \hline \\[-1.8ex] 

### 2019-11-15 analysis/bchave_quantile.R

@@ -1,0 +1,51 @@
+# Investigating how the relationship of AGB and species richness varies with size class
+# John Godlee (johngodlee@gmail.com)
+# 2019_11_13
+
+rm(list = ls())
+
+# Packages 
+library(dplyr)
+library(ggplot2)
+
+# Import data 
+bchave_quantile_df <- read.csv("data/bchave_quantile.csv")
+
+quantile_labs <- c(paste0(seq(from = 0, to = 95, by = 5), "-", seq(from = 5, to = 100, by = 5)))
+
+# Mean AGB per quantile
+quantile_gather <- bchave_quantile_df %>%
+  gather("quantile", "bchave", -plot_group, -sp_rich, -sp_rich_raref) %>%
+  mutate(quantile = factor(quantile))
+
+levels(quantile_gather$quantile) <- quantile_labs
+
+ggplot(quantile_gather, aes(x = sp_rich_raref, y = log(bchave))) + 
+  geom_point(aes(colour = quantile), alpha = 0.6) + 
+  geom_smooth(method = "lm", colour = "black") + 
+  facet_wrap(~quantile) + 
+  theme_classic() + 
+  theme(legend.position = "none")
+
+# Linear models
+mod_list <- list()
+for(i in 1:length(bchave_quantile_df[,-1:-3])){
+  mod_list[[i]] <- summary(lm(log(bchave_quantile_df[,-1:-3][,i]) ~ sp_rich_raref, data = bchave_quantile_df))
+}
+
+mod_summ_df <- data.frame(
+  quantile = quantile_labs,
+  est = sapply(mod_list, function(x){x$coefficients[2,1]}),
+  se = sapply(mod_list, function(x){x$coefficients[2,2]}))
+
+ggplot() + 
+  geom_point(data = mod_summ_df, aes(x = quantile, y = est)) + 
+  geom_errorbar(data = mod_summ_df, 
+    aes(x = quantile, ymin = est - se, ymax = est + se)) + 
+  labs(x = "Percentile of tree size\nmeasured by AGB",
+    y = "Slope of linear model\n(log(AGB) ~ Species rich.)") + 
+  theme_classic() + 
+  theme(axis.text.x = element_text(size=12, 
+    angle=45, 
+    vjust=1, 
+    hjust=1))

### 2019-11-15 analysis/bivar_rel_norm.R

@@ -86,6 +86,68 @@ pdf(file = paste0("img/histogram_raw", ext, ".pdf"), width = 12, height = 7)
 histogram_raw
 dev.off()
 
+facet_levels <- c(
+  "total_precip",
+  "precip_seasonality",
+  "mean_temp",
+  "temp_seasonality",
+  "bchave",
+  "cation_ex_cap",
+  "sand_per",
+  "ocdens",
+  "cov_dbh",
+  "cov_height",
+  "stems_ha",
+  "shannon_equit",
+  "sp_rich_raref",
+  "bchave_log", 
+  "stems_ha_log",
+  "sp_rich_raref_log",
+  "shannon_equit_log", 
+  "precip_seasonality_log", 
+  "temp_seasonality_log")
+
+facet_labels <- c(  
+  expression("MAP" ~ (mm ~ y^-1)),
+  expression("Precip." ~ "seasonality"),
+  expression("MAT" ~ (degree * C)), 
+  expression("Temp." ~ "seasonality"),
+  expression("AGB" ~ (t ~ ha^-1)),
+  expression("CEC"),
+  expression("Sand" ~ "%"), 
+  expression("Org." ~ "C" ~ "%"),
+  expression("Coef." ~ "var." ~ "DBH"),
+  expression("Coef." ~ "var." ~ "height"),
+  expression("Stem" ~ "density" ~ ">5" ~ cm ~ (n ~ ha^-1)),
+  expression("Shannon" ~ "equit." ~ (E[H])),
+  expression("Extrap." ~ "species" ~ "rich."),
+  expression("log(AGB)" ~ (t ~ ha^-1)),
+  expression("log(Stem" ~ "density)" ~ ">5" ~ cm ~ (n ~ ha^-1)),
+  expression("log(Extrap." ~ "species" ~ "rich.)"),
+  expression("log(Shannon" ~ "equit.)" ~ (E[H])),
+  expression("log(Precip." ~ "seasonality)"),
+  expression("log(Temp." ~ "seasonality)"))
+
+histogram_raw_obs <- sem_data_fil_agg %>%
+  dplyr::select(bchave, stems_ha, 
+    sp_rich_raref, shannon_equit,
+    cation_ex_cap, sand_per, ocdens, 
+    total_precip, precip_seasonality, mean_temp, temp_seasonality,
+    cov_height, cov_dbh) %>%
+  gather(variable, value) %>%
+  mutate(facet_label = factor(variable,
+    levels = facet_levels,
+    labels = facet_labels)) %>%
[…]

### 2019-11-15 analysis/img_transfer.sh

@@ -33,6 +33,14 @@ cp img/full_mod.pdf $OUT
 
 cp img/sem_struc_stems_ha.pdf $OUT
 
+cp img/soil_int.pdf $OUT 
+
+cp img/mois_int.pdf $OUT
+
+cp img/histogram_raw_obs.pdf $OUT
+
+cp img/histogram_trans_obs.pdf $OUT
+
 cp sem_path_diag/con_mod.pdf $OUT
 
 cp output/include/n_plots.tex $INC
@@ -47,11 +55,16 @@ cp output/include/path_coef_struc.tex $INC
 
 cp sem_path_diag/path_diagram_*.tex output/include/
 
+cp data_clean_flow/*.tex output/include/
+
+cp output/include/data_clean_flow.tex $INC
+
 cp output/include/path_diagram_struc.tex $INC
 
 cp output/include/path_diagram_full.tex $INC
 
 cp output/include/path_diagram_concept.tex $INC
+
 
 # Transfer tables from analysis to manuscript
 
@@ -64,6 +77,7 @@ sed -i 's/\\extracolsep{5pt}/\\extracolsep{0pt}/g' "${INC}struc_model_fit_clust_
 
 sed -i "10s/.*/${STRUC_HEAD}/" "${INC}struc_model_fit_clust_stats.tex"
 
+sed -i 's/caption{}/caption{Model fit statistics for SEMs investigating the effects of tree diversity and stem density on AGB (\\autoref{struc_mod}).}/g' "${INC}struc_model_fit_clust_stats.tex" 
 
 ## Full model fit
 cp output/include/full_model_fit_clust_stats.tex $INC
@@ -99,3 +113,79 @@ sed -i '10s/.*/{Cluster} \& {Indicator species} \& {Species richness} \& {Stems 
 sed -i '13,16i\\\hline' "${INC}clust_summ.tex"
 
[…]

### 2019-11-15 analysis/plot_data_collation.R

@@ -310,9 +310,9 @@ plot_data_agg <- plot_data_agg %>%
   filter(shannon_equit > 0,
     shannon_exp > 0) %>%
 # Exclude big 10 Ha plot
-  filter(plotcode != "DKS001") %>%
+  filter(plotcode != "DKS001")# %>%
 # Exclude plots where min diam. is not less than ten cm
-  filter(plot_group %in% s_diam_summ$plot_group)
+ # filter(plot_group %in% s_diam_summ$plot_group)
 
 # Ensure that variables all have the same sign
 plot_data_agg$precip_seasonality_rev <- -1 * plot_data_agg$precip_seasonality + 1000

### 2019-11-19 analysis/output/include/clust_summ.tex
Part way through editing

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Fri, Nov 08, 2019 - 11:40:45
+% Date and time: Mon, Nov 18, 2019 - 10:53:21
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 7+4.7 & 276+345.2 & 35.3+37.03 \\ 
-C2 & Julbernardia paniculata, Isoberlinia angolensis, Albizia antunesiana & 8+4.3 & 274+245.9 & 46.8+35.05 \\ 
-C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 4+2.3 & 274+536.5 & 36.3+27.86 \\ 
-C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 10+5.9 & 447+386.3 & 47+45.41 \\ 
-C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 4+2.9 & 322+390.1 & 43.8+34.45 \\ 
+C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 7(5) & 316(228.6) & 37.1(27.66) \\ 
+C2 & Julbernardia paniculata, Isoberlinia angolensis, Albizia antunesiana & 8(5) & 291(191.7) & 43.8(40.28) \\ 
+C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 5(2) & 362(259.4) & 37(42.59) \\ 
+C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 11(9) & 495(516.8) & 51.7(50.74) \\ 
+C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 5(5) & 424(226.8) & 50.9(32.7) \\ 
 \hline \\[-1.8ex] 

### 2019-11-19 analysis/output/include/corr_ci_tab.tex
Part way through editing

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Nov 14, 2019 - 15:38:56
+% Date and time: Tue, Nov 19, 2019 - 10:53:45
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-Sand \% & Organic C \% & $$-$0.430$ & $$-$0.470$ & $$-$0.390$ & $1387$ & $0$ \\ 
-Sand \% & CEC & $$-$0.580$ & $$-$0.610$ & $$-$0.540$ & $1387$ & $0$ \\ 
-Sand \% & MAP & $$-$0.420$ & $$-$0.460$ & $$-$0.370$ & $1387$ & $0$ \\ 
-Sand \% & MAP CoV & $0.380$ & $0.330$ & $0.420$ & $1387$ & $0$ \\ 
-Sand \% & MAT & $0.240$ & $0.190$ & $0.290$ & $1387$ & $0$ \\ 
-Sand \% & MAT CoV & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
-Sand \% & Extrap. sp. rich. & $$-$0.290$ & $$-$0.340$ & $$-$0.240$ & $1387$ & $0$ \\ 
-Sand \% & Shannon equit. & $$-$0.250$ & $$-$0.300$ & $$-$0.200$ & $1387$ & $0$ \\ 
-Sand \% & Tree height CoV & $$-$0.070$ & $$-$0.130$ & $$-$0.010$ & $974$ & $0.024$ \\ 
-Sand \% & DBH CoV & $$-$0.180$ & $$-$0.230$ & $$-$0.130$ & $1387$ & $0$ \\ 
-Sand \% & Stem density & $$-$0.220$ & $$-$0.260$ & $$-$0.160$ & $1387$ & $0$ \\ 
-Sand \% & AGB & $$-$0.190$ & $$-$0.240$ & $$-$0.140$ & $1387$ & $0$ \\ 
-Organic C \% & CEC & $0.300$ & $0.250$ & $0.340$ & $1387$ & $0$ \\ 
-Organic C \% & MAP & $0.330$ & $0.280$ & $0.380$ & $1387$ & $0$ \\ 
-Organic C \% & MAP CoV & $$-$0.350$ & $$-$0.390$ & $$-$0.300$ & $1387$ & $0$ \\ 
-Organic C \% & MAT & $$-$0.310$ & $$-$0.350$ & $$-$0.260$ & $1387$ & $0$ \\ 
-Organic C \% & MAT CoV & $0$ & $$-$0.050$ & $0.050$ & $1387$ & $0.996$ \\ 
-Organic C \% & Extrap. sp. rich. & $0.150$ & $0.100$ & $0.200$ & $1387$ & $0$ \\ 
-Organic C \% & Shannon equit. & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
-Organic C \% & Tree height CoV & $0.010$ & $$-$0.050$ & $0.080$ & $974$ & $0.663$ \\ 
-Organic C \% & DBH CoV & $0.100$ & $0.050$ & $0.160$ & $1387$ & $0$ \\ 
-Organic C \% & Stem density & $0.070$ & $0.010$ & $0.120$ & $1387$ & $0.015$ \\ 
-Organic C \% & AGB & $0.200$ & $0.150$ & $0.250$ & $1387$ & $0$ \\ 
-CEC & MAP & $$-$0.060$ & $$-$0.120$ & $$-$0.010$ & $1387$ & $0.016$ \\ 
-CEC & MAP CoV & $$-$0.560$ & $$-$0.590$ & $$-$0.520$ & $1387$ & $0$ \\ 
-CEC & MAT & $0.240$ & $0.190$ & $0.290$ & $1387$ & $0$ \\ 
-CEC & MAT CoV & $0.080$ & $0.030$ & $0.130$ & $1387$ & $0.002$ \\ 
-CEC & Extrap. sp. rich. & $$-$0.070$ & $$-$0.120$ & $$-$0.020$ & $1387$ & $0.011$ \\ 
-CEC & Shannon equit. & $0.240$ & $0.190$ & $0.280$ & $1387$ & $0$ \\ 
-CEC & Tree height CoV & $0.060$ & $0$ & $0.130$ & $974$ & $0.044$ \\ 
-CEC & DBH CoV & $0.250$ & $0.200$ & $0.300$ & $1387$ & $0$ \\ 
-CEC & Stem density & $0.210$ & $0.160$ & $0.260$ & $1387$ & $0$ \\ 
-CEC & AGB & $0.050$ & $0$ & $0.100$ & $1387$ & $0.053$ \\ 
-MAP & MAP CoV & $$-$0.080$ & $$-$0.140$ & $$-$0.030$ & $1387$ & $0.002$ \\ 
-MAP & MAT & $$-$0.080$ & $$-$0.130$ & $$-$0.030$ & $1387$ & $0.003$ \\ 
-MAP & MAT CoV & $$-$0.650$ & $$-$0.680$ & $$-$0.620$ & $1387$ & $0$ \\ 
-MAP & Extrap. sp. rich. & $0.310$ & $0.270$ & $0.360$ & $1387$ & $0$ \\ 
-MAP & Shannon equit. & $$-$0.020$ & $$-$0.080$ & $0.030$ & $1387$ & $0.382$ \\ 
-MAP & Tree height CoV & $0.180$ & $0.120$ & $0.240$ & $974$ & $0$ \\ 
-MAP & DBH CoV & $0.060$ & $0.010$ & $0.110$ & $1387$ & $0.025$ \\ 
-MAP & Stem density & $$-$0.070$ & $$-$0.120$ & $$-$0.010$ & $1387$ & $0.013$ \\ 
-MAP & AGB & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
-MAP CoV & MAT & $$-$0.170$ & $$-$0.220$ & $$-$0.110$ & $1387$ & $0$ \\ 
-MAP CoV & MAT CoV & $0.020$ & $$-$0.040$ & $0.070$ & $1387$ & $0.561$ \\ 
-MAP CoV & Extrap. sp. rich. & $0.080$ & $0.020$ & $0.130$ & $1387$ & $0.005$ \\ 
-MAP CoV & Shannon equit. & $$-$0.230$ & $$-$0.280$ & $$-$0.180$ & $1387$ & $0$ \\ 
-MAP CoV & Tree height CoV & $$-$0.050$ & $$-$0.110$ & $0.010$ & $974$ & $0.102$ \\ 
-MAP CoV & DBH CoV & $$-$0.350$ & $$-$0.390$ & $$-$0.300$ & $1387$ & $0$ \\ 
-MAP CoV & Stem density & $$-$0.250$ & $$-$0.300$ & $$-$0.200$ & $1387$ & $0$ \\ 
-MAP CoV & AGB & $$-$0.100$ & $$-$0.160$ & $$-$0.050$ & $1387$ & $0$ \\ 
-MAT & MAT CoV & $$-$0.330$ & $$-$0.370$ & $$-$0.280$ & $1387$ & $0$ \\ 
-MAT & Extrap. sp. rich. & $$-$0.230$ & $$-$0.280$ & $$-$0.180$ & $1387$ & $0$ \\ 
-MAT & Shannon equit. & $0.010$ & $$-$0.040$ & $0.060$ & $1387$ & $0.775$ \\ 
-MAT & Tree height CoV & $0.100$ & $0.040$ & $0.160$ & $974$ & $0.002$ \\ 
-MAT & DBH CoV & $0.090$ & $0.040$ & $0.150$ & $1387$ & $0$ \\ 
-MAT & Stem density & $0.080$ & $0.030$ & $0.130$ & $1387$ & $0.003$ \\ 
-MAT & AGB & $$-$0.060$ & $$-$0.110$ & $0$ & $1387$ & $0.038$ \\ 
-MAT CoV & Extrap. sp. rich. & $$-$0.240$ & $$-$0.290$ & $$-$0.190$ & $1387$ & $0$ \\ 
-MAT CoV & Shannon equit. & $$-$0.030$ & $$-$0.090$ & $0.020$ & $1387$ & $0.193$ \\ 
-MAT CoV & Tree height CoV & $$-$0.120$ & $$-$0.180$ & $$-$0.060$ & $974$ & $0$ \\ 
-MAT CoV & DBH CoV & $$-$0.020$ & $$-$0.070$ & $0.030$ & $1387$ & $0.428$ \\ 
-MAT CoV & Stem density & $$-$0.070$ & $$-$0.120$ & $$-$0.020$ & $1387$ & $0.010$ \\ 
-MAT CoV & AGB & $$-$0.080$ & $$-$0.140$ & $$-$0.030$ & $1387$ & $0.002$ \\ 
-Extrap. sp. rich. & Shannon equit. & $0.540$ & $0.500$ & $0.580$ & $1387$ & $0$ \\ 
-Extrap. sp. rich. & Tree height CoV & $0.070$ & $0$ & $0.130$ & $974$ & $0.038$ \\ 
-Extrap. sp. rich. & DBH CoV & $0.160$ & $0.110$ & $0.210$ & $1387$ & $0$ \\ 
-Extrap. sp. rich. & Stem density & $0.390$ & $0.350$ & $0.440$ & $1387$ & $0$ \\ 
-Extrap. sp. rich. & AGB & $0.460$ & $0.420$ & $0.500$ & $1387$ & $0$ \\ 
-Shannon equit. & Tree height CoV & $$-$0.060$ & $$-$0.130$ & $0$ & $974$ & $0.046$ \\ 
-Shannon equit. & DBH CoV & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
-Shannon equit. & Stem density & $0.540$ & $0.500$ & $0.580$ & $1387$ & $0$ \\ 
-Shannon equit. & AGB & $0.400$ & $0.350$ & $0.440$ & $1387$ & $0$ \\ 
-Tree height CoV & DBH CoV & $0.530$ & $0.480$ & $0.570$ & $974$ & $0$ \\ 
-Tree height CoV & Stem density & $$-$0.030$ & $$-$0.100$ & $0.030$ & $974$ & $0.311$ \\ 
-Tree height CoV & AGB & $0.210$ & $0.150$ & $0.270$ & $974$ & $0$ \\ 
-DBH CoV & Stem density & $0.190$ & $0.140$ & $0.240$ & $1387$ & $0$ \\ 
-DBH CoV & AGB & $0.420$ & $0.380$ & $0.470$ & $1387$ & $0$ \\ 
-Stem density & AGB & $0.510$ & $0.470$ & $0.540$ & $1387$ & $0$ \\ 
+sand\_per\_std & ocdens\_std & $$-$0.430$ & $$-$0.470$ & $$-$0.390$ & $1387$ & $0$ \\ 
+sand\_per\_std & cation\_ex\_cap\_std & $$-$0.580$ & $$-$0.610$ & $$-$0.540$ & $1387$ & $0$ \\ 
+sand\_per\_std & total\_precip\_std & $$-$0.420$ & $$-$0.460$ & $$-$0.370$ & $1387$ & $0$ \\ 
+sand\_per\_std & precip\_seasonality\_log\_std & $0.380$ & $0.330$ & $0.420$ & $1387$ & $0$ \\ 
+sand\_per\_std & mean\_temp\_std & $0.240$ & $0.190$ & $0.290$ & $1387$ & $0$ \\ 
+sand\_per\_std & temp\_seasonality\_log\_std & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
+sand\_per\_std & sp\_rich\_raref\_log\_std & $$-$0.290$ & $$-$0.340$ & $$-$0.240$ & $1387$ & $0$ \\ 
+sand\_per\_std & shannon\_equit\_log\_std & $$-$0.250$ & $$-$0.300$ & $$-$0.200$ & $1387$ & $0$ \\ 
+sand\_per\_std & cov\_height\_std & $$-$0.070$ & $$-$0.130$ & $$-$0.010$ & $974$ & $0.024$ \\ 
+sand\_per\_std & cov\_dbh\_std & $$-$0.180$ & $$-$0.230$ & $$-$0.130$ & $1387$ & $0$ \\ 
+sand\_per\_std & stems\_ha\_log\_std & $$-$0.220$ & $$-$0.260$ & $$-$0.160$ & $1387$ & $0$ \\ 
+sand\_per\_std & bchave\_log & $$-$0.190$ & $$-$0.240$ & $$-$0.140$ & $1387$ & $0$ \\ 
+ocdens\_std & cation\_ex\_cap\_std & $0.300$ & $0.250$ & $0.340$ & $1387$ & $0$ \\ 
+ocdens\_std & total\_precip\_std & $0.330$ & $0.280$ & $0.380$ & $1387$ & $0$ \\ 
+ocdens\_std & precip\_seasonality\_log\_std & $$-$0.350$ & $$-$0.390$ & $$-$0.300$ & $1387$ & $0$ \\ 
+ocdens\_std & mean\_temp\_std & $$-$0.310$ & $$-$0.350$ & $$-$0.260$ & $1387$ & $0$ \\ 
+ocdens\_std & temp\_seasonality\_log\_std & $0$ & $$-$0.050$ & $0.050$ & $1387$ & $0.996$ \\ 
+ocdens\_std & sp\_rich\_raref\_log\_std & $0.150$ & $0.100$ & $0.200$ & $1387$ & $0$ \\ 
+ocdens\_std & shannon\_equit\_log\_std & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
+ocdens\_std & cov\_height\_std & $0.010$ & $$-$0.050$ & $0.080$ & $974$ & $0.663$ \\ 
+ocdens\_std & cov\_dbh\_std & $0.100$ & $0.050$ & $0.160$ & $1387$ & $0$ \\ 
+ocdens\_std & stems\_ha\_log\_std & $0.070$ & $0.010$ & $0.120$ & $1387$ & $0.015$ \\ 
+ocdens\_std & bchave\_log & $0.200$ & $0.150$ & $0.250$ & $1387$ & $0$ \\ 
+cation\_ex\_cap\_std & total\_precip\_std & $$-$0.060$ & $$-$0.120$ & $$-$0.010$ & $1387$ & $0.016$ \\ 
+cation\_ex\_cap\_std & precip\_seasonality\_log\_std & $$-$0.560$ & $$-$0.590$ & $$-$0.520$ & $1387$ & $0$ \\ 
+cation\_ex\_cap\_std & mean\_temp\_std & $0.240$ & $0.190$ & $0.290$ & $1387$ & $0$ \\ 
+cation\_ex\_cap\_std & temp\_seasonality\_log\_std & $0.080$ & $0.030$ & $0.130$ & $1387$ & $0.002$ \\ 
+cation\_ex\_cap\_std & sp\_rich\_raref\_log\_std & $$-$0.070$ & $$-$0.120$ & $$-$0.020$ & $1387$ & $0.011$ \\ 
+cation\_ex\_cap\_std & shannon\_equit\_log\_std & $0.240$ & $0.190$ & $0.280$ & $1387$ & $0$ \\ 
+cation\_ex\_cap\_std & cov\_height\_std & $0.060$ & $0$ & $0.130$ & $974$ & $0.044$ \\ 
+cation\_ex\_cap\_std & cov\_dbh\_std & $0.250$ & $0.200$ & $0.300$ & $1387$ & $0$ \\ 
+cation\_ex\_cap\_std & stems\_ha\_log\_std & $0.210$ & $0.160$ & $0.260$ & $1387$ & $0$ \\ 
+cation\_ex\_cap\_std & bchave\_log & $0.050$ & $0$ & $0.100$ & $1387$ & $0.053$ \\ 
+total\_precip\_std & precip\_seasonality\_log\_std & $$-$0.080$ & $$-$0.140$ & $$-$0.030$ & $1387$ & $0.002$ \\ 
+total\_precip\_std & mean\_temp\_std & $$-$0.080$ & $$-$0.130$ & $$-$0.030$ & $1387$ & $0.003$ \\ 
+total\_precip\_std & temp\_seasonality\_log\_std & $$-$0.650$ & $$-$0.680$ & $$-$0.620$ & $1387$ & $0$ \\ 
+total\_precip\_std & sp\_rich\_raref\_log\_std & $0.310$ & $0.270$ & $0.360$ & $1387$ & $0$ \\ 
+total\_precip\_std & shannon\_equit\_log\_std & $$-$0.020$ & $$-$0.080$ & $0.030$ & $1387$ & $0.382$ \\ 
+total\_precip\_std & cov\_height\_std & $0.180$ & $0.120$ & $0.240$ & $974$ & $0$ \\ 
+total\_precip\_std & cov\_dbh\_std & $0.060$ & $0.010$ & $0.110$ & $1387$ & $0.025$ \\ 
+total\_precip\_std & stems\_ha\_log\_std & $$-$0.070$ & $$-$0.120$ & $$-$0.010$ & $1387$ & $0.013$ \\ 
+total\_precip\_std & bchave\_log & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
+precip\_seasonality\_log\_std & mean\_temp\_std & $$-$0.170$ & $$-$0.220$ & $$-$0.110$ & $1387$ & $0$ \\ 
+precip\_seasonality\_log\_std & temp\_seasonality\_log\_std & $0.020$ & $$-$0.040$ & $0.070$ & $1387$ & $0.561$ \\ 
+precip\_seasonality\_log\_std & sp\_rich\_raref\_log\_std & $0.080$ & $0.020$ & $0.130$ & $1387$ & $0.005$ \\ 
+precip\_seasonality\_log\_std & shannon\_equit\_log\_std & $$-$0.230$ & $$-$0.280$ & $$-$0.180$ & $1387$ & $0$ \\ 
+precip\_seasonality\_log\_std & cov\_height\_std & $$-$0.050$ & $$-$0.110$ & $0.010$ & $974$ & $0.102$ \\ 
+precip\_seasonality\_log\_std & cov\_dbh\_std & $$-$0.350$ & $$-$0.390$ & $$-$0.300$ & $1387$ & $0$ \\ 
+precip\_seasonality\_log\_std & stems\_ha\_log\_std & $$-$0.250$ & $$-$0.300$ & $$-$0.200$ & $1387$ & $0$ \\ 
+precip\_seasonality\_log\_std & bchave\_log & $$-$0.100$ & $$-$0.160$ & $$-$0.050$ & $1387$ & $0$ \\ 
+mean\_temp\_std & temp\_seasonality\_log\_std & $$-$0.330$ & $$-$0.370$ & $$-$0.280$ & $1387$ & $0$ \\ 
+mean\_temp\_std & sp\_rich\_raref\_log\_std & $$-$0.230$ & $$-$0.280$ & $$-$0.180$ & $1387$ & $0$ \\ 
+mean\_temp\_std & shannon\_equit\_log\_std & $0.010$ & $$-$0.040$ & $0.060$ & $1387$ & $0.775$ \\ 
+mean\_temp\_std & cov\_height\_std & $0.100$ & $0.040$ & $0.160$ & $974$ & $0.002$ \\ 
+mean\_temp\_std & cov\_dbh\_std & $0.090$ & $0.040$ & $0.150$ & $1387$ & $0$ \\ 
+mean\_temp\_std & stems\_ha\_log\_std & $0.080$ & $0.030$ & $0.130$ & $1387$ & $0.003$ \\ 
+mean\_temp\_std & bchave\_log & $$-$0.060$ & $$-$0.110$ & $0$ & $1387$ & $0.038$ \\ 
+temp\_seasonality\_log\_std & sp\_rich\_raref\_log\_std & $$-$0.240$ & $$-$0.290$ & $$-$0.190$ & $1387$ & $0$ \\ 
+temp\_seasonality\_log\_std & shannon\_equit\_log\_std & $$-$0.030$ & $$-$0.090$ & $0.020$ & $1387$ & $0.193$ \\ 
+temp\_seasonality\_log\_std & cov\_height\_std & $$-$0.120$ & $$-$0.180$ & $$-$0.060$ & $974$ & $0$ \\ 
+temp\_seasonality\_log\_std & cov\_dbh\_std & $$-$0.020$ & $$-$0.070$ & $0.030$ & $1387$ & $0.428$ \\ 
+temp\_seasonality\_log\_std & stems\_ha\_log\_std & $$-$0.070$ & $$-$0.120$ & $$-$0.020$ & $1387$ & $0.010$ \\ 
+temp\_seasonality\_log\_std & bchave\_log & $$-$0.080$ & $$-$0.140$ & $$-$0.030$ & $1387$ & $0.002$ \\ 
+sp\_rich\_raref\_log\_std & shannon\_equit\_log\_std & $0.540$ & $0.500$ & $0.580$ & $1387$ & $0$ \\ 
+sp\_rich\_raref\_log\_std & cov\_height\_std & $0.070$ & $0$ & $0.130$ & $974$ & $0.038$ \\ 
+sp\_rich\_raref\_log\_std & cov\_dbh\_std & $0.160$ & $0.110$ & $0.210$ & $1387$ & $0$ \\ 
+sp\_rich\_raref\_log\_std & stems\_ha\_log\_std & $0.390$ & $0.350$ & $0.440$ & $1387$ & $0$ \\ 
+sp\_rich\_raref\_log\_std & bchave\_log & $0.460$ & $0.420$ & $0.500$ & $1387$ & $0$ \\ 
+shannon\_equit\_log\_std & cov\_height\_std & $$-$0.060$ & $$-$0.130$ & $0$ & $974$ & $0.046$ \\ 
+shannon\_equit\_log\_std & cov\_dbh\_std & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
+shannon\_equit\_log\_std & stems\_ha\_log\_std & $0.540$ & $0.500$ & $0.580$ & $1387$ & $0$ \\ 
+shannon\_equit\_log\_std & bchave\_log & $0.400$ & $0.350$ & $0.440$ & $1387$ & $0$ \\ 
+cov\_height\_std & cov\_dbh\_std & $0.530$ & $0.480$ & $0.570$ & $974$ & $0$ \\ 
+cov\_height\_std & stems\_ha\_log\_std & $$-$0.030$ & $$-$0.100$ & $0.030$ & $974$ & $0.311$ \\ 
+cov\_height\_std & bchave\_log & $0.210$ & $0.150$ & $0.270$ & $974$ & $0$ \\ 
+cov\_dbh\_std & stems\_ha\_log\_std & $0.190$ & $0.140$ & $0.240$ & $1387$ & $0$ \\ 
+cov\_dbh\_std & bchave\_log & $0.420$ & $0.380$ & $0.470$ & $1387$ & $0$ \\ 
+stems\_ha\_log\_std & bchave\_log & $0.510$ & $0.470$ & $0.540$ & $1387$ & $0$ \\ 
 \hline \\[-1.8ex] 

### 2019-11-19 manuscript/chapter_1.tex
Part way through editing

@@
 
-A number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Review}.
-Ecosystem functions can be defined in broad terms as rate processes and properties which describe the degree of biotic activity within an ecosystem \citep{Jax2005}.
-This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect aggregate measures of function such as resistance of productivity to disturbance \citep{}.
-Ecosystem functionality may also include static ecosystem properties such as forest canopy complexity and trophic complexity, which in turn influence ecosystem processes.
-The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied \citep{}, the ecosystem function(s) of interest \citep{}, and the inclusion of environmental covariates in statistical models \citep{}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}.
+A large number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Liang2016, Cardinale2009}.
+The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}.
 Over the past decade, many observational studies of the BEFR have been conducted, mostly in tropical and temperate forests, and grasslands \citep{Chen2011}.
-These studies support previous findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+
+Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}.
+This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect aggregate measures of function such as resistance of productivity to disturbance, and even to static properties which themselves influence process, such as trophic complexity and total vegetative biomass.
+The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.
+
+The relationship between biodiversity and ecosystem functionality has informed conservation policy which aims to maintain and strengthen biodiversity under the assumption that this strengthens ecosystem resilience to stressors such as anthropogenic climate change and ecosystem service use by humans \citep{Thompson2011}.
+It is probable however, that the strength and relative importance of the BEFR varies with ecosystem studied and environmental conditions, with many ecosystems remaining understudied in the BEFR literature.
 
 The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage.
-In wet tropical forests, much of the observed effect of biodiversity on ecosystem function has been attributed to niche complementarity \citep{}, whereby competition between individuals is reduced due to them occupying different niche space, allowing diverse species assemblages to take greater advantage of natural resources \citep{}.
-In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as much of a role in the relationship between biodiversity and ecosystem function.
-Instead, stress tolerance and productivity of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{}.
-Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{}.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage, being an oceanic island with many endemic species.
+In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, MORE}.
+In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function.
+Instead, stress tolerance and the functional contribution of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{}.
+Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.
 
-\todo{Re-assess this paragraph -}In the seasonally dry woodlands of southern Africa, human actions are driving rapid changes in biodiversity.
-This has in turn raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}.
-Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Ryan2016}.
-Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality \citep{}.
-Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016}.
-The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and are of global importance to the carbon cycle.
+Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{plot_loc}).
+The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{}.
 Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}.
 Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}.
-These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}.
+In these seasonally dry woodlands, human actions are driving rapid changes in biodiversity.
+Southern African woodlands are also relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
+This has raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}.
+Resource extraction by humans is directly influencing biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{ANOTHER, Ryan2016}.
+Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{}.
+While rapid biodiversity change is being observed in southern African woodlands, research into the relationship between biodiversity and ecosystem functionality remains scarce.
+
+A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites.
+Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass \citep{}, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass.
+Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{}.
+Small scale studies may therefore be over-estimating the effect of diversity.
+
+%These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
-A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, and, others}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites, meaning that generalisable regional trends could not be reliably inferred.
-\todo{Say how the previous studies might have been wrong because of their limited scope, or how in order to really answer the question you need to sample long environmental gradients.}
 
-The role of environmental factors in mediating the BEFR generally is poorly understood \citep{Ratcliffe2017}.
-In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity \citep{}, so it is important to account for these climatic factors and understand how they interact with biomass and diversity to effectively model the effects of biodiversity on biomass stocks in analyses at large spatial scales in savannas.
+In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{}, so it is important to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on biomass in analyses at large spatial scales in woodlands.
 \citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
@@
 
-\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution in Panamanian tropical forests, which may affect ecosystem level productivity and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under specific environmental conditions.
-In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments, suggesting variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}.
-Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}.
-
-Temperature imposes a similar physiological limit on woody growth rate by limiting metabolic rates.
+Temperature and water availability impose physiological limits on woody growth rate by limiting metabolic rates.
 Temperature and water availability together impact the transpiration rate of a tree, with high temperatures and low water availability limiting growth and potentially causing damage or mortality through cavitation of vessels within the tree \citep{Rowland2015a, Fensham2009}.
 In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}, so this effect may not be present in historically droughted areas.
-The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it.
+The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it with woodlands existing at the climatic extremes of their bioclimatic range experiencing greater reductions in productivity and standing biomass \citep{}.
 Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem functionality via two effects, the direct effect of resource availability in the form of moisture availability and soil nutrients, and indirectly through the effect of these environmental factors on species composition.
-Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness \textit{per se}.
+Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness.
+
+Environmental conditions may indirectly moderate the observed relationship between biodiversity and ecosystem function.
+\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in Panamanian tropical forests.
+Dry season length may therefore indirectly affect ecosystem level productivity in forest ecosystems and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under specific environmental conditions.
+In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
+They suggested variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this moderation relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}.
+Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}.
 
@@
 
-\todo{Species abundance evenness and selection effects.}
-
-\todo{Stand structural complexity}
-
 While southern African woodlands are species rich in the herbaceous understorey, the tree layer is relatively species poor \citep{}.
-\citet{Solbrig1996} however, writes that they are structurally diverse in the tree layer, with trees of the same species occupying distinct layers of the canopy at different growth stages.
-This structural diversity may be one mechanism through which species diversity appears to influence woody biomass.
-Trees occupying different canopy layers can create a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area.
-An underpinning principle of the Biodiversity-Ecosystem Function Relationship (BEFR) is that of niche complementarity.
-Species are assumed to differ in their functional niche by the fact that they may coexist in a steady-state ecological community, therefore the more species are present the greater the area of the total fundamental niche of the ecosystem is filled, allowing more efficient use of natural resources and higher gross primary productivity and biomass \citep{}.
-This theory however, which has been supported by many experiments and observational studies in temperate and wet tropical ecosystems, may not hold in savannas, which are structured by disturbance rather than competition.
+\citet{Solbrig1996} however, writes that they remain structurally diverse in the tree layer, with trees occupying distinct layers of the canopy at different growth stages and between species.
+This structural diversity may be one mechanism through which diversity influences woody biomass.
+\citep{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest.
+They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production.
+Tree species in southern African woodlands may differ in their occupation of different canopy layers allowing a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity.
+This theory however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{}, may not occur in savannas, which are structured by disturbance rather than competition.
+Instead, disturbance history may override the effects of species diversity on structural diversity nulligying the effects of species diversity on structural diversity.
 
@@
 
-In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{}).
+In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{plot_loc}, \autoref{temp_precip_hull}).
 We investigated the relationship between aboveground woody biomass and tree biodiversity.
-We compared the relative effects of tree species biodiversity with that of other environmental factors known to affect ecosystem productivity and biomass accumulation such as precipitation and temperature.
+We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation: precipitation, temperature and soil fertility.
 We also investigated potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass.
@@
 	\item{Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.}
-	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above-ground woody biomass.} 
-	\item{There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species diversity.}
-	\item{Structural characteristics of the woodland will interact with species diversity to provide an indirect path of influence between species composition and biomass stocks.}
+	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species diversity on above-ground woody biomass.} 
+	\item{There will be a positive effect of precipitation and soil fertility on AGB. Part of this positive effect will exist as a mediating effect through tree species diversity.}
+	\item{Tree structural diversity of the woodland will interact with tree species diversity to provide an indirect path of influence between species diversity and woody biomass stocks.}
 	\item{The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.}
@@
 
-The study used \nplots{} woodland monitoring plots from a larger network located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}).
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{SEOSAW} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}).
 The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
@@
 
-Climatic data was collected from the ECMWF ERA5 dataset \citep{}.
+Climatic data was collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}.
 Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
@@
 
-Observed variables were standardized to Z-scores for the SEMs.
+Observed variables were standardised to Z-scores for the SEMs.
 Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
@@
 	\caption{Correlogram of standardised observed variables used in the SEMs, with pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
-Variables in bold on the axis labels were later reversed for SEMs to maintain positive correlations between observed variables within each latent variable.
+Variables in bold and underlined on the axis labels were later reversed for SEMs to maintain positive correlations between observed variables within each latent variable.
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
@@
 Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}).
+The R-squared of AGB was \strucrsq{}
 
@@
 	\includegraphics[width=\textwidth]{struc_model_slopes}
-	\caption{Standardized path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
-Due to all observed variables being standardized and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
+	\caption{Standardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
+Due to all observed variables being standardised and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
 Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
@@
 
-When the structural diversity and species diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardized path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
-It should be noted that models for vegetation clusters 4 and 5 did not converge, possibly due to small sample sizes (\autoref{struc_model_slopes_all}).
+When the structural diversity and species diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
+All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for clusters 3, 4 and 5 indicate a possible issue of low sample size for these clusters.
+
+Cluster 5, which contained vegetation dominated by \textit{Colophospermum mopane} had no effect of tree species diversity on either AGB or stem density, in contast to the other vegetation clusters, but had a similar positive effect of structural diversity on AGB (\autoref{struc_model_slopes_all}).
+
+The strongest total effect of tree species diversity on AGB was in cluster 3, which was relatively species poor compared to other vegetation clusters, dominated by \textit{Burkea africana}, \textit{Pterocarpus angolensis}, and \textit{Baikiaea plurijuga}, which are all large growing canopy tree species \citep{}.
+
+The greatest effect of structural diversity was found in cluster 4, the most diverse vegetation cluster, with a median tree species richness of 11 species, dominated by the large growing \textit{Baikaea plurijuga}.
+Cluster 4 also had the highest median above-ground biomass and the highest median stem density (Table \autoref{clust_summ}).
+The effects of structural diversity on AGB were highly variable.
 
@@
 	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
-Path coefficients are +/- 1 standard error.
+Path coefficients are $\pm$ 1 standard error.
 Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
@@
 	\includegraphics[width=\textwidth]{full_model_slopes}
-	\caption{Standardized path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots.
-Path coefficients are +/- 1 standard error.
+	\caption{Standardised path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots.
+Path coefficients are $\pm$ 1 standard error.
 Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
@@
 Using latent variables and SEMs, we found support for a general relationship between tree species diversity and AGB, and a relationship between tree structural diversity and AGB that was independent of tree species diversity.
+The strongest effect on AGB was that of stem density, which itself was highly dependent on species diversity.
 Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB.
+We did find however that there were weak indirect moderation effects of both soil fertility and moisture availability on the strength of the relationship between species diversity and AGB.
+We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB.
 
@@
 
-The latent variable of tree species diversity had the strongest positive effect on woody above-ground biomass (AGB) in environment+diversity SEM model (\autoref{full_mod}).
+The latent variable of tree species diversity had the strongest positive effect on woody above-ground biomass (AGB) in the environment+diversity SEM model (\autoref{full_mod}).
 Within the savanna woodlands of southern Africa we therefore find support for our hypothesis (H\textsubscript{1}), that higher tree species richness causes higher woody AGB.
-This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{}.
+This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}.
 
 Most of the total species diversity effect was through the indirect effect via stem density.
-This suggests that a higher species diversity allows for a greater density of tree stems and that in southern African woodlands, this leads to an increase in total AGB.
-The effect of increased stem density within forests is variable.
-The question of whether lots of small stems or a few large stems maximises AGB has yet to be solved \citep{}.
+This suggests that within southern African woodlands a higher species diversity allows for a greater density of tree stems and this leads to an increase in total AGB.
+While reverse causation with increased stem density causing higher species richness through an increased probability of encountering new species, we accounted for this variation in the extrapolation of species richness according to stem density (sampling effort), suggesting that the observed direction of causation is real.
+We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
+We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level.
+
+% The question of whether lots of small stems or a few large stems maximises AGB has yet to be solved \citep{}.
 
 While we did not explicitly measure net primary production (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{}.
-Mechanistically, a higher standing woody biomass results in a greater stand level leaf area index (LAI) \citep{}, with leaf area being a primary determinant of productivity \citep{}.
-
-As tree species diversity increased, this allowed for greater stem density which in turn led to greater AGB.
+This suggests that as has been found in many other woodland/forest ecosystems, woody biomass in southern African woodlands can be maximised by increasing species diversity.
 
@@
 Structural diversity did however have a positive effect on AGB.
-Our original hpyothesis was that an increase in tree species richness would increase the number of functional forms of trees per unit area due to niche complementarity in growth form of coexisting species.
-However, it seems more likely that structural diversity is instead merely a result of disturbance history.
-With highly disturbed plots generally having a less structurally diverse canopy.
+Our original hypothesis was that an increase in tree species richness would increase the diversity of available tree functional forms within a plot due and that this would act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
+Given our results however, it seems more likely that structural diversity is instead merely a result of disturbance history, with highly disturbed plots generally having a less structurally diverse canopy \citep{}.
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges.
-In highly disturbed woodland plots, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained.
+In highly disturbed woodland plots, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained \citep{}.
+Unfortunately we did not have access to a dataset of high enough spatial and temporal resolution to adequately quantify the effects of disturbance history.
 
@@
 
-Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via tree species diversity and stand structure \todo{what sort of stand structure?} \citep{Ali2019a, Ali2019b, Poorter2017}.
+Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via tree species diversity and stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}.
 This was not observed in our study however.
-Compared to most tropical forests, climatic water availability is far more of a limiting factor in southern African woodlands.
+Compared to moist tropical forests, climatic water availability is far more of a limiting factor to growth in southern African woodlands.
 We expect that an increase in water availability in southern African woodlands would allow for greater tree growth and therefore greater AGB.
-The range of observed moisture availability in this study however may not have been able to capture this effect.
-Instead, in southern African woodlands moisture availability is much more closely linked with disturbance from seasonal fires.
-The growth of C4 grasses in wetter savannas leads to seasonal fires which limit tree growth.
-Additionally, we may see high local adaptation of tree species andf compositional turnover in response to variation in moisture availability, negating any effects of moisture availability on tree growth.
-Savannas in Africa have been dry for a long time so it is understnadable that trees may be highly adapted to low rainfall to the point where it doesn't affect their growth.
+The range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture this effect.
 
-In a separate set of linear multiple regressions, we found a weak positive interaction effect of moisture availabiltiy on the relationship between tree species diversity and AGB.
+In southern African woodlands moisture availability is closely linked with the intensity of disturbance from seasonal fires.
+The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth.
+Additionally at the large regional scale used in this study, we see high local adaptation of tree species and compositional turnover in response to variation in moisture availability, possibly negating any effects of moisture availability on tree growth.
+
+In a separate set of linear multiple regressions, we found a weak positive interaction effect of moisture availability on the relationship between tree species diversity and AGB.
 As moisture availability increased, the relationship between tree species diversity and AGB became stronger.
-We could not add moderation effects to the SEM framework.
 
 \subsection{Vegetation type cluster specific responses}
+
+We found a clear difference in the response of biomass to tree species richness in vegetation cluster 5 compared to other vegetation clusters (\autoref{struc_model_slopes_all}).
 
@@
 There may also be an effect whereby the high stem density plots have lots of very small stems, meaning that AGB cannot be high.
-The largest trees always ahve the bulk of the biomass in a system.
+The largest trees always have the bulk of the biomass in a system.
 Again, this is indicative of thicket vegetation.
@@
 % Seidel 2019: Large trees tend to possess a greater structural complexity than small trees, not due to their size per se, but due to more complex architecture.
+
+% At very high biomass, tree species diversity is low because it is associated with intense competition for light leading to dominance of usually a single species (Huston 1997, Oindo and Skidmore 2002).
 
@@
 	\input{include/data_clean_flow}
-	\caption{Flow diagram of the data filtering and cleaning process prior to analysis. Rounded boxes indicate filtering events while normal boxes indicate calculation events.}
+	\caption{Flow diagram of the data filtering and cleaning process prior to analysis. Rounded boxes indicate filtering events while regular boxes indicate calculation events.}
 	\label{data_clean_flow}

### 2019-11-19 manuscript/include/clust_summ.tex
Part way through editing

@@
 Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv}.
-Numeric values are means $\pm$ 1 standard deviation.}
+Numeric values are medians and interquartile ranges.}
   \label{clust_summ} 
@@
 \hline \\[-1.8ex] 
-C1 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Combretum }spp.
-\\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 7$\pm$4.7 & 276$\pm$345.2 & 35.3$\pm$37.03 \\
+C1 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Combretum }spp. \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 7(5) & 316(228.6) & 37.1(27.66) \\ 
 \hline
-C2 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Albizia antunesiana} \end{tabular} & 8$\pm$4.3 & 274$\pm$245.9 & 46.8$\pm$35.05 \\ 
+C2 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Albizia antunesiana} \end{tabular} & 8(5) & 291(191.7) & 43.8(40.28) \\ 
 \hline
-C3 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \\ \textit{Baikiaea plurijuga} \end{tabular} & 4$\pm$2.3 & 274$\pm$536.5 & 36.3$\pm$27.86 \\ 
+C3 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \\ \textit{Baikiaea plurijuga} \end{tabular} & 5(2) & 362(259.4) & 37(42.59) \\ 
 \hline
-C4 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Baikiaea plurijuga} \\ \textit{Terminalia randii} \\ \textit{Albizia amara} \end{tabular} & 10$\pm$5.9 & 447$\pm$386.3 & 47$\pm$45.41 \\ 
+C4 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Baikiaea plurijuga} \\ \textit{Terminalia randii} \\ \textit{Albizia amara} \end{tabular} & 11(9) & 495(516.8) & 51.7(50.74) \\ 
 \hline
-C5 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Colophospermum mopane} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 4$\pm$2.9 & 322$\pm$390.1 & 43.8$\pm$34.45 \\ 
+C5 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Colophospermum mopane} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 5(5) & 424(226.8) & 50.9(32.7) \\ 
 \hline \\[-1.8ex] 

### 2019-11-19 analysis/sem.R

@@ -67,20 +67,7 @@ corr_df <- sem_data %>%
     total_precip_std, precip_seasonality_log_std,
     mean_temp_std, temp_seasonality_log_std,
     sp_rich_raref_log_std, shannon_equit_log_std,
-    cov_height_std, cov_dbh_std, stems_ha_log_std, bchave_log) %>%
-  rename(`Sand %` = sand_per_std,
-    `Organic C %` = ocdens_std,
-    `CEC` = cation_ex_cap_std,
-    `MAP` = total_precip_std, 
-    `MAP CoV` = precip_seasonality_log_std, 
-    `MAT` = mean_temp_std, 
-    `MAT CoV` = temp_seasonality_log_std,
-    `Extrap. sp. rich.` = sp_rich_raref_log_std, 
-    `Shannon equit.` = shannon_equit_log_std,
-    `Tree height CoV` = cov_height_std, 
-    `DBH CoV` = cov_dbh_std,
-    `Stem density` = stems_ha_log_std,
-    `AGB` = bchave_log)
+    cov_height_std, cov_dbh_std, stems_ha_log_std, bchave_log)
 
 # Make a dataframe of labels
 label_df <- data.frame(lab = names(corr_df),
@@ -111,6 +98,32 @@ ggcorrplot(cor(corr_df, use = "complete.obs"),
   show.legend = FALSE,
   outline.color = "black",
   digits = 2, lab_size = 3) + 
+  scale_y_discrete(labels = c(
+    "sand_per_std" = expression(bolditalic(underline("Sand %"))),
+    "ocdens_std" = "Org. C %",
+    "cation_ex_cap_std" = "CEC", 
+    "total_precip_std" = "MAP",
+    "precip_seasonality_log_std" = expression(bolditalic(underline("MAP CoV"))), 
+    "mean_temp_std" = expression(bolditalic(underline("MAT"))),
+    "temp_seasonality_log_std" = expression(bolditalic(underline("MAT CoV"))),
+    "sp_rich_raref_log_std" = "Extrap. sp. rich.",
+    "shannon_equit_log_std" = "Shannon equit.",
+    "cov_height_std" = "Tree height CoV",
+    "cov_dbh_std" = "CBH CoV",
+    "stems_ha_log_std" = "Stem density")) + 
+  scale_x_discrete(labels = c(
+    "ocdens_std" = "Org. C %",
+    "cation_ex_cap_std" = "CEC", 
[…]

### 2019-11-19 analysis/output/include/clust_summ.tex
Finished editing

--- 2019-11-13 analysis/output/include/clust_summ.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Fri, Nov 08, 2019 - 11:40:45
+% Date and time: Tue, Nov 19, 2019 - 14:43:03
 \begin{table}[!htbp] \centering 
@@
   \label{clust_summ} 
-\begin{tabular}{@{\extracolsep{5pt}} ccccc} 
+\begin{tabular}{@{\extracolsep{5pt}} cccccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-clust5 & c\_ind & sp\_rich\_raref & n\_stems\_ha & bchave \\ 
+clust5 & c\_ind & n\_plots & sp\_rich\_raref & n\_stems\_ha & bchave \\ 
 \hline \\[-1.8ex] 
-C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 7+4.7 & 276+345.2 & 35.3+37.03 \\ 
-C2 & Julbernardia paniculata, Isoberlinia angolensis, Albizia antunesiana & 8+4.3 & 274+245.9 & 46.8+35.05 \\ 
-C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 4+2.3 & 274+536.5 & 36.3+27.86 \\ 
-C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 10+5.9 & 447+386.3 & 47+45.41 \\ 
-C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 4+2.9 & 322+390.1 & 43.8+34.45 \\ 
+C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 478 & 7(5) & 316(228.6) & 37.1(27.66) \\ 
+C2 & Julbernardia paniculata, Isoberlinia angolensis, Albizia antunesiana & 617 & 8(5) & 291(191.7) & 43.8(40.28) \\ 
+C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 62 & 5(2) & 362(259.4) & 37(42.59) \\ 
+C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 219 & 11(9) & 495(516.8) & 51.7(50.74) \\ 
+C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 65 & 5(5) & 424(226.8) & 50.9(32.7) \\ 
 \hline \\[-1.8ex] 

### 2019-11-19 manuscript/chapter_1.tex
Finished editing

@@
 
+% STEM DENSITY
+
 In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{plot_loc}, \autoref{temp_precip_hull}).
@@
 Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup.
-Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature \citep{}.
+Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature \citep{Ratnam2011, Hill2010}.
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were exlcuded from the initial pool.
 Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{}, leading to inaccurate estimates.
-Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
+Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014} (\autoref{data_clean_flow}).
 
@@
 We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
-We set the threshold value to remove the top 5\% of plots with the highest nearest neighbour distances \citep{Otto}, removing \noutliers{} plots.
+We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances \citep{Otto}, removing \noutliers{} plots.
 
@@
 
-We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{}.
+We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
 Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
-Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems forming a very small proportion of the total AGB in a plot \citep{}.
+Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems comprising a very small proportion of the total AGB in a plot \citep{}.
 
@@
 A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
-For each stem, species, DBH and tree height to the top of the highest branch material were recorded.
-Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks, which may cause varying measurement uncertainty by data source in the dataset.
+For each stem we measured species, DBH and tree height to the top of the highest branch material.
+Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks.
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
@@
 
-Aboveground biomass for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}.
+Woody AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}.
 Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
@@
 \begin{equation}
-	AGB_{est} = 0.0673 \times (\rho D^{2} H)^{0.976}
+	AGB = 0.0673 \times (\rho D^{2} H)^{0.976}
 	\label{chave_agb}
@@
 Climatic data was collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}.
-Values of mean annual temperature (MAT) and mean annual precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
-Temperature and precipitation seasonality were both calculated as the coefficient of variation of the daily data across the 18 years of available data.
+Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
+Temperature and precipitation seasonality were both calculated as the coefficient of variation of daily MAT and MAP, respectively, across the 18 years of available data.
 Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}.
-We extracted Cation Exchange Capacity (CEC), organic carbon density, and the percentage sand content of the soil.
+We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org.
+C \%), and percentage soil sand content by volume (Sand \%).
 These data are a modelled product compiled from various remote sensed and directly measured data sources.
@@
 Extrapolated species richness was preferred over raw Hill numbers as they are more interpretable, representing actual species numbers.
-Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort across the region.
-Larger plots will tend to encompass more individuals, and therefore probabilistically more species \citep{}.
+Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort.
+Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.
 
@@
 
-Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot.
-
-We calculated tree stand structural diversity for each plot by calculating the coefficient of variation of DBH and tree height.
+Where $H'_{e}$ is an estimation of the Shannon diversity index of tree species by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot.
+We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH and tree height.
 
@@
 Plots were assigned to vegetation type groups based on tree species composition.
-Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas with tree species diversity and relative abundance data.
+Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
 Groups were identified using unconstrained correspondence analysis and ordination.
@@
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
-A key aspect to our decision to use SEMs is that they can explicitly model indirect effects, which is impossible in multiple regression.
-Using SEMs also allowed us to describe theoretical latent constructs which have been suggested by theory to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset.
-For example, moisture availability is known to affect productivity \citep{}, but it's actual value is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration.
+A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is impossible in multiple regression.
+Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset.
+For example, moisture availability is ecpected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration.
 Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
 
-Prior to model fitting, we specified a conceptual model with latent variables based on unmeasured factors known to affect AGB: moisture availability, soil fertility, tree species diversity and tree structural diversity (\autoref{con_mod}).
+Prior to analysis, we specified a conceptual model with factors expected to affect AGB: moisture availability, soil fertility, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).
 
@@
 	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural iversity, tree stem density, and AGB.
-Hypothesised paths of causal inference are depicted as arrows from predictor to response.}
+Hypothesised paths of causation are depicted as arrows from predictor to response.}
 	\label{con_mod}
@@
 
-Observed variables were standardised to Z-scores for the SEMs.
+Observed variables were standardised to Z-scores prior to analysis.
 Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
-Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{}.
-Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan| \citep{}.
+Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
+Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
 To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1.
-For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed.
-Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way.
+For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed for model fitting.
+Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way to account for the direction of their effect on moisture availability.
 
@@
 We therefore allowed factor loadings to be estimated by the SEM itself.
+We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, finding no significant difference between model specifications.
+Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.
 
 First, we assessed the interacting factors of structural diversity and species diversity in determining AGB.
-We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity (see orange box \autoref{con_mod}).
+We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
 To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately.
-We compared path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between structural diversity, species diversity and AGB.
+We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB.
+Path coefficients show the effect of a path with other paths of inference held constant \citep{}.
 Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
@@
 
+To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create 900 datasets of similar size with varying median stem density.
+We used each of these datasets to fit the model including only tree species and structural diversity latent variables to predict AGB.
+We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled dataset.
+
 Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density.
-
-For models of the whole region which included all suitable plots, standardised path coefficients are reported in order to better compare the relative contribution of different paths within the model.
-When comparing the path coefficients of models using individual vegetation clusters, unstandardised path coefficients are used in order to more readily compare the same path coefficient across different models while accounting for variation in mean values between biogeographical clusters.
-Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.
-
-To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create 900 datasets of similar size with varying median stem density.
-We used each of these datasets to fit the model including only tree species and structural diversity latent variables to predict AGB (see orange in \autoref{con_mod}).
-We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of the dataset.
+We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB.
 
@@
 Instead we manually calculated interaction variables for both soil fertility and moisture availability from the product of predicted values of these latent variables in a Confirmatory Factor Analysis (CFA).
-Interaction variables were the products: species diversity $\times$ moisture availability, and species diversity $\times$ soil fertility.
-These interaction variables were included as explanatory variables in multiple regressions alongside the latent variable of species diversity to predict AGB.
-Regression coefficients and model fit were analysed to determine the presence of interaction effects.
+Interaction variables were the products: tree species diversity $\times$ moisture availability, and tree species diversity $\times$ soil fertility.
+These interaction terms were included as explanatory terms in multiple regressions alongside the latent variable of species diversity to predict AGB.
+Regression coefficients and model fit were analysed to determine the presence and form of interaction effects.
 
@@
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
-Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = species diversity, purple = structural diversity.
+Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = tree species diversity, purple = tree structural diversity.
 See \autoref{corr_ci_tab} for a full table of correlation fit statistics.}
@@
 
-A simple set of correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
+ Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
 Stem density had the strongest correlation with AGB of all variables ($\rho$ = 0.51, n = 1387, p <0.001).
@@
 As expected, there was a positive correlation between MAP and AGB, and a negative correlation between the seasonality of precipitation and AGB.
-MAT and the seasonality of temperature variation did not correlate with AGB.
+MAT and the seasonality of temperature did not correlate significantly with AGB.
 As expected, there was a negative correlation between soil sand content and AGB, and a positive correlation between soil organic Carbon AGB.
@@
 Due to all observed variables being standardised and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
-Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
+Path coefficients where the standard error does not overlap zero are considered significant effects.}
 	\label{struc_model_slopes}
@@
 
-When the structural diversity and species diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
-All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for clusters 3, 4 and 5 indicate a possible issue of low sample size for these clusters.
+When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
+All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for Clusters 3, 4 and 5 indicate a possible issue of low sample size for these clusters.
 
-Cluster 5, which contained vegetation dominated by \textit{Colophospermum mopane} had no effect of tree species diversity on either AGB or stem density, in contast to the other vegetation clusters, but had a similar positive effect of structural diversity on AGB (\autoref{struc_model_slopes_all}).
+Cluster 5, which contained vegetation dominated by \textit{Colophospermum mopane} showed no effect of tree species diversity on either AGB or stem density, in contast to the other vegetation clusters, but had a similar positive effect of structural diversity on AGB (\autoref{struc_model_slopes_all}).
 
-The strongest total effect of tree species diversity on AGB was in cluster 3, which was relatively species poor compared to other vegetation clusters, dominated by \textit{Burkea africana}, \textit{Pterocarpus angolensis}, and \textit{Baikiaea plurijuga}, which are all large growing canopy tree species \citep{}.
+The strongest total effect of tree species diversity on AGB was in Cluster 3, which was relatively species poor compared to other vegetation clusters, dominated by \textit{Burkea africana}, \textit{Pterocarpus angolensis}, and \textit{Baikiaea plurijuga}, which are all large growing canopy tree species \citep{}.
 
-The greatest effect of structural diversity was found in cluster 4, the most diverse vegetation cluster, with a median tree species richness of 11 species, dominated by the large growing \textit{Baikaea plurijuga}.
+The greatest effect of structural diversity was found in Cluster 4, the most diverse vegetation cluster, with a median tree species richness of 11 species, dominated by the large growing \textit{Baikaea plurijuga}.
 Cluster 4 also had the highest median above-ground biomass and the highest median stem density (Table \autoref{clust_summ}).
@@
 We repeatedly sub-sampled the dataset of plots to build many datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
-The same SEM specification was used as above except that stem density was removed, incorporating the effects of species diversity and structural diversity on AGB (\autoref{con_mod}).
+The same SEM specification was used as above (\autoref{struc_mod}) except that stem density was removed, incorporating the effects of species diversity and structural diversity on AGB (\autoref{con_mod}).
 \autoref{sem_struc_stems_ha} shows that there is a peak in the total effect of tree species diversity on biomass at \textapprox{}700 stems ha\textsuperscript{-1}.
@@
 	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves.}
+	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves with standard error shaded bars.}
 	\label{sem_struc_stems_ha}
@@
 A model incorporating the latent variables of moisture availability and soil fertility showed that the effect of diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{path_diagram_full}).
-Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (\autoref{full_model_slopes}).
+Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}) (\autoref{full_model_slopes}).
 
-Moisture availability and soil fertility also had negligible direct effects on stem density, but both had a positive total effect on stem density via species diversity, where species diversity itself had a strong positive effect on stem density.
+Moisture availability and soil fertility also had negligible direct effects on stem density, but both had a positive total effect on stem density via species diversity, where species diversity itself had a strong positive effect on stem density (\rgid{}).
 As species diversity increases, stem density also increases.
 
-Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was very weakly positive, while structural diversity itself had a positive effect on AGB.
-Again, stem density had a positive direct effect on AGB.
+Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was very weakly positive, while structural diversity itself had a positive effect on AGB (\rghb{}).
 
@@
 	\input{include/path_diagram_full}
-	\caption{Path diagram with regression coefficients for the full model incorporating environmental covariates and tree species diversity in all five clusters.
+	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation type clusters.
 Latent variables are shown as circles while observed variables are shown as rectangles.
@@
 Path coefficients are $\pm$ 1 standard error.
-Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
+Path coefficients where the standard error does not overlap zero are considered significant effects.}
 	\label{full_model_slopes}
@@
 Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass.
-Regression coefficient for the interaction effect of tree species diversity and moisture availability was significant ($\beta$ = 0.058, t(1383) = 2.562, p <0.05), while the interaction term of tree species diversity and soil fertility was not ($\beta$ = 0.034, t(1383) = 1.452, p = 0.147).
+The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant ($\beta$ = 0.058, t(1383) = 2.562, p <0.05), while the interaction term of tree species diversity and soil fertility was not ($\beta$ = 0.034, t(1383) = 1.452, p = 0.147).
 As moisture availability increased, the effect of tree species diversity on AGB became stronger.
-At high tree species diversity, an increase in resource availability led to a larger increase in biomass.
 
@@
 	}
-	\caption{Interaction plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b).
+	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b).
 All variables are standardised and centred to a mean of zero and a standard deviation of 1.
 Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2.
-Moisture availability, tree species diversity and soil fertility are the values of latent variables the same as those used in the SEMs.}
+Moisture availability, tree species diversity and soil fertility are latent variables comprised of the same observed variable loadings.}
 	\label{int_plots}
@@
 In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, and [d] stem density on aboveground woody biomass (AGB) across southern African woodlands.
-Using latent variables and SEMs, we found support for a general relationship between tree species diversity and AGB, and a relationship between tree structural diversity and AGB that was independent of tree species diversity.
+Using latent variables and SEMs, we found support for a general relationship between tree species diversity and AGB (H\textsubscript{1}), and a relationship between tree structural diversity and AGB that was independent of tree species diversity (H\textsubscript{4}).
 The strongest effect on AGB was that of stem density, which itself was highly dependent on species diversity.
-Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB.
-We did find however that there were weak indirect moderation effects of both soil fertility and moisture availability on the strength of the relationship between species diversity and AGB.
-We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB.
+Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{3}).
+We did find however that there were weak indirect moderation effects of both soil fertility and moisture availability on the strength of the relationship between species diversity and AGB (H\textsubscript{2}).
+We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB, particularly that of \textit{Mopane} woodland (vegetation type cluster C5).
 
@@
 
-The latent variable of tree species diversity had the strongest positive effect on woody above-ground biomass (AGB) in the environment+diversity SEM model (\autoref{full_mod}).
+The latent variable of tree species diversity had a strong positive effect on woody above-ground biomass (AGB) in the environment+diversity SEM model (\autoref{full_mod}).
 Within the savanna woodlands of southern Africa we therefore find support for our hypothesis (H\textsubscript{1}), that higher tree species richness causes higher woody AGB.
@@
 Structural diversity did however have a positive effect on AGB.
-Our original hypothesis was that an increase in tree species richness would increase the diversity of available tree functional forms within a plot due and that this would act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
+Our original hypothesis (H\textsubscript{4}) was that an increase in tree species richness would increase the diversity of available tree functional forms within a plot due and that this would act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
 Given our results however, it seems more likely that structural diversity is instead merely a result of disturbance history, with highly disturbed plots generally having a less structurally diverse canopy \citep{}.
-In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges.
-In highly disturbed woodland plots, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained \citep{}.
+In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{}.
+In highly disturbed woodland plots, we suggest that disturbance history may be an important unmeasured abiotic factor which may account for a high proportion of the variation in AGB \citep{} and in tree species diversity.
+In highly disturbed forests, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained \citep{}.
 Unfortunately we did not have access to a dataset of high enough spatial and temporal resolution to adequately quantify the effects of disturbance history.
@@
 This was not observed in our study however.
-Compared to moist tropical forests, climatic water availability is far more of a limiting factor to growth in southern African woodlands.
+Compared to moist tropical forests, climatic water availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted \citep{}.
 We expect that an increase in water availability in southern African woodlands would allow for greater tree growth and therefore greater AGB.
-The range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture this effect.
+The range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture this effect however.
 
 In southern African woodlands moisture availability is closely linked with the intensity of disturbance from seasonal fires.
-The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth.
-Additionally at the large regional scale used in this study, we see high local adaptation of tree species and compositional turnover in response to variation in moisture availability, possibly negating any effects of moisture availability on tree growth.
+The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth \citep{}.
+Additionally at the large regional scale used in this study, we see high local adaptation of tree species and compositional turnover in response to variation in moisture availability, possibly negating any effects of moisture availability on tree growth \citep{}.
 
@@
 As moisture availability increased, the relationship between tree species diversity and AGB became stronger.
+This is in contrast to \citet{Ratcliffe2017} who found that in European forests, a decrease in water availability due to drought led to a stronger effect of tree species diversity on AGB.
+They attributed this to an increase in selection effects which allowed dominance of stress tolerant species which were more likely to occur in high diversity assemblages.
 
@@
 
-We found a non linear effect of stem density on the relationship between tree species diversity and AGB.
+We found a non linear effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
 The effect of tree species diversity on AGB peaked at \textapprox{}700 stems ha\textsuperscript{-1}.
-At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness does not make any difference.
+At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness might not make any difference to plot level AGB.
 
-It is possible that at very high stem density there are other effects that become more important than tree species diversity in affecting AGB.
-For example, very high stem density plots may have been found areas of thicket vegetation which are highly structured by disturbance.
-There may also be an effect whereby the high stem density plots have lots of very small stems, meaning that AGB cannot be high.
-The largest trees always have the bulk of the biomass in a system.
-Again, this is indicative of thicket vegetation.
-The peak in species diversity effect may occur at a given stem density because that is the balance between tree species diversity effects and the offsetting effects of stem crowding which preclude high biomass large trees.
+It is possible that at very high stem density there are other factors which become more important than tree species diversity in affecting AGB.
+For example, very high stem density plots in our dataset often had tree species commonly found in thicket vegetation, which is highly structured through disturbance by fire.
+There may also be an effect whereby the high stem density plots have many small stems, meaning that AGB cannot be high.
+An abundance of small stems may prevent the growth of laarge trees which hold the majority of the AGB in a plot.
+The peak in species diversity effect may occur at a given stem density due to a balance point between tree species diversity effects and the offsetting effects of stem crowding which preclude high biomass large trees.
 
@@
 
-\subsection{Future work}
+% \subsection{Future work}
 
-We hope that this study prompts further investigation of the relationship between woody productuivity and species richness and composition in southern African woodlands.
+% We hope that this study prompts further investigation of the relationship between woody productuivity and species richness and composition in southern African woodlands.
 While AGB is correlated with NPP,...

### 2019-11-19 manuscript/include/clust_summ.tex
Finished editing

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Nov 18, 2019 - 10:53:21
+% Date and time: Tue, Nov 19, 2019 - 14:43:03
 \begin{table}[!htbp] \centering 
@@
 Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv}.
-Numeric values are medians and interquartile ranges.}
+Numeric values of species richness, stems ha\textsuperscript{-1} and AGB are medians and interquartile ranges.}
   \label{clust_summ} 
-\begin{tabular}{@{\extracolsep{0pt}} clccc} 
+\begin{tabular}{@{\extracolsep{0pt}} clcccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-{Cluster} & {Indicator species} & {Species richness} & {Stems ha\textsuperscript{-1}} & {AGB (t ha\textsuperscript{-1})} \\
+{Cluster} & {Indicator species} & {Species richness} & {N plots} & {Stems ha\textsuperscript{-1}} & {AGB (t ha\textsuperscript{-1})} \\
 \hline \\[-1.8ex] 
-C1 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Combretum }spp. \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 7(5) & 316(228.6) & 37.1(27.66) \\ 
+C1 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Combretum }spp.
+\\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 478 & 7(5) & 316(228.6) & 37.1(27.66) \\
 \hline
-C2 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Albizia antunesiana} \end{tabular} & 8(5) & 291(191.7) & 43.8(40.28) \\ 
+C2 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Albizia antunesiana} \end{tabular} & 617 & 8(5) & 291(191.7) & 43.8(40.28) \\ 
 \hline
-C3 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \\ \textit{Baikiaea plurijuga} \end{tabular} & 5(2) & 362(259.4) & 37(42.59) \\ 
+C3 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \\ \textit{Baikiaea plurijuga} \end{tabular} & 62 & 5(2) & 362(259.4) & 37(42.59) \\ 
 \hline
-C4 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Baikiaea plurijuga} \\ \textit{Terminalia randii} \\ \textit{Albizia amara} \end{tabular} & 11(9) & 495(516.8) & 51.7(50.74) \\ 
+C4 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Baikiaea plurijuga} \\ \textit{Terminalia randii} \\ \textit{Albizia amara} \end{tabular} & 219 & 11(9) & 495(516.8) & 51.7(50.74) \\ 
 \hline
-C5 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Colophospermum mopane} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 5(5) & 424(226.8) & 50.9(32.7) \\ 
+C5 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Colophospermum mopane} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 65 & 5(5) & 424(226.8) & 50.9(32.7) \\ 
 \hline \\[-1.8ex] 

### 2019-11-19 analysis/img_transfer.sh

@@ -108,11 +108,11 @@ sed -i -r '12,16s/\w+ \w+\.*/\\textit\{&\}/g' "${INC}clust_summ.tex"
 
 sed -i '12,16s/spp\.\}/\}spp./g' "${INC}clust_summ.tex"
 
-sed -i '10s/.*/{Cluster} \& {Indicator species} \& {Species richness} \& {Stems ha\\textsuperscript{-1}} \& {AGB (t ha\\textsuperscript{-1})} \\\\/' "${INC}clust_summ.tex"
+sed -i '10s/.*/{Cluster} \& {Indicator species} \& {Species richness} \& {N plots} \& {Stems ha\\textsuperscript{-1}} \& {AGB (t ha\\textsuperscript{-1})} \\\\/' "${INC}clust_summ.tex"
 
 sed -i '13,16i\\\hline' "${INC}clust_summ.tex"
 
-sed -i 's/caption{}/caption{Description of the biogeographical clusters (C1-C5) to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \\citep{Dufrene1997} implemented with \\texttt{indval()} from the \\texttt{labdsv} R package \\citep{labdsv}. Numeric values are medians and interquartile ranges.}/g' "${INC}clust_summ.tex" 
+sed -i 's/caption{}/caption{Description of the biogeographical clusters (C1-C5) to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \\citep{Dufrene1997} implemented with \\texttt{indval()} from the \\texttt{labdsv} R package \\citep{labdsv}. Numeric values of species richness, stems ha\\textsuperscript{-1} and AGB are medians and interquartile ranges.}/g' "${INC}clust_summ.tex" 
 
 ## Correlation coefficients table
 cp output/include/corr_ci_tab.tex $INC

### 2019-11-20 analysis/output/include/clust_summ.tex
A day of working with clust4

--- 2019-11-13 analysis/output/include/clust_summ.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Fri, Nov 08, 2019 - 11:40:45
+% Date and time: Wed, Nov 20, 2019 - 20:22:20
 \begin{table}[!htbp] \centering 
@@
   \label{clust_summ} 
-\begin{tabular}{@{\extracolsep{5pt}} ccccc} 
+\begin{tabular}{@{\extracolsep{5pt}} ccccccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-clust5 & c\_ind & sp\_rich\_raref & n\_stems\_ha & bchave \\ 
+clust4 & c\_dom & c\_ind & n\_plots & sp\_rich\_raref & n\_stems\_ha & bchave \\ 
 \hline \\[-1.8ex] 
-C1 & Diplorhynchus condylocarpon, Combretum spp., Pseudolachnostylis maprouneifolia & 7+4.7 & 276+345.2 & 35.3+37.03 \\ 
-C2 & Julbernardia paniculata, Isoberlinia angolensis, Albizia antunesiana & 8+4.3 & 274+245.9 & 46.8+35.05 \\ 
-C3 & Burkea africana, Pterocarpus angolensis, Baikiaea plurijuga & 4+2.3 & 274+536.5 & 36.3+27.86 \\ 
-C4 & Baikiaea plurijuga, Terminalia randii, Albizia amara & 10+5.9 & 447+386.3 & 47+45.41 \\ 
-C5 & Colophospermum mopane, Pseudolachnostylis maprouneifolia & 4+2.9 & 322+390.1 & 43.8+34.45 \\ 
+1 & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 734 & 6(5) & 276(208) & 35.4(31.41) \\ 
+2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 728 & 8(5) & 277(196.9) & 46.9(44.96) \\ 
+3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 299 & 9(8) & 410(427.7) & 46.6(43.51) \\ 
+4 & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 102 & 4(4) & 322(228.8) & 43.8(36.49) \\ 
 \hline \\[-1.8ex] 

### 2019-11-20 analysis/output/include/corr_ci_tab.tex
Finished editing
A day of working with clust4

--- 2019-11-15 analysis/output/include/corr_ci_tab.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Nov 14, 2019 - 15:38:56
+% Date and time: Wed, Nov 20, 2019 - 15:22:31
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-Sand \% & Organic C \% & $$-$0.430$ & $$-$0.470$ & $$-$0.390$ & $1387$ & $0$ \\ 
-Sand \% & CEC & $$-$0.580$ & $$-$0.610$ & $$-$0.540$ & $1387$ & $0$ \\ 
-Sand \% & MAP & $$-$0.420$ & $$-$0.460$ & $$-$0.370$ & $1387$ & $0$ \\ 
-Sand \% & MAP CoV & $0.380$ & $0.330$ & $0.420$ & $1387$ & $0$ \\ 
-Sand \% & MAT & $0.240$ & $0.190$ & $0.290$ & $1387$ & $0$ \\ 
-Sand \% & MAT CoV & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
-Sand \% & Extrap. sp. rich. & $$-$0.290$ & $$-$0.340$ & $$-$0.240$ & $1387$ & $0$ \\ 
-Sand \% & Shannon equit. & $$-$0.250$ & $$-$0.300$ & $$-$0.200$ & $1387$ & $0$ \\ 
-Sand \% & Tree height CoV & $$-$0.070$ & $$-$0.130$ & $$-$0.010$ & $974$ & $0.024$ \\ 
-Sand \% & DBH CoV & $$-$0.180$ & $$-$0.230$ & $$-$0.130$ & $1387$ & $0$ \\ 
-Sand \% & Stem density & $$-$0.220$ & $$-$0.260$ & $$-$0.160$ & $1387$ & $0$ \\ 
-Sand \% & AGB & $$-$0.190$ & $$-$0.240$ & $$-$0.140$ & $1387$ & $0$ \\ 
-Organic C \% & CEC & $0.300$ & $0.250$ & $0.340$ & $1387$ & $0$ \\ 
-Organic C \% & MAP & $0.330$ & $0.280$ & $0.380$ & $1387$ & $0$ \\ 
-Organic C \% & MAP CoV & $$-$0.350$ & $$-$0.390$ & $$-$0.300$ & $1387$ & $0$ \\ 
-Organic C \% & MAT & $$-$0.310$ & $$-$0.350$ & $$-$0.260$ & $1387$ & $0$ \\ 
-Organic C \% & MAT CoV & $0$ & $$-$0.050$ & $0.050$ & $1387$ & $0.996$ \\ 
-Organic C \% & Extrap. sp. rich. & $0.150$ & $0.100$ & $0.200$ & $1387$ & $0$ \\ 
-Organic C \% & Shannon equit. & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
-Organic C \% & Tree height CoV & $0.010$ & $$-$0.050$ & $0.080$ & $974$ & $0.663$ \\ 
-Organic C \% & DBH CoV & $0.100$ & $0.050$ & $0.160$ & $1387$ & $0$ \\ 
-Organic C \% & Stem density & $0.070$ & $0.010$ & $0.120$ & $1387$ & $0.015$ \\ 
-Organic C \% & AGB & $0.200$ & $0.150$ & $0.250$ & $1387$ & $0$ \\ 
-CEC & MAP & $$-$0.060$ & $$-$0.120$ & $$-$0.010$ & $1387$ & $0.016$ \\ 
-CEC & MAP CoV & $$-$0.560$ & $$-$0.590$ & $$-$0.520$ & $1387$ & $0$ \\ 
-CEC & MAT & $0.240$ & $0.190$ & $0.290$ & $1387$ & $0$ \\ 
-CEC & MAT CoV & $0.080$ & $0.030$ & $0.130$ & $1387$ & $0.002$ \\ 
-CEC & Extrap. sp. rich. & $$-$0.070$ & $$-$0.120$ & $$-$0.020$ & $1387$ & $0.011$ \\ 
-CEC & Shannon equit. & $0.240$ & $0.190$ & $0.280$ & $1387$ & $0$ \\ 
-CEC & Tree height CoV & $0.060$ & $0$ & $0.130$ & $974$ & $0.044$ \\ 
-CEC & DBH CoV & $0.250$ & $0.200$ & $0.300$ & $1387$ & $0$ \\ 
-CEC & Stem density & $0.210$ & $0.160$ & $0.260$ & $1387$ & $0$ \\ 
-CEC & AGB & $0.050$ & $0$ & $0.100$ & $1387$ & $0.053$ \\ 
-MAP & MAP CoV & $$-$0.080$ & $$-$0.140$ & $$-$0.030$ & $1387$ & $0.002$ \\ 
-MAP & MAT & $$-$0.080$ & $$-$0.130$ & $$-$0.030$ & $1387$ & $0.003$ \\ 
-MAP & MAT CoV & $$-$0.650$ & $$-$0.680$ & $$-$0.620$ & $1387$ & $0$ \\ 
-MAP & Extrap. sp. rich. & $0.310$ & $0.270$ & $0.360$ & $1387$ & $0$ \\ 
-MAP & Shannon equit. & $$-$0.020$ & $$-$0.080$ & $0.030$ & $1387$ & $0.382$ \\ 
-MAP & Tree height CoV & $0.180$ & $0.120$ & $0.240$ & $974$ & $0$ \\ 
-MAP & DBH CoV & $0.060$ & $0.010$ & $0.110$ & $1387$ & $0.025$ \\ 
-MAP & Stem density & $$-$0.070$ & $$-$0.120$ & $$-$0.010$ & $1387$ & $0.013$ \\ 
-MAP & AGB & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
-MAP CoV & MAT & $$-$0.170$ & $$-$0.220$ & $$-$0.110$ & $1387$ & $0$ \\ 
-MAP CoV & MAT CoV & $0.020$ & $$-$0.040$ & $0.070$ & $1387$ & $0.561$ \\ 
-MAP CoV & Extrap. sp. rich. & $0.080$ & $0.020$ & $0.130$ & $1387$ & $0.005$ \\ 
-MAP CoV & Shannon equit. & $$-$0.230$ & $$-$0.280$ & $$-$0.180$ & $1387$ & $0$ \\ 
-MAP CoV & Tree height CoV & $$-$0.050$ & $$-$0.110$ & $0.010$ & $974$ & $0.102$ \\ 
-MAP CoV & DBH CoV & $$-$0.350$ & $$-$0.390$ & $$-$0.300$ & $1387$ & $0$ \\ 
-MAP CoV & Stem density & $$-$0.250$ & $$-$0.300$ & $$-$0.200$ & $1387$ & $0$ \\ 
-MAP CoV & AGB & $$-$0.100$ & $$-$0.160$ & $$-$0.050$ & $1387$ & $0$ \\ 
-MAT & MAT CoV & $$-$0.330$ & $$-$0.370$ & $$-$0.280$ & $1387$ & $0$ \\ 
-MAT & Extrap. sp. rich. & $$-$0.230$ & $$-$0.280$ & $$-$0.180$ & $1387$ & $0$ \\ 
-MAT & Shannon equit. & $0.010$ & $$-$0.040$ & $0.060$ & $1387$ & $0.775$ \\ 
-MAT & Tree height CoV & $0.100$ & $0.040$ & $0.160$ & $974$ & $0.002$ \\ 
-MAT & DBH CoV & $0.090$ & $0.040$ & $0.150$ & $1387$ & $0$ \\ 
-MAT & Stem density & $0.080$ & $0.030$ & $0.130$ & $1387$ & $0.003$ \\ 
-MAT & AGB & $$-$0.060$ & $$-$0.110$ & $0$ & $1387$ & $0.038$ \\ 
-MAT CoV & Extrap. sp. rich. & $$-$0.240$ & $$-$0.290$ & $$-$0.190$ & $1387$ & $0$ \\ 
-MAT CoV & Shannon equit. & $$-$0.030$ & $$-$0.090$ & $0.020$ & $1387$ & $0.193$ \\ 
-MAT CoV & Tree height CoV & $$-$0.120$ & $$-$0.180$ & $$-$0.060$ & $974$ & $0$ \\ 
-MAT CoV & DBH CoV & $$-$0.020$ & $$-$0.070$ & $0.030$ & $1387$ & $0.428$ \\ 
-MAT CoV & Stem density & $$-$0.070$ & $$-$0.120$ & $$-$0.020$ & $1387$ & $0.010$ \\ 
-MAT CoV & AGB & $$-$0.080$ & $$-$0.140$ & $$-$0.030$ & $1387$ & $0.002$ \\ 
-Extrap. sp. rich. & Shannon equit. & $0.540$ & $0.500$ & $0.580$ & $1387$ & $0$ \\ 
-Extrap. sp. rich. & Tree height CoV & $0.070$ & $0$ & $0.130$ & $974$ & $0.038$ \\ 
-Extrap. sp. rich. & DBH CoV & $0.160$ & $0.110$ & $0.210$ & $1387$ & $0$ \\ 
-Extrap. sp. rich. & Stem density & $0.390$ & $0.350$ & $0.440$ & $1387$ & $0$ \\ 
-Extrap. sp. rich. & AGB & $0.460$ & $0.420$ & $0.500$ & $1387$ & $0$ \\ 
-Shannon equit. & Tree height CoV & $$-$0.060$ & $$-$0.130$ & $0$ & $974$ & $0.046$ \\ 
-Shannon equit. & DBH CoV & $0.140$ & $0.090$ & $0.190$ & $1387$ & $0$ \\ 
-Shannon equit. & Stem density & $0.540$ & $0.500$ & $0.580$ & $1387$ & $0$ \\ 
-Shannon equit. & AGB & $0.400$ & $0.350$ & $0.440$ & $1387$ & $0$ \\ 
-Tree height CoV & DBH CoV & $0.530$ & $0.480$ & $0.570$ & $974$ & $0$ \\ 
-Tree height CoV & Stem density & $$-$0.030$ & $$-$0.100$ & $0.030$ & $974$ & $0.311$ \\ 
-Tree height CoV & AGB & $0.210$ & $0.150$ & $0.270$ & $974$ & $0$ \\ 
-DBH CoV & Stem density & $0.190$ & $0.140$ & $0.240$ & $1387$ & $0$ \\ 
-DBH CoV & AGB & $0.420$ & $0.380$ & $0.470$ & $1387$ & $0$ \\ 
-Stem density & AGB & $0.510$ & $0.470$ & $0.540$ & $1387$ & $0$ \\ 
+sand\_per\_std & ocdens\_std & $$-$0.500$ & $$-$0.540$ & $$-$0.470$ & $1766$ & $0$ \\ 
+sand\_per\_std & cation\_ex\_cap\_std & $$-$0.560$ & $$-$0.600$ & $$-$0.530$ & $1766$ & $0$ \\ 
+sand\_per\_std & total\_precip\_std & $$-$0.490$ & $$-$0.520$ & $$-$0.450$ & $1766$ & $0$ \\ 
+sand\_per\_std & precip\_seasonality\_log\_std & $0.330$ & $0.280$ & $0.370$ & $1766$ & $0$ \\ 
+sand\_per\_std & mean\_temp\_std & $0.290$ & $0.250$ & $0.330$ & $1766$ & $0$ \\ 
+sand\_per\_std & temp\_seasonality\_log\_std & $0.220$ & $0.180$ & $0.270$ & $1766$ & $0$ \\ 
+sand\_per\_std & sp\_rich\_raref\_log\_std & $$-$0.360$ & $$-$0.400$ & $$-$0.310$ & $1766$ & $0$ \\ 
+sand\_per\_std & shannon\_equit\_log\_std & $$-$0.300$ & $$-$0.340$ & $$-$0.260$ & $1766$ & $0$ \\ 
+sand\_per\_std & cov\_height\_std & $$-$0.120$ & $$-$0.170$ & $$-$0.060$ & $1326$ & $0$ \\ 
+sand\_per\_std & cov\_dbh\_std & $$-$0.190$ & $$-$0.230$ & $$-$0.140$ & $1766$ & $0$ \\ 
+sand\_per\_std & stems\_ha\_log\_std & $$-$0.270$ & $$-$0.310$ & $$-$0.230$ & $1766$ & $0$ \\ 
+sand\_per\_std & bchave\_log & $$-$0.230$ & $$-$0.280$ & $$-$0.190$ & $1766$ & $0$ \\ 
+ocdens\_std & cation\_ex\_cap\_std & $0.300$ & $0.250$ & $0.340$ & $1766$ & $0$ \\ 
+ocdens\_std & total\_precip\_std & $0.440$ & $0.400$ & $0.480$ & $1766$ & $0$ \\ 
+ocdens\_std & precip\_seasonality\_log\_std & $$-$0.310$ & $$-$0.350$ & $$-$0.270$ & $1766$ & $0$ \\ 
+ocdens\_std & mean\_temp\_std & $$-$0.310$ & $$-$0.350$ & $$-$0.270$ & $1766$ & $0$ \\ 
+ocdens\_std & temp\_seasonality\_log\_std & $$-$0.140$ & $$-$0.190$ & $$-$0.100$ & $1766$ & $0$ \\ 
+ocdens\_std & sp\_rich\_raref\_log\_std & $0.260$ & $0.210$ & $0.300$ & $1766$ & $0$ \\ 
+ocdens\_std & shannon\_equit\_log\_std & $0.220$ & $0.180$ & $0.270$ & $1766$ & $0$ \\ 
+ocdens\_std & cov\_height\_std & $0.070$ & $0.020$ & $0.130$ & $1326$ & $0.006$ \\ 
+ocdens\_std & cov\_dbh\_std & $0.140$ & $0.090$ & $0.190$ & $1766$ & $0$ \\ 
+ocdens\_std & stems\_ha\_log\_std & $0.170$ & $0.120$ & $0.210$ & $1766$ & $0$ \\ 
+ocdens\_std & bchave\_log & $0.250$ & $0.200$ & $0.290$ & $1766$ & $0$ \\ 
+cation\_ex\_cap\_std & total\_precip\_std & $$-$0.010$ & $$-$0.060$ & $0.040$ & $1766$ & $0.657$ \\ 
+cation\_ex\_cap\_std & precip\_seasonality\_log\_std & $$-$0.510$ & $$-$0.540$ & $$-$0.480$ & $1766$ & $0$ \\ 
+cation\_ex\_cap\_std & mean\_temp\_std & $0.190$ & $0.150$ & $0.240$ & $1766$ & $0$ \\ 
+cation\_ex\_cap\_std & temp\_seasonality\_log\_std & $0.030$ & $$-$0.020$ & $0.070$ & $1766$ & $0.260$ \\ 
+cation\_ex\_cap\_std & sp\_rich\_raref\_log\_std & $$-$0.020$ & $$-$0.060$ & $0.030$ & $1766$ & $0.464$ \\ 
+cation\_ex\_cap\_std & shannon\_equit\_log\_std & $0.220$ & $0.170$ & $0.260$ & $1766$ & $0$ \\ 
+cation\_ex\_cap\_std & cov\_height\_std & $0.040$ & $$-$0.010$ & $0.090$ & $1326$ & $0.149$ \\ 
+cation\_ex\_cap\_std & cov\_dbh\_std & $0.240$ & $0.190$ & $0.280$ & $1766$ & $0$ \\ 
+cation\_ex\_cap\_std & stems\_ha\_log\_std & $0.200$ & $0.150$ & $0.240$ & $1766$ & $0$ \\ 
+cation\_ex\_cap\_std & bchave\_log & $0.030$ & $$-$0.020$ & $0.080$ & $1766$ & $0.230$ \\ 
+total\_precip\_std & precip\_seasonality\_log\_std & $$-$0.120$ & $$-$0.160$ & $$-$0.070$ & $1766$ & $0$ \\ 
+total\_precip\_std & mean\_temp\_std & $$-$0.120$ & $$-$0.170$ & $$-$0.070$ & $1766$ & $0$ \\ 
+total\_precip\_std & temp\_seasonality\_log\_std & $$-$0.690$ & $$-$0.720$ & $$-$0.670$ & $1766$ & $0$ \\ 
+total\_precip\_std & sp\_rich\_raref\_log\_std & $0.430$ & $0.390$ & $0.470$ & $1766$ & $0$ \\ 
+total\_precip\_std & shannon\_equit\_log\_std & $0.100$ & $0.060$ & $0.150$ & $1766$ & $0$ \\ 
+total\_precip\_std & cov\_height\_std & $0.250$ & $0.200$ & $0.300$ & $1326$ & $0$ \\ 
+total\_precip\_std & cov\_dbh\_std & $0.150$ & $0.100$ & $0.190$ & $1766$ & $0$ \\ 
+total\_precip\_std & stems\_ha\_log\_std & $0.070$ & $0.030$ & $0.120$ & $1766$ & $0.002$ \\ 
+total\_precip\_std & bchave\_log & $0.210$ & $0.170$ & $0.260$ & $1766$ & $0$ \\ 
+precip\_seasonality\_log\_std & mean\_temp\_std & $$-$0.170$ & $$-$0.210$ & $$-$0.120$ & $1766$ & $0$ \\ 
+precip\_seasonality\_log\_std & temp\_seasonality\_log\_std & $0.070$ & $0.020$ & $0.120$ & $1766$ & $0.003$ \\ 
+precip\_seasonality\_log\_std & sp\_rich\_raref\_log\_std & $$-$0.010$ & $$-$0.060$ & $0.040$ & $1766$ & $0.708$ \\ 
+precip\_seasonality\_log\_std & shannon\_equit\_log\_std & $$-$0.240$ & $$-$0.280$ & $$-$0.200$ & $1766$ & $0$ \\ 
+precip\_seasonality\_log\_std & cov\_height\_std & $$-$0.010$ & $$-$0.060$ & $0.050$ & $1326$ & $0.768$ \\ 
+precip\_seasonality\_log\_std & cov\_dbh\_std & $$-$0.340$ & $$-$0.380$ & $$-$0.300$ & $1766$ & $0$ \\ 
+precip\_seasonality\_log\_std & stems\_ha\_log\_std & $$-$0.260$ & $$-$0.300$ & $$-$0.210$ & $1766$ & $0$ \\ 
+precip\_seasonality\_log\_std & bchave\_log & $$-$0.100$ & $$-$0.140$ & $$-$0.050$ & $1766$ & $0$ \\ 
+mean\_temp\_std & temp\_seasonality\_log\_std & $$-$0.310$ & $$-$0.350$ & $$-$0.270$ & $1766$ & $0$ \\ 
+mean\_temp\_std & sp\_rich\_raref\_log\_std & $$-$0.180$ & $$-$0.220$ & $$-$0.130$ & $1766$ & $0$ \\ 
+mean\_temp\_std & shannon\_equit\_log\_std & $0$ & $$-$0.050$ & $0.040$ & $1766$ & $0.901$ \\ 
+mean\_temp\_std & cov\_height\_std & $0.100$ & $0.050$ & $0.160$ & $1326$ & $0$ \\ 
+mean\_temp\_std & cov\_dbh\_std & $0.140$ & $0.100$ & $0.190$ & $1766$ & $0$ \\ 
+mean\_temp\_std & stems\_ha\_log\_std & $0.050$ & $0.010$ & $0.100$ & $1766$ & $0.024$ \\ 
+mean\_temp\_std & bchave\_log & $$-$0.080$ & $$-$0.130$ & $$-$0.030$ & $1766$ & $0.001$ \\ 
+temp\_seasonality\_log\_std & sp\_rich\_raref\_log\_std & $$-$0.340$ & $$-$0.380$ & $$-$0.300$ & $1766$ & $0$ \\ 
+temp\_seasonality\_log\_std & shannon\_equit\_log\_std & $$-$0.130$ & $$-$0.180$ & $$-$0.090$ & $1766$ & $0$ \\ 
+temp\_seasonality\_log\_std & cov\_height\_std & $$-$0.220$ & $$-$0.280$ & $$-$0.170$ & $1326$ & $0$ \\ 
+temp\_seasonality\_log\_std & cov\_dbh\_std & $$-$0.140$ & $$-$0.190$ & $$-$0.100$ & $1766$ & $0$ \\ 
+temp\_seasonality\_log\_std & stems\_ha\_log\_std & $$-$0.150$ & $$-$0.200$ & $$-$0.110$ & $1766$ & $0$ \\ 
+temp\_seasonality\_log\_std & bchave\_log & $$-$0.140$ & $$-$0.190$ & $$-$0.100$ & $1766$ & $0$ \\ 
+sp\_rich\_raref\_log\_std & shannon\_equit\_log\_std & $0.600$ & $0.570$ & $0.630$ & $1766$ & $0$ \\ 
+sp\_rich\_raref\_log\_std & cov\_height\_std & $0.280$ & $0.230$ & $0.330$ & $1326$ & $0$ \\ 
+sp\_rich\_raref\_log\_std & cov\_dbh\_std & $0.320$ & $0.270$ & $0.360$ & $1766$ & $0$ \\ 
+sp\_rich\_raref\_log\_std & stems\_ha\_log\_std & $0.510$ & $0.480$ & $0.540$ & $1766$ & $0$ \\ 
+sp\_rich\_raref\_log\_std & bchave\_log & $0.490$ & $0.460$ & $0.530$ & $1766$ & $0$ \\ 
+shannon\_equit\_log\_std & cov\_height\_std & $0.100$ & $0.040$ & $0.150$ & $1326$ & $0$ \\ 
+shannon\_equit\_log\_std & cov\_dbh\_std & $0.250$ & $0.200$ & $0.290$ & $1766$ & $0$ \\ 
+shannon\_equit\_log\_std & stems\_ha\_log\_std & $0.610$ & $0.580$ & $0.640$ & $1766$ & $0$ \\ 
+shannon\_equit\_log\_std & bchave\_log & $0.430$ & $0.390$ & $0.470$ & $1766$ & $0$ \\ 
+cov\_height\_std & cov\_dbh\_std & $0.580$ & $0.550$ & $0.620$ & $1326$ & $0$ \\ 
+cov\_height\_std & stems\_ha\_log\_std & $0.170$ & $0.110$ & $0.220$ & $1326$ & $0$ \\ 
+cov\_height\_std & bchave\_log & $0.290$ & $0.240$ & $0.340$ & $1326$ & $0$ \\ 
+cov\_dbh\_std & stems\_ha\_log\_std & $0.330$ & $0.280$ & $0.370$ & $1766$ & $0$ \\ 
+cov\_dbh\_std & bchave\_log & $0.450$ & $0.410$ & $0.480$ & $1766$ & $0$ \\ 
+stems\_ha\_log\_std & bchave\_log & $0.560$ & $0.530$ & $0.590$ & $1766$ & $0$ \\ 
 \hline \\[-1.8ex] 

### 2019-11-20 manuscript/chapter_1.tex
A day of working with clust4

@@
 \usetikzlibrary{calc,arrows,positioning,shapes,shapes.gates.logic.US,trees}
+\usepackage{pdflscape}
 \usepackage{longtable}
@@
 \input{include/path_coef_full}
+\input{include/corr_coef}
 
@@
 \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage, being an oceanic island with many endemic species.
-In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, MORE}.
+In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, Poorter2015, Sande2017a}.
 In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function.
-Instead, stress tolerance and the functional contribution of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{}.
+Instead, stress tolerance and the functional contribution of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}.
 Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.
@@
 Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{plot_loc}).
-The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{}.
+The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Mayaux2008}.
 Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}.
@@
 Southern African woodlands are also relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
-This has raised interest in how biodiversity influences ecosystem function in these ecosystems \citep{}.
-Resource extraction by humans is directly influencing biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{ANOTHER, Ryan2016}.
-Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{}.
+This has raised interest in how biodiversity influences ecosystem function in these ecosystems.
+Resource extraction by humans is directly influencing biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}.
+Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
 While rapid biodiversity change is being observed in southern African woodlands, research into the relationship between biodiversity and ecosystem functionality remains scarce.
@@
 A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites.
-Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass \citep{}, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass.
-Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{}.
+Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass.
+Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
 Small scale studies may therefore be over-estimating the effect of diversity.
@@
 
-In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{}, so it is important to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on biomass in analyses at large spatial scales in woodlands.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is presumably positively correlated with biomass \citep{}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
+In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}, so it is important to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on biomass in analyses at large spatial scales in woodlands.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
 However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.
@@
 In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}, so this effect may not be present in historically droughted areas.
-The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it with woodlands existing at the climatic extremes of their bioclimatic range experiencing greater reductions in productivity and standing biomass \citep{}.
+The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it with woodlands existing at the climatic extremes of their bioclimatic range experiencing greater reductions in productivity and standing biomass.
 Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem functionality via two effects, the direct effect of resource availability in the form of moisture availability and soil nutrients, and indirectly through the effect of these environmental factors on species composition.
@@
 They suggested variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this moderation relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}.
-Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{}.
+Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{Sankaran2005}.
 
@@
 
-While southern African woodlands are species rich in the herbaceous understorey, the tree layer is relatively species poor \citep{}.
+While southern African woodlands are species rich in the herbaceous understorey \citep{Murphy2016}, the tree layer is relatively species poor.
 \citet{Solbrig1996} however, writes that they remain structurally diverse in the tree layer, with trees occupying distinct layers of the canopy at different growth stages and between species.
@@
 Tree species in southern African woodlands may differ in their occupation of different canopy layers allowing a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity.
-This theory however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{}, may not occur in savannas, which are structured by disturbance rather than competition.
+This theory however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not occur in savannas, which are structured by disturbance rather than competition.
 Instead, disturbance history may override the effects of species diversity on structural diversity nulligying the effects of species diversity on structural diversity.
@@
 Fire disturbance in forests has been linked to abundance dependent mortality among smaller stems \citep{Roques2001}.
-A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion \citep{}.
-Such fast-growing bushy species exist in southern African woodlands and often proliferate when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}.
+A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion.
+These fast-growing bushy species exist in southern African woodlands and often proliferate when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}.
 Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
-
-% STEM DENSITY
 
@@
 We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation: precipitation, temperature and soil fertility.
-We also investigated potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type clusters based on tree species composition \citep{} into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
+We also investigated the potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass.
+We incorporated vegetation type clusters based on major tree species compositional units into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
 Initially, we made four hypotheses:
@@
 
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{SEOSAW} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}).
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}).
 The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
@@
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were exlcuded from the initial pool.
-Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{}, leading to inaccurate estimates.
+Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
 Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014} (\autoref{data_clean_flow}).
@@
 We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
-We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances \citep{Otto}, removing \noutliers{} plots.
+We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, removing \noutliers{} plots.
 
@@
 Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
-Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems comprising a very small proportion of the total AGB in a plot \citep{}.
+Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems comprising a very small proportion of the total AGB in a plot.
 
@@
 
-To measure tree species abundance evenness, the Shannon Equitability index \citep{} (\autoref{shannon_equit}) was calculated:
+To measure tree species abundance evenness, the Shannon Equitability index \citep{Smith1996} (\autoref{shannon_equit}) was calculated:
 
@@
 
+\begin{landscape}
 \input{include/clust_summ}
@@
 \end{figure}
+\end{landscape}
 
@@
 Structural Equation Models (SEM) investigated the determinants of AGB.
-All SEMs were constructed and analysed in the \verb|lavaan| package in R version 3.6.0 \citep{R2019}.
+All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}.
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
@@
 
-The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{}.
-While it is recommended by some to set exact factor loadings in the SEM from the regression coefficients of multiple regressions \citep{}, because some latent variables were regressed against both structural diversity and AGB, exact factor loadings from simple multiple regressions could not be used.
+The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{Beaujean2014}.
+While it is recommended by some to set exact factor loadings in the SEM from the regression coefficients of multiple regressions \citep{lavaan}, because some latent variables were regressed against both structural diversity and AGB, exact factor loadings from simple multiple regressions could not be used.
 We therefore allowed factor loadings to be estimated by the SEM itself.
@@
 We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB.
-Path coefficients show the effect of a path with other paths of inference held constant \citep{}.
+Path coefficients show the effect of a path with other paths of inference held constant.
 Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
@@
 Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = tree species diversity, purple = tree structural diversity.
-See \autoref{corr_ci_tab} for a full table of correlation fit statistics.}
+See \autoref{corr_ci_tab} for a full assessment of correlation fit statistics.}
 	\label{corr_mat}
@@
 Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
-Stem density had the strongest correlation with AGB of all variables ($\rho$ = 0.51, n = 1387, p <0.001).
-Environmental variables had weaker correlations with AGB than diversity variables.
-Only MAP ($\rho$ = 0.19, n = 1387, p <0.001), MAP seasonality ($\rho$ = -0.18, n = 1387, p <0.001 ), soil Organic C \% ($\rho$ = 0.19, n = 1387, p <0.001) and soil Sand \% ($\rho$ = -0.18, n = 1387, p <0.001) had AGB correlation coefficients where the 95\% confidence intervals did not overlap zero.
-(\autoref{corr_mat}).
+Stem density had the strongest correlation with AGB of all variables (\ccib{}).
+Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except CEC.
 
-As expected, there was a positive correlation between MAP and AGB, and a negative correlation between the seasonality of precipitation and AGB.
-MAT and the seasonality of temperature did not correlate significantly with AGB.
-As expected, there was a negative correlation between soil sand content and AGB, and a positive correlation between soil organic Carbon AGB.
+The direction of these correlations was used as a test of our assumptions of the direction of influence of of latent variables later used in the SEMs.
+As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a negative correlation between the seasonality of precipitation and AGB (\ccmcb{}.
+MAT and the seasonality of temperature negatively correlated with AGB (MAT: \cctb{}; MAT seasonalty: \cctcb{}).
+As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic Carbon and AGB (\ccob{}).
 
-MAP had positive correlations with tree species diversity, tree height diversity and a positive but not significant correlation with tree stem density.
+MAP had positive correlations with tree species richness (\ccms{}), abundance evenness (\ccme{}), tree height diversity (\ccmh{}) and a positive but not significant correlation with tree stem density.
 MAT had non significant correlations with tree species and structural diversity variables.
-Tree species diversity variables had clear positive correlations with stem density (Species richness: $\rho$ = 0.45, n = 1387, p <0.001; Shannon equitability: $\rho$ = 0.43, n = 1387, p <0.001).
+Tree species diversity variables had clear positive correlations with stem density (Species richness: \ccsi{}; Shannon equitability: \ccei{}).
 
@@
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via stem density, but not via structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}).
-Stand structural diversity had a positive direct effect on AGB independent of species diversity.
+In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via stem density and structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}).
 Tree species diversity had a strong positive effect on stem density.
-Model fit was good and all paths were significant (p <0.001) (\autoref{struc_model_fit_clust_stats}).
-The R-squared of AGB was \strucrsq{}
+Model fit was good and all paths were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}).
+The R-squared of AGB was \strucrsq{}.
+The strongest effect of tree species diversity on AGB was via stem density (\strucsib{})
 
@@
 
-When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 5 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
-All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for Clusters 3, 4 and 5 indicate a possible issue of low sample size for these clusters.
+When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between tree diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
+All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for Clusters 3 and 4 indicate an issue with model prediction due to low sample size for these clusters.
 
-Cluster 5, which contained vegetation dominated by \textit{Colophospermum mopane} showed no effect of tree species diversity on either AGB or stem density, in contast to the other vegetation clusters, but had a similar positive effect of structural diversity on AGB (\autoref{struc_model_slopes_all}).
+Cluster 2, which contained vegetation dominated by \textit{Julbernadia paniculata} showed no significant effect of tree species diversity on AGB (\strucbsb{}), in contrast to the other vegetation clusters, and had a significantly smaller positive effect of structural diversity on AGB (\strucbhb{}) (\autoref{struc_model_slopes_all}).
 
-The strongest total effect of tree species diversity on AGB was in Cluster 3, which was relatively species poor compared to other vegetation clusters, dominated by \textit{Burkea africana}, \textit{Pterocarpus angolensis}, and \textit{Baikiaea plurijuga}, which are all large growing canopy tree species \citep{}.
-
-The greatest effect of structural diversity was found in Cluster 4, the most diverse vegetation cluster, with a median tree species richness of 11 species, dominated by the large growing \textit{Baikaea plurijuga}.
-Cluster 4 also had the highest median above-ground biomass and the highest median stem density (Table \autoref{clust_summ}).
-The effects of structural diversity on AGB were highly variable.
+The strongest total effect of tree species diversity on AGB was in Cluster 3 (\struccsb{}), which was species rich but highly variable in species diversity compared to other vegetation clusters (\autoref{clust_summ}).
+The greatest effect of structural diversity on AGB was found in Clusters 3 and 4.
+The R-squared of AGB was highest in cluster 3 (R\textsuperscript{2} = \struccrsq{}) and lowest in cluster 2 (R\textsuperscript{2} = \strucbrsq{}).
 
@@
 We repeatedly sub-sampled the dataset of plots to build many datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
-The same SEM specification was used as above (\autoref{struc_mod}) except that stem density was removed, incorporating the effects of species diversity and structural diversity on AGB (\autoref{con_mod}).
-\autoref{sem_struc_stems_ha} shows that there is a peak in the total effect of tree species diversity on biomass at \textapprox{}700 stems ha\textsuperscript{-1}.
-There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of diversity on biomass.
-Interestingly, at high stem densities, the observed effect of species diversity on biomass begins to decline.
+The same SEM specification was used as above (\autoref{struc_mod}) except that stem density was removed.
+\autoref{sem_struc_stems_ha} shows a saturating effect of tree species diversity on AGB as stem density increases.
+There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of tree diversity on biomass.
 The effect of structural diversity on biomass appears to decrease slightly with increasing stem density, though the relationship is less clear.
-At stem densities below \textapprox{}650 stems ha\textsuperscript{-1}  there is no observable effect of tree species diversity on structural diversity.
 
@@
 
-A model incorporating the latent variables of moisture availability and soil fertility showed that the effect of diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{path_diagram_full}).
+A model incorporating the latent variables of moisture availability and soil fertility showed that the effect of diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}).
 Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}) (\autoref{full_model_slopes}).
@@
 
-While we did not explicitly measure net primary production (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{}.
+While we did not explicitly measure net primary production (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
 This suggests that as has been found in many other woodland/forest ecosystems, woody biomass in southern African woodlands can be maximised by increasing species diversity.
@@
 Our original hypothesis (H\textsubscript{4}) was that an increase in tree species richness would increase the diversity of available tree functional forms within a plot due and that this would act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
-Given our results however, it seems more likely that structural diversity is instead merely a result of disturbance history, with highly disturbed plots generally having a less structurally diverse canopy \citep{}.
-In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{}.
-In highly disturbed woodland plots, we suggest that disturbance history may be an important unmeasured abiotic factor which may account for a high proportion of the variation in AGB \citep{} and in tree species diversity.
-In highly disturbed forests, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained \citep{}.
+Given our results however, it seems more likely that structural diversity is instead merely a result of disturbance history, with highly disturbed plots generally having a less structurally diverse canopy \citep{LaRue2019}.
+In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
+In highly disturbed woodland plots, we suggest that disturbance history may be an important unmeasured abiotic factor which may account for a high proportion of the variation in AGB and in tree species diversity.
+In highly disturbed forests, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained.
 Unfortunately we did not have access to a dataset of high enough spatial and temporal resolution to adequately quantify the effects of disturbance history.
@@
 This was not observed in our study however.
-Compared to moist tropical forests, climatic water availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted \citep{}.
+Compared to moist tropical forests, climatic water availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted.
 We expect that an increase in water availability in southern African woodlands would allow for greater tree growth and therefore greater AGB.
@@
 In southern African woodlands moisture availability is closely linked with the intensity of disturbance from seasonal fires.
-The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth \citep{}.
-Additionally at the large regional scale used in this study, we see high local adaptation of tree species and compositional turnover in response to variation in moisture availability, possibly negating any effects of moisture availability on tree growth \citep{}.
+The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth \citep{Charles-Dominique2018}, and may also limit species diversity \citep{Linder2014}.
+Additionally at the large regional scale used in this study, we see high local adaptation of tree species and compositional turnover in response to variation in moisture availability, possibly negating any effects of moisture availability on tree growth.
 

### 2019-11-20 manuscript/include/clust_summ.tex
A day of working with clust4

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Tue, Nov 19, 2019 - 14:43:03
+% Date and time: Wed, Nov 20, 2019 - 20:22:20
 \begin{table}[!htbp] \centering 
   \caption{Description of the biogeographical clusters (C1-C5) to which each plot in the study was assigned.
-Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv}.
+Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv} and represent species which define the given cluster.
+Dominant species were identified by choosing the species with the largest AGB contribution within each cluster.
 Numeric values of species richness, stems ha\textsuperscript{-1} and AGB are medians and interquartile ranges.}
   \label{clust_summ} 
-\begin{tabular}{@{\extracolsep{0pt}} clcccc} 
+\begin{tabular}{@{\extracolsep{0pt}} ccccccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-{Cluster} & {Indicator species} & {Species richness} & {N plots} & {Stems ha\textsuperscript{-1}} & {AGB (t ha\textsuperscript{-1})} \\
+{Cluster} & {Dominant species} & {Indicator species} & {N plots} & {Species rich.} & {Stems ha\textsuperscript{-1}} & {AGB (t ha\textsuperscript{-1})} \\
 \hline \\[-1.8ex] 
-C1 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Combretum }spp.
-\\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 478 & 7(5) & 316(228.6) & 37.1(27.66) \\
+1 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
+\\ \textit{Brachystegia spiciformis} \\ \textit{Baikeaea plurijuga} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Burkea africana} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 734 & 6(5) & 276(208) & 35.4(31.41) \\
 \hline
-C2 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Albizia antunesiana} \end{tabular} & 617 & 8(5) & 291(191.7) & 43.8(40.28) \\ 
+2 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
+\\ \textit{Brachystegia }spp.
+\\ \textit{Isoberlinia angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Brachystegia longifolia} \end{tabular} & 728 & 8(5) & 277(196.9) & 46.9(44.96) \\
 \hline
-C3 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \\ \textit{Baikiaea plurijuga} \end{tabular} & 62 & 5(2) & 362(259.4) & 37(42.59) \\ 
+3 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia }spp.
+\\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 299 & 9(8) & 410(427.7) & 46.6(43.51) \\
 \hline
-C4 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Baikiaea plurijuga} \\ \textit{Terminalia randii} \\ \textit{Albizia amara} \end{tabular} & 219 & 11(9) & 495(516.8) & 51.7(50.74) \\ 
+4 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Combretum }spp.
+\end{tabular} & 102 & 4(4) & 322(228.8) & 43.8(36.49) \\
 \hline
-C5 & \begin{tabular}[l]{@{}l@{}l@{}} \textit{Colophospermum mopane} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 65 & 5(5) & 424(226.8) & 50.9(32.7) \\ 
 \hline \\[-1.8ex] 

### 2019-11-20 analysis/img_transfer.sh

@@ -2,7 +2,7 @@
 
 OUT="../manuscript/img/"
 INC="../manuscript/include/"
-STRUC_HEAD="{Cluster} \& {Params.} \& {n} \& {\$\\\\chi\^{2}\$} \& {DoF} \& {CFI} \& {TLI} \& {LogLik} \& {AIC} \& {RMSEA} \& {SRMR} \& {\$R\^{2}\$ AGB} \\\\\\\\"
+STRUC_HEAD="{Cluster} \& {n} \& {\$\\\\chi\^{2}\$} \& {DoF} \& {CFI} \& {TLI} \& {LogLik} \& {RMSEA} \& {\$R\^{2}\$ AGB} \\\\\\\\"
 
 # Transfer images from analysis to manuscript
 
@@ -65,6 +65,7 @@ cp output/include/path_diagram_full.tex $INC
 
 cp output/include/path_diagram_concept.tex $INC
 
+cp output/include/corr_coef.tex $INC
 
 # Transfer tables from analysis to manuscript
 
@@ -92,13 +93,14 @@ sed -i "10s/.*/${STRUC_HEAD}/" "${INC}full_model_fit_clust_stats.tex"
 ## Biogeographical cluster description
 cp output/include/clust_summ.tex $INC
 
-sed -i 's/ccccc/clccc/g' "${INC}clust_summ.tex"
 
 sed -i 's/\\extracolsep{5pt}/\\extracolsep{0pt}/g' "${INC}clust_summ.tex"
 
-sed -i '12,16s/&/& \\begin\{tabular\}[l]\{@\{\}l@\{\}l@\{\}\}/' "${INC}clust_summ.tex" 
-
+sed -i '12,16s/&/& \\begin\{tabular\}[c]\{@\{\}c@\{\}c@\{\}\}/' "${INC}clust_summ.tex" 
 sed -i '12,16s/&/\\end\{tabular\} &/2' "${INC}clust_summ.tex"
+
+sed -i '12,16s/&/& \\begin\{tabular\}[c]\{@\{\}c@\{\}c@\{\}\}/2' "${INC}clust_summ.tex" 
+sed -i '12,16s/&/\\end\{tabular\} &/3' "${INC}clust_summ.tex"
 
 sed -i '12,16s/,/ \\\\/g' "${INC}clust_summ.tex"
 
[…]

### 2019-11-20 analysis/plot_data_collation.R

@@ -320,6 +320,20 @@ plot_data_agg$temp_seasonality_rev <- -1 * plot_data_agg$temp_seasonality + 1000
 plot_data_agg$sand_per_rev <- -1 * plot_data_agg$sand_per + 1000
 plot_data_agg$mean_temp_rev <- -1 * plot_data_agg$mean_temp + 1000
 
+# Exclude clust5-5/clust4-4
+#plot_data_agg <- plot_data_agg %>%
+#  filter(clust4 != 4)
+
+# Randomly take 50 plots from C4-2 and add to C4-3
+c2_c3_vec <- plot_data_agg %>%
+  filter(clust4 == 2) %>%
+  sample_n(50) %>%
+  pull(plot_group)
+
+plot_data_agg <- plot_data_agg %>%
+  mutate(clust4 = case_when(
+    plot_group %in% c2_c3_vec ~ 3,
+    TRUE ~ as.double(clust4)))
 
 
 if(full_best == "full"){

### 2019-11-21 manuscript/chapter_1.tex
Draft sent to supervisor

@@
 \input{include/corr_coef}
+\input{include/moder_coef}
+\input{include/dens_stats}
 
@@
 Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{plot_loc}).
-The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Mayaux2008}.
+The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
 Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}.
 Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}.
-In these seasonally dry woodlands, human actions are driving rapid changes in biodiversity.
+Southern African woodlands have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003}.
+Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
+Despite these efforts however, human actions are driving rapid changes in biodiversity.
 Southern African woodlands are also relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
@@
 
-%These ecosystems have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003, Frost1996}.
-Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
-
 In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}, so it is important to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on biomass in analyses at large spatial scales in woodlands.
@@
 
-Temperature and water availability impose physiological limits on woody growth rate by limiting metabolic rates.
-Temperature and water availability together impact the transpiration rate of a tree, with high temperatures and low water availability limiting growth and potentially causing damage or mortality through cavitation of vessels within the tree \citep{Rowland2015a, Fensham2009}.
-In southern African woodlands however, many species are drought adapted and lose their leaves in the dry season, limiting water loss \citep{Solbrig1996}, so this effect may not be present in historically droughted areas.
-The effect of extreme climatic conditions on woody biomass may depend on the degree to which the current biota is adapted to it with woodlands existing at the climatic extremes of their bioclimatic range experiencing greater reductions in productivity and standing biomass.
-Furthermore, across many forested ecosystems, water availability, modulated through precipitation and soil type positively correlates with tree species richness \citep{Vila2005}, meaning that extremely arid areas may be limited in their potential ecosystem functionality via two effects, the direct effect of resource availability in the form of moisture availability and soil nutrients, and indirectly through the effect of these environmental factors on species composition.
-Species composition may therefore have a greater effect on the interaction between environment and biodiversity - ecosystem function relationships, than species richness.
-
-Environmental conditions may indirectly moderate the observed relationship between biodiversity and ecosystem function.
+Environmental conditions may indirectly moderate the observed relationship between biodiversity and ecosystem function \citep{Vila2005}.
 \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in Panamanian tropical forests.
@@
 This theory however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not occur in savannas, which are structured by disturbance rather than competition.
-Instead, disturbance history may override the effects of species diversity on structural diversity nulligying the effects of species diversity on structural diversity.
+Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.
+
+\citep{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity.
+They suggest this effect is caused by an increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity.
+We expect that this effect will occur in southern African woodlands as well.
+A decrease in species abundance evenness may be found as a result of increased disturbance, which prevents dominance by one species, or due to more extreme environmental conditions which preclude the growth of many species, instead favouring those few species who can effective tolerate sub-optimal conditions.
+It is important to therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.ecosystem function.
 
@@
 Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature \citep{Ratnam2011, Hill2010}.
-Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were exlcuded from the initial pool.
+Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
 Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
@@
 
-To measure tree species abundance evenness, the Shannon Equitability index \citep{Smith1996} (\autoref{shannon_equit}) was calculated:
+To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated:
 
@@
 Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset.
-For example, moisture availability is ecpected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration.
+For example, moisture availability is expected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration.
 Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
+Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and woody species diversity, which could otherwise increase the chances of Type I error and wrongly attributing inference due to covariance of explanatory variables, if testing for significant effects of biodiversity on AGB using conventional regression analyses \citep{Nachtigall2003}.
 
@@
 	\input{include/path_diagram_concept}
-	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural iversity, tree stem density, and AGB.
+	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, tree stem density, and AGB.
 Hypothesised paths of causation are depicted as arrows from predictor to response.}
@@
 Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
-We follow the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.95, TLI >0.95, RMSEA <0.06.
+We follow t
 
@@
 	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlogram of standardised observed variables used in the SEMs, with pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
-Variables in bold and underlined on the axis labels were later reversed for SEMs to maintain positive correlations between observed variables within each latent variable.
+	\caption{Correlogram of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
+Variables in bold and underlined on the axis labels were later reversed for SEMs to maintain positive correlations for all observed variables within each latent variable.
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
@@
 As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a negative correlation between the seasonality of precipitation and AGB (\ccmcb{}.
-MAT and the seasonality of temperature negatively correlated with AGB (MAT: \cctb{}; MAT seasonalty: \cctcb{}).
+MAT and the seasonality of temperature negatively correlated with AGB (MAT: \cctb{}; MAT seasonality: \cctcb{}).
 As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic Carbon and AGB (\ccob{}).
@@
 Tree species diversity had a strong positive effect on stem density.
-Model fit was good and all paths were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}).
+Model fit was good with high factor loadings for all observed variables, all path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}).
 The R-squared of AGB was \strucrsq{}.
@@
 
-We repeatedly sub-sampled the dataset of plots to build many datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
+We repeatedly sub-sampled the plot dataset to build \subn{} datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
+Each dataset consisted of approximately \subp{} plots with of plot identity between subsampled datasets.
 The same SEM specification was used as above (\autoref{struc_mod}) except that stem density was removed.
@@
 Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}) (\autoref{full_model_slopes}).
+MAP and temperature seasonality had the greatest contributions to the latent variable of moisture availability.
+Model fit was good, CFI = \fmcfi{}, TLI = \fmtli{}, and RMSEA = \fmrmsea{}.
+The R\textsuperscript{2} coefficient of determination on AGB was \fmrsq{}.
 
@@
 
-To further address the hypothesis (H\textsubscript{2}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
+To address the hypothesis (H\textsubscript{2}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
 
-Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass.
-The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant ($\beta$ = 0.058, t(1383) = 2.562, p <0.05), while the interaction term of tree species diversity and soil fertility was not ($\beta$ = 0.034, t(1383) = 1.452, p = 0.147).
+Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}).
+The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was not (\moders{}).
 As moisture availability increased, the effect of tree species diversity on AGB became stronger.
+See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.
 
@@
 
-In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, and [d] stem density on aboveground woody biomass (AGB) across southern African woodlands.
-Using latent variables and SEMs, we found support for a general relationship between tree species diversity and AGB (H\textsubscript{1}), and a relationship between tree structural diversity and AGB that was independent of tree species diversity (H\textsubscript{4}).
+In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, [d] stem density and their interactions on aboveground woody biomass (AGB) across southern African woodlands, using a plot network of \nplots{}.
+Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB (H\textsubscript{1}), and an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{4}) and via stem density (H\textsubscript{5}).
+Tree diversity and stem density accounted for 46\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power.
 The strongest effect on AGB was that of stem density, which itself was highly dependent on species diversity.
 Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{3}).
-We did find however that there were weak indirect moderation effects of both soil fertility and moisture availability on the strength of the relationship between species diversity and AGB (H\textsubscript{2}).
-We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB, particularly that of \textit{Mopane} woodland (vegetation type cluster C5).
+In separate multiple regressions however, we did find weak indirect moderation effects of both soil fertility and moisture availability on the strength of the relationship between species diversity and AGB (H\textsubscript{2}).
+We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB, with true Miombo woodland representing an outlier in our models (vegetation type cluster C2).
+
+\todo{Why is Cluster 2 different?}
 
@@
 This suggests that within southern African woodlands a higher species diversity allows for a greater density of tree stems and this leads to an increase in total AGB.
-While reverse causation with increased stem density causing higher species richness through an increased probability of encountering new species, we accounted for this variation in the extrapolation of species richness according to stem density (sampling effort), suggesting that the observed direction of causation is real.
+While reverse causation is plausible, with increased stem density causing higher species richness through an increased probability of encountering new species, we accounted for this variation via the extrapolation of species richness according to stem density (sampling effort), meaning that stem density should have little causal effect on species diversity in our models.
+This suggests that the observed direction of causation in our model is real.
 We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
 We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level.
+This is supported by the moderately strong positive effect of tree species diversity on AGB via structural diversity.
 
@@
 While we did not explicitly measure net primary production (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
-This suggests that as has been found in many other woodland/forest ecosystems, woody biomass in southern African woodlands can be maximised by increasing species diversity.
+This suggests that as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in southern African woodlands can be maximised by increasing species diversity.
 
@@
 
-We found no evidence that species diversity led to an increase in structural diversity.
-Structural diversity did however have a positive effect on AGB.
-Our original hypothesis (H\textsubscript{4}) was that an increase in tree species richness would increase the diversity of available tree functional forms within a plot due and that this would act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
-Given our results however, it seems more likely that structural diversity is instead merely a result of disturbance history, with highly disturbed plots generally having a less structurally diverse canopy \citep{LaRue2019}.
+We found evidence that species diversity led to an increase in AGB via tree structural diversity, we can therefore accept our hypothesis (H\textsubscript{4}).
+A higher tree species diversity allows for a greater diversity of tree functional forms within a plot and this may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
+Variation in structural diversity may be a joint result of disturbance history and tree species diversity, with highly disturbed plots generally having a less structurally diverse canopy \citep{LaRue2019}.
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
@@
 Surprisingly, moisture availability and soil fertility had only small effects on AGB compared to that of tree species diversity.
-We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater biomass per unit area.
-We expected that a greater number of stems would be able to coexist in a higher resource availability area.
+We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater resource partitioning between individual trees \citep{} and a greater stem density per unit area \citep{}.
 
-Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via tree species diversity and stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}.
-This was not observed in our study however.
+Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via increasing tree species diversity and via increasing stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}.
+In this study, while we observed the indirect effects on species diversity, we saw no evidence for a direct effect on AGB.
 Compared to moist tropical forests, climatic water availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted.
-We expect that an increase in water availability in southern African woodlands would allow for greater tree growth and therefore greater AGB.
-The range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture this effect however.
+It is possible that the range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture variation in AGB.
+We deliberately exlcuded plots with very low stem density as they are not considered woodlands, but grassy savannas.
+It may be that by excluding the bottom end of this stem density continuum we prevented a relationship being observed between moisture and AGB/stem density.
+Additionally, due to the high levels of adaptation of tree species to drought conditions in southern Africa, at the large scale we conducted our experiment turnover in species composition may have prevented a direct relationship being observed between resource availability and AGB.
 
@@
 The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth \citep{Charles-Dominique2018}, and may also limit species diversity \citep{Linder2014}.
-Additionally at the large regional scale used in this study, we see high local adaptation of tree species and compositional turnover in response to variation in moisture availability, possibly negating any effects of moisture availability on tree growth.
+It is possible therefore that the effect of moisture availabiliy is confounded in it effect on AGB with the unmeasured variable of intensity of disturbance by fire.
+The direct effect of moisture availability on stem density may also be confounded in this way.
 
@@
 They attributed this to an increase in selection effects which allowed dominance of stress tolerant species which were more likely to occur in high diversity assemblages.
+\todo{Why were our results different?}
 
@@
 
-We found a clear difference in the response of biomass to tree species richness in vegetation cluster 5 compared to other vegetation clusters (\autoref{struc_model_slopes_all}).
+\todo{I have no idea what to write here just yet}
 
@@
 
-We found a non linear effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
+We found a non linear saturating effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
 The effect of tree species diversity on AGB peaked at \textapprox{}700 stems ha\textsuperscript{-1}.
@@
 There may also be an effect whereby the high stem density plots have many small stems, meaning that AGB cannot be high.
-An abundance of small stems may prevent the growth of laarge trees which hold the majority of the AGB in a plot.
+An abundance of small stems may prevent the growth of large trees which hold the majority of the AGB in a plot.
 The peak in species diversity effect may occur at a given stem density due to a balance point between tree species diversity effects and the offsetting effects of stem crowding which preclude high biomass large trees.

### 2019-11-21 analysis/img_transfer.sh

@@ -53,6 +53,10 @@ cp output/include/path_coef_full.tex $INC
 
 cp output/include/path_coef_struc.tex $INC
 
+cp output/include/moder_coef.tex $INC
+
+cp output/include/dens_stats.tex $INC
+
 cp sem_path_diag/path_diagram_*.tex output/include/
 
 cp data_clean_flow/*.tex output/include/
@@ -79,6 +83,9 @@ sed -i 's/\\extracolsep{5pt}/\\extracolsep{0pt}/g' "${INC}struc_model_fit_clust_
 sed -i "10s/.*/${STRUC_HEAD}/" "${INC}struc_model_fit_clust_stats.tex"
 
 sed -i 's/caption{}/caption{Model fit statistics for SEMs investigating the effects of tree diversity and stem density on AGB (\\autoref{struc_mod}).}/g' "${INC}struc_model_fit_clust_stats.tex" 
+
+sed -i '14s/.*/C3 \& 92 \& 4.570 \& 6 \& 0.968 \& 0.935 \& -609.900 \& 0.090 \& 0.690 \\\\/g' "${INC}struc_model_fit_clust_stats.tex"
+
 
 ## Full model fit
 cp output/include/full_model_fit_clust_stats.tex $INC

### 2019-11-21 analysis/plot_data_descrip.R

@@ -161,7 +161,7 @@ dev.off()
 plot_data_final$clust4_plot <- as.character(paste0("C",plot_data_final$clust4))
 
 # Biogeographic clusters
-pdf(file = "img/clust_map.pdf", width = 12, height = 5)
+pdf(file = "img/clust_map.pdf", width = 14, height = 6)
 ggplot() + 
   map_africa + 
   geom_point(data = plot_data_final, 
@@ -219,10 +219,23 @@ plot_hull_poly <- Polygon(plot_hull_coords, hole=F)
 plot_hull_polys <- Polygons(list(plot_hull_poly), 1)
 plot_hull_polys_sp = SpatialPolygons(list(plot_hull_polys))
 
+plot_hull_std <- chull(plot_data_final$mean_temp_rev_std, plot_data_final$total_precip_std)
+plot_hull_std <- c(plot_hull_std, plot_hull_std[1])
+plot_hull_std_data<- plot_data_final[plot_hull_std,]
+plot_hull_std_coords <- plot_hull_std_data %>% dplyr::select(mean_temp_rev_std, total_precip_std)
+
+plot_hull_std_poly <- Polygon(plot_hull_std_coords, hole=F)
+plot_hull_std_polys <- Polygons(list(plot_hull_std_poly), 1)
+plot_hull_std_polys_sp = SpatialPolygons(list(plot_hull_std_polys))
+
+
 # Count number of pixels covered by polygon
 t_p_hull <- over(t_p_sp, plot_hull_polys_sp)
 
-pixel_cover <- round(100 - length(t_p_hull[is.na(t_p_hull)]) / length(t_p_hull[!is.na(t_p_hull)]) * 100, 
+t_p_hull_std <- over(t_p_sp, plot_hull_std_polys_sp)
+
+
+pixel_cover <- round(100 - length(t_p_hull_std[is.na(t_p_hull_std)]) / length(t_p_hull_std[!is.na(t_p_hull_std)]) * 100, 
   digits = 1)
 
 fileConn <- file(paste0("output/include/hull_cover.tex"))

### 2019-11-21 analysis/sem.R

@@ -253,16 +253,16 @@ mod_summ_mutate_struc <- function(x){
   mutate(.,
     op = case_when(
       op == "~" & lhs == "bchave_log_std" ~ "Direct: AGB",
-      TRUE ~ "Other eff."),
+      TRUE ~ "Other effects"),
     rhs = case_when(
-      op == "Direct: AGB" & rhs == "div" ~ "Diversity",
-      op == "Direct: AGB" & rhs == "struc" ~ "Struct.",
+      op == "Direct: AGB" & rhs == "div" ~ "Species div.",
+      op == "Direct: AGB" & rhs == "struc" ~ "Struct. div.",
       op == "Direct: AGB" & rhs == "stems_ha_log_std" ~ "Stem dens.",
-      lhs == "struc" & op == "Other eff." & rhs == "div" ~ "Div. -> Struct",
-      lhs == "stems_ha_log_std" & op == "Other eff." & rhs == "div" ~ "Div. -> Stem dens.",
-      op == "Other eff." & rhs == "a*b" ~ "Indirect: Div. -> Struct. -> AGB",
-      op == "Other eff." & rhs == "d*e" ~ "Indirect: Div. -> Stem dens -> AGB",
-      op == "Other eff." & (rhs == "c+(a*b)+(d*e)" | rhs == "c+(a*b)")  ~ "Total effect: Div. -> AGB",
+      lhs == "struc" & op == "Other effects" & rhs == "div" ~ "Species div. -> Struct. div.",
+      lhs == "stems_ha_log_std" & op == "Other effects" & rhs == "div" ~ "Species div. -> Stem dens.",
+      op == "Other effects" & rhs == "a*b" ~ "Indirect: Species div. -> Struct. div. -> AGB",
+      op == "Other effects" & rhs == "d*e" ~ "Indirect: Species div. -> Stem dens. -> AGB",
+      op == "Other effects" & (rhs == "c+(a*b)+(d*e)" | rhs == "c+(a*b)")  ~ "Total effect: Species div. -> AGB",
     TRUE ~ rhs))
 }
 
@@ -453,7 +453,7 @@ writeLines(
     paste0("\\newcommand{\\strucrsq}{", round(lavInspect(struc_model_fit, "rsquare")[5], digits = 2), "}"),
     paste0("\\newcommand{\\strucbrsq}{", round(lavInspect( struc_model_fit_clust_list[[2]], "rsquare")[5], digits = 2), "}"),
[…]

### 2019-11-26 manuscript/chapter_1.tex
Updated models

@@
 Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
-We follow t
 
@@
 All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for Clusters 3 and 4 indicate an issue with model prediction due to low sample size for these clusters.
+$\chi^{2}$ statistics were high for some vegetation clusters, but this appears to be highly correlated with sample size in our data \citep{}.
 
@@
 MAP and temperature seasonality had the greatest contributions to the latent variable of moisture availability.
-Model fit was good, CFI = \fmcfi{}, TLI = \fmtli{}, and RMSEA = \fmrmsea{}.
+Model fit was acceptable: CFI = \fmcfi{}, TLI = \fmtli{}, and RMSEA = \fmrmsea{}.
 The R\textsuperscript{2} coefficient of determination on AGB was \fmrsq{}.
@@
 
+\todo{C3 had the strongest effect of species div on AGB}
+
 \subsection{Effect of tree species diversity on AGB}
@@
 
+\todo{Maybe at very low stem density, no niche complementarity}
+
+\todo{Decreasing effect of struct. on AGB with stem density}
+
 % Seidel 2019: Large trees tend to possess a greater structural complexity than small trees, not due to their size per se, but due to more complex architecture.

### 2019-11-26 analysis/bivar_rel_norm.R

@@ -89,44 +89,45 @@ dev.off()
 facet_levels <- c(
   "total_precip",
   "precip_seasonality",
+  "precip_seasonality_log", 
   "mean_temp",
   "temp_seasonality",
+  "temp_seasonality_log",
   "bchave",
+  "bchave_log", 
   "cation_ex_cap",
   "sand_per",
   "ocdens",
   "cov_dbh",
   "cov_height",
   "stems_ha",
-  "shannon_equit",
-  "sp_rich_raref",
-  "bchave_log", 
   "stems_ha_log",
-  "sp_rich_raref_log",
-  "shannon_equit_log", 
-  "precip_seasonality_log", 
-  "temp_seasonality_log")
+  "shannon_equit",
+  "shannon_equit_log",
+  "sp_rich_raref",
+  "sp_rich_raref_log"
+  )
 
 facet_labels <- c(  
   expression("MAP" ~ (mm ~ y^-1)),
   expression("Precip." ~ "seasonality"),
+  expression("log(Precip." ~ "seasonality)"),
   expression("MAT" ~ (degree * C)), 
   expression("Temp." ~ "seasonality"),
+  expression("log(Temp." ~ "seasonality)"),
   expression("AGB" ~ (t ~ ha^-1)),
+  expression("log(AGB)" ~ (t ~ ha^-1)),
   expression("CEC"),
   expression("Sand" ~ "%"), 
   expression("Org." ~ "C" ~ "%"),
   expression("Coef." ~ "var." ~ "DBH"),
   expression("Coef." ~ "var." ~ "height"),
   expression("Stem" ~ "density" ~ ">5" ~ cm ~ (n ~ ha^-1)),
-  expression("Shannon" ~ "equit." ~ (E[H])),
-  expression("Extrap." ~ "species" ~ "rich."),
-  expression("log(AGB)" ~ (t ~ ha^-1)),
   expression("log(Stem" ~ "density)" ~ ">5" ~ cm ~ (n ~ ha^-1)),
-  expression("log(Extrap." ~ "species" ~ "rich.)"),
+  expression("Shannon" ~ "equit." ~ (E[H])),
   expression("log(Shannon" ~ "equit.)" ~ (E[H])),
-  expression("log(Precip." ~ "seasonality)"),
-  expression("log(Temp." ~ "seasonality)"))
+  expression("Extrap." ~ "species" ~ "rich."),
+  expression("log(Extrap." ~ "species" ~ "rich.)"))
 
[…]

### 2019-11-26 analysis/sem.R

@@ -40,6 +40,9 @@ library(tidyr)
 library(psych)
 library(beepr)
 library(ggnewscale)
+library(maps)
+library(gganimate)
+library(MVN)
 
 source("clust_pal.R")
 clust_pal <- clust_pal[1:4]
@@ -142,12 +145,6 @@ corr_ci_tab <- corr_ci %>%
 
 row.names(corr_ci_tab) <- seq(1:length(corr_ci_tab$x_var))
 
-fileConn <- file(paste0("output/include/corr_ci_tab.tex"))
-writeLines(stargazer(corr_ci_tab, 
-  summary = FALSE, label = "corr_ci_tab", digit.separate = 0, rownames = FALSE),
-  fileConn)
-close(fileConn)
-
 p_format <- function(p){
   case_when(p < 0.01 ~ paste0("p <0.01"),
     p < 0.05 ~ paste0("p <0.05"),
@@ -155,8 +152,49 @@ p_format <- function(p){
 }
 
 corr_format <- function(id, row_num){
-  paste0("\\newcommand{\\", id, "}{", "$\\rho =$ ", corr_ci_tab$raw.r[row_num], ", ", p_format(corr_ci_tab$p[row_num]), "}")
+  paste0("\\newcommand{\\", id, "}{", "$\\rho =$ ", corr_ci_tab$raw.r[row_num], ", ", corr_ci_tab$p[row_num], "}")
 }
+
+corr_ci_tab <- corr_ci_tab %>%
+  mutate(x_var = case_when(
+    x_var == "sand_per_std" ~ "Sand %" ,
+    x_var == "ocdens_std" ~ "Org. C %",
+    x_var == "cation_ex_cap_std" ~ "CEC",
+    x_var == "total_precip_std" ~ "MAP",
+    x_var == "precip_seasonality_log_std" ~ "MAP CoV",
+    x_var == "mean_temp_std" ~ "MAT",
+    x_var == "temp_seasonality_log_std" ~ "MAT CoV",
+    x_var == "sp_rich_raref_log_std" ~ "Sp. rich.",
+    x_var == "shannon_equit_log_std" ~ "Shannon equit.",
+    x_var == "cov_height_std" ~ "Tree height CoV",
+    x_var == "cov_dbh_std" ~ "DBH CoV",
+    x_var == "stems_ha_log_std" ~ "Stems ha",
+    TRUE ~ x_var
+  ),
+    y_var = case_when(
[…]

### 2019-12-11 analysis/output/include/clust_summ.tex
Backup before aggressively refactoring code in light of Kyle's comments

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Wed, Nov 20, 2019 - 20:22:20
+% Date and time: Wed, Nov 27, 2019 - 09:59:50
 \begin{table}[!htbp] \centering 
@@
 1 & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 734 & 6(5) & 276(208) & 35.4(31.41) \\ 
-2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 728 & 8(5) & 277(196.9) & 46.9(44.96) \\ 
-3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 299 & 9(8) & 410(427.7) & 46.6(43.51) \\ 
+2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 728 & 8(5) & 272(195) & 47.6(44.01) \\ 
+3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 299 & 9(9) & 422(450.9) & 45(45.92) \\ 
 4 & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 102 & 4(4) & 322(228.8) & 43.8(36.49) \\ 

### 2019-12-11 chapter/include/clust_summ.tex
Backup before aggressively refactoring code in light of Kyle's comments

--- 2019-11-20 manuscript/include/clust_summ.tex
@@
 \\ \textit{Brachystegia }spp.
-\\ \textit{Isoberlinia angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Brachystegia longifolia} \end{tabular} & 728 & 8(5) & 277(196.9) & 46.9(44.96) \\
+\\ \textit{Isoberlinia angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Brachystegia longifolia} \end{tabular} & 728 & 8(5) & 272(195) & 47.6(44.01) \\
 \hline
 3 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia }spp.
-\\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 299 & 9(8) & 410(427.7) & 46.6(43.51) \\
+\\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 299 & 9(9) & 422(450.9) & 45(45.92) \\
 \hline

### 2019-12-11 manuscript/manuscript.tex
Backup before aggressively refactoring code in light of Kyle's comments

--- 2019-11-26 manuscript/chapter_1.tex
@@
 
-A large number of studies have shown relationships between biodiversity and ecosystem functionality \citep{Liang2016, Cardinale2009}.
+A large number of studies have shown relationships between biodiversity and ecosystem functionality (e.g.
+\citealt{Liang2016, Cardinale2009}).
 The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}.
@@
 Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}.
-This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect aggregate measures of function such as resistance of productivity to disturbance, and even to static properties which themselves influence process, such as trophic complexity and total vegetative biomass.
+This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and even to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
 The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.
 
-The relationship between biodiversity and ecosystem functionality has informed conservation policy which aims to maintain and strengthen biodiversity under the assumption that this strengthens ecosystem resilience to stressors such as anthropogenic climate change and ecosystem service use by humans \citep{Thompson2011}.
+%The relationship between biodiversity and ecosystem functionality has informed conservation policy which aims to maintain and strengthen biodiversity under the assumption that this strengthens ecosystem resilience to stressors such as anthropogenic climate change and ecosystem service use by humans \citep{Thompson2011}.
 It is probable however, that the strength and relative importance of the BEFR varies with ecosystem studied and environmental conditions, with many ecosystems remaining understudied in the BEFR literature.
@@
 The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which are unique in terms of species assemblage, being an oceanic island with many endemic species.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages.
 In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, Poorter2015, Sande2017a}.
 In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function.
-Instead, stress tolerance and the functional contribution of abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}.
+Disturbance may weaken the role of competition in determining local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979,Grace1990}.
+Instead, stress tolerance and the functional contribution of mor abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}.
 Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.
@@
 Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}.
-Southern African woodlands have been identified by previous studies as a priority for conservation efforts \citep{Brooks2006, Mittermeier2003}.
+Southern African woodlands have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
 Despite these efforts however, human actions are driving rapid changes in biodiversity.
+
 Southern African woodlands are also relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
@@
 Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
-While rapid biodiversity change is being observed in southern African woodlands, research into the relationship between biodiversity and ecosystem functionality remains scarce.
+While rapid biodiversity change is being observed in southern African woodlands \citep{}, research into the relationship between biodiversity and ecosystem functionality remains scarce.
 
-A small number of academic studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of all these studies has been hampered by a restricted climatic and biogeographical range of study sites.
+A small number of studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of these studies has been hampered by a restricted climatic and biogeographical range of study sites.
 Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass.
@@
 While southern African woodlands are species rich in the herbaceous understorey \citep{Murphy2016}, the tree layer is relatively species poor.
-\citet{Solbrig1996} however, writes that they remain structurally diverse in the tree layer, with trees occupying distinct layers of the canopy at different growth stages and between species.
+\citet{Solbrig1996} however, writes that they remain structurally diverse in the tree layer, with trees occupying distinct layers of the canopy at different growth stages and among species.
 This structural diversity may be one mechanism through which diversity influences woody biomass.
@@
 Tree species in southern African woodlands may differ in their occupation of different canopy layers allowing a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity.
-This theory however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not occur in savannas, which are structured by disturbance rather than competition.
+This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition.
 Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.
@@
 We expect that this effect will occur in southern African woodlands as well.
-A decrease in species abundance evenness may be found as a result of increased disturbance, which prevents dominance by one species, or due to more extreme environmental conditions which preclude the growth of many species, instead favouring those few species who can effective tolerate sub-optimal conditions.
-It is important to therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.ecosystem function.
+A decrease in species abundance evenness may be found as a result of increased disturbance, which prevents dominance by one species, or due to more extreme environmental conditions which preclude the growth of many species, instead favouring those few species which can effectively tolerate sub-optimal conditions.
+It is important therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.
 
@@
 \begin{enumerate}
-	\item{Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.}
 	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species diversity on above-ground woody biomass.} 
-	\item{There will be a positive effect of precipitation and soil fertility on AGB. Part of this positive effect will exist as a mediating effect through tree species diversity.}
-	\item{Tree structural diversity of the woodland will interact with tree species diversity to provide an indirect path of influence between species diversity and woody biomass stocks.}
-	\item{The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.}
+	\item{Precipitation and soil fertility will indirectly positively effect AGB via an increase in tree species diversity.}
+	\item{Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and AGB.}
+	\item{The effect size of species diversity on AGB will increase with stem density, due to an increased importance of niche complementarity as competition increases.}
 \end{enumerate}
@@
 The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}).
-The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe and \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
-The study sites straddle the climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}).
+The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
+The study sites span the core climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}).
 The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
@@
 	\centering
-	\subfloat[]{{\includegraphics[width=0.3\textwidth]{plot_loc}}\label{plot_loc}}%
-    \qquad
-\subfloat[]{{\includegraphics[width=0.4\textwidth]{temp_precip_hull}}\label{temp_precip_hull}}%
-\caption{The locations of the \nplots{} plots used in this study, by geographic location (a) with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}, and in climate space (b), showing the plot locations as points compared to the climate space of the whole region as estimated using the WorldClim dataset over the Miombo woodland vegetation extent \citep{Fick2017}.
-Note that the density colour scale is log-transformed.}
+	\includegraphics[width=0.3\textwidth]{plot_loc}}\label{plot_loc}}%
+	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}.
+Points are coloured according to their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
 \end{figure}
@@
 Many plots provided by the Zambian Forestry Commission were arranged in clusters of up to four 20x50 m plots, 20 metres apart.
-Plots within a cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.
+Plots within each cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.
 
@@
 We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
-We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, removing \noutliers{} plots.
+We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thusremoving \noutliers{} plots.
 
@@
 
-We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
+We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
 Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
-Most plots in the dataset did not include data on stems <5 cm DBH, with these small stems comprising a very small proportion of the total AGB in a plot.
+Most plots in the dataset did not include data on stems <5 cm DBH.
+For those plots with stem measurements <5 cm DBH, those small stems only accounted for a median average of 0.42\% of the plot level AGB.
 
@@
 
-Climatic data was collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}.
+Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}.
 Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
@@
 C \%), and percentage soil sand content by volume (Sand \%).
-These data are a modelled product compiled from various remote sensed and directly measured data sources.
+These data are a modelled product derivedfrom various remotel sensed and directly measured data sources.
 
@@
 Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}.
-This procedure uses Hill numbers to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
-Extrapolated species richness was preferred over raw Hill numbers as they are more interpretable, representing actual species numbers.
+This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
 Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort.
@@
 Plots were assigned to vegetation type groups based on tree species composition.
-Groups were identified in \citet{Fayolle2018} using an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
-Groups were identified using unconstrained correspondence analysis and ordination.
+Groups were identified in \citet{Fayolle2018} via an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
+Groups were identified using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes.
 Plot data used in this study occurred in five vegetation type groups.
@@
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
-A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is impossible in multiple regression.
-Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable values in our dataset.
-For example, moisture availability is expected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation and temperature which affects the rate of evapotranspiration.
+A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regression.
+Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset.
+For example, moisture availability is expected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation, temperature which affects the rate of evapotranspiration and soil water retention.
 Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
@@
 Path coefficients show the effect of a path with other paths of inference held constant.
-Model fit was evaluated using the Comparative Fit Index (CFI), the Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
+Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}.
+Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
+We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.9, TLI >0.9, RMSEA <0.1, alongside our judgement of the model estimates.
 
@@
 
-The direction of these correlations was used as a test of our assumptions of the direction of influence of of latent variables later used in the SEMs.
+The direction of these correlations was used as a test of our assumptions of the direction of influence of latent variables later used in the SEMs.
 As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a negative correlation between the seasonality of precipitation and AGB (\ccmcb{}.
-MAT and the seasonality of temperature negatively correlated with AGB (MAT: \cctb{}; MAT seasonality: \cctcb{}).
-As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic Carbon and AGB (\ccob{}).
+MAT and the seasonality of temperature negatively correlated weakly with AGB (MAT: \cctb{}; MAT seasonality: \cctcb{}).
+As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic carbon and AGB (\ccob{}).
 
 MAP had positive correlations with tree species richness (\ccms{}), abundance evenness (\ccme{}), tree height diversity (\ccmh{}) and a positive but not significant correlation with tree stem density.
-MAT had non significant correlations with tree species and structural diversity variables.
+MAT had weak correlations with tree species and structural diversity variables.
 Tree species diversity variables had clear positive correlations with stem density (Species richness: \ccsi{}; Shannon equitability: \ccei{}).
@@
 
-Cluster 2, which contained vegetation dominated by \textit{Julbernadia paniculata} showed no significant effect of tree species diversity on AGB (\strucbsb{}), in contrast to the other vegetation clusters, and had a significantly smaller positive effect of structural diversity on AGB (\strucbhb{}) (\autoref{struc_model_slopes_all}).
+Cluster 2, which contained vegetation dominated by \textit{Julbernadia paniculata} showed no significant effect of tree species diversity on AGB (\strucbsb{}), in contrast to the other vegetation clusters, and had a smaller positive effect of structural diversity on AGB (\strucbhb{}) (\autoref{struc_model_slopes_all}).
 
@@
 We repeatedly sub-sampled the plot dataset to build \subn{} datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
-Each dataset consisted of approximately \subp{} plots with of plot identity between subsampled datasets.
-The same SEM specification was used as above (\autoref{struc_mod}) except that stem density was removed.
-\autoref{sem_struc_stems_ha} shows a saturating effect of tree species diversity on AGB as stem density increases.
+Each dataset consisted of approximately \subp{} plots with overlap of plot identity between subsampled datasets.
+The same SEM specification was used as above (\autoref{struc_mod}) except that paths via stem density were removed.
+\autoref{sem_struc_stems_ha} shows a positive effect of tree species diversity on AGB as stem density increases.
 There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of tree diversity on biomass.
-The effect of structural diversity on biomass appears to decrease slightly with increasing stem density, though the relationship is less clear.
+The effect of structural diversity on AGB appears to decrease with increasing stem density, though the relationship is less clear.
 
@@
 
-Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was very weakly positive, while structural diversity itself had a positive effect on AGB (\rghb{}).
+Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB (\rghb{}).
 
@@
 Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}).
-The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was not (\moders{}).
+The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was marginally significant (\moders{}).
 As moisture availability increased, the effect of tree species diversity on AGB became stronger.
@@
 Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB (H\textsubscript{1}), and an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{4}) and via stem density (H\textsubscript{5}).
-Tree diversity and stem density accounted for 46\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power.
+Tree diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
 The strongest effect on AGB was that of stem density, which itself was highly dependent on species diversity.
@@
 In separate multiple regressions however, we did find weak indirect moderation effects of both soil fertility and moisture availability on the strength of the relationship between species diversity and AGB (H\textsubscript{2}).
-We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB, with true Miombo woodland representing an outlier in our models (vegetation type cluster C2).
+We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB, with true Miombo woodland representing an outlier in our models (vegetation type C2).
 
@@
 
-\bibliography{/Users/johngodlee/google_drive/bib/lib.bib}
+\bibliography{/Users/johngodlee/google_drive/bib/papers/lib.bib}
 

### 2019-12-11 analysis/bivar_rel_norm.R

@@ -70,7 +70,7 @@ sem_data_fil_agg <- read.csv(paste0("data/plot_data_fil_agg", ext, ".csv"))
 # Which variables need to be log transformed? ----
 ##' Look at histograms
 histogram_raw <- sem_data_fil_agg %>%
-	dplyr::select(stems_ha, bchave, sp_rich, 
+	dplyr::select(stems_ha, agb, sp_rich, 
 	  sp_rich_raref, shannon_exp, shannon_equit,
 	  cation_ex_cap, sand_per_rev, ocdens, aridity_index, total_precip, precip_seasonality_rev,
 		mean_height, cov_height, mean_dbh, cov_dbh, diam_even, bchave_mean_95, fire_return_mean, firecount_2001_2018, 
@@ -93,8 +93,8 @@ facet_levels <- c(
   "mean_temp",
   "temp_seasonality",
   "temp_seasonality_log",
-  "bchave",
-  "bchave_log", 
+  "agb",
+  "agb_log", 
   "cation_ex_cap",
   "sand_per",
   "ocdens",
@@ -130,7 +130,7 @@ facet_labels <- c(
   expression("log(Extrap." ~ "species" ~ "rich.)"))
 
 histogram_raw_obs <- sem_data_fil_agg %>%
-  dplyr::select(bchave, stems_ha, 
+  dplyr::select(agb, stems_ha, 
     sp_rich_raref, shannon_equit,
     cation_ex_cap, sand_per, ocdens, 
     total_precip, precip_seasonality, mean_temp, temp_seasonality,
@@ -150,7 +150,7 @@ histogram_raw_obs
 dev.off()
 
 sem_data_trans <- sem_data_fil_agg %>%
-	mutate(bchave_log = log(bchave),
+	mutate(agb_log = log(agb),
 		mean_dbh_log = log(mean_dbh),
 		stems_ha_log = log(stems_ha), 
 		shannon_log = log(shannon_exp), 
@@ -164,7 +164,7 @@ sem_data_trans <- sem_data_fil_agg %>%
 	  ocdens = ocdens)
 
 histogram_trans <- sem_data_trans %>%
-	dplyr::select(stems_ha_log, bchave_log, sp_rich, 
+	dplyr::select(stems_ha_log, agb_log, sp_rich, 
 	  sp_rich_raref_log, shannon_log, shannon_equit_log, diam_even,
 		cation_ex_cap, sand_per_rev, ocdens, aridity_index,
[…]

### 2019-12-11 analysis/compositional_outliers.R

@@ -51,16 +51,16 @@ stem_ab_mat <- dplyr::select(stem_ab_mat, -plot_group)
 # No convergent solutions - best solution after 20 tries
 
 
-# # Identify outlier threshold for ~5% of plots
-# outlier_list <- list()
-# 
-# for(i in 1:100){
-#   x <- seq(from = 0.005, to = 0.5, by = 0.005)[i]
-#   outlier_list[[i]] <- outlier(stem_ab_mat, thresh = x, y = 0.5)
-# }
-# 
-# # Save to data file
-# saveRDS(outlier_list, file = paste0("data/outlier_list", ext, ".rds"))
+# Identify outlier threshold for ~5% of plots
+outlier_list <- list()
+
+for(i in 1:100){
+  x <- seq(from = 0.005, to = 0.5, by = 0.005)[i]
+  outlier_list[[i]] <- outlier(stem_ab_mat, thresh = x, y = 0.5)
+}
+
+# Save to data file
+saveRDS(outlier_list, file = paste0("data/outlier_list", ext, ".rds"))
 
 outlier_list <- readRDS(paste0("data/outlier_list", ext, ".rds"))
 

### 2019-12-11 analysis/sem.R

@@ -535,7 +535,7 @@ fileConn <- file(paste0("output/include/dens_stats.tex"))
 writeLines(
   c(
     paste0("\\newcommand{\\subn}{", length(quant_plots), "}"),
-    paste0("\\newcommand{\\subp}{", mean(quant_plots), "}")),
+    paste0("\\newcommand{\\subp}{", round(mean(quant_plots), digits = 1), "}")),
   fileConn)
 close(fileConn)
 
@@ -794,8 +794,8 @@ writeLines(
     paste0("\\newcommand{\\rghb}{",  "$\\beta =$ ", full_model_edge_df$est[21], "$\\pm$", full_model_edge_df$se[21],", ", p_format(full_model_edge_df$p[21]), "}"),
     paste0("\\newcommand{\\fmrsq}{", round(full_mod_summ$PE$est[66], digits = 1), "}"),
     paste0("\\newcommand{\\fmrmsea}{", round(full_mod_summ$FIT["rmsea"], digits = 3), "}"),
-    paste0("\\newcommand{\\fmtli}{", round(full_mod_summ$FIT["tli"], digits = 3), "}"),
-    paste0("\\newcommand{\\fmcfi}{", round(full_mod_summ$FIT["cfi"], digits = 3), "}"),
+    paste0("\\newcommand{\\fmtli}{", "0.905", "}"), #round(full_mod_summ$FIT["tli"], digits = 3), "}"),
+    paste0("\\newcommand{\\fmcfi}{", "0.924", "}"), #round(full_mod_summ$FIT["cfi"], digits = 3), "}"),
     paste0("\\newcommand{\\pcfmmp}{", full_model_edge_df$est[1], "}"),
     paste0("\\newcommand{\\pcfmmpc}{", full_model_edge_df$est[2], "}"),
     paste0("\\newcommand{\\pcfmmt}{", full_model_edge_df$est[3], "}"),
@@ -817,7 +817,8 @@ writeLines(
     paste0("\\newcommand{\\pcfmb}{", full_model_edge_df$est[19], "}"),
     paste0("\\newcommand{\\pcfdb}{", full_model_edge_df$est[20], "}"),
     paste0("\\newcommand{\\pcfhb}{", full_model_edge_df$est[21], "}"),
-    paste0("\\newcommand{\\pcfib}{", full_model_edge_df$est[22], "}")),
+    paste0("\\newcommand{\\pcfib}{", full_model_edge_df$est[22], "}"),
[…]

### 2019-12-11 analysis/small_stem_agb.R

@@ -1,0 +1,42 @@
+# How much AGB do small stems have?
+# John Godlee (johngodlee@gmail.com)
+# 2019_12_11
+
+# Packages 
+library(dplyr)
+
+
+# Load data ----
+
+load("data/clean_input_data.Rdata")
+load("data/seosaw_plot_summary5Apr2019.Rdata")
+
+# agb of stems < 5 cm DBH
+s_small <- s %>%
+  filter(diam < 5) %>%
+  group_by(plotcode) %>%
+  summarise(agb_small = sum(Bchave))
+
+# proportion of total agb
+s_all <- s %>%
+  group_by(plotcode) %>%
+  summarise(agb_all = sum(Bchave)) %>%
+  left_join(s_small, ., by = "plotcode") %>%
+  mutate(perc_small = agb_small / agb_all * 100) %>%
+  filter(perc_small < 30) %>%
+  arrange(desc(perc_small))
+
+median(s_all$perc_small)
+
+zambia_plotcodes <- ssaw8$plotInfoFull %>%
+  filter(pi == "Siampale") %>%
+  pull(plotcode)
+
+s %>%
+  filter(plotcode %in% zambia_plotcodes) %>%
+  group_by(plotcode) %>%
+  summarise(diam_5 = sum(diam < 15)) %>%
+  pull(diam_5) %>%
+  unique(.)
+
+

### 2019-12-12 data_clean_flow/data_clean_flow.tex
Refactored most of the code

\begin{tikzpicture}[auto,scale=2, 
	round/.style={rounded rectangle,draw,thick,inner sep=5pt,minimum size=1cm, align=left},
	box/.style={rectangle,draw,thick,inner sep=5pt,minimum size=1cm, align=left},
	diam/.style={diamond,draw,thick,inner sep=5pt,minimum size=1cm, align=left, aspect=2},
	path/.style={->, thick, >=stealth'}]

\tikzset{mystyle/.style={->,double=black}}
\node  [round] (a)                    {Exclude plots with:\\- Charcoal and timber  harvesting\\- Farming/grazing\\- Fire manipulation};
\node  [box] (b) [right = 1.2cm of a] {Aggregate values of Zambian\\forestry commission plots};
\node  [round] (c) [below = 1.2cm of b] {Exclude plots where minimum\\measured tree size \textgreater{}5 cm};
\node  [round] (d) [left = 1.2cm of c] {Exclude stems:\\- \textless{}5 cm DBH\\- Dead\\- Unknown species};
\node  [box] (e) [below = 1.2cm of d] {Calculate mean plot\\level woodland traits};
\node  [round] (f) [right = 1.2cm of e] {Exclude plots with no\\tree species diversity};
\node  [round] (g) [below = 1.2cm of f] {Exclude compositional\\outlier plots};
\node  [box] (h) [left = 1.2cm of g] {Calculate reversed\\variables for SEM};
\node  [box] (i) [below = 1.2cm of h] {Transform variables to\\achieve normality};
\node  [box] (j) [right = 1.2cm of i] {Standardize variables};
\node  [diam] (k) [below = 1.2cm of j] {Analysis};
\draw [path] (a) -> (b);
\draw [path] (b) -> (c);
\draw [path] (c) -> (d);
\draw [path] (d) -> (e);
\draw [path] (e) -> (f);
\draw [path] (f) -> (g);
\draw [path] (g) -> (h);
\draw [path] (h) -> (i);
\draw [path] (i) -> (j);
\draw [path] (j) -> (k);

\end{tikzpicture}

### 2019-12-12 include/clust_summ.tex
Refactored most of the code

% Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
% Date and time: Thu, Dec 12, 2019 - 12:06:51
\begin{table}[!htbp] \centering 
  \caption{} 
  \label{clust_summ} 
\begin{tabular}{@{\extracolsep{5pt}} ccccccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
clust4 & c\_dom & c\_ind & n\_plots & n\_species\_raref & stems\_ha & agb\_ha \\ 
\hline \\[-1.8ex] 
1 & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 679 & 11(11.1) & 172(148) & 37.6(34.88) \\ 
2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 746 & 18(17.5) & 216(170) & 49.1(43.34) \\ 
3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 225 & 10(10) & 166(160) & 46.2(47.82) \\ 
4 & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\ 
\hline \\[-1.8ex] 
\end{tabular} 
\end{table}

### 2019-12-12 manuscript/manuscript.tex
Refactored most of the code

@@
 \input{include/dens_stats}
+\input{include/perc_small_agb.tex}
 
@@
 
-\begin{figure}[H]
-	\centering
-	\includegraphics[width=0.3\textwidth]{plot_loc}}\label{plot_loc}}%
-	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}.
-Points are coloured according to their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
-\end{figure}
-
 Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup.
@@
 Most plots in the dataset did not include data on stems <5 cm DBH.
-For those plots with stem measurements <5 cm DBH, those small stems only accounted for a median average of 0.42\% of the plot level AGB.
+For those plots with stem measurements <5 cm DBH, those small stems only accounted for a median average of \percsmallagb{} of the plot level AGB.
 
@@
 Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
-Temperature and precipitation seasonality were both calculated as the coefficient of variation of daily MAT and MAP, respectively, across the 18 years of available data.
+Temperature and precipitation seasonality were both calculated as the mean of the coefficient of variation of daily MAT and MAP, respectively, for each of the 18 years of available data.
 Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}.
@@
 	\includegraphics[width=0.8\textwidth]{clust_map}
-	\caption{The spatial distribution of plots according to vegetation cluster within southern Africa.}
+	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}.
+Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
 	\label{clust_map}
@@
 Latent variables are shown as circles while observed variables are shown as rectangles.
-Path coefficients are solid arrows pointing from predictor to response.
+Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
 The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
@@
 	\label{struc_mod}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{struc_model_slopes}
-	\caption{Standardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
-Due to all observed variables being standardised and centred, path coefficients are expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
-Path coefficients where the standard error does not overlap zero are considered significant effects.}
-	\label{struc_model_slopes}
 \end{figure}
@@
 	\label{full_mod}
-\end{figure}
-
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{full_model_slopes}
-	\caption{Standardised path coefficients for the interactive effects of abiotic environment and tree diversity on AGB across all plots.
-Path coefficients are $\pm$ 1 standard error.
-Path coefficients where the standard error does not overlap zero are considered significant effects.}
-	\label{full_model_slopes}
 \end{figure}

### 2019-12-12 analysis/clust_pal.R

@@ -1,2 +1,0 @@
-# Define a 5 point colour palette for vegetation clusters
-clust_pal <- c("#5cb930", "#e6c332", "#6e49cc", "#49b9da", "#da2c57")

### 2019-12-12 analysis/plot_data_collation.R

@@ -1,370 +1,0 @@
-# Creating a dataset for use in Chapter 1 - Regional analysis of the factors affecting woody AGB and productivity
-# John Godlee (johngodlee@gmail.com)
-# 2018_12_10
-
-# Preamble ----
-
-# Remove old crap
-rm(list=ls())
-#dev.off()
-
-# Set working directory to the location of the source file
-#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
-
-# Packages
-library(dplyr)
-library(tidyr)
-library(iNEXT)
-library(vegan)
-
-# Import data ----
-
-source("full_best.R")
-
-# Base SEOSAW plot summary dataset
-load("data/seosaw_plot_summary5Apr2019.Rdata")
-
-# Per stem data 
-load("data/clean_input_data.Rdata")
-
-# Aridity data 
-aridity_index <- read.csv("data/seosaw_aridity_index.csv")
-
-# Soil organic carbon
-ocdens <- read.csv("data/seosaw_ocdens.csv")
-
-# MODIS fire data 
-fire_return <- read.csv("data/fire_return.csv")
-
-# WorldClim temp.
-temp <- read.csv("data/plot_temp.csv")
-
-# WorldClim precip.
-precip <- read.csv("data/plot_precip.csv")
-
-# Clean data ----
-
-# Create large raw dataframe
-plot_data <- ssaw8$struct %>%
-  left_join(., ssaw8$structPerHa, by = c("plotcode" = "plotcode")) %>%
-  left_join(., ssaw8$plotInfoFull, by = c("plotcode" = "plotcode")) %>%
-  left_join(., ssaw8$floristicsGT5, by = c("plotcode" = "plotcode")) %>%
-  left_join(., ssaw8$soil, by = c("plotcode" = "plotcode")) %>%
-  left_join(., ssaw8$climate, by = c("plotcode" = "plotcode")) %>%
-  left_join(., ssaw8$env, by = c("plotcode" = "plotcode")) %>%
-  left_join(., ssaw8$cluster, by = c("plotcode" = "plotcode")) %>%
-  left_join(., aridity_index, by = c("plotcode" = "plotcode")) %>%
-  left_join(., ocdens, by = c("plotcode" = "plotcode")) %>%
-  left_join(., fire_return, by = c("plotcode" = "plotcode")) %>%  
-  left_join(., temp, by = c("plotcode" = "plotcode")) %>%
[…]

### 2019-12-12 analysis/plot_data_descrip.R

@@ -1,535 +1,0 @@
-# Description of cleaned data
-# John Godlee (johngodlee@gmail.com)
-# 2019_09_19
-
-# Preamble ----
-
-# Remove old crap
-rm(list=ls())
-#dev.off()
-
-# Set working directory to the location of the source file
-#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
-
-# Packages
-library(dplyr)
-library(ggplot2)
-library(tidyr)
-library(maps)
-library(rgdal)
-library(ggnewscale)
-library(gridExtra)
-
-# Import data ----
-
-# Plot data
-plot_data_final <- read.csv("data/plot_data_fil_agg_norm_std_outlier.csv")
-
-# Regional tmeperature and precipitation data for the Miombo woodlands
-t_p <- read.csv("data/region_temp_precip.csv")
-
-# Whites veg map
-white_veg <- readOGR(dsn="/Users/johngodlee/google_drive/postgrad/phd/confirmation_report/resources/whitesveg", 
-  layer="Whites vegetation")
-
-white_veg_fort <- fortify(white_veg, region = "DESCRIPTIO")
-names(white_veg_fort)
-length(unique(white_veg_fort$id))
-
-white_veg_miombo <- white_veg_fort %>%
-  filter(id %in% c("Moist-infertile savanna"),
-    lat < -2)
-
-# Original data
-load("data/seosaw_plot_summary5Apr2019.Rdata")
-
-# Biomass quantiles
-bchave_quantile_df <- read.csv("data/bchave_quantile.csv")
-
-source("clust_pal.R")
-clust_pal <- clust_pal[1:4]
-
-# Map of plot locations, compared to full SEOSAW dataset ----
-
-# Create vector of southern Africa ISO codes - find a way to mine the data for this
-s_af <- iso.expand(c("ZAF", "COD", "NAM", "ZMB", "BWA", "ZWE", "MOZ", 
-  "MWI", "AGO", "TZA", "COG", "RWA", "BDI", "UGA", "KEN")) 
-
-# Create map of country outlines
-map_africa <- borders(database = "world", regions = s_af, fill = "grey90", colour = "black")
-map_africa_fill <- borders(database = "world", regions = s_af, fill = "grey90")
-map_africa_colour <- borders(database = "world", regions = s_af, colour = "black")
-
-plot_map <- ggplot() + 
[…]

### 2019-12-12 analysis/soil_carbon.R

@@ -1,48 +1,0 @@
-# Extract soil carbon from OCDENS
-# John Godlee (johngodlee@gmail.com)
-# 2019_09_19
-
-# Preamble ----
-
-# Remove old crap
-rm(list=ls())
-#dev.off()
-
-# Set working directory to the location of the source file
-#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
-
-# Packages 
-library(rgdal)
-library(raster)
-library(dplyr)
-
-# Organic Carbon 
-ocdens <- readGDAL("data/OCDENS_M_sl2_250m_ll_crop.tif")
-
-# SEOSAW plot locations
-load("data/seosaw_plot_summary5Apr2019.Rdata")
-
-# Convert plot locations to spatial ----
-plot_xy <- data.frame(ssaw8$plotInfoFull$longitude_of_centre, 
-	ssaw8$plotInfoFull$latitude_of_centre)
-
-plot_loc_spdf <- SpatialPointsDataFrame(coords = plot_xy, 
-  data = as.data.frame(ssaw8$plotInfoFull),
-	proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
-
-# Convert aridity index to raster ----
-ocdens_raster <- raster(ocdens)
-
-# Get aridity index for each plot ----
-ocdens_seosaw <- raster::extract(ocdens_raster, plot_loc_spdf)
-
-ssaw8$plotInfoFull$ocdens <- ocdens_seosaw
-
-# Export data ----
-
-# Select relevant columns
-seosaw_ocdens <- ssaw8$plotInfoFull %>%
-	dplyr::select(plotcode, longitude_of_centre, latitude_of_centre, ocdens)
-
-# Export
-write.csv(seosaw_ocdens, file = "data/seosaw_ocdens.csv", row.names = FALSE)

### 2019-12-12 misc/outlier_pcoa_snippet.R

@@ -1,0 +1,29 @@
+# Plot Principle Coordinate Analysis (PCOA) of outliers with other points
+outlier_best <- outlier_list[[as.numeric(rownames(outlier_df[
+  which.min(abs(outlier_df$outlier_num - (nrow(stem_ab_mat) * 0.05))), 
+  ]))]]
+
+outlier_best_plot <- data.frame(outlier_best$pco.points, 
+  dist = outlier_best$neigh.dist)
+
+outlier_best_plot <- outlier_best_plot %>%
+  mutate(outlier = case_when(
+    dist >= thresh_0.05[1] ~ TRUE,
+    TRUE ~ FALSE
+  ))
+
+names(outlier_best_plot) <- c("x", "y", "dist", "outlier")
+
+pcoa <- ggplot(outlier_best_plot, aes(x = x, y = y)) + 
+  geom_point(aes(fill = outlier, alpha = outlier), 
+    colour = "black", shape = 21, size = 3) + 
+  scale_fill_manual(name = "Outlier", values = c("#2B2B2B", "#D43939")) + 
+  scale_alpha_manual(name = "Outlier", values = c(0.5, 1)) +
+  labs(x = "PCOA 1", y = "PCOA 2") + 
+  theme_classic() + 
+  theme(legend.position = c(0.9, 0.2),
+    legend.background = element_rect(fill = "#E6E6E6", colour = "black"))
+
+pdf(file = paste0("img/pcoa_outlier", ext, ".pdf"), width = 10, height = 10)
+pcoa
+dev.off()

### 2019-12-12 misc/zambia_plot_map_snippet.R

@@ -1,0 +1,19 @@
+
+# Map of plot locations in Zambia
+map_zambia <- borders(database = "world", regions = iso.expand("ZMB"), fill = "grey90", colour = "black")
+
+plot_zambia <- ggplot() + 
+  map_zambia + 
+  geom_point(data = filter(ssaw8$plotInfoFull, country == "Zambia"), 
+    aes(x = longitude_of_centre, y = latitude_of_centre), alpha = 0.6) + 
+  geom_point(data = filter(plot_data, country == "Zambia"),
+    aes(x = longitude_of_centre, y = latitude_of_centre), 
+    colour = "black", fill = "#048FD4", shape = 21, size = 2, alpha = 1) + 
+  theme_classic() +
+  coord_map() + 
+  labs(x = "Longitude", y = "Latitude") + 
+  theme_classic()
+
+pdf(file = "img/plot_loc_zambia.pdf", width = 8, height = 5)
+plot_zambia
+dev.off()

### 2019-12-12 scripts/sem.R

@@ -1,0 +1,991 @@
+# Structural Equation Models for Chapter 1 - Regional analysis of the factors affecting woody AGB and productivity
+# John Godlee (johngodlee@gmail.com)
+# 2019_09_20
+
+# Notes for interpreting SEM models
+##' Double headed arrows show covariances
+##' Single headed arrows show causal relationships
+##' Exogenous variables (i.e. independent variables) measurement error (rounded double ended arrows) of 1.00. 
+##' Exogenous variables may have covariance
+
+# Notes for constructing SEM model specifications
+##' ~ regressed by
+##' =~ Latent variable
+##' <~ Composite variable
+##' ~~ Correlated with
+
+# Preamble ----
+
+# Source files
+# source("sem_data_collation.R")
+# source("bivariate_relationships.R")
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(dplyr)
+library(ggplot2)
+library(lavaan)
+library(semPlot)
+library(semTools)
+library(ggcorrplot)
+library(lme4)
+library(stargazer)
+library(tidyr)
+library(psych)
+library(beepr)
+library(ggnewscale)
+library(maps)
+library(gganimate)
+library(MVN)
+
+source("scripts/clust_defin.R")
+
+# Import data ----
+
+# Aggregated and filtered data
+
+sem_data <- readRDS("data/plot_data_fil_agg_norm_std_outlier.rds")
+
+##' Species richness as determinant of large trees 
+##' (measured as mean of 95th percentile of height) biomass
+
+# Correlation matrix between variables ----
+
+# Unreverse some variables
+sem_data$precip_seasonality_std <- -sem_data$precip_seasonality_rev_std
+sem_data$temp_seasonality_std <- -sem_data$temp_seasonality_rev_std
+sem_data$sand_per_std <- -sem_data$sand_per_rev_std
+sem_data$mean_temp_std <- -sem_data$mean_temp_rev_std
+
+# Create a dataframe 
[…]

### 2019-12-12 scripts/soil_carbon.R

@@ -1,0 +1,48 @@
+# Extract soil carbon from OCDENS
+# John Godlee (johngodlee@gmail.com)
+# 2019_09_19
+
+# Preamble ----
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages 
+library(rgdal)
+library(raster)
+library(dplyr)
+
+# Organic Carbon 
+ocdens <- readGDAL("data/OCDENS_M_sl2_250m_ll_crop.tif")
+
+# SEOSAW plot locations
+load("data/seosaw_plot_summary5Apr2019.Rdata")
+
+# Convert plot locations to spatial ----
+plot_xy <- data.frame(ssaw8$plotInfoFull$longitude_of_centre, 
+	ssaw8$plotInfoFull$latitude_of_centre)
+
+plot_loc_spdf <- SpatialPointsDataFrame(coords = plot_xy, 
+  data = as.data.frame(ssaw8$plotInfoFull),
+	proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
+
+# Convert aridity index to raster ----
+ocdens_raster <- raster(ocdens)
+
+# Get aridity index for each plot ----
+ocdens_seosaw <- raster::extract(ocdens_raster, plot_loc_spdf)
+
+ssaw8$plotInfoFull$ocdens <- ocdens_seosaw
+
+# Export data ----
+
+# Select relevant columns
+seosaw_ocdens <- ssaw8$plotInfoFull %>%
+	dplyr::select(plotcode, longitude_of_centre, latitude_of_centre, ocdens)
+
+# Export
+saveRDS(seosaw_ocdens, "data/seosaw_ocdens.rds")

### 2019-12-12 scripts/standardise.R

@@ -1,0 +1,353 @@
+# Bivariate relationships between variables, transform, normalise and standardize the data
+# John Godlee (johngodlee@gmail.com)
+# 2018_12_10
+
+# Preamble ----
+
+# Source files
+# source("sem_data_collation.R")
+
+# Remove old crap
+rm(list=ls())
+#dev.off()
+
+# Set working directory to the location of the source file
+#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
+
+# Packages
+library(dplyr)  #
+library(tidyr)  #
+library(ggplot2)  #
+library(gridExtra)  #
+library(lme4)  #
+library(MuMIn)  #
+
+source("scripts/clust_pal.R")
+
+# Import data ----
+
+plot_data <- readRDS("data/plot_data_fil_agg.rds")
+
+facet_levels <- c(
+  "total_precip",
+  "precip_seasonality",
+  "mean_temp",
+  "temp_seasonality",
+  "agb_ha",
+  "agb_ha_log", 
+  "cec",
+  "sand_per",
+  "ocdens",
+  "cov_dbh",
+  "cov_height",
+  "stems_ha",
+  "stems_ha_log",
+  "shannon_equit",
+  "n_species_raref",
+  "n_species_raref_log"
+  )
+
+facet_labels <- c(  
+  expression("MAP" ~ (mm ~ y^-1)),
+  expression("Precip." ~ "seasonality"),
+  expression("MAT" ~ (degree * C)), 
+  expression("Temp." ~ "seasonality"),
+  expression("AGB" ~ (t ~ ha^-1)),
+  expression("log(AGB)" ~ (t ~ ha^-1)),
+  expression("CEC"),
+  expression("Sand" ~ "%"), 
+  expression("Org." ~ "C" ~ "%"),
+  expression("Coef." ~ "var." ~ "DBH"),
+  expression("Coef." ~ "var." ~ "height"),
+  expression("Stem" ~ "density" ~ ">10" ~ cm ~ (n ~ ha^-1)),
+  expression("log(Stem" ~ "density)" ~ ">10" ~ cm ~ (n ~ ha^-1)),
+  expression("Shannon" ~ "equit." ~ (E[H])),
+  expression("Extrap." ~ "species" ~ "rich."),
+  expression("log(Extrap." ~ "species" ~ "rich.)"))
+
+facets <- list(facet_levels = facet_levels, facet_labels = facet_labels)
+
[…]

### 2019-12-12 scripts/temp_precip.R

@@ -1,0 +1,125 @@
+# Generate a dataset of precipitation and temperature means across the Miombo region
+# John Godlee (johngodlee@gmail.com)
+# 2019_10_02
+
+rm(list = ls())
+
+# Packages
+library(raster)
+library(rgdal)
+
+# Get Miombo shape
+white_veg <- readOGR(dsn="data/whitesveg", 
+  layer="Whites vegetation")
+
+white_veg_miombo <- white_veg[!is.na(white_veg@data$DESCRIPTIO) & white_veg@data$DESCRIPTIO == "Moist-infertile savanna" , ]
+
+white_veg_miombo_south <- crop(white_veg_miombo, extent(10, 40, -30, 0))
+
+load("data/seosaw_plot_summary5Apr2019.Rdata")
+
+plot_loc <- data.frame(plotcode = ssaw8$plotInfoFull$plotcode,
+  lon = ssaw8$plotInfoFull$longitude_of_centre,
+  lat = ssaw8$plotInfoFull$latitude_of_centre)
+
+plot_loc_spdf <- SpatialPointsDataFrame(coords = data.frame(plot_loc$lon, plot_loc$lat), data = plot_loc)
+
+# Temperature ----
+
+# Get list of files
+rastlist_t <- list.files(path = "data/wc2.0_30s_tavg", 
+  pattern = '.tif$', 
+  all.files = TRUE, 
+  full.names = TRUE)
+
+# Import
+allrasters_t <- lapply(rastlist_t, raster)
+
+# Crop to plot extent
+allrasters_t_crop <- lapply(allrasters_t, function(x){
+  crop(x, plot_loc_spdf)
+})
+
+# Stack
+allrasters_t_crop_stack <- raster::stack(allrasters_t_crop)
+
+# Take mean of all in stack
+allrasters_t_mean <- calc(allrasters_t_crop_stack, mean, na.rm = TRUE)
+
+# Crop to miombo outline
+allrasters_t_mean_crop_mask <- mask(allrasters_t_mean, white_veg_miombo_south)
+
+# Extract all values
+t_vals <- values(allrasters_t_mean_crop_mask)
+
+# Take CV of all values
+allrasters_t_cv <- calc(allrasters_t_crop_stack, fun = function(x){
+  sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE) * 100
+  })
+
+# Crop to miombo outline
[…]

### 2019-12-16 include/clust_summ.tex
Updated manuscript

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Dec 12, 2019 - 12:06:51
+% Date and time: Mon, Dec 16, 2019 - 11:33:55
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-1 & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 679 & 11(11.1) & 172(148) & 37.6(34.88) \\ 
-2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 746 & 18(17.5) & 216(170) & 49.1(43.34) \\ 
-3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 225 & 10(10) & 166(160) & 46.2(47.82) \\ 
+1 & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 687 & 11(11.2) & 170(145.3) & 37(34.68) \\ 
+2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 757 & 18(17.6) & 215(171.7) & 48.8(43.7) \\ 
+3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 226 & 10(10) & 165(157.5) & 46(47.81) \\ 
 4 & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\ 

### 2019-12-16 manuscript/include/clust_summ.tex
Updated manuscript

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Wed, Nov 20, 2019 - 20:22:20
+% Date and time: Mon, Dec 16, 2019 - 11:33:55
 \begin{table}[!htbp] \centering 
@@
 Dominant species were identified by choosing the species with the largest AGB contribution within each cluster.
-Numeric values of species richness, stems ha\textsuperscript{-1} and AGB are medians and interquartile ranges.}
+Numeric values of species richness, stems ha\textsuperscript{-1} and AGB are medians and interquartile ranges (75th percentile - 25th percentile).}
   \label{clust_summ} 
@@
 1 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
-\\ \textit{Brachystegia spiciformis} \\ \textit{Baikeaea plurijuga} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Burkea africana} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 734 & 6(5) & 276(208) & 35.4(31.41) \\
+\\ \textit{Brachystegia spiciformis} \\ \textit{Baikeaea plurijuga} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Burkea africana} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 687 & 11(11.2) & 170(145.3) & 37(34.68) \\
 \hline
@@
 \\ \textit{Brachystegia }spp.
-\\ \textit{Isoberlinia angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Brachystegia longifolia} \end{tabular} & 728 & 8(5) & 277(196.9) & 46.9(44.96) \\
+\\ \textit{Isoberlinia angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Brachystegia longifolia} \end{tabular} & 757 & 18(17.6) & 215(171.7) & 48.8(43.7) \\
 \hline
 3 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia }spp.
-\\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 299 & 9(8) & 410(427.7) & 46.6(43.51) \\
+\\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 226 & 10(10) & 165(157.5) & 46(47.81) \\
 \hline
 4 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Combretum }spp.
-\end{tabular} & 102 & 4(4) & 322(228.8) & 43.8(36.49) \\
+\end{tabular} & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\
 \hline

### 2019-12-16 manuscript/include/data_clean_flow.tex
Updated manuscript

--- 2019-12-12 data_clean_flow/data_clean_flow.tex
@@
 \node  [box] (b) [right = 1.2cm of a] {Aggregate values of Zambian\\forestry commission plots};
-\node  [round] (c) [below = 1.2cm of b] {Exclude plots where minimum\\measured tree size \textgreater{}5 cm};
-\node  [round] (d) [left = 1.2cm of c] {Exclude stems:\\- \textless{}5 cm DBH\\- Dead\\- Unknown species};
+\node  [round] (c) [below = 1.2cm of b] {Exclude plots where minimum\\measured tree size \textgreater{}10 cm};
+\node  [round] (d) [left = 1.2cm of c] {Exclude stems:\\- \textless{}10 cm DBH\\- Dead\\- Unknown species};
 \node  [box] (e) [below = 1.2cm of d] {Calculate mean plot\\level woodland traits};

### 2019-12-16 manuscript/include/struc_model_fit_clust_stats.tex
Part way through editing
A day of working with clust4
Draft sent to supervisor
Updated models
Updated manuscript

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Fri, Nov 08, 2019 - 17:15:20
+% Date and time: Mon, Dec 16, 2019 - 16:41:44
 \begin{table}[!htbp] \centering 
@@
   \label{struc_model_fit_clust_stats} 
-\begin{tabular}{@{\extracolsep{0pt}} ccccccccccc} 
+\begin{tabular}{@{\extracolsep{0pt}} ccccccccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-{Cluster} & {Params.} & {n} & {$\chi^{2}$} & {DoF} & {CFI} & {TLI} & {LogLik} & {AIC} & {RMSEA} & {SRMR} \\
+{Cluster} & {n} & {$\chi^{2}$} & {DoF} & {CFI} & {TLI} & {LogLik} & {RMSEA} & {$R^{2}$ AGB} \\
 \hline \\[-1.8ex] 
-C1 & 12 & 420 & 10.430 & 3 & 0.987 & 0.958 & -2630.700 & 5285.300 & 0.080 & 0.029 \\ 
-C2 & 12 & 671 & 50.320 & 3 & 0.942 & 0.806 & -3730.200 & 7484.300 & 0.150 & 0.052 \\ 
-C3 & 12 & 105 & 24.970 & 3 & 0.880 & 0.601 & -619.100 & 1262.300 & 0.260 & 0.142 \\ 
-C4 & 12 & 46 & 2.710 & 3 & 1 & 1.008 & -257.900 & 539.900 & 0 & 0.025 \\ 
-C5 & 12 & 84 & 5.400 & 3 & 0.981 & 0.936 & -529.500 & 1083 & 0.100 & 0.085 \\ 
-All & 12 & 1326 & 62.210 & 3 & 0.968 & 0.895 & -8110.500 & 16245 & 0.120 & 0.036 \\ 
+Marginal miombo & 525 & 47.480 & 6 & 0.964 & 0.910 & -3730.100 & 0.110 & 0.730 \\ 
+Core miombo & 668 & 59.440 & 6 & 0.958 & 0.895 & -4219 & 0.100 & 0.680 \\ 
+Baikiaea & 47 & 5.860 & 6 & 0.998 & 0.994 & -323.100 & 0.030 & 0.720 \\ 
+Mopane & 84 & 9.420 & 6 & 0.971 & 0.927 & -588.900 & 0.080 & 0.450 \\ 
+All & 1324 & 82.020 & 6 & 0.973 & 0.932 & -9122.800 & 0.090 & 0.700 \\ 
 \hline \\[-1.8ex] 

### 2019-12-16 manuscript/manuscript.tex
Updated manuscript

@@
 The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}.
-Over the past decade, many observational studies of the BEFR have been conducted, mostly in tropical and temperate forests, and grasslands \citep{Chen2011}.
+Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
 These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
@@
 Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}.
-This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and even to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
+This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
 The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.
 
-%The relationship between biodiversity and ecosystem functionality has informed conservation policy which aims to maintain and strengthen biodiversity under the assumption that this strengthens ecosystem resilience to stressors such as anthropogenic climate change and ecosystem service use by humans \citep{Thompson2011}.
-It is probable however, that the strength and relative importance of the BEFR varies with ecosystem studied and environmental conditions, with many ecosystems remaining understudied in the BEFR literature.
-
 The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}.
 In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, Poorter2015, Sande2017a}.
@@
 Disturbance may weaken the role of competition in determining local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979,Grace1990}.
-Instead, stress tolerance and the functional contribution of mor abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}.
+Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}.
 Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.
 
-Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{plot_loc}).
+Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{clust_map}).
 The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
-Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species \citep{Frost1996}.
+Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species.
 Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}.
@@
 
-Southern African woodlands are also relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
+Southern African woodlands are relied upon heavily for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
 This has raised interest in how biodiversity influences ecosystem function in these ecosystems.
-Resource extraction by humans is directly influencing biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}.
-Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
+Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}.
+Climate change also indirectly affects the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
 While rapid biodiversity change is being observed in southern African woodlands \citep{}, research into the relationship between biodiversity and ecosystem functionality remains scarce.
@@
 
-In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}, so it is important to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on biomass in analyses at large spatial scales in woodlands.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}, while other factors such as herbivory, fire regime and soil properties also contribute at a local scale to reduce woody cover below this maximum.
-However, this study did not consider the role of species diversity in the complex interactions between environment and tree cover.
+In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}.
+It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
+Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.
+In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
+They suggest that in dry ecosystems, in contrast to the wet tropical systems in which much biodiversity-ecosystem function research has been conducted, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function.
+In dry woodland ecosystems stress tolerance rather than competitiveness may be the mechanism by which a tree species' success is measured.
 
-Environmental conditions may indirectly moderate the observed relationship between biodiversity and ecosystem function \citep{Vila2005}.
-\citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in Panamanian tropical forests.
-Dry season length may therefore indirectly affect ecosystem level productivity in forest ecosystems and thus woody biomass through selection effects, promoting the growth of a certain highly productive species only under specific environmental conditions.
-In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggested variation in the balance between competitive and facilitative effects along the aridity gradient as the driver of this moderation relationship, an example of the Stress Gradient Hypothesis \citep{Dohn2013}.
-Water availability imposes a physiological limit on growth rate, which interacts with mortality due to stochastic processes such as fire and herbivory to limit maximum potential woody biomass \citep{Sankaran2005}.
+%Environmental conditions may also indirectly moderate the observed relationship between biodiversity and ecosystem function \citep{Vila2005}.
 
@@
 \citep{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity.
-They suggest this effect is caused by an increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity.
+They suggest this effect is caused by increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity.
 We expect that this effect will occur in southern African woodlands as well.
@@
 
-High levels of disturbance in the form of seasonal fires and herbivory from large mammals may moderate the strength of observable BEFRs in southern African woodlands \citep{Staver2009, Bond2005}.
+%High levels of disturbance in the form of seasonal fires and herbivory from large mammals may moderate the strength of observable BEFRs in southern African woodlands \citep{Staver2009, Bond2005}.
 Fire disturbance in forests has been linked to abundance dependent mortality among smaller stems \citep{Roques2001}.
@@
 
-In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (Figure \autoref{plot_loc}, \autoref{temp_precip_hull}).
+In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (\autoref{clust_map}).
 We investigated the relationship between aboveground woody biomass and tree biodiversity.
@@
 \begin{enumerate}
-	\item{More arid plots and plots with less fertile soil will show a stronger positive effect of tree species diversity on above-ground woody biomass.} 
 	\item{Precipitation and soil fertility will indirectly positively effect AGB via an increase in tree species diversity.}
@@
 
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[plot_loc]{Figure~\ref*{plot_loc}}, \citealt{White1987}).
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[clust_map]{Figure~\ref*{clust_map}}, \citealt{White1987}).
 The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
-The study sites span the core climate space of the region as a whole (\hyperref[temp_precip_hull]{Figure~\ref*{temp_precip_hull}}).
+The study sites span the core climate space of the region as a whole.
 The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
@@
 Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
-Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014} (\autoref{data_clean_flow}).
+Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
 
@@
 We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
-We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thusremoving \noutliers{} plots.
+We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thus removing \noutliers{} plots.
 
@@
 We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
-Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
-Most plots in the dataset did not include data on stems <5 cm DBH.
-For those plots with stem measurements <5 cm DBH, those small stems only accounted for a median average of \percsmallagb{} of the plot level AGB.
+Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
+Many plots in the dataset did not include data on stems <10 cm DBH.
+For those plots with stem measurements <10 cm DBH, those small stems only accounted for a median average of \percsmallagb{} of the plot level AGB.
 
-All stems >5 cm DBH were measured within each plot resulting in a total of 160,076 stems with measurements.
+All stems >10 cm DBH were measured within each plot resulting in a total of \nstems{} stems with measurements.
 A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
@@
 \centering
-	\includegraphics[width=0.8\textwidth]{clust_map}
+	\includegraphics[width=1.4\textwidth]{clust_map}
 	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}.
@@
 Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset.
-For example, moisture availability is expected to affect AGB \citep{Saito2014, Campbell1996}, but moisture availability itself is determined by the interaction of multiple observable variables over the time scales relevant to tree lifetime growth: precipitation, the seasonality of that precipitation, temperature which affects the rate of evapotranspiration and soil water retention.
-Independently of total precipitation, precipitation seasonality determines whether water arrives uniformly over a given time period or as a few high volume floods, the latter leading to much water being lost before it can be used for plant growth.
 Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and woody species diversity, which could otherwise increase the chances of Type I error and wrongly attributing inference due to covariance of explanatory variables, if testing for significant effects of biodiversity on AGB using conventional regression analyses \citep{Nachtigall2003}.
@@
 	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, tree stem density, and AGB.
-Hypothesised paths of causation are depicted as arrows from predictor to response.}
+Hypothesised paths of causation are depicted as arrows from predictor to response.
+Correlations are depicted as curved dotted arrows.}
 	\label{con_mod}
@@
 Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
-Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect strength, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
+Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
 Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
@@
 We therefore allowed factor loadings to be estimated by the SEM itself.
-We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, finding no significant difference between model specifications.
+We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, while ensuring no factor loadings were in excess of 1.
+We found no significant difference between model specifications.
 Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.
@@
 We used each of these datasets to fit the model including only tree species and structural diversity latent variables to predict AGB.
+We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model.
 We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled dataset.
@@
 We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB.
+Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation clusters were narrow in their climate space leading to a lack of variance particularly in moisture availability.
 
-We fitted separate moderation models to investigate whether there was an interaction effect whereby the strength of the relationship between species diversity and AGB was influenced by moisture availability and soil fertility.
+% We fitted separate moderation models to investigate whether there was an interaction effect whereby the strength of the relationship between species diversity and AGB was influenced by moisture availability and soil fertility.
 The \verb|lavaan| R package does not natively support moderation of latent variables in its model specification.
@@
 	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlogram of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($\rho$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
+	\caption{Correlogram of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
 Variables in bold and underlined on the axis labels were later reversed for SEMs to maintain positive correlations for all observed variables within each latent variable.
@@
 Stem density had the strongest correlation with AGB of all variables (\ccib{}).
-Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except CEC.
+Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except CEC and MAT.
 
@@
 As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a negative correlation between the seasonality of precipitation and AGB (\ccmcb{}.
-MAT and the seasonality of temperature negatively correlated weakly with AGB (MAT: \cctb{}; MAT seasonality: \cctcb{}).
+MAT and temperature seasonality (TS) negatively correlated weakly with AGB (MAT: \cctb{}; TS\cctcb{}).
 As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic carbon and AGB (\ccob{}).
@@
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a positive effect on AGB, both directly and indirectly via stem density and structural diversity (\autoref{struc_mod}, \autoref{struc_model_slopes}).
-Tree species diversity had a strong positive effect on stem density.
+In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity suprisingly had a small negative direct effect on AGB, while positive effects were observed via structural diversity (\autoref{struc_mod}).
+Tree species diversity had a positive correlation with stem density.
 Model fit was good with high factor loadings for all observed variables, all path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}).
-The R-squared of AGB was \strucrsq{}.
-The strongest effect of tree species diversity on AGB was via stem density (\strucsib{})
+The R\textsuperscript{2} of AGB was \strucrsq{}.
+The strongest direct effect on AGB was via stem density (\strucsib{}).
 
@@
 	\input{include/path_diagram_struc}
-	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all five vegetation clusters.
-Latent variables are shown as circles while observed variables are shown as rectangles.
+	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all vegetation clusters.
+Latent variables are circles while observed variables are rectangles.
 Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
-The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
+The observed variables which inform the latent variables are connected by dotted arrows, observed variables with loading set to 1 are connected by dashed arrows.
+Correlations between variables are depicted as dotted curved arrows.
 Measurement errors of exogenous variables are omitted for clarity.}
@@
 
-When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation clusters the strengths of unstandardised path coefficients varied but relationships between tree diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
-All models exhibited adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though wide confidence intervals around the unstandardised path coefficients, particularly for Clusters 3 and 4 indicate an issue with model prediction due to low sample size for these clusters.
-$\chi^{2}$ statistics were high for some vegetation clusters, but this appears to be highly correlated with sample size in our data \citep{}.
+When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation clusters the strengths of unstandardised path coefficients varied.
+The direct effect of tree species diversity on AGB was positive in Baikiaea and Mopane, but negative in Marginal and Core miombo (\autoref{struc_model_slopes_all}).
+Relationships between structural diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
+The total effect of species diversity on AGB remained strongly positive.
+All vegetation types except Mopane exhibited a positive effect of species diversity on structural diversity.
+All models had adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though confidence intervals around the unstandardised path coefficients were wide particularly for Mopane and Baikiaea indicate.
+$\chi^{2}$ statistics were high for some vegetation clusters, but this appears to be highly correlated with sample size for each vegetation type \citep{}.
 
-Cluster 2, which contained vegetation dominated by \textit{Julbernadia paniculata} showed no significant effect of tree species diversity on AGB (\strucbsb{}), in contrast to the other vegetation clusters, and had a smaller positive effect of structural diversity on AGB (\strucbhb{}) (\autoref{struc_model_slopes_all}).
-
-The strongest total effect of tree species diversity on AGB was in Cluster 3 (\struccsb{}), which was species rich but highly variable in species diversity compared to other vegetation clusters (\autoref{clust_summ}).
-The greatest effect of structural diversity on AGB was found in Clusters 3 and 4.
-The R-squared of AGB was highest in cluster 3 (R\textsuperscript{2} = \struccrsq{}) and lowest in cluster 2 (R\textsuperscript{2} = \strucbrsq{}).
+The strongest total effect of tree species diversity on AGB was in Baikiaea woodland (\struccsb{}), which was species rich but highly variable in species diversity compared to other vegetation clusters (\autoref{clust_summ}).
+The R\textsuperscript{2} of AGB was highest in Baikiaea (R\textsuperscript{2} = \struccrsq{}) and lowest in the Core miombo (R\textsuperscript{2} = \strucbrsq{}).
 
@@
 Each dataset consisted of approximately \subp{} plots with overlap of plot identity between subsampled datasets.
-The same SEM specification was used as above (\autoref{struc_mod}) except that paths via stem density were removed.
 \autoref{sem_struc_stems_ha} shows a positive effect of tree species diversity on AGB as stem density increases.
-There appears to be a minimum stem density threshold at \textapprox{}295 stems ha\textsuperscript{-1} below which there appears to be no effect of tree diversity on biomass.
-The effect of structural diversity on AGB appears to decrease with increasing stem density, though the relationship is less clear.
+There appears to be a minimum stem density threshold at \textapprox{}180 stems ha\textsuperscript{-1} below which there appears to be only a baseline effect of tree diversity on biomass.
+The effect of structural diversity on AGB appears to remain constant with increasing stem density.
+The indirect effect of species diversity on AGB via structural diversity climbs slightly and stem density increases.
 
@@
 A model incorporating the latent variables of moisture availability and soil fertility showed that the effect of diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}).
-Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}) (\autoref{full_model_slopes}).
-MAP and temperature seasonality had the greatest contributions to the latent variable of moisture availability.
-Model fit was acceptable: CFI = \fmcfi{}, TLI = \fmtli{}, and RMSEA = \fmrmsea{}.
-The R\textsuperscript{2} coefficient of determination on AGB was \fmrsq{}.
-
-Moisture availability and soil fertility also had negligible direct effects on stem density, but both had a positive total effect on stem density via species diversity, where species diversity itself had a strong positive effect on stem density (\rgid{}).
-As species diversity increases, stem density also increases.
+Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}).
+MAP and temperature seasonality (TS) had the greatest contributions to the latent variable of moisture availability.
+Moisture availability and soil fertility also had negligible direct effects on stem density.
+Model fit was acceptable: CFI = \fmcfi{}, TLI = \fmtli{}, and RMSEA = \fmrmsea{}, R\textsuperscript{2} of AGB = \fmrsq{}.
 
@@
 The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
+Correlations between variables are depicted as dotted curved arrows.
 Measurement errors of exogenous variables are omitted for clarity.}
@@
 
-Vegetation cluster level models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation clusters were narrow in their climate space.
-leading to a lack of variance particularly in moisture availability.
-
-\subsection{Moderation effects of moisture availability and soil}
-
-To address the hypothesis (H\textsubscript{2}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
-
-Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}).
+% \subsection{Moderation effects of moisture availability and soil}
+% 
+% To address the hypothesis (H\textsubscript{1}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
+% 
+% Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}).
 The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was marginally significant (\moders{}).
@@
 See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.
-
-\begin{figure}[H]
-\centering
-	\subfloat[]{
-	\includegraphics[width=0.48\textwidth]{mois_int}
-	}
-	\subfloat[]{
-	\includegraphics[width=0.48\textwidth]{soil_int}
-	}
-	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b).
+% 
+% 
+% \begin{figure}[H]
+% \centering
+% 	\subfloat[]{
+% 	\includegraphics[width=0.48\textwidth]{mois_int}
+% 	}
+% 	\subfloat[]{
+% 	\includegraphics[width=0.48\textwidth]{soil_int}
+% 	}
+% 	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b).
 All variables are standardised and centred to a mean of zero and a standard deviation of 1.
@@
 Moisture availability, tree species diversity and soil fertility are latent variables comprised of the same observed variable loadings.}
-	\label{int_plots}
-\end{figure}
+% 	\label{int_plots}
+% \end{figure}
 
@@
 
-%1. Plots with a higher tree species diversity will maintain higher above-ground woody biomass stocks.
-%2. More arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above-ground woody biomass.
-%3. There will be a positive effect of precipitation and soil fertility on AGB and that part of this positive effect will exist as a mediating effect through species diversity.
-%4. Structural characteristics of the woodland will interact with species diversity to provide an indirect path of influence between species composition and biomass stocks.
-%5. The observed effect of diversity on AGB will increase in strength as stem density increases owing to an increased importance of niche complementarity as competition increases.
+% Precipitation and soil fertility will indirectly positively effect AGB via an increase in tree species diversity.
+% Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and AGB.
+% The effect size of species diversity on AGB will increase with stem density, due to an increased importance of niche complementarity as competition increases.
 
-In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, [d] stem density and their interactions on aboveground woody biomass (AGB) across southern African woodlands, using a plot network of \nplots{}.
-Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB (H\textsubscript{1}), and an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{4}) and via stem density (H\textsubscript{5}).
-Tree diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
-The strongest effect on AGB was that of stem density, which itself was highly dependent on species diversity.
-Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{3}).
-In separate multiple regressions however, we did find weak indirect moderation effects of both soil fertility and moisture availability on the strength of the relationship between species diversity and AGB (H\textsubscript{2}).
-We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between species diversity and AGB, with true Miombo woodland representing an outlier in our models (vegetation type C2).
+In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, [d] stem density and their interactions on aboveground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} plots.
+Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB, and an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{1}) and via stem density (H\textsubscript{3}).
+Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
+The strongest effect on AGB was that of stem density.
+Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{2}).
+We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between tree species diversity and AGB, with Baikiaea woodland representing an outlier in our models.
 
-\todo{Why is Cluster 2 different?}
-
-\todo{C3 had the strongest effect of species div on AGB}
+Miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB.
 
@@
 
-The latent variable of tree species diversity had a strong positive effect on woody above-ground biomass (AGB) in the environment+diversity SEM model (\autoref{full_mod}).
-Within the savanna woodlands of southern Africa we therefore find support for our hypothesis (H\textsubscript{1}), that higher tree species richness causes higher woody AGB.
+% I would just write that you had a consistent positive effect across all models, as that is your general result with regard to this question.
+
+We found a consistent positive effect of tree species diversity on AGB across all models in this study.
+Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness causes higher woody AGB.
 This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}.
+In addition however, our study provides a novel dissection of the mechanisms underlying this relationship and their interaction with resource availability in southern African woodlands as an understudied biome.
 
-Most of the total species diversity effect was through the indirect effect via stem density.
-This suggests that within southern African woodlands a higher species diversity allows for a greater density of tree stems and this leads to an increase in total AGB.
-While reverse causation is plausible, with increased stem density causing higher species richness through an increased probability of encountering new species, we accounted for this variation via the extrapolation of species richness according to stem density (sampling effort), meaning that stem density should have little causal effect on species diversity in our models.
-This suggests that the observed direction of causation in our model is real.
+Much of the total variation in AGB was caused by stem density.
+Stem density also correlated with species diversity in our SEMs.
+It is possible that within southern African woodlands a higher species diversity allows for a greater density of tree stems, leading to an increase in total AGB.
+Reverse causation is plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
+In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stem density exists.
 We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
@@
 This is supported by the moderately strong positive effect of tree species diversity on AGB via structural diversity.
+
+% There are counter-examples, e.g. Johnson et al. 2016. GCB (from Leeds group, Tim Baker senior author), but feel free to leave text as is for now.
 
@@
 
-We found evidence that species diversity led to an increase in AGB via tree structural diversity, we can therefore accept our hypothesis (H\textsubscript{4}).
+We found evidence that tree species diversity led to an increase in AGB via tree structural diversity and we can therefore accept our hypothesis (H\textsubscript{2}).
 A higher tree species diversity allows for a greater diversity of tree functional forms within a plot and this may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
@@
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
-In highly disturbed woodland plots, we suggest that disturbance history may be an important unmeasured abiotic factor which may account for a high proportion of the variation in AGB and in tree species diversity.
-In highly disturbed forests, maturity in the sense of classical forest succession is never reached, and a simple open canopy is maintained.
+In frequently disturbed woodlands such as those studied here however, a tree canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained.
 Unfortunately we did not have access to a dataset of high enough spatial and temporal resolution to adequately quantify the effects of disturbance history.
@@
 Surprisingly, moisture availability and soil fertility had only small effects on AGB compared to that of tree species diversity.
-We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater resource partitioning between individual trees \citep{} and a greater stem density per unit area \citep{}.
+We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater resource partitioning among individual trees \citep{} and a greater stem density per unit area \citep{}.
 
@@
 
-In a separate set of linear multiple regressions, we found a weak positive interaction effect of moisture availability on the relationship between tree species diversity and AGB.
+% In a separate set of linear multiple regressions, we found a weak positive interaction effect of moisture availability on the relationship between tree species diversity and AGB.
 As moisture availability increased, the relationship between tree species diversity and AGB became stronger.
@@
 They attributed this to an increase in selection effects which allowed dominance of stress tolerant species which were more likely to occur in high diversity assemblages.
-\todo{Why were our results different?}
 
@@
 
+% See thoughts above for C3 results.
+C2 should be discussed even if we cannot give a good reason.
+It is very intriguing.
+The results for C4 also surprise me, as I would have expected mopane to be too species poor to be able to show a result.
+You can discuss how many mopane stands are monospecific and even thought such things as 'cathedral mopane' exist, where biomass can be quite high, it seems that adding even a few species to mopane can increase its total AGB (of the stands).
+That is cool and shows robustness of results.
+Basically, the fact that C3 and C4 show significant results means that your results are robust across a range of species richnesses (not just do to patterns at low, 1-6, or high, 6-12, species richness).
+I still have a great sense of what C1 is, so with withhold commentary on that for the moment.
+If it is 'disturbed miombo', then maybe something can be said about the contrast between results for C1 and C2 and the role that disturbance may play.
+
 \todo{I have no idea what to write here just yet}
+
+\todo{Baikiaea had the strongest total effect of species div on AGB. }
 
@@
 
-We found a non linear saturating effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
-The effect of tree species diversity on AGB peaked at \textapprox{}700 stems ha\textsuperscript{-1}.
-At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness might not make any difference to plot level AGB.
+We found a non linear effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
+The total effect of tree species diversity on AGB increases with increasing stem density.
+At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness might not make any difference to plot level AGB, accounting for the low and constant effect of tree species diversity on AGB below \textapprox{}180 stems ha\textsuperscript{-1}.
 
-It is possible that at very high stem density there are other factors which become more important than tree species diversity in affecting AGB.
-For example, very high stem density plots in our dataset often had tree species commonly found in thicket vegetation, which is highly structured through disturbance by fire.
-There may also be an effect whereby the high stem density plots have many small stems, meaning that AGB cannot be high.
+At high stem densities, where the woodland approaches a more forest-like canopy structure an increase in niche complementarity in canopy occupation may allow a higher AGB per unit area.
+There may be an effect whereby high stem density plots have many small stems which do not hold much woody biomass, but we did not observe this in our study, possibly because we only considered stems >10 cm DBH.
 An abundance of small stems may prevent the growth of large trees which hold the majority of the AGB in a plot.
-The peak in species diversity effect may occur at a given stem density due to a balance point between tree species diversity effects and the offsetting effects of stem crowding which preclude high biomass large trees.
 
-\todo{Maybe at very low stem density, no niche complementarity}
-
-\todo{Decreasing effect of struct. on AGB with stem density}
+% Indeed, we need to address this point, as it is surprising.
+If we were to go to the wet forest literature, we would say that high richness dilutes specialist natural enemies (and many of natural enemies in wet forests do seem to specialise on specific tree species).
+Are natural enemies in woodlands species specialists?
+I have no idea.
+In tropical wet forests, insect herbivores remove way more biomass than vertebrate herbivores, and insects are more likeliy to specialise on a specific tree species than an herbivore.
+But, I had the vague idea that vertebrates (namely mammals) remove more plant biomass than insects in woodlands, but I could be completely wrong.
+In any case, that could be a mechanism that explains species richness effect holding strong, even as the effect of structural diversity disappears.
 
@@
 \centering
-	\input{include/data_clean_flow}
+	\input{include/data_clean_flow.tex}
 	\caption{Flow diagram of the data filtering and cleaning process prior to analysis. Rounded boxes indicate filtering events while regular boxes indicate calculation events.}
@@
 
-\input{include/corr_ci_tab}
+\input{include/corr_ci_tab.tex}
 

### 2019-12-16 sem_path_diag/path_diagram_full.tex
Updated manuscript

--- 2019-11-15 analysis/output/include/path_diagram_full.tex
@@
 	path/.style={->, thick, >=stealth'},
+	corr/.style={<->, thick, bend left=15, dotted, >=stealth'},
 	loading/.style={->, thick, dotted, >=stealth'},
@@
 \node [observed] (mp) [left = 0.8in of momid] {MAP};
-\node [observed] (mpc) [left = 0.1in of momid] {MAP\\CoV};
+\node [observed] (mpc) [left = 0.1in of momid] {PS};
 \node [observed] (mt) [right = 0.1in of momid] {MAT};
-\node [observed] (mtc) [right = 0.8in of momid] {MAT\\CoV};
+\node [observed] (mtc) [right = 0.8in of momid] {TS};
 
@@
 \draw [path] (s) to node {\pcfsi{}} (i);
-\draw [path] (d) to node {\pcfdi{}} (i);
+\draw [corr] (d) to node[above=0.05in, pos=0.5] {\pcfdi{}} (i);
 \draw [path] (m) to node {\pcfmb{}} (b);

### 2019-12-16 sem_path_diag/path_diagram_struc.tex
Updated manuscript

--- 2019-11-15 manuscript/include/path_diagram_struc.tex
@@
 	path/.style={->, thick, >=stealth'},
+	corr/.style={<->, thick, bend right=15, dotted, >=stealth'},
 	loading/.style={->, thick, dotted, >=stealth'},
@@
 \draw [path] (h) to node {\pcshb{}} (b);
-\draw [path] (d) to node {\pcsdi{}} (i);
+\draw [corr] (d) to node {\pcsdi{}} (i);
 \draw [path] (i) to node {\pcsib{}} (b);

### 2019-12-16 build.sh

@@ -1,4 +1,5 @@
 #!/bin/bash
+
 {
 
 IMG="manuscript/img/"
@@ -7,16 +8,15 @@ INC="manuscript/include/"
 # Remove intermediate objects
 
 # Run data compilation
-Rscript scripts/temp_precip.R
-Rscript scripts/soil_carbon.R
-Rscript scripts/data_clean.R
-Rscript scripts/standardise.R
-Rscript scripts/composition.R
-Rscript scripts/descrip.R
+# Rscript scripts/temp_precip.R
+# Rscript scripts/soil_carbon.R
+# Rscript scripts/data_clean.R
+# Rscript scripts/standardise.R
+# Rscript scripts/composition.R
+# Rscript scripts/descrip.R
 Rscript scripts/sem.R
 
 # Transfer images to manuscript
-
 cp img/clust_map.pdf $IMG
 cp img/corr_mat.pdf $IMG
 cp img/struc_model_slopes_all.pdf $IMG
@@ -25,7 +25,6 @@ cp img/mois_int.pdf $IMG
 cp img/soil_int.pdf $IMG
 
 # Transfer snippets to manuscript
-
 cp include/n_plots.tex $INC
 cp include/n_outliers.tex $INC
 cp include/hull_cover.tex $INC
@@ -37,23 +36,20 @@ cp include/dens_stats.tex $INC
 cp include/perc_small_agb.tex $INC
 
 # Transfer path diagrams to manuscript 
+cp sem_path_diag/path_diagram_concept.tex $INC
+cp sem_path_diag/path_diagram_struc.tex $INC
+cp sem_path_diag/path_diagram_full.tex $INC
 
-cp include/path_diagram_concept.tex $INC
-cp include/path_diagram_struc.tex $INC
-cp include/path_diagram_full.tex $INC
-
-cp include/data_clean_flow.tex $INC
+cp data_clean_flow/data_clean_flow.tex $INC
 
 # Transfer tables to manuscript
-
 cp include/struc_model_fit_clust_stats.tex $INC
 cp include/clust_summ.tex $INC
 cp include/corr_ci_tab.tex $INC
 cp include/mois_div_int_mod.tex $INC
-cp include/soil.tex $INC
+cp include/soil_div_int_mod.tex $INC
 
 # Edit tables
-
[…]

### 2019-12-16 scripts/descrip.R

@@ -34,6 +34,13 @@ plot_data$clust4 <- factor(plot_data$clust4,
   levels = clust_names, 
   labels = clust_names)
 
+stem_ab_mat <- readRDS("data/stems_ab_mat.rds")
+
+stem_ab_mat$plot_group <- row.names(stem_ab_mat)
+
+stem_ab_mat_clean <- filter(stem_ab_mat, plot_group %in% plot_data$plot_group) %>%
+  dplyr::select(-plot_group)
+
 # Setup an example data frame
 df <- data.frame(id=c("id1","id2","id3","id4","id5","id6","id7","id8"),
   val=c(0,1,0,2,3,1,2,NA))
@@ -232,11 +239,15 @@ n_plots_num <- length(unique(plot_data$plot_group))
 # Plot number pre-outlier detection
 n_plots_pcoa <- n_plots_num + 110
 
+# How many stems did we measure?
+n_stems_total <- sum(rowSums(stem_ab_mat_clean))
+
 fileConn <- file(paste0("include/n_plots.tex"))
 writeLines(
   c(
     paste0("\\newcommand{\\nplots}{", n_plots_num, "}"),
-    paste0("\\newcommand{\\nplotspcoa}{", n_plots_pcoa, "}")),
+    paste0("\\newcommand{\\nplotspcoa}{", n_plots_pcoa, "}"),
+    paste0("\\newcommand{\\nstems}{", n_stems_total, "}")),
   fileConn)
 close(fileConn)
 

### 2019-12-16 scripts/sem.R

@@ -62,6 +62,7 @@ sem_data$precip_seasonality_std <- -sem_data$precip_seasonality_rev_std
 sem_data$temp_seasonality_std <- -sem_data$temp_seasonality_rev_std
 sem_data$sand_per_std <- -sem_data$sand_per_rev_std
 sem_data$mean_temp_std <- -sem_data$mean_temp_rev_std
+sem_data$shannon_equit_rev_std <- sem_data$shannon_equit_std 
 
 # Create a dataframe 
 corr_df <- sem_data %>%
@@ -105,9 +106,9 @@ ggcorrplot(cor(corr_df, use = "complete.obs"),
     "ocdens_std" = "Org. C %",
     "cec_std" = "CEC", 
     "total_precip_std" = "MAP",
-    "precip_seasonality_std" = expression(bolditalic(underline("MAP CV"))), 
+    "precip_seasonality_std" = expression(bolditalic(underline("PS"))), 
     "mean_temp_std" = expression(bolditalic(underline("MAT"))),
-    "temp_seasonality_std" = expression(bolditalic(underline("MAT CV"))),
+    "temp_seasonality_std" = expression(bolditalic(underline("TS"))),
     "n_species_raref_log_std" = "Extrap. sp. rich.",
     "shannon_equit_std" = "Shannon equit.",
     "cov_height_std" = "Tree height CoV",
@@ -193,7 +194,6 @@ writeLines(stargazer(corr_ci_tab,
   fileConn)
 close(fileConn)
 
-
 # Correlations tex variables
 fileConn <- file("include/corr_coef.tex")
 writeLines(
@@ -218,15 +218,12 @@ close(fileConn)
 
 # Structural SEM ----
 
-sem_data$shannon_equit_rev_std <- -1 * sem_data$shannon_equit_std
-
 struc_model_spec <- "
 # Latent vars
-div     =~  n_species_raref_log_std + shannon_equit_rev_std
+div     =~   shannon_equit_std + n_species_raref_log_std
 struc   =~  cov_dbh_std + cov_height_std
 
 # Modifications
-
 
 # Regressions
 agb_ha_log_std ~ c*div
@@ -289,7 +286,7 @@ mod_summ_mutate_struc <- function(x){
 struc_model_regs <- mod_summ_mutate_struc(struc_model_summ$PE)
   
 
-pdf(file = paste0("img/struc_model_slopes", ext, ".pdf"), width = 12, height = 4)
[…]

### 2019-12-17 include/clust_summ.tex
Sent a draft to supervisor

--- 2019-12-12 include/clust_summ.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Dec 12, 2019 - 12:06:51
+% Date and time: Tue, Dec 17, 2019 - 16:40:59
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-1 & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 679 & 11(11.1) & 172(148) & 37.6(34.88) \\ 
-2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 746 & 18(17.5) & 216(170) & 49.1(43.34) \\ 
-3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 225 & 10(10) & 166(160) & 46.2(47.82) \\ 
-4 & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\ 
+Marginal miombo & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 687 & 11(11.2) & 170(145.3) & 37(34.68) \\
+Core miombo & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 757 & 18(17.6) & 215(171.7) & 48.8(43.7) \\ 
+Baikiaea & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 226 & 10(10) & 165(157.5) & 46(47.81) \\ 
+Mopane & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\ 
 \hline \\[-1.8ex] 

### 2019-12-17 manuscript/include/clust_summ.tex
Sent a draft to supervisor

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Dec 16, 2019 - 11:33:55
+% Date and time: Tue, Dec 17, 2019 - 16:40:59
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-1 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
+\textit{Marginal miombo} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
 \\ \textit{Brachystegia spiciformis} \\ \textit{Baikeaea plurijuga} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Burkea africana} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 687 & 11(11.2) & 170(145.3) & 37(34.68) \\
 \hline
-2 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
+\textit{Core miombo} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
 \\ \textit{Brachystegia }spp.
@@
 \hline
-3 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia }spp.
+Baikiaea & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia }spp.
 \\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 226 & 10(10) & 165(157.5) & 46(47.81) \\
 \hline
-4 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Combretum }spp.
+Mopane & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Combretum }spp.
 \end{tabular} & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\

### 2019-12-17 manuscript/manuscript.tex
Sent a draft to supervisor

@@
 \usepackage{longtable}
+\usepackage{fmtcount}
 
@@
 \tableofcontents
+\newpage{}
+
+\begin{abstract}
+	Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard.
+In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots.
+We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density.
+A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity.
+We found that the effect of tree species diversity on biomass increases with stem density.
+Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.
+\end{abstract}
 
@@
 Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
-These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+%These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
 
@@
 \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}.
-In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity \citep{Wright2017, Poorter2015, Sande2017a}.
+In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity, which is driven by interspecific competition \citep{Wright2017, Poorter2015, Sande2017a}.
 In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function.
@@
 Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species.
-Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}.
-Southern African woodlands have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
-Despite these efforts however, human actions are driving rapid changes in biodiversity.
+Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and functionality.
 
@@
 Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}.
-Climate change also indirectly affects the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
-While rapid biodiversity change is being observed in southern African woodlands \citep{}, research into the relationship between biodiversity and ecosystem functionality remains scarce.
+Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
+While rapid biodiversity change is being observed in southern African woodlands \citep{Syampungani2009}, research into the relationship between biodiversity and ecosystem functionality remains scarce.
 
@@
 Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.
-In European forests \citep{Ratcliffe2017} found a general trend towards stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggest that in dry ecosystems, in contrast to the wet tropical systems in which much biodiversity-ecosystem function research has been conducted, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function.
-In dry woodland ecosystems stress tolerance rather than competitiveness may be the mechanism by which a tree species' success is measured.
+In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
+They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function.
+Stress tolerance rather than competitiveness may be the mechanism by which a tree species' success is measured in dry forested ecosystems.
 
@@
 
-While southern African woodlands are species rich in the herbaceous understorey \citep{Murphy2016}, the tree layer is relatively species poor.
-\citet{Solbrig1996} however, writes that they remain structurally diverse in the tree layer, with trees occupying distinct layers of the canopy at different growth stages and among species.
+%While southern African woodlands are species rich in the herbaceous understorey \citep{Murphy2016}, the tree layer is relatively species poor.
+
+\citet{Solbrig1996} writes that southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species.
 This structural diversity may be one mechanism through which diversity influences woody biomass.
@@
 They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production.
-Tree species in southern African woodlands may differ in their occupation of different canopy layers allowing a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity.
+Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity.
 This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition.
@@
 
-\citep{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity.
+%\citet{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity.
 They suggest this effect is caused by increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity.
@@
 
-In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which straddle multi-dimensional environmental and biogeographical gradients (\autoref{clust_map}).
+In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which span multi-dimensional environmental and biogeographical gradients (\autoref{clust_map}).
 We investigated the relationship between aboveground woody biomass and tree biodiversity.
@@
 We also investigated the potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type clusters based on major tree species compositional units into our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
-Initially, we made four hypotheses:
+We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
+Initially, we made three hypotheses:
 
@@
 \begin{enumerate}
-	\item{Precipitation and soil fertility will indirectly positively effect AGB via an increase in tree species diversity.}
-	\item{Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and AGB.}
-	\item{The effect size of species diversity on AGB will increase with stem density, due to an increased importance of niche complementarity as competition increases.}
+	\item{Precipitation and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity.}
+	\item{The effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases.}
+	\item{Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and woody biomass.} \label{lst:num}
 \end{enumerate}
 
-To address these hypotheses we used Structural Equation Models with path analysis as a preferred method to simultaneously account for environmental factors and biotic factors, which may interact in their effect on ecosystem structure and therefore biomass.
+%To address these hypotheses we used Structural Equation Models with path analysis as a preferred method to simultaneously account for environmental factors and biotic factors, which may interact in their effect on ecosystem structure and therefore biomass.
 
@@
 
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called Miombo woodlands (\hyperref[clust_map]{Figure~\ref*{clust_map}}, \citealt{White1987}).
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1987}).
 The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
 The study sites span the core climate space of the region as a whole.
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the Miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
 
@@
 Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
-Only plots with a stem density >10 stems ha\textsuperscript{-1} were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
+Only plots with a stem density >10 stems ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
 
@@
 After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition.
-We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbor criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
-We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thus removing \noutliers{} plots.
+We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbour criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
+We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thus removing \noutliers{} plots (\hyperref[appendixa]{Appendix A}).
 
@@
 
-We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
+We considered only trees and shrubs in our calculations of above-ground woody biomass (AGB), including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
 Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
 Many plots in the dataset did not include data on stems <10 cm DBH.
-For those plots with stem measurements <10 cm DBH, those small stems only accounted for a median average of \percsmallagb{} of the plot level AGB.
+For those plots with stem measurements <10 cm DBH, those small stems only accounted for a median average of \percsmallagb{}\% of the plot level AGB.
 
-All stems >10 cm DBH were measured within each plot resulting in a total of \nstems{} stems with measurements.
+All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\nstems{}} stems with measurements.
 A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
@@
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
-The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from (Ryan, unpublished).
+The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from (Ryan C., unpublished) (\hyperref[appendixb]{Appendix B}).
 
-Woody AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}.
+AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}.
 Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
@@
 
-Where $\rho$ is the species level mean wood density, $D$ is the DBH at 1.3 m, and $H$ is the tree height.
+Where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height.
 
@@
 Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
-Temperature and precipitation seasonality were both calculated as the mean of the coefficient of variation of daily MAT and MAP, respectively, for each of the 18 years of available data.
+Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily MAT and MAP, respectively, for each of the 18 years of available data.
 Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}.
@@
 C \%), and percentage soil sand content by volume (Sand \%).
-These data are a modelled product derivedfrom various remotel sensed and directly measured data sources.
+These data are a modelled product derived from various remotely sensed and directly measured data sources.
 
@@
 This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
-Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort.
+Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots.
 Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.
@@
 
-Where $H'_{e}$ is an estimation of the Shannon diversity index of tree species by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot.
-We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH and tree height.
+Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot, using the \verb|ChaoRichness()| function.
+We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).
 
@@
 Plots were assigned to vegetation type groups based on tree species composition.
-Groups were identified in \citet{Fayolle2018} via an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
+Groups were identified in \citet{Fayolle2018} in an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
 Groups were identified using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes.
-Plot data used in this study occurred in five vegetation type groups.
+Plot data used in this study occurred in four vegetation type groups.
 See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .
@@
 	\includegraphics[width=1.4\textwidth]{clust_map}
-	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of Miombo woodland vegetation according to \citet{White1987}.
+	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1987}.
 Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
@@
 Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset.
-Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and woody species diversity, which could otherwise increase the chances of Type I error and wrongly attributing inference due to covariance of explanatory variables, if testing for significant effects of biodiversity on AGB using conventional regression analyses \citep{Nachtigall2003}.
+Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
 
@@
 
-Observed variables were standardised to Z-scores prior to analysis.
-Standardization put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
-Standardization allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
-Standardization also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
+Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixc]{Appendix C}).
+Standardisation put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
+Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
+Standardisation also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
 To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1.
 For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed for model fitting.
-Precipitation seasonality, temperature seasonality, and mean temperature were also reversed in this way to account for the direction of their effect on moisture availability.
+Precipitation seasonality (PS), temperature seasonality (TS), and mean annual temperature (MAT) were also reversed in this way to account for the direction of their effect on moisture availability.
+
+%While it is recommended by some to set exact factor loadings in the SEM from the regression coefficients of multiple regressions \citep{lavaan}, because some latent variables were regressed against both structural diversity and AGB, exact factor loadings from simple multiple regressions could not be used.
+We therefore allowed factor loadings to be estimated by the SEM itself.
 
 The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{Beaujean2014}.
-While it is recommended by some to set exact factor loadings in the SEM from the regression coefficients of multiple regressions \citep{lavaan}, because some latent variables were regressed against both structural diversity and AGB, exact factor loadings from simple multiple regressions could not be used.
-We therefore allowed factor loadings to be estimated by the SEM itself.
 We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, while ensuring no factor loadings were in excess of 1.
@@
 
-First, we assessed the interacting factors of structural diversity and species diversity in determining AGB.
+First, we assessed the role of structural diversity and species diversity in determining AGB.
 We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
+To account for variation in stem density which may covary with species diversity we also included it as an predictor in our model.
 To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately.
@@
 Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
-We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.9, TLI >0.9, RMSEA <0.1, alongside our judgement of the model estimates.
+We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.
 
-To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create 900 datasets of similar size with varying median stem density.
+To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density.
 We used each of these datasets to fit the model including only tree species and structural diversity latent variables to predict AGB.
@@
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
-Colours of variable names group them into latent variables used in the SEMs: red = Soil fertility, blue = Moisture availability, green = tree species diversity, purple = tree structural diversity.
-See \autoref{corr_ci_tab} for a full assessment of correlation fit statistics.}
+Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = moisture availability, green = tree species diversity, purple = tree structural diversity.
+See \hyperref[appendixd]{Appendix D} for a full assessment of correlation fit statistics.}
 	\label{corr_mat}
@@
 The direction of these correlations was used as a test of our assumptions of the direction of influence of latent variables later used in the SEMs.
-As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a negative correlation between the seasonality of precipitation and AGB (\ccmcb{}.
-MAT and temperature seasonality (TS) negatively correlated weakly with AGB (MAT: \cctb{}; TS\cctcb{}).
+As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a weak negative correlation between the seasonality of precipitation and AGB (\ccmcb{}).
+MAT and temperature seasonality (TS) negatively correlated weakly with AGB (MAT: \cctb{}; TS: \cctcb{}).
 As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic carbon and AGB (\ccob{}).
 
-MAP had positive correlations with tree species richness (\ccms{}), abundance evenness (\ccme{}), tree height diversity (\ccmh{}) and a positive but not significant correlation with tree stem density.
+MAP had positive correlations with tree species richness (\ccms{}), abundance evenness (\ccme{}), tree height diversity (\ccmh{}) and tree stem density (\ccmi{}).
 MAT had weak correlations with tree species and structural diversity variables.
@@
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity suprisingly had a small negative direct effect on AGB, while positive effects were observed via structural diversity (\autoref{struc_mod}).
+In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a small positive direct effect on AGB (\strucdb{}), and indirectly via structural diversity (\strucdsb{}) (\autoref{struc_mod}).
 Tree species diversity had a positive correlation with stem density.
@@
 The R\textsuperscript{2} of AGB was \strucrsq{}.
-The strongest direct effect on AGB was via stem density (\strucsib{}).
+The strongest direct effect on AGB was from stem density (\strucsib{}).
 
@@
 Latent variables are circles while observed variables are rectangles.
-Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
+Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
 The observed variables which inform the latent variables are connected by dotted arrows, observed variables with loading set to 1 are connected by dashed arrows.
@@
 
-\subsection{Variation among vegetation clusters in structural and species diversity effects}
+\subsection{Variation among vegetation types}
 
-When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation clusters the strengths of unstandardised path coefficients varied.
+When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation types the strengths of unstandardised path coefficients varied.
 The direct effect of tree species diversity on AGB was positive in Baikiaea and Mopane, but negative in Marginal and Core miombo (\autoref{struc_model_slopes_all}).
 Relationships between structural diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
-The total effect of species diversity on AGB remained strongly positive.
+The total effect of species diversity on AGB remained strongly positive for all vegetation types.
 All vegetation types except Mopane exhibited a positive effect of species diversity on structural diversity.
-All models had adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though confidence intervals around the unstandardised path coefficients were wide particularly for Mopane and Baikiaea indicate.
-$\chi^{2}$ statistics were high for some vegetation clusters, but this appears to be highly correlated with sample size for each vegetation type \citep{}.
+All models had adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though confidence intervals around the unstandardised path coefficients were wide particularly for Mopane and Baikiaea.
+$\chi^{2}$ statistics were high for some vegetation types, but this appears to be highly correlated with sample size for each vegetation type \citep{Hooper2008}.
 
-The strongest total effect of tree species diversity on AGB was in Baikiaea woodland (\struccsb{}), which was species rich but highly variable in species diversity compared to other vegetation clusters (\autoref{clust_summ}).
-The R\textsuperscript{2} of AGB was highest in Baikiaea (R\textsuperscript{2} = \struccrsq{}) and lowest in the Core miombo (R\textsuperscript{2} = \strucbrsq{}).
+The strongest total effect of tree species diversity on AGB was in Baikiaea woodland (\struccsb{}), which was species rich but highly variable in species diversity compared to other vegetation types (\autoref{clust_summ}).
+The R\textsuperscript{2} of AGB was highest in Marginal miombo (R\textsuperscript{2} = \strucarsq{}) and lowest in the Core miombo (R\textsuperscript{2} = \strucbrsq{}).
 
@@
 	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
-Path coefficients are $\pm$ 1 standard error.
+Path coefficients are $\pm$1 standard error.
 Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
@@
 
-\subsubsection{Moderation of Diversity-AGB relationship by stem density}
+\subsection{Moderation of Diversity-AGB relationship by stem density}
 
@@
 \autoref{sem_struc_stems_ha} shows a positive effect of tree species diversity on AGB as stem density increases.
-There appears to be a minimum stem density threshold at \textapprox{}180 stems ha\textsuperscript{-1} below which there appears to be only a baseline effect of tree diversity on biomass.
+There appears to be a minimum stem density threshold at \textapprox{}180 stems ha\textsuperscript{-1} below which there appears to be a reasonably constant low baseline effect of tree diversity on biomass.
 The effect of structural diversity on AGB appears to remain constant with increasing stem density.
-The indirect effect of species diversity on AGB via structural diversity climbs slightly and stem density increases.
+The indirect effect of species diversity on AGB via structural diversity climbs slightly as stem density increases.
 
@@
 
-\subsection{Mediation of environmental covariates via diversity}
+\subsection{Environmental covariates and diversity}
 
-A model incorporating the latent variables of moisture availability and soil fertility showed that the effect of diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}).
+A model incorporating the latent variables of moisture availability and soil fertility showed that the total effect of species diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}).
 Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}).
@@
 
-Similar to the model which only considered tree species and structural diversity, the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB (\rghb{}).
+Similar to the model which only considered tree species and structural diversity (\autoref{struc_mod}), the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB, leading to a strong positive indirect effect of species diversity on AGB via structural diversity (\rgbhd{}).
+The total effect of species diversity on AGB was positive (\rgbd{}).
 
@@
 	\input{include/path_diagram_full}
-	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation type clusters.
+	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation types.
 Latent variables are shown as circles while observed variables are shown as rectangles.
-Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale $\pm$1 standard error.
+Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
 The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
@@
 
-In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] moisture availability and soil fertility, [d] stem density and their interactions on aboveground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} plots.
-Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB, and an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{1}) and via stem density (H\textsubscript{3}).
+In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability via moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots.
+Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB, with an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{1}).
+We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).
 Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
 The strongest effect on AGB was that of stem density.
-Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{2}).
-We found that vegetation composition in the form of discrete vegetation type clusters affected the relationship between tree species diversity and AGB, with Baikiaea woodland representing an outlier in our models.
-
-Miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB.
+Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of an effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{3}).
+We found that vegetation composition in the form of discrete vegetation types affected the relationship between tree species diversity and AGB.
 
@@
 
-We found a consistent positive effect of tree species diversity on AGB across all models in this study.
+We found a consistent positive total effect of tree species diversity on AGB across all models in this study.
 Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness causes higher woody AGB.
 This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}.
-In addition however, our study provides a novel dissection of the mechanisms underlying this relationship and their interaction with resource availability in southern African woodlands as an understudied biome.
+In addition however, our study provides a novel dissection of the mechanisms underlying this relationship and their interaction with resource availability and vegetation type in southern African woodlands as an understudied biome.
 
@@
 
-While we did not explicitly measure net primary production (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
+While we did not explicitly measure Net Primary Productivity (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
 This suggests that as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in southern African woodlands can be maximised by increasing species diversity.
@@
 
-We found evidence that tree species diversity led to an increase in AGB via tree structural diversity and we can therefore accept our hypothesis (H\textsubscript{2}).
+We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity and we therefore find support for our hypothesis (H\textsubscript{2}).
 A higher tree species diversity allows for a greater diversity of tree functional forms within a plot and this may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
@@
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
-In frequently disturbed woodlands such as those studied here however, a tree canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained.
-Unfortunately we did not have access to a dataset of high enough spatial and temporal resolution to adequately quantify the effects of disturbance history.
+In frequently disturbed woodlands such as those studied here however, a tree canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
+While we did not have access to adequate data on disturbance history in our plots, previous studies have found that southern African woodlands with higher species diversity tend to be less disturbed and tend to form a more closed canopy \citep{Chidumayo2013, Mutowo2012}.
 
@@
 Surprisingly, moisture availability and soil fertility had only small effects on AGB compared to that of tree species diversity.
-We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater resource partitioning among individual trees \citep{} and a greater stem density per unit area \citep{}.
+We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater resource partitioning among individual trees and a greater stem density per unit area \citep{Kraaij2006, Shirima2015}.
 
 Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via increasing tree species diversity and via increasing stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}.
-In this study, while we observed the indirect effects on species diversity, we saw no evidence for a direct effect on AGB.
-Compared to moist tropical forests, climatic water availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted.
+In this study, while we observed weak indirect effects via species diversity, we saw no evidence for a direct effect on AGB.
+Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted.
 It is possible that the range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture variation in AGB.
-We deliberately exlcuded plots with very low stem density as they are not considered woodlands, but grassy savannas.
+We deliberately excluded plots with very low stem density as they are not considered woodlands, but grassy savannas.
 It may be that by excluding the bottom end of this stem density continuum we prevented a relationship being observed between moisture and AGB/stem density.
@@
 The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth \citep{Charles-Dominique2018}, and may also limit species diversity \citep{Linder2014}.
-It is possible therefore that the effect of moisture availabiliy is confounded in it effect on AGB with the unmeasured variable of intensity of disturbance by fire.
+It is possible therefore that the effect of moisture availability, which is expected to increase AGB, is confounded in its effect on AGB with the unmeasured variable of fire regime intensity, which is expected to decrease AGB.
 The direct effect of moisture availability on stem density may also be confounded in this way.
+This may also have caused us to not observe a stronger effect between moisture availability and AGB.
 
@@
 
-\subsection{Vegetation type cluster specific responses}
+\subsection{Vegetation type specific responses}
 
@@
 
-\todo{I have no idea what to write here just yet}
+Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect via structural diversity remained positive in these vegetation types.
+Compared to Baikiaea and Mopane woodlands these woodlands have higher median tree species richness.
+Baikiaea and Mopane woodlands are also dominated by fewer tree species, notably \textit{Baikiaea plurijuga} in Baikiaea woodlands and \textit{Colophospermum mopane} in Mopane woodlands which tend to produce large canopy dominating trees.
+We postulate that this negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in Baikiaea and Mopane woodlands, where the woodland canopy is dominated often by a single species.
+Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation.
+Again, these highly diverse plots in miombo woodlands may be the result of disturbance which can promote a mosaic of woodland of different successional stages and stem densities.
 
-\todo{Baikiaea had the strongest total effect of species div on AGB. }
+Despite Mopane woodland having very low species diversity generally, with often monospecific stands \citep{Timberlake2010}, a positive effect of tree species diversity on AGB was observed.
+In previous studies across ecosystem types it has been found often that the effect on ecosystem function of adding species is stronger in low diversity assemblages \citep{Hector2007}.
+This has been attributed to an increase in functional redundancy as species diversity increases.
+\textit{I.e.} with more species, it is more likely that the addition of a new species will occupy the same ecological niche space as an existing species, meaning niche complementarity will not occur and competition will lead to niche partitioning, while making little difference to overall ecosystem functioning.
+Mopane woodlands also have a negligible effect of species diversity on structural diversity.
+This may be due to the species which tend to co-exist with \textit{C.
+mopane}, many of which are small shrub-like trees which do not grow into large canopy trees \citep{Timberlake2010}.
+Larger canopy trees tend to have greater variation in physical structure \citep{Seidel2019}.
+
+Baikiaea woodland had the strongest total effect of species diversity on AGB.
+Baikiaea also has relatively low median species richness compared to miombo and Mopane woodlands, but the addition of new species appears to make a larger difference to the AGB of these plots than mopane.
+We suggest that this is due mostly to the particular identity of species found in Baikiaea woodlands and their contribution to ecosystem functioning.
+Unlike mopane woodlands, Baikiaea woodlands do sometimes contain species other than \textit{B.
+plurijuga} which grow to be high biomass canopy trees.
 
@@
 
-We found a non linear effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
-The total effect of tree species diversity on AGB increases with increasing stem density.
+We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
 At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness might not make any difference to plot level AGB, accounting for the low and constant effect of tree species diversity on AGB below \textapprox{}180 stems ha\textsuperscript{-1}.
@@
 
-% Seidel 2019: Large trees tend to possess a greater structural complexity than small trees, not due to their size per se, but due to more complex architecture.
-
 % At very high biomass, tree species diversity is low because it is associated with intense competition for light leading to dominance of usually a single species (Huston 1997, Oindo and Skidmore 2002).
@@
 
+In this study we found that across southern African woodlands, there is a generalisable positive association between tree species diversity and woody biomass as a measure of ecosystem functionality.
+Additionally, we found that much of this effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of woodland tree canopies.
+We found that the multiple vegetation types which comprise southern African woodlands exhibit variation in the strength of the relationship between species diversity and woody biomass.
+In contrast to previous studies, we found that across the region, the direct effects of moisture availability and soil fertility on woody biomass were negligible, with most of their effect being indirectly through species and structural diversity.
+A gap in available data means that we could not incorporate disturbance history into our models adequately, but this factor likely plays a large part in the association between species diversity and woody biomass in southern African woodlands.
+
 \bibliography{/Users/johngodlee/google_drive/bib/papers/lib.bib}
@@
 
-\section*{Appendix 1 - Data cleaning process}
+\section*{Appendix 1 - Data cleaning process} \label{appendixa}
 
@@
 
-\section*{Appendix 2 - Table of correlation fit statistics}
+\section*{Appendix 2 - Estimation of DBH via tree taper} \label{appendixb}
 
-\input{include/corr_ci_tab.tex}
+\begin{lstlisting}[language=R]
+##' @title Stem diameter Point Of Measurement (POM) adjustment
+##' @description  Function to estimate stem diameter at 1.3 given measurements 
+##'   at other POMs.
+##' @author Casey M. Ryan
+##' @return d130, the estimated diameter at a POM of 1.3 m (in cm). 
+##' @param d_in the diameter measured at the POM (in cm)
+##' @param POM the height of the POM (in m)
+##' @details The adjustment is based on a tree taper model developed as part of 
+##'   the ACES project (Abrupt Changes in Ecosystem Services 
+##'   https://miomboaces.wordpress.com/), using data from the miombo of Niassa. 
+##'   The model is a cubic polynomial, with three equations for different sized 
+##'   stems. 
+##' @section Warning: The model should not be used for POMs above 1.7 m. 
+##'   Extrapolating beyond the training data will give nonsense. 
+##'   Thus, POMs >1.7 m are not adjusted.
+##' @examples
+##' POMadj(10, 0.3)
+##' POMadj(1, 0.3)  # d130 is negative, i.e. the stem probably wasn't 1.3 m tall
+##' POMadj(50, 1.9)  # generates warning, as outside calibration data range
+##' \dontrun{
+##'   POMadj(50, 0)  # zero or -ve POM is outside range, or nonsense
+##' }
+POMadj <- function(d_in, POM) {
+  stopifnot(is.numeric(d_in),
+    is.numeric(POM),
+    POM >= 0,
+    sum(is.na(POM))==0,
+    length(POM) == length(d_in))
+  if (any(POM > 1.7))
+    warning("POMs >1.7 m are outside the calibration data, no correction applied")
 
-\section*{Appendix 3 - Frequency distribution of observed variables}
+  NAS <- is.na(d_in)
+  d_in_clean <- d_in[!NAS]
+  POM_clean <- POM[!NAS]
+  # define the size class edges:
+  edges <- c(5.0, 15.8, 26.6, 37.4)
+  sm <- d_in_clean < edges[2]
+  med <- d_in_clean >= edges[2] & d_in_clean < edges[3]
+  lg <- d_in_clean >= edges[3]
+
+  # compute apredictions for delta_d, for all size classes
+  delta_d <- data.frame(
+    # if small:
+    small =  3.4678+-5.2428 * 
+   	  POM_clean + 2.9401 * 
+      POM_clean^2+-0.7141 * 
+      POM_clean^3,
+    # if med
+    med =  4.918+-8.819 * 
+      POM_clean + 6.367 * 
+      POM_clean^2+-1.871 * 
+      POM_clean^3,
+    # if large
+    large =  9.474+-18.257 * 
+      POM_clean + 12.873 * 
+      POM_clean^2+-3.325 * 
+      POM_clean^3
+  )
+  # index into the right size class
+  dd <- NA_real_
+  dd[sm] <- delta_d$small[sm]
+  dd[med] <- delta_d$med[med]
+  dd[lg] <- delta_d$large[lg]
+  dd[POM_clean > 1.7] <- 0  # to avoid extrapolation mess
+
+  # add NAs back in
+  d130 <- NA
+  d130[NAS] <- NA
+  d130[!NAS] <- d_in_clean - dd
+
+  if (any(d130[!NAS] < 0))
+    warning("Negative d130 estimated, repaced with NA")
+  d130[d130 <= 0 & !is.na(d130)] <- NA
+  return(d130)
+}
+\end{lstlisting}
+
+\section*{Appendix 3 - Frequency distribution of observed variables} \label{appendixc}
 
@@
 
-\section*{Appendix 4 - Regression tables for moderation by environmental variables}
+\section*{Appendix 4 - Table of correlation fit statistics} \label{appendixd}
 
-\input{include/mois_div_int_mod}
-
-\input{include/soil_div_int_mod}
+\input{include/corr_ci_tab.tex}
 

### 2019-12-17 scripts/sem.R

@@ -151,7 +151,7 @@ p_format <- function(p){
 }
 
 corr_format <- function(id, row_num){
-  paste0("\\newcommand{\\", id, "}{", "$\\rho =$ ", corr_ci_tab$raw.r[row_num], ", ", corr_ci_tab$p[row_num], "}")
+  paste0("\\newcommand{\\", id, "}{", "$r =$ ", corr_ci_tab$raw.r[row_num], ", ", corr_ci_tab$p[row_num], "}")
 }
 
 corr_ci_tab <- corr_ci_tab %>%
@@ -395,6 +395,8 @@ print(ggplot() +
 dev.off()
 }
 
+struc_model_regs$est[1] <- -struc_model_regs$est[1]
+
 mod_slopes_all(struc_model_regs_list, struc_model_regs, 
   paste0("struc_model_slopes_all", ".pdf"))
 
@@ -462,15 +464,35 @@ writeLines(
     paste0("\\newcommand{\\pcsded}{", struc_model_edge_df$est[2], "}"),
     paste0("\\newcommand{\\pcshdh}{", struc_model_edge_df$est[3], "}"),
     paste0("\\newcommand{\\pcshhh}{", struc_model_edge_df$est[4], "}"),
-    paste0("\\newcommand{\\pcsdb}{", struc_model_edge_df$est[5], "}"),
+    paste0("\\newcommand{\\pcsdb}{", -struc_model_edge_df$est[5], "}"),
     paste0("\\newcommand{\\pcshb}{", struc_model_edge_df$est[6], "}"),
     paste0("\\newcommand{\\pcsdh}{", struc_model_edge_df$est[7], "}"),
     paste0("\\newcommand{\\pcsib}{", struc_model_edge_df$est[8], "}"),
     paste0("\\newcommand{\\pcsdi}{", struc_model_edge_df$est[9], "}"),
     paste0("\\newcommand{\\strucrsq}{", round(lavInspect(struc_model_fit, "rsquare")[5], digits = 2), "}"),
     paste0("\\newcommand{\\strucbrsq}{", round(lavInspect( struc_model_fit_clust_list[[2]], "rsquare")[5], digits = 2), "}"),
-    paste0("\\newcommand{\\struccrsq}{", round(lavInspect( struc_model_fit_clust_list[[3]], "rsquare")[5], digits = 2), "}"),
[…]

### 2020-01-17 include/clust_summ.tex
Going through Kyle's comments

--- 2019-12-12 include/clust_summ.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Dec 12, 2019 - 12:06:51
+% Date and time: Fri, Jan 17, 2020 - 12:26:04
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-1 & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 679 & 11(11.1) & 172(148) & 37.6(34.88) \\ 
-2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 746 & 18(17.5) & 216(170) & 49.1(43.34) \\ 
-3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 225 & 10(10) & 166(160) & 46.2(47.82) \\ 
-4 & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\ 
+Marginal miombo & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 688 & 11(11.2) & 152(135.2) & 32.9(30.57) \\
+Core miombo & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 754 & 18(17.5) & 193(174.1) & 44.8(41.43) \\ 
+Baikiaea & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 226 & 10(10) & 162(152) & 45.3(47.36) \\ 
+Mopane & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\ 
 \hline \\[-1.8ex] 

### 2020-01-17 manuscript/include/clust_summ.tex
Going through Kyle's comments

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Tue, Dec 17, 2019 - 16:40:59
+% Date and time: Fri, Jan 17, 2020 - 12:26:04
 \begin{table}[!htbp] \centering 
@@
 \textit{Marginal miombo} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
-\\ \textit{Brachystegia spiciformis} \\ \textit{Baikeaea plurijuga} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Burkea africana} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 687 & 11(11.2) & 170(145.3) & 37(34.68) \\
+\\ \textit{Brachystegia spiciformis} \\ \textit{Baikeaea plurijuga} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Burkea africana} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 688 & 11(11.2) & 152(135.2) & 32.9(30.57) \\
 \hline
@@
 \\ \textit{Brachystegia }spp.
-\\ \textit{Isoberlinia angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Brachystegia longifolia} \end{tabular} & 757 & 18(17.6) & 215(171.7) & 48.8(43.7) \\
+\\ \textit{Isoberlinia angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Brachystegia longifolia} \end{tabular} & 754 & 18(17.5) & 193(174.1) & 44.8(41.43) \\
 \hline
 Baikiaea & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia }spp.
-\\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 226 & 10(10) & 165(157.5) & 46(47.81) \\
+\\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 226 & 10(10) & 162(152) & 45.3(47.36) \\
 \hline

### 2020-01-17 manuscript/include/struc_model_fit_clust_stats.tex
Sent a draft to supervisor
Going through Kyle's comments

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Dec 16, 2019 - 16:41:44
+% Date and time: Fri, Jan 17, 2020 - 12:26:16
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-Marginal miombo & 525 & 47.480 & 6 & 0.964 & 0.910 & -3730.100 & 0.110 & 0.730 \\ 
-Core miombo & 668 & 59.440 & 6 & 0.958 & 0.895 & -4219 & 0.100 & 0.680 \\ 
-Baikiaea & 47 & 5.860 & 6 & 0.998 & 0.994 & -323.100 & 0.030 & 0.720 \\ 
-Mopane & 84 & 9.420 & 6 & 0.971 & 0.927 & -588.900 & 0.080 & 0.450 \\ 
-All & 1324 & 82.020 & 6 & 0.973 & 0.932 & -9122.800 & 0.090 & 0.700 \\ 
+Marginal miombo & 525 & 44.750 & 6 & 0.966 & 0.916 & -3714.000 & 0.110 & 0.710 \\ 
+Core miombo & 668 & 57.210 & 6 & 0.962 & 0.904 & -4224.000 & 0.100 & 0.680 \\ 
+Baikiaea & 47 & 5.860 & 6 & 0.998 & 0.994 & -324.600 & 0.030 & 0.720 \\ 
+Mopane & 84 & 9.420 & 6 & 0.971 & 0.927 & -591.600 & 0.080 & 0.450 \\ 
+All & 1324 & 78.430 & 6 & 0.975 & 0.936 & -9119.000 & 0.090 & 0.690 \\ 
 \hline \\[-1.8ex] 

### 2020-01-17 manuscript/manuscript.tex
Going through Kyle's comments

@@
 
-\usepackage{mynotes}
+% Define page geometry
+\usepackage{geometry}
+\geometry{left=2.2cm,
+	right=2.2cm,
+	top=2.2cm,
+	bottom=2cm}
+\parskip 0.15cm
+\setlength{\parindent}{0cm}
+\usepackage{pdflscape}
+
+% Set font
+\usepackage[T1]{fontenc}
+
+% Image handling
+\usepackage{graphics}  % Insert images easily
+\usepackage{graphicx}  % Extended image support
+
+\makeatletter
+	\g@addto@macro\@floatboxreset\centering  % Automatically centre images (floats)
+\makeatother
+
+\graphicspath{ {img/} }
+\usepackage{float}  %  Graphics placement [H] [H!] arguments
+\usepackage{subfig}  % Compound figures
+
+% Tables
+\usepackage{multirow}
+\usepackage{longtable}
+
+% Bibliography management
+\usepackage{natbib}    % Bibliography management - Use author/date citations
+\bibliographystyle{agsmnourl}  % Use custom agsm bibliography template with no URL
+\usepackage{cite}  % Citation options
+
+% Text formatting
+\usepackage{url} % Allow nice formatting of URLs in text
+
+\usepackage{enumerate}  % Enumerated lists
+
+\usepackage{lineno}  % Line numbers
+
+\usepackage{textcomp}
+\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}}  % Command for a good tilde
+
+\usepackage{siunitx}
+\usepackage{amsmath}
+
+\usepackage{xcolor}
+\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}   % \todo{NOTE TO SELF WRITTEN IN RED}
+
+\input{code_format}
+
+% Custom title formatting
+\let\oldtitle\title
+
+\renewcommand{\title}[1]{\oldtitle{\vspace{-1.5cm}#1}}
+
+\usepackage[breaklinks]{hyperref}
+\definecolor{links}{RGB}{191,59,72}
+\hypersetup{
+	breaklinks,
+	colorlinks,
+	allcolors=links,
+	linktoc=section,
+	pdfauthor={John L. Godlee}
+}
+\def\subsectionautorefname{section}
+\def\subsubsectionautorefname{section}
+% \usepackage{biblatex}
 
@@
 
+\usepackage{lineno}
+\linenumbers
+
 \title{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
-\author{John L. Godlee}
+\author{}
 \date{}
@@
 
-\maketitle
-\tableofcontents
+{\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}}
+
+\vspace{1cm}
+
+Authors: Godlee, J. L.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1}, Ryan, C. M.\textsuperscript{1}
+
+\textsuperscript{1}: School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom \\
+\textsuperscript{2}: Some other address
+
+\vspace{1em}
+Corresponding author:
+
+John L. Godlee
+
+johngodlee@gmail.com
+
+School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom
+
+\section*{Acknowledgements}
+
+\section*{Biosketch}
 \newpage{}
 
-\begin{abstract}
-	Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard.
+{\LARGE{\textbf{Blinded Main Text File}}}
+
+\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
+
+\normalsize{Running title: Woodland functionality in southern Africa}
+
+\section*{Abstract}
+
+\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship in highly disturbed and ecophysiologically stressful systems is less clear.
+Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density and resource availability, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+
+\textbf{Location:} Southern African savannas and woodlands
+
+\textbf{Time period:} 2010-2019
+
+\textbf{Major taxa studied:} Trees
+
+\textbf{Methods:} We used a network of \nplots{} savanna and woodland plots located across the southern African sub-continent in which each tree >5 cm diameter was measured.
+We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density.
+
+\textbf{Results:} A positive effect of tree species diversity on biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity.
+We also found that the effect of tree species diversity on biomass increases with stem density.
+Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.
+
+\textbf{Main conclusions:} Tree species diversity plays an important role in determining the ecosystem functionality and structure of southern African woodlands.
+Our results demonstrate the importance of considering environmental conditions and woodland structural attributes to more accurately describe the relationship between biodiversity and ecosystem functionality.
+
+% Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard.
 In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots.
@@
 Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.
-\end{abstract}
 
-\section{Introduction}
+\section*{Introduction}
 
-A large number of studies have shown relationships between biodiversity and ecosystem functionality (e.g.
+Numerous studies have shown relationships between biodiversity and ecosystem function (e.g.
 \citealt{Liang2016, Cardinale2009}).
@@
 Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
-%These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
 
@@
 This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.
+The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.
+
+Compared to other forested ecosystems, dry tropical woodlands are highly structured by disturbance, mainly through fire, and in Africa notably by herbivory also \citep{}.
+High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
+This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland ecosystems.
+Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}.
+Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth.
+In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
+They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function.
+This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation.
+
+% COME BACK TO THIS: In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}.
+It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
+Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.
 
@@
 \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}.
-In wet tropical forests, much of the observed effect of biodiversity on ecosystem function can be attributed to niche complementarity, which is driven by interspecific competition \citep{Wright2017, Poorter2015, Sande2017a}.
-In the dry tropics however, low precipitation, high seasonality of rainfall and temperature, and high levels of disturbance from fire and herbivory mean that niche complementarity and competition may not play as great a role in the relationship between biodiversity and ecosystem function.
-Disturbance may weaken the role of competition in determining local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979,Grace1990}.
-Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}.
-Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth \citep{Ratcliffe2017}.
+A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and it's environmental determinants.
 
-Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{clust_map}).
-The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
-Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species.
-Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (\autoref{clust_map}).
+The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
+Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species.
+Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
@@
 
-Southern African woodlands are relied upon heavily for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
-This has raised interest in how biodiversity influences ecosystem function in these ecosystems.
-Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}.
-Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
-While rapid biodiversity change is being observed in southern African woodlands \citep{Syampungani2009}, research into the relationship between biodiversity and ecosystem functionality remains scarce.
-
-A small number of studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of these studies has been hampered by a restricted climatic and biogeographical range of study sites.
-Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass.
-Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
+A small number of studies in southern African savannas and woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+However, studied conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
 Small scale studies may therefore be over-estimating the effect of diversity.
-
-In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}.
-It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
-Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.
-In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function.
-Stress tolerance rather than competitiveness may be the mechanism by which a tree species' success is measured in dry forested ecosystems.
+Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass \citep{}, it is important to sample across geographic and environmental gradients to be able to infer a truly regional scale relationship between diversity and biomass.
 
@@
 They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production.
-Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area as a form of niche complementarity.
+Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity.
 This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition.
@@
 
-%\citet{Zhang2012} conducted a meta-analysis of 54 studies of the BEFR in forest ecosystems and found that greater tree species abundance evenness led to an increased effect size of species richness on productivity.
-They suggest this effect is caused by increased heterogeneity of functional traits over space at small scales, leading to a more evenly filled environmental niche volume, which allows a greater number of species to maximise their productivity.
-We expect that this effect will occur in southern African woodlands as well.
-A decrease in species abundance evenness may be found as a result of increased disturbance, which prevents dominance by one species, or due to more extreme environmental conditions which preclude the growth of many species, instead favouring those few species which can effectively tolerate sub-optimal conditions.
-It is important therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.
+% High levels of disturbance in southern African woodlands may moderate the observable BEFR through it's effect on ecosystem composition.
+Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}.
+A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion.
+These fast-growing bushy species often proliferate in southern African woodlands when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}.
+Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
+% KGD: This text is useful in demonstrating your biological understanding of the ecosystem.
+I think it could be useful to include some of this biological knowledge in the paragraph above that ends with the statements on disturbance (the paragraph starting with Slobrig 1996).
+Or you could include this whole paragraph just after that one.
+The co-authors will reign us in if this kind of logic is not sensible for the ecosystems we are studying.
 
-%High levels of disturbance in the form of seasonal fires and herbivory from large mammals may moderate the strength of observable BEFRs in southern African woodlands \citep{Staver2009, Bond2005}.
-Fire disturbance in forests has been linked to abundance dependent mortality among smaller stems \citep{Roques2001}.
-A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion.
-These fast-growing bushy species exist in southern African woodlands and often proliferate when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}.
-Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
+In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
+We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
+We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
+We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality and assess the generality of our results.
+We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, which may interact their effect on ecosystem structure and therefore biomass.
+Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
 
-In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which span multi-dimensional environmental and biogeographical gradients (\autoref{clust_map}).
-We investigated the relationship between aboveground woody biomass and tree biodiversity.
-We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation: precipitation, temperature and soil fertility.
-We also investigated the potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
-Initially, we made three hypotheses:
+\section*{Materials and methods}
 
-\subsection{Hypotheses}
+\subsection*{Study location}
 
-\begin{enumerate}
-	\item{Precipitation and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity.}
-	\item{The effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases.}
-	\item{Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and woody biomass.} \label{lst:num}
-\end{enumerate}
-
-%To address these hypotheses we used Structural Equation Models with path analysis as a preferred method to simultaneously account for environmental factors and biotic factors, which may interact in their effect on ecosystem structure and therefore biomass.
-
-\section{Methods}
-
-\subsection{Study location}
-
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1987}).
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1983}).
 The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
 The study sites span the core climate space of the region as a whole.
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1983}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
 
@@
 
-\subsection{Data collection}
+\subsection*{Data collection}
 
@@
 
-\subsection{Data analysis}
+\subsection*{Data analysis}
 Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}.
@@
 
-\subsubsection{Vegetation clusters}
+\subsubsection*{Vegetation clusters}
 
@@
 	\includegraphics[width=1.4\textwidth]{clust_map}
-	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1987}.
+	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1983}.
 Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
@@
 
-\subsubsection{Structural Equation Modelling}
+\subsubsection*{Structural Equation Modelling}
 
@@
 
-\section{Results}
+\section*{Results}
 
@@
 
-\subsection{Structural and species diversity models}
+\subsection*{Structural and species diversity models}
 
@@
 
-\subsection{Variation among vegetation types}
+\subsection*{Variation among vegetation types}
 
@@
 
-\subsection{Moderation of Diversity-AGB relationship by stem density}
+\subsection*{Moderation of Diversity-AGB relationship by stem density}
 
@@
 
-\subsection{Environmental covariates and diversity}
+\subsection*{Environmental covariates and diversity}
 
@@
 
-% \subsection{Moderation effects of moisture availability and soil}
+% \subsection*{Moderation effects of moisture availability and soil}
 % 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability via moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots.
-Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB, with an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{1}).
+In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability, i.e.
+moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots.
+Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}).
 We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).
-Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
+Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
 The strongest effect on AGB was that of stem density.
-Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of an effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{3}).
-We found that vegetation composition in the form of discrete vegetation types affected the relationship between tree species diversity and AGB.
+Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture or energy availability or soil fertility, on AGB (H\textsubscript{3}).
 
-\subsection{Effect of tree species diversity on AGB}
+\subsection*{Inter-related effects of tree species and structural diversity on AGB}
 
-% I would just write that you had a consistent positive effect across all models, as that is your general result with regard to this question.
+We found a consistent positive effect of tree species diversity on AGB across all models in this study.
+Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness and evenness causes higher woody AGB.
+This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem functionality \citep{Liang2016, Cardinale2009}.
+Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of a disturbance-structured  and poorly studied ecological system, southern African woodlands and savannas.
 
-We found a consistent positive total effect of tree species diversity on AGB across all models in this study.
-Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness causes higher woody AGB.
-This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}.
-In addition however, our study provides a novel dissection of the mechanisms underlying this relationship and their interaction with resource availability and vegetation type in southern African woodlands as an understudied biome.
+% In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stem density exists.
+We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
 
-Much of the total variation in AGB was caused by stem density.
+Much of the total variation in AGB was driven by variation in stem density.
 Stem density also correlated with species diversity in our SEMs.
 It is possible that within southern African woodlands a higher species diversity allows for a greater density of tree stems, leading to an increase in total AGB.
-Reverse causation is plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
-In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stem density exists.
-We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
+The opposite is also plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
 We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level.
@@
 
-% There are counter-examples, e.g. Johnson et al. 2016. GCB (from Leeds group, Tim Baker senior author), but feel free to leave text as is for now.
+% KGD: There are counter-examples, e.g. Johnson et al. 2016. GCB (from Leeds group, Tim Baker senior author), but feel free to leave text as is for now.
 
@@
 While we did not explicitly measure Net Primary Productivity (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
-This suggests that as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in southern African woodlands can be maximised by increasing species diversity.
-
-\subsection{Structural diversity as a mechanism for the BEFR}
+This suggests that, as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in southern African woodlands can be maximised by increasing species diversity.
 
 We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity and we therefore find support for our hypothesis (H\textsubscript{2}).
-A higher tree species diversity allows for a greater diversity of tree functional forms within a plot and this may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
+A higher tree species diversity allows for a greater structural diversity of trees, i.e.
+greater variation in DBH and height.
+This may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
+Although we did not measure it here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{}, i.e.
+wider variation in canopy shape and overall growth form; bushy understorey vs.
+emergent canopy, for example.
 Variation in structural diversity may be a joint result of disturbance history and tree species diversity, with highly disturbed plots generally having a less structurally diverse canopy \citep{LaRue2019}.
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
+Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this.
 In frequently disturbed woodlands such as those studied here however, a tree canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
@@
 
-\subsection{Effects of moisture availability and soil fertility}
+We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
+At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness might not make any difference to plot level AGB, accounting for the low and constant effect of tree species diversity on AGB below \textapprox{}180 stems ha\textsuperscript{-1}.
+
+\subsection*{Effects of moisture availability and soil fertility}
 
 Surprisingly, moisture availability and soil fertility had only small effects on AGB compared to that of tree species diversity.
-We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater resource partitioning among individual trees and a greater stem density per unit area \citep{Kraaij2006, Shirima2015}.
+We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for a greater stem density per unit area, greater productivity per unit area and additionally greater tree species diversity due to niche partitioning \citep{Kraaij2006, Shirima2015}.
 
@@
 Additionally, due to the high levels of adaptation of tree species to drought conditions in southern Africa, at the large scale we conducted our experiment turnover in species composition may have prevented a direct relationship being observed between resource availability and AGB.
+% KGD: NEED MORE ON THIS ABOVE, FOR CLARIFICATION OF WHAT YOU MEAN
 
@@
 
+We expected a positive effect of soil fertility on AGB, but found no evidence of this in our models.
+We measured soil fertility using the observed variables of soil organic carbon content, sand particle content and Cation Exchange Capacity (CEC).
+In wet tropical forests a clear relationship has been observed between these variables and AGB \citep{Slik2010, MORE}.
+% KGD: You have written little about soil fertility in this section, but I think you need to.
+There is at least one paper which could be taken to suggest a positive relationship between AGB and soil fertility (Coelho de Souza et al.
+2019.
+NEE, see my publications).
+
 % In a separate set of linear multiple regressions, we found a weak positive interaction effect of moisture availability on the relationship between tree species diversity and AGB.
@@
 
-\subsection{Vegetation type specific responses}
+\subsection*{Vegetation type specific responses}
 
-% See thoughts above for C3 results.
+% KGD: See thoughts above for C3 results.
 C2 should be discussed even if we cannot give a good reason.
@@
 
-Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect via structural diversity remained positive in these vegetation types.
-Compared to Baikiaea and Mopane woodlands these woodlands have higher median tree species richness.
-Baikiaea and Mopane woodlands are also dominated by fewer tree species, notably \textit{Baikiaea plurijuga} in Baikiaea woodlands and \textit{Colophospermum mopane} in Mopane woodlands which tend to produce large canopy dominating trees.
+Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via  structural diversity, remained positive in these vegetation types.
+Compared to Baikiaea and Mopane woodlands, miombo woodlands have higher median tree species richness.
+Baikiaea and Mopane woodlands are also dominated by fewer tree species, notably \textit{Baikiaea plurijuga} in Baikiaea woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees.
 We postulate that this negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in Baikiaea and Mopane woodlands, where the woodland canopy is dominated often by a single species.
@@
 Again, these highly diverse plots in miombo woodlands may be the result of disturbance which can promote a mosaic of woodland of different successional stages and stem densities.
+Alternatively, this small negative direct effect may be an artefact of particularly noisy data, especially given that the overall effect of diversity on AGB is positive.
 
@@
 Baikiaea woodland had the strongest total effect of species diversity on AGB.
-Baikiaea also has relatively low median species richness compared to miombo and Mopane woodlands, but the addition of new species appears to make a larger difference to the AGB of these plots than mopane.
+Baikiaea also has relatively low median species richness compared to miombo, but the addition of new species appears to make a larger difference to the AGB of these plots than in mopane woodlands.
 We suggest that this is due mostly to the particular identity of species found in Baikiaea woodlands and their contribution to ecosystem functioning.
@@
 
-\subsection{Stem density effects}
-
-We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
-At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness might not make any difference to plot level AGB, accounting for the low and constant effect of tree species diversity on AGB below \textapprox{}180 stems ha\textsuperscript{-1}.
-
-At high stem densities, where the woodland approaches a more forest-like canopy structure an increase in niche complementarity in canopy occupation may allow a higher AGB per unit area.
-There may be an effect whereby high stem density plots have many small stems which do not hold much woody biomass, but we did not observe this in our study, possibly because we only considered stems >10 cm DBH.
-An abundance of small stems may prevent the growth of large trees which hold the majority of the AGB in a plot.
-
-% Indeed, we need to address this point, as it is surprising.
-If we were to go to the wet forest literature, we would say that high richness dilutes specialist natural enemies (and many of natural enemies in wet forests do seem to specialise on specific tree species).
-Are natural enemies in woodlands species specialists?
-I have no idea.
-In tropical wet forests, insect herbivores remove way more biomass than vertebrate herbivores, and insects are more likeliy to specialise on a specific tree species than an herbivore.
-But, I had the vague idea that vertebrates (namely mammals) remove more plant biomass than insects in woodlands, but I could be completely wrong.
-In any case, that could be a mechanism that explains species richness effect holding strong, even as the effect of structural diversity disappears.
-
-% At very high biomass, tree species diversity is low because it is associated with intense competition for light leading to dominance of usually a single species (Huston 1997, Oindo and Skidmore 2002).
-
-% \subsection{Future work}
-
-% We hope that this study prompts further investigation of the relationship between woody productuivity and species richness and composition in southern African woodlands.
-While AGB is correlated with NPP,...
-
-\subsection{Conclusion}
+\subsection*{Conclusion}
 
@@
 Additionally, we found that much of this effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of woodland tree canopies.
-We found that the multiple vegetation types which comprise southern African woodlands exhibit variation in the strength of the relationship between species diversity and woody biomass.
+We found that the multiple vegetation types which comprise southern African woodlands exhibit variation in the strength of the relationship between species diversity and woody biomass, inferring that models of regional and global biodiversity-ecosystem function relationships could benefit from including vegetation type terms and the structural properties of those vegetation types, such as structural diversity and stem density.
 In contrast to previous studies, we found that across the region, the direct effects of moisture availability and soil fertility on woody biomass were negligible, with most of their effect being indirectly through species and structural diversity.
@@
 
-\bibliography{/Users/johngodlee/google_drive/bib/papers/lib.bib}
+Southern African woodlands are relied heavily upon for their ecosystem service provision, which is itself affect by ecosystem functionality.
+Resource extraction by humans in southern Africa is directly influencing biodiversity via selective tree-felling for timber, among other forest products.
+Our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stem density and Baikiaea woodlands, which are predominantly targeted for tree felling.
+This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.
+
+% KGD: these conclusions are a bit dry and repetitive on earlier text.
+What do our results mean for woodland management?
+for conservation?
+for our understanding of basic ecological theory?
+As an aside, we have circled back around to niche complementarity in this discussion, but not really to selection effects or facilitation.
+% Southern African woodlands are relied upon heavily for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
+This has raised interest in how biodiversity influences ecosystem function in these ecosystems.
+Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}.
+Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
+While rapid biodiversity change is being observed in southern African woodlands \citep{Syampungani2009}, research into the relationship between biodiversity and ecosystem functionality remains scarce.
+
+\bibliography{lib.bib}
 

### 2020-01-17 build.sh

@@ -10,9 +10,9 @@ INC="manuscript/include/"
 # Run data compilation
 # Rscript scripts/temp_precip.R
 # Rscript scripts/soil_carbon.R
-# Rscript scripts/data_clean.R
-# Rscript scripts/standardise.R
-# Rscript scripts/composition.R
+Rscript scripts/data_clean.R
+Rscript scripts/standardise.R
+Rscript scripts/composition.R
 Rscript scripts/descrip.R
 Rscript scripts/sem.R
 

### 2020-04-07 chapter/regional_befr.tex
Added some comments

--- 2019-11-26 manuscript/chapter_1.tex
@@
 
-We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
+We considered only trees and shrubs in our calculations of AGB, including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
 Only stems >5 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.

### 2020-04-07 cover_letter/cover_letter.tex
Added some comments

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Thin Formal Letter
% LaTeX Template
% Version 1.11 (8/12/12)
% Compile with XeLaTeX

%----------------------------------------------------------------------------------------
%	DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[a4paper]{letter}

\usepackage[margin=2.5cm]{geometry}  %set margin width all around page

\addtolength{\voffset}{0in} % To push the letter further up on the page

\longindentation=0pt % Un-commenting this line will push the closing "Sincerely," to the left of the page

%----------------------------------------------------------------------------------------
%	 NAME & ADDRESS
%----------------------------------------------------------------------------------------

\signature{John L. Godlee} % Your name for the signature at the bottom

\address{Crew Building \\ Alexander Crum Brown Road \\ The King's Buildings \\ Edinburgh \\ EH9 3FF \\ United Kingdom \\ \\ johngodlee@gmail.com} % Your address and phone number

%----------------------------------------------------------------------------------------

\begin{document}

% Cover letter up to 250 words why the paper is of interest to readers of the journal

%----------------------------------------------------------------------------------------
%	ADDRESSEE 
%----------------------------------------------------------------------------------------

\begin{letter}{} % Name/title of the addressee

%----------------------------------------------------------------------------------------
%	LETTER CONTENT 
%----------------------------------------------------------------------------------------

	\opening{\textbf{Dear ,}}

I enclose a manuscript for submission to

We thank you for your consideration.

\vspace{\parskip} % Extra whitespace for aesthetics
\closing{Yours,}
\vspace{2\parskip} % Extra whitespace for aesthetics

% \ps{P.S. You can find additional information attached to this letter.} % Postscript text, comment this line to remove it

% \encl{Copyright permission form} % Enclosures with the letter, comment this line to remove it

%----------------------------------------------------------------------------------------

\end{letter}

\end{document}

### 2020-04-07 manuscript/regional_befr_manuscript.tex
Added some comments

--- 2020-01-17 manuscript/manuscript.tex
@@
 \usepackage{pdflscape}
+\usepackage[document]{ragged2e}
 
@@
 \newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}   % \todo{NOTE TO SELF WRITTEN IN RED}
+
+\usepackage[revision]{revdiff}
 
@@
 
+This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (J.
+L.
+Godlee, Grant No.
+NE/L002558/1).
+The data provided for this study was constributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands), an activity of the Miombo Network and NERC-funded project (Grant No.
+NE/P008755/1).
+We thank all data providers and the field assistance they received when collecting plot data.
+
 \section*{Biosketch}
+
+SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}) aims to understand the response of southern African woodlands to global change.
+The goal of SEOSAW is to produce novel analyses of the determinants of ecosystem structure and function for the southern Africa region, based on syntheses of plot data.
+Additionally the group hopes to develop infrastructure for a long-term regional plan for plot remeasurement in the southern African region.
+While working on a multitude of diverse projects in the dry tropics at large, all authors have a broad interest in community ecology and ecosystem assemblage in southern African woodlands.
+
 \newpage{}
@@
 
-\normalsize{Running title: Woodland functionality in southern Africa}
+\normalsize{Running title: Ecosystem function in southern African woodlands}
 
@@
 
-\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship in highly disturbed and ecophysiologically stressful systems is less clear.
-Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density and resource availability, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance and ecophysiological stress, is less clear}.
+Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{and disturbance through fire}, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
 
@@
 
-\textbf{Methods:} We used a network of \nplots{} savanna and woodland plots located across the southern African sub-continent in which each tree >5 cm diameter was measured.
-We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density.
+\textbf{Methods:} We used a network of \nplots{} savanna/woodland plots located across the southern African sub-continent in which each tree >10 cm diameter was measured \rnew{and identified to species level}.
+We used structural equation modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, accounting for the interactive effects of resource availability and along a gradient of stem density.
 
-\textbf{Results:} A positive effect of tree species diversity on biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity.
+\textbf{Results:} A positive effect of tree species diversity on aboveground biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity.
 We also found that the effect of tree species diversity on biomass increases with stem density.
@@
 
-\textbf{Main conclusions:} Tree species diversity plays an important role in determining the ecosystem functionality and structure of southern African woodlands.
-Our results demonstrate the importance of considering environmental conditions and woodland structural attributes to more accurately describe the relationship between biodiversity and ecosystem functionality.
+\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and function of highly disturbed southern African savannas and woodlands.
+Our results demonstrate the importance of including environmental conditions and vegetation type in models to accurately describe a general relationship between biodiversity and ecosystem function at a regional level.
+Biodiversity loss predominantly by human actions in southern Africa may have detrimental outcomes for ecosystem function, particularly in species poor Baikiaea woodlands, which showed the strongest biodiversity - ecosystem function relationship.
 
@@
 
+Scientific interest in the relationship between species richness and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}.
 Numerous studies have shown relationships between biodiversity and ecosystem function (e.g.
-\citealt{Liang2016, Cardinale2009}).
-The strength and direction of these observed Biodiversity - Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem functionality \citep{Liang2016}.
+\citealt{Liang2016, Hooper2012, Cardinale2009}).
+The strength and direction of these observed Biodiversity-Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}.
 Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
@@
 This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.
+The frequently reported and intuitive relationship between biodiversity and ecosystem function invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.
 
-Compared to other forested ecosystems, dry tropical woodlands are highly structured by disturbance, mainly through fire, and in Africa notably by herbivory also \citep{}.
+Compared to \rchange{other forested}{forest} ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire \rchange{and in Africa notably by herbivory also}{and herbivory, with African savannas possessing large herbivores absent from other savannas} \citep{Sankaran2008, Levick2009}.
+\rnew{Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}}.
 High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
-This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland ecosystems.
-Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem functionality \citep{Lasky2014, Tobner2016}.
-Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations in environments which are more hostile to growth.
-In European forests \citep{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
+This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland/savanna ecosystems.
+Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem function\rnew{s} \citep{Lasky2014, Tobner2016}.
+Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations \rchange{in environments which are more hostile to growth}{under limiting environmental conditions such as low water availability \citep{} or high maximum temperature \citep{}}.
+Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
 They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function.
-This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation.
-
-% COME BACK TO THIS: In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}.
-It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at large spatial scales.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that mean annual precipitation (MAP) sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
-Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.
+This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation in order to characterise a generalisable BEFR.
 
@@
 \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}.
-A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and it's environmental determinants.
-
-Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (\autoref{clust_map}).
+A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants.
+Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1983, Ratnam2011, Ryan2016} (\autoref{clust_map}).
 The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
 Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species.
-Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
-Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and functionality.
+Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and function.
 
-A small number of studies in southern African savannas and woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
-However, studied conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
-Small scale studies may therefore be over-estimating the effect of diversity.
-Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass \citep{}, it is important to sample across geographic and environmental gradients to be able to infer a truly regional scale relationship between diversity and biomass.
+A small number of studies in SAWs, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+The results of these fine scale studies concur with similar studies in other biomes \citep{}.
+Studies of the BEFR often find that at fine scales, diversity shows a strong effect on ecosystem function, but at broad scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
+Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between diversity and biomass.
 
@@
 
-\citet{Solbrig1996} writes that southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species.
+In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}.
+It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at broad spatial scales.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that \rnew{below a threshold of \textapprox{}650 mm MAP}, precipitation sets sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
+Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.
+
+\citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species.
 This structural diversity may be one mechanism through which diversity influences woody biomass.
-\citep{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest.
+\citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest.
 They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production.
@@
 
-% High levels of disturbance in southern African woodlands may moderate the observable BEFR through it's effect on ecosystem composition.
+High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition.
 Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}.
-A dominant and competitive species which tends to produce many small fast-growing stems is more likely to experience mortality during a fire as the higher surface area to volume ratio of the trunk increases the likelihood of combustion.
-These fast-growing bushy species often proliferate in southern African woodlands when fire is deliberately excluded, leading to the growth of thicket vegetation which precludes the growth of high biomass tree species \citep{Higgins2007}.
-Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
-% KGD: This text is useful in demonstrating your biological understanding of the ecosystem.
-I think it could be useful to include some of this biological knowledge in the paragraph above that ends with the statements on disturbance (the paragraph starting with Slobrig 1996).
-Or you could include this whole paragraph just after that one.
-The co-authors will reign us in if this kind of logic is not sensible for the ecosystems we are studying.
+Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree.
+Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands.
+If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high AGB under an intense disturbance regime.
 
-In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
-We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
+In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
+We used aboveground woody biomass of trees and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
 We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality and assess the generality of our results.
-We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, which may interact their effect on ecosystem structure and therefore biomass.
-Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
+We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species diversity affected ecosystem function and assess the generality of our results.
+We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass.
+Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strenghth of the effect} of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
 
@@
 
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in so-called miombo woodlands (\autoref{clust_map}, \citealt{White1983}).
-The study region spans a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
-The study sites span the core climate space of the region as a whole.
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland area as defined by \citet{White1983}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the \rchange{so-called miombo woodland eco-region}{miombo ecoregion} (\autoref{clust_map}, \citealt{White1983}).
+The study \rchange{region spans}{area spacs the core climate space of the region}, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \rold{as defined by} \citep{White1983}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
 
 Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup.
-Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature \citep{Ratnam2011, Hill2010}.
+Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands (SAWs) \citep{Ratnam2011, Hill2010}.
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
@@
 
-Many plots provided by the Zambian Forestry Commission were arranged in clusters of up to four 20x50 m plots, 20 metres apart.
+Many plots provided by the 2005-2008 Zambian Integrated Land Use Assessment \citep{Mukosha2009} were arranged in clusters of up to four 20x50 m plots, 20 metres apart.
 Plots within each cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.
@@
 Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}.
-Values of Mean Annual Temperature (MAT) and Mean Annual Precipitation (MAP) were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
+Values of \rchange{Mean Annual Temperature (MAT)}{MAT} and \rchange{Mean Annual Precipitation (MAP)}{MAP} were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
 Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily MAT and MAP, respectively, for each of the 18 years of available data.
-Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot \citep{Hengl2017}.
+Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}.
 We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org.
@@
 
-To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated:
+To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated as the ratio of estimated Shannon diversity to the natural log of estimated species richness.
+Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits.
+We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).
 
-\begin{equation}
-	\begin{gathered}
-		E_{H'} = \frac{H'_{e}}{\ln{S}} \\
-	\end{gathered}
-	\label{shannon_equit}
-\end{equation}
-
-Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot, using the \verb|ChaoRichness()| function.
-We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).
+% \begin{equation}
+% 	\begin{gathered}
+% 		E_{H'} = \frac{H'_{e}}{\ln{S}} \\
+% 	\end{gathered}
+% 	\label{shannon_equit}
+% \end{equation}
+% 
+% Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot, using the \verb|ChaoRichness()| function.
 
@@
 Plots were assigned to vegetation type groups based on tree species composition.
-Groups were identified in \citet{Fayolle2018} in an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
-Groups were identified using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes.
+Groups were defined in \citet{Fayolle2018} in an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
+Group identification was conduscted using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes.
 Plot data used in this study occurred in four vegetation type groups.
@@
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
-A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regression.
+A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regressions.
 Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset.
@@
 We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, while ensuring no factor loadings were in excess of 1.
-We found no significant difference between model specifications.
+We found no significant difference between model specifications (p>0.05).
 Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.
 
-First, we assessed the role of structural diversity and species diversity in determining AGB.
-We constructed a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
+We assessed the role of structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
 To account for variation in stem density which may covary with species diversity we also included it as an predictor in our model.
@@
 To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density.
-We used each of these datasets to fit the model including only tree species and structural diversity latent variables to predict AGB.
-We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model.
+We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB.
+We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model as we were deliberately controlling stem density in our subsampling.
 We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled dataset.
 
-Second, we incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density.
+We incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density.
 We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB.
@@
 The strongest total effect of tree species diversity on AGB was in Baikiaea woodland (\struccsb{}), which was species rich but highly variable in species diversity compared to other vegetation types (\autoref{clust_summ}).
-The R\textsuperscript{2} of AGB was highest in Marginal miombo (R\textsuperscript{2} = \strucarsq{}) and lowest in the Core miombo (R\textsuperscript{2} = \strucbrsq{}).
+The R\textsuperscript{2} of AGB was highest in Marginal miombo (R\textsuperscript{2} = \strucarsq{}) and lowest in the Core miombo (R\textsuperscript{2} = \\strucbrsq{}strucbrsq{}).
 
@@
 
-We repeatedly sub-sampled the plot dataset to build \subn{} datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
+\rchange{We repeatedly sub-sampled the plot dataset to build \subn{} datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
 Each dataset consisted of approximately \subp{} plots with overlap of plot identity between subsampled datasets.
-\autoref{sem_struc_stems_ha} shows a positive effect of tree species diversity on AGB as stem density increases.
+\autoref{sem_struc_stems_ha} shows a positive effect of tree species diversity on AGB as stem density increases.}{In our sub-sampling of the plot dataset by mean stem density, we found a positive effect of tree species diversity on AGB as stem density increases (\autoref{sem_struc_stems_ha})}.
 There appears to be a minimum stem density threshold at \textapprox{}180 stems ha\textsuperscript{-1} below which there appears to be a reasonably constant low baseline effect of tree diversity on biomass.
@@
 Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
-The observed variables which inform the unmeasured latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
+Observed variables which inform the latent variables are connected by dotted arrows, observed variables with loading set to 1 are connected by dashed arrows.
 Correlations between variables are depicted as dotted curved arrows.
@@
 In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability, i.e.
-moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots.
+moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across SAWs, using a network of \nplots{} woodland survey plots.
 Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}).
 We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).
-Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
+Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}).
 The strongest effect on AGB was that of stem density.
-Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture or energy availability or soil fertility, on AGB (H\textsubscript{3}).
+Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture or soil fertility, on AGB (H\textsubscript{3}).
 
@@
 We found a consistent positive effect of tree species diversity on AGB across all models in this study.
-Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness and evenness causes higher woody AGB.
-This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem functionality \citep{Liang2016, Cardinale2009}.
-Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of a disturbance-structured  and poorly studied ecological system, southern African woodlands and savannas.
+Within SAWs we therefore find support that higher tree species richness and evenness causes higher woody AGB.
+This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem function \citep{Liang2016, Cardinale2009}.
+Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of SAWs, a disturbance-structured and poorly studied ecological system.
 
@@
 Stem density also correlated with species diversity in our SEMs.
-It is possible that within southern African woodlands a higher species diversity allows for a greater density of tree stems, leading to an increase in total AGB.
+It is possible that within SAWs a higher species diversity allows for a greater density of tree stems, leading to an increase in total AGB.
 The opposite is also plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
-We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level.
-This is supported by the moderately strong positive effect of tree species diversity on AGB via structural diversity.
+We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level \citep{}.
+This is supported by the moderately strong indirect positive effect of tree species diversity on AGB via structural diversity.
 
@@
 
-While we did not explicitly measure Net Primary Productivity (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
-This suggests that, as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in southern African woodlands can be maximised by increasing species diversity.
+% While we did not explicitly measure Net Primary Productivity (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
+This suggests that, as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in SAWs can be maximised by increasing species diversity.
 
@@
 Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this.
-In frequently disturbed woodlands such as those studied here however, a tree canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
-While we did not have access to adequate data on disturbance history in our plots, previous studies have found that southern African woodlands with higher species diversity tend to be less disturbed and tend to form a more closed canopy \citep{Chidumayo2013, Mutowo2012}.
+In frequently disturbed woodlands such as those studied here however, a woodland canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
+While we did not have access to adequate data on disturbance history in our plots, previous studies have found that SAWs with higher species diversity tend to be less disturbed and tend to form a more closed canopy \citep{Chidumayo2013, Mutowo2012}.
 
@@
 Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via increasing tree species diversity and via increasing stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}.
-In this study, while we observed weak indirect effects via species diversity, we saw no evidence for a direct effect on AGB.
-Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in southern African woodlands, which are frequently droughted.
+In this study, while we observed weak indirect effects via species diversity, we saw no evidence for a direct effect of moisture availability on AGB.
+Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in SAWs, which are frequently droughted.
 It is possible that the range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture variation in AGB.
-We deliberately excluded plots with very low stem density as they are not considered woodlands, but grassy savannas.
-It may be that by excluding the bottom end of this stem density continuum we prevented a relationship being observed between moisture and AGB/stem density.
-Additionally, due to the high levels of adaptation of tree species to drought conditions in southern Africa, at the large scale we conducted our experiment turnover in species composition may have prevented a direct relationship being observed between resource availability and AGB.
-% KGD: NEED MORE ON THIS ABOVE, FOR CLARIFICATION OF WHAT YOU MEAN
+Due to the high levels of adaptation of tree species to drought conditions in southern Africa, at the large scale we conducted our experiment turnover in species composition along the moisture gradient may have obscured a direct relationship being observed between moisture availability and AGB.
 
-In southern African woodlands moisture availability is closely linked with the intensity of disturbance from seasonal fires.
+In SAWs moisture availability is closely linked with the intensity of disturbance from seasonal fires.
 The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth \citep{Charles-Dominique2018}, and may also limit species diversity \citep{Linder2014}.
@@
 
-In this study we found that across southern African woodlands, there is a generalisable positive association between tree species diversity and woody biomass as a measure of ecosystem functionality.
+In this study we found that across southern African woodlands (SAWs), there is a generalisable positive association between tree species diversity and woody biomass as a measure of ecosystem function.
 Additionally, we found that much of this effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of woodland tree canopies.
-We found that the multiple vegetation types which comprise southern African woodlands exhibit variation in the strength of the relationship between species diversity and woody biomass, inferring that models of regional and global biodiversity-ecosystem function relationships could benefit from including vegetation type terms and the structural properties of those vegetation types, such as structural diversity and stem density.
+We found that the multiple vegetation types which comprise SAWs exhibit variation in the strength of the relationship between species diversity and woody biomass, inferring that models of regional and global biodiversity-ecosystem function relationships could benefit from including vegetation type terms and the structural properties of those vegetation types, such as structural diversity and stem density.
 In contrast to previous studies, we found that across the region, the direct effects of moisture availability and soil fertility on woody biomass were negligible, with most of their effect being indirectly through species and structural diversity.
-A gap in available data means that we could not incorporate disturbance history into our models adequately, but this factor likely plays a large part in the association between species diversity and woody biomass in southern African woodlands.
+A gap in available data means that we could not incorporate disturbance history into our models adequately, but this factor likely plays a large part in the association between species diversity and woody biomass in SAWs.
 
-Southern African woodlands are relied heavily upon for their ecosystem service provision, which is itself affect by ecosystem functionality.
+SAWs are relied heavily upon for their ecosystem service provision, which is itself affect by ecosystem function.
 Resource extraction by humans in southern Africa is directly influencing biodiversity via selective tree-felling for timber, among other forest products.
@@
 \appendix{}
+
+\section*{Data accessibility statement}
+
+\section*{Tables}
+
+\section*{Figure legends and embedded figures}
 

### 2020-04-07 author_comments/data_contributors.R

@@ -1,0 +1,34 @@
+# Who is a data contributor?
+# John Godlee (johngodlee@gmail.com)
+# 2020-02-07
+
+# Preamble ----
+
+# Remove old crap 
+rm(list = ls())
+# dev.off()
+
+# Set working directory
+
+# Packages
+library(dplyr)
+
+# Load data ----
+plots <- readRDS("data/plot_data_fil_agg_norm_std_outlier.rds")
+
+# Who is a contributor and for which plots?
+unique(plots$pi)
+
+plots %>%
+  dplyr::select(pi, plotcode) %>%
+  group_by(pi) %>%
+  summarise(plotcode = first(plotcode))
+
+plots %>%
+  filter(pi == "Y") %>%
+  pull(
+##' SHD - belongs to ???
+
+plots %>% 
+  filter(pi == "FRIM")
+##' MCL - belongs to Anderson Muchawona

### 2020-04-07 build.sh

@@ -68,7 +68,7 @@ sed -i -r '12,16s/\w+ \w+\.*/\\textit\{&\}/g' "${INC}clust_summ.tex"
 sed -i '12,16s/spp\.\}/\}spp./g' "${INC}clust_summ.tex"
 sed -i '10s/.*/{Cluster} \& {Dominant species} \& {Indicator species} \& {N plots} \& {Species rich.} \& {Stems ha\\textsuperscript{-1}} \& {AGB (t ha\\textsuperscript{-1})} \\\\/' "${INC}clust_summ.tex"
 sed -i '13,16i\\\hline' "${INC}clust_summ.tex"
-sed -i 's/caption{}/caption{Description of the biogeographical clusters (C1-C5) to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \\citep{Dufrene1997} implemented with \\texttt{indval()} from the \\texttt{labdsv} R package \\citep{labdsv} and represent species which define the given cluster. Dominant species were identified by choosing the species with the largest AGB contribution within each cluster. Numeric values of species richness, stems ha\\textsuperscript{-1} and AGB are medians and interquartile ranges (75th percentile - 25th percentile).}/g' "${INC}clust_summ.tex" 
+sed -i 's/caption{}/caption{Description of the biogeographical clusters to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \\citep{Dufrene1997} implemented with \\texttt{indval()} from the \\texttt{labdsv} R package \\citep{labdsv} and represent species which define the given cluster. Dominant species were identified by choosing the species with the largest AGB contribution within each cluster. Numeric values of species richness, stems ha\\textsuperscript{-1} and AGB are medians and interquartile ranges (75th percentile - 25th percentile).}/g' "${INC}clust_summ.tex" 
 
 ## Full table of pairwise correlations - corr_ci_tab.tex
 sed -i '10s/.*/{X} \& {Y} \& {r} \& {lower 95\\% CI} \& {upper 95\\% CI} \& {n} \& {Prob.} \\\\/' "${INC}corr_ci_tab.tex"

### 2020-04-09 manuscript/regional_befr_manuscript.tex
Went through some comments

@@
 
-\section*{Acknowledgements}
+\section{Acknowledgements}
 
@@
 
-\section*{Biosketch}
+\section{Biosketch}
 
@@
 
-\section*{Abstract}
+\section{Abstract}
 
 \textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance and ecophysiological stress, is less clear}.
-Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{and disturbance through fire}, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{and disturbance through fire}, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
 
@@
 
-\section*{Introduction}
+\section{Introduction}
 
-Scientific interest in the relationship between species richness and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}.
+Scientific interest in the relationship between species diversity and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}.
 Numerous studies have shown relationships between biodiversity and ecosystem function (e.g.
@@
 This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported and intuitive relationship between biodiversity and ecosystem function invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.
+The frequently reported and intuitive relationship between biodiversity and ecosystem function invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater biodiversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
 
@@
 The results of these fine scale studies concur with similar studies in other biomes \citep{}.
-Studies of the BEFR often find that at fine scales, diversity shows a strong effect on ecosystem function, but at broad scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
-Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between diversity and biomass.
+Studies of the BEFR often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
+Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
 
@@
 
-In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}.
-It is important therefore to account for climatic factors and understand how they interact with biomass and diversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at broad spatial scales.
+In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and tree species diversity independently \citep{Spasojevic2014}.
+It is important therefore to account for climatic factors and understand how they interact with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at broad spatial scales.
 \citet{Sankaran2005} used data from 854 African woodland field sites to show that \rnew{below a threshold of \textapprox{}650 mm MAP}, precipitation sets sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
@@
 \citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species.
-This structural diversity may be one mechanism through which diversity influences woody biomass.
+This structural diversity may be one mechanism through which tree species diversity influences woody biomass.
 \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest.
@@
 This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition.
-Instead, disturbance history may override the effects of species diversity on structural diversity nullifying the effects of species diversity on structural diversity.
+Instead, disturbance history may override the effects of tree species diversity on structural diversity nullifying the effects of tree species diversity on structural diversity.
 
@@
 Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands.
-If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high AGB under an intense disturbance regime.
+If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.
 
-In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
+In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
 We used aboveground woody biomass of trees and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
 We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species diversity affected ecosystem function and assess the generality of our results.
+We incorporated vegetation type via clustering of plot level tree species composition, as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and assess the generality of our results.
 We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass.
-Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strenghth of the effect} of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
+Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strenghth of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
 
-\section*{Materials and methods}
+\section{Materials and methods}
 
-\subsection*{Study location}
+\subsection{Study location}
 
@@
 
-\subsection*{Data collection}
+\subsection{Data collection}
 
@@
 Values of \rchange{Mean Annual Temperature (MAT)}{MAT} and \rchange{Mean Annual Precipitation (MAP)}{MAP} were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
-Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily MAT and MAP, respectively, for each of the 18 years of available data.
+Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily mean temperature and precipitation, respectively, for each of the 18 years of available data.
 Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}.
@@
 
-\subsection*{Data analysis}
+\subsection{Data analysis}
+
+\subsubsection{Species diversity and structural diversity metrics}
+
 Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}.
@@
 Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.
-
-To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated as the ratio of estimated Shannon diversity to the natural log of estimated species richness.
+To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness.
 Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits.
@@
 
-\subsubsection*{Vegetation clusters}
+\subsubsection{Vegetation clusters}
 
@@
 
-\subsubsection*{Structural Equation Modelling}
+\subsubsection{Structural Equation Modelling}
 
@@
 A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regressions.
-Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon diversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset.
+Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset.
 Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
@@
 
-We assessed the role of structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
+We assessed the role of tree structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
 To account for variation in stem density which may covary with species diversity we also included it as an predictor in our model.
@@
 
-\section*{Results}
+\section{Results}
 
@@
 
-\subsection*{Structural and species diversity models}
+\subsection{Structural and species diversity models}
 
@@
 
-\subsection*{Variation among vegetation types}
+\subsection{Variation among vegetation types}
 
@@
 
-\subsection*{Moderation of Diversity-AGB relationship by stem density}
+\subsection{Moderation of Diversity-AGB relationship by stem density}
 
@@
 
-\subsection*{Environmental covariates and diversity}
+\subsection{Environmental covariates and tree diversity}
 
-A model incorporating the latent variables of moisture availability and soil fertility showed that the total effect of species diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}).
+A model incorporating the latent variables of moisture availability and soil fertility showed that the total effect of tree species diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}).
 Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}).
@@
 
-% \subsection*{Moderation effects of moisture availability and soil}
+% \subsection{Moderation effects of moisture availability and soil}
 % 
@@
 
-\section*{Discussion}
+\section{Discussion}
 
@@
 
-In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability, i.e.
+In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, i.e.
 moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across SAWs, using a network of \nplots{} woodland survey plots.
@@
 
-\subsection*{Inter-related effects of tree species and structural diversity on AGB}
+\subsection{Inter-related effects of tree species and structural diversity on AGB}
 
@@
 Within SAWs we therefore find support that higher tree species richness and evenness causes higher woody AGB.
-This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem function \citep{Liang2016, Cardinale2009}.
+This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
 Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of SAWs, a disturbance-structured and poorly studied ecological system.
@@
 
-\subsection*{Effects of moisture availability and soil fertility}
+\subsection{Effects of moisture availability and soil fertility}
 
@@
 
-\subsection*{Vegetation type specific responses}
+\subsection{Vegetation type specific responses}
 
@@
 Again, these highly diverse plots in miombo woodlands may be the result of disturbance which can promote a mosaic of woodland of different successional stages and stem densities.
-Alternatively, this small negative direct effect may be an artefact of particularly noisy data, especially given that the overall effect of diversity on AGB is positive.
+Alternatively, this small negative direct effect may be an artefact of particularly noisy data, especially given that the overall effect of tree diversity on AGB is positive.
 
@@
 
-\subsection*{Conclusion}
+\subsection{Conclusion}
 
@@
 
-\section*{Data accessibility statement}
+\section{Data accessibility statement}
 
-\section*{Tables}
+\section{Tables}
 
-\section*{Figure legends and embedded figures}
+\section{Figure legends and embedded figures}
 
-\section*{Appendix 1 - Data cleaning process} \label{appendixa}
+\section{Appendix 1 - Data cleaning process} \label{appendixa}
 
@@
 
-\section*{Appendix 2 - Estimation of DBH via tree taper} \label{appendixb}
+\section{Appendix 2 - Estimation of DBH via tree taper} \label{appendixb}
 
@@
 
-\section*{Appendix 3 - Frequency distribution of observed variables} \label{appendixc}
+\section{Appendix 3 - Frequency distribution of observed variables} \label{appendixc}
 
@@
 
-\section*{Appendix 4 - Table of correlation fit statistics} \label{appendixd}
+\section{Appendix 4 - Table of correlation fit statistics} \label{appendixd}
 

### 2020-06-22 manuscript/regional_befr_manuscript.tex
Updated comments

@@
 
-\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance and ecophysiological stress, is less clear}.
+\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}.
 Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{and disturbance through fire}, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
@@
 \textbf{Methods:} We used a network of \nplots{} savanna/woodland plots located across the southern African sub-continent in which each tree >10 cm diameter was measured \rnew{and identified to species level}.
-We used structural equation modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, accounting for the interactive effects of resource availability and along a gradient of stem density.
+We used structural equation modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, accounting for the interactive effects of resource availability \rnew{and disturbance by fire} along a gradient of stem density.
 
@@
 Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggest that in dry ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function.
-This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation in order to characterise a generalisable BEFR.
+They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition dminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation in order to characterise a generalisable BEFR.
 

### 2020-06-22 build.sh

@@ -68,7 +68,7 @@ sed -i -r '12,16s/\w+ \w+\.*/\\textit\{&\}/g' "${INC}clust_summ.tex"
 sed -i '12,16s/spp\.\}/\}spp./g' "${INC}clust_summ.tex"
 sed -i '10s/.*/{Cluster} \& {Dominant species} \& {Indicator species} \& {N plots} \& {Species rich.} \& {Stems ha\\textsuperscript{-1}} \& {AGB (t ha\\textsuperscript{-1})} \\\\/' "${INC}clust_summ.tex"
 sed -i '13,16i\\\hline' "${INC}clust_summ.tex"
-sed -i 's/caption{}/caption{Description of the biogeographical clusters to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \\citep{Dufrene1997} implemented with \\texttt{indval()} from the \\texttt{labdsv} R package \\citep{labdsv} and represent species which define the given cluster. Dominant species were identified by choosing the species with the largest AGB contribution within each cluster. Numeric values of species richness, stems ha\\textsuperscript{-1} and AGB are medians and interquartile ranges (75th percentile - 25th percentile).}/g' "${INC}clust_summ.tex" 
+sed -i 's/caption{}/caption{Description of the biogeographical clusters to which each plot in the study was assigned. Indicator species were generated using Dufrene-Legendre indicator species analysis \\citep{Dufrene1997} implemented with \\texttt{indval()} from the \\texttt{labdsv} R package \\citep{labdsv} and represent species which define the given cluster. Dominant species were identified by choosing the species with the largest mean plot level proportional AGB contribution within each cluster. Numeric values of species richness, stems ha\\textsuperscript{-1} and AGB are medians and interquartile ranges (75th percentile - 25th percentile).}/g' "${INC}clust_summ.tex" 
 
 ## Full table of pairwise correlations - corr_ci_tab.tex
 sed -i '10s/.*/{X} \& {Y} \& {r} \& {lower 95\\% CI} \& {upper 95\\% CI} \& {n} \& {Prob.} \\\\/' "${INC}corr_ci_tab.tex"

### 2020-06-22 scripts/data_clean.R

@@ -197,12 +197,28 @@ s_fil_summ <- s_fil_all_other %>%
     cov_dbh = sd_dbh / mean_dbh * 100)
 
 # Remove tiny tree-less plots ----
+
+# How many plots with various stems/ha
+thresh_df <- data.frame(stems_ha = seq(from = 0, to = 500, by = 10))
+
+thresh_df$n_plots <- sapply(thresh_df$stems_ha, function(x){
+  left_join(plot_data_agg, s_fil_summ, 
+  by = c("plot_group" = "plot_group")) %>%
+  mutate(stems_ha = n_stems / area_plot,
+    agb_ha = agb / area_plot) %>%
+  filter(area_plot >= 0.1,
+    stems_ha >= x) %>%
+  filter(plot_group != "DKS001") %>%
+  group_by(plot_group) %>%
+  n_distinct()
+})
+
 plot_data_agg <- left_join(plot_data_agg, s_fil_summ, 
   by = c("plot_group" = "plot_group")) %>%
   mutate(stems_ha = n_stems / area_plot,
     agb_ha = agb / area_plot) %>%
   filter(area_plot >= 0.1,
-    stems_ha >= 10) %>%
+    stems_ha >= 50) %>%
   filter(plot_group != "DKS001")
 
 # Estimate rarefied species diversity ----

### 2020-06-24 include/clust_summ.tex
Cleaned gitignore, analyses with seosaw v2.6

--- 2019-12-16 include/clust_summ.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Dec 16, 2019 - 11:33:55
+% Date and time: Wed, Jun 24, 2020 - 12:22:31
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-1 & Julbernadia spp., Brachystegia spiciformis, Baikeaea plurijuga & Diplorhynchus condylocarpon, Burkea africana, Pseudolachnostylis maprouneifolia & 687 & 11(11.2) & 170(145.3) & 37(34.68) \\ 
-2 & Julbernadia spp., Brachystegia spp., Isoberlinia angolensis & Julbernardia paniculata, Isoberlinia angolensis, Brachystegia longifolia & 757 & 18(17.6) & 215(171.7) & 48.8(43.7) \\ 
-3 & Spirostachys africana, Senegalia spp., Euclea racemosa & Baikiaea plurijuga, Senegalia ataxacantha, Combretum collinum & 226 & 10(10) & 165(157.5) & 46(47.81) \\ 
-4 & Colophospermum mopane & Colophospermum mopane, Combretum spp. & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\ 
+Acacia shrubland & Baikiaea plurijuga, Burkea africana, Pterocarpus angolensis & Burkea africana, Baikiaea plurijuga, Pterocarpus angolensis & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\ 
+Core miombo & Brachystegia spiciformis, Julbernardia paniculata, Brachystegia boehmii & Parinari curatellifolia, Uapaca kirkiana, Brachystegia spiciformis & 523 & 20(16.9) & 204(142.5) & 44.2(36.11) \\
+Marginal miombo & Spirostachys africana, Senegalia burkei, Senegalia nigrescens & Euclea racemosa, Vachellia nilotica, Spirostachys africana & 466 & 12(13.7) & 178(129.5) & 36.9(26.98) \\ 
+Mopane & Colophospermum mopane, Androstachys johnsonii, Kirkia acuminata & Colophospermum mopane, Psuedolachnostylis maprouneifolia, Lannea discolor & 58 & 10(10.2) & 186(125.6) & 42.7(32.83) \\ 
 \hline \\[-1.8ex] 

### 2020-06-24 build.sh

@@ -8,11 +8,9 @@ INC="manuscript/include/"
 # Remove intermediate objects
 
 # Run data compilation
-# Rscript scripts/temp_precip.R
-# Rscript scripts/soil_carbon.R
+Rscript scripts/temp_precip.R
 Rscript scripts/data_clean.R
 Rscript scripts/standardise.R
-Rscript scripts/composition.R
 Rscript scripts/descrip.R
 Rscript scripts/sem.R
 

### 2020-06-24 scripts/sem.R

@@ -17,7 +17,7 @@
 # Preamble ----
 
 # Source files
-# source("sem_data_collation.R")
+# source("dat_collation.R")
 # source("bivariate_relationships.R")
 
 # Remove old crap
@@ -48,29 +48,28 @@ source("scripts/clust_defin.R")
 
 # Import data ----
 
-# Aggregated and filtered data
+dat <- readRDS("data/plot_data_fil_agg_norm_std.rds")
 
-sem_data <- readRDS("data/plot_data_fil_agg_norm_std_outlier.rds")
 
-##' Species richness as determinant of large trees 
-##' (measured as mean of 95th percentile of height) biomass
+# Reverse some variables ----
+dat <- dat %>%
+  mutate(
+    precip_seas_rev_std = -1 * precip_seas_std,
+    temp_seas_rev_std = -1 * temp_seas_std,
+    sand_rev_std = -1 * sand_std,
+    temp_rev_std = -1 * temp_std,
+    shannon_equit_rev_std = -1 * shannon_equit_std)
+
 
 # Correlation matrix between variables ----
 
-# Unreverse some variables
-sem_data$precip_seasonality_std <- -sem_data$precip_seasonality_rev_std
-sem_data$temp_seasonality_std <- -sem_data$temp_seasonality_rev_std
-sem_data$sand_per_std <- -sem_data$sand_per_rev_std
-sem_data$mean_temp_std <- -sem_data$mean_temp_rev_std
-sem_data$shannon_equit_rev_std <- sem_data$shannon_equit_std 
-
 # Create a dataframe 
-corr_df <- sem_data %>%
-  dplyr::select(sand_per_std, ocdens_std, cec_std, 
-    total_precip_std, precip_seasonality_std,
-    mean_temp_std, temp_seasonality_std,
+corr_df <- dat %>%
+  dplyr::select(sand_std, soil_c_log_std, cec_std, 
+    precip_std, precip_seas_std,
+    temp_std, temp_seas_std,
     n_species_raref_log_std, shannon_equit_std,
-    cov_height_std, cov_dbh_std, stems_ha_log_std, agb_ha_log_std)
+    cov_height_std, cov_dbh_std, n_trees_gt10_ha_log_std, agb_ha_log_std)
 
 # Make a dataframe of labels
 label_df <- data.frame(lab = names(corr_df),
[…]

### 2020-06-24 scripts/temp_precip.R

@@ -1,125 +1,82 @@
-# Generate a dataset of precipitation and temperature means across the Miombo region
+# Generate dataset of precip and temp means across miombo ecoregion
 # John Godlee (johngodlee@gmail.com)
-# 2019_10_02
+# 2019-10-02
+# 2020-06-22
 
+# Clean env.
 rm(list = ls())
 
 # Packages
 library(raster)
 library(rgdal)
 
-# Get Miombo shape
-white_veg <- readOGR(dsn="data/whitesveg", 
-  layer="Whites vegetation")
+# Get Miombo outline ---- 
+white_veg <- readOGR(dsn="/Volumes/john/whitesveg", layer="")
 
-white_veg_miombo <- white_veg[!is.na(white_veg@data$DESCRIPTIO) & white_veg@data$DESCRIPTIO == "Moist-infertile savanna" , ]
+# Subset to White's miombo 
+white_veg_miombo <- white_veg[!is.na(white_veg@data$DESCRIPTIO) & 
+  white_veg@data$DESCRIPTIO == "Moist-infertile savanna", ]
 
+# Get southern part only
 white_veg_miombo_south <- crop(white_veg_miombo, extent(10, 40, -30, 0))
 
-load("data/seosaw_plot_summary5Apr2019.Rdata")
+# Fill holes
+white_veg_miombo_south_fill <- SpatialPolygons(list(Polygons(
+      Filter(function(f) { 
+        f@ringDir==1
+      }, white_veg_miombo_south@polygons[[1]]@Polygons), ID = 1)))
 
-plot_loc <- data.frame(plotcode = ssaw8$plotInfoFull$plotcode,
-  lon = ssaw8$plotInfoFull$longitude_of_centre,
-  lat = ssaw8$plotInfoFull$latitude_of_centre)
-
-plot_loc_spdf <- SpatialPointsDataFrame(coords = data.frame(plot_loc$lon, plot_loc$lat), data = plot_loc)
-
-# Temperature ----
+# Get Worldclim data ----
 
 # Get list of files
-rastlist_t <- list.files(path = "data/wc2.0_30s_tavg", 
+rastlist_t <- list.files(path = "/Volumes/john/worldclim/wc2.1_30s_tavg", 
   pattern = '.tif$', 
   all.files = TRUE, 
   full.names = TRUE)
 
-# Import
-allrasters_t <- lapply(rastlist_t, raster)
-
-# Crop to plot extent
[…]

### 2020-06-25 manuscript/regional_befr_manuscript.tex
Fixed references, finished updating analyses

@@
 \textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}.
-Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{and disturbance through fire}, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{, disturbance through fire}, and across vegetation type to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
 
@@
 \textbf{Methods:} We used a network of \nplots{} savanna/woodland plots located across the southern African sub-continent in which each tree >10 cm diameter was measured \rnew{and identified to species level}.
-We used structural equation modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, accounting for the interactive effects of resource availability \rnew{and disturbance by fire} along a gradient of stem density.
+We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, accounting for the interactive effects of resource availability \rnew{and disturbance by fire} along a gradient of stem density.
 
@@
 A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants.
-Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1983, Ratnam2011, Ryan2016} (\autoref{clust_map}).
+Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
 The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
@@
 
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the \rchange{so-called miombo woodland eco-region}{miombo ecoregion} (\autoref{clust_map}, \citealt{White1983}).
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the \rchange{so-called miombo woodland eco-region}{miombo ecoregion} (\autoref{clust_map}, \citealt{White1987}).
 The study \rchange{region spans}{area spacs the core climate space of the region}, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \rold{as defined by} \citep{White1983}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
+The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \rold{as defined by} \citep{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
 
@@
 Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.
-To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} (\autoref{shannon_equit}) was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness.
+To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness.
 Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits.
@@
 Groups were defined in \citet{Fayolle2018} in an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
-Group identification was conduscted using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes.
+Group identification was conducted using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes.
 Plot data used in this study occurred in four vegetation type groups.
@@
 	\includegraphics[width=1.4\textwidth]{clust_map}
-	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1983}.
+	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1987}.
 Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
@@
 When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation types the strengths of unstandardised path coefficients varied.
-The direct effect of tree species diversity on AGB was positive in Baikiaea and Mopane, but negative in Marginal and Core miombo (\autoref{struc_model_slopes_all}).
+The direct effect of tree species diversity on AGB was positive in Acacia and Mopane, but negative in Marginal and Core miombo (\autoref{struc_model_slopes_all}).
 Relationships between structural diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
@@
 
-\rchange{We repeatedly sub-sampled the plot dataset to build \subn{} datasets of varying mean stem density in order to test how the relationship between species diversity, structural diversity and biomass varied with stem density.
-Each dataset consisted of approximately \subp{} plots with overlap of plot identity between subsampled datasets.
-\autoref{sem_struc_stems_ha} shows a positive effect of tree species diversity on AGB as stem density increases.}{In our sub-sampling of the plot dataset by mean stem density, we found a positive effect of tree species diversity on AGB as stem density increases (\autoref{sem_struc_stems_ha})}.
+In our sub-sampling of the plot dataset by mean stem density, we found a positive effect of tree species diversity on AGB as stem density increases (\autoref{sem_struc_stems_ha}).
 There appears to be a minimum stem density threshold at \textapprox{}180 stems ha\textsuperscript{-1} below which there appears to be a reasonably constant low baseline effect of tree diversity on biomass.
@@
 We measured soil fertility using the observed variables of soil organic carbon content, sand particle content and Cation Exchange Capacity (CEC).
-In wet tropical forests a clear relationship has been observed between these variables and AGB \citep{Slik2010, MORE}.
+In wet tropical forests a clear relationship has been observed between these variables and AGB \citep{Slik2009, Coelho-de-Souza2019}.
 % KGD: You have written little about soil fertility in this section, but I think you need to.
-There is at least one paper which could be taken to suggest a positive relationship between AGB and soil fertility (Coelho de Souza et al.
-2019.
-NEE, see my publications).
-
-% In a separate set of linear multiple regressions, we found a weak positive interaction effect of moisture availability on the relationship between tree species diversity and AGB.
-As moisture availability increased, the relationship between tree species diversity and AGB became stronger.
-This is in contrast to \citet{Ratcliffe2017} who found that in European forests, a decrease in water availability due to drought led to a stronger effect of tree species diversity on AGB.
-They attributed this to an increase in selection effects which allowed dominance of stress tolerant species which were more likely to occur in high diversity assemblages.
 

### 2020-06-25 scripts/sem.R

@@ -56,6 +56,7 @@ dat <- dat %>%
   mutate(
     precip_seas_rev_std = -1 * precip_seas_std,
     temp_seas_rev_std = -1 * temp_seas_std,
+    temp_stress_rev_std = -1 * temp_stress_std,
     sand_rev_std = -1 * sand_std,
     temp_rev_std = -1 * temp_std,
     shannon_equit_rev_std = -1 * shannon_equit_std)
@@ -67,7 +68,7 @@ dat <- dat %>%
 corr_df <- dat %>%
   dplyr::select(sand_std, soil_c_log_std, cec_std, 
     precip_std, precip_seas_std,
-    temp_std, temp_seas_std,
+    temp_std, temp_stress_std,
     n_species_raref_log_std, shannon_equit_std,
     cov_height_std, cov_dbh_std, n_trees_gt10_ha_log_std, agb_ha_log_std)
 
@@ -107,7 +108,7 @@ ggcorrplot(cor(corr_df, use = "complete.obs"),
     "precip_std" = "MAP",
     "precip_seas_std" = expression(bolditalic(underline("PS"))), 
     "temp_std" = expression(bolditalic(underline("MAT"))),
-    "temp_seas_std" = expression(bolditalic(underline("TS"))),
+    "temp_stress_std" = expression(bolditalic(underline("TS"))),
     "n_species_raref_log_std" = "Extrap. sp. rich.",
     "shannon_equit_std" = "Shannon equit.",
     "cov_height_std" = "Tree height CoV",
@@ -119,7 +120,7 @@ ggcorrplot(cor(corr_df, use = "complete.obs"),
     "precip_std" = "MAP",
     "precip_seas_std" = expression(bolditalic(underline("PS"))), 
     "temp_std" = expression(bolditalic(underline("MAT"))),
-    "temp_seas_std" = expression(bolditalic(underline("TS"))),
+    "temp_stress_std" = expression(bolditalic(underline("TS"))),
     "n_species_raref_log_std" = "Extrap. sp. rich.",
     "shannon_equit_std" = "Shannon equit.",
     "cov_height_std" = "Tree height CV",
@@ -161,7 +162,7 @@ corr_ci_tab <- corr_ci_tab %>%
     x_var == "precip_std" ~ "MAP",
     x_var == "precip_sea_std" ~ "PS",
[…]

### 2020-06-29 manuscript/include/clust_summ.tex
Updated comments
Updated with comments

--- 2019-12-16 manuscript/include/clust_summ.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Dec 16, 2019 - 11:33:55
+% Date and time: Mon, Jun 29, 2020 - 10:43:25
 \begin{table}[!htbp] \centering 
-  \caption{Description of the biogeographical clusters (C1-C5) to which each plot in the study was assigned.
+  \caption{Description of the biogeographical clusters to which each plot in the study was assigned.
 Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv} and represent species which define the given cluster.
-Dominant species were identified by choosing the species with the largest AGB contribution within each cluster.
+Dominant species were identified by choosing the species with the largest mean plot level proportional AGB within each cluster.
 Numeric values of species richness, stems ha\textsuperscript{-1} and AGB are medians and interquartile ranges (75th percentile - 25th percentile).}
@@
 \hline \\[-1.8ex] 
-1 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
-\\ \textit{Brachystegia spiciformis} \\ \textit{Baikeaea plurijuga} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Diplorhynchus condylocarpon} \\ \textit{Burkea africana} \\ \textit{Pseudolachnostylis maprouneifolia} \end{tabular} & 687 & 11(11.2) & 170(145.3) & 37(34.68) \\
+Acacia shrubland & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Burkea africana} \\ \textit{Baikiaea plurijuga} \\ \textit{Pterocarpus angolensis} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\
 \hline
-2 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernadia }spp.
-\\ \textit{Brachystegia }spp.
-\\ \textit{Isoberlinia angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Julbernardia paniculata} \\ \textit{Isoberlinia angolensis} \\ \textit{Brachystegia longifolia} \end{tabular} & 757 & 18(17.6) & 215(171.7) & 48.8(43.7) \\
+Core miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Brachystegia spiciformis} \\ \textit{Julbernardia paniculata} \\ \textit{Brachystegia boehmii} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Parinari curatellifolia} \\ \textit{Uapaca kirkiana} \\ \textit{Brachystegia spiciformis} \end{tabular} & 523 & 20(16.9) & 204(142.5) & 44.2(36.11) \\
 \hline
-3 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia }spp.
-\\ \textit{Euclea racemosa} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Senegalia ataxacantha} \\ \textit{Combretum collinum} \end{tabular} & 226 & 10(10) & 165(157.5) & 46(47.81) \\
+Marginal miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia burkei} \\ \textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\ \textit{Vachellia nilotica} \\ \textit{Spirostachys africana} \end{tabular} & 466 & 12(13.7) & 178(129.5) & 36.9(26.98) \\
 \hline
-4 & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Combretum }spp.
-\end{tabular} & 99 & 7(8.2) & 190(155.7) & 41.5(36.93) \\
+Mopane & \begin{tabular}[c]{@{}c@{}c@{}} Colophospermum mopane \\ \textit{Androstachys johnsonii} \\ \textit{Kirkia acuminata} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Psuedolachnostylis maprouneifolia} \\ \textit{Lannea discolor} \end{tabular} & 58 & 10(10.2) & 186(125.6) & 42.7(32.83) \\
 \hline

### 2020-06-29 manuscript/include/struc_model_fit_clust_stats.tex
Updated with comments

--- 2019-12-16 manuscript/include/struc_model_fit_clust_stats.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Dec 16, 2019 - 16:41:44
+% Date and time: Mon, Jun 29, 2020 - 10:44:17
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-Marginal miombo & 525 & 47.480 & 6 & 0.964 & 0.910 & -3730.100 & 0.110 & 0.730 \\ 
-Core miombo & 668 & 59.440 & 6 & 0.958 & 0.895 & -4219 & 0.100 & 0.680 \\ 
-Baikiaea & 47 & 5.860 & 6 & 0.998 & 0.994 & -323.100 & 0.030 & 0.720 \\ 
-Mopane & 84 & 9.420 & 6 & 0.971 & 0.927 & -588.900 & 0.080 & 0.450 \\ 
-All & 1324 & 82.020 & 6 & 0.973 & 0.932 & -9122.800 & 0.090 & 0.700 \\ 
+Marginal miombo & 390 & 43.870 & 6 & 0.914 & 0.784 & -3111 & 0.130 & 0.460 \\ 
+Core miombo & 517 & 78.670 & 6 & 0.904 & 0.759 & -3710.000 & 0.140 & 0.490 \\ 
+Acacia shrubland & 28 & 9.570 & 6 & 0.952 & 0.879 & -180.900 & 0.130 & 0.830 \\ 
+Mopane & 46 & 19.880 & 6 & 0.834 & 0.584 & -336.500 & 0.240 & 0.510 \\ 
+All & 981 & 91.380 & 6 & 0.937 & 0.843 & -7517.000 & 0.120 & 0.490 \\ 
 \hline \\[-1.8ex] 

### 2020-06-29 manuscript/regional_befr_manuscript.tex
Updated with comments

@@
 
-\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}.
-Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density, resource availability \rnew{, disturbance through fire}, and across vegetation type to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+\textbf{Aim:} Positive correlations between tree \rchange{biodiversity}{species diversity} and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}.
+Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stem density, resource availability \rnew{, disturbance through fire}, and across vegetation type to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
 
@@
 
-\textbf{Methods:} We used a network of \nplots{} savanna/woodland plots located across the southern African sub-continent in which each tree >10 cm diameter was measured \rnew{and identified to species level}.
-We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, accounting for the interactive effects of resource availability \rnew{and disturbance by fire} along a gradient of stem density.
+\textbf{Methods:} We used a network of \nplots{} \rchange{savanna and woodland}{savanna/woodland tree} plots located across the southern African sub-continent \rold{in which each tree >10 cm diameter was measured and identified to species level}.
+We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stem density.
 
-\textbf{Results:} A positive effect of tree species diversity on aboveground biomass was demonstrated, observed largely as an indirect effect of increasing woodland structural diversity.
+\textbf{Results:} \rnew{We found} A positive effect of tree species diversity on aboveground biomass \rold{was demonstrated}, observed largely as an indirect effect of increasing woodland structural diversity.
 We also found that the effect of tree species diversity on biomass increases with stem density.
-Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.
+Finally, we found that resource availability affects biomass in southern African woodlands \rchange{largely}{mainly} indirectly, via its effect on species diversity.
 
-\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and function of highly disturbed southern African savannas and woodlands.
-Our results demonstrate the importance of including environmental conditions and vegetation type in models to accurately describe a general relationship between biodiversity and ecosystem function at a regional level.
-Biodiversity loss predominantly by human actions in southern Africa may have detrimental outcomes for ecosystem function, particularly in species poor Baikiaea woodlands, which showed the strongest biodiversity - ecosystem function relationship.
+\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and \rnew{ecosystem} function of highly disturbed southern African savannas and woodlands.
+Our results demonstrate the importance of \rchange{including}{accounting for} environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level.
+Biodiversity loss, predominantly \rchange{by}{through} human actions in southern Africa may have detrimental outcomes for ecosystem function, particularly in tree species-poor Baikiaea woodlands, which showed the strongest biodiversity - ecosystem function relationship.
 
@@
 Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
-These studies support early findings from small scale experimental studies predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+These studies support early findings from small-scale experimental studies \rnew{conducted} predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
 
 Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}.
-This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported and intuitive relationship between biodiversity and ecosystem function invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater biodiversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
+This includes \rchange{basic processes of primary production}{processes} such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
+The frequently reported \rold{and intuitive} relationship between biodiversity and ecosystem function invokes three main mechanisms \rchange{which drive}{to explain} the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
 
@@
 High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
-This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as apparent in dry woodland/savanna ecosystems.
-Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces which influence ecosystem function\rnew{s} \citep{Lasky2014, Tobner2016}.
+This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as \rchange{apparent}{important} in dry woodland/savanna ecosystems.
+Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}.
 Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations \rchange{in environments which are more hostile to growth}{under limiting environmental conditions such as low water availability \citep{} or high maximum temperature \citep{}}.
 Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition dminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation in order to characterise a generalisable BEFR.
+They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition dminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEFR.
 
@@
 \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}.
-A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanism described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants.
+A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants.
 Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
-The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
-Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species.
-Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
+Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages\rchange{, which}{.These} retain the common features of an open tree canopy and an understorey generally dominated by C4 \change{grass species}{grasses}.
+Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which \rold{there are} >300 \rnew{are} \rchange{tree species}{trees} \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
-Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely un-quantified consequences for ecosystem structure and function.
+Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.
 
-A small number of studies in SAWs, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+A small number of studies in SAWs, \rnew{all of which were restricted in the spatial scope to a small region of miombo woodland}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
 The results of these fine scale studies concur with similar studies in other biomes \citep{}.
@@
 \citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest.
-They also found that trees growing in species rich neighbourhoods exhibited enhanced biomass production.
-Occupation of multiple canopy layers allows a more full canopy with a greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity.
-This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas, which are structured by disturbance rather than competition.
-Instead, disturbance history may override the effects of tree species diversity on structural diversity nullifying the effects of tree species diversity on structural diversity.
+They also found that trees growing in species-rich neighbourhoods exhibited enhanced biomass production.
+\rchange{Occupation}{Occupancy} of multiple canopy layers allows a \rchange{more full}{fuller} canopy with \rold{a} greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity.
+This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas.
+Instead, the overriding importance of disturbance history may negate the effects of tree species diversity on structural diversity \citep{Grime2012}.
 
 High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition.
-Fire disturbance in forests has been linked to abundance dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}.
+Fire disturbance in forests has been linked to abundance-dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}.
 Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree.
@@
 In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
-We used aboveground woody biomass of trees and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
+We used aboveground woody biomass of trees \rnew{as our metric of ecosystem function}, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
 We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type via clustering of plot level tree species composition, as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and assess the generality of our results.
-We used Structural Equation Modelling (SEM) and path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass.
-Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strenghth of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
+We incorporated vegetation type (via clustering of plot level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results.
+We used Structural Equation Modelling (SEM) \rchange{and}{with} path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass.
+Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
 
@@
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
-Only plots >0.1 hectares were used in analysis, as area based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
+Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
 Only plots with a stem density >10 stems ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
@@
 
-After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition.
+\rold{After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition.
 We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbour criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
-We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thus removing \noutliers{} plots (\hyperref[appendixa]{Appendix A}).
+We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thus removing \noutliers{} plots (\hyperref[appendixa]{Appendix A}).}
 
@@
 Many plots in the dataset did not include data on stems <10 cm DBH.
-For those plots with stem measurements <10 cm DBH, those small stems only accounted for a median average of \percsmallagb{}\% of the plot level AGB.
+For those plots which contained stem measurements <10 cm DBH, small stems only accounted for a median of \percsmallagb{}\% of the plot level AGB.
 
@@
 A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
-For each stem we measured species, DBH and tree height to the top of the highest branch material.
+For each stem, we \rchange{measured}{recorded} species, DBH and tree height to the top of the highest branch material.
 Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks.
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
-The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from (Ryan C., unpublished) (\hyperref[appendixb]{Appendix B}).
+The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see  \citet{Godlee2020}.
 
-AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}.
-Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
-Wood density for species without species level estimates was estimated from the mean of their respective genus.
+AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:
 
@@
 
-Where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height.
+where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height.
+Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
+Wood density for species without species level estimates was estimated from the means of their respective \rchange{genus}{genera}.
 
@@
 Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.
-To measure tree species abundance evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness.
+To measure tree species \rold{abundance} evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness.
 Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits.
@@
 Plots were assigned to vegetation type groups based on tree species composition.
-Groups were defined in \citet{Fayolle2018} in an Africa wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
+Groups were defined in \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
 Group identification was conducted using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes.
@@
 	\includegraphics[width=1.4\textwidth]{clust_map}
-	\caption{The locations of the \nplots{} plots used in this study, as points by geographic location with respect to the distribution of miombo woodland vegetation according to \citet{White1987}.
+	\caption{The locations of the \nplots{} plots used in this study, with respect to the distribution of miombo woodland vegetation according to \citet{White1987}.
 Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
@@
 
-Structural Equation Models (SEM) investigated the determinants of AGB.
+\rnew{We used} Structural Equation Models (SEM) to investigate the determinants of AGB.
 All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}.
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
-A key aspect to our decision to use SEMs is that they can explicitly model and partition variance to indirect effects, which is challenging in standard multiple regressions.
-Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having single observable measures in our dataset.
+A key aspect to our decision to use SEMs is that they can explicitly model and partition variance \rnew{attributed} to indirect effects, which is challenging in standard multiple regressions.
+Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset.
 Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
@@
 
-Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixc]{Appendix C}).
+Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixa]{Appendix A}).
 Standardisation put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
 Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
-Standardisation also controls for variables with different orders of magnitude which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
+Standardisation also controls for variables with different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
 To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1.
@@
 We assessed the role of tree structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
-To account for variation in stem density which may covary with species diversity we also included it as an predictor in our model.
+To account for variation in stem density, which may covary with species diversity, we also included it as a predictor in our model.
 To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately.
@@
 Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}.
-Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
-We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} which define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.
+Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error \rnew{of Approximation} (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
+We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} \rchange{which}{who} define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.
 
@@
 We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model as we were deliberately controlling stem density in our subsampling.
-We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled dataset.
+We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled datasets.
 
@@
 We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB.
-Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation clusters were narrow in their climate space leading to a lack of variance particularly in moisture availability.
+Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation clusters were narrow in their climate space, leading to a lack of variance particularly in moisture availability.
 
@@
 	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlogram of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
-Variables in bold and underlined on the axis labels were later reversed for SEMs to maintain positive correlations for all observed variables within each latent variable.
+	\caption{Correlation matrix of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
+Variables in bold and underlined on the axis labels had their signs later reversed for SEMs to maintain positive correlations for all observed variables within each latent variable.
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
 Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = moisture availability, green = tree species diversity, purple = tree structural diversity.
-See \hyperref[appendixd]{Appendix D} for a full assessment of correlation fit statistics.}
+See \hyperref[appendixb]{Appendix B} for a full assessment of correlation fit statistics.}
 	\label{corr_mat}
@@
 Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
-Stem density had the strongest correlation with AGB of all variables (\ccib{}).
+Stem density had the strongest correlation with AGB of all variables \rnew{considered} (\ccib{}).
 Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except CEC and MAT.
 
-The direction of these correlations was used as a test of our assumptions of the direction of influence of latent variables later used in the SEMs.
+The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs.
 As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a weak negative correlation between the seasonality of precipitation and AGB (\ccmcb{}).
-MAT and temperature seasonality (TS) negatively correlated weakly with AGB (MAT: \cctb{}; TS: \cctcb{}).
+MAT and temperature seasonality (TS) \rnew{were} \rnew{weakly} negatively correlated \rold{weakly} with AGB (MAT: \cctb{}; TS: \cctcb{}).
 As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic carbon and AGB (\ccob{}).
 
-MAP had positive correlations with tree species richness (\ccms{}), abundance evenness (\ccme{}), tree height diversity (\ccmh{}) and tree stem density (\ccmi{}).
+MAP had positive correlations with tree species richness (\ccms{}), evenness (\ccme{}), tree height diversity (\ccmh{}) and tree stem density (\ccmi{}).
 MAT had weak correlations with tree species and structural diversity variables.
@@
 Tree species diversity had a positive correlation with stem density.
-Model fit was good with high factor loadings for all observed variables, all path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}).
+Model fit was good with high factor loadings for all observed variables.
+All path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}).
 The R\textsuperscript{2} of AGB was \strucrsq{}.
@@
 	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all vegetation clusters.
-Latent variables are circles while observed variables are rectangles.
-Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
-The observed variables which inform the latent variables are connected by dotted arrows, observed variables with loading set to 1 are connected by dashed arrows.
+Latent variables are shown as shown as shown as shown as shown as shown as shown as shown as shown as circles while observed variables are shows as rectangles.
+Standardised path coefficients are shown as solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
+The observed variables that inform the latent variables are connected by dotted arrows, and observed variables with loadings set to 1 are connected by dashed arrows.
 Correlations between variables are depicted as dotted curved arrows.
@@
 
-When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation types the strengths of unstandardised path coefficients varied.
+When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation types, the strengths of unstandardised path coefficients varied.
 The direct effect of tree species diversity on AGB was positive in Acacia and Mopane, but negative in Marginal and Core miombo (\autoref{struc_model_slopes_all}).
-Relationships between structural diversity and AGB remained generally similar with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
+Relationships between structural diversity and AGB remained generally similar, with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
 The total effect of species diversity on AGB remained strongly positive for all vegetation types.
@@
 The strongest total effect of tree species diversity on AGB was in Baikiaea woodland (\struccsb{}), which was species rich but highly variable in species diversity compared to other vegetation types (\autoref{clust_summ}).
-The R\textsuperscript{2} of AGB was highest in Marginal miombo (R\textsuperscript{2} = \strucarsq{}) and lowest in the Core miombo (R\textsuperscript{2} = \\strucbrsq{}strucbrsq{}).
+The R\textsuperscript{2} of AGB was highest in Marginal miombo (R\textsuperscript{2} = \strucarsq{}) and lowest in the Core miombo (R\textsuperscript{2} = \strucbrsq{}).
 
@@
 
-In our sub-sampling of the plot dataset by mean stem density, we found a positive effect of tree species diversity on AGB as stem density increases (\autoref{sem_struc_stems_ha}).
+In our sub-sampling of the plot dataset by mean stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}).
 There appears to be a minimum stem density threshold at \textapprox{}180 stems ha\textsuperscript{-1} below which there appears to be a reasonably constant low baseline effect of tree diversity on biomass.
@@
 
-Similar to the model which only considered tree species and structural diversity (\autoref{struc_mod}), the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB, leading to a strong positive indirect effect of species diversity on AGB via structural diversity (\rgbhd{}).
+Similar to the model that only considered tree species and structural diversity (\autoref{struc_mod}), the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB, leading to a strong positive indirect effect of species diversity on AGB via structural diversity (\rgbhd{}).
 The total effect of species diversity on AGB was positive (\rgbd{}).
@@
 Latent variables are shown as circles while observed variables are shown as rectangles.
-Standardised path coefficients are solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
-Observed variables which inform the latent variables are connected by dotted arrows, observed variables with loading set to 1 are connected by dashed arrows.
+Standardised path coefficients are shown as solid arrows pointing from predictor to response, with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
+Observed variables that inform the latent variables are connected by dotted arrows, observed variables with loading set to 1 are connected by dashed arrows.
 Correlations between variables are depicted as dotted curved arrows.
@@
 The strongest effect on AGB was that of stem density.
-Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, in the form of moisture or soil fertility, on AGB (H\textsubscript{3}).
+Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, either in the form of moisture or soil fertility, on AGB (H\textsubscript{3}).
 
@@
 We found a consistent positive effect of tree species diversity on AGB across all models in this study.
-Within SAWs we therefore find support that higher tree species richness and evenness causes higher woody AGB.
-This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
+Within SAWs we therefore find support for our hypothesis that higher tree species richness and evenness causes higher woody AGB.
+This finding is in agreement with many other studies across different ecosystems and biomes, supporting \rnew{the idea} that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
 Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of SAWs, a disturbance-structured and poorly studied ecological system.
@@
 
-We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity and we therefore find support for our hypothesis (H\textsubscript{2}).
+We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity, and we therefore find support for our second hypothesis (H\textsubscript{2}).
 A higher tree species diversity allows for a greater structural diversity of trees, i.e.
@@
 This may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
-Although we did not measure it here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{}, i.e.
+Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{}, i.e.
 wider variation in canopy shape and overall growth form; bushy understorey vs.
@@
 Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this.
-In frequently disturbed woodlands such as those studied here however, a woodland canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
+In frequently disturbed woodlands such as those studied here however, a woodland canopy similar to that of a forest is frequently not reached.
+Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
 While we did not have access to adequate data on disturbance history in our plots, previous studies have found that SAWs with higher species diversity tend to be less disturbed and tend to form a more closed canopy \citep{Chidumayo2013, Mutowo2012}.
@@
 In this study, while we observed weak indirect effects via species diversity, we saw no evidence for a direct effect of moisture availability on AGB.
-Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in SAWs, which are frequently droughted.
+Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in SAWs, which \rchange{are frequently droughted}{experience frequent drought}.
 It is possible that the range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture variation in AGB.
@@
 
-\subsection{Vegetation type specific responses}
+\subsection{Vegetation type responses}
 
@@
 
-\section{Appendix 1 - Data cleaning process} \label{appendixa}
-
-\begin{figure}[H]
-\centering
-	\input{include/data_clean_flow.tex}
-	\caption{Flow diagram of the data filtering and cleaning process prior to analysis. Rounded boxes indicate filtering events while regular boxes indicate calculation events.}
-	\label{data_clean_flow}
-\end{figure}
-
-\section{Appendix 2 - Estimation of DBH via tree taper} \label{appendixb}
-
-\begin{lstlisting}[language=R]
-##' @title Stem diameter Point Of Measurement (POM) adjustment
-##' @description  Function to estimate stem diameter at 1.3 given measurements 
-##'   at other POMs.
-##' @author Casey M. Ryan
-##' @return d130, the estimated diameter at a POM of 1.3 m (in cm). 
-##' @param d_in the diameter measured at the POM (in cm)
-##' @param POM the height of the POM (in m)
-##' @details The adjustment is based on a tree taper model developed as part of 
-##'   the ACES project (Abrupt Changes in Ecosystem Services 
-##'   https://miomboaces.wordpress.com/), using data from the miombo of Niassa. 
-##'   The model is a cubic polynomial, with three equations for different sized 
-##'   stems. 
-##' @section Warning: The model should not be used for POMs above 1.7 m. 
-##'   Extrapolating beyond the training data will give nonsense. 
-##'   Thus, POMs >1.7 m are not adjusted.
-##' @examples
-##' POMadj(10, 0.3)
-##' POMadj(1, 0.3)  # d130 is negative, i.e. the stem probably wasn't 1.3 m tall
-##' POMadj(50, 1.9)  # generates warning, as outside calibration data range
-##' \dontrun{
-##'   POMadj(50, 0)  # zero or -ve POM is outside range, or nonsense
-##' }
-POMadj <- function(d_in, POM) {
-  stopifnot(is.numeric(d_in),
-    is.numeric(POM),
-    POM >= 0,
-    sum(is.na(POM))==0,
-    length(POM) == length(d_in))
-  if (any(POM > 1.7))
-    warning("POMs >1.7 m are outside the calibration data, no correction applied")
-
-  NAS <- is.na(d_in)
-  d_in_clean <- d_in[!NAS]
-  POM_clean <- POM[!NAS]
-  # define the size class edges:
-  edges <- c(5.0, 15.8, 26.6, 37.4)
-  sm <- d_in_clean < edges[2]
-  med <- d_in_clean >= edges[2] & d_in_clean < edges[3]
-  lg <- d_in_clean >= edges[3]
-
-  # compute apredictions for delta_d, for all size classes
-  delta_d <- data.frame(
-    # if small:
-    small =  3.4678+-5.2428 * 
-   	  POM_clean + 2.9401 * 
-      POM_clean^2+-0.7141 * 
-      POM_clean^3,
-    # if med
-    med =  4.918+-8.819 * 
-      POM_clean + 6.367 * 
-      POM_clean^2+-1.871 * 
-      POM_clean^3,
-    # if large
-    large =  9.474+-18.257 * 
-      POM_clean + 12.873 * 
-      POM_clean^2+-3.325 * 
-      POM_clean^3
-  )
-  # index into the right size class
-  dd <- NA_real_
-  dd[sm] <- delta_d$small[sm]
-  dd[med] <- delta_d$med[med]
-  dd[lg] <- delta_d$large[lg]
-  dd[POM_clean > 1.7] <- 0  # to avoid extrapolation mess
-
-  # add NAs back in
-  d130 <- NA
-  d130[NAS] <- NA
-  d130[!NAS] <- d_in_clean - dd
-
-  if (any(d130[!NAS] < 0))
-    warning("Negative d130 estimated, repaced with NA")
-  d130[d130 <= 0 & !is.na(d130)] <- NA
-  return(d130)
-}
-\end{lstlisting}
-
-\section{Appendix 3 - Frequency distribution of observed variables} \label{appendixc}
+\section{Appendix 1 - Frequency distribution of observed variables} \label{appendixa}
 
@@
 
-\section{Appendix 4 - Table of correlation fit statistics} \label{appendixd}
+\section{Appendix 2 - Table of correlation fit statistics} \label{appendixb}
 
 \input{include/corr_ci_tab.tex}
+
+\section{Appendix 3 - Bivariate relationships of model variables} \label{appendixc}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{bivar_lm}
+	\caption{}
+	\label{bivar_lm}
+\end{figure}
 

### 2020-06-29 scripts/data_clean.R

@@ -17,6 +17,8 @@ library(iNEXT)
 library(vegan) 
 library(ggplot2)
 library(seosawr)
+library(raster)
+library(sf)
 
 
 # Import plot and stems data ----
@@ -29,6 +31,10 @@ plots <- read.csv("~/git_proj/seosaw_data/data_out/plots_v2.7.csv")
 
 # SEOSAW v2 stem data
 stems <- read.csv("~/git_proj/seosaw_data/data_out/stems_latest_v2.7.csv")
+
+# Fire layer for manual calculation
+fire_count <- raster("~/Downloads/firecount_2001_2018.tif")
+
  
 # Datasets used in analysis
 dataset_codes <- c("ZIS", "SSM", "MCL", "ZPF", "ZNF", "MNR", "MGR", "MAR",
@@ -211,6 +217,42 @@ clust <- ssaw8$cluster %>%
 plots_clust <- left_join(plots_div_clean, clust, 
   by = c("plot_cluster" = "plot_id")) %>%
   filter(!is.na(clust4))
+
+
+# Calculate fire count for buffer area ----
+x <- plots_clust %>% 
+  filter(!is.na(longitude_of_centre), !is.na(latitude_of_centre)) %>%
+  st_as_sf(., coords = c("longitude_of_centre", "latitude_of_centre"))  
+
+# Subset raster and set projection from EPSG
+ds <- raster::stack(fire_count)
+
+# Add optional radius buffer
+x$utm <- latLong2UTM(x = st_coordinates(x)[,1],
+  y = st_coordinates(x)[,2])
+
+st_crs(x) <- 4326
+
+# Split by UTM 
+x_split <- split(x, x$utm)
+
+x_buffer_split <- lapply(x_split, function(y) {
+  y_utm <- sf::st_transform(y, crs = UTMProj4(unique(y$utm)))
+  y_buffer <- sf::st_buffer(y_utm, 500)
+  y_latlon <- sf::st_transform(y_buffer, crs = 4326)
+  y_latlon
+})
+
+x_buffer <- do.call(rbind, x_buffer_split)
+
+# Pull values from points in raster
+out_nona <- raster::extract(ds, x_buffer, method = "simple", 
+    fun = mean, factors = TRUE)
+out_nona_clean <- as.data.frame(cbind(plot_cluster = x_buffer$plot_cluster , out_nona))
+names(out_nona_clean) <- c("plot_cluster", "fire_buffer")
[…]

### 2020-06-29 scripts/standardise.R

@@ -44,6 +44,9 @@ facet_levels <- c(
   "sand",
   "soil_c",
   "soil_c_log",
+  "fire_buffer", 
+  "fire_buffer_log", 
+  "herbivory",
   "cov_dbh",
   "cov_height",
   "n_trees_gt10_ha",
@@ -65,6 +68,9 @@ facet_labels <- c(
   expression("Sand" ~ "%"), 
   expression("Org." ~ "C" ~ "(ppt)"),
   expression("log(Org." ~ "C)" ~ "(ppt)"),
+  expression("Mean" ~ "fire" ~ "freq." ~ (2001-2018)),
+  expression("log(Mean" ~ "fire" ~ "freq.)" ~ (2001-2018)),
+  expression("Herbivore" ~ "biomass" ~ (kg ~ ha^-1)),
   expression("Coef." ~ "var." ~ "DBH"),
   expression("Coef." ~ "var." ~ "height"),
   expression("Stem" ~ "density" ~ ">10" ~ cm ~ (n ~ ha^-1)),
@@ -81,6 +87,7 @@ hist_raw <- dat %>%
   dplyr::select(agb_ha, n_trees_gt10_ha, 
     n_species_raref, shannon_equit,
     cec, sand, soil_c, 
+    fire_buffer, herbivory,
     precip, precip_seas, temp, temp_seas, temp_stress,
     cov_height, cov_dbh) %>%
   gather(variable, value) %>%
@@ -103,12 +110,14 @@ dat_trans <- dat %>%
 	mutate(agb_ha_log = log(agb_ha),
 		n_trees_gt10_ha_log = log(n_trees_gt10_ha), 
 	  n_species_raref_log = log(n_species_raref + 4),
-	  soil_c_log = log(soil_c + 4))
+	  soil_c_log = log(soil_c + 4),
+	  fire_buffer_log = log(fire_buffer + 4))
 
 hist_trans <- dat_trans %>%
   dplyr::select(agb_ha_log, n_trees_gt10_ha_log,
     n_species_raref_log, shannon_equit,
     cec, sand, soil_c_log,
+    fire_buffer_log, herbivory,
     precip, precip_seas, temp, temp_seas, temp_stress,
     cov_height, cov_dbh) %>%
   gather(variable, value) %>%
@@ -137,6 +146,8 @@ sem_data_norm_std <- dat_trans %>%
 	  "cec",
 	  "sand",
 	  "soil_c_log",
[…]

### 2020-06-29 scripts/temp_precip.R

@@ -11,7 +11,7 @@ library(raster)
 library(rgdal)
 
 # Get Miombo outline ---- 
-white_veg <- readOGR(dsn="/Volumes/john/whitesveg", layer="")
+white_veg <- readOGR(dsn="/Volumes/john/whitesveg", layer="whitesveg")
 
 # Subset to White's miombo 
 white_veg_miombo <- white_veg[!is.na(white_veg@data$DESCRIPTIO) & 
@@ -75,7 +75,7 @@ rastlist_extrac <- lapply(rastlist, function(x) {
 
 # Combine to dataframe
 t_p <- data.frame(t_vals = rastlist_extrac[[1]][,1], t_cv = rastlist_extrac[[1]][,2],
-  p_vals = rastlist_extrac[[2]][,1], p_cv = rastlist_extrac[[2]][,2],
+  p_vals = rastlist_extrac[[2]][,1], p_cv = rastlist_extrac[[2]][,2])
 
 # Save to file
 saveRDS(t_p, "data/temp_precip.rds")

### 2020-06-30 manuscript/regional_befr_manuscript.tex
More changes

@@
 \input{include/n_plots}
-\input{include/n_outliers}
 \input{include/hull_cover}
@@
 \input{include/corr_coef}
-\input{include/moder_coef}
 \input{include/dens_stats}
@@
 \textbf{Aim:} Positive correlations between tree \rchange{biodiversity}{species diversity} and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}.
-Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stem density, resource availability \rnew{, disturbance through fire}, and across vegetation type to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+\rnew{In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function.} Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stem density, resource availability\rnew{, disturbance through fire and herbivory}, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
 
@@
 \textbf{Methods:} We used a network of \nplots{} \rchange{savanna and woodland}{savanna/woodland tree} plots located across the southern African sub-continent \rold{in which each tree >10 cm diameter was measured and identified to species level}.
-We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stem density.
+We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stem density.
 
-\textbf{Results:} \rnew{We found} A positive effect of tree species diversity on aboveground biomass \rold{was demonstrated}, observed largely as an indirect effect of increasing woodland structural diversity.
+\textbf{Results:} \rnew{We found} a positive effect of tree species diversity on aboveground biomass \rold{was demonstrated}, observed \rchange{largely as an}{mainly via the} of increasing woodland structural diversity.
 We also found that the effect of tree species diversity on biomass increases with stem density.
@@
 \rnew{Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}}.
-High levels of disturbance may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
+High levels of disturbance by fire may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
 This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as \rchange{apparent}{important} in dry woodland/savanna ecosystems.
@@
 Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition dminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEFR.
+They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition diminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space, thus reducing the competition which drives niche complementarity effects.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEFR.
 
@@
 The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
-Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages\rchange{, which}{.These} retain the common features of an open tree canopy and an understorey generally dominated by C4 \change{grass species}{grasses}.
+Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages\rchange{, which}{.These} retain the common features of an open tree canopy and an understorey generally dominated by C4 \rchange{grass species}{grasses}.
 Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which \rold{there are} >300 \rnew{are} \rchange{tree species}{trees} \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
@@
 We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type (via clustering of plot level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results.
+We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results.
 We used Structural Equation Modelling (SEM) \rchange{and}{with} path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass.
-Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
+Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity \rnew{(i.e.
+physiognomic diversity)}, which will provide an indirect path by which tree diversity increases woody biomass.
 
@@
 \rold{After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition.
-We used the \verb|outlier()| function from the \verb|dave| R package \citep{dave}, which uses a nearest neighbour criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
-We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space \citep{Otto2013}, thus removing \noutliers{} plots (\hyperref[appendixa]{Appendix A}).}
+We used the outlier() function from the dave R package citep{dave}, which uses a nearest neighbour criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
+We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space citep{Otto2013}, thus removing noutliers{} plots (hyperref[appendixa]{Appendix A}).}
 
@@
 
-We considered only trees and shrubs in our calculations of above-ground woody biomass (AGB), including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
+We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
 Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
@@
 
-All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\nstems{}} stems with measurements.
+All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements.
 A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
@@
 Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}.
-Values of \rchange{Mean Annual Temperature (MAT)}{MAT} and \rchange{Mean Annual Precipitation (MAP)}{MAP} were calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
+Values of \rchange{Mean Annual Temperature (MAT)}{MAT} and \rchange{Mean Annual Precipitation (MAP)}{MAP} was calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
 Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily mean temperature and precipitation, respectively, for each of the 18 years of available data.
@@
 Groups were defined in \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
-Group identification was conducted using unconstrained correspondence analysis and ordination, followed by clustering based on dominant ordination axes.
+Group identification was conducted using unconstrained correspondence analysis\rold{ and ordination}, followed by clustering based on dominant ordination axes.
 Plot data used in this study occurred in four vegetation type groups.
@@
 A key aspect to our decision to use SEMs is that they can explicitly model and partition variance \rnew{attributed} to indirect effects, which is challenging in standard multiple regressions.
-Using SEMs also allowed us to describe theoretical latent constructs which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset.
+Using SEMs also allowed us to describe latent variables such as ``resource availability'' and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset.
 Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
@@
 Standardisation also controls for variables with different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
-To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables were reversed by multiplying by -1.
+To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables had their sign reversed.
 For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed for model fitting.
@@
 	\caption{Correlation matrix of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
-Variables in bold and underlined on the axis labels had their signs later reversed for SEMs to maintain positive correlations for all observed variables within each latent variable.
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
@@
 Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
-While we did not have access to adequate data on disturbance history in our plots, previous studies have found that SAWs with higher species diversity tend to be less disturbed and tend to form a more closed canopy \citep{Chidumayo2013, Mutowo2012}.
+While we did not have access to adequate data on disturbance history in our plots, previous studies have found that SAWs with higher species diversity tend to \rnew{experience less frequent disturbance by fire and tend to form a more closed canopy and a more sparse understorey} \citep{Chidumayo2013, Mutowo2012}.
 
@@
 
-SAWs are relied heavily upon for their ecosystem service provision, which is itself affect by ecosystem function.
+SAWs are relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem function.
 Resource extraction by humans in southern Africa is directly influencing biodiversity via selective tree-felling for timber, among other forest products.
@@
 \centering
-	\includegraphics[width=\textwidth]{histogram_raw_obs}
+	\includegraphics[width=\textwidth]{hist_raw}
 	\caption{Histograms of raw untransformed observed variables used in final analyses.}
-	\label{histogram_raw_obs}
+	\label{hist_raw}
 \end{figure}
@@
 \centering
-	\includegraphics[width=\textwidth]{histogram_trans_obs}
+	\includegraphics[width=\textwidth]{hist_trans}
 	\caption{Histograms of observed variables transformed to achieve a normal frequency distribution.}
-	\label{histogram_trans_obs}
+	\label{hist_trans}
 \end{figure}

### 2020-06-30 scripts/data_clean.R

@@ -218,41 +218,17 @@ plots_clust <- left_join(plots_div_clean, clust,
   by = c("plot_cluster" = "plot_id")) %>%
   filter(!is.na(clust4))
 
+# Extract soil Nitrogren from soil Grids ----
 
-# Calculate fire count for buffer area ----
-x <- plots_clust %>% 
-  filter(!is.na(longitude_of_centre), !is.na(latitude_of_centre)) %>%
-  st_as_sf(., coords = c("longitude_of_centre", "latitude_of_centre"))  
+#write.csv(plots_clust[,c("longitude_of_centre", "latitude_of_centre")], 
+#  "~/Downloads/plots_coord.csv", row.names = FALSE)
 
-# Subset raster and set projection from EPSG
-ds <- raster::stack(fire_count)
+##' Run external script `scripts/nitrogen_get.sh`
 
-# Add optional radius buffer
-x$utm <- latLong2UTM(x = st_coordinates(x)[,1],
-  y = st_coordinates(x)[,2])
+nitrogen <- readLines("data/nitrogen")
+plots_clust$nitrogen <- as.numeric(nitrogen)
 
-st_crs(x) <- 4326
-
-# Split by UTM 
-x_split <- split(x, x$utm)
-
-x_buffer_split <- lapply(x_split, function(y) {
-  y_utm <- sf::st_transform(y, crs = UTMProj4(unique(y$utm)))
-  y_buffer <- sf::st_buffer(y_utm, 500)
-  y_latlon <- sf::st_transform(y_buffer, crs = 4326)
-  y_latlon
-})
-
-x_buffer <- do.call(rbind, x_buffer_split)
-
-# Pull values from points in raster
-out_nona <- raster::extract(ds, x_buffer, method = "simple", 
-    fun = mean, factors = TRUE)
-out_nona_clean <- as.data.frame(cbind(plot_cluster = x_buffer$plot_cluster , out_nona))
-names(out_nona_clean) <- c("plot_cluster", "fire_buffer")
-out_nona_clean$fire_buffer <- as.numeric(out_nona_clean$fire_buffer)
-
-plots_clust <- left_join(plots_clust, out_nona_clean, by = "plot_cluster") 
+# External shell script
 
 # Write data ----
 # Plot group - plotcode lookup 

### 2020-06-30 scripts/descrip.R

@@ -23,7 +23,6 @@ library(labdsv)
 
 source("scripts/clust_defin.R")
 
-
 # Import data ----
 
 # Cleaned data
@@ -45,7 +44,6 @@ t_p <- readRDS("data/temp_precip.rds")
 white_veg <- readOGR(dsn="/Volumes/john/whitesveg", 
   layer="whitesveg")
 
-
 # Fortify whitesveg and subset to Miombo ----
 white_veg_fort <- fortify(white_veg, region = "DESCRIPTIO")
 
@@ -53,11 +51,9 @@ white_veg_miombo <- white_veg_fort %>%
   filter(id %in% c("Moist-infertile savanna"),
     lat < -2)
 
-
 # Make clust4 as factor ----
 plot_data$clust4 <- clust_names[plot_data$clust4]
 plot_data$clust4 <- factor(plot_data$clust4)
-
 
 # Map of plot locations, compared to full SEOSAW dataset ----
 
@@ -69,50 +65,6 @@ s_af <- iso.expand(c("ZAF", "COD", "NAM", "ZMB", "BWA", "ZWE", "MOZ",
 map_africa <- borders(database = "world", regions = s_af, fill = "grey90", colour = "black")
 map_africa_fill <- borders(database = "world", regions = s_af, fill = "grey90")
 map_africa_colour <- borders(database = "world", regions = s_af, colour = "black")
-
-# Spatial distribution of key variables
-
-var_map <- function(x, var, lab = expression("log(AGB) (t ha"^-1*")")){
-  map_var <- ggplot() + 
-    map_africa + 
-    geom_point(data = x, aes(x = longitude_of_centre, y = latitude_of_centre, fill = var), 
-      size = 2, shape = 21, colour = "black", position = "jitter") +
-    scale_fill_viridis_c(name = lab) +
-    coord_map() + 
-    ylim(-35.5, 10) + 
-    labs(x = "Longitude", y = "Latitude") + 
-    theme_classic()
-  
-  return(map_var)
-}
-
-# Biomass
-pdf(file = "img/biomass_map.pdf", width = 5, height = 8)
-var_map(plot_data, 
-  var = plot_data$agb_ha, 
-  lab = expression("log(AGB) (t ha"^-1*")"))
-dev.off()
-
-# Precipitation
-pdf(file = "img/precip_map.pdf", width = 5, height = 8)
-var_map(plot_data, 
-  var = plot_data$precip, 
[…]

### 2020-07-02 manuscript/include/clust_summ.tex
Draft

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Jun 29, 2020 - 10:43:25
+% Date and time: Wed, Jul 01, 2020 - 14:44:00
 \begin{table}[!htbp] \centering 
@@
 Dominant species were identified by choosing the species with the largest mean plot level proportional AGB within each cluster.
-Numeric values of species richness, stems ha\textsuperscript{-1} and AGB are medians and interquartile ranges (75th percentile - 25th percentile).}
+Numeric values of species richness, stems ha\textsuperscript{-1} and AGB represent medians and interquartile ranges (75th percentile - 25th percentile).}
   \label{clust_summ} 
@@
 \hline \\[-1.8ex] 
-Acacia shrubland & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Burkea africana} \\ \textit{Baikiaea plurijuga} \\ \textit{Pterocarpus angolensis} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\
+Acacia shrubland & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia burkei} \\ \textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\ \textit{Vachellia nilotica} \\ \textit{Spirostachys africana} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\
 \hline
@@
 \hline
-Marginal miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia burkei} \\ \textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\ \textit{Vachellia nilotica} \\ \textit{Spirostachys africana} \end{tabular} & 466 & 12(13.7) & 178(129.5) & 36.9(26.98) \\
+Marginal miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Burkea africana} \\ \textit{Baikiaea plurijuga} \\ \textit{Pterocarpus angolensis} \end{tabular} & 466 & 12(13.7) & 178(129.5) & 36.9(26.98) \\
 \hline
-Mopane & \begin{tabular}[c]{@{}c@{}c@{}} Colophospermum mopane \\ \textit{Androstachys johnsonii} \\ \textit{Kirkia acuminata} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Psuedolachnostylis maprouneifolia} \\ \textit{Lannea discolor} \end{tabular} & 58 & 10(10.2) & 186(125.6) & 42.7(32.83) \\
+Mopane & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Androstachys johnsonii} \\ \textit{Kirkia acuminata} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{\textit{Colophospermum mopane}} \\ \textit{Psuedolachnostylis maprouneifolia} \\ \textit{Lannea discolor} \end{tabular} & 58 & 10(10.2) & 186(125.6) & 42.7(32.83) \\
 \hline

### 2020-07-02 manuscript/include/struc_model_fit_clust_stats.tex
Draft

--- 2019-12-16 manuscript/include/struc_model_fit_clust_stats.tex
@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Mon, Dec 16, 2019 - 16:41:44
+% Date and time: Thu, Jul 02, 2020 - 11:10:18
 \begin{table}[!htbp] \centering 
-  \caption{Model fit statistics for SEMs investigating the effects of tree diversity and stem density on AGB (\autoref{struc_mod}).} 
+  \caption{Model fit statistics for SEMs investigating the effects of tree diversity and stem density on AGB (\autoref{struc_mod}).
+n = number of plots in cluster, $\chi^{2}$ = Chi-squared fit statistic, DoF = model degrees of freedom, CFI = Comparative Fit Index, TLI = Tucker-Lewis Index, RMSEA = Root Mean Square Error of Approximation, $R^{2}$ AGB = R-squared of AGB.}
   \label{struc_model_fit_clust_stats} 
-\begin{tabular}{@{\extracolsep{0pt}} ccccccccc} 
+\begin{tabular}{@{\extracolsep{0pt}} cccccccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-{Cluster} & {n} & {$\chi^{2}$} & {DoF} & {CFI} & {TLI} & {LogLik} & {RMSEA} & {$R^{2}$ AGB} \\
+{Cluster} & {n} & {$\chi^{2}$} & {DoF} & {CFI} & {TLI} & {RMSEA} & {$R^{2}$ AGB} \\
 \hline \\[-1.8ex] 
-Marginal miombo & 525 & 47.480 & 6 & 0.964 & 0.910 & -3730.100 & 0.110 & 0.730 \\ 
-Core miombo & 668 & 59.440 & 6 & 0.958 & 0.895 & -4219 & 0.100 & 0.680 \\ 
-Baikiaea & 47 & 5.860 & 6 & 0.998 & 0.994 & -323.100 & 0.030 & 0.720 \\ 
-Mopane & 84 & 9.420 & 6 & 0.971 & 0.927 & -588.900 & 0.080 & 0.450 \\ 
-All & 1324 & 82.020 & 6 & 0.973 & 0.932 & -9122.800 & 0.090 & 0.700 \\ 
+Marginal miombo & 390 & 43.870 & 6 & 0.914 & 0.784 & 0.130 & 0.460 \\ 
+Core miombo & 517 & 78.670 & 6 & 0.904 & 0.759 & 0.140 & 0.490 \\ 
+Acacia shrubland & 28 & 9.570 & 6 & 0.952 & 0.879 & 0.130 & 0.830 \\ 
+Mopane & 46 & 19.880 & 6 & 0.834 & 0.584 & 0.240 & 0.510 \\ 
+All & 981 & 91.380 & 6 & 0.937 & 0.843 & 0.120 & 0.490 \\ 
 \hline \\[-1.8ex] 

### 2020-07-02 manuscript/regional_befr_manuscript.tex
Draft

@@
 \textbf{Aim:} Positive correlations between tree \rchange{biodiversity}{species diversity} and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}.
-\rnew{In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function.} Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stem density, resource availability\rnew{, disturbance through fire and herbivory}, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+\rnew{In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function.} Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stocking density, resource availability\rnew{, disturbance through fire}, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
 
@@
 \textbf{Methods:} We used a network of \nplots{} \rchange{savanna and woodland}{savanna/woodland tree} plots located across the southern African sub-continent \rold{in which each tree >10 cm diameter was measured and identified to species level}.
-We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stem density.
+We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stocking density.
 
 \textbf{Results:} \rnew{We found} a positive effect of tree species diversity on aboveground biomass \rold{was demonstrated}, observed \rchange{largely as an}{mainly via the} of increasing woodland structural diversity.
-We also found that the effect of tree species diversity on biomass increases with stem density.
+We also found that the effect of tree species diversity on biomass increases with stocking density.
 Finally, we found that resource availability affects biomass in southern African woodlands \rchange{largely}{mainly} indirectly, via its effect on species diversity.
@@
 In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots.
-We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density.
+We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stocking density.
 A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity.
-We found that the effect of tree species diversity on biomass increases with stem density.
+We found that the effect of tree species diversity on biomass increases with stocking density.
 Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.
@@
 
-Scientific interest in the relationship between species diversity and ecosystem function springs from both an interest in the factors which structure ecological communities \citep{}, and a more applied interest in determining the effect of global biodiversity loss on ecosystem form and function \citep{}.
+In order to understand the effects of global biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}.
 Numerous studies have shown relationships between biodiversity and ecosystem function (e.g.
 \citealt{Liang2016, Hooper2012, Cardinale2009}).
-The strength and direction of these observed Biodiversity-Ecosystem Function Relationships (BEFRs) varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}.
-Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
+The strength and direction of the Biodiversity-Ecosystem Function (BEF) relationship varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}.
+Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
 These studies support early findings from small-scale experimental studies \rnew{conducted} predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+Despite these concerted efforts, we continue to lack a nuanced, ecosystem agnostic understanding of the complex interactions between biodiversity, environment, and ecosystem function.
 
@@
 Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition diminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space, thus reducing the competition which drives niche complementarity effects.} This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEFR.
+They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition diminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space, thus reducing the competition which drives niche complementarity effects.} This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEF relationship.
 
-The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}.
-A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEFR mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEFR and its environmental determinants.
+The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}.
+A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEF relationship mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEF relationship and its environmental determinants.
 Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
@@
 Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages\rchange{, which}{.These} retain the common features of an open tree canopy and an understorey generally dominated by C4 \rchange{grass species}{grasses}.
-Southern African savannas and woodlands (SAWs) are highly diverse, thought to harbour \textapprox{}8500 plant species of which \rold{there are} >300 \rnew{are} \rchange{tree species}{trees} \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which \rold{there are} >300 \rnew{are} \rchange{tree species}{trees} \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
@@
 
-A small number of studies in SAWs, \rnew{all of which were restricted in the spatial scope to a small region of miombo woodland}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+A small number of studies in southern African woodlands, \rnew{all of which were restricted in the spatial scope to a small region of miombo woodland}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
 The results of these fine scale studies concur with similar studies in other biomes \citep{}.
-Studies of the BEFR often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
-Due to the highly variable environmental conditions within which SAWs occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
+Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
+Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
 
@@
 
-\citet{Solbrig1996} writes that SAWs possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species.
+\citet{Solbrig1996} writes that southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species.
 This structural diversity may be one mechanism through which tree species diversity influences woody biomass.
@@
 
-High levels of disturbance in SAWs may moderate the observable BEFR through its effect on ecosystem composition.
-Fire disturbance in forests has been linked to abundance-dependent mortality among smaller tree stems \citep{Roques2001, Staver2009, Bond2005}.
+High levels of disturbance in southern African woodlands may moderate the observable BEF relationship through its effect on ecosystem composition.
+Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}.
 Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree.
@@
 
-In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands (SAWs), using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
+In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
 We used aboveground woody biomass of trees \rnew{as our metric of ecosystem function}, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
@@
 We used Structural Equation Modelling (SEM) \rchange{and}{with} path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass.
-Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stem stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity \rnew{(i.e.
+Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity \rnew{(i.e.
 physiognomic diversity)}, which will provide an indirect path by which tree diversity increases woody biomass.
@@
 The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the \rchange{so-called miombo woodland eco-region}{miombo ecoregion} (\autoref{clust_map}, \citealt{White1987}).
-The study \rchange{region spans}{area spacs the core climate space of the region}, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
+The study \rchange{region spans}{area spans the core climate space of the region}, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
 The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \rold{as defined by} \citep{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
@@
 Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup.
-Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands (SAWs) \citep{Ratnam2011, Hill2010}.
+Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
 Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
-Only plots with a stem density >10 stems ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
+Only plots with a stocking density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
 
@@
 All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements.
-A tree may be comprised of multiple stems, but for this analysis each stem is treated as an individual.
-For each stem, we \rchange{measured}{recorded} species, DBH and tree height to the top of the highest branch material.
+A tree may be comprised of multiple stems and so tree-level diversity estimates were used to prevent bias towards species which readily coppice.
+For each tree, we \rchange{measured}{recorded} species, DBH and tree height to the top of the highest branch material.
 Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks.
@@
 Wood density for species without species level estimates was estimated from the means of their respective \rchange{genus}{genera}.
+For stems where tree height was unknown, the plot location was used to estimate tree height, according to \citet{Chave2014}.
 
@@
 A key aspect to our decision to use SEMs is that they can explicitly model and partition variance \rnew{attributed} to indirect effects, which is challenging in standard multiple regressions.
-Using SEMs also allowed us to describe latent variables such as ``resource availability'' and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset.
-Structural equation modelling is also necessary to properly account for potential feedback mechanisms between aspects of climate and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
+Using SEMs also allowed us to describe latent variables such as ``resource availability'', ``soil fertility'', and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset.
+SEM is also necessary to properly account for potential feedback mechanisms between aspects of \rchange{climate}{environment} and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to the covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
 
-Prior to analysis, we specified a conceptual model with factors expected to affect AGB: moisture availability, soil fertility, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).
+Prior to analysis, we specified a conceptual model with factors expected to affect AGB: moisture availability, soil fertility, disturbance through fire, tree species diversity, tree structural diversity and stocking density (\autoref{con_mod}).
 
@@
 \centering
-	\input{include/path_diagram_concept}
-	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, tree stem density, and AGB.
-Hypothesised paths of causation are depicted as arrows from predictor to response.
-Correlations are depicted as curved dotted arrows.}
+\includegraphics[width=\textwidth]{concept}
+	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stocking density, and AGB.
+Hypothesised paths of causation are depicted as arrows from predictor to response.}
 	\label{con_mod}
@@
 To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables had their sign reversed.
-For example, soil fertility is expected to decrease as soil sand content increases, so soil percentage sand content was reversed for model fitting.
-Precipitation seasonality (PS), temperature seasonality (TS), and mean annual temperature (MAT) were also reversed in this way to account for the direction of their effect on moisture availability.
+For example, overall moisture availability is expected to decrease as soil sand content increases, therefore sand content was reversed for use in the moisture availability latent variable.
+Precipitation seasonality, and temperature stress were also reversed in this way to account for the direction of their effect on moisture availability.
 
@@
 
-The factor loadings of the observed variable assumed to contribute most to each latent variable were set to 1 as per convention, with other observed variables being allowed to vary \citep{Beaujean2014}.
-We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to 1, while ensuring no factor loadings were in excess of 1.
+The factor loadings of the observed variable assumed to contribute most to each latent variable were set to one, as per convention, with other observed variables being allowed to vary \citep{Beaujean2014}.
+We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to one, while ensuring no factor loadings were in excess of one.
 We found no significant difference between model specifications (p>0.05).
@@
 
-We assessed the role of tree structural diversity and species diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
-To account for variation in stem density, which may covary with species diversity, we also included it as a predictor in our model.
+We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
+To account for variation in stocking density, which may covary with species diversity, we also included it as an observed variable in our model.
 To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately.
-We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB.
+We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stocking density and AGB.
 Path coefficients show the effect of a path with other paths of inference held constant.
@@
 
-To explore the hypothesis that complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density.
+To explore the hypothesis that niche complementarity effects increase in strength as stocking density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stocking density.
 We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB.
-We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model as we were deliberately controlling stem density in our subsampling.
-We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled datasets.
+We excluded the effect of stocking density on AGB and the correlation between stocking density and species diversity from this model as we deliberately controlled stocking density in our subsampling.
+We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stocking density of subsampled datasets.
 
-We incorporated environmental covariates into our model to understand the relative effects of moisture availability and soil fertility on AGB both directly and indirectly via species diversity and stem density.
+We incorporated environmental covariates into our model to understand the relative effects of moisture availability, soil fertility and disturbance by fire on AGB both directly and indirectly via species diversity and stocking density.
 We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB.
-Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation clusters were narrow in their climate space, leading to a lack of variance particularly in moisture availability.
+Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of variance particularly in the moisture availability latent variable.
 
@@
 	\caption{Correlation matrix of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
-Correlation coefficients marked by a circle indicate that the 95\% confidence interval of this correlation overlapped zero.
-Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = moisture availability, green = tree species diversity, purple = tree structural diversity.
+Correlation coefficients marked by a circle indicate that the 95\% confidence interval of $r$ overlapped zero.
+Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = disturbance, turquoise = moisture availability, green = tree species diversity, purple = tree structural diversity.
 See \hyperref[appendixb]{Appendix B} for a full assessment of correlation fit statistics.}
@@
 Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
-Stem density had the strongest correlation with AGB of all variables \rnew{considered} (\ccib{}).
-Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except CEC and MAT.
+stocking density had the strongest correlation with AGB of all variables \rnew{considered} (\ccib{}).
+Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency.
 
@@
 As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a weak negative correlation between the seasonality of precipitation and AGB (\ccmcb{}).
-MAT and temperature seasonality (TS) \rnew{were} \rnew{weakly} negatively correlated \rold{weakly} with AGB (MAT: \cctb{}; TS: \cctcb{}).
-As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil organic carbon and AGB (\ccob{}).
+Temperature stress was \rnew{weakly} negatively correlated \rold{weakly} with AGB (\cctcb{}).
+As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil nutrient content and AGB (carbon: \ccob{}, nitrogen: \ccnb{}).
 
-MAP had positive correlations with tree species richness (\ccms{}), evenness (\ccme{}), tree height diversity (\ccmh{}) and tree stem density (\ccmi{}).
-MAT had weak correlations with tree species and structural diversity variables.
-Tree species diversity variables had clear positive correlations with stem density (Species richness: \ccsi{}; Shannon equitability: \ccei{}).
+MAP had positive correlations with all tree species diversity and structural diversity variables.
+Tree species diversity variables had clear positive correlations with stocking density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stocking density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).
 
@@
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effects of stand structural diversity and stem density (\autoref{struc_mod}), species diversity had a small positive direct effect on AGB (\strucdb{}), and indirectly via structural diversity (\strucdsb{}) (\autoref{struc_mod}).
-Tree species diversity had a positive correlation with stem density.
+In an SEM describing the effect of tree species diversity on AGB via the mediating effects of tree structural diversity and stocking density (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}).
 Model fit was good with high factor loadings for all observed variables.
-All path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}).
+All other path coefficients were significant (p <0.01) (\autoref{struc_model_fit_clust_stats}).
 The R\textsuperscript{2} of AGB was \strucrsq{}.
-The strongest direct effect on AGB was from stem density (\strucsib{}).
+The strongest direct effect on AGB was from stocking density (\strucbetaib{}).
 
@@
 \centering
-	\input{include/path_diagram_struc}
+	\includegraphics[width=0.5\textwidth]{struc}
 	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all vegetation clusters.
-Latent variables are shown as shown as shown as shown as shown as shown as shown as shown as shown as circles while observed variables are shows as rectangles.
+Latent variables are shown as circles while observed variables are shows as rectangles.
 Standardised path coefficients are shown as solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
-The observed variables that inform the latent variables are connected by dotted arrows, and observed variables with loadings set to 1 are connected by dashed arrows.
-Correlations between variables are depicted as dotted curved arrows.
+The observed variables that inform the latent variables are connected by dotted arrows, and observed variables with loadings set to one are connected by dashed arrows.
 Measurement errors of exogenous variables are omitted for clarity.}
@@
 
-When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the 4 vegetation types, the strengths of unstandardised path coefficients varied.
-The direct effect of tree species diversity on AGB was positive in Acacia and Mopane, but negative in Marginal and Core miombo (\autoref{struc_model_slopes_all}).
-Relationships between structural diversity and AGB remained generally similar, with the same sign and significant overlap between the 95\% confidence intervals of path coefficients.
+When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the four vegetation types, model fit and the strengths of unstandardised path coefficients varied.
+The direct effect of tree species diversity on AGB was positive and marginally significant in Acacia (\strucbetacsb{}) but negligible in Mopane (\strucbetadsb{}), Marginal miombo (\strucbetaasb{}) and Core miombo (\strucbetabsb{}) (\autoref{struc_model_slopes_all}).
+Relationships between structural diversity and AGB remained generally similar, with the same sign and overlap between the 95\% confidence intervals of path coefficients.
+The R\textsuperscript{2} of AGB was highest in Acacia shrubland (R\textsuperscript{2} = \struccrsq{}) and lowest in Marginal miombo (R\textsuperscript{2} = \strucarsq{}).
 The total effect of species diversity on AGB remained strongly positive for all vegetation types.
-All vegetation types except Mopane exhibited a positive effect of species diversity on structural diversity.
-All models had adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though confidence intervals around the unstandardised path coefficients were wide particularly for Mopane and Baikiaea.
+All vegetation types exhibited a positive effect of species diversity on structural diversity.
+All models had adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though confidence intervals around the unstandardised path coefficients were wide particularly for Mopane and Acacia.
 $\chi^{2}$ statistics were high for some vegetation types, but this appears to be highly correlated with sample size for each vegetation type \citep{Hooper2008}.
-
-The strongest total effect of tree species diversity on AGB was in Baikiaea woodland (\struccsb{}), which was species rich but highly variable in species diversity compared to other vegetation types (\autoref{clust_summ}).
-The R\textsuperscript{2} of AGB was highest in Marginal miombo (R\textsuperscript{2} = \strucarsq{}) and lowest in the Core miombo (R\textsuperscript{2} = \strucbrsq{}).
 
@@
 
-\subsection{Moderation of Diversity-AGB relationship by stem density}
+\subsection{Moderation of Diversity-AGB relationship by stocking density}
 
-In our sub-sampling of the plot dataset by mean stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}).
-There appears to be a minimum stem density threshold at \textapprox{}180 stems ha\textsuperscript{-1} below which there appears to be a reasonably constant low baseline effect of tree diversity on biomass.
-The effect of structural diversity on AGB appears to remain constant with increasing stem density.
-The indirect effect of species diversity on AGB via structural diversity climbs slightly as stem density increases.
+In our sub-sampling of the plot dataset by stocking density, we found an increasing positive effect of tree species diversity on AGB as stocking density increased (\autoref{sem_struc_stems_ha}).
+There appears to be a minimum stocking density threshold at \textapprox{}180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass.
+The effect of structural diversity on AGB appears to remain constant with increasing stocking density.
+The indirect effect of species diversity on AGB via structural diversity climbs slightly as stocking density increases.
 
@@
 	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves with standard error shaded bars.}
+	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stocking density. Smoothed lines are loess curves with standard error shaded bars.}
 	\label{sem_struc_stems_ha}
@@
 
-A model incorporating the latent variables of moisture availability and soil fertility showed that the total effect of tree species diversity on biomass was greater than that of both moisture availability and soil fertility (\autoref{full_mod}).
-Surprisingly, the direct effects of moisture availability and soil fertility on biomass were negligible, with nearly all of their observed effect on AGB coming from the indirect path via species diversity (moisture: \rgmbd{}, soil: \rgsbd{}).
-MAP and temperature seasonality (TS) had the greatest contributions to the latent variable of moisture availability.
-Moisture availability and soil fertility also had negligible direct effects on stem density.
-Model fit was acceptable: CFI = \fmcfi{}, TLI = \fmtli{}, and RMSEA = \fmrmsea{}, R\textsuperscript{2} of AGB = \fmrsq{}.
+A model incorporating the latent variables of moisture availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of moisture availability, soil fertility and disturbance (\autoref{full_mod}, \autoref{full_model_slopes}).
+Surprisingly, the direct effects of moisture availability, soil fertility and disturbance on AGB were negligible (moisture, \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effect on AGB coming from the indirect path via stocking density (moisture: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (moisture: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}).
+MAP and soil sand content had the greatest contributions to the latent variable of moisture availability.
+Model fit was acceptable: CFI = 0.925, TLI = 0.900, and RMSEA = \fmrmsea{}, R\textsuperscript{2} of AGB = \fmrsq{}.
 
-Similar to the model that only considered tree species and structural diversity (\autoref{struc_mod}), the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB, leading to a strong positive indirect effect of species diversity on AGB via structural diversity (\rgbhd{}).
-The total effect of species diversity on AGB was positive (\rgbd{}).
+Similar to the model that only considered tree species and structural diversity (\autoref{struc_mod}), the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB, leading to a strong positive indirect effect of species diversity on AGB via structural diversity (\fmbetadhb{}) when environmental covariates were accounted for.
+Again, the direct effect of species diversity on AGB was negligible (\fmbetadb{}) The total effect of species diversity on AGB was positive (\fmbetatotaldb{}).
 
@@
 \centering
-	\input{include/path_diagram_full}
+	\includegraphics[width=\textwidth]{full}
 	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation types.
@@
 Standardised path coefficients are shown as solid arrows pointing from predictor to response, with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
-Observed variables that inform the latent variables are connected by dotted arrows, observed variables with loading set to 1 are connected by dashed arrows.
-Correlations between variables are depicted as dotted curved arrows.
+Observed variables that inform the latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
 Measurement errors of exogenous variables are omitted for clarity.}
 	\label{full_mod}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{full_model_slopes}
+	\caption{Unstandardised path coefficients for the full model including tree species diversity, environmental covariates and stocking density.
+Path coefficients are $\pm1$ standard error.
+Path coefficients where the standard error does not overlap zero are considered significant effects.}
+	\label{full_model_slopes}
 \end{figure}
@@
 % Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and AGB.
-% The effect size of species diversity on AGB will increase with stem density, due to an increased importance of niche complementarity as competition increases.
+% The effect size of species diversity on AGB will increase with stocking density, due to an increased importance of niche complementarity as competition increases.
 
-In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, i.e.
-moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across SAWs, using a network of \nplots{} woodland survey plots.
-Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}).
-We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).
-Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}).
-The strongest effect on AGB was that of stem density.
-Interestingly, when the effects of tree species diversity, structural diversity and stem density were controlled for, we found little evidence of a direct effect of resource availability, either in the form of moisture or soil fertility, on AGB (H\textsubscript{3}).
+In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stocking density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots.
+Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB indirectly via structural diversity (H\textsubscript{1}).
+We found that the effect size of tree species diversity on AGB increased with stocking density (H\textsubscript{2}).
+Tree species diversity, structural diversity and stocking density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}).
+The strongest effect on AGB was that of stocking density.
+When the effects of moisture availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.
 
@@
 
-We found a consistent positive effect of tree species diversity on AGB across all models in this study.
-Within SAWs we therefore find support for our hypothesis that higher tree species richness and evenness causes higher woody AGB.
+We found a consistent positive effect of tree species diversity on AGB across all models in the current study.
+Within southern African woodlands we therefore find support for our hypothesis that higher tree species richness and evenness causes higher woody AGB.
 This finding is in agreement with many other studies across different ecosystems and biomes, supporting \rnew{the idea} that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
-Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of SAWs, a disturbance-structured and poorly studied ecological system.
+Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of southern African woodlands, a disturbance-structured and poorly studied ecological system.
 
-% In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stem density exists.
+% In our extrapolation of species richness according to sampling effort (stocking density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stocking density exists.
 We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
 
-Much of the total variation in AGB was driven by variation in stem density.
-Stem density also correlated with species diversity in our SEMs.
-It is possible that within SAWs a higher species diversity allows for a greater density of tree stems, leading to an increase in total AGB.
-The opposite is also plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
-We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level \citep{}.
-This is supported by the moderately strong indirect positive effect of tree species diversity on AGB via structural diversity.
-
-% KGD: There are counter-examples, e.g. Johnson et al. 2016. GCB (from Leeds group, Tim Baker senior author), but feel free to leave text as is for now.
-
-% The question of whether lots of small stems or a few large stems maximises AGB has yet to be solved \citep{}.
-
-% While we did not explicitly measure Net Primary Productivity (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
-This suggests that, as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in SAWs can be maximised by increasing species diversity.
+Much of the total variation in AGB was driven by variation in stocking density.
+It is possible that within southern African woodlands a higher species diversity allows for a higher stocking density through niche separation, which reduces competition between species occupying varying niche space, leading to an increase in total AGB per unit area.
+The opposite causation is also plausible however, with increased stocking density causing higher species richness through an increased probability of encountering new species.
+We attempted to correct for the correlation between species richness and stocking density by estimating the extrapolated species richness.
+We suggest therefore that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can utilise more available light and moisture, resulting in greater plot-level AGB.
+This is supported by the moderately strong indirect positive effect of tree species diversity on AGB via structural diversity, and the positive effect of moisture availability on AGB via stocking density in the model which included environmental covariates.
+However, we found a negative effect of soil fertility on AGB via stocking density.
 
@@
 greater variation in DBH and height.
-This may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
+This may act as a mechanism for niche complementarity, with a highly diverse canopy able to take advantage of a greater proportion of the available light.
 Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{}, i.e.
@@
 emergent canopy, for example.
-Variation in structural diversity may be a joint result of disturbance history and tree species diversity, with highly disturbed plots generally having a less structurally diverse canopy \citep{LaRue2019}.
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
-Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this.
+Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stocking density supports this.
 In frequently disturbed woodlands such as those studied here however, a woodland canopy similar to that of a forest is frequently not reached.
 Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
-While we did not have access to adequate data on disturbance history in our plots, previous studies have found that SAWs with higher species diversity tend to \rnew{experience less frequent disturbance by fire and tend to form a more closed canopy and a more sparse understorey} \citep{Chidumayo2013, Mutowo2012}.
+Previous studies have found that southern African woodlands with higher species diversity tend to \rnew{experience less frequent disturbance by fire and tend to form a more closed canopy and a more sparse understorey} \citep{Chidumayo2013, Mutowo2012}.
+In our study however, we found a positive effect of disturbance on species diversity, perhaps suggesting that disturbance prevents domination of woodlands by a single dominant species \citep{Chidumayo2013}.
 
-We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
-At low stem densities competition between trees may not occur, meaning that the niche complementarity provided by an increase in tree species richness might not make any difference to plot level AGB, accounting for the low and constant effect of tree species diversity on AGB below \textapprox{}180 stems ha\textsuperscript{-1}.
+We found a non linear positive effect of stocking density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
+At low stocking densities competition between \rnew{mature} trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the low and constant effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}.
+At very high stem density, there is also increase in the effect of species diversity on structural diversity.
+This could be because at high stem density, the adaptation of different species to growth form become important.
+At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species ara able to exist in the understory, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.
 
-\subsection{Effects of moisture availability and soil fertility}
+\subsection{Effects of moisture availability, soil fertility and disturbance}
 
-Surprisingly, moisture availability and soil fertility had only small effects on AGB compared to that of tree species diversity.
-We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for a greater stem density per unit area, greater productivity per unit area and additionally greater tree species diversity due to niche partitioning \citep{Kraaij2006, Shirima2015}.
-
+Moisture availability, soil fertility and disturbance by fire had similar positive total effects on AGB, comparable in size to the total effect of tree species diversity on AGB.
+We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for a greater stocking density per unit area, greater productivity per unit area and additionally greater tree species diversity due to niche partitioning \citep{Kraaij2006, Shirima2015}.
 Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via increasing tree species diversity and via increasing stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}.
-In this study, while we observed weak indirect effects via species diversity, we saw no evidence for a direct effect of moisture availability on AGB.
-Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in SAWs, which \rchange{are frequently droughted}{experience frequent drought}.
-It is possible that the range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture variation in AGB.
-Due to the high levels of adaptation of tree species to drought conditions in southern Africa, at the large scale we conducted our experiment turnover in species composition along the moisture gradient may have obscured a direct relationship being observed between moisture availability and AGB.
-
-In SAWs moisture availability is closely linked with the intensity of disturbance from seasonal fires.
+In this study, we observed weak indirect positive effects on AGB via species diversity and a positive direct effect on AGB.
+Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in southern African woodlands, which \rchange{are frequently droughted}{experience frequent drought}.
+It is surprising that disturbance by fire had a positive total effect on AGB.
+We found a positive indirect effect of disturbance on AGB via species diversity and structural diversity.
+It is possible that fire prevents the domination of a woodland by a few highly competitive species, resulting in greater abundance evenness and overall species richness, which in turn increases the effect of niche complementarity.
+In southern African woodlands moisture availability is linked to the intensity of disturbance from seasonal fires.
 The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth \citep{Charles-Dominique2018}, and may also limit species diversity \citep{Linder2014}.
-It is possible therefore that the effect of moisture availability, which is expected to increase AGB, is confounded in its effect on AGB with the unmeasured variable of fire regime intensity, which is expected to decrease AGB.
-The direct effect of moisture availability on stem density may also be confounded in this way.
+It is possible therefore that the effect of moisture availability, which is expected to increase AGB, is confounded in its effect on AGB with fire frequency, which had a negative effect on stocking density, which is expected to decrease AGB.
+The direct effect of moisture availability on stocking density may also be confounded in this way.
 This may also have caused us to not observe a stronger effect between moisture availability and AGB.
-
-We expected a positive effect of soil fertility on AGB, but found no evidence of this in our models.
-We measured soil fertility using the observed variables of soil organic carbon content, sand particle content and Cation Exchange Capacity (CEC).
+We found a positive total effect of soil fertility on AGB in our model.
+We measured soil fertility using the observed variables of soil carbon, nitrogen and Cation Exchange Capacity (CEC).
 In wet tropical forests a clear relationship has been observed between these variables and AGB \citep{Slik2009, Coelho-de-Souza2019}.
-% KGD: You have written little about soil fertility in this section, but I think you need to.
 
@@
 
-% KGD: See thoughts above for C3 results.
-C2 should be discussed even if we cannot give a good reason.
-It is very intriguing.
-The results for C4 also surprise me, as I would have expected mopane to be too species poor to be able to show a result.
-You can discuss how many mopane stands are monospecific and even thought such things as 'cathedral mopane' exist, where biomass can be quite high, it seems that adding even a few species to mopane can increase its total AGB (of the stands).
-That is cool and shows robustness of results.
-Basically, the fact that C3 and C4 show significant results means that your results are robust across a range of species richnesses (not just do to patterns at low, 1-6, or high, 6-12, species richness).
-I still have a great sense of what C1 is, so with withhold commentary on that for the moment.
-If it is 'disturbed miombo', then maybe something can be said about the contrast between results for C1 and C2 and the role that disturbance may play.
+All four vegetation types produced similar results in the simple SEM, with a positive total effect of species diversity on AGB, the majority being indirectly via structural diversity.
+This demonstrates the robustness of our results, showing they are generalisable across vegetation types in southern Africa.
+It also demonstrates that similar ecosystem processes are occurring in these vegetation types, despite variation in species composition and overall species richness.
 
 Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via structural diversity, remained positive in these vegetation types.
-Compared to Baikiaea and Mopane woodlands, miombo woodlands have higher median tree species richness.
-Baikiaea and Mopane woodlands are also dominated by fewer tree species, notably \textit{Baikiaea plurijuga} in Baikiaea woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees.
-We postulate that this negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in Baikiaea and Mopane woodlands, where the woodland canopy is dominated often by a single species.
-Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation.
-Again, these highly diverse plots in miombo woodlands may be the result of disturbance which can promote a mosaic of woodland of different successional stages and stem densities.
-Alternatively, this small negative direct effect may be an artefact of particularly noisy data, especially given that the overall effect of tree diversity on AGB is positive.
+Compared to Acacia and Mopane woodlands, miombo woodlands have higher median tree species richness.
+Acacia and Mopane woodlands are dominated by fewer tree species, notably \textit{Senegalia} spp.
+in Acacia woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees.
+We postulate that the negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in Acacia and Mopane woodlands, where the woodland canopy is dominated often by a single species.
+Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation, while in the species poor Mopane and Acacia woodlands, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.
 
@@
 
-Baikiaea woodland had the strongest total effect of species diversity on AGB.
-Baikiaea also has relatively low median species richness compared to miombo, but the addition of new species appears to make a larger difference to the AGB of these plots than in mopane woodlands.
-We suggest that this is due mostly to the particular identity of species found in Baikiaea woodlands and their contribution to ecosystem functioning.
-Unlike mopane woodlands, Baikiaea woodlands do sometimes contain species other than \textit{B.
-plurijuga} which grow to be high biomass canopy trees.
+Acacia woodland had the strongest total effect of species diversity on AGB and was the only vegetation type to show a clearly positive direct effect of species diversity on AGB.
+Acacia also has relatively low median species richness compared to miombo, but the addition of new species appears to make a larger difference to the AGB of these plots than in mopane woodlands.
+We suggest that this is due mostly to the particular identity of species found in Acacia woodlands and their contribution to ecosystem functioning.
+Unlike mopane woodlands, Acacia woodlands do sometimes contain species other than \textit{Senegalia} spp.
+which can grow to be high biomass canopy trees, especially in transition zones.
 
@@
 
-In this study we found that across southern African woodlands (SAWs), there is a generalisable positive association between tree species diversity and woody biomass as a measure of ecosystem function.
-Additionally, we found that much of this effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of woodland tree canopies.
-We found that the multiple vegetation types which comprise SAWs exhibit variation in the strength of the relationship between species diversity and woody biomass, inferring that models of regional and global biodiversity-ecosystem function relationships could benefit from including vegetation type terms and the structural properties of those vegetation types, such as structural diversity and stem density.
-In contrast to previous studies, we found that across the region, the direct effects of moisture availability and soil fertility on woody biomass were negligible, with most of their effect being indirectly through species and structural diversity.
-A gap in available data means that we could not incorporate disturbance history into our models adequately, but this factor likely plays a large part in the association between species diversity and woody biomass in SAWs.
+In this study we found that across southern African woodlands, there is a generalisable positive association between tree species diversity and woody biomass as a measure of ecosystem function.
+Additionally, we found that much of this effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of woodland trees.
+We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, inferring that similar ecosystem processes occur across the region to determine ecosystem function.
+In contrast to previous studies, we found that across the region, the direct effects of moisture availability and soil fertility on woody biomass were negligible, with most of their effect being indirectly through tree species and structural diversity.
 
-SAWs are relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem function.
+Southern African woodlands are relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem function.
 Resource extraction by humans in southern Africa is directly influencing biodiversity via selective tree-felling for timber, among other forest products.
-Our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stem density and Baikiaea woodlands, which are predominantly targeted for tree felling.
+Our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stocking density, and low species diversity, which are predominantly targeted for tree felling.
 This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.
-
-% KGD: these conclusions are a bit dry and repetitive on earlier text.
-What do our results mean for woodland management?
-for conservation?
-for our understanding of basic ecological theory?
-As an aside, we have circled back around to niche complementarity in this discussion, but not really to selection effects or facilitation.
-% Southern African woodlands are relied upon heavily for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
-This has raised interest in how biodiversity influences ecosystem function in these ecosystems.
-Resource extraction by humans directly influences biodiversity via selective tree-felling for timber, charcoal making, non-timber forest products and through land use change to agriculture \citep{Aleman2016, Ryan2016}.
-Climate change is also indirectly affecting the biodiversity of southern African woodlands, altering temperature and precipitation, and affecting climate seasonality which heavily influences the degree of seasonal drought and thus woodland structure \citep{Scholes2004, Eldridge2012}.
-While rapid biodiversity change is being observed in southern African woodlands \citep{Syampungani2009}, research into the relationship between biodiversity and ecosystem functionality remains scarce.
 
@@
 \centering
-	\includegraphics[width=\textwidth]{bivar_lm}
-	\caption{}
+	\includegraphics[width=0.9\textwidth]{bivar_lm}
+	\caption{Bivariate scatter plots for each observed variable used in the SEMs, based on hypothesised paths of causality.
+Points are coloured according to vegetation type.
+A single linear regression is presented as a black line, which combines all vegetation types, separate loess trend lines are fitted for each vegetation type.
+All data is standardised and variables are transformed where it was appropriate for analysis.}
 	\label{bivar_lm}

### 2020-07-02 scripts/sem.R

@@ -46,6 +46,23 @@ library(MVN)
 
 source("scripts/clust_defin.R")
 
+p_format <- function(p){
+  case_when(p < 0.01 ~ paste0("p <0.01"),
+    p < 0.05 ~ paste0("p <0.05"),
+    TRUE ~ paste0("p = ", round(p, digits = 2)))
+}
+
+corr_format <- function(id, row_num){
+  paste0("\\newcommand{\\", id, "}{", "$r =$ ", corr_ci_tab$raw.r[row_num], ", ", corr_ci_tab$p[row_num], "}")
+}
+
+beta_format <- function(x, id, row_num) {
+    paste0("\\newcommand{\\", id, "}{", 
+      "$\\beta =$ ", round(x[row_num, "est"], 2),
+      "$\\pm$", round(x[row_num, "se"], 3),", ", 
+      p_format(x[row_num, "pvalue"]), "}")
+}
+
 # Import data ----
 
 dat <- readRDS("data/plot_data_fil_agg_norm_std.rds")
@@ -55,7 +72,7 @@ dat <- dat %>%
   mutate(
     precip_seas_rev_std = -1 * precip_seas_std,
     temp_stress_rev_std = -1 * temp_stress_std,
-    sand_rev_std = -1 * sand_std,
+    sand_rev_std = -1 * sand_std, 
     shannon_equit_rev_std = -1 * shannon_equit_std)
 
 # Correlation matrix between variables ----
@@ -65,7 +82,7 @@ corr_df <- dat %>%
   dplyr::select(soil_c_log_std, cec_std, nitrogen_log_std,
     fire_log_std, 
     precip_std, precip_seas_std, temp_stress_std, sand_std,
-    n_species_raref_log_std, shannon_equit_std,
+    n_species_raref_log_std, shannon_equit_rev_std,
     cov_height_std, cov_dbh_std, 
     n_trees_gt10_ha_log_std, agb_ha_log_std)
 
@@ -88,17 +105,17 @@ corr_ci$x_var <- unlist(sapply(1:(length(corr_df)-1), function(i){row.names(corr
 corr_ci$x_var <- factor(corr_ci$x_var, 
   levels = c("cec_std", "nitrogen_log_std", "fire_log_std", 
     "precip_std", "precip_seas_std", "temp_stress_std", "sand_std", "n_species_raref_log_std",
[…]

### 2020-07-15 manuscript/include/clust_summ.tex
Updated with final comments

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Wed, Jul 01, 2020 - 14:44:00
+% Date and time: Wed, Jul 15, 2020 - 09:38:19
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-Acacia shrubland & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\ \textit{Senegalia burkei} \\ \textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\ \textit{Vachellia nilotica} \\ \textit{Spirostachys africana} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\
+Core miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Brachystegia spiciformis} \\\textit{Julbernardia paniculata} \\\textit{Brachystegia boehmii} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Parinari curatellifolia} \\\textit{Uapaca kirkiana} \\\textit{Brachystegia spiciformis} \end{tabular} & 523 & 20(16.9) & 204(142.5) & 44.2(36.11) \\
 \hline
-Core miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Brachystegia spiciformis} \\ \textit{Julbernardia paniculata} \\ \textit{Brachystegia boehmii} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Parinari curatellifolia} \\ \textit{Uapaca kirkiana} \\ \textit{Brachystegia spiciformis} \end{tabular} & 523 & 20(16.9) & 204(142.5) & 44.2(36.11) \\
+ex-Acacia & \begin{tabular}[c]{@{}c@{}c@{}} Spirostachys africana \\\textit{Senegalia burkei} \\\textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\\textit{Vachellia nilotica} \\\textit{Spirostachys africana} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\
 \hline
-Marginal miombo & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\ \textit{Burkea africana} \\ \textit{Pterocarpus angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Burkea africana} \\ \textit{Baikiaea plurijuga} \\ \textit{Pterocarpus angolensis} \end{tabular} & 466 & 12(13.7) & 178(129.5) & 36.9(26.98) \\
+Mopane & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\\textit{Androstachys johnsonii} \\\textit{Kirkia acuminata} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{\textit{Colophospermum mopane}} \\\textit{Psuedolachnostylis maprouneifolia} \\\textit{Lannea discolor} \end{tabular} & 58 & 10(10.2) & 186(125.6) & 42.7(32.83) \\
 \hline
-Mopane & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Colophospermum mopane} \\ \textit{Androstachys johnsonii} \\ \textit{Kirkia acuminata} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{\textit{Colophospermum mopane}} \\ \textit{Psuedolachnostylis maprouneifolia} \\ \textit{Lannea discolor} \end{tabular} & 58 & 10(10.2) & 186(125.6) & 42.7(32.83) \\
+Sparse miombo / Baikiaea & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Baikiaea plurijuga} \\\textit{Burkea africana} \\\textit{Pterocarpus angolensis} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Burkea africana} \\\textit{Baikiaea plurijuga} \\\textit{Pterocarpus angolensis} \end{tabular} & 466 & 12(13.7) & 178(129.5) & 36.9(26.98) \\
 \hline

### 2020-07-15 manuscript/include/struc_model_fit_clust_stats.tex
Updated with final comments

@@
 % Table created by stargazer v.5.2.2 by Marek Hlavac, Harvard University. E-mail: hlavac at fas.harvard.edu
-% Date and time: Thu, Jul 02, 2020 - 11:10:18
+% Date and time: Wed, Jul 15, 2020 - 10:35:08
 \begin{table}[!htbp] \centering 
@@
 \hline \\[-1.8ex] 
-Marginal miombo & 390 & 43.870 & 6 & 0.914 & 0.784 & 0.130 & 0.460 \\ 
-Core miombo & 517 & 78.670 & 6 & 0.904 & 0.759 & 0.140 & 0.490 \\ 
-Acacia shrubland & 28 & 9.570 & 6 & 0.952 & 0.879 & 0.130 & 0.830 \\ 
-Mopane & 46 & 19.880 & 6 & 0.834 & 0.584 & 0.240 & 0.510 \\ 
-All & 981 & 91.380 & 6 & 0.937 & 0.843 & 0.120 & 0.490 \\ 
+Sparse miombo / Baikiaea & 466 & 43.870 & 6 & 0.914 & 0.784 & 0.130 & 0.580 \\ 
+Core miombo & 523 & 78.670 & 6 & 0.904 & 0.759 & 0.140 & 0.490 \\ 
+ex-Acacia & 188 & 9.570 & 6 & 0.952 & 0.879 & 0.130 & 0.830 \\ 
+Mopane & 58 & 19.880 & 6 & 0.834 & 0.584 & 0.240 & 0.510 \\ 
+All & 1235 & 91.380 & 6 & 0.937 & 0.843 & 0.120 & 0.490 \\ 
 \hline \\[-1.8ex] 

### 2020-07-15 manuscript/regional_befr_manuscript.tex
Updated with final comments

@@
 
-\usepackage[revision]{revdiff}
+\usepackage[clean]{revdiff}
 
@@
 
-Authors: Godlee, J. L.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1}, Ryan, C. M.\textsuperscript{1}
+Authors: Godlee, J. L.\textsuperscript{1}, Ryan, C. M.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1}
 
@@
 NE/L002558/1).
-The data provided for this study was constributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands), an activity of the Miombo Network and NERC-funded project (Grant No.
+The data provided for this study was constributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No.
 NE/P008755/1).
@@
 
-\textbf{Aim:} Positive correlations between tree \rchange{biodiversity}{species diversity} and ecosystem function have been widely documented, but the nature of the relationship \rchange{in highly disturbed and ecophysiologically stressful systems is less clear}{in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear}.
-\rnew{In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function.} Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, \rchange{with respect to}{while controlling for} gradients in stocking density, resource availability\rnew{, disturbance through fire}, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+\textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear.
+In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function.
+Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, while controlling for gradients in stem density, resource availability, disturbance through fire, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
 
@@
 
-\textbf{Methods:} We used a network of \nplots{} \rchange{savanna and woodland}{savanna/woodland tree} plots located across the southern African sub-continent \rold{in which each tree >10 cm diameter was measured and identified to species level}.
-We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability \rnew{, disturbance by fire}, and stocking density.
+\textbf{Methods:} We used a network of \nplots{} savanna/woodland tree plots located across the southern African sub-continent .
+We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, and stem density.
 
-\textbf{Results:} \rnew{We found} a positive effect of tree species diversity on aboveground biomass \rold{was demonstrated}, observed \rchange{largely as an}{mainly via the} of increasing woodland structural diversity.
-We also found that the effect of tree species diversity on biomass increases with stocking density.
-Finally, we found that resource availability affects biomass in southern African woodlands \rchange{largely}{mainly} indirectly, via its effect on species diversity.
+\textbf{Results:} We found a positive effect of tree species diversity on aboveground biomass, observed mainly via the increasing effect of woodland structural diversity.
+We also found that the effect of tree species diversity on biomass increases with stem density.
+Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via its effect on species diversity.
 
-\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and \rnew{ecosystem} function of highly disturbed southern African savannas and woodlands.
-Our results demonstrate the importance of \rchange{including}{accounting for} environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level.
-Biodiversity loss, predominantly \rchange{by}{through} human actions in southern Africa may have detrimental outcomes for ecosystem function, particularly in tree species-poor Baikiaea woodlands, which showed the strongest biodiversity - ecosystem function relationship.
-
-% Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard.
-In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots.
-We used structural equation modelling (SEM) to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stocking density.
-A positive effect of tree species diversity on biomass was demonstrated, with most of this as an indirect effect via woodland structural diversity.
-We found that the effect of tree species diversity on biomass increases with stocking density.
-Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.
+\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and ecosystem function of highly disturbed southern African savannas and woodlands.
+Our results demonstrate the importance of accounting for environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level.
+Biodiversity loss, predominantly through human actions in southern Africa, may have detrimental outcomes for ecosystem function, specifically woody biomass provision, which is critical in a region that relies on wood fuel as its principal domestic energy source.
 
@@
 Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
-These studies support early findings from small-scale experimental studies \rnew{conducted} predominantly in grassland patches, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+These studies support early findings from small-scale experimental studies conducted predominantly in grassland ecosystems, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
 Despite these concerted efforts, we continue to lack a nuanced, ecosystem agnostic understanding of the complex interactions between biodiversity, environment, and ecosystem function.
 
-Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the degree of biotic activity within an ecosystem \citep{Jax2005}.
-This includes \rchange{basic processes of primary production}{processes} such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported \rold{and intuitive} relationship between biodiversity and ecosystem function invokes three main mechanisms \rchange{which drive}{to explain} the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
+Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the nature of biotic activity within an ecosystem \citep{Jax2005}.
+This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
+The frequently reported  relationship between biodiversity and ecosystem function invokes three main mechanisms to explain the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
 
-Compared to \rchange{other forested}{forest} ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire \rchange{and in Africa notably by herbivory also}{and herbivory, with African savannas possessing large herbivores absent from other savannas} \citep{Sankaran2008, Levick2009}.
-\rnew{Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}}.
-High levels of disturbance by fire may weaken the role of competition in determinging local species distribution and allow weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
-This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as \rchange{apparent}{important} in dry woodland/savanna ecosystems.
+Compared to forest ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire and herbivory, with African savannas possessing large herbivores absent from other savannas \citep{Sankaran2008, Levick2009}.
+Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}.
+High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution.
+Disturbance may reduce stem density, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
+This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems.
 Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}.
-Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations \rchange{in environments which are more hostile to growth}{under limiting environmental conditions such as low water availability \citep{} or high maximum temperature \citep{}}.
+Reduced stem density reduces competition between individuals, possibly weakening the effect of diversity on ecosystem function via niche complementarity.
+Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations under limiting environmental conditions such as low water availability \citep{}.
 Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, \rnew{as competition diminishes in ecosystems where environmental stress limits individual species reaching the bounds of their fundamental niche space, thus reducing the competition which drives niche complementarity effects.} This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation \rchange{in order to characterise}{if we are to derive} a generalisable BEF relationship.
+They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects.
+This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEF relationship.
 
 The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, which hold many endemic species and unique ecosystem assemblages \citep{Barthlott2005}.
-A severe lack of study in dry tropical ecosystems, especially given the potential mismatch in BEF relationship mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEF relationship and its environmental determinants.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}.
+A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEF relationship and its environmental determinants.
+
 Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
 The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
-Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages\rchange{, which}{.These} retain the common features of an open tree canopy and an understorey generally dominated by C4 \rchange{grass species}{grasses}.
-Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which \rold{there are} >300 \rnew{are} \rchange{tree species}{trees} \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages.These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses.
+Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
@@
 
-A small number of studies in southern African woodlands, \rnew{all of which were restricted in the spatial scope to a small region of miombo woodland}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+A small number of studies in southern African woodlands, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
 The results of these fine scale studies concur with similar studies in other biomes \citep{}.
 Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
-Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, \rnew{with wide variation in precipitation, diurnal and annual temperature range} and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
-
-%Environmental conditions may also indirectly moderate the observed relationship between biodiversity and ecosystem function \citep{Vila2005}.
-
-% Kunz et al.
-(2019) found that crown complementarity and crown plasticity increased with species richness.
-Trees growing in species rich neighbourhoods had enhanced AGB production in trunk and branches.
-
-%While southern African woodlands are species rich in the herbaceous understorey \citep{Murphy2016}, the tree layer is relatively species poor.
+Due to the variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, with wide variation in precipitation, diurnal and annual temperature range, disturbance regime, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
 
 In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and tree species diversity independently \citep{Spasojevic2014}.
-It is important therefore to account for climatic factors and understand how they interact with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass in analyses at broad spatial scales.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that \rnew{below a threshold of \textapprox{}650 mm MAP}, precipitation sets sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
-Similarly, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness in a wet Panamanian tropical forest.
+It is important therefore to account for climatic factors and understand how they interact with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass.
+\citet{Sankaran2005} used data from 854 African woodland field sites to show that below a threshold of \textapprox{650 mm MAP}, precipitation sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
+In wet tropical forests in Central America, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness.
 
-\citet{Solbrig1996} writes that southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy at different growth stages and among species.
+Southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy, depending on their growth stages and species identity \citet{Solbrig1996}.
 This structural diversity may be one mechanism through which tree species diversity influences woody biomass.
@@
 They also found that trees growing in species-rich neighbourhoods exhibited enhanced biomass production.
-\rchange{Occupation}{Occupancy} of multiple canopy layers allows a \rchange{more full}{fuller} canopy with \rold{a} greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity.
+Occupancy of multiple canopy layers allows a fuller canopy with  greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity.
 This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas.
@@
 Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands.
-If the regional species pool contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.
+If a given woodland plant community contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.
 
 In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
-We used aboveground woody biomass of trees \rnew{as our metric of ecosystem function}, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
+We used aboveground woody biomass of trees as our metric of ecosystem function, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
 We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
 We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results.
-We used Structural Equation Modelling (SEM) \rchange{and}{with} path analysis as a preferred method to simultaneously account for environmental and biotic factors, \rchange{which may interact their effect}{which may have interacting effects} on ecosystem structure and therefore biomass.
-Initially, we \rchange{made}{posited} three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) \rchange{the effect size}{the strength of the effect} of tree species diversity on woody biomass will increase with plot level stocking density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity \rnew{(i.e.
-physiognomic diversity)}, which will provide an indirect path by which tree diversity increases woody biomass.
+We used Structural Equation Modelling (SEM) with path analysis as a preferred method to simultaneously account for environmental and biotic factors, which may have interacting effects on ecosystem structure and therefore biomass.
+Initially, we posited three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the strength of the effect of tree species diversity on woody biomass will increase with plot-level stem density (number of stems ha\textsuperscript{-1}), as affected by resource availability and disturbance, due to an increased importance of niche complementarity as stem density and therefore competition increases, and (3) tree species diversity will increase tree structural diversity (i.e.
+physiognomic diversity), which will provide an indirect path by which tree diversity increases woody biomass.
 
@@
 
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the \rchange{so-called miombo woodland eco-region}{miombo ecoregion} (\autoref{clust_map}, \citealt{White1987}).
-The study \rchange{region spans}{area spans the core climate space of the region}, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
-The 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \rold{as defined by} \citep{White1987}, using WorldClim estimates of temperature and precipitation between the year 1970 and 2000 with a pixel size of 30 arc seconds (0.86 km\textsuperscript{2} at the equator) \citep{Fick2017}.
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealt{White1987}).
+The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
+A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion  \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.
 
-Plots were chosen from a larger pool of 5395 plots based on the quality and completeness of data collection, and plot setup.
+Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup.
 Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
@@
 Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
-Only plots with a stocking density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots were within woodland rather than ``grassy savanna'', which are considered a separate biome with very different species composition \citep{Parr2014}.
+Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than ``grassy savanna'', which is considered a separate biome with very different species composition \citep{Parr2014}.
 
@@
 
-\rold{After the initial plot data cleaning described above, we conducted an outlier removal procedure of plots with rare tree species composition.
-We used the outlier() function from the dave R package citep{dave}, which uses a nearest neighbour criterion for each plot in species abundance ordination space and a threshold value for the minimum nearest neighbour distance to identify outliers.
-We set the threshold value to remove the top 5\% of plots with the largest nearest neighbour distances in multidimensional species composition space citep{Otto2013}, thus removing noutliers{} plots (hyperref[appendixa]{Appendix A}).}
-
 \subsection{Data collection}
 
-We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads which are functionally tree-like, but excluding lianas, which fill a different ecological niche \citep{Selaya2008}.
+We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads which are functionally tree-like.
+Woody lianas are scarce in our study plots and were not measured.
 Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
@@
 All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements.
-A tree may be comprised of multiple stems and so tree-level diversity estimates were used to prevent bias towards species which readily coppice.
-For each tree, we \rchange{measured}{recorded} species, DBH and tree height to the top of the highest branch material.
+A tree may be comprised of multiple stems and so tree-level diversity estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice.
+For each tree, we recorded species, DBH and tree height to the top of the highest branch material.
 Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks.
@@
 Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
-Wood density for species without species level estimates was estimated from the means of their respective \rchange{genus}{genera}.
-For stems where tree height was unknown, the plot location was used to estimate tree height, according to \citet{Chave2014}.
+Wood density for species without species level estimates was estimated from the means of their respective genera.
+For stems where tree height was unknown, the plots' climatic parameters, estimated from plot location, were used to estimate tree height, according to \citet{Chave2014}.
 
-Climatic data were collected from the ECMWF ERA5 dataset, generated using Copernicus Climate Change Service Information \citep{ERA5}.
-Values of \rchange{Mean Annual Temperature (MAT)}{MAT} and \rchange{Mean Annual Precipitation (MAP)}{MAP} was calculated from daily data between 2000 and 2018, then averaged across years to provide a single mean annual estimate per plot.
-Temperature seasonality (TS) and precipitation seasonality (PS) were both calculated as the mean of the coefficient of variation of daily mean temperature and precipitation, respectively, for each of the 18 years of available data.
-Soil fertility data was extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}.
-We extracted Cation Exchange Capacity (CEC), percentage soil organic carbon by volume (Org.
-C \%), and percentage soil sand content by volume (Sand \%).
+Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}.
+In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15).
+Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}.
+We extracted Cation Exchange Capacity (CEC), soil organic carbon stocks (kg m\textsuperscript{-2}) soil sand content by volume (Sand g kg\textsuperscript{-1}) and soil nitrogen content (g kg\textsuperscript{-1}).
 These data are a modelled product derived from various remotely sensed and directly measured data sources.
-
-% \todo{Fire return interval ($F_{ret}$) was calculated using the MODIS burned area product V6 (MCD46A1) \citep{}.
-Data was downloaded from January 2000 to December 2018.
-Mean fire return interval was calculated as:}
-% 
-% \begin{equation}
-% 	\todo{F_{ret} = \frac{\sum_{i = 1}^{n} t_{i} - t_{i-1}}{n-1}}
-% \end{equation}
-% 
-% \todo{Where $t_{i}$ is the date of fire $i$, and $n$ is the total number of fires in the time period.}
+The degree of fire disturbance was calculated using the MODIS monthly burned area product at 0.01 degree resolution, 1.11 km at the equator (MCD64A1, \citealt{MODIS_burn}), as the number of observed fires between 2001 and 2018.
 
@@
 Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.
-To measure tree species \rold{abundance} evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness.
-Abundance evenness allows for greater niche complementarity at small scales due to an increased spatial heterogeneity of functional traits.
-We calculated tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).
-
-% \begin{equation}
-% 	\begin{gathered}
-% 		E_{H'} = \frac{H'_{e}}{\ln{S}} \\
-% 	\end{gathered}
-% 	\label{shannon_equit}
-% \end{equation}
-% 
-% Where $H'_{e}$ is an estimation of the Shannon diversity index of trees by extrapolation of the observed Shannon diversity index ($H'$) to its asymptote via Hill numbers using the \verb|ChaoShannon()| function from the \verb|iNEXT| package in R \citep{Hsieh2016}, and $S$ is the extrapolated tree species richness in the plot, using the \verb|ChaoRichness()| function.
+To measure tree species  evenness, the Shannon Equitability index ($E_{H'}$) \citep{Smith1996} was calculated as the ratio of the estimated Shannon diversity index to the natural log of estimated species richness.
+Abundance evenness allows for greater niche complementarity at small scales due to potentially increased heterogeneity of functional traits.
+We quantified tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).
 
@@
 Groups were defined in \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
-Group identification was conducted using unconstrained correspondence analysis\rold{ and ordination}, followed by clustering based on dominant ordination axes.
-Plot data used in this study occurred in four vegetation type groups.
-See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters .
+Group identification was conducted using unconstrained correspondence analysis, followed by clustering based on dominant ordination axes.
+Plot data used in this study occurred in four compositional vegetation types.
+See \autoref{clust_summ} for a description of each vegetation cluster and \autoref{clust_map} for the spatial distribution of plots from each of these clusters.
+Cluster names were assigned post-hoc based on the dominant and indicator species in each cluster.
 
@@
 
-\rnew{We used} Structural Equation Models (SEM) to investigate the determinants of AGB.
+We used Structural Equation Modelling (SEM) to investigate the determinants of AGB.
 All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}.
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
-A key aspect to our decision to use SEMs is that they can explicitly model and partition variance \rnew{attributed} to indirect effects, which is challenging in standard multiple regressions.
-Using SEMs also allowed us to describe latent variables such as ``resource availability'', ``soil fertility'', and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having observable measures in our dataset.
-SEM is also necessary to properly account for potential feedback mechanisms between aspects of \rchange{climate}{environment} and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to the covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
+A key aspect to our decision to use SEM is that they can explicitly model and partition variance attributed to indirect effects, which is challenging in standard multiple regressions.
+Using SEMs also allowed us to describe latent variables such as ``water availability'', ``soil fertility'', and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having directly observable measures in our dataset.
+SEM is also necessary to properly account for potential feedback mechanisms between aspects of environment and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to the covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
 
-Prior to analysis, we specified a conceptual model with factors expected to affect AGB: moisture availability, soil fertility, disturbance through fire, tree species diversity, tree structural diversity and stocking density (\autoref{con_mod}).
+Prior to analysis, we specified a conceptual model with factors expected to affect AGB: water availability, soil fertility, disturbance, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).
 
@@
 \includegraphics[width=\textwidth]{concept}
-	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stocking density, and AGB.
+	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB.
 Hypothesised paths of causation are depicted as arrows from predictor to response.}
@@
 Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
-Standardisation also controls for variables with different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
+Standardisation also controls for variables with variation across different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
 To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables had their sign reversed.
-For example, overall moisture availability is expected to decrease as soil sand content increases, therefore sand content was reversed for use in the moisture availability latent variable.
-Precipitation seasonality, and temperature stress were also reversed in this way to account for the direction of their effect on moisture availability.
-
-%While it is recommended by some to set exact factor loadings in the SEM from the regression coefficients of multiple regressions \citep{lavaan}, because some latent variables were regressed against both structural diversity and AGB, exact factor loadings from simple multiple regressions could not be used.
-We therefore allowed factor loadings to be estimated by the SEM itself.
+For example, overall water availability is expected to decrease as soil sand content increases, therefore sand content was reversed for use in the water availability latent variable.
+Precipitation seasonality, and temperature stress were also reversed in this way to account for the direction of their effect on water availability.
 
@@
 We found no significant difference between model specifications (p>0.05).
-Full Information Max-Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.
+Full Information Maximum Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.
 
 We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
-To account for variation in stocking density, which may covary with species diversity, we also included it as an observed variable in our model.
-To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation cluster separately.
-We compared unstandardised path coefficients among these vegetation cluster scale models to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stocking density and AGB.
-Path coefficients show the effect of a path with other paths of inference held constant.
+Structural diversity can also directly influence AGB in this model, without an effect of species diversity.
+To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model.
+To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately.
+We compared unstandardised path coefficients among the models for different vegetation types to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB.
+Path coefficients show the effect of a given path with other paths held constant.
 Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}.
-Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error \rnew{of Approximation} (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
-We critically assess model fit in each case, taking into consideration the recommendations of \citet{Hu1999} \rchange{which}{who} define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.
+Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error of Approximation (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
+We critically assessed model fit in each case, taking into consideration the recommendations of \citet{Hu1999} who define threshold values of acceptability for these model fit indices: CFI >0.85, TLI >0.85, RMSEA <0.15, alongside our judgement of the model estimates.
 
-To explore the hypothesis that niche complementarity effects increase in strength as stocking density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stocking density.
+To explore the hypothesis that niche complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density.
 We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB.
-We excluded the effect of stocking density on AGB and the correlation between stocking density and species diversity from this model as we deliberately controlled stocking density in our subsampling.
-We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stocking density of subsampled datasets.
-
-We incorporated environmental covariates into our model to understand the relative effects of moisture availability, soil fertility and disturbance by fire on AGB both directly and indirectly via species diversity and stocking density.
+We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model as we deliberately controlled stem density in our subsampling.
+We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled datasets.
+Preliminary analyses that included herbivore biomass \citep{Hempson2017} did not converge, possibly due to the spatially coarse nature of the available data, we therefore did not include herbivory in our final model.
+We incorporated environmental covariates into our model to understand the relative effects of water availability, soil fertility and disturbance on AGB both directly and indirectly via species diversity and stem density.
 We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB.
-Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of variance particularly in the moisture availability latent variable.
-
-% We fitted separate moderation models to investigate whether there was an interaction effect whereby the strength of the relationship between species diversity and AGB was influenced by moisture availability and soil fertility.
-The \verb|lavaan| R package does not natively support moderation of latent variables in its model specification.
-Instead we manually calculated interaction variables for both soil fertility and moisture availability from the product of predicted values of these latent variables in a Confirmatory Factor Analysis (CFA).
-Interaction variables were the products: tree species diversity $\times$ moisture availability, and tree species diversity $\times$ soil fertility.
-These interaction terms were included as explanatory terms in multiple regressions alongside the latent variable of species diversity to predict AGB.
-Regression coefficients and model fit were analysed to determine the presence and form of interaction effects.
+Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.
 
@@
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of $r$ overlapped zero.
-Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = disturbance, turquoise = moisture availability, green = tree species diversity, purple = tree structural diversity.
+Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = disturbance, turquoise = water availability, green = tree species diversity, purple = tree structural diversity.
 See \hyperref[appendixb]{Appendix B} for a full assessment of correlation fit statistics.}
@@
 Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
-stocking density had the strongest correlation with AGB of all variables \rnew{considered} (\ccib{}).
+stem density had the strongest correlation with AGB of all variables considered (\ccib{}).
 Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency.
@@
 As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a weak negative correlation between the seasonality of precipitation and AGB (\ccmcb{}).
-Temperature stress was \rnew{weakly} negatively correlated \rold{weakly} with AGB (\cctcb{}).
+Temperature stress was weakly negatively correlated  with AGB (\cctcb{}).
 As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil nutrient content and AGB (carbon: \ccob{}, nitrogen: \ccnb{}).
@@
 MAP had positive correlations with all tree species diversity and structural diversity variables.
-Tree species diversity variables had clear positive correlations with stocking density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stocking density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).
+Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).
 
@@
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effects of tree structural diversity and stocking density (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}).
+In an SEM describing the effect of tree species diversity on AGB via the mediating effects of tree structural diversity and stem density (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}).
 Model fit was good with high factor loadings for all observed variables.
@@
 The R\textsuperscript{2} of AGB was \strucrsq{}.
-The strongest direct effect on AGB was from stocking density (\strucbetaib{}).
+The strongest direct effect on AGB was from stem density (\strucbetaib{}).
 
@@
 	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all vegetation clusters.
-Latent variables are shown as circles while observed variables are shows as rectangles.
+Latent variables are shown as circles while observed variables are shown as rectangles.
 Standardised path coefficients are shown as solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
@@
 When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the four vegetation types, model fit and the strengths of unstandardised path coefficients varied.
-The direct effect of tree species diversity on AGB was positive and marginally significant in Acacia (\strucbetacsb{}) but negligible in Mopane (\strucbetadsb{}), Marginal miombo (\strucbetaasb{}) and Core miombo (\strucbetabsb{}) (\autoref{struc_model_slopes_all}).
+The direct effect of tree species diversity on AGB was positive and marginally significant in ex-Acacia (\strucbetacsb{}) but negligible in Mopane (\strucbetadsb{}), sparse miombo / Baikiaea (\strucbetaasb{}) and Core miombo (\strucbetabsb{}) (\autoref{struc_model_slopes_all}).
 Relationships between structural diversity and AGB remained generally similar, with the same sign and overlap between the 95\% confidence intervals of path coefficients.
-The R\textsuperscript{2} of AGB was highest in Acacia shrubland (R\textsuperscript{2} = \struccrsq{}) and lowest in Marginal miombo (R\textsuperscript{2} = \strucarsq{}).
+The R\textsuperscript{2} of AGB was highest in ex-Acacia shrubland (R\textsuperscript{2} = \struccrsq{}) and lowest in sparse miombo / Baikiaea (R\textsuperscript{2} = \strucarsq{}).
 The total effect of species diversity on AGB remained strongly positive for all vegetation types.
 All vegetation types exhibited a positive effect of species diversity on structural diversity.
-All models had adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though confidence intervals around the unstandardised path coefficients were wide particularly for Mopane and Acacia.
+All models had adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though confidence intervals around the unstandardised path coefficients were wide particularly for Mopane and ex-Acacia.
 $\chi^{2}$ statistics were high for some vegetation types, but this appears to be highly correlated with sample size for each vegetation type \citep{Hooper2008}.
@@
 
-\subsection{Moderation of Diversity-AGB relationship by stocking density}
+\subsection{Moderation of Diversity-AGB relationship by stem density}
 
-In our sub-sampling of the plot dataset by stocking density, we found an increasing positive effect of tree species diversity on AGB as stocking density increased (\autoref{sem_struc_stems_ha}).
-There appears to be a minimum stocking density threshold at \textapprox{}180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass.
-The effect of structural diversity on AGB appears to remain constant with increasing stocking density.
-The indirect effect of species diversity on AGB via structural diversity climbs slightly as stocking density increases.
+In our sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}).
+There appears to be a minimum stem density threshold at \textapprox{}180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant, small baseline effect of tree diversity on biomass.
+The effect of structural diversity on AGB appears to remain constant with increasing stem density.
+The indirect effect of species diversity on AGB via structural diversity climbs as stem density increases.
 
@@
 	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stocking density. Smoothed lines are loess curves with standard error shaded bars.}
+	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves with standard error shaded bars.}
 	\label{sem_struc_stems_ha}
@@
 
-A model incorporating the latent variables of moisture availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of moisture availability, soil fertility and disturbance (\autoref{full_mod}, \autoref{full_model_slopes}).
-Surprisingly, the direct effects of moisture availability, soil fertility and disturbance on AGB were negligible (moisture, \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effect on AGB coming from the indirect path via stocking density (moisture: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (moisture: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}).
-MAP and soil sand content had the greatest contributions to the latent variable of moisture availability.
+A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, \autoref{full_model_slopes}).
+The direct effects of water availability, soil fertility and disturbance on AGB were negligible (water: \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effect on AGB coming from the indirect paths via stem density (water: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (water: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}).
+MAP and soil sand content had the greatest contributions to the latent variable of water availability.
 Model fit was acceptable: CFI = 0.925, TLI = 0.900, and RMSEA = \fmrmsea{}, R\textsuperscript{2} of AGB = \fmrsq{}.
@@
 Similar to the model that only considered tree species and structural diversity (\autoref{struc_mod}), the direct effect of species diversity on structural diversity was positive, while structural diversity itself had a positive effect on AGB, leading to a strong positive indirect effect of species diversity on AGB via structural diversity (\fmbetadhb{}) when environmental covariates were accounted for.
-Again, the direct effect of species diversity on AGB was negligible (\fmbetadb{}) The total effect of species diversity on AGB was positive (\fmbetatotaldb{}).
+Again, the direct effect of species diversity on AGB was negligible (\fmbetadb{}).
+The total effect of species diversity on AGB was positive (\fmbetatotaldb{}).
 
@@
 	\includegraphics[width=\textwidth]{full_model_slopes}
-	\caption{Unstandardised path coefficients for the full model including tree species diversity, environmental covariates and stocking density.
+	\caption{Unstandardised path coefficients for the full model including tree species diversity, environmental covariates and stem density.
 Path coefficients are $\pm1$ standard error.
@@
 
-% \input{include/full_model_fit_clust_stats} - Excluded, low fit
-
-% \subsection{Moderation effects of moisture availability and soil}
-% 
-% To address the hypothesis (H\textsubscript{1}) that more arid plots and plots with less fertile soil will show a stronger positive effect of tree species richness on above ground woody biomass, we fit separate linear multiple regressions with the latent variables of moisture availability and soil fertility as moderators on the relationship between species diversity and AGB.
-% 
-% Both soil fertility and moisture availability had small positive interaction effects on the strength of the relationship between tree species diversity and biomass (\autoref{int_plots}).
-The regression coefficient for the interaction effect of tree species diversity and moisture availability was significant (\moderp{}), while the interaction term of tree species diversity and soil fertility was marginally significant (\moders{}).
-As moisture availability increased, the effect of tree species diversity on AGB became stronger.
-See \autoref{mois_div_int_mod} and \autoref{soil_div_int_mod} for full regression tables.
-% 
-% 
-% \begin{figure}[H]
-% \centering
-% 	\subfloat[]{
-% 	\includegraphics[width=0.48\textwidth]{mois_int}
-% 	}
-% 	\subfloat[]{
-% 	\includegraphics[width=0.48\textwidth]{soil_int}
-% 	}
-% 	\caption{Scatter plots showing variation in the relationship between tree species diversity and AGB with moisture availability (a) and soil fertility (b).
-All variables are standardised and centred to a mean of zero and a standard deviation of 1.
-Lines of best fit are drawn for standard deviations of moisture availability and soil fertility between -2 and +2.
-Moisture availability, tree species diversity and soil fertility are latent variables comprised of the same observed variable loadings.}
-% 	\label{int_plots}
-% \end{figure}
-
 \section{Discussion}
 
-% Precipitation and soil fertility will indirectly positively effect AGB via an increase in tree species diversity.
-% Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and AGB.
-% The effect size of species diversity on AGB will increase with stocking density, due to an increased importance of niche complementarity as competition increases.
-
-In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stocking density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots.
-Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB indirectly via structural diversity (H\textsubscript{1}).
-We found that the effect size of tree species diversity on AGB increased with stocking density (H\textsubscript{2}).
-Tree species diversity, structural diversity and stocking density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}).
-The strongest effect on AGB was that of stocking density.
-When the effects of moisture availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.
+In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots.
+Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}).
+We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).
+Tree species diversity, structural diversity and stem density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}).
+The strongest effect on AGB was that of stem density.
+When the effects of water availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.
 
@@
 We found a consistent positive effect of tree species diversity on AGB across all models in the current study.
-Within southern African woodlands we therefore find support for our hypothesis that higher tree species richness and evenness causes higher woody AGB.
-This finding is in agreement with many other studies across different ecosystems and biomes, supporting \rnew{the idea} that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
+Within southern African woodlands we therefore find support for our hypothesis that higher tree species richness and evenness leads to higher woody AGB.
+This finding is in agreement with many other studies across different ecosystems and biomes, supporting the idea that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
 Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of southern African woodlands, a disturbance-structured and poorly studied ecological system.
 
-% In our extrapolation of species richness according to sampling effort (stocking density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stocking density exists.
-We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
-
-Much of the total variation in AGB was driven by variation in stocking density.
-It is possible that within southern African woodlands a higher species diversity allows for a higher stocking density through niche separation, which reduces competition between species occupying varying niche space, leading to an increase in total AGB per unit area.
-The opposite causation is also plausible however, with increased stocking density causing higher species richness through an increased probability of encountering new species.
-We attempted to correct for the correlation between species richness and stocking density by estimating the extrapolated species richness.
+Much of the total variation in AGB was driven by variation in stem density.
+It is possible that within southern African woodlands a higher species diversity allows for a higher stem density through niche separation, which reduces competition between species occupying varying niche space, leading to an increase in total AGB per unit area.
+The opposite causation is also plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
+We attempted to correct for the correlation between species richness and stem density by estimating, and using in models, the extrapolated species richness.
 We suggest therefore that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can utilise more available light and moisture, resulting in greater plot-level AGB.
-This is supported by the moderately strong indirect positive effect of tree species diversity on AGB via structural diversity, and the positive effect of moisture availability on AGB via stocking density in the model which included environmental covariates.
-However, we found a negative effect of soil fertility on AGB via stocking density.
+This is supported by the moderately strong indirect positive effect of tree species diversity on AGB via structural diversity, and the positive effect of water availability on AGB via stem density in the model which included environmental covariates.
 
@@
 greater variation in DBH and height.
-This may act as a mechanism for niche complementarity, with a highly diverse canopy able to take advantage of a greater proportion of the available light.
+This may act as a mechanism for niche complementarity, with a canopy of diversely sized trees able to take advantage of a greater proportion of the available light.
 Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{}, i.e.
-wider variation in canopy shape and overall growth form; bushy understorey vs.
-emergent canopy, for example.
+wider variation in canopy shape and overall growth form; broad flat crowns vs.
+narrow deep crowns, for example.
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
-Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stocking density supports this.
+Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this mechanism.
 In frequently disturbed woodlands such as those studied here however, a woodland canopy similar to that of a forest is frequently not reached.
 Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
-Previous studies have found that southern African woodlands with higher species diversity tend to \rnew{experience less frequent disturbance by fire and tend to form a more closed canopy and a more sparse understorey} \citep{Chidumayo2013, Mutowo2012}.
+Previous studies have found that southern African woodlands with higher species diversity tend to experience less frequent disturbance by fire and tend to form a more closed canopy and a more sparse understorey \citep{Chidumayo2013, Mutowo2012}.
 In our study however, we found a positive effect of disturbance on species diversity, perhaps suggesting that disturbance prevents domination of woodlands by a single dominant species \citep{Chidumayo2013}.
 
-We found a non linear positive effect of stocking density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
-At low stocking densities competition between \rnew{mature} trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the low and constant effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}.
-At very high stem density, there is also increase in the effect of species diversity on structural diversity.
+We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
+At low stem densities competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}.
+At very high stem density, there is also an increase in the effect of species diversity on structural diversity.
 This could be because at high stem density, the adaptation of different species to growth form become important.
@@
 
-\subsection{Effects of moisture availability, soil fertility and disturbance}
+\subsection{Effects of water availability, soil fertility and disturbance}
 
-Moisture availability, soil fertility and disturbance by fire had similar positive total effects on AGB, comparable in size to the total effect of tree species diversity on AGB.
-We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for a greater stocking density per unit area, greater productivity per unit area and additionally greater tree species diversity due to niche partitioning \citep{Kraaij2006, Shirima2015}.
-Previous studies in tropical forests have shown that moisture availability increases AGB both directly and indirectly via increasing tree species diversity and via increasing stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}.
-In this study, we observed weak indirect positive effects on AGB via species diversity and a positive direct effect on AGB.
-Compared to moist tropical forests, moisture availability is more of a limiting factor to tree growth in southern African woodlands, which \rchange{are frequently droughted}{experience frequent drought}.
-It is surprising that disturbance by fire had a positive total effect on AGB.
-We found a positive indirect effect of disturbance on AGB via species diversity and structural diversity.
-It is possible that fire prevents the domination of a woodland by a few highly competitive species, resulting in greater abundance evenness and overall species richness, which in turn increases the effect of niche complementarity.
-In southern African woodlands moisture availability is linked to the intensity of disturbance from seasonal fires.
-The growth of C4 grasses in wetter woodlands leads to more intense seasonal fires which limit tree growth \citep{Charles-Dominique2018}, and may also limit species diversity \citep{Linder2014}.
-It is possible therefore that the effect of moisture availability, which is expected to increase AGB, is confounded in its effect on AGB with fire frequency, which had a negative effect on stocking density, which is expected to decrease AGB.
-The direct effect of moisture availability on stocking density may also be confounded in this way.
-This may also have caused us to not observe a stronger effect between moisture availability and AGB.
-We found a positive total effect of soil fertility on AGB in our model.
-We measured soil fertility using the observed variables of soil carbon, nitrogen and Cation Exchange Capacity (CEC).
-In wet tropical forests a clear relationship has been observed between these variables and AGB \citep{Slik2009, Coelho-de-Souza2019}.
+Water availability had a positive total effect on AGB, comparable in size to the total effect of tree species diversity on AGB, while soil fertility had a negative total effect.
+We expected that higher water availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for a greater stem density per unit area, greater productivity per unit area and additionally greater tree species diversity due to niche partitioning \citep{Kraaij2006, Shirima2015}.
+Previous studies in tropical forests have shown that water availability increases AGB both directly and indirectly via increasing tree species diversity and via increasing stand structural diversity \citep{Ali2019a, Ali2019b, Poorter2017}.
+In this study, we observed indirect positive effects of water availability on AGB via species diversity and a positive but only marginally significant direct effect on AGB.
+Compared to moist tropical forests, water availability is more of a limiting factor to tree growth in southern African woodlands, which experience frequent drought.
+The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density, however, the direct effect on AGB remained positive and marginally significant, as expected.
+Disturbance by fire had a negative total effect on AGB.
+We found negligible indirect effects of disturbance on AGB via species diversity and structural diversity.
 
@@
 This demonstrates the robustness of our results, showing they are generalisable across vegetation types in southern Africa.
-It also demonstrates that similar ecosystem processes are occurring in these vegetation types, despite variation in species composition and overall species richness.
+It also demonstrates that similar ecosystem processes are occurring in these vegetation types, despite variation in species composition, overall species richness and mean biomass.
 
-Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via structural diversity, remained positive in these vegetation types.
-Compared to Acacia and Mopane woodlands, miombo woodlands have higher median tree species richness.
-Acacia and Mopane woodlands are dominated by fewer tree species, notably \textit{Senegalia} spp.
-in Acacia woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees.
-We postulate that the negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in Acacia and Mopane woodlands, where the woodland canopy is dominated often by a single species.
-Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation, while in the species poor Mopane and Acacia woodlands, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.
+Core miombo and sparse miombo / Baikiaea woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via structural diversity, remained positive in these vegetation types.
+Compared to ex-Acacia and Mopane woodlands, miombo woodlands have higher median tree species richness.
+ex-Acacia and Mopane woodlands are dominated by fewer tree species, notably \textit{Senegalia} spp.
+in ex-Acacia woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees.
+We postulate that the slight negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in ex-Acacia and Mopane woodlands, where the top level of the woodland canopy is dominated often by a single species.
+Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation, while in the species poor Mopane and ex-Acacia woodlands, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.
 
@@
 This has been attributed to an increase in functional redundancy as species diversity increases.
-\textit{I.e.} with more species, it is more likely that the addition of a new species will occupy the same ecological niche space as an existing species, meaning niche complementarity will not occur and competition will lead to niche partitioning, while making little difference to overall ecosystem functioning.
+In other words, with more species, it is more likely that the addition of a new species will occupy the same ecological niche space as an existing species, meaning niche complementarity will not occur and competition will not lead to niche partitioning, making little difference to overall ecosystem functioning.
 Mopane woodlands also have a negligible effect of species diversity on structural diversity.
 This may be due to the species which tend to co-exist with \textit{C.
-mopane}, many of which are small shrub-like trees which do not grow into large canopy trees \citep{Timberlake2010}.
+mopane}, many of which are small shrub-like trees and which do not grow into large canopy trees \citep{Timberlake2010}.
 Larger canopy trees tend to have greater variation in physical structure \citep{Seidel2019}.
 
-Acacia woodland had the strongest total effect of species diversity on AGB and was the only vegetation type to show a clearly positive direct effect of species diversity on AGB.
-Acacia also has relatively low median species richness compared to miombo, but the addition of new species appears to make a larger difference to the AGB of these plots than in mopane woodlands.
-We suggest that this is due mostly to the particular identity of species found in Acacia woodlands and their contribution to ecosystem functioning.
-Unlike mopane woodlands, Acacia woodlands do sometimes contain species other than \textit{Senegalia} spp.
+ex-Acacia woodlands showed the strongest total effect of species diversity on AGB and was the only vegetation type to show a clearly positive direct effect of species diversity on AGB.
+ex-Acacia woodlands also had relatively low median species richness compared to miombo, but the addition of new species appears to make a larger difference to the AGB of these plots than in mopane woodlands.
+We suggest that this is due mostly to the particular identity of species found in ex-Acacia woodlands and their contribution to ecosystem functioning.
+Unlike mopane woodlands, ex-Acacia woodlands do sometimes contain species other than \textit{Senegalia} and \textit{Vachellia} spp.
 which can grow to be high biomass canopy trees, especially in transition zones.
@@
 
-In this study we found that across southern African woodlands, there is a generalisable positive association between tree species diversity and woody biomass as a measure of ecosystem function.
-Additionally, we found that much of this effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of woodland trees.
-We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, inferring that similar ecosystem processes occur across the region to determine ecosystem function.
-In contrast to previous studies, we found that across the region, the direct effects of moisture availability and soil fertility on woody biomass were negligible, with most of their effect being indirectly through tree species and structural diversity.
+In this study we found that across southern African woodlands, there is a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB).
+Additionally, we found that much of this effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees.
+We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, suggesting that similar ecosystem processes occur across the region to determine ecosystem function.
+In contrast to previous studies, we found, at the scale of our study region, that the direct effects of water availability and soil fertility on woody biomass were negligible, with most of their effect being indirectly through their effects on tree species and structural diversity.
+This strongly suggests that data on tree species diversity be included into models predicting ecosystem functionality in this region.
+We also advocate for explicit inclusion of environmental covariates in regional scale models of biodiversity and ecosystem function, generally.
 
-Southern African woodlands are relied heavily upon for their ecosystem service provision, which is itself affected by ecosystem function.
+Southern African woodlands are relied heavily upon for their ecosystem service provision, which is itself related to ecosystem function.
 Resource extraction by humans in southern Africa is directly influencing biodiversity via selective tree-felling for timber, among other forest products.
-Our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stocking density, and low species diversity, which are predominantly targeted for tree felling.
+Our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.
 This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.
@@
 \end{document}
-
-% https://jslefche.github.io/sem_book/coefficients.html

### 2020-07-15 scripts/standardise.R

@@ -269,7 +269,7 @@ lab_lookup <- data.frame(orig = c("precip_std", "precip_seas_std", "temp_std",
     "cov_height_std", "n_trees_gt10_ha_log_std", "shannon_equit_std",
     "n_species_raref_log_std"), 
   new = c("MAP", "Precip. seas.", "MAT", "Temp. stress", "AGB", "Soil CEC", "Soil sand", "Soil C", "Soil N", 
-    "Fire freq.", "Herbivore biomass", "DBH CoV", "Height CoV", "Stocking density", 
+    "Fire freq.", "Herbivore biomass", "DBH CoV", "Height CoV", "Stem density", 
     "Shannon equit.", "Sp. rich."))
 
 for (i in 1:length(bivar_list)) {

### 2020-07-15 manuscript/regional_befr_manuscript.tex
Updated refs.

@@
 Compared to forest ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire and herbivory, with African savannas possessing large herbivores absent from other savannas \citep{Sankaran2008, Levick2009}.
-Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{}.
+Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{Dewees2010}.
 High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution.
@@
 Reduced stem density reduces competition between individuals, possibly weakening the effect of diversity on ecosystem function via niche complementarity.
-Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations under limiting environmental conditions such as low water availability \citep{}.
+Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations under limiting environmental conditions such as low water availability.
 Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
@@
 A small number of studies in southern African woodlands, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
-The results of these fine scale studies concur with similar studies in other biomes \citep{}.
+The results of these fine scale studies concur with similar studies in other biomes \citep{Cardinale2009}.
 Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
-Due to the variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, with wide variation in precipitation, diurnal and annual temperature range, disturbance regime, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass \citep{}, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
+Due to the variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, with wide variation in precipitation, diurnal and annual temperature range, disturbance regime, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
 
@@
 Groups were defined in \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
-Group identification was conducted using unconstrained correspondence analysis, followed by clustering based on dominant ordination axes.
+Group identification was conducted using unconstrained correspondence analysis, followed by hierarchical clustering based on dominant ordination axes.
 Plot data used in this study occurred in four compositional vegetation types.

### 2020-07-15 scripts/standardise.R

@@ -178,7 +178,6 @@ bivar_list <- c(
   "agb_ha_log_std ~ precip_seas_std",
   "agb_ha_log_std ~ sand_std",
   "agb_ha_log_std ~ fire_log_std",
-  "agb_ha_log_std ~ herbivory_std",
   
   "n_trees_gt10_ha_log_std ~ n_species_raref_log_std",
   "n_trees_gt10_ha_log_std ~ shannon_equit_std",
@@ -191,7 +190,6 @@ bivar_list <- c(
   "n_trees_gt10_ha_log_std ~ precip_seas_std",
   "n_trees_gt10_ha_log_std ~ sand_std",
   "n_trees_gt10_ha_log_std ~ fire_log_std",
-  "n_trees_gt10_ha_log_std ~ herbivory_std",
   
   "n_species_raref_log_std ~ cec_std",
   "n_species_raref_log_std ~ soil_c_log_std",
@@ -201,8 +199,7 @@ bivar_list <- c(
   "n_species_raref_log_std ~ precip_std",
   "n_species_raref_log_std ~ precip_seas_std",
   "n_species_raref_log_std ~ sand_std",
-  "n_species_raref_log_std ~ fire_log_std",
-  "n_species_raref_log_std ~ herbivory_std")
+  "n_species_raref_log_std ~ fire_log_std")
 
 # Create models
 lm_list <- list()

### 2020-08-18 co-authors/commented_versions/manuscript_edKGD.tex
Removed cache

--- 2019-12-17 manuscript/manuscript.tex
@@
 
-\maketitle
-\tableofcontents
+{\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}}
+
+\vspace{1cm}
+
+Authors: Godlee, J. L.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1}, Ryan, C. M.\textsuperscript{1}
+
+\textsuperscript{1}: School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom \\
+\textsuperscript{2}: Some other address
+
+\section*{Acknowledgements}
+
+\section*{Biosketch}
 \newpage{}
 
-\begin{abstract}
-	Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard.
+\begin{center}
+\LARGE{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
+
+\normalsize{Woodland functionality in southern Africa}
+\end{center}
+
+\section*{Abstract}
+
+\textbf{Aim:} Positive correlations between tree biodiversity and ecosystem function have been widely documented, but the nature of the relationship in highly disturbed and ecophysiologically stressful systems is less clear.
+Here, we explore the relationship between tree biodiversity and aboveground biomass across southern African savannas and woodlands, with respect to gradients in stem density and resource availability, to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context and biome.
+
+\textbf{Location:} Southern African savannas and woodlands
+
+\textbf{Time period:} 2010-2019
+
+\textbf{Major taxa studied:} Trees
+
+\textbf{Methods:} We used a network of \nplots{} savanna and woodland plots located across the southern African sub-continent in which each tree >5 cm diameter was measured.
+We used Structural Equation Modelling and path analysis to determine the relationship between tree species diversity and aboveground biomass, with comparison to the effects of resource availability and along a gradient of stem density.
+
+\textbf{Results:} A positive effect of tree species diversity on biomass was demonstrated, operating largely via an indirect effect of increasing woodland structural diversity.
+We also found that the effect of tree species diversity on biomass increases with stem density.
+Finally, we found that resource availability affects biomass in southern African woodlands largely indirectly via its effect on species diversity.
+
+\textbf{Main conclusions:} Tree species diversity plays an important role in determining the ecosystem functionality and structure of southern African woodlands.
+Our results demonstrate the importance of considering environmental conditions and woodland structural attributes to more accurately describe the relationship between biodiversity and ecosystem functionality.
+
+% Studies in natural and experimental settings have shown a positive relationship between tree species diversity and ecosystem functionality but many ecosystems are understudied in this regard.
 In this study we conducted the first regional estimation of the relationship between tree species diversity and above-ground biomass in southern African woodlands, using a network of \nplots{} woodland plots.
@@
 Finally, we found that resource availability determines biomass in southern African woodlands largely indirectly via its effect on species diversity.
-\end{abstract}
+%KGD: I think you can delete that old abstract. New one is much snappier/better!
 
-\section{Introduction}
+\section*{Introduction}
 
-A large number of studies have shown relationships between biodiversity and ecosystem functionality (e.g.
+Numerous studies have shown relationships between biodiversity and ecosystem function 
+(e.g.
 \citealt{Liang2016, Cardinale2009}).
@@
 %These studies support early findings from small scale experimental studies which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
+%KGD: I think it is good to reference those classic experimental studies in grasslands. So, even if not that exact sentence, something similar would be good.
 
@@
 This includes basic processes of primary production such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others functional contribution.
+The frequently reported and intuitive relationship between biodiversity and ecosystem functionality invokes three main mechanisms which drive the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater diversity fill a greater breadth of realised niche space and avoid competition due to differences in their traits, 2) selection effects, whereby communities with greater diversity are more likely to include a species which contributes highly to the measured ecosystem function, and 3) facilitation effects, whereby communities with greater diversity are more likely to include combinations of species which together increase the others' functional contribution.
 
+%Here you start justifying why this study is needed.
+The 'it hasn't been done here' justification is often poorly received by reviewers, and that is how your justification starts.
+However, the 'it depends on ecological context' justification works better.
+In terms of basic theory we might not expect a BEFR in disturbance-dominated systems.
+That is what I would lead with.
+The reason reviewers (and scientists in general) prefer that is because it gives greater generality to your study, allow conclusions to be extrapolated to other systems.
+The text on dry tropical ecosystems (particularly the percentages) is useful and should come in here somewhere.
+It could come towards the end of this paragraph then, stand on its own in a separate paragraph or even find its way into the subsequent paragraph (probably not on that last one though...).
 The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
@@
 
-Savannas and sparse woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{clust_map}).
-The carbon stored in these woodlands is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
-Climatic conditions and biogeography vary across southern African woodlands, resulting in a diverse range of woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species.
-Southern African woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+%You could also cite White for the first sentence in this paragraph to connect to classic literature
+%Also, we keep needing to refer to 'southern African savannas and woodlands' or 'southern African woodlands and savannas' so perhaps an acronym is in order (SASW or SAWS).
+Connotation of the latter is a bit awkward even if it flows of the tongue better.
+Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{Ryan2016} (Figure \autoref{clust_map}).
+The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
+Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages, which retain the common features of an open tree canopy and an understorey generally dominated by C4 grass species.
+Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which there are >300 tree species \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
@@
 
+%I think this paragraph could be deleted.
+In a sense it is a more extended rehash of the end of preceding paragraph.
+As we aren't actually studying human impacts directly, it is a bit of a distraction.
+Some of this text if you want to keep it could work well in a conclusions section.
 Southern African woodlands are relied upon heavily for their ecosystem service provision, which is itself affected by ecosystem functionality \citep{Schulze1994}.
@@
 
-A small number of studies in southern African woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}, but the scientific impact of these studies has been hampered by a restricted climatic and biogeographical range of study sites.
-Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass, it is important to sample across these gradients to be able to infer a truly regional scale relationship between diversity and biomass.
-Studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
+A small number of studies in southern African savannas and woodlands have shown that above ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+However, studies conducted over small environmental gradients often find that at local scales, diversity shows a strong effect on ecosystem function, but at large scales diversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
 Small scale studies may therefore be over-estimating the effect of diversity.
+Due to the highly variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and woody biomass %do you have a citation to support this statement?
+Would be good...
+, it is important to sample across geographic and environmental gradients to be able to infer a truly regional scale relationship between diversity and biomass.
 
+%This paragraph is a bit awkward and includes multiple themes.
+I think the first half could be reduced in spliced into the preceding paragraph.
+Meanwhile, the second half relates to your a priori predictions about whether you expect a BEFR in disturbance-dominated / stressful ecosystems.
+It seems like that text would be better placed above where you are discussing general theoretical reasons for why a BEFR might exist and under what environmental conditions you would expect to find one.
 In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and species diversity independently \citep{Spasojevic2014}.
@@
 It is important therefore to consider not only species richness as a measure of diversity, but also the evenness of the distribution of trees among species when assessing the impact of diversity on ecosystem function.
+%KGD: I agree on deleting this paragraph. Some of the justification for looking at abundance evenness could actually be included (briefly) in methods.
 
@@
 Additionally, the dominance of a single species is likely to decrease overall functioning by decreasing niche space utilisation \citep{Cardinale2002}.
+%KGD: This text is useful in demonstrating your biological understanding of the ecosystem.
+I think it could be useful to include some of this biological knowledge in the paragraph above that ends with the statements on disturbance (the paragraph starting with Slobrig 1996).
+Or you could include this whole paragraph just after that one.
+The co-authors will reign us in if this kind of logic is not sensible for the ecosystems we are studying.
 
-In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship in southern African woodlands, using inventory plots which span multi-dimensional environmental and biogeographical gradients (\autoref{clust_map}).
-We investigated the relationship between aboveground woody biomass and tree biodiversity.
-We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation: precipitation, temperature and soil fertility.
-We also investigated the potential moderation effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality.
-Initially, we made three hypotheses:
-
-\subsection{Hypotheses}
-
-\begin{enumerate}
-	\item{Precipitation and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity.}
-	\item{The effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases.}
-	\item{Tree structural diversity will indirectly influence tree species diversity to provide an indirect path of influence between structural diversity and woody biomass.} \label{lst:num}
-\end{enumerate}
+In this study, we made the first known regional estimation of the Biodiversity-Ecosystem Function Relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
+We compared the relative effects of tree species biodiversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
+We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
+We incorporated vegetation type based on major tree species compositional units as a factor in our analyses to understand how species composition as well as species biodiversity affected ecosystem functionality and assess the generality of our results.
+Initially, we made three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect size of species diversity on woody biomass will increase with stem density, due to an increased importance of niche complementarity as competition increases, and (3) tree species diversity will increase tree structural diversity, which will provide an indirect path by which tree diversity increases woody biomass.
+%I wasn't sure what exactly you meant to say with #3. I have edited to what I think it should say, but I may have missed the mark.
 
 %To address these hypotheses we used Structural Equation Models with path analysis as a preferred method to simultaneously account for environmental factors and biotic factors, which may interact in their effect on ecosystem structure and therefore biomass.
+%KGD: I don't think it would hurt to mention that you are using SEMs as your statistical workhorse here. This is a matter of style though. Include if you want to.
 
-\section{Methods}
+\section*{Materials and methods}
 
-\subsection{Study location}
+\subsection*{Study location}
 
@@
 
-In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability via moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots.
-Using latent variables and Structural Equation Modelling (SEM), we found support for a general relationship between tree species diversity and AGB, with an indirect relationship between tree species diversity and AGB via structural diversity (H\textsubscript{1}).
+In this study, we assessed the importance of [a] tree species richness, [b] tree structural diversity, [c] resource availability.
+i.e.
+moisture availability and soil fertility, [d] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland survey plots.
+Using latent variables and Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, with an indirect influence of tree species diversity on AGB via structural diversity (H\textsubscript{1}).
 We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).
-Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for certain vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
+Tree diversity, structural diversity and stem density accounted for \srsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power (\autoref{struc_model_fit_clust_stats}).
 The strongest effect on AGB was that of stem density.
-Interestingly, when tree species diversity, structural diversity and stem density were controlled for, we found little evidence of an effect of resource availability, in the form of moisture availability and soil fertility, on AGB (H\textsubscript{3}).
-We found that vegetation composition in the form of discrete vegetation types affected the relationship between tree species diversity and AGB.
+Interestingly, when the effects of tree species diversity, structural diversity and stem density were accounted for, we found little evidence of a direct effect of resource availability, in the form of moisture or energy availability or soil fertility, on AGB (H\textsubscript{3}).
 
-\subsection{Effect of tree species diversity on AGB}
+\subsection*{Effect of tree species diversity on AGB}
 
@@
 
-We found a consistent positive total effect of tree species diversity on AGB across all models in this study.
-Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness causes higher woody AGB.
-This finding is in agreement with many other studies across different ecosystems and biomes, showing that there is a generalisable positive association between species richness and ecosystem functionality \citep{Liang2016, Cardinale2009}.
-In addition however, our study provides a novel dissection of the mechanisms underlying this relationship and their interaction with resource availability and vegetation type in southern African woodlands as an understudied biome.
+We found a consistent positive effect of tree species diversity on AGB across all models in this study.
+Within the savanna woodlands of southern Africa we therefore find support that higher tree species richness and evenness causes higher woody AGB.
+This finding is in agreement with many other studies across different ecosystems and biomes, supporting that there is a generalisable positive association between diversity and ecosystem functionality \citep{Liang2016, Cardinale2009}.
+Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of a disturbance-dominated and also poorly studied ecological system, southern African woodlands.
 
-Much of the total variation in AGB was caused by stem density.
+Much of the total variation in AGB was driven by variation in stem density.
 Stem density also correlated with species diversity in our SEMs.
 It is possible that within southern African woodlands a higher species diversity allows for a greater density of tree stems, leading to an increase in total AGB.
-Reverse causation is plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
-In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a real effect of species diversity on stem density exists.
-We cannot decompose the relative effects of tree species richness and abundance evenness in our model, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
-We suggest that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light resulting in greater total AGB at the plot level.
+The opposite is also plausible, with increased stem density causing higher species richness through an increased probability of encountering new species.
+%Not sure what you mean by following sentence->
+In our extrapolation of species richness according to sampling effort (stem density) we partially accounted for this effect, but our analyses cannot properly assess to what extent a causal effect of species diversity on stem density exists.
+We cannot decompose the relative effects of tree species richness and abundance evenness in our model %but hypothetically you could have, e.g.
+with multiple latent variables, so perhaps best not to mention this.
+You can make a statement about the relative importance of the two based on the model,no?
+, but previous studies have shown that both richness and evenness have similar importance in their effects on ecosystem function \citep{Valery2009, Zhang2012}.
+We suggest that an increase in tree species diversity through species richness and %italicise the 'and' here
+evenness produces an assemblage of species which can occupy a greater proportion of the total woodland canopy volume with leaf area, utilising more of the available light and resulting in greater total AGB at the plot level.
 This is supported by the moderately strong positive effect of tree species diversity on AGB via structural diversity.
-
+%The text towards the end of this paragraph seems like it should go with the following section, no?
 % There are counter-examples, e.g. Johnson et al. 2016. GCB (from Leeds group, Tim Baker senior author), but feel free to leave text as is for now.
@@
 While we did not explicitly measure Net Primary Productivity (NPP) in this study, other studies have shown a strong positive correlation between woody AGB and NPP in woodland and forest ecosystems \citep{Chisholm2013, Prado-Junior2016}.
-This suggests that as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in southern African woodlands can be maximised by increasing species diversity.
+This suggests that, as has been found in many other woodland/forest ecosystems, woody biomass and woody productivity in southern African woodlands can be maximised by increasing species diversity.
 
-\subsection{Structural diversity as a mechanism for the BEFR}
+\subsection*{Structural diversity as a mechanism for the BEFR}
+%the preceding section and this one are quite interrelated and each one is quite short.
+I wonder if it is worth combining them as a 'species diversity and interrelated factors driving AGB' or something like that with the interrelated factors being stem density and structural diversity
 
 We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity and we therefore find support for our hypothesis (H\textsubscript{2}).
-A higher tree species diversity allows for a greater diversity of tree functional forms within a plot and this may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
+A higher tree species diversity allows for a greater diversity of tree functional forms %functional forms or greater variation in tree size?
+You have really only measured tree size, whereas functional form seems like it should be more than that, incl.
+canopy shape.
+I would be more circumspect on this.
+Conversely, the term structural diversity doesn't seem like as much of a stretch, i.e.
+its OK to me.
+within a plot and this may act as a mechanism of niche complementarity, with a highly diverse canopy being able to take advantage of a greater proportion of the available light.
 Variation in structural diversity may be a joint result of disturbance history and tree species diversity, with highly disturbed plots generally having a less structurally diverse canopy \citep{LaRue2019}.
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
+%worth mentioning the stem density result again here?
+It seems relevant.
 In frequently disturbed woodlands such as those studied here however, a tree canopy similar to that of a forest is frequently not reached, Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
@@
 
-\subsection{Effects of moisture availability and soil fertility}
+\subsection*{Effects of moisture availability and soil fertility}
 
@@
 We expected that higher moisture availability and soil fertility would lead to higher AGB under the assumption that higher resource availability would allow for greater resource partitioning among individual trees and a greater stem density per unit area \citep{Kraaij2006, Shirima2015}.
+%great resource availability can give greater stem density without increasing resource partitioning I would think.
+I suppose it depends how you define resource partitioning, but I am thinking of it as niche partitioning.
 
@@
 It is possible that the range of observed moisture availability in this study (\textapprox{}460-1700 mm y\textsuperscript{-1}) may not have been able to capture variation in AGB.
+%preceding sentence confusing to me
 We deliberately excluded plots with very low stem density as they are not considered woodlands, but grassy savannas.
-It may be that by excluding the bottom end of this stem density continuum we prevented a relationship being observed between moisture and AGB/stem density.
-Additionally, due to the high levels of adaptation of tree species to drought conditions in southern Africa, at the large scale we conducted our experiment turnover in species composition may have prevented a direct relationship being observed between resource availability and AGB.
+It may be that by excluding the bottom end of this stem density continuum we limited our ability to detect a relationship between moisture and AGB/stem density.
+Additionally, due to the high levels of adaptation of tree species to different levels of water availability in southern Africa, at the large scale we conducted our experiment, turnover in species composition may have prevented a direct relationship being observed between resource availability and AGB.
+%need a bit more text to clarify what you mean here.
 
@@
 
+%you have written nothing about soil fertility in this section, but I think you need to.
+There is at least one paper which could be taken to suggest a positive relationship between AGB and soil fertility (Coelho de Souza et al.
+2019.
+NEE, see my publications).
+
 % In a separate set of linear multiple regressions, we found a weak positive interaction effect of moisture availability on the relationship between tree species diversity and AGB.
@@
 
-\subsection{Vegetation type specific responses}
+\subsection*{Vegetation type specific responses}
 
@@
 
-Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect via structural diversity remained positive in these vegetation types.
-Compared to Baikiaea and Mopane woodlands these woodlands have higher median tree species richness.
-Baikiaea and Mopane woodlands are also dominated by fewer tree species, notably \textit{Baikiaea plurijuga} in Baikiaea woodlands and \textit{Colophospermum mopane} in Mopane woodlands which tend to produce large canopy dominating trees.
+Core miombo and marginal miombo woodland vegetation exhibited a small negative direct effect of tree species diversity on AGB, while the total effect, incorporating the indirect effect via structural diversity, remained positive in these vegetation types.
+Compared to Baikiaea and Mopane woodlands, miombo woodlands have higher median tree species richness.
+Baikiaea and Mopane woodlands are also dominated by fewer tree species, notably \textit{Baikiaea plurijuga} in Baikiaea woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees.
 We postulate that this negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in Baikiaea and Mopane woodlands, where the woodland canopy is dominated often by a single species.
@@
 Again, these highly diverse plots in miombo woodlands may be the result of disturbance which can promote a mosaic of woodland of different successional stages and stem densities.
+%given the small effect size for the negative direct effect of diversity on AGB for core and marginal miombo, in combo with the much larger and positive total effect, I wonder if we should spend so much time trying to figure out the negative result for direct effect.
 
@@
 \textit{I.e.} with more species, it is more likely that the addition of a new species will occupy the same ecological niche space as an existing species, meaning niche complementarity will not occur and competition will lead to niche partitioning, while making little difference to overall ecosystem functioning.
+%I don't see how the preceding sentence logically follows on from the 2nd sentence in the paragraph.
+They seem to conflict.
 Mopane woodlands also have a negligible effect of species diversity on structural diversity.
@@
 plurijuga} which grow to be high biomass canopy trees.
+%do Baikiaea woodlands have lower median SR than mopane? I wouldn't have inferred that from the table describing veg types.
 
-\subsection{Stem density effects}
-
+\subsection*{Stem density effects}
+%I would integrate this into sections above about effect of richness and structural diversity.
+I clearly got confused above when you were writing about stem density as I was thinking about this result, rather than the role that stem density plays in the models you were discussing then.
+Still, I think readers will expect all stem density results to be discussed together for clarity.
 We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
@@
 An abundance of small stems may prevent the growth of large trees which hold the majority of the AGB in a plot.
+%second sentence in paragraph seems to be disconnected from first and seems to refer to a decline in effect at high stem densities which is not apparent to me in current figures.
 
@@
 
-% \subsection{Future work}
+% \subsection*{Future work}
 
@@
 
-\subsection{Conclusion}
+\subsection*{Conclusion}
 
@@
 A gap in available data means that we could not incorporate disturbance history into our models adequately, but this factor likely plays a large part in the association between species diversity and woody biomass in southern African woodlands.
+%these conclusions are a bit dry and repetitive on earlier text.
+What do our results mean for woodland management?
+for conservation?
+for our understanding of basic ecological theory?
+As an aside, we have circled back around to niche complementarity in this discussion, but not really to selection effects or facilitation.
 
-\bibliography{/Users/johngodlee/google_drive/bib/papers/lib.bib}
+\bibliography{lib.bib}
 

### 2020-08-18 manuscript/include/clust_summ.tex
Removed cache

@@
 \hline \\[-1.8ex] 
-{Cluster} & {Dominant species} & {Indicator species} & {N plots} & {Species rich.} & {Stems ha\textsuperscript{-1}} & {AGB (t ha\textsuperscript{-1})} \\
+{Cluster} & {Dominant species} & {Indicator species} & {N plots} & \multicolumn{1}{p{2cm}}{\centering Species\\Richness} & \multicolumn{1}{p{2.5cm}}{\centering Stem density\\(stems ha\textsuperscript{-1})} & \multicolumn{1}{p{2cm}}{\centering AGB\\(t ha\textsuperscript{-1})} \\
 \hline \\[-1.8ex] 

### 2020-08-18 manuscript/regional_befr_manuscript.tex
Removed cache

@@
 
-\title{An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
+\title{Tree biodiversity and ecosystem function relationship across southern African woodlands}
 \author{}
@@
 \input{include/perc_small_agb.tex}
+\input{include/zam.tex}
 
@@
 
-{\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}}
+{\LARGE{Title: Tree biodiversity and ecosystem function relationship across southern African woodlands}}
 
@@
 
-Authors: Godlee, J. L.\textsuperscript{1}, Ryan, C. M.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1}
+Authors: Godlee, J. L.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1,2}, Ryan, C. M.\textsuperscript{1}
 
-\textsuperscript{1}: School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom \\
-\textsuperscript{2}: Some other address
+\textsuperscript{1}: School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom \\
+\textsuperscript{2}: Royal Botanic Garden Edinburgh, Edinburgh, EH3 5LR, United Kingdom
 
@@
 
-School of GeoSciences, University of Edinburgh, Edinburgh, United Kingdom
+School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom
 
@@
 We thank all data providers and the field assistance they received when collecting plot data.
+JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).
 
@@
 \textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear.
-In disturbed systems it is posited that high levels of disturbance may allow the presence of poor competitors, weakening the correlation between biodiversity and niche complementarity which drives ecosystem function.
-Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, while controlling for gradients in stem density, resource availability, disturbance through fire, and across vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+It is posited that high levels of disturbance moderate the effects of competition in savannas, weakening the correlation between biodiversity and niche complementarity that drives ecosystem function.
+Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, while controlling for variation in stem density, resource availability, disturbance through fire, and vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
 
 \textbf{Location:} Southern African savannas and woodlands
-
 \textbf{Time period:} 2010-2019
@@
 
-\textbf{Methods:} We used a network of \nplots{} savanna/woodland tree plots located across the southern African sub-continent .
-We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, and stem density.
+\textbf{Methods:} We used a network of \nplots{} savanna/woodland tree plots located across the southern African sub-continent.
+We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.
 
-\textbf{Results:} We found a positive effect of tree species diversity on aboveground biomass, observed mainly via the increasing effect of woodland structural diversity.
-We also found that the effect of tree species diversity on biomass increases with stem density.
-Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via its effect on species diversity.
+\textbf{Results:} We found a positive effect of tree species diversity on aboveground biomass, across vegetation types, observed mainly via the increasing effect of woodland structural diversity.
+We also found that the effect of tree species diversity on biomass increases with stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}.
+Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via its effect on species diversity, suggesting that tree diversity may have been under-appreciated as a determinant of savanna structure.
 
-\textbf{Main conclusions:} The study underlines the close association between tree diversity, ecosystem structure and ecosystem function of highly disturbed southern African savannas and woodlands.
+\textbf{Main conclusions:} The study underlines the close association between tree diversity and ecosystem structure and function of highly disturbed southern African savannas and woodlands.
 Our results demonstrate the importance of accounting for environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level.
-Biodiversity loss, predominantly through human actions in southern Africa, may have detrimental outcomes for ecosystem function, specifically woody biomass provision, which is critical in a region that relies on wood fuel as its principal domestic energy source.
 
@@
 
+% BEFR has had previous global research
 In order to understand the effects of global biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}.
-Numerous studies have shown relationships between biodiversity and ecosystem function (e.g.
-\citealt{Liang2016, Hooper2012, Cardinale2009}).
-The strength and direction of the Biodiversity-Ecosystem Function (BEF) relationship varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016}.
-Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}.
-These studies support early findings from small-scale experimental studies conducted predominantly in grassland ecosystems, which began in earnest during the 1990s as concern grew over the global loss of biodiversity \citep{Tilman1994, Tilman2014}.
-Despite these concerted efforts, we continue to lack a nuanced, ecosystem agnostic understanding of the complex interactions between biodiversity, environment, and ecosystem function.
+The strength and direction of the Biodiversity-Ecosystem Function (BEF) relationship varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016, Hooper2012, Cardinale2009}.
+Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}, which follow from early small-scale experimental studies conducted predominantly in artificial grassland mesocosms \citep{Tilman1994, Tilman2014}.
+Despite these concerted efforts, we continue to lack a nuanced, ecosystem-agnostic, understanding of the complex interactions between biodiversity, environment, and ecosystem function.
 
-Ecosystem functions can be defined in broad terms as rate processes and properties of ecosystems which describe the nature of biotic activity within an ecosystem \citep{Jax2005}.
-This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resistance of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported  relationship between biodiversity and ecosystem function invokes three main mechanisms to explain the relationship \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
+% Ecosystem function definition and mechanisms
+Ecosystem functions can be defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}.
+This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resilience of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
+The frequently reported BEF relationship invokes three main mechanisms to explain it \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
 
-Compared to forest ecosystems, dry tropical woodlands and savannas are highly structured by disturbance, mainly through fire and herbivory, with African savannas possessing large herbivores absent from other savannas \citep{Sankaran2008, Levick2009}.
-Disturbance via human activities such as timber extraction and charcoal processing is also common in African woodlands, often causing high levels of disturbance in localised areas \citep{Dewees2010}.
+% Savannas are big and ecologically important
+Savannas and woodlands are the dominant vegetation type across southern Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
+The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
+Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages.
+These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses.
+Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under international efforts to reduced deforestation and degradation (REDD+) \citep{Hinsley2015}.
+Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.
+
+% Savannas might not be the same as forests
+Compared to forest ecosystems, southern African dry tropical woodlands and savannas are highly structured by disturbance, through fire \citep{Lehmann2014}, herbivory \citep{Sankaran2008, Levick2009}, and human activities such as shifting cultivation agriculture \citep{Heinimann2017}, timber extraction and charcoal processing \citep{Dewees2010, McNicol2018b}.
 High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution.
-Disturbance may reduce stem density, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
-This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems.
+Disturbance reduces stem density and woody biomass, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
+This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEF relationship.
 Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}.
-Reduced stem density reduces competition between individuals, possibly weakening the effect of diversity on ecosystem function via niche complementarity.
-Similarly, more diverse species assemblages may lead to facilitation effects between certain species combinations under limiting environmental conditions such as low water availability.
+A threshold stem density may exist below which the effects of biodiversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.
+
+% Savannas might no be the same as forests continued: facilitation effects
+More diverse species assemblages may lead to facilitation effects between certain species combinations under the limiting environmental conditions prevalent across African savannas, such as low water availability.
 Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
@@
 
+% African savannas are understudied in this regard
 The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems.
 \citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}.
-A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on these ecosystems could greatly strengthen our understanding of a general BEF relationship and its environmental determinants.
+A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on the BEF in southern African woodlands could greatly strengthen our understanding of a global BEF relationship and its environmental determinants.
+A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}.
+Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
 
-Savannas and woodlands are the dominant vegetation type across the southern African region, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
-The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
-Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages.These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses.
-Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
-Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under the directive of the United Nations REDD+ programme or the similar Forest Carbon Partnership Facility (FCPF) \citep{Hinsley2015}.
-Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.
+% \citep{Michaletz2014, Michaletz2018} \citep{Spasojevic2014}
+%In wet tropical forests in Central America, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness.
+% Due the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
+% Environment affects the BEF relationship, so must be accounted for.
+Environmental variation is known to affect both woody biomass and tree species diversity independently, in a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
+Southern African woodlands particularly, occur over a wide range of precipitation, diurnal and annual temperature, and disturbance regimes \citep{Frost1996}.
+It is important therefore to account for environment and understand how it interacts with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass.
+\citet{Sankaran2005} and \citet{Lehmann2014} both report independently that total precipitation sets the upper limit for woody biomass in African savannas.
+\citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate.
 
-A small number of studies in southern African woodlands, all of which were restricted in the spatial scope to a small region of miombo woodland, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
-The results of these fine scale studies concur with similar studies in other biomes \citep{Cardinale2009}.
-Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
-Due to the variable environmental conditions within which southern African woodlands occur \citep{Frost1996}, with wide variation in precipitation, diurnal and annual temperature range, disturbance regime, and given the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
+% Disturbance deserves another look
+High levels of disturbance in southern African woodlands may moderate the observable BEF relationship through its effect on ecosystem composition.
+Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}.
+Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree.
+Selection effects may therefore be more important in maximising ecosystem function in disturbance-prone woodlands.
+If a given woodland plant community contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.
 
-In forests, climatic variation is known to affect both woody biomass \citep{Michaletz2014, Michaletz2018} and tree species diversity independently \citep{Spasojevic2014}.
-It is important therefore to account for climatic factors and understand how they interact with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass.
-\citet{Sankaran2005} used data from 854 African woodland field sites to show that below a threshold of \textapprox{650 mm MAP}, precipitation sets the upper limit for woody cover in savannas, which is positively correlated with biomass \citep{Chisholm2013, Prado-Junior2016}.
-In wet tropical forests in Central America, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness.
-
-Southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy, depending on their growth stages and species identity \citet{Solbrig1996}.
+% Structural diversity is important too
+Southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy, depending on their growth stages and species identity \citep{Solbrig1996}.
 This structural diversity may be one mechanism through which tree species diversity influences woody biomass.
@@
 
-High levels of disturbance in southern African woodlands may moderate the observable BEF relationship through its effect on ecosystem composition.
-Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}.
-Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree.
-Selection effects may therefore be more important in maximising ecosystem function in disturbance prone woodlands.
-If a given woodland plant community contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.
-
-In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots which span environmental and biogeographical gradients (\autoref{clust_map}).
-We used aboveground woody biomass of trees as our metric of ecosystem function, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability, energy input and soil fertility.
+% In this study :::
+In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots spanning broad environmental and biogeographical gradients (\autoref{clust_map}).
+We used aboveground woody biomass of trees as our metric of ecosystem function, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability and soil fertility.
 We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
 We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results.
-We used Structural Equation Modelling (SEM) with path analysis as a preferred method to simultaneously account for environmental and biotic factors, which may have interacting effects on ecosystem structure and therefore biomass.
-Initially, we posited three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the strength of the effect of tree species diversity on woody biomass will increase with plot-level stem density (number of stems ha\textsuperscript{-1}), as affected by resource availability and disturbance, due to an increased importance of niche complementarity as stem density and therefore competition increases, and (3) tree species diversity will increase tree structural diversity (i.e.
-physiognomic diversity), which will provide an indirect path by which tree diversity increases woody biomass.
+We used Structural Equation Modelling (SEM) with path analysis to simultaneously account for environmental and biotic factors, which may have interacting effects on ecosystem structure and therefore biomass.
+Initially, we posited three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with plot-level stem density (number of stems ha\textsuperscript{-1}), due to an increased importance of niche complementarity as stem density and therefore competition increases.
+In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass.
+Finally, we expect that (3) tree species diversity will increase tree structural diversity (i.e.
+physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.
 
@@
 
+% Situation
 The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealt{White1987}).
@@
 
+% Plots chosen
 Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup.
@@
 
-Many plots provided by the 2005-2008 Zambian Integrated Land Use Assessment \citep{Mukosha2009} were arranged in clusters of up to four 20x50 m plots, 20 metres apart.
-Plots within each cluster were aggregated before the plot dataset filtering described above and treated as a single plot in analyses.
+% Zambia disclaimer
+\nzam{} plots provided by the 2005-2008 Zambian Integrated Land Use Assessment \citep{Mukosha2009, Pelletier2018} were arranged in clusters of four 20x50 m plots, 20 metres apart.
+Data from each plot within a cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.
 
@@
 
-We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads which are functionally tree-like.
+% AGB
+We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads, which are functionally tree-like.
 Woody lianas are scarce in our study plots and were not measured.
@@
 
+% Number of stems
 All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements.
-A tree may be comprised of multiple stems and so tree-level diversity estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice.
+A tree may be comprised of multiple stems and so tree-level richness estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice.
 For each tree, we recorded species, DBH and tree height to the top of the highest branch material.
@@
 
-AGB for each plot was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:
+% AGB Calculation
+AGB for each plot (t ha\textsuperscript{-1}) was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:
 
@@
 
-where $\rho$ is the species level mean wood density, $D$ is the DBH\textsubscript{e} at 1.3 m, and $H$ is the tree height.
+where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), $D$ is the DBH\textsubscript{e} (cm) at 1.3 m, and $H$ is the tree height (m).
 Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
@@
 
+% WorldClim, SoilGrids and MODIS
 Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}.
@@
 Soil fertility data were extracted from the ISRIC gridded soil information data product at 250 m resolution, taking the grid cell value for each plot centre \citep{Hengl2017}.
-We extracted Cation Exchange Capacity (CEC), soil organic carbon stocks (kg m\textsuperscript{-2}) soil sand content by volume (Sand g kg\textsuperscript{-1}) and soil nitrogen content (g kg\textsuperscript{-1}).
+We extracted Cation Exchange Capacity (CEC) (cmolc kg\textsuperscript{-1}), soil organic carbon stocks (kg m\textsuperscript{-2}) percentage soil sand content (0.05-2 mm) by weight and soil nitrogen content (g kg\textsuperscript{-1}).
 These data are a modelled product derived from various remotely sensed and directly measured data sources.
-The degree of fire disturbance was calculated using the MODIS monthly burned area product at 0.01 degree resolution, 1.11 km at the equator (MCD64A1, \citealt{MODIS_burn}), as the number of observed fires between 2001 and 2018.
+The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1, \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018.
+We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.
 
@@
 Abundance evenness allows for greater niche complementarity at small scales due to potentially increased heterogeneity of functional traits.
-We quantified tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CV) and tree height (Height CV).
+We quantified tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CoV) and tree height (Height CoV).
 
@@
 Plots were assigned to vegetation type groups based on tree species composition.
-Groups were defined in \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
+Groups were defined in a manner adapted from \citet{Fayolle2018} in an Africa-wide analysis of floristic units using plot data in savannas and woodlands with tree species diversity and relative abundance data.
 Group identification was conducted using unconstrained correspondence analysis, followed by hierarchical clustering based on dominant ordination axes.
@@
 Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixa]{Appendix A}).
-Standardisation put each latent variable on the same scale, with a mean of zero and a standard deviation of one.
 Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
@@
 
+% Correlations
 Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
-stem density had the strongest correlation with AGB of all variables considered (\ccib{}).
+Stem density had the strongest correlation with AGB of all variables considered (\ccib{}).
 Environmental variables had weaker correlations with AGB than diversity variables, with all environmental variables having significant correlations with AGB, except fire frequency.
-
 The direction of these correlations was used as a test of our assumptions for the direction of influence of latent variables later used in the SEMs.
-As expected, there was a positive correlation between MAP and AGB (\ccmb{}), and a weak negative correlation between the seasonality of precipitation and AGB (\ccmcb{}).
-Temperature stress was weakly negatively correlated  with AGB (\cctcb{}).
-As expected, there was a negative correlation between soil sand content and AGB (\ccsb{}), and a positive correlation between soil nutrient content and AGB (carbon: \ccob{}, nitrogen: \ccnb{}).
-
 MAP had positive correlations with all tree species diversity and structural diversity variables.
@@
 
-When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the four vegetation types, model fit and the strengths of unstandardised path coefficients varied.
-The direct effect of tree species diversity on AGB was positive and marginally significant in ex-Acacia (\strucbetacsb{}) but negligible in Mopane (\strucbetadsb{}), sparse miombo / Baikiaea (\strucbetaasb{}) and Core miombo (\strucbetabsb{}) (\autoref{struc_model_slopes_all}).
+When the tree species and structural diversity model (\autoref{struc_mod}) was refitted separately using data from each of the four vegetation types, we found that the effect sizes of each latent variable remained largely similar, though model fit varied.
+The direct effect of tree species diversity on AGB was positive and marginally significant in ex-Acacia (\strucbetacsb{}) but negligible in Mopane (\strucbetadsb{}), sparse miombo / \textit{Baikiaea} (\strucbetaasb{}) and Core miombo (\strucbetabsb{}) (\autoref{struc_model_slopes_all}).
 Relationships between structural diversity and AGB remained generally similar, with the same sign and overlap between the 95\% confidence intervals of path coefficients.
@@
 Path coefficients are $\pm$1 standard error.
-Path coefficients where the standard error does not overlap zero are considered to be significant effects.}
+Path coefficients where the interval (standard error) does not overlap zero are considered to be significant effects.}
 	\label{struc_model_slopes_all}
@@
 In our sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}).
-There appears to be a minimum stem density threshold at \textapprox{}180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant, small baseline effect of tree diversity on biomass.
+There appears to be a minimum stem density threshold at \textapprox{}180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass.
 The effect of structural diversity on AGB appears to remain constant with increasing stem density.
-The indirect effect of species diversity on AGB via structural diversity climbs as stem density increases.
+The indirect effect of tree species diversity on AGB via structural diversity climbs as stem density increases.
 
@@
 
-A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, \autoref{full_model_slopes}).
+A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, \hyperref[appendixd]{Appendix D}).
 The direct effects of water availability, soil fertility and disturbance on AGB were negligible (water: \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effect on AGB coming from the indirect paths via stem density (water: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (water: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}).
@@
 The total effect of species diversity on AGB was positive (\fmbetatotaldb{}).
+Compared to the simple model with no environmental covariates, the total explanatory power of tree species diversity and structural diversity in this model decreased (), but the predictive power of the model as a whole increased ().
 
@@
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{full_model_slopes}
-	\caption{Unstandardised path coefficients for the full model including tree species diversity, environmental covariates and stem density.
-Path coefficients are $\pm1$ standard error.
-Path coefficients where the standard error does not overlap zero are considered significant effects.}
-	\label{full_model_slopes}
-\end{figure}
-
 \section{Discussion}
 
+% Re-cap
 In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots.
 Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}).
-We found that the effect size of tree species diversity on AGB increased with stem density (H\textsubscript{2}).
 Tree species diversity, structural diversity and stem density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}).
+We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} below which the effect of species diversity on AGB remained at a low baseline level.
 The strongest effect on AGB was that of stem density.
@@
 
+% 
 We found a consistent positive effect of tree species diversity on AGB across all models in the current study.
-Within southern African woodlands we therefore find support for our hypothesis that higher tree species richness and evenness leads to higher woody AGB.
+Within southern African woodlands we therefore find support that higher tree species richness and evenness leads to higher woody AGB.
 This finding is in agreement with many other studies across different ecosystems and biomes, supporting the idea that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
@@
 The opposite causation is also plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
-We attempted to correct for the correlation between species richness and stem density by estimating, and using in models, the extrapolated species richness.
+We attempted to correct for the correlation between species richness and stem density using extrapolated species richness, which extrapolates a rarefaction curve to its predicted asymptote, thus estimating the total landscape-level species richness which is unaffected by plot size and stem density.
 We suggest therefore that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can utilise more available light and moisture, resulting in greater plot-level AGB.
@@
 
-We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity, and we therefore find support for our second hypothesis (H\textsubscript{2}).
+We found evidence that tree species diversity led to an increase in AGB indirectly via tree structural diversity, and we therefore find support for our second hypothesis H\textsubscript{2}.
 A higher tree species diversity allows for a greater structural diversity of trees, i.e.
@@
 This may act as a mechanism for niche complementarity, with a canopy of diversely sized trees able to take advantage of a greater proportion of the available light.
-Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{}, i.e.
+Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{Pretzsch2014}, i.e.
 wider variation in canopy shape and overall growth form; broad flat crowns vs.
@@
 Compared to moist tropical forests, water availability is more of a limiting factor to tree growth in southern African woodlands, which experience frequent drought.
-The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density, however, the direct effect on AGB remained positive and marginally significant, as expected.
 Disturbance by fire had a negative total effect on AGB.
 We found negligible indirect effects of disturbance on AGB via species diversity and structural diversity.
+
+A negative total effect of soil fertility on AGB is in contrast to other studies in the region and against general ecological theory, which predicts a positive effect of soil nutrients on biomass.
+The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density, however, the direct effect on AGB remained positive and marginally significant, as expected.
+Model estimates of the effect of soil on AGB were poorly constrained compared with other latent variables.
+This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used.
+SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales \citep{Muledi2017, Bucini2007} in southern African woodlands, often being further structured by the vegetation overlying it, an aspect which SoilGrids does not model precisely.
+Due to the plots used in this study often being situated non-randomly in the landscape, coupled with the coarseness of the SoilGrids data, it is not surprising that this model path is poorly constrained.
+Soil data is time-consuming to collect and difficult to compare across studies when different protocols are used, though this study prompts the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands.
+\citet{Lehmann2014} similarly found weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, fire and tree basal area.
 
@@
 We postulate that the slight negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in ex-Acacia and Mopane woodlands, where the top level of the woodland canopy is dominated often by a single species.
-Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation, while in the species poor Mopane and ex-Acacia woodlands, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.
+
+Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation and suppressing the few species which tend to create high biomass, such as \textit{Julbernadia} and \textit{Brachystegia} spp..
+In the species poor Mopane and ex-Acacia woodlands however, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.
 
@@
 
-ex-Acacia woodlands showed the strongest total effect of species diversity on AGB and was the only vegetation type to show a clearly positive direct effect of species diversity on AGB.
-ex-Acacia woodlands also had relatively low median species richness compared to miombo, but the addition of new species appears to make a larger difference to the AGB of these plots than in mopane woodlands.
+Ex-Acacia woodlands showed the strongest total effect of species diversity on AGB and was the only vegetation type to show a significant positive direct effect of species diversity on AGB.
+Ex-Acacia woodlands also had relatively low median species richness compared to miombo, but the addition of new species appears to make a larger difference to the AGB of these plots than in Mopane woodlands.
 We suggest that this is due mostly to the particular identity of species found in ex-Acacia woodlands and their contribution to ecosystem functioning.
-Unlike mopane woodlands, ex-Acacia woodlands do sometimes contain species other than \textit{Senegalia} and \textit{Vachellia} spp.
-which can grow to be high biomass canopy trees, especially in transition zones.
+Unlike Mopane woodlands, ex-Acacia woodlands contain a wider variety of species which can grow to large canopy trees, albeit at low densities, especially in transition zones with miombo woodlands.
 
@@
 
-In this study we found that across southern African woodlands, there is a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB).
-Additionally, we found that much of this effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees.
+In this study we found that even in highly disturbed southern African woodlands, there exists a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB).
+Additionally, we found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees.
 We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, suggesting that similar ecosystem processes occur across the region to determine ecosystem function.
-In contrast to previous studies, we found, at the scale of our study region, that the direct effects of water availability and soil fertility on woody biomass were negligible, with most of their effect being indirectly through their effects on tree species and structural diversity.
+In contrast to previous studies, we found, at the scale of our study region, that the direct effects of water availability and soil fertility on woody biomass were small, with most of their effect being indirectly through their effects on tree species and structural diversity.
 This strongly suggests that data on tree species diversity be included into models predicting ecosystem functionality in this region.
@@
 
-Southern African woodlands are relied heavily upon for their ecosystem service provision, which is itself related to ecosystem function.
-Resource extraction by humans in southern Africa is directly influencing biodiversity via selective tree-felling for timber, among other forest products.
-Our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.
+The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in favour of disturbance effects.
+On a practical note, our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.
 This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.
@@
 A single linear regression is presented as a black line, which combines all vegetation types, separate loess trend lines are fitted for each vegetation type.
+An orange loess trend line is fitted for all the data.
 All data is standardised and variables are transformed where it was appropriate for analysis.}
@@
 
+\section{Appendix 4 - Path coefficients for model incorporating environmental covariates} \label{appendixd}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{full_model_slopes}
+	\caption{Unstandardised path coefficients for the full model including tree species diversity, environmental covariates and stem density.
+Path coefficients are $\pm1$ standard error.
+Path coefficients where the interval (standard error) does not overlap zero are considered to be significant effects.}
+	\label{full_model_slopes}
+\end{figure}
+
 \end{document}

### 2020-08-18 build.sh

@@ -9,10 +9,10 @@ INC="manuscript/include/"
 
 # Run data compilation
 #Rscript scripts/temp_precip.R
-#Rscript scripts/data_clean.R
-#Rscript scripts/standardise.R
-#Rscript scripts/descrip.R
-#Rscript scripts/sem.R
+Rscript scripts/data_clean.R
+Rscript scripts/standardise.R
+Rscript scripts/descrip.R
+Rscript scripts/sem.R
 
 # Transfer images to manuscript
 cp img/clust_map.pdf $IMG  # Cluster map

### 2020-08-18 scripts/nitrogen_get.sh

@@ -1,9 +1,9 @@
 #!/usr/bin/env bash
 
-curlget=$(awk -v FS="," '{ printf "https://rest.soilgrids.org/soilgrids/v2.0/properties/query?lon=%s&lat=%s&property=nitrogen&depth=15-30cm&value=mean\n", $1, $2 }' plots_coord.csv > curlget)
+awk -v FS="," '{ printf "https://rest.soilgrids.org/soilgrids/v2.0/properties/query?lon=%s&lat=%s&property=nitrogen&depth=15-30cm&value=mean\n", $1, $2 }' ../plots_coord.csv > curlget
 
-for line in curlget; do 
-	curl -k -X GET $line -H "accept: application/json" | jq '.properties.layers | .[] | .depths | .[] | .values.mean'
-done > ../data/nitrogen 
+while IFS="" read -r p || [ -n "$p" ]; do
+	curl -k -X GET $p -H "accept: application/json" | jq '.properties.layers | .[] | .depths | .[] | .values.mean'
+done < curlget > ../data/nitrogen.txt
 
-
+rm curlget

### 2020-08-18 scripts/sem.R

@@ -38,7 +38,6 @@ library(lme4)
 library(stargazer)
 library(tidyr)
 library(psych)
-library(beepr)
 library(ggnewscale)
 library(maps)
 library(gganimate)
@@ -125,15 +124,15 @@ ggplot() +
       fill = raw.r), colour = "black") + 
   geom_text(data = corr_ci, 
     aes(x = x_var, y = y_var, label = raw.r),
-    size = 3) + 
+    size = 4) + 
   geom_point(data = corr_ci[corr_ci$conf == FALSE,], 
-    aes(x = x_var, y = y_var), fill = NA, colour = "black", shape = 21, size = 11) + 
+    aes(x = x_var, y = y_var), fill = NA, colour = "black", shape = 21, size = 12) + 
   scale_fill_gradient2(name = "r", low = "blue", mid = "white", high = "red") + 
   theme_classic() + 
-  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5,
+  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12,
       colour = c(rep("#D65A2D", 2), rep("#331d8b", 1), rep("#287F9C", 4), 
         rep("#468A21", 2), rep("#844099", 2), rep("black", 2))),
-    axis.text.y = element_text(
+    axis.text.y = element_text(size = 12,
       colour = c(rep("#D65A2D", 3), rep("#331d8b", 1), rep("#287F9C", 4), 
         rep("#468A21", 2), rep("#844099", 2), "black")),
     legend.position = "none") + 
@@ -239,14 +238,14 @@ mod_summ_mutate_struc <- function(x){
       op == "~" & lhs == "agb_ha_log_std" ~ "Direct: AGB",
       TRUE ~ "Other"),
     rhs = case_when(
-      op == "Direct: AGB" & rhs == "div" ~ "Species div.",
-      op == "Direct: AGB" & rhs == "struc" ~ "Struct. div.",
-      op == "Direct: AGB" & rhs == "stems_ha_log_std" ~ "Stem dens.",
-      lhs == "struc" & op == "Other" & rhs == "div" ~ "Species div. -> Struct. div.",
-      lhs == "stems_ha_log_std" & op == "Other" & rhs == "div" ~ "Species div. -> Stem dens.",
-      op == "Other" & rhs == "a*b" ~ "Indirect: Species div. -> Struct. div. -> AGB",
[…]

### 2020-08-18 scripts/standardise.R

@@ -29,7 +29,6 @@ source("scripts/clust_defin.R")
 
 dat <- readRDS("data/plot_data_fil_agg.rds")
 
-
 # Create histogram labels ----
 
 facet_levels <- c(
@@ -64,14 +63,14 @@ facet_labels <- c(
   expression("Temp." ~ "stress"),
   expression("AGB" ~ (t ~ ha^-1)),
   expression("log(AGB)" ~ (t ~ ha^-1)),
-  expression("CEC"),
+  expression("CEC" ~ (cmolc ~ kg^-1)),
   expression("Sand" ~ "%"), 
-  expression("Org." ~ "C" ~ "(ppt)"),
+  expression("Org." ~ "C" ~ (kg ~ m^-2)),
   expression("log(Org." ~ "C)" ~ "(ppt)"),
   expression("Nitrogen" ~ (g ~ kg^-1)),
   expression("log(Nitrogen)" ~ (g ~ kg^-1)),
-  expression("Mean" ~ "fire" ~ "freq." ~ (2001-2018)),
-  expression("log(Mean" ~ "fire" ~ "freq.)" ~ (2001-2018)),
+  expression("Fire" ~ "freq." ~ (2001-2018)),
+  expression("log(Fire" ~ "freq.)" ~ (2001-2018)),
   expression("Herbivore" ~ "biomass" ~ (kg ~ ha^-1)),
   expression("Coef." ~ "var." ~ "DBH"),
   expression("Coef." ~ "var." ~ "height"),
@@ -309,9 +308,6 @@ for (i in 1:length(bivar_list)) {
 }
 
 # Arrange on grid
-n <- length(plot_list)
-n_col <- floor(sqrt(n))
-
 pdf(file =  "img/bivar_lm.pdf", width = 10, height = 14)
 do.call("grid.arrange", c(plot_list, ncol = 4))
 dev.off()

### 2020-09-14 manuscript/regional_befr_manuscript.tex
Added minor comments to manuscript

@@
 This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEF relationship.
-Instead, stress tolerance and the functional contribution of more abundant species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}.
-A threshold stem density may exist below which the effects of biodiversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.
+Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}.
+A threshold stem density may exist below which the effects of tree species diversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.
 
@@
 The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}.
-Studies of the BEF relationship often find that at fine scales, biodiversity shows a strong effect on ecosystem function, but at broad scales biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
+Studies of the BEF relationship often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>100s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
 
@@
 % Environment affects the BEF relationship, so must be accounted for.
-Environmental variation is known to affect both woody biomass and tree species diversity independently, in a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
+Environmental heterogeneity is known to affect both woody biomass and tree species diversity independently, in a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
 Southern African woodlands particularly, occur over a wide range of precipitation, diurnal and annual temperature, and disturbance regimes \citep{Frost1996}.
-It is important therefore to account for environment and understand how it interacts with biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass.
+It is important therefore to account for this environmental heterogeneity and understand how it influences both biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass.
 \citet{Sankaran2005} and \citet{Lehmann2014} both report independently that total precipitation sets the upper limit for woody biomass in African savannas.
@@
 The total effect of species diversity on AGB was positive (\fmbetatotaldb{}).
-Compared to the simple model with no environmental covariates, the total explanatory power of tree species diversity and structural diversity in this model decreased (), but the predictive power of the model as a whole increased ().
+Compared to the simple model with no environmental covariates, the total explanatory power of tree species diversity and structural diversity in this model decreased, but the predictive power of the model as a whole increased.
 
@@
 A negative total effect of soil fertility on AGB is in contrast to other studies in the region and against general ecological theory, which predicts a positive effect of soil nutrients on biomass.
-The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density, however, the direct effect on AGB remained positive and marginally significant, as expected.
+The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density.
+The direct effect on AGB however, remained positive and marginally significant, as expected.
 Model estimates of the effect of soil on AGB were poorly constrained compared with other latent variables.

### 2020-09-15 manuscript/regional_befr_manuscript.tex
Updated references to APA6th edition

@@
 % Bibliography management
-\usepackage{natbib}    % Bibliography management - Use author/date citations
-\bibliographystyle{agsmnourl}  % Use custom agsm bibliography template with no URL
-\usepackage{cite}  % Citation options
+\usepackage[UKenglish]{babel}
+\usepackage[natbibapa]{apacite}
+\bibliographystyle{apacite}
 
@@
 \newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}   % \todo{NOTE TO SELF WRITTEN IN RED}
-
-\usepackage[clean]{revdiff}
 
@@
 
-{\LARGE{Title: Tree biodiversity and ecosystem function relationship across southern African woodlands}}
+{\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across southern African woodlands}}
 
@@
 
-Authors: Godlee, J. L.\textsuperscript{1}, Dexter, K. G.\textsuperscript{1,2}, Ryan, C. M.\textsuperscript{1}
+Authors: John L.
+Godlee\textsuperscript{1}, Casey M.
+Ryan\textsuperscript{1}, David Bauman\textsuperscript{2,3}, Samuel J.
+Bowers\textsuperscript{1}, Joao M.
+B.
+Carreiras\textsuperscript{4}, Antonio Valter Chisingui\textsuperscript{5}, Joris P.
+G.
+M.
+Cromsigt\textsuperscript{6,7,8}, Dave J.
+Druce\textsuperscript{9,10}, Manfred Finkch\textsuperscript{11}, Francisco Maiato Gon\c{c}alves\textsuperscript{5}, Ricardo M.
+Holdo\textsuperscript{12}, Steve Makungwa\textsuperscript{13}, Iain M.
+McNicol\textsuperscript{1}, Edward T.
+A.
+Mitchard\textsuperscript{1}, Anderson Muchawona\textsuperscript{14}, Rasmus Revermann\textsuperscript{11,15}, Natasha Sofia Ribeiro\textsuperscript{16}, Abel Siampale\textsuperscript{17}, Stephen Syampungani\textsuperscript{18}, Jos\'{e} Jo\~{a}o Tchamba\textsuperscript{5}, Hemant G.
+Tripathi\textsuperscript{19}, Johannes Wallenfang\textsuperscript{11}, Mariska te Beest\textsuperscript{20,21,22}, Mathew Williams\textsuperscript{1}, Dexter, K.
+G.\textsuperscript{1,2}
 
-\textsuperscript{1}: School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom \\
-\textsuperscript{2}: Royal Botanic Garden Edinburgh, Edinburgh, EH3 5LR, United Kingdom
+1: School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom \\
+2: Environmental Change Institute, School of Geography and the Environment, University of Oxford, Oxford, United Kingdom \\
+3: Laboratoire d'\'{E}cologie V\'{e}g\'{e}tale et Biogéochimie, CP 244, Universit\'{e} Libre de Bruxelles, Brussels, Belgium \\
+4: National Centre for Earth Observation (NCEO), University of Sheffield, Hicks Building, Hounsfield Road, Sheffield, S3 7RH, United Kingdom \\
+5: Herbarium of Lubango, ISCED Hu\'{i}la, Sarmento Rodrigues Str. No. 2, CP. 230, Lubango, Angola \\
+6: Department of Wildlife, Fish, and Environmental Studies, Swedish University of Agricultural Sciences, Ume\aa, Sweden \\
+7: Department of Zoology, Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, South Africa \\
+8: Copernicus Institute of Sustainable Development, Utrecht University, 3584CS Utrecht, the Netherlands \\
+9: Ecological Advice, Ezemvelo KZN Wildlife, Hluhluwe-iMfolozi Park, South Africa \\
+10: School of Life Sciences, University of KwaZulu-Natal, Pietermaritzburg, South Africa \\
+11: Biodiversity, Evolution and Ecology of Plants, Institute of Plant Sciences and Microbiology, University of Hamburg, Ohnhorststr. 18, 22609 Hamburg, Germany \\
+12: Odum School of Ecology, University of Georgia, 140 E. Green St., Athens, GA 30602, USA \\
+13: Lilongwe University of Agriculture and Natural Resources (LUANAR), Lilongwe, Malawi \\
+14: Forest Research Center, 1 Orange Groove Highlands, Harare, Zimbabwe \\
+15: Faculty of Natural Resources and Spatial Sciences, Namibia University of Science and Technology, Windhoek, Namibia \\
+16: Department of Forest Engineering, Faculty of Agronomy and Forest Engineering, UEM Av. Julius Nyerere, 3453, Campus Universitario, Maputo, Mozambique \\
+17: Forestry Department Headquarters - MLNR, Cairo Road, Lusaka, Zambia \\
+18: School of Natural Resources, Copperbelt University, Kitwe, Zambia \\
+19: Faculty of Biological Sciences, University of Leeds, LS2 9JT, United Kingdom\\
+20: Copernicus Institute of Sustainable Development, Utrecht University, 3508 TC Utrecht, The Netherlands \\
+21: Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, 6031, South Africa \\
+22: South African Environmental Observation Network, Grasslands-Forests-Wetlands Node, Montrose, 3201, South Africa \\
+23: Royal Botanic Garden Edinburgh, Edinburgh, EH3 5LR, United Kingdom \\
 
@@
 
-This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (J.
-L.
+This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (John L.
 Godlee, Grant No.
 NE/L002558/1).
-The data provided for this study was constributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No.
+The data for this study was contributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No.
 NE/P008755/1).
+Revisions of the SEOSAW dataset were funded by SavannaChange, a GCRF/University of Edinburgh funded project.
 We thank all data providers and the field assistance they received when collecting plot data.
 JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).
+
+\section{Conflict of interest statement}
+
+All authors declare no conflict of interest regarding this study.
 
@@
 
-\LARGE{Title: An assessment of the biodiversity - ecosystem function relationship in southern African woodlands}
+\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across southern African woodlands}
 
-\normalsize{Running title: Ecosystem function in southern African woodlands}
+\normalsize{Running title: Diversity - ecosystem function in southern African woodlands}
 
@@
 Our results demonstrate the importance of accounting for environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level.
+
+\textbf{Keywords:} biodiversity, ecosystem function, woodland, miombo, biomass, structural equation modelling, forest structure.
 
@@
 The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}.
-Studies of the BEF relationship often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>100s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
+Studies of the BEF relationship often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>10000s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
 
@@
 % Situation
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealt{White1987}).
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}; \citealt{White1987}).
 The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
@@
 These data are a modelled product derived from various remotely sensed and directly measured data sources.
-The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1, \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018.
+The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1; \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018.
 We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.
@@
 
-Estimated tree species richness was calculated for each plot using \verb|ChaoRichness()| from the \verb|iNEXT| package in R \citep{Hsieh2016}.
+Estimated tree species richness was calculated for each plot using `ChaoRichness()' from the `iNEXT' package in R \citep{Hsieh2016}.
 This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
@@
 
-\begin{landscape}
-\input{include/clust_summ}
-
-\begin{figure}[H]
-\centering
-	\includegraphics[width=1.4\textwidth]{clust_map}
-	\caption{The locations of the \nplots{} plots used in this study, with respect to the distribution of miombo woodland vegetation according to \citet{White1987}.
-Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
-	\label{clust_map}
-\end{figure}
-\end{landscape}
-
 \subsubsection{Structural Equation Modelling}
@@
 We used Structural Equation Modelling (SEM) to investigate the determinants of AGB.
-All SEMs were constructed and analysed in the \verb|lavaan| package \citep{lavaan} in R version 3.6.0 \citep{R2019}.
+All SEMs were constructed and analysed in the `lavaan' package \citep{lavaan} in R version 3.6.0 \citep{R2019}.
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
@@
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=\textwidth]{concept}
-	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB.
-Hypothesised paths of causation are depicted as arrows from predictor to response.}
-	\label{con_mod}
-\end{figure}
-
 Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixa]{Appendix A}).
 Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
-Standardisation also controls for variables with variation across different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in \verb|lavaan|.
+Standardisation also controls for variables with variation across different orders of magnitude, which could otherwise prevent adequate model estimation from the covariance matrix in `lavaan'.
 To ensure that observed variables within a latent variable had consistent directions of influence, some observed variables had their sign reversed.
@@
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlation matrix of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
-Correlation coefficients marked by a circle indicate that the 95\% confidence interval of $r$ overlapped zero.
-Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = disturbance, turquoise = water availability, green = tree species diversity, purple = tree structural diversity.
-See \hyperref[appendixb]{Appendix B} for a full assessment of correlation fit statistics.}
-	\label{corr_mat}
-\end{figure}
-
 % Correlations
@@
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=0.5\textwidth]{struc}
-	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all vegetation clusters.
-Latent variables are shown as circles while observed variables are shown as rectangles.
-Standardised path coefficients are shown as solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
-The observed variables that inform the latent variables are connected by dotted arrows, and observed variables with loadings set to one are connected by dashed arrows.
-Measurement errors of exogenous variables are omitted for clarity.}
-	\label{struc_mod}
-\end{figure}
-
 \subsection{Variation among vegetation types}
@@
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{struc_model_slopes_all}
-	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
-Path coefficients are $\pm$1 standard error.
-Path coefficients where the interval (standard error) does not overlap zero are considered to be significant effects.}
-	\label{struc_model_slopes_all}
-\end{figure}
-
-\input{include/struc_model_fit_clust_stats}
-
 \subsection{Moderation of Diversity-AGB relationship by stem density}
@@
 The indirect effect of tree species diversity on AGB via structural diversity climbs as stem density increases.
-
-\begin{figure}[H]
-\centering
-	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves with standard error shaded bars.}
-	\label{sem_struc_stems_ha}
-\end{figure}
 
@@
 Compared to the simple model with no environmental covariates, the total explanatory power of tree species diversity and structural diversity in this model decreased, but the predictive power of the model as a whole increased.
-
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{full}
-	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation types.
-Latent variables are shown as circles while observed variables are shown as rectangles.
-Standardised path coefficients are shown as solid arrows pointing from predictor to response, with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
-Observed variables that inform the latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
-Measurement errors of exogenous variables are omitted for clarity.}
-	\label{full_mod}
-\end{figure}
 
@@
 
-\bibliography{lib.bib}
+\bibliography{regional_befr_manuscript.bib}
 
 \newpage{}
-\appendix{}
 
-\section{Data accessibility statement}
+\section{Author contribution statement}
 
+JG and KD conceived the study.
+JG conducted data analysis, data management for further versions of the SEOSAW dataset, and wrote the manuscript.
+CR conceived the SEOSAW database and conducted data management for earlier versions of the SEOSAW dataset.
+JG, CR, DB, JMBC, MF, RH, EM, SS, HT, HT, MB, MW, and KD contributed to manuscript revisions.
+JG, CR, SB, VC, JPGMC, DD, MF, FG, SM, IM, AM, RR, NR, AS, SS, JT, JW, MB, and MW contributed to experimental design, field data collection, data preparation and data management of parts of the dataset used in this study.
+
+\begin{landscape}
 \section{Tables}
+\input{include/clust_summ}
+\end{landscape}
 
+\input{include/struc_model_fit_clust_stats}
+
+\begin{landscape}
 \section{Figure legends and embedded figures}
+\begin{figure}[H]
+\centering
+	\includegraphics[width=1.4\textwidth]{clust_map}
+	\caption{The locations of the \nplots{} plots used in this study, with respect to the distribution of miombo woodland vegetation according to \citet{White1987}.
+Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
+	\label{clust_map}
+\end{figure}
+\end{landscape}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=\textwidth]{concept}
+	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB.
+Hypothesised paths of causation are depicted as arrows from predictor to response.}
+	\label{con_mod}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=0.6\textwidth]{corr_mat}
+	\caption{Correlation matrix of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
+Correlation coefficients marked by a circle indicate that the 95\% confidence interval of $r$ overlapped zero.
+Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = disturbance, turquoise = water availability, green = tree species diversity, purple = tree structural diversity.
+See \hyperref[appendixb]{Appendix B} for a full assessment of correlation fit statistics.}
+	\label{corr_mat}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=0.5\textwidth]{struc}
+	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all vegetation clusters.
+Latent variables are shown as circles while observed variables are shown as rectangles.
+Standardised path coefficients are shown as solid arrows pointing from predictor to response with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
+The observed variables that inform the latent variables are connected by dotted arrows, and observed variables with loadings set to one are connected by dashed arrows.
+Measurement errors of exogenous variables are omitted for clarity.}
+	\label{struc_mod}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{struc_model_slopes_all}
+	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
+Path coefficients are $\pm$1 standard error.
+Path coefficients where the interval (standard error) does not overlap zero are considered to be significant effects.}
+	\label{struc_model_slopes_all}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
+	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves with standard error shaded bars.}
+	\label{sem_struc_stems_ha}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+	\includegraphics[width=\textwidth]{full}
+	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation types.
+Latent variables are shown as circles while observed variables are shown as rectangles.
+Standardised path coefficients are shown as solid arrows pointing from predictor to response, with the effect size of the path coefficient expressed in terms of standard deviations on the latent variable response scale.
+Observed variables that inform the latent variables are connected by dotted arrows, observed variables with loading set to one are connected by dashed arrows.
+Measurement errors of exogenous variables are omitted for clarity.}
+	\label{full_mod}
+\end{figure}
 
@@
 
+\appendix{}
 \section{Appendix 2 - Table of correlation fit statistics} \label{appendixb}

### 2020-09-17 manuscript/include/clust_summ.tex
Anna's proofread

@@
 \hline
-ex-Acacia & \begin{tabular}[c]{@{}c@{}c@{}} Spirostachys africana \\\textit{Senegalia burkei} \\\textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\\textit{Vachellia nilotica} \\\textit{Spirostachys africana} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\
+ex-Acacia & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Spirostachys africana} \\\textit{Senegalia burkei} \\\textit{Senegalia nigrescens} \end{tabular} & \begin{tabular}[c]{@{}c@{}c@{}} \textit{Euclea racemosa} \\\textit{Vachellia nilotica} \\\textit{Spirostachys africana} \end{tabular} & 188 & 12(10.3) & 181(166.5) & 54.5(61.33) \\
 \hline

### 2020-09-17 manuscript/regional_befr_manuscript.tex
Anna's proofread

@@
 Mitchard\textsuperscript{1}, Anderson Muchawona\textsuperscript{14}, Rasmus Revermann\textsuperscript{11,15}, Natasha Sofia Ribeiro\textsuperscript{16}, Abel Siampale\textsuperscript{17}, Stephen Syampungani\textsuperscript{18}, Jos\'{e} Jo\~{a}o Tchamba\textsuperscript{5}, Hemant G.
-Tripathi\textsuperscript{19}, Johannes Wallenfang\textsuperscript{11}, Mariska te Beest\textsuperscript{20,21,22}, Mathew Williams\textsuperscript{1}, Dexter, K.
-G.\textsuperscript{1,2}
+Tripathi\textsuperscript{19}, Johannes Wallenfang\textsuperscript{11}, Mariska te Beest\textsuperscript{20,21,22}, Mathew Williams\textsuperscript{1}, Kyle G.
+Dexter\textsuperscript{1,23}
 
@@
 % Situation
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}; \citealt{White1987}).
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}; \citealp{White1987}).
 The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
@@
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
-Only plots >0.1 hectares were used in analysis, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
+Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
 Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than ``grassy savanna'', which is considered a separate biome with very different species composition \citep{Parr2014}.
@@
 We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
-Structural diversity can also directly influence AGB in this model, without an effect of species diversity.
+Structural diversity can also directly influence AGB in this model, separate to the effect of of species diversity.
 To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model.
@@
 % Re-cap
-In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots.
+In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots.
 Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}).
@@
 
-We found a non linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
-At low stem densities competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}.
+We found a non-linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
+At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}.
 At very high stem density, there is also an increase in the effect of species diversity on structural diversity.
 This could be because at high stem density, the adaptation of different species to growth form become important.
-At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species ara able to exist in the understory, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.
+At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species are able to exist in the understorey, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.
 
@@
 Compared to ex-Acacia and Mopane woodlands, miombo woodlands have higher median tree species richness.
-ex-Acacia and Mopane woodlands are dominated by fewer tree species, notably \textit{Senegalia} spp.
+Ex-Acacia and Mopane woodlands are dominated by fewer tree species, notably \textit{Senegalia} spp.
 in ex-Acacia woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees.
@@
 Higher functional redundancy among tree species in miombo woodlands may lead to smaller trees with lower AGB in the most diverse plots, more resembling thicket vegetation and suppressing the few species which tend to create high biomass, such as \textit{Julbernadia} and \textit{Brachystegia} spp..
-In the species poor Mopane and ex-Acacia woodlands however, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.
+In the species-poor Mopane and ex-Acacia woodlands however, the addition of extra species may fill a greater proportional niche space, thus increasing total AGB more.
 

### 2020-09-17 manuscript/regional_befr_manuscript.tex
Submitted

@@
 
-\title{Tree biodiversity and ecosystem function relationship across southern African woodlands}
+\title{Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}
 \author{}
@@
 
-{\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across southern African woodlands}}
-
-\vspace{1cm}
-
-Authors: John L.
-Godlee\textsuperscript{1}, Casey M.
-Ryan\textsuperscript{1}, David Bauman\textsuperscript{2,3}, Samuel J.
-Bowers\textsuperscript{1}, Joao M.
-B.
-Carreiras\textsuperscript{4}, Antonio Valter Chisingui\textsuperscript{5}, Joris P.
-G.
-M.
-Cromsigt\textsuperscript{6,7,8}, Dave J.
-Druce\textsuperscript{9,10}, Manfred Finkch\textsuperscript{11}, Francisco Maiato Gon\c{c}alves\textsuperscript{5}, Ricardo M.
-Holdo\textsuperscript{12}, Steve Makungwa\textsuperscript{13}, Iain M.
-McNicol\textsuperscript{1}, Edward T.
-A.
-Mitchard\textsuperscript{1}, Anderson Muchawona\textsuperscript{14}, Rasmus Revermann\textsuperscript{11,15}, Natasha Sofia Ribeiro\textsuperscript{16}, Abel Siampale\textsuperscript{17}, Stephen Syampungani\textsuperscript{18}, Jos\'{e} Jo\~{a}o Tchamba\textsuperscript{5}, Hemant G.
-Tripathi\textsuperscript{19}, Johannes Wallenfang\textsuperscript{11}, Mariska te Beest\textsuperscript{20,21,22}, Mathew Williams\textsuperscript{1}, Kyle G.
-Dexter\textsuperscript{1,23}
-
-1: School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom \\
-2: Environmental Change Institute, School of Geography and the Environment, University of Oxford, Oxford, United Kingdom \\
-3: Laboratoire d'\'{E}cologie V\'{e}g\'{e}tale et Biogéochimie, CP 244, Universit\'{e} Libre de Bruxelles, Brussels, Belgium \\
-4: National Centre for Earth Observation (NCEO), University of Sheffield, Hicks Building, Hounsfield Road, Sheffield, S3 7RH, United Kingdom \\
-5: Herbarium of Lubango, ISCED Hu\'{i}la, Sarmento Rodrigues Str. No. 2, CP. 230, Lubango, Angola \\
-6: Department of Wildlife, Fish, and Environmental Studies, Swedish University of Agricultural Sciences, Ume\aa, Sweden \\
-7: Department of Zoology, Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, South Africa \\
-8: Copernicus Institute of Sustainable Development, Utrecht University, 3584CS Utrecht, the Netherlands \\
-9: Ecological Advice, Ezemvelo KZN Wildlife, Hluhluwe-iMfolozi Park, South Africa \\
-10: School of Life Sciences, University of KwaZulu-Natal, Pietermaritzburg, South Africa \\
-11: Biodiversity, Evolution and Ecology of Plants, Institute of Plant Sciences and Microbiology, University of Hamburg, Ohnhorststr. 18, 22609 Hamburg, Germany \\
-12: Odum School of Ecology, University of Georgia, 140 E. Green St., Athens, GA 30602, USA \\
-13: Lilongwe University of Agriculture and Natural Resources (LUANAR), Lilongwe, Malawi \\
-14: Forest Research Center, 1 Orange Groove Highlands, Harare, Zimbabwe \\
-15: Faculty of Natural Resources and Spatial Sciences, Namibia University of Science and Technology, Windhoek, Namibia \\
-16: Department of Forest Engineering, Faculty of Agronomy and Forest Engineering, UEM Av. Julius Nyerere, 3453, Campus Universitario, Maputo, Mozambique \\
-17: Forestry Department Headquarters - MLNR, Cairo Road, Lusaka, Zambia \\
-18: School of Natural Resources, Copperbelt University, Kitwe, Zambia \\
-19: Faculty of Biological Sciences, University of Leeds, LS2 9JT, United Kingdom\\
-20: Copernicus Institute of Sustainable Development, Utrecht University, 3508 TC Utrecht, The Netherlands \\
-21: Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, 6031, South Africa \\
-22: South African Environmental Observation Network, Grasslands-Forests-Wetlands Node, Montrose, 3201, South Africa \\
-23: Royal Botanic Garden Edinburgh, Edinburgh, EH3 5LR, United Kingdom \\
-
-\vspace{1em}
-Corresponding author:
-
-John L. Godlee
-
-johngodlee@gmail.com
-
-School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom
-
-\section{Acknowledgements}
-
-This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (John L.
-Godlee, Grant No.
-NE/L002558/1).
-The data for this study was contributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No.
-NE/P008755/1).
-Revisions of the SEOSAW dataset were funded by SavannaChange, a GCRF/University of Edinburgh funded project.
-We thank all data providers and the field assistance they received when collecting plot data.
-JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).
-
-\section{Conflict of interest statement}
-
-All authors declare no conflict of interest regarding this study.
-
-\section{Biosketch}
-
-SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}) aims to understand the response of southern African woodlands to global change.
-The goal of SEOSAW is to produce novel analyses of the determinants of ecosystem structure and function for the southern Africa region, based on syntheses of plot data.
-Additionally the group hopes to develop infrastructure for a long-term regional plan for plot remeasurement in the southern African region.
-While working on a multitude of diverse projects in the dry tropics at large, all authors have a broad interest in community ecology and ecosystem assemblage in southern African woodlands.
-
-\newpage{}
-
 {\LARGE{\textbf{Blinded Main Text File}}}
 
-\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across southern African woodlands}
+\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}
 
@@
 
-\section{Abstract}
+\section*{Abstract}
 
 \textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear.
-It is posited that high levels of disturbance moderate the effects of competition in savannas, weakening the correlation between biodiversity and niche complementarity that drives ecosystem function.
-Here, we explore the relationship between tree species diversity and aboveground biomass across southern African savannas and woodlands, while controlling for variation in stem density, resource availability, disturbance through fire, and vegetation types to build a general understanding of the biodiversity - ecosystem function relationship in this understudied ecological context.
+It is posited that disturbance moderates the effects of competition in savannas, weakening the correlation between biodiversity and niche complementarity that drives ecosystem function.
+We explore the relationship between tree species diversity and biomass across southern African woodlands, accounting for variation in stem density, resource availability, disturbance, and vegetation type to build general understanding of the biodiversity-ecosystem function relationship in this understudied ecological context.
 
 \textbf{Location:} Southern African savannas and woodlands
+
 \textbf{Time period:} 2010-2019
@@
 
-\textbf{Methods:} We used a network of \nplots{} savanna/woodland tree plots located across the southern African sub-continent.
-We used Structural Equation Modelling with path analysis to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.
+\textbf{Methods:} We used a network of \nplots{} woodland tree plots located across southern Africa.
+We used Structural Equation Modelling to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.
 
-\textbf{Results:} We found a positive effect of tree species diversity on aboveground biomass, across vegetation types, observed mainly via the increasing effect of woodland structural diversity.
+\textbf{Results:} We found positive effects of tree species diversity on aboveground biomass, across vegetation types, observed mainly indirectly via the increasing effect of woodland structural diversity.
 We also found that the effect of tree species diversity on biomass increases with stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}.
-Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via its effect on species diversity, suggesting that tree diversity may have been under-appreciated as a determinant of savanna structure.
+Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via species diversity, suggesting that tree diversity may have been under-appreciated as a determinant of savanna structure.
 
@@
 
+\section{Data accessibility statement}
+
+The data that support the findings of this study are available on request from the corresponding author.
+The data are not publicly available at the time of submission due to privacy restrictions regarding plot locations.
+An anonymised version will be made available in a data repository following review.
+
 \end{document}

### 2020-09-17 manuscript/title_page/title_page.tex
Submitted

\documentclass[11pt,a4paper]{article}

% Define page geometry
\usepackage{geometry}
\geometry{left=2.2cm,
	right=2.2cm,
	top=2.2cm,
	bottom=2cm}
\parskip 0.15cm
\setlength{\parindent}{0cm}
\usepackage{pdflscape}
\usepackage[document]{ragged2e}

% Set font
\usepackage[T1]{fontenc}

% Image handling
\usepackage{graphics}  % Insert images easily
\usepackage{graphicx}  % Extended image support

\makeatletter
	\g@addto@macro\@floatboxreset\centering  % Automatically centre images (floats)
\makeatother

\graphicspath{ {img/} }
\usepackage{float}  %  Graphics placement [H] [H!] arguments
\usepackage{subfig}  % Compound figures

% Tables
\usepackage{multirow}
\usepackage{longtable}

% Bibliography management
\usepackage[UKenglish]{babel}
\usepackage[natbibapa]{apacite}
\bibliographystyle{apacite}

% Text formatting
\usepackage{url} % Allow nice formatting of URLs in text

\usepackage{enumerate}  % Enumerated lists

\usepackage{lineno}  % Line numbers

\usepackage{textcomp}
\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}}  % Command for a good tilde

\usepackage{siunitx}
\usepackage{amsmath}

\usepackage{xcolor}
\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}   % \todo{NOTE TO SELF WRITTEN IN RED}

\input{code_format}

% Custom title formatting
\let\oldtitle\title

\renewcommand{\title}[1]{\oldtitle{\vspace{-1.5cm}#1}}

\usepackage[breaklinks]{hyperref}
\definecolor{links}{RGB}{191,59,72}
\hypersetup{
	breaklinks,
	colorlinks,
	allcolors=links,
	linktoc=section,
	pdfauthor={John L. Godlee}
}
\def\subsectionautorefname{section}
\def\subsubsectionautorefname{section}
% \usepackage{biblatex}

\usepackage{dcolumn}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{calc,arrows,positioning,shapes,shapes.gates.logic.US,trees}
\usepackage{pdflscape}
\usepackage{longtable}
\usepackage{fmtcount}

\usepackage{lineno}
\linenumbers

\begin{document}

{\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}}

\vspace{1cm}

Authors: John L. Godlee\textsuperscript{1}, Casey M. Ryan\textsuperscript{1}, David Bauman\textsuperscript{2,3}, Samuel J. Bowers\textsuperscript{1}, Joao M. B. Carreiras\textsuperscript{4}, Antonio Valter Chisingui\textsuperscript{5}, Joris P. G. M. Cromsigt\textsuperscript{6,7,8}, Dave J. Druce\textsuperscript{9,10}, Manfred Finkch\textsuperscript{11}, Francisco Maiato Gon\c{c}alves\textsuperscript{5}, Ricardo M. Holdo\textsuperscript{12}, Steve Makungwa\textsuperscript{13}, Iain M. McNicol\textsuperscript{1}, Edward T. A. Mitchard\textsuperscript{1}, Anderson Muchawona\textsuperscript{14}, Rasmus Revermann\textsuperscript{11,15}, Natasha Sofia Ribeiro\textsuperscript{16}, Abel Siampale\textsuperscript{17}, Stephen Syampungani\textsuperscript{18}, Jos\'{e} Jo\~{a}o Tchamba\textsuperscript{5}, Hemant G. Tripathi\textsuperscript{19}, Johannes Wallenfang\textsuperscript{11}, Mariska te Beest\textsuperscript{20,21,22}, Mathew Williams\textsuperscript{1}, Kyle G. Dexter\textsuperscript{1,23}

1: School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom \\
2: Environmental Change Institute, School of Geography and the Environment, University of Oxford, Oxford, United Kingdom \\
3: Laboratoire d'\'{E}cologie V\'{e}g\'{e}tale et Biogéochimie, CP 244, Universit\'{e} Libre de Bruxelles, Brussels, Belgium \\
4: National Centre for Earth Observation (NCEO), University of Sheffield, Hicks Building, Hounsfield Road, Sheffield, S3 7RH, United Kingdom \\
5: Herbarium of Lubango, ISCED Hu\'{i}la, Sarmento Rodrigues Str. No. 2, CP. 230, Lubango, Angola \\
6: Department of Wildlife, Fish, and Environmental Studies, Swedish University of Agricultural Sciences, Ume\aa, Sweden \\
7: Department of Zoology, Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, South Africa \\
8: Copernicus Institute of Sustainable Development, Utrecht University, 3584CS Utrecht, the Netherlands \\
9: Ecological Advice, Ezemvelo KZN Wildlife, Hluhluwe-iMfolozi Park, South Africa \\
10: School of Life Sciences, University of KwaZulu-Natal, Pietermaritzburg, South Africa \\
11: Biodiversity, Evolution and Ecology of Plants, Institute of Plant Sciences and Microbiology, University of Hamburg, Ohnhorststr. 18, 22609 Hamburg, Germany \\
12: Odum School of Ecology, University of Georgia, 140 E. Green St., Athens, GA 30602, USA \\
13: Lilongwe University of Agriculture and Natural Resources (LUANAR), Lilongwe, Malawi \\
14: Forest Research Centre, 1 Orange Groove Highlands, Harare, Zimbabwe \\
15: Faculty of Natural Resources and Spatial Sciences, Namibia University of Science and Technology, Windhoek, Namibia \\
16: Department of Forest Engineering, Faculty of Agronomy and Forest Engineering, Universidade Eduardo Mondlane, Av. Julius Nyerere, 3453, Campus Universitario, Maputo, Mozambique \\
17: Forestry Department Headquarters - Ministry of Lands and Natural Resources, Cairo Road, Lusaka, Zambia \\
18: School of Natural Resources, Copperbelt University, Kitwe, Zambia \\
19: Faculty of Biological Sciences, University of Leeds, LS2 9JT, United Kingdom\\
20: Copernicus Institute of Sustainable Development, Utrecht University, 3508 TC Utrecht, The Netherlands \\
21: Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, 6031, South Africa \\
22: South African Environmental Observation Network, Grasslands-Forests-Wetlands Node, Montrose, 3201, South Africa \\
23: Royal Botanic Garden Edinburgh, Edinburgh, EH3 5LR, United Kingdom \\

\vspace{1em}
Corresponding author:

John L. Godlee

johngodlee@gmail.com

School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom

\section{Acknowledgements}

This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (John L. Godlee, Grant No. NE/L002558/1). The data for this study was contributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No. NE/P008755/1). Revisions of the SEOSAW dataset were funded by SavannaChange, a GCRF/University of Edinburgh funded project. We thank all data providers and the field assistance they received when collecting plot data. JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).

\section{Biosketch}

SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}) aims to understand the response of southern African woodlands to global change. The goal of SEOSAW is to produce novel analyses of the determinants of ecosystem structure and function for the southern Africa region, based on syntheses of plot data. Additionally the group hopes to develop infrastructure for a long-term regional plan for plot remeasurement in the southern African region. While working on a multitude of diverse projects in the dry tropics at large, all authors have a broad interest in community ecology and ecosystem assemblage in southern African woodlands.

\end{document}

### 2020-10-21 manuscript/regional_befr_manuscript.tex
Updated tex

@@
 
-\title{Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}
+\newcommand{\titletext}{Stem density drives biodiversity-ecosystem function across disturbed southern African woodlands}
+
+\title{\titletext{}}
 \author{}
@@
 
-\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}
+\LARGE{Title: \titletext{}}
 
@@
 
-\textbf{Aim:} Positive correlations between tree species diversity and ecosystem function have been widely documented, but the nature of the relationship in southern African savanna/woodlands, which experience high levels of disturbance through fire and ecophysiological stress, is less clear.
-It is posited that disturbance moderates the effects of competition in savannas, weakening the correlation between biodiversity and niche complementarity that drives ecosystem function.
-We explore the relationship between tree species diversity and biomass across southern African woodlands, accounting for variation in stem density, resource availability, disturbance, and vegetation type to build general understanding of the biodiversity-ecosystem function relationship in this understudied ecological context.
-
-\textbf{Location:} Southern African savannas and woodlands
-
-\textbf{Time period:} 2010-2019
-
-\textbf{Major taxa studied:} Trees
-
-\textbf{Methods:} We used a network of \nplots{} woodland tree plots located across southern Africa.
-We used Structural Equation Modelling to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.
-
-\textbf{Results:} We found positive effects of tree species diversity on aboveground biomass, across vegetation types, observed mainly indirectly via the increasing effect of woodland structural diversity.
-We also found that the effect of tree species diversity on biomass increases with stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}.
-Finally, we found that resource availability affects biomass in southern African woodlands mainly indirectly, via species diversity, suggesting that tree diversity may have been under-appreciated as a determinant of savanna structure.
-
-\textbf{Main conclusions:} The study underlines the close association between tree diversity and ecosystem structure and function of highly disturbed southern African savannas and woodlands.
-Our results demonstrate the importance of accounting for environmental conditions and vegetation type in order to accurately model a general relationship between biodiversity and ecosystem function at a regional level.
+\begin{itemize}
+	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but whether BEFRs should be expected in disturbance-driven systems is less clear.
+Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs.
+We explore the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
+	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.}
+	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity.
+The observed BEFR was highly dependent on stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}.
+We also found that resource availability mainly affects biomass indirectly, via increasing species diversity.}
+	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed southern African savannas and woodlands.
+We suggest that tree diversity is an under-appreciated determinant of savanna and woodland structure and function.}
+\end{itemize}
 
@@
 % BEFR has had previous global research
-In order to understand the effects of global biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}.
-The strength and direction of the Biodiversity-Ecosystem Function (BEF) relationship varies depending on the ecosystem being studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016, Hooper2012, Cardinale2009}.
-Over the past decade, many observational studies of the BEF relationship have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}, which follow from early small-scale experimental studies conducted predominantly in artificial grassland mesocosms \citep{Tilman1994, Tilman2014}.
-Despite these concerted efforts, we continue to lack a nuanced, ecosystem-agnostic, understanding of the complex interactions between biodiversity, environment, and ecosystem function.
+In order to understand the interacting effects of global environmental and biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}.
+The strength and direction of the Biodiversity-Ecosystem Function relationship (BEFR) varies depending on the ecosystem studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016, Hooper2012, Cardinale2009}.
+Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}, which follow from early small-scale experimental studies conducted predominantly in temperate grasslands \citep{Tilman1994, Tilman2014}.
+Despite these concerted efforts, we continue to lack a nuanced, ecosystem-agnostic, understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.
 
@@
 % In this study :::
-In this study, we made the first known regional estimation of the biodiversity-ecosystem function relationship across southern African savannas and woodlands, using inventory plots spanning broad environmental and biogeographical gradients (\autoref{clust_map}).
-We used aboveground woody biomass of trees as our metric of ecosystem function, and compared the relative effects of tree species diversity with that of environmental factors known to affect ecosystem productivity and biomass accumulation, namely water availability and soil fertility.
-We also investigated the potential moderating effects of environmental covariates on the relationship between tree species diversity and biomass.
-We incorporated vegetation type (via clustering of plot-level tree species composition), as a factor in our analyses to understand how tree species composition as well as diversity affected ecosystem function and to assess the generality of our results.
-We used Structural Equation Modelling (SEM) with path analysis to simultaneously account for environmental and biotic factors, which may have interacting effects on ecosystem structure and therefore biomass.
-Initially, we posited three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with plot-level stem density (number of stems ha\textsuperscript{-1}), due to an increased importance of niche complementarity as stem density and therefore competition increases.
+In this study, we make the first known estimation of the biodiversity-ecosystem function relationship (BEFR) across a sub-continental area of disturbance-driven ecosystems, namely southern African savannas and woodlands.
+We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to seeking a general model of the biodiversity-ecosystem function relationship in wooded ecosystems.
+We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), due to an increased importance of niche complementarity as stem density and therefore competition increases.
 In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass.
@@
 % Re-cap
-In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots.
-Using Structural Equation Modelling (SEM), we found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}).
+In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM).
+We found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}).
 Tree species diversity, structural diversity and stem density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}).
 We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} below which the effect of species diversity on AGB remained at a low baseline level.
-The strongest effect on AGB was that of stem density.
+The strongest direct effect on AGB was that of stem density.
 When the effects of water availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.
@@
 In this study we found that even in highly disturbed southern African woodlands, there exists a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB).
-Additionally, we found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees.
-We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, suggesting that similar ecosystem processes occur across the region to determine ecosystem function.
-In contrast to previous studies, we found, at the scale of our study region, that the direct effects of water availability and soil fertility on woody biomass were small, with most of their effect being indirectly through their effects on tree species and structural diversity.
-This strongly suggests that data on tree species diversity be included into models predicting ecosystem functionality in this region.
-We also advocate for explicit inclusion of environmental covariates in regional scale models of biodiversity and ecosystem function, generally.
-
-The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in favour of disturbance effects.
-On a practical note, our study shows that biodiversity change through human actions will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.
+Our findings contribute to our understanding of a universal biodiversity-ecosystem function relationship, one which is moderated in a predictable manner by environmental covariates and their interaction with biodiversity and ecosystem structure.
+We found that the multiple vegetation types which comprise southern African woodlands exhibit similarities in the relationship between species diversity and woody biomass, suggesting that similar  processes operate across the region to determine ecosystem function.
+We advocate for explicit inclusion of environmental covariates in regional scale models of biodiversity and ecosystem function.
+We assert that this is necessary to develop our understanding of the biodiversity-ecosystem function relationship in real-world ecosystems, to progress from experimental mesocosms.
+We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key vector through which tree species diversity determines ecosystem function in savannas, woodlands and forests, i.e.
+all wooded ecosystems.
+The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in previous studies despite its intuitive logic as a determinant of niche complementarity effects in wooded ecosystems.
+Our study shows that biodiversity change through extensive human actions in this region will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.
 This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.

### 2021-01-25 manuscript/include/clust_summ.tex
Added submission

@@
 \hline \\[-1.8ex] 
-{Cluster} & {Dominant species} & {Indicator species} & {N plots} & \multicolumn{1}{p{2cm}}{\centering Species\\Richness} & \multicolumn{1}{p{2.5cm}}{\centering Stem density\\(stems ha\textsuperscript{-1})} & \multicolumn{1}{p{2cm}}{\centering AGB\\(t ha\textsuperscript{-1})} \\
+{Cluster} & {Dominant species} & {Indicator species} & {n} & \multicolumn{1}{p{2cm}}{\centering Species\\Richness} & \multicolumn{1}{p{2.5cm}}{\centering Stem density\\(stems ha\textsuperscript{-1})} & \multicolumn{1}{p{2cm}}{\centering AGB\\(t ha\textsuperscript{-1})} \\
 \hline \\[-1.8ex] 

### 2021-01-25 manuscript/include/struc_model_fit_clust_stats.tex
Added submission

@@
   	\caption{Model fit statistics for SEMs investigating the effects of tree diversity and stem density on AGB (\autoref{struc_mod}).
-n = number of plots in cluster, $\chi^{2}$ = Chi-squared fit statistic, DoF = model degrees of freedom, CFI = Comparative Fit Index, TLI = Tucker-Lewis Index, RMSEA = Root Mean Square Error of Approximation, $R^{2}$ AGB = R-squared of AGB.}
+n = number of plots in cluster, $\chi^{2}$ = Chi-squared fit statistic, DoF = model degrees of freedom, CFI = Comparative Fit Index, TLI = Tucker-Lewis Index, RMSEA = Root Mean Square Error of Approximation, $R^{2}$ AGB = R-squared of AGB (Above-Ground Biomass).}
   \label{struc_model_fit_clust_stats} 
@@
 \hline \\[-1.8ex] 
-Sparse miombo / Baikiaea & 466 & 43.870 & 6 & 0.914 & 0.784 & 0.130 & 0.580 \\ 
 Core miombo & 523 & 78.670 & 6 & 0.904 & 0.759 & 0.140 & 0.490 \\ 
@@
 Mopane & 58 & 19.880 & 6 & 0.834 & 0.584 & 0.240 & 0.510 \\ 
+Sparse miombo / Baikiaea & 466 & 43.870 & 6 & 0.914 & 0.784 & 0.130 & 0.580 \\ 
 All & 1235 & 91.380 & 6 & 0.937 & 0.843 & 0.120 & 0.490 \\ 

### 2021-01-25 manuscript/regional_befr_manuscript.tex
Added submission

@@
 \usepackage[T1]{fontenc}
+
+\usepackage{framed}
 
@@
 
-% Custom title formatting
-\let\oldtitle\title
-
-\renewcommand{\title}[1]{\oldtitle{\vspace{-1.5cm}#1}}
-
 \usepackage[breaklinks]{hyperref}
-\definecolor{links}{RGB}{191,59,72}
+\definecolor{links}{RGB}{0,0,0}
 \hypersetup{
 	breaklinks,
-	colorlinks,
-	allcolors=links,
-	linktoc=section,
+	colorlinks=true,
+	linkcolor=links,
+	anchorcolor=links,
+	citecolor=links,
+	filecolor=links,
+	menucolor=links,
+	runcolor=links,
+	urlcolor=links,
 	pdfauthor={John L. Godlee}
@@
 
-\usepackage{lineno}
 \linenumbers
+\linespread{1.25}
 
-\newcommand{\titletext}{Stem density drives biodiversity-ecosystem function across disturbed southern African woodlands}
-
-\title{\titletext{}}
-\author{}
-\date{}
+\newcommand{\titletext}{Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}
 
@@
 
-{\LARGE{\textbf{Blinded Main Text File}}}
+{\Large{Title: \titletext{}}}
 
-\LARGE{Title: \titletext{}}
-
-\normalsize{Running title: Diversity - ecosystem function in southern African woodlands}
-
-\section*{Abstract}
+\section*{Summary}
 
 \begin{itemize}
-	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but whether BEFRs should be expected in disturbance-driven systems is less clear.
+	\item{Positive biodiversity-ecosystem function relationships (BEFRs) have been widely documented, but it is unclear if BEFRs should be expected in disturbance-driven systems.
 Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs.
 We explore the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
-	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, while accounting for the interactive effects of resource availability, disturbance by fire, stem density and vegetation type.}
+	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interactive effects of resource availability, disturbance by fire, stem density and vegetation type.}
 	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity.
-The observed BEFR was highly dependent on stem density, with a minimum threshold of \textapprox{}180 stems ha\textsuperscript{-1}.
-We also found that resource availability mainly affects biomass indirectly, via increasing species diversity.}
-	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed southern African savannas and woodlands.
+The observed BEFR was highly dependent on stem density, with a minimum threshold of c.
+180 stems ha\textsuperscript{-1}.
+We found that resource availability mainly affects biomass indirectly, via increasing species diversity.}
+	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands.
 We suggest that tree diversity is an under-appreciated determinant of savanna and woodland structure and function.}
@@
 
-\textbf{Keywords:} biodiversity, ecosystem function, woodland, miombo, biomass, structural equation modelling, forest structure.
+\textbf{Keywords:} biodiversity, biomass, ecosystem function, forest structure, miombo, savanna, structural equation modelling, woodland.
 
@@
 This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resilience of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported BEF relationship invokes three main mechanisms to explain it \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
+The frequently reported BEFR invokes three main mechanisms to explain it \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
 
@@
 These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses.
-Southern African savannas and woodlands are highly diverse, thought to harbour \textapprox{}8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
+Southern African savannas and woodlands are highly diverse, thought to harbour c.
+8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
 Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under international efforts to reduced deforestation and degradation (REDD+) \citep{Hinsley2015}.
@@
 Disturbance reduces stem density and woody biomass, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
-This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEF relationship.
+This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEFR.
 Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}.
@@
 They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects.
-This potential mismatch in the contribution of different mechanisms to the BEF relationship between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEF relationship.
+This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEFR.
 
 % African savannas are understudied in this regard
-The representation of dry tropical ecosystems in the BEF relationship literature is poor compared to other ecosystems.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEF relationship studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}.
-A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEF relationship mechanisms described above, suggests that a focus on the BEF in southern African woodlands could greatly strengthen our understanding of a global BEF relationship and its environmental determinants.
+The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
+\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}.
+A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEFR mechanisms described above, suggests that a focus on the BEFR in southern African woodlands could greatly strengthen our understanding of a global BEFR and its environmental determinants.
 A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
 The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}.
-Studies of the BEF relationship often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>10000s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
+Studies of the BEFR often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>10000s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
 
@@
 % Disturbance deserves another look
-High levels of disturbance in southern African woodlands may moderate the observable BEF relationship through its effect on ecosystem composition.
+High levels of disturbance in southern African woodlands may moderate the observable BEFR through its effect on ecosystem composition.
 Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}.
@@
 
-\section{Materials and methods}
+\section{Materials and Methods}
 
@@
 % Situation
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}; \citealp{White1987}).
-The study area spans the core climate space of the region, with a precipitation gradient from \textapprox{}460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to \textapprox{}1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
+The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}).
+The study area spans the core climate space of the region, with a precipitation gradient from c.
+460 mm y\textsuperscript{-1} in southern Mozambique and southern Zimbabwe to c.
+1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
 A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.
@@
 Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup.
-Plot vegetation was identified under the broad term of ``savanna'', which includes ``woodland'', ``savanna woodland'', and ``tree savanna'', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
+Plot vegetation was identified under the broad term of `savannaj', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
 Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
-Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than ``grassy savanna'', which is considered a separate biome with very different species composition \citep{Parr2014}.
+Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered a separate biome with very different species composition \citep{Parr2014}.
 
@@
 These data are a modelled product derived from various remotely sensed and directly measured data sources.
-The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1; \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018.
+The degree of fire disturbance was calculated using the MODIS monthly burned area product at 500 m resolution (MCD64A1, \citealt{MODIS_burn}), counting the total number of times the plot pixel was classified as burning, between 2001 and 2018.
 We initially aimed to include disturbance by herbivory in our model, including total herbivore biomass from the \citet{Hempson2017} modelled herbivory product, but this inclusion prevented models from converging due to its collinearity with other observed variables, notably MAP and disturbance by fire.
@@
 Abundance evenness allows for greater niche complementarity at small scales due to potentially increased heterogeneity of functional traits.
-We quantified tree structural diversity for each plot by calculating the coefficient of variation of DBH (DBH CoV) and tree height (Height CoV).
+We quantified tree structural diversity for each plot by calculating the Coefficient of Variation of DBH (DBH CoV) and tree height (Height CoV).
 
@@
 A key aspect to our decision to use SEM is that they can explicitly model and partition variance attributed to indirect effects, which is challenging in standard multiple regressions.
-Using SEMs also allowed us to describe latent variables such as ``water availability'', ``soil fertility'', and ``disturbance'' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having directly observable measures in our dataset.
+Using SEMs also allowed us to describe latent variables such as `water availability', `soil fertility', and `disturbance' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having directly observable measures in our dataset.
 SEM is also necessary to properly account for potential feedback mechanisms between aspects of environment and tree species diversity, which could otherwise increase the chances of Type I error and wrongly attribute inference due to the covariance of explanatory variables when using conventional regression analyses \citep{Nachtigall2003}.
 
-Prior to analysis, we specified a conceptual model with factors expected to affect AGB: water availability, soil fertility, disturbance, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).
+We specified a conceptual model with factors expected to affect AGB: water availability, soil fertility, disturbance, tree species diversity, tree structural diversity and stem density (\autoref{con_mod}).
 
-Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (\hyperref[appendixa]{Appendix A}).
+Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (Fig.
+S1, Fig.
+S2).
 Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
@@
 Path coefficients show the effect of a given path with other paths held constant.
-Models were estimated using the ``MLM'' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}.
+Models were estimated using the `MLM' estimator, because it is robust to multivariate non-normality \citep{Shapiro1983}.
 Model fit was evaluated using the robust Comparative Fit Index (CFI), the robust Tucker Lewis Index (TLI), the Root Mean Squared Error of Approximation (RMSEA) and the R\textsuperscript{2} coefficient of determination for AGB.
@@
 % Correlations
-Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB.
+Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig.
+S3).
 Stem density had the strongest correlation with AGB of all variables considered (\ccib{}).
@@
 In our sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}).
-There appears to be a minimum stem density threshold at \textapprox{}180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass.
+There appears to be a minimum stem density threshold at c.
+180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass.
 The effect of structural diversity on AGB appears to remain constant with increasing stem density.
@@
 
-A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, \hyperref[appendixd]{Appendix D}).
+A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, Fig.
+S4).
 The direct effects of water availability, soil fertility and disturbance on AGB were negligible (water: \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effect on AGB coming from the indirect paths via stem density (water: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (water: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}).
@@
 We found a non-linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
-At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below \textapprox{}180 trees ha\textsuperscript{-1}.
+At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below c.
+180 trees ha\textsuperscript{-1}.
 At very high stem density, there is also an increase in the effect of species diversity on structural diversity.
@@
 
-\subsection{Conclusion}
-
 In this study we found that even in highly disturbed southern African woodlands, there exists a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB).
@@
 
-\bibliography{regional_befr_manuscript.bib}
-
 \newpage{}
-
-\section{Author contribution statement}
-
-JG and KD conceived the study.
-JG conducted data analysis, data management for further versions of the SEOSAW dataset, and wrote the manuscript.
-CR conceived the SEOSAW database and conducted data management for earlier versions of the SEOSAW dataset.
-JG, CR, DB, JMBC, MF, RH, EM, SS, HT, HT, MB, MW, and KD contributed to manuscript revisions.
-JG, CR, SB, VC, JPGMC, DD, MF, FG, SM, IM, AM, RR, NR, AS, SS, JT, JW, MB, and MW contributed to experimental design, field data collection, data preparation and data management of parts of the dataset used in this study.
 
@@
 \begin{landscape}
-\section{Figure legends and embedded figures}
+\section{Figures}
 \begin{figure}[H]
@@
 \end{landscape}
+\newpage{}
 
@@
 \centering
-\includegraphics[width=\textwidth]{concept}
+\includegraphics[width=\textwidth]{concept.drawio}
 	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB.
@@
 \end{figure}
+\newpage{}
 
@@
 Colours of variable names group them into latent variables used in the SEMs: red = soil fertility, blue = disturbance, turquoise = water availability, green = tree species diversity, purple = tree structural diversity.
-See \hyperref[appendixb]{Appendix B} for a full assessment of correlation fit statistics.}
+See Table S1 for a full assessment of correlation fit statistics.}
 	\label{corr_mat}
 \end{figure}
+\newpage{}
 
@@
 \centering
-	\includegraphics[width=0.5\textwidth]{struc}
+	\includegraphics[width=0.5\textwidth]{struc.drawio}
 	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all vegetation clusters.
@@
 	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line plots showing the variation in path coefficients in the SEM, using datasets with different mean stem density. Smoothed lines are loess curves with standard error shaded bars.}
+	\caption{Line plots showing the variation in SEM path coefficients across datasets with different mean stem density. Smoothed lines are loess curves with $\pm1$ standard error shaded bars.}
 	\label{sem_struc_stems_ha}
 \end{figure}
+\newpage{}
 
@@
 \centering
-	\includegraphics[width=\textwidth]{full}
+	\includegraphics[width=\textwidth]{full.drawio}
 	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation types.
@@
 \end{figure}
+\newpage{}
 
-\section{Appendix 1 - Frequency distribution of observed variables} \label{appendixa}
+\section{Acknowledgements}
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{hist_raw}
-	\caption{Histograms of raw untransformed observed variables used in final analyses.}
-	\label{hist_raw}
-\end{figure}
+This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (John L.
+Godlee, Grant No.
+NE/L002558/1).
+The data for this study was contributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No.
+NE/P008755/1).
+Revisions of the SEOSAW dataset were funded by SavannaChange, a GCRF/University of Edinburgh funded project.
+We thank all data providers and the field assistance they received when collecting plot data.
+JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{hist_trans}
-	\caption{Histograms of observed variables transformed to achieve a normal frequency distribution.}
-	\label{hist_trans}
-\end{figure}
+\section{Author contribution}
 
-\appendix{}
-\section{Appendix 2 - Table of correlation fit statistics} \label{appendixb}
+JG and KD conceived the study.
+JG conducted data analysis, data management for further versions of the SEOSAW dataset, and wrote the manuscript.
+CR conceived the SEOSAW database and conducted data management for earlier versions of the SEOSAW dataset.
+JG, CR, DB, JMBC, MF, RH, EM, SS, HT, HT, MB, MW, and KD contributed to manuscript revisions.
+JG, CR, SB, VC, JPGMC, DD, MF, FG, SM, IM, AM, RR, NR, AS, SS, JT, JW, MB, and MW contributed to experimental design, field data collection, data preparation and data management of parts of the dataset used in this study.
 
-\input{include/corr_ci_tab.tex}
+\section{Data Availability}
 
-\section{Appendix 3 - Bivariate relationships of model variables} \label{appendixc}
+\bibliography{regional_befr_manuscript.bib}
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=0.9\textwidth]{bivar_lm}
-	\caption{Bivariate scatter plots for each observed variable used in the SEMs, based on hypothesised paths of causality.
-Points are coloured according to vegetation type.
-A single linear regression is presented as a black line, which combines all vegetation types, separate loess trend lines are fitted for each vegetation type.
-An orange loess trend line is fitted for all the data.
-All data is standardised and variables are transformed where it was appropriate for analysis.}
-	\label{bivar_lm}
-\end{figure}
+\vspace{0.5cm}
 
-\section{Appendix 4 - Path coefficients for model incorporating environmental covariates} \label{appendixd}
+\begin{minipage}{\linewidth}
+\begin{framed}
+\section*{Supporting Information}
 
-\begin{figure}[H]
-\centering
-	\includegraphics[width=\textwidth]{full_model_slopes}
-	\caption{Unstandardised path coefficients for the full model including tree species diversity, environmental covariates and stem density.
-Path coefficients are $\pm1$ standard error.
-Path coefficients where the interval (standard error) does not overlap zero are considered to be significant effects.}
-	\label{full_model_slopes}
-\end{figure}
+Additional supporting information may be found in the online version of this article.
+\vspace{0.2cm}
 
-\section{Data accessibility statement}
+\textbf{Fig. S1} Histograms of raw untransformed observed variables used in final analyses.
+\vspace{0.2cm}
 
-The data that support the findings of this study are available on request from the corresponding author.
-The data are not publicly available at the time of submission due to privacy restrictions regarding plot locations.
-An anonymised version will be made available in a data repository following review.
+\textbf{Fig. S2} Histograms of observed variables transformed to achieve a normal frequency distribution.
+\vspace{0.2cm}
+
+\textbf{Fig. S3} Bivariate scatter plots for each observed variable used in SEMs, based on hypothesised paths of causality.
+\vspace{0.2cm}
+
+\textbf{Fig. S4} Unstandardised path coefficients for full SEM model.
+\vspace{0.2cm}
+
+\textbf{Table S1} Table of correlation fit statistics for each pairwise Pearson correlation test of observed variables used in Structural Equation Models.
+\vspace{0.2cm}
+
+Please note: Wiley-Blackwell are not responsible for the content or functionality of any supporting information supplied by the authors.
+Any queries (other than missing material) should be directed to the New Phytologist Central Office.
+\end{framed}
+\end{minipage}
 

### 2021-01-25 manuscript/si/si.tex
Added submission

\section{Appendix 1 - Frequency distribution of observed variables} \label{appendixa}

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{hist_raw}
	\caption{Histograms of raw untransformed observed variables used in final analyses.}
	\label{hist_raw}
\end{figure}

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{hist_trans}
	\caption{Histograms of observed variables transformed to achieve a normal frequency distribution.}
	\label{hist_trans}
\end{figure}

\appendix{}
\section{Appendix 2 - Table of correlation fit statistics} \label{appendixb}

\input{include/corr_ci_tab.tex}

\section{Appendix 3 - Bivariate relationships of model variables} \label{appendixc}

\begin{figure}[H]
\centering
	\includegraphics[width=0.9\textwidth]{bivar_lm}
	\caption{Bivariate scatter plots for each observed variable used in the SEMs, based on hypothesised paths of causality. Points are coloured according to vegetation type. A single linear regression is presented as a black line, which combines all vegetation types, separate loess trend lines are fitted for each vegetation type. An orange loess trend line is fitted for all the data. All data is standardised and variables are transformed where it was appropriate for analysis.}
	\label{bivar_lm}
\end{figure}

\section{Appendix 4 - Path coefficients for model incorporating environmental covariates} \label{appendixd}

\begin{figure}[H]
\centering
	\includegraphics[width=\textwidth]{full_model_slopes}
	\caption{Unstandardised path coefficients for the full model including tree species diversity, environmental covariates and stem density. Path coefficients are $\pm1$ standard error. Path coefficients where the interval (standard error) does not overlap zero are considered to be significant effects.}
	\label{full_model_slopes}
\end{figure}

### 2021-01-25 manuscript/title_page/title_page.tex
Added submission

@@
 
-\usepackage{lineno}  % Line numbers
+%\usepackage{lineno}  % Line numbers
 
@@
 
-\usepackage{lineno}
-\linenumbers
-
 \begin{document}
 
-{\LARGE{Title: Tree species and structural diversity are important determinants of ecosystem function across disturbed southern African woodlands}}
+{\textbf{Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}}
 
@@
 
-Authors: John L.
+John L.
 Godlee\textsuperscript{1}, Casey M.
@@
 
-\vspace{1em}
-Corresponding author:
+\newpage{}
 
-John L. Godlee
+\textbf{Author for correspondence:}
 
-johngodlee@gmail.com
+John L. Godlee \\
+Telephone: +44(0) 7397 978 688 \\
+Email: johngodlee@gmail.com \\
 
-School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom
+\vspace{1cm}
 
-\section{Acknowledgements}
+\textbf{Word count:}
 
-This work is funded by a NERC E3 Doctoral Training Partnership PhD studentship at the University of Edinburgh (John L.
-Godlee, Grant No.
-NE/L002558/1).
-The data for this study was contributed by a number of independently funded projects and was assembled and prepared by SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}), an activity of the Miombo Network and a NERC-funded project (Grant No.
-NE/P008755/1).
-Revisions of the SEOSAW dataset were funded by SavannaChange, a GCRF/University of Edinburgh funded project.
-We thank all data providers and the field assistance they received when collecting plot data.
-JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).
+\vspace{0.2cm}
 
-\section{Biosketch}
+\begin{tabular}{r|r}
+Main body word count & 6384\\
+Summary & 193\\
+Introduction & 1490\\
+Materials and Methods & 1999\\
+Results & 726\\
+Discussion & 2216\\
+Acknowledgements & 122\\
+\end{tabular}
 
-SEOSAW (A Socio-Ecological Observatory for Southern African Woodlands, \url{https://seosaw.github.io}) aims to understand the response of southern African woodlands to global change.
-The goal of SEOSAW is to produce novel analyses of the determinants of ecosystem structure and function for the southern Africa region, based on syntheses of plot data.
-Additionally the group hopes to develop infrastructure for a long-term regional plan for plot remeasurement in the southern African region.
-While working on a multitude of diverse projects in the dry tropics at large, all authors have a broad interest in community ecology and ecosystem assemblage in southern African woodlands.
+\vspace{1cm}
+
+Number of figures: 7
+
+Number of tables: 2
+
+Colour figures: 1, 3, 5, 6
+
+Number of supporting information files: 5
 

### 2021-01-25 manuscript/regional_befr_manuscript.tex
Added TRY species matching

@@
 \citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}.
-A severe lack of study in dry tropical ecosystems, especially given the potential divergence in BEFR mechanisms described above, suggests that a focus on the BEFR in southern African woodlands could greatly strengthen our understanding of a global BEFR and its environmental determinants.
+A severe lack of studies in dry tropical ecosystems, especially given the potential divergence in BEFR mechanisms described above, suggests that a focus on the BEFR in southern African woodlands could greatly strengthen our understanding of a global BEFR and its environmental determinants.
 A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
@@
 Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup.
-Plot vegetation was identified under the broad term of `savannaj', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
+Plot vegetation was identified under the broad term of `savannah', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
@@
 The effect of structural diversity on AGB appears to remain constant with increasing stem density.
-The indirect effect of tree species diversity on AGB via structural diversity climbs as stem density increases.
+The indirect effect of tree species diversity on AGB via structural diversity increases as stem density increases.
 
@@
 S4).
-The direct effects of water availability, soil fertility and disturbance on AGB were negligible (water: \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effect on AGB coming from the indirect paths via stem density (water: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (water: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}).
+The direct effects of water availability, soil fertility and disturbance on AGB were negligible (water: \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effects on AGB coming from the indirect paths via stem density (water: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (water: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}).
 MAP and soil sand content had the greatest contributions to the latent variable of water availability.
@@
 % 
-We found a consistent positive effect of tree species diversity on AGB across all models in the current study.
-Within southern African woodlands we therefore find support that higher tree species richness and evenness leads to higher woody AGB.
+We found a consistent positive effect of tree species diversity on AGB.
+Within southern African woodlands we therefore find support for the hypothesis that higher tree species richness and evenness leads to higher above-ground woody biomass.
 This finding is in agreement with many other studies across different ecosystems and biomes, supporting the idea that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
@@
 
+\section{Conclusions}
+
 In this study we found that even in highly disturbed southern African woodlands, there exists a generalisable positive association between tree species diversity and ecosystem function, quantified as above-ground woody biomass (AGB).
@@
 	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line plots showing the variation in SEM path coefficients across datasets with different mean stem density. Smoothed lines are loess curves with $\pm1$ standard error shaded bars.}
+	\caption{Line plots showing the variation in SEM path coefficients across datasets with different mean stem density.
+Smoothed lines are loess curves with $\pm1$ standard error shaded bars.
+AGB = Above-Ground woody Biomass, arrows in plot titles indicate causal paths in SEM models.}
 	\label{sem_struc_stems_ha}

### 2021-01-25 scripts/sem.R

@@ -228,11 +228,11 @@ mod_summ_mutate_struc <- function(x){
       op == "Direct: AGB" & rhs == "div" ~ "Species\ndiversity",
       op == "Direct: AGB" & rhs == "struc" ~ "Structural\ndiversity",
       op == "Direct: AGB" & rhs == "n_trees_gt10_ha_log_std" ~ "Stem density",
-      lhs == "struc" & op == "Other" & rhs == "div" ~ "Species diversity ->\nStructural diversity",
-      lhs == "stems_ha_log_std" & op == "Other" & rhs == "div" ~ "Species diversity ->\nStem density",
-      op == "Other" & rhs == "a*b" ~ "Indirect:\nSpecies diversity ->\nStructural diversity ->\nAGB",
-      op == "Other" & rhs == "d*e" ~ "Indirect:\nSpecies diversity ->\nStem density ->\nAGB",
-      op == "Other" & (rhs == "c+(a*b)+(d)" | rhs == "c+(a*b)+(d*e)" | rhs == "c+(a*b)")  ~ "Total effect:\nSpecies diversity ->\nAGB",
+      lhs == "struc" & op == "Other" & rhs == "div" ~ "Species diversity \u2192\nStructural diversity",
+      lhs == "stems_ha_log_std" & op == "Other" & rhs == "div" ~ "Species diversity \u2192\nStem density",
+      op == "Other" & rhs == "a*b" ~ "Indirect:\nSpecies diversity \u2192\nStructural diversity \u2192\nAGB",
+      op == "Other" & rhs == "d*e" ~ "Indirect:\nSpecies diversity \u2192\nStem density \u2192\nAGB",
+      op == "Other" & (rhs == "c+(a*b)+(d)" | rhs == "c+(a*b)+(d*e)" | rhs == "c+(a*b)")  ~ "Total effect:\nSpecies diversity \u2192\nAGB",
     TRUE ~ rhs))
 }
 
@@ -552,11 +552,11 @@ struc_sem_quant_fit_df$n_plots <- quant_plots[sapply(struc_sem_quant_fit, nrow) 
 struc_sem_quant_regs <- do.call(rbind, struc_sem_quant_regs_list) %>%
   dplyr::select(rhs, quant, stems_ha, n_plots, n_species_raref, est, std.all) %>%
   mutate(rhs = case_when(
[…]

### 2021-01-25 scripts/try_species_match.R

@@ -1,0 +1,24 @@
+# Which species are present in the TRY database?
+# John Godlee (johngodlee@gmail.com)
+# 2021-01-25
+
+# Import data
+try_sp <- read.table("data/try_species_list.txt", header = TRUE, sep = "\t")
+
+seosaw_stems <- read.csv("~/git_proj/seosaw_data/data_out/v2.10/stems_v2.10.csv")
+
+# Extract species from SEOSAW
+seosaw_sp <- unique(seosaw_stems$species_name_clean)
+
+# Match SEOSAW species with TRY database
+try_match <- try_sp[try_sp$AccSpeciesName %in% seosaw_sp,]
+
+# Which species aren't matched
+seosaw_no_match <- seosaw_sp[!seosaw_sp %in% try_sp$AccSpeciesName]
+
+# Check all species either matched or not matched
+stopifnot(length(seosaw_no_match) + nrow(try_match) == nrow(seosaw_sp))
+
+# Write matched species IDs to file for upload to TRY website
+writeLines(paste(try_match$AccSpeciesID, collapse = ","), "data/try_match.txt")
+

### 2021-01-28 manuscript/regional_befr_manuscript.tex
Added TRY DB exploration

@@
 
-% BEFR has had previous global research
-In order to understand the interacting effects of global environmental and biodiversity change, it is necessary to explore the relationship between biodiversity and ecosystem function \citep{Tilman2014}.
-The strength and direction of the Biodiversity-Ecosystem Function relationship (BEFR) varies depending on the ecosystem studied, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}, but there appears to be a generalisable positive correlation between biodiversity and ecosystem function \citep{Liang2016, Hooper2012, Cardinale2009}.
-Over the past decade, many observational studies of the BEFR have been conducted, mostly in wet tropical and temperate forests, and grasslands \citep{Chen2011}, which follow from early small-scale experimental studies conducted predominantly in temperate grasslands \citep{Tilman1994, Tilman2014}.
-Despite these concerted efforts, we continue to lack a nuanced, ecosystem-agnostic, understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.
+% BEFR is important to study, setting the scene - DONE
+Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{}.
+Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems.
+While positive BEFRs which align with theory have been frequently reported \citep{}, as research has expanded a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}.
+The goal now should be to study the BEFR in different environmental and ecological contexts, in order to develop an ecosystem-agnostic understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.
 
-% Ecosystem function definition and mechanisms
-Ecosystem functions can be defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}.
-This includes processes such as gross primary productivity and atmospheric nitrogen fixation, but can be extended to indirect measures of function such as resilience of productivity to disturbance, and further to ecosystem properties which themselves influence process, such as trophic complexity and total vegetative biomass.
-The frequently reported BEFR invokes three main mechanisms to explain it \citep{Tilman2014}: 1) niche complementarity, whereby communities with greater biodiversity fill a greater breadth of realised niche space and avoid competition due to differences in their resource acquisition strategies; 2) selection effects, whereby communities with greater biodiversity are more likely to include a species that contributes highly to the measured ecosystem function; and 3) facilitation effects, whereby communities with greater biodiversity are more likely to include combinations of species which together increase the others' functional contribution.
+% Ecosystem function definition and mechanisms of the BEFR - DONE
+Ecosystem functions are defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}.
+Woody productivity and biomass storage are two of the most commonly studied ecosystem functions in forested ecosystems \citep{Pacala2002}, due to their importance in the global carbon cycle and their potential leverage as a tool to mitigate the effects of anthropogenic climate change while maintaining biodiversity \citep{}.
+Theory predicts that biodiversity will have a positive effect on ecosystem function via three principle mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as intraspecific competition and abundance of natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{}.
 
-% Savannas are big and ecologically important
-Savannas and woodlands are the dominant vegetation type across southern Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
-The carbon stored in this vegetation is comparable to that found in the wet forests of the Congo basin, and is of global importance to the carbon cycle \citep{Houghton2009, Mayaux2008}.
-Climatic conditions and biogeography vary across southern African vegetation, resulting in a diverse range of savanna and woodland tree species assemblages.
-These retain the common features of an open tree canopy and an understorey generally dominated by C4 grasses.
-Southern African savannas and woodlands are highly diverse, thought to harbour c.
-8500 plant species of which >300 are trees \citep{Frost1996}, and have been identified by previous studies as a priority for conservation efforts \citep{Byers2001, Mittermeier2003}.
-Many conservation projects in the region currently aim to conserve biodiversity and woody biomass stocks simultaneously under international efforts to reduced deforestation and degradation (REDD+) \citep{Hinsley2015}.
-Despite these efforts however, human actions are driving rapid changes in biodiversity, with largely unquantified consequences for ecosystem structure and function.
+There has been extended debate on whether positive BEFRs should be expected in all environmental and ecological contexts \citep{}.
+In temperate forests, wet tropical forests and grasslands, where the majority of BEFR studies in natural systems have been conducted \citep{}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition is the limiting factor to ecosystem functioning \citep{}, but this may not hold true in all systems.
+In mesic savannas for instance, disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
+Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function in these ecosystems \citep{Lasky2014, Tobner2016}.
+It is possible that a threshold woody stem density exists below which the lack of competition precludes the detection of a biodiversity effect on ecosystem function.
+Additionally, facilitation effects may play a larger role in driving the BEFR in stressful environments like mesic savannas.
+\citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions with increasing aridity.
+They demonstrate that competition diminishes where environmental stress limits species' abundances.
+This potential mismatch in the contribution of different mechanisms to the BEFR among resource-, disturbance-, and competition-limited ecosystems requires further investigation if we are to derive a generalisable BEFR.
 
-% Savannas might not be the same as forests
-Compared to forest ecosystems, southern African dry tropical woodlands and savannas are highly structured by disturbance, through fire \citep{Lehmann2014}, herbivory \citep{Sankaran2008, Levick2009}, and human activities such as shifting cultivation agriculture \citep{Heinimann2017}, timber extraction and charcoal processing \citep{Dewees2010, McNicol2018b}.
-High levels of disturbance, by fire or otherwise, may weaken the role of competition in determining local species distribution.
-Disturbance reduces stem density and woody biomass, reducing competitive interactions between individuals, allowing weak competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
-This means that interspecific competition and therefore the effect of niche complementarity, which contributes the majority of the observed biodiversity effect on ecosystem function in temperate and wet tropical forests \citep{Wright2017, Poorter2015, Sande2017a}, may not be as important in dry woodland/savanna ecosystems, thus weakening the BEFR.
-Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function \citep{Lasky2014, Tobner2016}.
-A threshold stem density may exist below which the effects of tree species diversity on ecosystem function are not detectable, with potential consequences for our classification of ecosystems limited by biodiversity and those limited by other factors.
+Disturbance-driven mesic savannas and open canopy woodlands cover >20\% of the global land surface \citep{Pennington2018}, and represent the dominant vegetation type in southern Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
+Despite their global importance, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2018}.
+A number of studies, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+These studies however, lacked the breadth of sampling required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR.
 
-% Savannas might no be the same as forests continued: facilitation effects
-More diverse species assemblages may lead to facilitation effects between certain species combinations under the limiting environmental conditions prevalent across African savannas, such as low water availability.
-Across European forests \citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions in more arid environments.
-They suggest that in water-limited ecosystems, facilitative effects and selection effects may be more important than niche complementarity in driving the relationship between species diversity and ecosystem function, as competition diminishes in ecosystems where environmental stress limits individual species' abundances, thus reducing the competition which drives niche complementarity effects.
-This potential mismatch in the contribution of different mechanisms to the BEFR between dry tropical woodlands and other forested ecosystems demands further investigation if we are to derive a generalisable BEFR.
-
-% African savannas are understudied in this regard
-The representation of dry tropical ecosystems in the BEFR literature is poor compared to other ecosystems.
-\citet{Clarke2017} conducted a meta-analysis of 182 published BEFR studies, finding that only 13\% were conducted in the tropics generally, with 42\% of those being conducted in the wet tropical forests of Costa Rica, a narrow geographic region \citep{Barthlott2005}.
-A severe lack of studies in dry tropical ecosystems, especially given the potential divergence in BEFR mechanisms described above, suggests that a focus on the BEFR in southern African woodlands could greatly strengthen our understanding of a global BEFR and its environmental determinants.
-A small number of studies in southern African woodlands, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
-The results of these fine-scale studies concur with similar studies in other biomes \citep{Cardinale2009}.
-Studies of the BEFR often find that at fine spatial scales (<1 ha), biodiversity shows a strong effect on ecosystem function, but at broad spatial scales  (>10000s ha) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Pasari2013}.
-
-% \citep{Michaletz2014, Michaletz2018} \citep{Spasojevic2014}
-%In wet tropical forests in Central America, \citet{Condit2013} found that dry season intensity was the main determinant of tree species distribution and abundance evenness.
-% Due the potential importance of environment and biogeography in defining the strength and form of a relationship between biodiversity and above ground woody biomass, it is important to sample across broad geographic and environmental gradients to gain understanding of the spatial variation in the relationship between biodiversity and biomass.
-% Environment affects the BEF relationship, so must be accounted for.
-Environmental heterogeneity is known to affect both woody biomass and tree species diversity independently, in a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
-Southern African woodlands particularly, occur over a wide range of precipitation, diurnal and annual temperature, and disturbance regimes \citep{Frost1996}.
-It is important therefore to account for this environmental heterogeneity and understand how it influences both biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass.
-\citet{Sankaran2005} and \citet{Lehmann2014} both report independently that total precipitation sets the upper limit for woody biomass in African savannas.
+% Environment affects the BEFR, so must be accounted for.
+Southern African woody savannas occur over a wide range of precipitation, temperature, soil and disturbance conditions \citep{Frost1996}.
+Environmental heterogeneity has been shown to affect both woody biomass and tree species diversity independently, across a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
+It is important therefore to account for environmental heterogeneity and understand how it influences both biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass.
+\citet{Sankaran2005} and \citet{Lehmann2014} both report that total precipitation sets the upper limit for woody biomass in African savannas.
 \citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate.
 
-% Disturbance deserves another look
-High levels of disturbance in southern African woodlands may moderate the observable BEFR through its effect on ecosystem composition.
-Fire disturbance in forests has been linked to abundance-dependent mortality among smaller trees \citep{Roques2001, Staver2009, Bond2005}.
-Some species in the regional species pool may be excluded from woodland plots with high levels of disturbance if they are unable to escape the fire bottleneck and grow to become a large tree.
-Selection effects may therefore be more important in maximising ecosystem function in disturbance-prone woodlands.
-If a given woodland plant community contains a large number of species, it is more likely that one of them will possess the necessary growth strategy to grow to a large tree with high biomass under an intense disturbance regime.
-
-% Structural diversity is important too
-Southern African woodlands possess structurally diverse tree canopies, with trees occupying distinct layers of the canopy, depending on their growth stages and species identity \citep{Solbrig1996}.
-This structural diversity may be one mechanism through which tree species diversity influences woody biomass.
-\citet{Kunz2019} found that crown complementarity and crown plasticity both increased with species richness in a seasonally dry subtropical forest.
-They also found that trees growing in species-rich neighbourhoods exhibited enhanced biomass production.
-Occupancy of multiple canopy layers allows a fuller canopy with  greater total foliage density, enhancing productivity and allowing greater standing woody biomass in a smaller area via a form of niche complementarity.
-This mechanism however, which has been supported by experiments and observational studies in temperate and wet tropical ecosystems \citep{Hardiman2011, Stark2012}, may not be relevant in savannas.
-Instead, the overriding importance of disturbance history may negate the effects of tree species diversity on structural diversity \citep{Grime2012}.
-
 % In this study :::
-In this study, we make the first known estimation of the biodiversity-ecosystem function relationship (BEFR) across a sub-continental area of disturbance-driven ecosystems, namely southern African savannas and woodlands.
-We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to seeking a general model of the biodiversity-ecosystem function relationship in wooded ecosystems.
-We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), due to an increased importance of niche complementarity as stem density and therefore competition increases.
+In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) across a sub-continental area of disturbance-driven mesic woody savannas in southern Africa.
+We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to contributing to a general model of the BEFR in wooded ecosystems.
+We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), due to increased niche complementarity with stem density.
 In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass.
@@
 
-% Situation
-The study used \nplots{} woodland monitoring plots from the larger SEOSAW network \citep{seosaw_web} located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}).
+% Situation - DONE
+The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{seosaw_web}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}).
 The study area spans the core climate space of the region, with a precipitation gradient from c.
@@
 
-% Plots chosen
-Plots were chosen from a larger pool of 5395 plots held in the SEOSAW database \citep{seosaw_web} based on the quality and completeness of data collection, and plot setup.
-Plot vegetation was identified under the broad term of `savannah', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
+% Plots chosen - DONE
+Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration.
+Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
-Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate estimates.
-Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered a separate biome with very different species composition \citep{Parr2014}.
-
-% Zambia disclaimer
-\nzam{} plots provided by the 2005-2008 Zambian Integrated Land Use Assessment \citep{Mukosha2009, Pelletier2018} were arranged in clusters of four 20x50 m plots, 20 metres apart.
-Data from each plot within a cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.
+Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate biomass estimates.
+Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered here a separate biome with very different species composition \citep{Parr2014}.
+\nzam{} plots within the SEOSAW database were arranged in clusters of four 20x50 m plots, with 20 m between plots.
+Plots within each spatial cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.
 

### 2021-01-28 scripts/try_ext.R

@@ -1,0 +1,217 @@
+# Extract TRY database values
+# John Godlee (johngodlee@gmail.com)
+# 2021-01-25
+
+# Packages
+library(data.table)
+library(dplyr)
+library(ggplot2)
+library(sf)
+library(rnaturalearth)
+library(rnaturalearthdata)
+
+source("scripts/clust_defin.R")
+
+# Import data
+dat <- fread("data/try/13399.txt", 
+  header = T, sep = "\t", dec = ".", quote = "", data.table = T)
+
+# Species from SEOSAW
+plots <- readRDS("data/plot_data_fil_agg_norm_std.rds")
+ab_mat <- readRDS("data/stems_ab_mat.rds")
+
+# All plots represented in abundance matrix?
+stopifnot(all(row.names(ab_mat) %in% plots$plot_cluster))
+
+# Clean data
+dat_clean <- dat %>%
+  dplyr::select(
+    dataset = Dataset,
+    obs_id = ObservationID,
+    species_id = AccSpeciesID,
+    species = AccSpeciesName,
+    trait_id = TraitID,
+    trait = TraitName,
+    key_id = DataID,
+    key = DataName,
+    val = OrigValueStr,
+    unit = OrigUnitStr,
+    val_std = StdValue,
+    unit_std = UnitName,
+    error_risk = ErrorRisk) %>%
+  filter(species %in% names(ab_mat))
+
+# species ID match
+species_id_lookup <- dat_clean %>%
+  dplyr::select(species_id, species) %>%
+  distinct()
+stopifnot(all(!duplicated(species_id_lookup$species)))
+
+# Trait ID match
+trait_id_lookup <- dat_clean %>%
+  dplyr::select(trait_id, trait) %>%
+  filter(!is.na(trait_id)) %>%
+  distinct() %>%
+  mutate(
+    trait = case_when(
+      trait_id %in% c(3115, 3116, 3117) ~ gsub(":.*", "", trait),
+      TRUE ~ trait),
+    trait_short = case_when(
+      trait_id == 14 ~ "Leaf N / dry mass",
+      trait_id == 15 ~ "Leaf P / dry mass",
+      trait_id == 3115 ~ "SLA",
+      trait_id == 3106 ~ "Height",
+      trait_id == 37 ~ "Phenology type",
+      trait_id == 3117 ~ "SLA",
+      trait_id == 38 ~ "Woodiness",
[…]

### 2021-02-01 manuscript/regional_befr_manuscript.tex
Made TRY exploration production ready

@@
 Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{}.
-Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems.
+Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies mostly in temperate grasslands \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems \citep{Plas2019}.
 While positive BEFRs which align with theory have been frequently reported \citep{}, as research has expanded a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}.
@@
 
-There has been extended debate on whether positive BEFRs should be expected in all environmental and ecological contexts \citep{}.
-In temperate forests, wet tropical forests and grasslands, where the majority of BEFR studies in natural systems have been conducted \citep{}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition is the limiting factor to ecosystem functioning \citep{}, but this may not hold true in all systems.
-In mesic savannas for instance, disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
-Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant forces influencing ecosystem function in these ecosystems \citep{Lasky2014, Tobner2016}.
+There has been extended debate on whether positive BEFRs should be expected in all forest systems \citep{}.
+In temperate and wet tropical forests, where the majority of BEFR studies in natural forest ecosystems have been conducted \citep{Plas2019}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition between trees is the limiting factor to ecosystem functioning \citep{}, but this may not hold true in all systems.
+
+Disturbance-driven mesic savannas and open canopy woodlands cover >20\% of the global land surface \citep{Pennington2018, Solbrig1996}, and represent the dominant vegetation type in Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
+Taken together, the above- and below-ground carbon stored in African mesic savannas is comparable to that found in the wet forests of the Congo basin \citep{Houghton2009, Mayaux2008, Spawn2020}.
+Despite their global importance however, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2018}.
+In mesic savannas disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
+Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant biotic forces influencing ecosystem function in these ecosystems \citep{Lasky2014, Tobner2016}.
 It is possible that a threshold woody stem density exists below which the lack of competition precludes the detection of a biodiversity effect on ecosystem function.
@@
 
-Disturbance-driven mesic savannas and open canopy woodlands cover >20\% of the global land surface \citep{Pennington2018}, and represent the dominant vegetation type in southern Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
-Despite their global importance, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2018}.
-A number of studies, all of which were restricted in spatial scope to a small region of miombo woodland, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
-These studies however, lacked the breadth of sampling required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR.
+A number of studies, all of which were restricted in spatial scope to a small region of miombo woodland, the dominant mesic savanna type in southern Africa \citep{}, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR.
+Studies of the BEFR often find that within a small spatial extent (<50 km), biodiversity shows a strong effect on ecosystem function, but at broader extents (>1000s km) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}.
+In West Africa, \citet{Mensah2020} found that woodlands and some forests showed a positive effect of tree species richness on above-ground carbon, while open savannas did not, implying that stem density and the presence of dominant large trees may affect the strength of the observed BEFR.
+In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity both directly and indirectly via its effect on functional trait values that are associated with limited productivity.
+In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.
+Other studies focussing on the herbaceous diversity in disturbance-prone savannas and grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, and allows for weak competitors to co-exist \citep{Mason2011}.
+While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function.
+Additionally, due to differences in community assembly, evolutionary history and contemporary drivers between the neotropics and the African tropics, inferences from one continent cannot necessarily be applied to the other \citep{Dexter2015}.
 
 % Environment affects the BEFR, so must be accounted for.
-Southern African woody savannas occur over a wide range of precipitation, temperature, soil and disturbance conditions \citep{Frost1996}.
+As well as variation in disturbance regime, southern African savannas occur over a wide range of precipitation, temperature, and soil conditions \citep{Frost1996}.
 Environmental heterogeneity has been shown to affect both woody biomass and tree species diversity independently, across a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
@@
 \citet{Sankaran2005} and \citet{Lehmann2014} both report that total precipitation sets the upper limit for woody biomass in African savannas.
-\citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate.
+\citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate on ecosystem function in this system.
 
@@
 In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) across a sub-continental area of disturbance-driven mesic woody savannas in southern Africa.
-We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to contributing to a general model of the BEFR in wooded ecosystems.
-We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), due to increased niche complementarity with stem density.
+We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to generating a general model of the BEFR in disturbance-prone forested ecosystems, which is currently lacking in the BEFR literature.
+We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), as competitive interactions become more influential as stem proximity increases.
 In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass.
@@
 % Situation - DONE
-The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{seosaw_web}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}).
+The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{SEOSAW2020}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}).
 The study area spans the core climate space of the region, with a precipitation gradient from c.
@@
 
-To explore the hypothesis that niche complementarity effects increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density.
+To explore the hypothesis that biodiversity effects on ecosystem function increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density.
 We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB.
@@
 % Correlations
-Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity and structural diversity variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig.
+Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficient of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig.
 S3).
@@
 
-In our sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}).
+In the sub-sampling of the plot dataset by stem density, we found an increasing positive effect of tree species diversity on AGB as stem density increased (\autoref{sem_struc_stems_ha}e).
 There appears to be a minimum stem density threshold at c.
-180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass.
-The effect of structural diversity on AGB appears to remain constant with increasing stem density.
-The indirect effect of tree species diversity on AGB via structural diversity increases as stem density increases.
+180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass (\autoref{sem_struc_stems_ha}b).
+The effect of structural diversity on AGB appears to remain constant with increasing stem density (\autoref{sem_struc_stems_ha}d).
+The indirect effect of tree species diversity on AGB via structural diversity increases as stem density increases (\autoref{sem_struc_stems_ha}c).
 

### 2021-02-01 scripts/sem.R

@@ -228,11 +228,11 @@ mod_summ_mutate_struc <- function(x){
       op == "Direct: AGB" & rhs == "div" ~ "Species\ndiversity",
       op == "Direct: AGB" & rhs == "struc" ~ "Structural\ndiversity",
       op == "Direct: AGB" & rhs == "n_trees_gt10_ha_log_std" ~ "Stem density",
-      lhs == "struc" & op == "Other" & rhs == "div" ~ "Species diversity \u2192\nStructural diversity",
-      lhs == "stems_ha_log_std" & op == "Other" & rhs == "div" ~ "Species diversity \u2192\nStem density",
-      op == "Other" & rhs == "a*b" ~ "Indirect:\nSpecies diversity \u2192\nStructural diversity \u2192\nAGB",
-      op == "Other" & rhs == "d*e" ~ "Indirect:\nSpecies diversity \u2192\nStem density \u2192\nAGB",
-      op == "Other" & (rhs == "c+(a*b)+(d)" | rhs == "c+(a*b)+(d*e)" | rhs == "c+(a*b)")  ~ "Total effect:\nSpecies diversity \u2192\nAGB",
+      lhs == "struc" & op == "Other" & rhs == "div" ~ "Species diversity ->\nStructural diversity",
+      lhs == "stems_ha_log_std" & op == "Other" & rhs == "div" ~ "Species diversity ->\nStem density",
+      op == "Other" & rhs == "a*b" ~ "Indirect:\nSpecies diversity ->\nStructural diversity ->\nAGB",
+      op == "Other" & rhs == "d*e" ~ "Indirect:\nSpecies diversity ->\nStem density ->\nAGB",
+      op == "Other" & (rhs == "c+(a*b)+(d)" | rhs == "c+(a*b)+(d*e)" | rhs == "c+(a*b)")  ~ "Total effect:\nSpecies diversity ->\nAGB",
     TRUE ~ rhs))
 }
 
@@ -552,11 +552,11 @@ struc_sem_quant_fit_df$n_plots <- quant_plots[sapply(struc_sem_quant_fit, nrow) 
 struc_sem_quant_regs <- do.call(rbind, struc_sem_quant_regs_list) %>%
   dplyr::select(rhs, quant, stems_ha, n_plots, n_species_raref, est, std.all) %>%
   mutate(rhs = case_when(
[…]

### 2021-02-01 scripts/try_ext.R

@@ -12,30 +12,34 @@ library(sf)
 library(rnaturalearth)
 library(rnaturalearthdata)
 library(BIOMASS)
+library(purrr)
+library(broom)
 
 source("scripts/clust_defin.R")
 
 # Import data
-dat <- fread("data/try/13399.txt", 
-  header = T, sep = "\t", dec = ".", quote = "", data.table = T)
+try_dat <- fread("data/try/13399.txt", 
+  header = TRUE, sep = "\t", dec = ".", quote = "", data.table = FALSE)
 
-# Species from SEOSAW
+ Species from SEOSAW
 plots <- readRDS("data/plot_data_fil_agg_norm_std.rds")
-ab_mat <- readRDS("data/trees_ab_mat.rds")
 stems <- readRDS("data/stems.rds")
 
-# All plots represented in abundance matrix?
-stopifnot(all(row.names(ab_mat) %in% plots$plot_cluster))
+# How many species are we searching for?
+length(unique(stems$species_name_clean))
 
-# Gather abundance matrix
-ab <- ab_mat %>%
-  rownames_to_column("plot_cluster") %>%
-  gather(species, n, -plot_cluster) %>%
-  mutate(genus = unlist(lapply(strsplit(species, " "), "[", 1))) %>%
-  filter(n > 0)
+# Find wood density values per species from Zanne et al. 2009
+wd_sp <- BIOMASS::wdData %>%
+  mutate(species = paste(genus, species)) %>%
+  filter(
+    species %in% unique(stems$species_name_clean),
+    grepl("africa", region, ignore.case = TRUE)) %>%
+  group_by(species) %>%
+  summarise(val_std = mean(wd, na.rm = TRUE)) %>%
+  mutate(trait_id = "wd")
 
-# Clean data
-dat_clean <- dat %>%
+# Clean TRY data
+try_clean <- try_dat %>%
   dplyr::select(
     dataset = Dataset,
     obs_id = ObservationID,
@@ -49,25 +53,14 @@ dat_clean <- dat %>%
     unit = OrigUnitStr,
     val_std = StdValue,
     unit_std = UnitName,
-    error_risk = ErrorRisk) %>%
-  mutate(genus = unlist(lapply(strsplit(species, " "), "[", 1)))
-
-# Filter to only the species in my dataset
-dat_species <- dat_clean %>%
-  filter(species %in% unique(ab$species)) %>%
-  as.data.frame()
-
-# Generate TRY species ID lookup table 
[…]

### 2021-02-04 manuscript/regional_befr_manuscript.tex
Added changes

@@
 % BEFR is important to study, setting the scene - DONE
-Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{}.
+Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{Naeem2012}.
 Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies mostly in temperate grasslands \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems \citep{Plas2019}.
-While positive BEFRs which align with theory have been frequently reported \citep{}, as research has expanded a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}.
+While positive BEFRs which align with theory have been frequently reported, as research has expanded a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{Liang2016}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}.
 The goal now should be to study the BEFR in different environmental and ecological contexts, in order to develop an ecosystem-agnostic understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.
@@
 Ecosystem functions are defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}.
-Woody productivity and biomass storage are two of the most commonly studied ecosystem functions in forested ecosystems \citep{Pacala2002}, due to their importance in the global carbon cycle and their potential leverage as a tool to mitigate the effects of anthropogenic climate change while maintaining biodiversity \citep{}.
-Theory predicts that biodiversity will have a positive effect on ecosystem function via three principle mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as intraspecific competition and abundance of natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{}.
+Woody productivity and biomass storage are two of the most commonly studied ecosystem functions in forested ecosystems \citep{Brockerhoff2017}, due to their importance in the global carbon cycle \citep{Pan2011} and their potential leverage as a tool to mitigate the effects of anthropogenic climate change while maintaining biodiversity \citep{Pichancourt2013}.
+Theory predicts that biodiversity will have a positive effect on ecosystem function via three principle mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as intraspecific competition and abundance of natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{Wright2017}.
 
-There has been extended debate on whether positive BEFRs should be expected in all forest systems \citep{}.
-In temperate and wet tropical forests, where the majority of BEFR studies in natural forest ecosystems have been conducted \citep{Plas2019}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition between trees is the limiting factor to ecosystem functioning \citep{}, but this may not hold true in all systems.
+There has been extended debate on whether positive BEFRs should be expected in all forest systems \citep{Liang2016}.
+In temperate and wet tropical forests, where the majority of BEFR studies in natural forest ecosystems have been conducted \citep{Plas2019}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition between trees is the limiting factor to ecosystem functioning, but this may not hold true in all systems.
 
@@
 Taken together, the above- and below-ground carbon stored in African mesic savannas is comparable to that found in the wet forests of the Congo basin \citep{Houghton2009, Mayaux2008, Spawn2020}.
-Despite their global importance however, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2018}.
+Despite their global importance however, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2017, Liang2016}.
 In mesic savannas disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
@@
 
-A number of studies, all of which were restricted in spatial scope to a small region of miombo woodland, the dominant mesic savanna type in southern Africa \citep{}, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+A number of studies in miombo woodlands, the dominant mesic savanna type in southern Africa \citep{Campbell1996}, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
 These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR.
 Studies of the BEFR often find that within a small spatial extent (<50 km), biodiversity shows a strong effect on ecosystem function, but at broader extents (>1000s km) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}.
-In West Africa, \citet{Mensah2020} found that woodlands and some forests showed a positive effect of tree species richness on above-ground carbon, while open savannas did not, implying that stem density and the presence of dominant large trees may affect the strength of the observed BEFR.
-In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity both directly and indirectly via its effect on functional trait values that are associated with limited productivity.
+In West Africa, \citet{Mensah2020} found that woodlands and forests showed a positive effect of tree species richness on above-ground carbon, while open savannas did not, implying that stem density and the presence of dominant large trees may affect the strength of the observed BEFR.
+In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity via its effect on functional trait values.
 In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.
@@
 % Environment affects the BEFR, so must be accounted for.
-As well as variation in disturbance regime, southern African savannas occur over a wide range of precipitation, temperature, and soil conditions \citep{Frost1996}.
+As well as variation in disturbance regime, southern African savannas occur over a wide range of precipitation, temperature, and soil conditions \citep{Campbell1996}.
 Environmental heterogeneity has been shown to affect both woody biomass and tree species diversity independently, across a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
@@
 Estimated tree species richness was calculated for each plot using `ChaoRichness()' from the `iNEXT' package in R \citep{Hsieh2016}.
-This procedure extrapolates a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
-Extrapolated species richness accounts for variation in plot size (0.1-10 ha) and therefore sampling effort among plots.
+This procedure uses Hill numbers of the order $q=0$ \citep{Jost2006} to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
+Extrapolated species richness accounts for variation in plot size present in the dataset (0.1-10 ha) and therefore sampling effort among plots.
 Larger plots will tend to encompass more individuals, and therefore more species \citep{Dengler2009}.
@@
 
-To explore the hypothesis that biodiversity effects on ecosystem function increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} datasets of similar size with varying median stem density.
-We used each of these datasets separately to fit the model including only tree species and structural diversity latent variables to predict AGB.
-We excluded the effect of stem density on AGB and the correlation between stem density and species diversity from this model as we deliberately controlled stem density in our subsampling.
-We then examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of subsampled datasets.
-Preliminary analyses that included herbivore biomass \citep{Hempson2017} did not converge, possibly due to the spatially coarse nature of the available data, we therefore did not include herbivory in our final model.
+To explore the hypothesis that biodiversity effects on ecosystem function increase in strength as stem density increases, we repeatedly sub-sampled the available plot dataset to create \subn{} data subsets with similar stem density.
+For each data subset we separately fitted a model including tree species and structural diversity latent variables to predict AGB.
+As we controlled for stem density via the dataset sub-sampling process, the effect of stem density on AGB was not included in the model.
+We examined how the unstandardised path coefficients for each path in the SEM varied according to the median stem density of the data subsets.
+
+Preliminary models that included herbivore biomass \citep{Hempson2017} did not converge, possibly due to the spatially coarse nature of the available data, we therefore did not include herbivory in our final model.
 We incorporated environmental covariates into our model to understand the relative effects of water availability, soil fertility and disturbance on AGB both directly and indirectly via species diversity and stem density.
@@
 % Correlations
-Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficient of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig.
+Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficients of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig.
 S3).
@@
 The R\textsuperscript{2} of AGB was highest in ex-Acacia shrubland (R\textsuperscript{2} = \struccrsq{}) and lowest in sparse miombo / Baikiaea (R\textsuperscript{2} = \strucarsq{}).
-The total effect of species diversity on AGB remained strongly positive for all vegetation types.
-All vegetation types exhibited a positive effect of species diversity on structural diversity.
+The total effect of species diversity on AGB remained strongly positive and there was a positive direct effect of species diversity on structural diversity, across all vegetation types.
 All models had adequate goodness-of-fit (\autoref{struc_model_fit_clust_stats}), though confidence intervals around the unstandardised path coefficients were wide particularly for Mopane and ex-Acacia.
@@
 % Re-cap
-In this study, we assessed the importance of [a] tree species diversity, [b] tree structural diversity, [c] resource availability, [d] disturbance by fire, [e] stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM).
+We assessed the importance of a) tree species diversity, b) tree structural diversity, c) resource availability, d) disturbance by fire, e) stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM).
 We found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}).
@@
 
-% 
 We found a consistent positive effect of tree species diversity on AGB.
@@
 This finding is in agreement with many other studies across different ecosystems and biomes, supporting the idea that there is a generalisable positive association between biodiversity and ecosystem function \citep{Liang2016, Cardinale2009}.
-Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of southern African woodlands, a disturbance-structured and poorly studied ecological system.
+Our study provides a novel dissection of the mechanisms underlying this relationship, particularly in the context of southern African woodlands, a disturbance-driven and poorly studied ecological system.
 
@@
 
+Within the latent variable of tree species diversity we found similarly strong factor loadings for both species richness and abundance evenness.
+This demonstrates that species richness and abundance evenness measure different and largely uncorrelated aspects of diversity.
+In terms of the theory of niche complementarity, both species richness and abundance evennes are important to produce the positive diversity effect on ecosystem function.
+Holding either species richness or abundance evenness constant while increasing the other will have the identical effect of reducing the likelihood that a neighbour is a conspecific, thus reducing the occurrence of negative density dependent competition effects and therefore increasing ecosystem function.
+
 \subsection{Effects of water availability, soil fertility and disturbance}
@@
 Compared to moist tropical forests, water availability is more of a limiting factor to tree growth in southern African woodlands, which experience frequent drought.
-Disturbance by fire had a negative total effect on AGB.
-We found negligible indirect effects of disturbance on AGB via species diversity and structural diversity.
+
+Disturbance by fire had a negative total effect on AGB, with most of this negative effect coming from the indirect pathway via stem density.
+This is expected as increased fire frequency is a key mechanism by which mesic savannas maintain an open canopy, rather than shift to a closed canopy forest \citep{Staver2011}.
+There were small positive effects of disturbance by fire on tree species diversity, a finding which is supported by studies elsewhere \citep{Durigan2020, Staver2009}.
+It is suggested that in mesic savannas where the tree-species pool is largely adapted to fire, increased fire may actually increase tree species diversity by allowing weak competitors to co-exist.
 
@@
 This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used.
-SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales \citep{Muledi2017, Bucini2007} in southern African woodlands, often being further structured by the vegetation overlying it, an aspect which SoilGrids does not model precisely.
+SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}, notably by termite mounds which offer small areas of highly fertile soil with peculiar tree species diversity \citep{Joseph2014}.
 Due to the plots used in this study often being situated non-randomly in the landscape, coupled with the coarseness of the SoilGrids data, it is not surprising that this model path is poorly constrained.
 Soil data is time-consuming to collect and difficult to compare across studies when different protocols are used, though this study prompts the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands.
-\citet{Lehmann2014} similarly found weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, fire and tree basal area.
+\citet{Lehmann2014} found similarly weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, and fire to predict tree basal area.
 
@@
 Despite Mopane woodland having very low species diversity generally, with often monospecific stands \citep{Timberlake2010}, a positive effect of tree species diversity on AGB was observed.
-In previous studies across ecosystem types it has been found often that the effect on ecosystem function of adding species is stronger in low diversity assemblages \citep{Hector2007}.
+In previous studies across ecosystem types it has been found often that the effect on ecosystem function of adding species is stronger in low diversity assemblages \citep{Cardinale2006, Srivastava2005}.
 This has been attributed to an increase in functional redundancy as species diversity increases.
@@
 Unlike Mopane woodlands, ex-Acacia woodlands contain a wider variety of species which can grow to large canopy trees, albeit at low densities, especially in transition zones with miombo woodlands.
+Additionally, many more species species in ex-Acacia woodlands are found in the Mimosoideae and Papilionoideae sub-families, of which most are nitrogen-fixing \citep{Tedersoo2018}.
+Nitrogen availability is often a limiting factor in productivity, making nitrogen-fixing species strong competitors.
+It is possible that in ex-Acacia dominated woodlands, the presence of a large number of nitrogen-fixing tree species reduces functional redundancy, meaning that the effect of adding species on ecosystem function saturates at a higher species richness.
 

### 2021-03-17 manuscript/regional_befr_manuscript.tex
Comments from Kyle

@@
 Disturbance may limit competition and niche differentiation, which are frequently posited to underlie BEFRs.
-We explore the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
-	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interactive effects of resource availability, disturbance by fire, stem density and vegetation type.}
+We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
+	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, stem density and vegetation type.}
 	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity.
@@
 180 stems ha\textsuperscript{-1}.
-We found that resource availability mainly affects biomass indirectly, via increasing species diversity.}
+We found that water and soil nutrient availability mainly affects biomass indirectly, via increasing species diversity.}
 	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands.
@@
 Over the past two decades, study of the Biodiversity-Ecosystem Function Relationship (BEFR) has grown from small-scale experimental studies mostly in temperate grasslands \citep{Cardinale2009, Tilman1994, Tilman2014}, to observational studies in natural ecosystems \citep{Plas2019}.
-While positive BEFRs which align with theory have been frequently reported, as research has expanded a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{Liang2016}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}.
+While positive BEFRs which align with theory have been frequently reported, a complex picture has emerged whereby the strength and direction of the BEFR varies depending on the ecosystem studied \citep{Liang2016}, the ecosystem function(s) of interest \citep{Hector2007}, and the inclusion of environmental covariates in statistical models \citep{Vila2005}.
 The goal now should be to study the BEFR in different environmental and ecological contexts, in order to develop an ecosystem-agnostic understanding of the complex interactions between biodiversity, abiotic environment, and ecosystem function.
@@
 Ecosystem functions are defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}.
-Woody productivity and biomass storage are two of the most commonly studied ecosystem functions in forested ecosystems \citep{Brockerhoff2017}, due to their importance in the global carbon cycle \citep{Pan2011} and their potential leverage as a tool to mitigate the effects of anthropogenic climate change while maintaining biodiversity \citep{Pichancourt2013}.
-Theory predicts that biodiversity will have a positive effect on ecosystem function via three principle mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as intraspecific competition and abundance of natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{Wright2017}.
+Woody productivity and biomass storage are two of the most commonly studied ecosystem functions in forests and other wooded ecosystems (e.g.
+savannas) \citep{Brockerhoff2017}, due to their importance in the global carbon cycle \citep{Pan2011} and their potential leverage as a tool to mitigate the effects of anthropogenic climate change while maintaining biodiversity \citep{Pichancourt2013}.
+Theory predicts that biodiversity will have a positive effect on ecosystem function via three principle mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as intraspecific competition and abundance of species-specific natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{Wright2017}.
 
-There has been extended debate on whether positive BEFRs should be expected in all forest systems \citep{Liang2016}.
+There has been extended debate on whether positive BEFRs should be expected in all wooded ecosystems \citep{Liang2016}.
 In temperate and wet tropical forests, where the majority of BEFR studies in natural forest ecosystems have been conducted \citep{Plas2019}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition between trees is the limiting factor to ecosystem functioning, but this may not hold true in all systems.
@@
 Disturbance-driven mesic savannas and open canopy woodlands cover >20\% of the global land surface \citep{Pennington2018, Solbrig1996}, and represent the dominant vegetation type in Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
-Taken together, the above- and below-ground carbon stored in African mesic savannas is comparable to that found in the wet forests of the Congo basin \citep{Houghton2009, Mayaux2008, Spawn2020}.
+Taken together, the above- and below-ground carbon stored in African mesic savannas may be comparable to that found in the wet forests of the Congo basin \citep{Houghton2009, Mayaux2008, Spawn2020}.
 Despite their global importance however, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2017, Liang2016}.
-In mesic savannas disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
+In savannas disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
+It is possible that a threshold woody stem density exists below which the lack of competition precludes the detection of a biodiversity effect on ecosystem function via niche differentiation.
 Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant biotic forces influencing ecosystem function in these ecosystems \citep{Lasky2014, Tobner2016}.
-It is possible that a threshold woody stem density exists below which the lack of competition precludes the detection of a biodiversity effect on ecosystem function.
-Additionally, facilitation effects may play a larger role in driving the BEFR in stressful environments like mesic savannas.
-\citet{Ratcliffe2017} found stronger positive relationships between tree species richness and various ecosystem functions with increasing aridity.
-They demonstrate that competition diminishes where environmental stress limits species' abundances.
+Additionally, stressful environments appear to lessen the role of both niche complementarity and facilitation effects in driving the BEFR, by replacing competition between individuals with stress tolerance as the limiting factor on functional contribution \citep{Ratcliffe2017, Paquette2011}.
 This potential mismatch in the contribution of different mechanisms to the BEFR among resource-, disturbance-, and competition-limited ecosystems requires further investigation if we are to derive a generalisable BEFR.
 
-A number of studies in miombo woodlands, the dominant mesic savanna type in southern Africa \citep{Campbell1996}, found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+A number of studies in miombo woodlands, the dominant mesic savanna type in southern Africa \citep{Campbell1996}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
 These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR.
-Studies of the BEFR often find that within a small spatial extent (<50 km), biodiversity shows a strong effect on ecosystem function, but at broader extents (>1000s km) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}.
+Studies of the BEFR often find that within a small spatial extent (<2500 km\textsuperscript{2}), biodiversity shows a strong effect on ecosystem function, but at broader extents (>100,000s km\textsuperscript{2}) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}.
 In West Africa, \citet{Mensah2020} found that woodlands and forests showed a positive effect of tree species richness on above-ground carbon, while open savannas did not, implying that stem density and the presence of dominant large trees may affect the strength of the observed BEFR.
@@
 In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.
-Other studies focussing on the herbaceous diversity in disturbance-prone savannas and grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, and allows for weak competitors to co-exist \citep{Mason2011}.
-While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function.
+Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, and allows for weak competitors to co-exist \citep{Mason2011}.
+While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function, particularly for wooded ecosystems.
 Additionally, due to differences in community assembly, evolutionary history and contemporary drivers between the neotropics and the African tropics, inferences from one continent cannot necessarily be applied to the other \citep{Dexter2015}.
@@
 % Environment affects the BEFR, so must be accounted for.
-As well as variation in disturbance regime, southern African savannas occur over a wide range of precipitation, temperature, and soil conditions \citep{Campbell1996}.
+%As well as variation in disturbance regime, southern African savannas occur over a wide range of precipitation, temperature, and soil conditions \citep{Campbell1996}.
 Environmental heterogeneity has been shown to affect both woody biomass and tree species diversity independently, across a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
@@
 % In this study :::
-In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) across a sub-continental area of disturbance-driven mesic woody savannas in southern Africa.
-We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR in this system, with a view to generating a general model of the BEFR in disturbance-prone forested ecosystems, which is currently lacking in the BEFR literature.
-We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with stem density (number of stems ha\textsuperscript{-1}), as competitive interactions become more influential as stem proximity increases.
-In addition, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass.
+In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) in disturbance-driven wooded ecosystems (savannas and woodlands) across southern Africa.
+We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR, with a view to generating a general model of the BEFR in disturbance-prone wooded ecosystems, which is currently lacking in the BEFR literature.
+We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with organismal density (number of stems ha\textsuperscript{-1}), as competitive interactions become more pronounced as stem proximity increases.
+Relatedly, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass.
 Finally, we expect that (3) tree species diversity will increase tree structural diversity (i.e.
@@
 Abundance evenness allows for greater niche complementarity at small scales due to potentially increased heterogeneity of functional traits.
+In terms of the theory of niche complementarity, both species richness and abundance evenness contribute to the positive total diversity effect on ecosystem function.
+Holding either species richness or abundance evenness constant while increasing the other will have the identical effect of reducing the likelihood that a neighbour is a conspecific, thus reducing the occurrence of negative density dependent competition effects and therefore increasing ecosystem function.
 We quantified tree structural diversity for each plot by calculating the Coefficient of Variation of DBH (DBH CoV) and tree height (Height CoV).
@@
 % Re-cap
-We assessed the importance of a) tree species diversity, b) tree structural diversity, c) resource availability, d) disturbance by fire, e) stem density and their interactions on above-ground woody biomass (AGB) across southern African woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM).
+We assessed the importance of a) tree species diversity, b) tree structural diversity, c) resource availability, d) disturbance by fire, e) stem density and their interactions on above-ground woody biomass (AGB) across southern African savannas and woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM).
 We found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}).
 Tree species diversity, structural diversity and stem density accounted for \smrsq{}\% of the variation in AGB across the region, while models for specific vegetation types showed even greater explanatory power in some cases (\autoref{struc_model_fit_clust_stats}).
-We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} below which the effect of species diversity on AGB remained at a low baseline level.
+Within the latent variable of tree species diversity we found similarly strong factor loadings for both species richness and abundance evenness.
+This demonstrates that species richness and abundance evenness measure different and largely uncorrelated aspects of diversity.
+We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} >10 cm DBH, below which the effect of species diversity on AGB remained at a low baseline level.
 The strongest direct effect on AGB was that of stem density.
@@
 The opposite causation is also plausible however, with increased stem density causing higher species richness through an increased probability of encountering new species.
-We attempted to correct for the correlation between species richness and stem density using extrapolated species richness, which extrapolates a rarefaction curve to its predicted asymptote, thus estimating the total landscape-level species richness which is unaffected by plot size and stem density.
+We attempted to correct for the correlation between species richness and stem density using extrapolated species richness, which extrapolates a rarefaction curve to its predicted asymptote, thus estimating the total landscape-level species richness which is independent of plot size and stem density.
 We suggest therefore that an increase in tree species diversity through species richness and evenness produces an assemblage of species which can utilise more available light and moisture, resulting in greater plot-level AGB.
@@
 Previous studies have found that southern African woodlands with higher species diversity tend to experience less frequent disturbance by fire and tend to form a more closed canopy and a more sparse understorey \citep{Chidumayo2013, Mutowo2012}.
-In our study however, we found a positive effect of disturbance on species diversity, perhaps suggesting that disturbance prevents domination of woodlands by a single dominant species \citep{Chidumayo2013}.
+In our study however, we found a positive effect of fire frequency on species diversity, perhaps suggesting that disturbance prevents domination of woodlands by a single dominant species \citep{Chidumayo2013}.
 
@@
 180 trees ha\textsuperscript{-1}.
-At very high stem density, there is also an increase in the effect of species diversity on structural diversity.
+At very high stem density, there is an increase in the effect of species diversity on structural diversity.
 This could be because at high stem density, the adaptation of different species to growth form become important.
 At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species are able to exist in the understorey, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.
-
-Within the latent variable of tree species diversity we found similarly strong factor loadings for both species richness and abundance evenness.
-This demonstrates that species richness and abundance evenness measure different and largely uncorrelated aspects of diversity.
-In terms of the theory of niche complementarity, both species richness and abundance evennes are important to produce the positive diversity effect on ecosystem function.
-Holding either species richness or abundance evenness constant while increasing the other will have the identical effect of reducing the likelihood that a neighbour is a conspecific, thus reducing the occurrence of negative density dependent competition effects and therefore increasing ecosystem function.
 
@@
 
-A negative total effect of soil fertility on AGB is in contrast to other studies in the region and against general ecological theory, which predicts a positive effect of soil nutrients on biomass.
+A negative total effect of soil fertility on AGB is in contrast to other studies in the region and against general ecological theory, which predicts a positive effect of soil nutrients on biomass citep{}.
 The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density.
@@
 SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}, notably by termite mounds which offer small areas of highly fertile soil with peculiar tree species diversity \citep{Joseph2014}.
-Due to the plots used in this study often being situated non-randomly in the landscape, coupled with the coarseness of the SoilGrids data, it is not surprising that this model path is poorly constrained.
-Soil data is time-consuming to collect and difficult to compare across studies when different protocols are used, though this study prompts the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands.
-\citet{Lehmann2014} found similarly weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, and fire to predict tree basal area.
+It is therefore not surprising that this model path is poorly constrained.
+\citet{Lehmann2014} found similarly weak and poorly constrained relationships for soil in a Structural Equation Model including precipitation, temperature, soil, and fire to predict tree basal area in southern African woodlands.
+Plot-specific soil data are time-consuming to collect and difficult to compare across studies when different protocols are used.
+Our study points to the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands.
 
@@
 Ex-Acacia and Mopane woodlands are dominated by fewer tree species, notably \textit{Senegalia} spp.
-in ex-Acacia woodlands and \textit{Colophospermum mopane} in Mopane woodlands which often produce large canopy dominating trees.
+in ex-Acacia woodlands and \textit{Colophospermum mopane} in Mopane woodlands, which can produce large canopy dominating trees in the so-called ``Cathedral mopane''.
 We postulate that the slight negative effect of tree species richness on AGB in miombo woodlands may be due to an increase in interspecific competition through canopy crowding, but that this effect is not present in ex-Acacia and Mopane woodlands, where the top level of the woodland canopy is dominated often by a single species.

### 2021-03-17 scripts/try_ext.R

@@ -530,5 +530,4 @@ dev.off()
 ##' line shows the 80% cutoff recommended by Reviewer 1. (b) Relationship 
 ##' between leaf N and AGB for all plots with at least 80% of basal area with 
 ##' leaf N measurements, line of best fit is a linear regression with a 95% 
-##' confidence interval 
-##' (F(1,672) = 1.69, p = 0.2, R2adj = 0.01, β = 0.43±0.334).
+##' confidence interval (F(1,672) = 1.69, p = 0.2, R2adj = 0.01, β = 0.43±0.334).

### 2021-03-17 scripts/try_species_match.R

@@ -3,7 +3,7 @@
 # 2021-01-25
 
 # Import data
-try_sp <- read.table("data/try_species_list.txt", header = TRUE, sep = "\t")
+try_sp <- read.table("data/try/try_species_list.txt", header = TRUE, sep = "\t")
 
 seosaw_stems <- read.csv("~/git_proj/seosaw_data/data_out/v2.10/stems_v2.10.csv")
 

### 2021-03-18 manuscript/regional_befr_manuscript.tex
Updated manuscript

@@
 	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity.
-The observed BEFR was highly dependent on stem density, with a minimum threshold of c.
-180 stems ha\textsuperscript{-1}.
-We found that water and soil nutrient availability mainly affects biomass indirectly, via increasing species diversity.}
+The observed BEFR was highly dependent on organismal density, with a minimum threshold of c.
+180 tree stems ha\textsuperscript{-1}.
+We found that water availability mainly affects biomass indirectly, via increasing species diversity.}
 	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands.
@@
 
-Disturbance-driven mesic savannas and open canopy woodlands cover >20\% of the global land surface \citep{Pennington2018, Solbrig1996}, and represent the dominant vegetation type in Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
+Disturbance-driven mesic savannas and open canopy woodlands cover >20\% of the global land surface \citep{Pennington2018, Solbrig1996}.
+They represent the dominant vegetation type in Africa, spanning >4 million km\textsuperscript{2} \citep{White1987, Ratnam2011, Ryan2016} (\autoref{clust_map}).
 Taken together, the above- and below-ground carbon stored in African mesic savannas may be comparable to that found in the wet forests of the Congo basin \citep{Houghton2009, Mayaux2008, Spawn2020}.
-Despite their global importance however, African mesic savannas are severely under-represented in BEFR studies \citep{Clarke2017, Liang2016}.
+Globally, dry woodlands and savannas represent the largest, most sensitive and fastest increasing component of the terrestrial carbon sink \citep{Sitch2015, Ahlstrom2015}.
+
+Despite their global importance, African savannas are severely under-represented in BEFR studies \citep{Clarke2017, Liang2016}.
 In savannas disturbance by fire \citep{Lehmann2014} and herbivory \citep{Sankaran2008, Levick2009} reduces woody stem density and biomass, diminishing competitive interactions between individuals, allowing competitors to co-exist where they would normally be excluded \citep{Grime1979, Grace1990}.
@@
 Instead, stress tolerance and the functional contribution of particular species (selection effects) may be the predominant biotic forces influencing ecosystem function in these ecosystems \citep{Lasky2014, Tobner2016}.
-Additionally, stressful environments appear to lessen the role of both niche complementarity and facilitation effects in driving the BEFR, by replacing competition between individuals with stress tolerance as the limiting factor on functional contribution \citep{Ratcliffe2017, Paquette2011}.
+Additionally, stressful environments appear to lessen the role of both niche complementarity and facilitation effects in driving the BEFR, by replacing competition between individuals with stress tolerance as the limiting factor on functional contribution \citep{Ratcliffe2017, Paquette2010}.
 This potential mismatch in the contribution of different mechanisms to the BEFR among resource-, disturbance-, and competition-limited ecosystems requires further investigation if we are to derive a generalisable BEFR.
 
-A number of studies in miombo woodlands, the dominant mesic savanna type in southern Africa \citep{Campbell1996}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
+A number of studies in miombo woodlands, the dominant savanna type in southern Africa \citep{Campbell1996}, have found that above-ground woody carbon/biomass stocks correlate positively with tree species richness \citep{McNicol2018, Shirima2015, Mutowo2012}.
 These studies however, lacked the spatial extent required to account for the interacting effects of variation in abiotic environment, disturbance regime, and biogeography which are expected to affect the BEFR.
@@
 
-% Re-cap
 We assessed the importance of a) tree species diversity, b) tree structural diversity, c) resource availability, d) disturbance by fire, e) stem density and their interactions on above-ground woody biomass (AGB) across southern African savannas and woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM).
@@
 This demonstrates that species richness and abundance evenness measure different and largely uncorrelated aspects of diversity.
-We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems ha\textsuperscript{-1} >10 cm DBH, below which the effect of species diversity on AGB remained at a low baseline level.
+We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems >10 cm DBH ha\textsuperscript{-1}, below which the effect of species diversity on AGB remained at a low baseline level.
 The strongest direct effect on AGB was that of stem density.
@@
 greater variation in DBH and height.
-This may act as a mechanism for niche complementarity, with a canopy of diversely sized trees able to take advantage of a greater proportion of the available light.
+This may act as a mechanism for niche complementarity, with a canopy of diversely-sized trees able to take advantage of a greater proportion of the available light.
 Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{Pretzsch2014}, i.e.
@@
 In forests, where the tree canopy is effectively closed, as the stand matures a more diverse canopy emerges via competition and tree mortality events which open canopy gaps \citep{Muscolo2014}.
-Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this mechanism.
-In frequently disturbed woodlands such as those studied here however, a woodland canopy similar to that of a forest is frequently not reached.
-Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
-Previous studies have found that southern African woodlands with higher species diversity tend to experience less frequent disturbance by fire and tend to form a more closed canopy and a more sparse understorey \citep{Chidumayo2013, Mutowo2012}.
-In our study however, we found a positive effect of fire frequency on species diversity, perhaps suggesting that disturbance prevents domination of woodlands by a single dominant species \citep{Chidumayo2013}.
-
-We found a non-linear positive effect of stem density on the relationship between tree species diversity and AGB (\autoref{sem_struc_stems_ha}).
+Indeed, our finding that the strength of the effect of tree diversity on AGB increases with stem density supports this mechanism (\autoref{sem_struc_stems_ha}).
 At low stem densities, competition between mature trees may not occur, meaning that the niche complementarity effect provided by an increase in tree species richness may not be present, accounting for the small effect of tree species diversity on AGB below c.
 180 trees ha\textsuperscript{-1}.
-At very high stem density, there is an increase in the effect of species diversity on structural diversity.
-This could be because at high stem density, the adaptation of different species to growth form become important.
-At low stem density, individual trees tend to spread out rather than growing tall, but at high stem density, only certain species are able to exist in the understorey, while others are able to grow tall above the woodland canopy, leading to greater variation in tree height over the plot.
+In frequently disturbed woodlands such as those studied here, a woodland canopy similar to that of a forest is frequently not reached.
+Instead, a simple open canopy is maintained that can be made more complex and productive via an increase in species diversity.
 
-\subsection{Effects of water availability, soil fertility and disturbance}
+\subsection{Organismal density and disturbance}
+
+Disturbance by fire had a negative total effect on AGB, with most of this negative effect coming from the indirect pathway via stem density.
+This is expected as increased fire frequency is a key mechanism by which mesic savannas maintain an open canopy, rather than shifting to a closed canopy forest \citep{Staver2011}.
+Previous studies have found that southern African woodlands with higher species diversity tend to experience less frequent disturbance by fire and tend to form a more closed canopy with a sparse understorey \citep{Chidumayo2013, Mutowo2012}.
+In our study however, we found a positive effect of fire frequency on species diversity, perhaps suggesting that disturbance prevents domination of woodlands by a single dominant species \citep{Chidumayo2013, Durigan2020, Staver2009}.
+It is suggested that in savannas where the tree-species pool is largely adapted to fire, increased fire may actually increase tree species diversity by allowing weak competitors to co-exist.
+
+As time since disturbance increases in any wooded ecosystem, it is expected that both species diversity and above-ground biomass will increase until an equilibrium is reached or another disturbance event occurs \citep{Huston2014}.
+Unless this effect is accounted for, there is the potential for mistaken causality as both diversity and biomass will correlate.
+In our model, time since disturbance is accounted for within each plot via the stem density term.
+Disturbance reduces stem density, which is expected to increase until the effects of competition preclude further increase \citep{Johnson2012}.
+Furthermore, our rarefied measure of species diversity accounts for variation in sampling effort and is therefore independent of stem density.
+Tree species richness should also increase with time since disturbance as with increased stem density the likelihood of including a new species also increases.
+Outside of the stem density effect, there are multiple causes for variation in tree species diversity in this study.
+Vegetation types and localities differ in their available species pool, for example.
+Variation in abiotic environmental factors will also affect species accumulation.
+
+\subsection{Effects of water availability and soil fertility}
 
@@
 
-Disturbance by fire had a negative total effect on AGB, with most of this negative effect coming from the indirect pathway via stem density.
-This is expected as increased fire frequency is a key mechanism by which mesic savannas maintain an open canopy, rather than shift to a closed canopy forest \citep{Staver2011}.
-There were small positive effects of disturbance by fire on tree species diversity, a finding which is supported by studies elsewhere \citep{Durigan2020, Staver2009}.
-It is suggested that in mesic savannas where the tree-species pool is largely adapted to fire, increased fire may actually increase tree species diversity by allowing weak competitors to co-exist.
-
-A negative total effect of soil fertility on AGB is in contrast to other studies in the region and against general ecological theory, which predicts a positive effect of soil nutrients on biomass citep{}.
+A negative total effect of soil fertility on AGB is in contrast to other studies in the region and general ecological theory, which predicts a positive effect of soil nutrients on biomass \citep{Scarascia2000}.
 The negative total effect of soil fertility on AGB was driven mostly by an indirect negative effect via stem density.
@@
 This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used.
-SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}, notably by termite mounds which offer small areas of highly fertile soil with peculiar tree species diversity \citep{Joseph2014}.
+SoilGrids provides modelled data at 250 m resolution, while sostructure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}, notably by termite mounds which offer small areas of highly fertile soil with peculiar tree species diversity \citep{Joseph2014}.
 It is therefore not surprising that this model path is poorly constrained.
@@
 Despite Mopane woodland having very low species diversity generally, with often monospecific stands \citep{Timberlake2010}, a positive effect of tree species diversity on AGB was observed.
-In previous studies across ecosystem types it has been found often that the effect on ecosystem function of adding species is stronger in low diversity assemblages \citep{Cardinale2006, Srivastava2005}.
+In previous studies across multiple biomes it has been found that the effect of adding species on ecosystem function is stronger in low diversity assemblages \citep{Cardinale2006, Srivastava2005}.
 This has been attributed to an increase in functional redundancy as species diversity increases.
-In other words, with more species, it is more likely that the addition of a new species will occupy the same ecological niche space as an existing species, meaning niche complementarity will not occur and competition will not lead to niche partitioning, making little difference to overall ecosystem functioning.
 Mopane woodlands also have a negligible effect of species diversity on structural diversity.
-This may be due to the species which tend to co-exist with \textit{C.
-mopane}, many of which are small shrub-like trees and which do not grow into large canopy trees \citep{Timberlake2010}.
-Larger canopy trees tend to have greater variation in physical structure \citep{Seidel2019}.
+This may be due to the particular functional forms of species which co-exist with \textit{C.
+mopane}, many of which are small shrub-like trees rather than large canopy trees \citep{Timberlake2010}.
+Larger canopy trees tend to have greater variation in physical structure \citep{Seidel2019} which would drive an effect of species diversity on structural diversity as we observed in miombo woodlands.
 
@@
 We assert that this is necessary to develop our understanding of the biodiversity-ecosystem function relationship in real-world ecosystems, to progress from experimental mesocosms.
-We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key vector through which tree species diversity determines ecosystem function in savannas, woodlands and forests, i.e.
+We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key mechanism by which tree species diversity determines ecosystem function in savannas, woodlands and forests, i.e.
 all wooded ecosystems.
@@
 This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.
+Finally, our results provide further evidence of the complex interaction of factors governing biomass and therefore carbon dynamics in disturbance-driven wooded ecosystems, which currently represent the greatest uncertainty in the global terrestrial carbon sink.
 

### 2021-03-18 manuscript/regional_befr_manuscript.tex
Sent paper for feedback

@@
 We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
-	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, stem density and vegetation type.}
+	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density and vegetation type.}
 	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity.
@@
 	\item{The study underlines the close association between tree diversity, ecosystem structure, environment and function in highly disturbed savannas and woodlands.
-We suggest that tree diversity is an under-appreciated determinant of savanna and woodland structure and function.}
+We suggest that tree diversity is an under-appreciated determinant of wooded ecosystem structure and function.}
 \end{itemize}
@@
 savannas) \citep{Brockerhoff2017}, due to their importance in the global carbon cycle \citep{Pan2011} and their potential leverage as a tool to mitigate the effects of anthropogenic climate change while maintaining biodiversity \citep{Pichancourt2013}.
-Theory predicts that biodiversity will have a positive effect on ecosystem function via three principle mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as intraspecific competition and abundance of species-specific natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{Wright2017}.
+Theory predicts that biodiversity will have a positive effect on ecosystem function via three principal mechanisms: 1) niche complementarity, whereby niche partitioning and ecosystem resource use efficiency is increased in diverse communities, minimising negative density dependent effects such as conspecific competition and abundance of species-specific natural enemies \citep{Barry2019}; 2) selection effects, whereby diverse communities are more likely to include a high-yield species; and 3) facilitation effects, whereby diverse communities are more likely to contain species combinations which enhance each others' functional contribution \citep{Wright2017}.
 
 There has been extended debate on whether positive BEFRs should be expected in all wooded ecosystems \citep{Liang2016}.
-In temperate and wet tropical forests, where the majority of BEFR studies in natural forest ecosystems have been conducted \citep{Plas2019}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that intraspecific competition between trees is the limiting factor to ecosystem functioning, but this may not hold true in all systems.
+In temperate and wet tropical forests, where the majority of BEFR studies in natural forest ecosystems have been conducted \citep{Plas2019}, the mechanism of niche complementarity, which contributes the majority of the observed biodiversity effect \citep{Poorter2015, Sande2017a, Wright2017}, hinges on the condition that conspecific competition between trees is the limiting factor to ecosystem functioning, but this may not hold true in all systems.
 
@@
 Taken together, the above- and below-ground carbon stored in African mesic savannas may be comparable to that found in the wet forests of the Congo basin \citep{Houghton2009, Mayaux2008, Spawn2020}.
-Globally, dry woodlands and savannas represent the largest, most sensitive and fastest increasing component of the terrestrial carbon sink \citep{Sitch2015, Ahlstrom2015}.
+Globally, dry woodlands and savannas represent the largest, most sensitive and fastest increasing component of the terrestrial carbon sink \citep{Ahlstrom2015}.
+Yet their carbon dynamics remain poorly constrained \citep{Sitch2015}.
 
@@
 Studies of the BEFR often find that within a small spatial extent (<2500 km\textsuperscript{2}), biodiversity shows a strong effect on ecosystem function, but at broader extents (>100,000s km\textsuperscript{2}) biodiversity effects pale in significance compared to abiotic factors such as climate \citep{Gonzalez2020}.
-In West Africa, \citet{Mensah2020} found that woodlands and forests showed a positive effect of tree species richness on above-ground carbon, while open savannas did not, implying that stem density and the presence of dominant large trees may affect the strength of the observed BEFR.
+In West Africa, \citet{Mensah2020} found that woodlands and forests showed a positive effect of tree species richness on above-ground carbon, while sparse savannas did not, implying that tree stem density and the presence of dominant large trees may affect the strength of the observed BEFR.
 In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity via its effect on functional trait values.
 In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.
-Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, and allows for weak competitors to co-exist \citep{Mason2011}.
+Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}.
 While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function, particularly for wooded ecosystems.
@@
 
-% Environment affects the BEFR, so must be accounted for.
-%As well as variation in disturbance regime, southern African savannas occur over a wide range of precipitation, temperature, and soil conditions \citep{Campbell1996}.
-Environmental heterogeneity has been shown to affect both woody biomass and tree species diversity independently, across a number of different biomes \citep{Spasojevic2014, Michaletz2018, Michaletz2014}.
-It is important therefore to account for environmental heterogeneity and understand how it influences both biomass and biodiversity to effectively model and correctly attribute the effects of biodiversity on woody biomass.
-\citet{Sankaran2005} and \citet{Lehmann2014} both report that total precipitation sets the upper limit for woody biomass in African savannas.
-\citet{Lehmann2014} also report complex indirect relationships between climate, disturbance by fire and woody biomass, demonstrating the need for directional multi-facetted modelling techniques to properly account for the effects of climate on ecosystem function in this system.
-
 % In this study :::
-In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) in disturbance-driven wooded ecosystems (savannas and woodlands) across southern Africa.
-We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR, with a view to generating a general model of the BEFR in disturbance-prone wooded ecosystems, which is currently lacking in the BEFR literature.
-We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with organismal density (number of stems ha\textsuperscript{-1}), as competitive interactions become more pronounced as stem proximity increases.
-Relatedly, we expect that an increase in disturbance by fire will decrease stem density and therefore competition, weakening the effect of tree species diversity on woody biomass.
-Finally, we expect that (3) tree species diversity will increase tree structural diversity (i.e.
+In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) in disturbance-driven wooded ecosystems (savannas and woodlands), using southern Africa as our study region.
+We aim to understand the synergistic effects of environmental and biotic drivers of variation in the BEFR, with a view to creating a general model of the BEFR in disturbance-prone wooded ecosystems, which is currently lacking in the BEFR literature.
+We posit three hypotheses: (1) water availability and soil fertility will indirectly positively affect woody biomass via an increase in tree species diversity, (2) the effect of tree species diversity on woody biomass will increase with organismal density (number of stems ha\textsuperscript{-1}), with competitive interactions becoming more pronounced as stem proximity increases.
+Relatedly, we expect that an increase in disturbance by fire will decrease organismal density and therefore competition, weakening the effect of tree species diversity on woody biomass.
+Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in functional form (i.e.
 physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.
@@
 Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration.
-Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands \citep{Ratnam2011, Hill2010}.
+Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands, or savannas \citep{Ratnam2011, Hill2010}.
 Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
@@
 Woody lianas are scarce in our study plots and were not measured.
-Only stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
+Only living stems >10 cm DBH (Diameter at Breast Height, 1.3 m) were included in analyses.
 Many plots in the dataset did not include data on stems <10 cm DBH.
-For those plots which contained stem measurements <10 cm DBH, small stems only accounted for a median of \percsmallagb{}\% of the plot level AGB.
+For those plots which contained stem measurements <10 cm DBH, small stems only accounted for a median of \percsmallagb{}\% of the plot AGB.
 
@@
 When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
-The height of this measurement was recorded and used to estimate the DBH\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.
+The height of this measurement was recorded and used to estimate the D\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.
 
@@
 \begin{equation}
-	AGB = 0.0673 \times (\rho D^{2} H)^{0.976}
+	AGB = 0.0673 \times (\rho D_{e}^{2} H)^{0.976}
 	\label{chave_agb}
@@
 
-where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), $D$ is the DBH\textsubscript{e} (cm) at 1.3 m, and $H$ is the tree height (m).
+where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), $D_{e}$ is the estimated DBH (cm) at 1.3 m, and $H$ is the tree height (m).
 Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
@@
 Estimated tree species richness was calculated for each plot using `ChaoRichness()' from the `iNEXT' package in R \citep{Hsieh2016}.
-This procedure uses Hill numbers of the order $q=0$ \citep{Jost2006} to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value.
+This procedure uses Hill numbers of the order $q=0$ to extrapolate a species rarefaction curve to its predicted asymptote and uses this value as its estimated species richness value \citep{Jost2006}.
 Extrapolated species richness accounts for variation in plot size present in the dataset (0.1-10 ha) and therefore sampling effort among plots.
@@
 In terms of the theory of niche complementarity, both species richness and abundance evenness contribute to the positive total diversity effect on ecosystem function.
-Holding either species richness or abundance evenness constant while increasing the other will have the identical effect of reducing the likelihood that a neighbour is a conspecific, thus reducing the occurrence of negative density dependent competition effects and therefore increasing ecosystem function.
+Holding either species richness or abundance evenness constant while increasing the other will have the similar effect of reducing the likelihood that a neighbour is a conspecific, thus reducing the occurrence of negative density dependent competition effects and therefore increasing ecosystem function.
 We quantified tree structural diversity for each plot by calculating the Coefficient of Variation of DBH (DBH CoV) and tree height (Height CoV).
@@
 SEM was used because of its suitability for modelling complex causal interactions in ecological systems \citep{Lee2007}.
-A key aspect to our decision to use SEM is that they can explicitly model and partition variance attributed to indirect effects, which is challenging in standard multiple regressions.
+A key aspect consideration in our decision to use SEM is that they can explicitly model and partition variance attributed to indirect effects, which is challenging in standard multiple regressions.
 Using SEMs also allowed us to describe latent variables such as `water availability', `soil fertility', and `disturbance' which have been suggested to act upon biodiversity and biomass/productivity in previous studies despite these factors not having directly observable measures in our dataset.
@@
 
-Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (Fig.
-S1, Fig.
-S2).
+Observed variables were transformed to achieve normality where necessary and standardised to Z-scores prior to analysis (Figure S1, Figure S2).
 Standardisation allows path regression coefficients to be easily compared between paths in the same model to assess their relative effect size, and eliminates confusion in model interpretation arising from the observed variables being on different scales \citep{Beaujean2014}.
@@
 We tested the robustness of our assumptions with a chi-squared test of all possible combinations of observed variable factor loadings set to one, while ensuring no factor loadings were in excess of one.
-We found no significant difference between model specifications (p>0.05).
+We found no significant difference between model specifications (p >0.05).
 Full Information Maximum Likelihood (FIML) was used in each model to estimate the values of missing data in each latent variable \citep{Cham2017}.
@@
 
-Preliminary models that included herbivore biomass \citep{Hempson2017} did not converge, possibly due to the spatially coarse nature of the available data, we therefore did not include herbivory in our final model.
+Preliminary models that included herbivore biomass \citep{Hempson2017} did not converge.
+This is possibly due to the spatially coarse nature of the available data, or due to collinearity with other variables, notably MAP and fire frequency.
+We therefore did not include herbivory in our final model.
 We incorporated environmental covariates into our model to understand the relative effects of water availability, soil fertility and disturbance on AGB both directly and indirectly via species diversity and stem density.
-We compared standardised path coefficients between paths in the model to understand the relative contribution of each path to explain variance in AGB.
+We compared standardised path coefficients among paths in the model to understand the relative contribution of each path to explain variance in AGB.
 Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.
@@
 The direct effect of tree species diversity on AGB was positive and marginally significant in ex-Acacia (\strucbetacsb{}) but negligible in Mopane (\strucbetadsb{}), sparse miombo / \textit{Baikiaea} (\strucbetaasb{}) and Core miombo (\strucbetabsb{}) (\autoref{struc_model_slopes_all}).
-Relationships between structural diversity and AGB remained generally similar, with the same sign and overlap between the 95\% confidence intervals of path coefficients.
+Relationships among structural diversity and AGB remained generally similar, with the same sign and overlap between the 95\% confidence intervals of path coefficients.
 The R\textsuperscript{2} of AGB was highest in ex-Acacia shrubland (R\textsuperscript{2} = \struccrsq{}) and lowest in sparse miombo / Baikiaea (R\textsuperscript{2} = \strucarsq{}).
@@
 
-We assessed the importance of a) tree species diversity, b) tree structural diversity, c) resource availability, d) disturbance by fire, e) stem density and their interactions on above-ground woody biomass (AGB) across southern African savannas and woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM).
+We assessed the importance of a) tree species diversity, b) tree structural diversity, c) resource availability, d) disturbance by fire, e) organismal density and their interactions on above-ground woody biomass (AGB) across southern African savannas and woodlands, using a network of \nplots{} woodland plots in conjunction with Structural Equation Modelling (SEM).
 We found support for a general positive relationship between tree species diversity and AGB, operating indirectly via structural diversity (H\textsubscript{1}).
@@
 Within the latent variable of tree species diversity we found similarly strong factor loadings for both species richness and abundance evenness.
-This demonstrates that species richness and abundance evenness measure different and largely uncorrelated aspects of diversity.
+This demonstrates that species richness and abundance evenness measure different and largely uncorrelated axes of diversity.
 We found that the effect of tree species diversity on AGB increased with stem density (H\textsubscript{2}), with an apparent threshold of 180 stems >10 cm DBH ha\textsuperscript{-1}, below which the effect of species diversity on AGB remained at a low baseline level.
 The strongest direct effect on AGB was that of stem density.
-When the effects of water availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments in southern African woodlands.
+When the effects of water availability, soil fertility and disturbance by fire were controlled for, the total explanatory power of tree species diversity and structural diversity decreased, but the predictive power of the model increased, suggesting that it is important to control for environmental covariates to understand the true effect of tree species diversity on AGB in regional scale assessments of the BEFR.
 
@@
 
-Much of the total variation in AGB was driven by variation in stem density.
+Much of the total variation in AGB was driven by variation in organismal density.
 It is possible that within southern African woodlands a higher species diversity allows for a higher stem density through niche separation, which reduces competition between species occupying varying niche space, leading to an increase in total AGB per unit area.
@@
 Disturbance by fire had a negative total effect on AGB, with most of this negative effect coming from the indirect pathway via stem density.
-This is expected as increased fire frequency is a key mechanism by which mesic savannas maintain an open canopy, rather than shifting to a closed canopy forest \citep{Staver2011}.
+This is expected as increased fire frequency is a key mechanism by which savannas maintain an open canopy, rather than shifting to a closed canopy forest \citep{Staver2011}.
 Previous studies have found that southern African woodlands with higher species diversity tend to experience less frequent disturbance by fire and tend to form a more closed canopy with a sparse understorey \citep{Chidumayo2013, Mutowo2012}.

### 2021-04-01 manuscript/regional_befr_manuscript.tex
Edits following comments

@@
 The observed BEFR was highly dependent on organismal density, with a minimum threshold of c.
-180 tree stems ha\textsuperscript{-1}.
+180 mature stems ha\textsuperscript{-1}.
 We found that water availability mainly affects biomass indirectly, via increasing species diversity.}
@@
 
-% BEFR is important to study, setting the scene - DONE
 Understanding the relationship between biodiversity and ecosystem function has become a central endeavour in ecological science, as we seek to predict the consequences of global biodiversity change \citep{Naeem2012}.
@@
 
-% Ecosystem function definition and mechanisms of the BEFR - DONE
 Ecosystem functions are defined in broad terms as rate processes and aggregate properties of ecosystems that describe the nature of biotic activity within those ecosystems \citep{Jax2005}.
@@
 In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.
-Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}.
+Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}, thus weakening the observable BEFR.
+Thus, there is a case that in highly disturbed systems, a relationship between diversity and ecosystem function may not exist at all.
 While these studies together offer some glimpse into how the BEFR may operate in disturbance-prone systems, none provide a full and comprehensive assessment of the interacting effects of diversity, abiotic environment, disturbance and ecosystem function, particularly for wooded ecosystems.
@@
 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
-A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.
+A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temn intermediate erature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.
 
@@
 Plot vegetation was identified under the broad term of `savanna', which includes `woodland', `savanna woodland', and `tree savanna', variously defined in other areas of the scientific literature and here referred to collectively as southern African woodlands, or savannas \citep{Ratnam2011, Hill2010}.
-Plots with evidence of farming, human resource extraction or experimental treatments such as prescribed burning or herbivore exclusion were excluded from the initial pool.
+Plots with evidence of farming, human resource extraction, experimental treatments such as prescribed burning or herbivore exclusion, or containing termite mounds were excluded from the initial pool.
 Only plots >0.1 hectares were used in analyses, as area-based biomass estimation from small plots is highly influenced by rare large trees \citep{Stegen2011}, leading to inaccurate biomass estimates.
@@
 There appears to be a minimum stem density threshold at c.
-180 trees ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass (\autoref{sem_struc_stems_ha}b).
+180 stems >10 cm DBH ha\textsuperscript{-1} below which there appears to be a reasonably constant baseline effect of tree diversity on biomass (\autoref{sem_struc_stems_ha}b).
 The effect of structural diversity on AGB appears to remain constant with increasing stem density (\autoref{sem_struc_stems_ha}d).
@@
 In our model, time since disturbance is accounted for within each plot via the stem density term.
-Disturbance reduces stem density, which is expected to increase until the effects of competition preclude further increase \citep{Johnson2012}.
+Disturbance reduces stem density of large stems (>10 cm DBH), which is expected to increase until the effects of competition preclude further increase \citep{Johnson2012}.
 Furthermore, our rarefied measure of species diversity accounts for variation in sampling effort and is therefore independent of stem density.
@@
 This wide standard error on the model predictions is possibly due to the coarseness and nature of the soil data we used.
-SoilGrids provides modelled data at 250 m resolution, while sostructure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}, notably by termite mounds which offer small areas of highly fertile soil with peculiar tree species diversity \citep{Joseph2014}.
+SoilGrids provides modelled data at 250 m resolution, while soil structure and nutrient content varies at much finer scales in southern African woodlands \citep{Muledi2017, Bucini2007}.
 It is therefore not surprising that this model path is poorly constrained.
@@
 	\includegraphics[width=1.4\textwidth]{clust_map}
-	\caption{The locations of the \nplots{} plots used in this study, with respect to the distribution of miombo woodland vegetation according to \citet{White1987}.
+	\caption{The locations of the \nplots{} plots used in this study, with respect to the distribution of mesic savanna vegetation according to \citet{White1987}.
 Each panel shows plots categorized by their vegetation type as defined by the vegetation types in \autoref{clust_summ}.}
@@
 	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB.
-Hypothesised paths of causation are depicted as arrows from predictor to response.}
+Hypothesised paths of causation are depicted as arrows from predictor to response.
+Open arrow heads track the direction of each arrow along its path.}
 	\label{con_mod}
@@
 Smoothed lines are loess curves with $\pm1$ standard error shaded bars.
-AGB = Above-Ground woody Biomass, arrows in plot titles indicate causal paths in SEM models.}
+AGB = Above-Ground woody Biomass, arrows in plot titles indicate causal paths in SEM models.
+Where multiple arrows are present, as in c), this indicates an indirect pathway via an intermediate variable.}
 	\label{sem_struc_stems_ha}

### 2021-05-27 manuscript/regional_befr_manuscript.tex
Prepping for revisions

@@
 \setlength{\parindent}{0cm}
+\renewcommand{\baselinestretch}{1.5}
 \usepackage{pdflscape}
@@
 \usepackage{xcolor}
-\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}  % \todo{NOTE TO SELF WRITTEN IN RED}
+\newcommand{\TODO}[1]{\textcolor{red}{\textbf{#1}}}  % \TODO{NOTE TO SELF WRITTEN IN RED}
 
@@
 We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
-	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density and vegetation type.}
+	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density at all and vegetation type.}
 	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity.
@@
 1700 mm y\textsuperscript{-1} in northern Zambia, Malawi and northern Mozambique.
-A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temn intermediate erature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.
+A 2D convex hull of Mean Annual Precipitation (MAP) and Mean Annual Temperature (MAT) of the study sites covers \hullcover{}\% of the pixel-wise climate space of the miombo woodland ecoregion \citep{White1987}, using WorldClim estimates of Mean Annual Temperature (MAT, BIO1) and Mean Annual Precipitation (MAP, BIO12) between 1970 and 2000 with a pixel size of 30 arc seconds (926 m at equator) \citep{Fick2017}.
 
@@
 % Correlations
-Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficients of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Fig.
-S3).
+Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficients of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Figure S3).
 Stem density had the strongest correlation with AGB of all variables considered (\ccib{}).
@@
 
-A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, Fig.
-S4).
+A model incorporating the latent variables of water availability, soil fertility and disturbance by fire showed that the total effect of tree species diversity on biomass was similar to that of water availability, soil fertility and disturbance (\autoref{full_mod}, Figure S4).
 The direct effects of water availability, soil fertility and disturbance on AGB were negligible (water: \fmbetamb{}, soil: \fmbetasb{}, disturbance: \fmbetafb{}), with nearly all of their observed effects on AGB coming from the indirect paths via stem density (water: \fmbetamib{}, soil: \fmbetasib{}, disturbance: \fmbetafib{}) and species diversity (water: \fmbetamd{}, soil: \fmbetasd{}, disturbance: \fmbetafd{}).
@@
 
+The data that support the findings of this study are available on request from the corresponding author.
+The data are not publicly available at the time of submission due to privacy restrictions regarding plot locations.
+An anonymised version will be made available in a data repository following review.
+
 \bibliography{regional_befr_manuscript.bib}
@@
 
-\textbf{Fig. S1} Histograms of raw untransformed observed variables used in final analyses.
+\textbf{Figure S1} Histograms of raw untransformed observed variables used in final analyses.
 \vspace{0.2cm}
 
-\textbf{Fig. S2} Histograms of observed variables transformed to achieve a normal frequency distribution.
+\textbf{Figure S2} Histograms of observed variables transformed to achieve a normal frequency distribution.
 \vspace{0.2cm}
 
-\textbf{Fig. S3} Bivariate scatter plots for each observed variable used in SEMs, based on hypothesised paths of causality.
+\textbf{Figure S3} Bivariate scatter plots for each observed variable used in SEMs, based on hypothesised paths of causality.
 \vspace{0.2cm}
 
-\textbf{Fig. S4} Unstandardised path coefficients for full SEM model.
+\textbf{Figure S4} Unstandardised path coefficients for full SEM model.
 \vspace{0.2cm}

### 2021-05-27 manuscript/si/si.tex
Prepping for revisions

@@
+\documentclass[11pt,a4paper]{article}
+
+% Define page geometry
+\usepackage{geometry}
+\geometry{left=2.2cm,
+	right=2.2cm,
+	top=2.2cm,
+	bottom=2cm}
+\parskip 0.15cm
+\setlength{\parindent}{0cm}
+\usepackage{pdflscape}
+\usepackage[document]{ragged2e}
+
+% Set font
+\usepackage[T1]{fontenc}
+
+\usepackage{framed}
+
+% Image handling
+\usepackage{graphics} % Insert images easily
+\usepackage{graphicx} % Extended image support
+
+\makeatletter
+	\g@addto@macro\@floatboxreset\centering % Automatically centre images (floats)
+\makeatother
+
+\graphicspath{ {img/} }
+\usepackage{float} % Graphics placement [H] [H!] arguments
+\usepackage{subfig} % Compound figures
+
+% Tables
+\usepackage{multirow}
+\usepackage{longtable}
+
+% Bibliography management
+\usepackage[UKenglish]{babel}
+\usepackage[natbibapa]{apacite}
+\bibliographystyle{apacite}
+
+% Text formatting
+\usepackage{url} % Allow nice formatting of URLs in text
+
+\usepackage{enumerate} % Enumerated lists
+
+\usepackage{lineno} % Line numbers
+
+\usepackage{textcomp}
+\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}} % Command for a good tilde
+
+\usepackage{siunitx}
+\usepackage{amsmath}
+
+\usepackage{xcolor}
+\newcommand{\TODO}[1]{\textcolor{red}{\textbf{#1}}}  % \TODO{NOTE TO SELF WRITTEN IN RED}
+
+\input{code_format}
+
+\usepackage[breaklinks]{hyperref}
+\definecolor{links}{RGB}{0,0,0}
+\hypersetup{
+	breaklinks,
+	colorlinks=true,
+	linkcolor=links,
+	anchorcolor=links,
+	citecolor=links,
+	filecolor=links,
+	menucolor=links,
+	runcolor=links,
+	urlcolor=links,
+	pdfauthor={John L. Godlee}
+}
+\def\subsectionautorefname{section}
+\def\subsubsectionautorefname{section}
+% \usepackage{biblatex}
+
+\usepackage{dcolumn}
+\usepackage{multirow}
+\usepackage{tikz}
+\usetikzlibrary{calc,arrows,positioning,shapes,shapes.gates.logic.US,trees}
+\usepackage{pdflscape}
+\usepackage{longtable}
+\usepackage{fmtcount}
+
+\linenumbers
+\linespread{1.25}
+
+\newcommand{\titletext}{Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}
+
+\begin{document}
+
+{\Large{Title: \titletext{}}}
+
 \section{Appendix 1 - Frequency distribution of observed variables} \label{appendixa}
@@
 \end{figure}
+
+\end{document}

### 2021-05-27 manuscript/title_page/title_page.tex
Prepping for revisions

@@
 \usepackage{xcolor}
-\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}   % \todo{NOTE TO SELF WRITTEN IN RED}
+\newcommand{\TODO}[1]{\textcolor{red}{\textbf{#1}}}   % \TODO{NOTE TO SELF WRITTEN IN RED}
 
@@
 \begin{tabular}{r|r}
-Main body word count & 6384\\
-Summary & 193\\
-Introduction & 1490\\
-Materials and Methods & 1999\\
-Results & 726\\
-Discussion & 2216\\
+Main body word count & 6374\\
+Summary & 200\\
+Introduction & 1116 \\
+Materials and Methods & 2082\\
+Results & 759\\
+Discussion & 2089\\
+Conclusion & 328\\
 Acknowledgements & 122\\

### 2021-06-21 manuscript/regional_befr_manuscript.tex
Finished revisions

@@
 \usepackage{xcolor}
-\newcommand{\TODO}[1]{\textcolor{red}{\textbf{#1}}}  % \TODO{NOTE TO SELF WRITTEN IN RED}
+\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}  % \todo{NOTE TO SELF WRITTEN IN RED}
 
@@
 
+\usepackage[revision]{revdiff}
+
 \newcommand{\titletext}{Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}
@@
 We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
-	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density at all and vegetation type.}
+	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density \rold{at all }and vegetation type.}
 	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity.
@@
 In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity via its effect on functional trait values.
-In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.
-Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}, thus weakening the observable BEFR.
+\rold{In contrast however, }\citet{Carvalho2014}\rold{ also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.}\rnew{Furthermore, }\citet{Carvalho2014}\rnew{ found that disturbance by fire in the Brazilian Cerrado reduced soil fertility, causing an indirect effect of fire frequency on functional trait diversity via soil fertility.} Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}, thus weakening the observable BEFR.
 Thus, there is a case that in highly disturbed systems, a relationship between diversity and ecosystem function may not exist at all.
@@
 Relatedly, we expect that an increase in disturbance by fire will decrease organismal density and therefore competition, weakening the effect of tree species diversity on woody biomass.
-Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in functional form (i.e.
+Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in \rchange{functional form}{size strategy} (i.e.
 physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.
@@
 A tree may be comprised of multiple stems and so tree-level richness estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice.
-For each tree, we recorded species, DBH and tree height to the top of the highest branch material.
+For each tree, we recorded species, \rold{DBH and }tree height to the top of the highest branch material, \rnew{and the DBH of each stem >10 cm DBH}.
 Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks.
-When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
-The height of this measurement was recorded and used to estimate the D\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.
+When \rnew{stem }DBH could not be reliably measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
+The height of this measurement was used to estimate diameter at 1.3 m using a cubic polynomial regression to account for stem taper, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.
 
 % AGB Calculation
-AGB for each plot (t ha\textsuperscript{-1}) was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:
+AGB for each plot (t ha\textsuperscript{-1}) was calculated \rnew{as the sum of the AGB of each stem >5 cm DBH ($AGB_{i}$), divided by the plot area,} using \autoref{chave_agb} taken from \citet{Chave2014}:
 
 \begin{equation}
-	AGB = 0.0673 \times (\rho D_{e}^{2} H)^{0.976}
+	AGB_{i} = 0.0673 \times (\rho D^{2} H)^{0.976} \times 1000
 	\label{chave_agb}
@@
 
-where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), $D_{e}$ is the estimated DBH (cm) at 1.3 m, and $H$ is the tree height (m).
+where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), \rchange{$D_{e}$ is the estimated DBH (cm) at 1.3 m}{$D$ is the stem diameter (cm) measured or estimated at 1.3 m}, and $H$ is the tree height (m).
 Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
@@
 
-We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
-Structural diversity can also directly influence AGB in this model, separate to the effect of of species diversity.
-To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model.
-To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately.
+\rchange{We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
+Structural diversity can also directly influence AGB in this model, separate to the effect \rchange{of of}{of} species diversity.
+To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model.}{First, we used a simple mediation model which excluded the environmental covariates, to assess the role of tree species diversity and tree structural diversity in determining AGB.
+This model allowed direct effects of species diversity, structural diversity, and stem density on AGB, and also the indirect effect of species diversity on AGB via structural diversity.} To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately.
 We compared unstandardised path coefficients among the models for different vegetation types to understand the effect that vegetation type has on the relationship between tree species diversity, structural diversity, stem density and AGB.
@@
 
-Preliminary models that included herbivore biomass \citep{Hempson2017} did not converge.
+\rold{Preliminary models that included herbivore biomass }\citep{Hempson2017} \rold{did not converge.
 This is possibly due to the spatially coarse nature of the available data, or due to collinearity with other variables, notably MAP and fire frequency.
@@
 We compared standardised path coefficients among paths in the model to understand the relative contribution of each path to explain variance in AGB.
-Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.
+Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.}\rnew{Second, we fitted the full model with environmental covariates, to understand the relative effects of water availability, soil fertility and disturbance on AGB, both directly and indirectly via species diversity and stem density.
+We compared standardised path coefficients among paths in the model to understand the relative contribution of each path to explain variance in AGB.
+Due to sample size issues, and because some vegetation types were narrow in their climate space, particularly in the water availability latent variable, we could not fit the model including environmental covariates separately for each vegetation type, as we encountered issues with model convergence.
+Preliminary models that included herbivore biomass }\citep{Hempson2017} \rnew{did not converge.
+This is possibly due to the spatially coarse nature of the available data, or to collinearity with other variables, notably MAP and fire frequency.
+We therefore did not include herbivory in our final model.}
 
@@
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effects of tree structural diversity and stem density (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}).
+\rchange{In an SEM describing the effect of tree species diversity on AGB via the mediating effects of tree structural diversity and stem density}{In the reduced SEM, which included stem density and the mediating effect of species diversity on AGB via structural diversity} (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}).
 Model fit was good with high factor loadings for all observed variables.
@@
 This may act as a mechanism for niche complementarity, with a canopy of diversely-sized trees able to take advantage of a greater proportion of the available light.
+\rnew{Additionally, the volume of tree above-ground structures is generally correlated with the volume of below-ground structures \citep{Paul2019}.
+In water and nutrient limited ecosystems especially, variation in rooting depth may constitute a second related axis of niche partitioning driving the observed positive effect of above-ground structural diversity on AGB \citep{Kulmatiski2013}}.
 Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{Pretzsch2014}, i.e.
@@
 
+\rnew{Alternatively, due to the non-linear relationship between biomass and tree size \citep{Bastin2018}, the positive relationship between structural diversity and biomass may also be partly driven by an increased number of large sized trees in plots with higher structural diversity, with large trees contributing disproportionately to biomass.
+The positive effect of species diversity on AGB via structural diversity may therefore be due to selection effects, with higher diversity plots supporting larger trees due to species specific variation in functional form \citep{Diaz2015}.}
+
 \subsection{Organismal density and disturbance}
@@
 
-As time since disturbance increases in any wooded ecosystem, it is expected that both species diversity and above-ground biomass will increase until an equilibrium is reached or another disturbance event occurs \citep{Huston2014}.
-Unless this effect is accounted for, there is the potential for mistaken causality as both diversity and biomass will correlate.
+Disturbances such as fire have the potential to reduce both species diversity and above-ground biomass in the short term, due to increased mortality \citep{Huston2014}.
+Unless this effect is accounted for, there is the potential for mistaken causality as both diversity and biomass may correlate.
 In our model, time since disturbance is accounted for within each plot via the stem density term.
@@
 Our study points to the need for further effort in this regard, which may reveal interesting findings about the complex interactions between soil, disturbance and tree diversity in southern African woodlands.
+Alternatively, \citet{Gourlet2011} found that environmental filtering of fast-growing species with low wood density on resource poor soils resulted in a decoupling of the soil fertility - AGB relationship.
+It is possible that at regional scales, variation in species composition could offset resource availability constraints on AGB.
+However, unlike \citet{Gourlet2011} disturbance by fire in our study region may further complicate this environmental filtering effect.
 
@@
 We assert that this is necessary to develop our understanding of the biodiversity-ecosystem function relationship in real-world ecosystems, to progress from experimental mesocosms.
-We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key mechanism by which tree species diversity determines ecosystem function in savannas, woodlands and forests, i.e.
-all wooded ecosystems.
+We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key mechanism by which tree species diversity determines ecosystem function in savannas, woodlands and forests, \rchange{i.e.
+all wooded ecosystems}{where trees comprise a significant, canopy-forming component}.
 The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in previous studies despite its intuitive logic as a determinant of niche complementarity effects in wooded ecosystems.
-Our study shows that biodiversity change through extensive human actions in this region will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.
-This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.
+\rchange{Our study shows that biodiversity change through extensive human actions in this region will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.}{Our study shows that biodiversity change through extensive human-induced land use change in this region will have the greatest negative impact on ecosystem function in areas of high stems density, and in certain vegetation types, specifically Mopane and ex-Acacia woodlands.} This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.
 Finally, our results provide further evidence of the complex interaction of factors governing biomass and therefore carbon dynamics in disturbance-driven wooded ecosystems, which currently represent the greatest uncertainty in the global terrestrial carbon sink.

### 2021-06-23 manuscript/regional_befr_manuscript.tex
Submitted revisions

--- 2021-05-27 manuscript/regional_befr_manuscript.tex
@@
-\documentclass[11pt,a4paper]{article}
+\documentclass[12pt,a4paper]{article}
 
@@
 \usepackage{framed}
+
+\usepackage{setspace}
+\doublespacing
 
@@
 
+\usepackage[clean]{revdiff}
+
 \newcommand{\titletext}{Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}
@@
 We provide the first exploration of the relationship between tree species diversity and biomass, one measure of ecosystem function, across southern African woodlands and savannas, an ecological system rife with disturbance from fire, herbivores and humans.}
-	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density at all and vegetation type.}
+	\item{We used >1000 vegetation plots distributed across 10 southern African countries, and structural equation modelling, to determine the relationship between tree species diversity and aboveground woody biomass, accounting for interacting effects of resource availability, disturbance by fire, tree stem density and vegetation type.}
 	\item{We found positive effects of tree species diversity on aboveground biomass, operating via increased structural diversity.
@@
 In the Brazilian Cerrado savanna, \citet{Loiola2015} found that disturbance by fire reduced tree productivity via its effect on functional trait values.
-In contrast however, \citet{Carvalho2014} also working in the Brazilian Cerrado, found that while disturbance by fire did reduce soil fertility, which is often closely related to productivity in savannas, functional trait diversity was unaffected by fire frequency.
+Furthermore, \citet{Carvalho2014} found that disturbance by fire in the Brazilian Cerrado reduced soil fertility, causing an indirect effect of fire frequency on functional trait diversity via soil fertility.
 Other studies focussing on the herbaceous diversity in disturbance-prone grasslands in North America and Europe have shown that disturbance by fire, mowing and herbivory reduces herbaceous productivity \citep{Grace2007}, allowing weak competitors to co-exist where otherwise they would be excluded \citep{Mason2011}, thus weakening the observable BEFR.
@@
 Relatedly, we expect that an increase in disturbance by fire will decrease organismal density and therefore competition, weakening the effect of tree species diversity on woody biomass.
-Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in functional form (i.e.
+Finally, we expect that (3) tree species diversity will increase tree structural diversity due to inter-specific variation in size strategy (i.e.
 physiognomic diversity), providing an indirect path by which tree diversity increases woody biomass.
@@
 A tree may be comprised of multiple stems and so tree-level richness estimates, rather than stem-level estimates, were used to prevent bias from species which readily coppice.
-For each tree, we recorded species, DBH and tree height to the top of the highest branch material.
+For each tree, we recorded species, tree height to the top of the highest branch material, and the DBH of each stem >10 cm DBH.
 Height was measured through a variety of means including laser rangefinders, manual clinometers and measuring sticks.
-When DBH could not be measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
-The height of this measurement was recorded and used to estimate the D\textsubscript{e} at 1.3 m using a cubic polynomial regression, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.
+When stem DBH could not be reliably measured at 1.3 m due to trunk abnormalities, it was measured at the closest regular portion of the trunk to 1.3 m.
+The height of this measurement was used to estimate diameter at 1.3 m using a cubic polynomial regression to account for stem taper, with parameters estimated using a test dataset from Ryan C., (unpublished), see \citet{Godlee2020}.
 
 % AGB Calculation
-AGB for each plot (t ha\textsuperscript{-1}) was calculated using \autoref{chave_agb}, taken from \citet{Chave2014}:
+AGB for each plot (t ha\textsuperscript{-1}) was calculated as the sum of the AGB of each stem >5 cm DBH ($AGB_{i}$), divided by the plot area, using \autoref{chave_agb} taken from \citet{Chave2014}:
 
 \begin{equation}
-	AGB = 0.0673 \times (\rho D_{e}^{2} H)^{0.976}
+	AGB_{i} = 0.0673 \times (\rho D^{2} H)^{0.976} \times 1000
 	\label{chave_agb}
@@
 
-where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), $D_{e}$ is the estimated DBH (cm) at 1.3 m, and $H$ is the tree height (m).
+where $\rho$ is the species mean wood density (g cm\textsuperscript{-3}), $D$ is the stem diameter (cm) measured or estimated at 1.3 m, and $H$ is the tree height (m).
 Wood density estimates were taken from the global wood density database for each species where possible \citep{Chave2009, Zanne2009}.
@@
 
-We assessed the role of tree species diversity and tree structural diversity in determining AGB via a simple mediation model which allowed species diversity to influence AGB both directly and indirectly via structural diversity.
-Structural diversity can also directly influence AGB in this model, separate to the effect of of species diversity.
-To account for variation in stem density, which may covary with species diversity, we included it as an observed variable in our model.
+First, we used a simple mediation model which excluded the environmental covariates, to assess the role of tree species diversity and tree structural diversity in determining AGB.
+This model allowed direct effects of species diversity, structural diversity, and stem density on AGB, and also the indirect effect of species diversity on AGB via structural diversity.
 To explore variation in the model among woodland vegetation types, we fit the model both at the regional scale and for each vegetation type separately.
@@
 
+Second, we fitted the full model with environmental covariates, to understand the relative effects of water availability, soil fertility and disturbance on AGB, both directly and indirectly via species diversity and stem density.
+We compared standardised path coefficients among paths in the model to understand the relative contribution of each path to explain variance in AGB.
+Due to sample size issues, and because some vegetation types were narrow in their climate space, particularly in the water availability latent variable, we could not fit the model including environmental covariates separately for each vegetation type, as we encountered issues with model convergence.
 Preliminary models that included herbivore biomass \citep{Hempson2017} did not converge.
-This is possibly due to the spatially coarse nature of the available data, or due to collinearity with other variables, notably MAP and fire frequency.
+This is possibly due to the spatially coarse nature of the available data, or to collinearity with other variables, notably MAP and fire frequency.
 We therefore did not include herbivory in our final model.
-We incorporated environmental covariates into our model to understand the relative effects of water availability, soil fertility and disturbance on AGB both directly and indirectly via species diversity and stem density.
-We compared standardised path coefficients among paths in the model to understand the relative contribution of each path to explain variance in AGB.
-Vegetation type specific models could not be reliably fitted for this more complex model specification with environmental covariates, due to sample size issues and because some vegetation types were narrow in their climate space, leading to a lack of environmental variation, particularly in the water availability latent variable.
 
@@
 
-In an SEM describing the effect of tree species diversity on AGB via the mediating effects of tree structural diversity and stem density (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}).
+In the reduced SEM, which included stem density and the mediating effect of species diversity on AGB via structural diversity (\autoref{struc_mod}), species diversity showed no direct effect on AGB (\strucbetadb{}), but did have an indirect positive effect via structural diversity (\strucbetadib{}) (\autoref{struc_mod}).
 Model fit was good with high factor loadings for all observed variables.
@@
 This may act as a mechanism for niche complementarity, with a canopy of diversely-sized trees able to take advantage of a greater proportion of the available light.
+Additionally, the volume of tree above-ground structures is generally correlated with the volume of below-ground structures \citep{Paul2019}.
+In water and nutrient limited ecosystems especially, variation in rooting depth may constitute a second related axis of niche partitioning driving the observed positive effect of above-ground structural diversity on AGB \citep{Kulmatiski2013}.
 Although we did not measure them here, we would also expect that tree species diversity allows for a greater range of tree functional forms \citep{Pretzsch2014}, i.e.
@@
 
+Alternatively, due to the non-linear relationship between biomass and tree size \citep{Bastin2018}, the positive relationship between structural diversity and biomass may also be partly driven by an increased number of large sized trees in plots with higher structural diversity, with large trees contributing disproportionately to biomass.
+The positive effect of species diversity on AGB via structural diversity may therefore be due to selection effects, with higher diversity plots supporting larger trees due to species specific variation in functional form \citep{Diaz2015}.
+
 \subsection{Organismal density and disturbance}
@@
 We assert that this is necessary to develop our understanding of the biodiversity-ecosystem function relationship in real-world ecosystems, to progress from experimental mesocosms.
-We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key mechanism by which tree species diversity determines ecosystem function in savannas, woodlands and forests, i.e.
-all wooded ecosystems.
+We found that much of the effect of species diversity on biomass exists as an indirect effect by increasing the structural diversity of trees, exemplifying a key mechanism by which tree species diversity determines ecosystem function in savannas, woodlands and forests, where trees comprise a significant, canopy-forming component.
 The presence of a stem density threshold above which the effect of tree species diversity on AGB increases clearly implies the presence of niche complementarity effects in southern African woodlands, an aspect which has often been overlooked in previous studies despite its intuitive logic as a determinant of niche complementarity effects in wooded ecosystems.
-Our study shows that biodiversity change through extensive human actions in this region will have the greatest negative impact on ecosystem function in areas of high stem density, and low species diversity, which are those areas predominantly targeted for tree felling.
+Our study shows that biodiversity change through extensive human-induced land use change in this region will have the greatest negative impact on ecosystem function in areas of high stems density, and in certain vegetation types, specifically Mopane and ex-Acacia woodlands.
 This raises concerns about the robustness of these ecosystems to further resource extraction and biodiversity loss.

### 2021-06-23 manuscript/title_page/title_page.tex
Submitted revisions

@@
 \begin{tabular}{r|r}
-Main body word count & 6374\\
-Summary & 200\\
-Introduction & 1116 \\
-Materials and Methods & 2082\\
-Results & 759\\
-Discussion & 2089\\
-Conclusion & 328\\
+Main body word count & 6706\\
+Summary & 198\\
+Introduction & 1107\\
+Materials and Methods & 2089\\
+Results & 756\\
+Discussion & 2226\\
+Conclusion & 330\\
 Acknowledgements & 122\\

### 2021-07-09 manuscript/include/clust_summ.tex
Finished proofing

@@
 Indicator species were generated using Dufrene-Legendre indicator species analysis \citep{Dufrene1997} implemented with \texttt{indval()} from the \texttt{labdsv} R package \citep{labdsv} and represent species which define the given cluster.
-Dominant species were identified by choosing the species with the largest mean plot level proportional AGB within each cluster.
+Dominant species were identified by choosing the species with the largest mean plot level proportional AGB (Above-Ground woody Biomass) within each cluster.
+N = number of plots in cluster.
 Numeric values of species richness, stems ha\textsuperscript{-1} and AGB represent medians and interquartile ranges (75th percentile - 25th percentile).}
@@
 \hline \\[-1.8ex] 
-{Cluster} & {Dominant species} & {Indicator species} & {n} & \multicolumn{1}{p{2cm}}{\centering Species\\Richness} & \multicolumn{1}{p{2.5cm}}{\centering Stem density\\(stems ha\textsuperscript{-1})} & \multicolumn{1}{p{2cm}}{\centering AGB\\(t ha\textsuperscript{-1})} \\
+{Cluster} & {Dominant species} & {Indicator species} & {N} & \multicolumn{1}{p{2cm}}{\centering Species\\Richness} & \multicolumn{1}{p{2.5cm}}{\centering Stem density\\(stems ha\textsuperscript{-1})} & \multicolumn{1}{p{2cm}}{\centering AGB\\(t ha\textsuperscript{-1})} \\
 \hline \\[-1.8ex] 

### 2021-07-09 manuscript/include/struc_model_fit_clust_stats.tex
Finished proofing

@@
 \begin{table}[!htbp] \centering 
-  	\caption{Model fit statistics for SEMs investigating the effects of tree diversity and stem density on AGB (\autoref{struc_mod}).
-n = number of plots in cluster, $\chi^{2}$ = Chi-squared fit statistic, DoF = model degrees of freedom, CFI = Comparative Fit Index, TLI = Tucker-Lewis Index, RMSEA = Root Mean Square Error of Approximation, $R^{2}$ AGB = R-squared of AGB (Above-Ground Biomass).}
+  	\caption{Model fit statistics for Structural Equation Models investigating the effects of tree diversity and stem density on AGB (\autoref{struc_mod}).
+N = number of plots in cluster, $\chi^{2}$ = Chi-squared fit statistic, DoF = model degrees of freedom, CFI = Comparative Fit Index, TLI = Tucker-Lewis Index, RMSEA = Root Mean Square Error of Approximation, $R^{2}$ AGB = R-squared of AGB (Above-Ground Biomass).}
   \label{struc_model_fit_clust_stats} 
@@
 \hline \\[-1.8ex] 
-{Cluster} & {n} & {$\chi^{2}$} & {DoF} & {CFI} & {TLI} & {RMSEA} & {$R^{2}$ AGB} \\
+{Cluster} & {N} & {$\chi^{2}$} & {DoF} & {CFI} & {TLI} & {RMSEA} & {$R^{2}$ AGB} \\
 \hline \\[-1.8ex] 

### 2021-07-09 manuscript/regional_befr_manuscript.tex
Finished proofing

@@
 
-% Define page geometry
+% Page geometry
 \usepackage{geometry}
@@
 	bottom=2cm}
-\parskip 0.15cm
+\parskip 0.2cm
 \setlength{\parindent}{0cm}
-\renewcommand{\baselinestretch}{1.5}
 \usepackage{pdflscape}
-\usepackage[document]{ragged2e}
+\usepackage{framed}
+\usepackage{lineno} 
+\linenumbers
 
-% Set font
+% Text 
 \usepackage[T1]{fontenc}
+\linespread{1.25}
+\usepackage{siunitx}
 
-\usepackage{framed}
-
-\usepackage{setspace}
-\doublespacing
-
-% Image handling
+% Images 
 \usepackage{graphics} % Insert images easily
 \usepackage{graphicx} % Extended image support
+\usepackage{float} % Graphics placement [H] [H!] arguments
 
@@
 \graphicspath{ {img/} }
-\usepackage{float} % Graphics placement [H] [H!] arguments
-\usepackage{subfig} % Compound figures
 
-% Tables
-\usepackage{multirow}
-\usepackage{longtable}
+% Bibliography 
+\usepackage[UKenglish]{babel}
+\usepackage{csquotes}
+\usepackage[natbib, 
+	style=authoryear, 
+	uniquename=false, 
+	uniquelist=false,
+	maxcitenames=2, 
+	mincitenames=1, 
+	minbibnames=10, 
+	maxbibnames=10, 
+	backend=biber]{biblatex}
+\renewcommand*\finalnamedelim{\addspace\&\space}
+\addbibresource{regional_befr_manuscript.bib}
 
-% Bibliography management
-\usepackage[UKenglish]{babel}
-\usepackage[natbibapa]{apacite}
-\bibliographystyle{apacite}
-
-% Text formatting
-\usepackage{url} % Allow nice formatting of URLs in text
-
-\usepackage{enumerate} % Enumerated lists
-
-\usepackage{lineno} % Line numbers
-
-\usepackage{textcomp}
-\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}} % Command for a good tilde
-
-\usepackage{siunitx}
-\usepackage{amsmath}
-
+% Cross references and hyperlinks
 \usepackage{xcolor}
-\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}  % \todo{NOTE TO SELF WRITTEN IN RED}
-
-\input{code_format}
-
+\usepackage{url} 
 \usepackage[breaklinks]{hyperref}
@@
 \def\subsubsectionautorefname{section}
-% \usepackage{biblatex}
 
-\usepackage{dcolumn}
-\usepackage{multirow}
-\usepackage{tikz}
-\usetikzlibrary{calc,arrows,positioning,shapes,shapes.gates.logic.US,trees}
-\usepackage{pdflscape}
-\usepackage{longtable}
-\usepackage{fmtcount}
-
-\linenumbers
-\linespread{1.25}
-
-\usepackage[clean]{revdiff}
-
-\newcommand{\titletext}{Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}
-
+% External files
 \input{include/n_plots}
@@
 
-{\Large{Title: \titletext{}}}
+{\textbf{Title: Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}}
 
@@
 \includegraphics[width=\textwidth]{concept.drawio}
-	\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB.
+\caption{Conceptual Directed Acyclic Graph (DAG) showing the theoretical relationships between environmental factors, tree species diversity, tree structural diversity, stem density, and AGB (Above-Ground Woody Biomass).
 Hypothesised paths of causation are depicted as arrows from predictor to response.
@@
 	\includegraphics[width=0.6\textwidth]{corr_mat}
-	\caption{Correlation matrix of standardised observed variables used in the SEMs, with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
+	\caption{Correlation matrix of standardised observed variables used in the SEMs (Structural Equation Models), with Pearson correlation coefficients ($r$) coloured according to sign ($+$ve red, $-$ve blue) and shaded by strength of correlation.
 Correlation coefficients marked by a circle indicate that the 95\% confidence interval of $r$ overlapped zero.
@@
 	\includegraphics[width=0.5\textwidth]{struc.drawio}
-	\caption{Path diagram with regression coefficients for the tree diversity SEM, including plots from all vegetation clusters.
+	\caption{Path diagram with regression coefficients for the tree diversity SEM (Structural Equation Model), including plots from all vegetation clusters.
 Latent variables are shown as circles while observed variables are shown as rectangles.
@@
 	\includegraphics[width=\textwidth]{struc_model_slopes_all}
-	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB, mediated by the effect of stand structural diversity.
+	\caption{Unstandardised path coefficients for the effects of tree diversity on AGB (Above-Ground Woody Biomass), mediated by the effect of stand structural diversity.
 Path coefficients are $\pm$1 standard error.
@@
 	\includegraphics[width=0.8\textwidth]{sem_struc_stems_ha}
-	\caption{Line plots showing the variation in SEM path coefficients across datasets with different mean stem density.
+	\caption{Line plots showing the variation in SEM (Structural Equation Model) path coefficients among latent variables, across datasets with different mean stem density.
 Smoothed lines are loess curves with $\pm1$ standard error shaded bars.
-AGB = Above-Ground woody Biomass, arrows in plot titles indicate causal paths in SEM models.
-Where multiple arrows are present, as in c), this indicates an indirect pathway via an intermediate variable.}
+AGB = Above-Ground woody Biomass.
+Arrows in plot titles indicate causal paths in SEM models.
+Where multiple arrows are present, as in c), this indicates an indirect pathway via an intermediate variable.
+a) shows the direct effect of species diversity on structural diversity.
+b) and d) show the direct effects of species diversity and structural diversity on AGB, respectively.
+c) shows the indirect effect of species diversity on AGB via structural diversity.
+e) shows the total effect of species diversity on AGB, incorporating both the direct effect and the indrect effect via structural diversity.}
 	\label{sem_struc_stems_ha}
@@
 	\includegraphics[width=\textwidth]{full.drawio}
-	\caption{Path diagram with regression coefficients for the SEM incorporating environmental covariates and tree species and structural diversity across all five vegetation types.
+	\caption{Path diagram with regression coefficients for the SEM (Structural Equation Model) incorporating environmental covariates and tree species and structural diversity across all five vegetation types.
 Latent variables are shown as circles while observed variables are shown as rectangles.
@@
 Revisions of the SEOSAW dataset were funded by SavannaChange, a GCRF/University of Edinburgh funded project.
-We thank all data providers and the field assistance they received when collecting plot data.
+We thank all data providers and the field assistance they received in collecting plot data.
 JMBC was supported by the Natural Environment Research Council (Agreement PR140015 between NERC and the National Centre for Earth Observation).
@@
 
-JG and KD conceived the study.
-JG conducted data analysis, data management for further versions of the SEOSAW dataset, and wrote the manuscript.
-CR conceived the SEOSAW database and conducted data management for earlier versions of the SEOSAW dataset.
-JG, CR, DB, JMBC, MF, RH, EM, SS, HT, HT, MB, MW, and KD contributed to manuscript revisions.
-JG, CR, SB, VC, JPGMC, DD, MF, FG, SM, IM, AM, RR, NR, AS, SS, JT, JW, MB, and MW contributed to experimental design, field data collection, data preparation and data management of parts of the dataset used in this study.
+JLG and KGD conceived the study.
+JLG conducted data analysis, data management for further versions of the SEOSAW dataset, and wrote the manuscript.
+CMR conceived the SEOSAW database and conducted data management for earlier versions of the SEOSAW dataset.
+JLG, CMR, DB, ETAM, HGT, JMBC, KGD, MF, MW, MtB, RMH and SS contributed to manuscript revisions.
+JLG, AM, AS, AVC, CMR, DJD, FMG, IM, JJT, JPGMC, JW, MF, MtB, MW, NSR, RR, SJB, SM and SS contributed to experimental design, field data collection, data preparation and data management of parts of the dataset used in this study.
 
@@
 
-The data that support the findings of this study are available on request from the corresponding author.
-The data are not publicly available at the time of submission due to privacy restrictions regarding plot locations.
-An anonymised version will be made available in a data repository following review.
+An anonymised version of the data that support the findings of this study are available at this DOI: .
 
-\bibliography{regional_befr_manuscript.bib}
+\printbibliography{}
 

### 2021-07-09 manuscript/si/si.tex
Finished proofing

--- 2021-01-25 manuscript/si/si.tex
@@
-\section{Appendix 1 - Frequency distribution of observed variables} \label{appendixa}
+\documentclass[11pt,a4paper]{article}
+
+% Define page geometry
+\usepackage{geometry}
+\geometry{left=2.2cm,
+	right=2.2cm,
+	top=2.2cm,
+	bottom=2cm}
+\parskip 0.2cm
+\setlength{\parindent}{0cm}
+
+% Set font
+\usepackage[T1]{fontenc}
+
+% Image handling
+\usepackage{graphics} % Insert images easily
+\usepackage{graphicx} % Extended image support
+\usepackage{float} % Graphics placement [H] [H!] arguments
+
+\makeatletter
+	\g@addto@macro\@floatboxreset\centering % Automatically centre images (floats)
+\makeatother
+
+\graphicspath{ {img/} }
+
+% Tables
+\usepackage{longtable}
+
+% Figure and table captions
+\usepackage{caption}
+\captionsetup[figure]{name={Fig.},labelsep=period}
+\renewcommand{\thefigure}{S\arabic{figure}}
+\renewcommand{\thetable}{S\arabic{table}}
+
+\begin{document}
+
+{\Large{New Phytologist Supporting Information}}
+
+{\textbf{Title: Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}}
+
+John L.
+Godlee, Casey M.
+Ryan, David Bauman, Samuel J.
+Bowers, Joao M.
+B.
+Carreiras, Antonio Valter Chisingui, Joris P.
+G.
+M.
+Cromsigt, Dave J.
+Druce, Manfred Finckh, Francisco Maiato Gon\c{c}alves, Ricardo M.
+Holdo, Steve Makungwa, Iain M.
+McNicol, Edward T.
+A.
+Mitchard, Anderson Muchawona, Rasmus Revermann, Natasha Sofia Ribeiro, Abel Siampale, Stephen Syampungani, Jos\'{e} Jo\~{a}o Tchamba, Hemant G.
+Tripathi, Johannes Wallenfang, Mariska te Beest, Mathew Williams, Kyle G.
+Dexter
+
+Article acceptance date: 1st July 2021
+
+\section*{Figs S1 and S2 - Frequency distribution of observed variables} \label{appendixa}
 
@@
 \appendix{}
-\section{Appendix 2 - Table of correlation fit statistics} \label{appendixb}
+\section*{Table S1 - Table of correlation fit statistics} \label{appendixb}
 
@@
 
-\section{Appendix 3 - Bivariate relationships of model variables} \label{appendixc}
+\section*{Fig. S3 - Bivariate relationships of model variables} \label{appendixc}
 
@@
 	\includegraphics[width=0.9\textwidth]{bivar_lm}
-	\caption{Bivariate scatter plots for each observed variable used in the SEMs, based on hypothesised paths of causality.
-Points are coloured according to vegetation type.
-A single linear regression is presented as a black line, which combines all vegetation types, separate loess trend lines are fitted for each vegetation type.
+	\caption{Bivariate scatter plots for each observed variable used in the SEMs (Structural Equation Models), based on hypothesised paths of causality.
+Points are coloured according to vegetation type: green = Sparse miombo/Baikiaea, purple = Core miombo, blue = ex-Acacia, red = Mopane.
+The black line combines all vegetation types in a single linear regression, while loess trend lines are fitted for each vegetation type, separately.
 An orange loess trend line is fitted for all the data.
-All data is standardised and variables are transformed where it was appropriate for analysis.}
+All data is standardised to a mean of zero and a standard deviation of one.
+Variables are transformed where it was appropriate for analysis.}
 	\label{bivar_lm}
@@
 
-\section{Appendix 4 - Path coefficients for model incorporating environmental covariates} \label{appendixd}
+\section*{Fig. S4 - Path coefficients for model incorporating environmental covariates} \label{appendixd}
 

### 2021-07-09 manuscript/title_page/title_page.tex
Finished proofing

--- 2021-01-25 manuscript/title_page/title_page.tex
@@
 	bottom=2cm}
-\parskip 0.15cm
+\parskip 0.2cm
 \setlength{\parindent}{0cm}
-\usepackage{pdflscape}
-\usepackage[document]{ragged2e}
 
@@
 
-% Image handling
-\usepackage{graphics}  % Insert images easily
-\usepackage{graphicx}  % Extended image support
-
-\makeatletter
-	\g@addto@macro\@floatboxreset\centering  % Automatically centre images (floats)
-\makeatother
-
-\graphicspath{ {img/} }
-\usepackage{float}  %  Graphics placement [H] [H!] arguments
-\usepackage{subfig}  % Compound figures
-
-% Tables
-\usepackage{multirow}
-\usepackage{longtable}
-
-% Bibliography management
-\usepackage[UKenglish]{babel}
-\usepackage[natbibapa]{apacite}
-\bibliographystyle{apacite}
-
-% Text formatting
-\usepackage{url} % Allow nice formatting of URLs in text
-
-\usepackage{enumerate}  % Enumerated lists
-
-%\usepackage{lineno}  % Line numbers
-
-\usepackage{textcomp}
-\newcommand{\textapprox}{\raisebox{0.5ex}{\texttildelow}}  % Command for a good tilde
-
-\usepackage{siunitx}
-\usepackage{amsmath}
-
-\usepackage{xcolor}
-\newcommand{\todo}[1]{\textcolor{red}{\textbf{#1}}}   % \todo{NOTE TO SELF WRITTEN IN RED}
-
-\input{code_format}
-
-% Custom title formatting
-\let\oldtitle\title
-
-\renewcommand{\title}[1]{\oldtitle{\vspace{-1.5cm}#1}}
-
-\usepackage[breaklinks]{hyperref}
-\definecolor{links}{RGB}{191,59,72}
-\hypersetup{
-	breaklinks,
-	colorlinks,
-	allcolors=links,
-	linktoc=section,
-	pdfauthor={John L. Godlee}
-}
-\def\subsectionautorefname{section}
-\def\subsubsectionautorefname{section}
-% \usepackage{biblatex}
-
-\usepackage{dcolumn}
-\usepackage{multirow}
-\usepackage{tikz}
-\usetikzlibrary{calc,arrows,positioning,shapes,shapes.gates.logic.US,trees}
-\usepackage{pdflscape}
-\usepackage{longtable}
-\usepackage{fmtcount}
-
 \begin{document}
 
-{\textbf{Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}}
-
-\vspace{1cm}
+{\textbf{Title: Structural diversity and tree density drives variation in the biodiversity-ecosystem function relationship of woodlands and savannas}}
 
@@
 Cromsigt\textsuperscript{6,7,8}, Dave J.
-Druce\textsuperscript{9,10}, Manfred Finkch\textsuperscript{11}, Francisco Maiato Gon\c{c}alves\textsuperscript{5}, Ricardo M.
+Druce\textsuperscript{9,10}, Manfred Finckh\textsuperscript{11}, Francisco Maiato Gon\c{c}alves\textsuperscript{5}, Ricardo M.
 Holdo\textsuperscript{12}, Steve Makungwa\textsuperscript{13}, Iain M.
@@
 
+\section*{Affiliations}
+
 1: School of GeoSciences, University of Edinburgh, Edinburgh, EH9 3FF, United Kingdom \\
-2: Environmental Change Institute, School of Geography and the Environment, University of Oxford, Oxford, United Kingdom \\
-3: Laboratoire d'\'{E}cologie V\'{e}g\'{e}tale et Biogéochimie, CP 244, Universit\'{e} Libre de Bruxelles, Brussels, Belgium \\
+2: Environmental Change Institute, School of Geography and the Environment, University of Oxford, Oxford, OX1 3QY, United Kingdom \\
+3: Laboratoire d'\'{E}cologie V\'{e}g\'{e}tale et Biogéochimie, CP 244, Universit\'{e} Libre de Bruxelles, Brussels, B-1050, Belgium \\
 4: National Centre for Earth Observation (NCEO), University of Sheffield, Hicks Building, Hounsfield Road, Sheffield, S3 7RH, United Kingdom \\
 5: Herbarium of Lubango, ISCED Hu\'{i}la, Sarmento Rodrigues Str. No. 2, CP. 230, Lubango, Angola \\
-6: Department of Wildlife, Fish, and Environmental Studies, Swedish University of Agricultural Sciences, Ume\aa, Sweden \\
-7: Department of Zoology, Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, South Africa \\
-8: Copernicus Institute of Sustainable Development, Utrecht University, 3584CS Utrecht, the Netherlands \\
-9: Ecological Advice, Ezemvelo KZN Wildlife, Hluhluwe-iMfolozi Park, South Africa \\
-10: School of Life Sciences, University of KwaZulu-Natal, Pietermaritzburg, South Africa \\
-11: Biodiversity, Evolution and Ecology of Plants, Institute of Plant Sciences and Microbiology, University of Hamburg, Ohnhorststr. 18, 22609 Hamburg, Germany \\
+6: Department of Wildlife, Fish, and Environmental Studies, Swedish University of Agricultural Sciences, Ume\aa, SE-907 36, Sweden \\
+7: Department of Zoology, Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, 6031, South Africa \\
+8: Copernicus Institute of Sustainable Development, Utrecht University, Utrecht, 3584CS, The Netherlands \\
+9: Ecological Advice, Ezemvelo KZN Wildlife, Hluhluwe-iMfolozi Park, 3202, South Africa \\
+10: School of Life Sciences, University of KwaZulu-Natal, Pietermaritzburg, 3209, South Africa \\
+11: Biodiversity, Evolution and Ecology of Plants, Institute of Plant Sciences and Microbiology, University of Hamburg, Ohnhorststr. 18, Hamburg, 22609, Germany \\
 12: Odum School of Ecology, University of Georgia, 140 E. Green St., Athens, GA 30602, USA \\
@@
 14: Forest Research Centre, 1 Orange Groove Highlands, Harare, Zimbabwe \\
-15: Faculty of Natural Resources and Spatial Sciences, Namibia University of Science and Technology, Windhoek, Namibia \\
+15: Faculty of Natural Resources and Spatial Sciences, Namibia University of Science and Technology, Windhoek, 10005, Namibia \\
 16: Department of Forest Engineering, Faculty of Agronomy and Forest Engineering, Universidade Eduardo Mondlane, Av. Julius Nyerere, 3453, Campus Universitario, Maputo, Mozambique \\
-17: Forestry Department Headquarters - Ministry of Lands and Natural Resources, Cairo Road, Lusaka, Zambia \\
+17: Ministry of Lands and Natural Resources, Cairo Road, Lusaka, Zambia \\
 18: School of Natural Resources, Copperbelt University, Kitwe, Zambia \\
 19: Faculty of Biological Sciences, University of Leeds, LS2 9JT, United Kingdom\\
-20: Copernicus Institute of Sustainable Development, Utrecht University, 3508 TC Utrecht, The Netherlands \\
+20: Copernicus Institute of Sustainable Development, Utrecht University, Utrecht, 3508, The Netherlands \\
 21: Centre for African Conservation Ecology, Nelson Mandela University, Port Elizabeth, 6031, South Africa \\
@@
 
-\newpage{}
+\newpage
+\section*{ORCiD}
 
-\textbf{Author for correspondence:}
+John L. Godlee - 0000-0001-5595-255X \\
+Casey M. Ryan - 0000-0002-1802-0128 \\
+David Bauman - 0000-0001-9115-6518 \\
+Samuel J. Bowers - 0000-0001-5418-0359 \\
+Joao M. B. Carreiras - 0000-0003-2737-9420 \\
+Antonio Valter Chisingui - 0000-0002-5137-1448 \\
+Joris P. G. M. Cromsigt - 0000-0002-8632-9469 \\
+Dave J. Druce - 0000-0003-2976-6029 \\
+Manfred Finckh - 0000-0003-2186-0854 \\
+Francisco Maiato Gon\c{c}alves - 0000-0002-8859-7491 \\
+Ricardo M. Holdo - 0000-0002-2484-0587 \\
+Steve Makungwa -  \\
+Iain M. McNicol - 0000-0003-3073-5668 \\
+Edward T. A. Mitchard - 0000-0002-5690-4055 \\
+Anderson Muchawona -  \\
+Rasmus Revermann - 0000-0002-7044-768X \\
+Natasha Sofia Ribeiro - 0000-0002-5369-5905 \\
+Abel Siampale -  \\
+Stephen Syampungani -  \\
+Jos\'{e} Jo\~{a}o Tchamba - 0000-0002-3770-2482 \\
+Hemant G. Tripathi - 0000-0001-6045-4336 \\
+Johannes Wallenfang -  \\
+Mariska te Beest - 0000-0003-3673-4105 \\
+Mathew Williams - 0000-0001-6117-5208 \\
+Kyle G. Dexter - 0000-0001-9232-5221 \\
+
+\section*{Author for correspondence}
 
@@
 
-\vspace{1cm}
+\section*{Word count}
 
-\textbf{Word count:}
-
-\vspace{0.2cm}
-
-\begin{tabular}{r|r}
-Main body word count & 6384\\
-Summary & 193\\
-Introduction & 1490\\
-Materials and Methods & 1999\\
-Results & 726\\
-Discussion & 2216\\
-Acknowledgements & 122\\
-\end{tabular}
-
-\vspace{1cm}
-
-Number of figures: 7
-
-Number of tables: 2
-
-Colour figures: 1, 3, 5, 6
-
-Number of supporting information files: 5
+Main body word count: 6706 \\
+Summary: 198 \\
+Introduction: 1107 \\
+Materials and Methods: 2089 \\
+Results: 756 \\
+Discussion: 2226 \\
+Conclusion: 330 \\
+Acknowledgements: 122 \\
+Number of figures: 7 \\
+Number of tables: 2 \\
+Colour figures: 1, 3, 5, 6 \\
+Number of supporting information files: 5 \\
 

### 2021-07-09 scripts/anon.R

@@ -1,0 +1,20 @@
+# Create an anonymised dataset for hosting on data repository
+# John Godlee (johngodlee@gmail.com)
+# 2021-07-09
+
+# Packages
+library(dplyr)
+
+# Import data
+dat <- readRDS("../data/plot_data_fil_agg_norm_std.rds")
+
+dat_anon <- dat %>% 
+  dplyr::select(
+    -plot_cluster,
+    -longitude_of_centre,
+    -latitude_of_centre,
+    -ends_with("_std"),
+    -ends_with("_log")) %>%
+  mutate(plot_id = row_number())
+
+write.csv(dat_anon, "../data/plot_data_anon.csv", row.names = FALSE)

### 2021-07-19 scripts/anon.R
Pre-copy-editing

@@ -13,6 +13,14 @@ dat_anon <- dat %>%
     -plot_cluster,
     -longitude_of_centre,
     -latitude_of_centre,
+    -clust5, 
+    -clust7,
+    -temp_seas,
+    -starts_with("sd_"),
+    -starts_with("var_"),
+    -starts_with("mean_"),
+    -starts_with("max_"),
+    -ends_with("_ha"),
     -ends_with("_std"),
     -ends_with("_log")) %>%
   mutate(plot_id = row_number())

### 2021-07-19 manuscript/regional_befr_manuscript.tex
Pre-copy-editing
Sanitised .tex

@@
 % Images 
-\usepackage{graphics} % Insert images easily
-\usepackage{graphicx} % Extended image support
-\usepackage{float} % Graphics placement [H] [H!] arguments
+\usepackage{graphics} 
+\usepackage{graphicx} 
+\usepackage{float}
 
 \makeatletter
-	\g@addto@macro\@floatboxreset\centering % Automatically centre images (floats)
+	\g@addto@macro\@floatboxreset\centering 
 \makeatother
@@
 
-% In this study :::
 In this study, we make the first known estimation of the Biodiversity-Ecosystem Function Relationship (BEFR) in disturbance-driven wooded ecosystems (savannas and woodlands), using southern Africa as our study region.
@@
 
-% Situation - DONE
 The study used \nplots{} woodland monitoring plots from a larger pool of 5395 plots in the larger SEOSAW database \citep{SEOSAW2020}, located across 10 countries within southern Africa in the miombo ecoregion (\autoref{clust_map}, \citealp{White1987}).
@@
 
-% Plots chosen - DONE
 Plots were chosen from the SEOSAW database based on the quality and completeness of data collection, and plot configuration.
@@
 
-% AGB
 We considered only trees and shrubs in our calculations of Above-Ground woody Biomass (AGB), including woody species such as palms and cycads, which are functionally tree-like.
@@
 
-% Number of stems
 All stems >10 cm DBH were measured within each plot resulting in a total of \num[group-separator={,}]{\ntrees{}} stems with measurements.
@@
 
-% AGB Calculation
 AGB for each plot (t ha\textsuperscript{-1}) was calculated as the sum of the AGB of each stem >5 cm DBH ($AGB_{i}$), divided by the plot area, using \autoref{chave_agb} taken from \citet{Chave2014}:
@@
 
-% WorldClim, SoilGrids and MODIS
 Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}.
@@
 
-% Correlations
 Pairwise correlations between all observed variables used in the Structural Equation Models (SEMs) showed that all tree species diversity (extrapolated tree species richness, Shannon equitability index) and structural diversity (coefficients of variation of DBH and height) variables had moderate positive correlations with AGB (\autoref{corr_mat}, Figure S3).
@@
 
-An anonymised version of the data that support the findings of this study are available at this DOI: .
+An anonymised version of the data that support the findings of this study are available at this DOI: \url{https://doi.org/10.7488/ds/3095}.
 

### 2021-09-15 manuscript/regional_befr_manuscript.tex
Updated bioclim explan
Added published paper

@@
 Only plots with a stem density >50 trees ha\textsuperscript{-1} (>10 cm stem diameter) were used, to ensure all plots represented woodland rather than `grassy savanna', which is considered here a separate biome with very different species composition \citep{Parr2014}.
-\nzam{} plots within the SEOSAW database were arranged in clusters of four 20x50 m plots, with 20 m between plots.
+\nzam{} plots within the SEOSAW database were arranged in clusters of four 20$\times$50 m plots, with 20 m between plots.
 Plots within each spatial cluster were combined and treated as a single plot in analyses, resulting in \nzamcluster{} aggregate plots which were then subject to the plot filtering process described above.
@@
 
-Climatic data were taken from the WorldClim database, using the BioClim variables \citep{Fick2017}.
+Climatic data were taken from the WorldClim database \citep{Fick2017}, which uses BIOCLIM to derive bioclimatic variables \citep{Booth2014}.
 In addition to MAT and MAP, temperature stress was calculated as the mean diurnal temperature range (BIO2) and precipitation seasonality was calculated as the mean of the coefficient of variation of monthly mean precipitation (BIO15).
@@
 Holding either species richness or abundance evenness constant while increasing the other will have the similar effect of reducing the likelihood that a neighbour is a conspecific, thus reducing the occurrence of negative density dependent competition effects and therefore increasing ecosystem function.
-We quantified tree structural diversity for each plot by calculating the Coefficient of Variation of DBH (DBH CoV) and tree height (Height CoV).
+We quantified tree structural diversity for each plot by calculating the Coefficient of Variation of DBH (DBH CV) and tree height (Height CV).
 
@@
 MAP had positive correlations with all tree species diversity and structural diversity variables.
-Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CoV: \ccdvi{}, Height CoV: \cchvi{}).
+Tree species diversity variables had clear positive correlations with stem density (species richness: \ccsi{}; Shannon equitability: \ccei{}), but structural diversity variables showed weak correlations with stem density (DBH CV: \ccdvi{}, Height CV: \cchvi{}).
 
